### 2019-08-26 content.tex
initial commit

\begin{quote}
	\textbf{Abstract:} 
\end{quote}

\section{Introduction}
Continuous technological advancements to high-throughput profiling of single cells
are having profound effects on how researchers can validate biological hypotheses. 
For example, single-cell RNA sequencing (scRNA-seq) directly resulted in the development
of a new type of computational method called trajectory inference (TI). By profiling
the transcriptomics profiles of developing cells, TI methods attempt to reconstruct 
and characterise the underlying dynamic processes \cite{cannoodt_computationalmethodstrajectory_2016}.
While early experimental technologies allowed to profile one single modality (e.g. DNA sequence, 
RNA or protein expression), recent developments permit profiling multiple modalities simultaneously.

An ideal experiment would be able to observe all aspects of a cell, including a full history of its 
molecular states, spatial positions and environmental interactions \cite{stuart_integrativesinglecellanalysis_2019}. 
While this falls outside the reach of current experimental technologies, \textit{in silico} simulations
of single cells would allow developing the next wave of computational techniques
in anticipation of new experimental technologies.

A few generators of scRNA-seq profiles have already been developed (e.g. splatter \cite{zappia_splattersimulationsinglecell_2017} and PROSSTT \cite{papadopoulos_prossttprobabilisticsimulation_2018}). % BoolODE and dyntoy are also examples.
These can be used to evaluate the performance of computational tools, and to explore their strengths and weaknesses. A limitation of directly simulating a scRNA-seq profile (instead of a single cell) is that extending the simulation to other aspects of the cell -- such as tracking the full history of molecular states -- becomes difficult.

We introduce dyngen, a multi-modality simulator of single cells (Figure \ref{fig:dyngen}).
dyngen was initially developed as part of a comprehensive benchmark of TI methods \cite{saelens_comparisonsinglecelltrajectory_2019} but has since been extended to be applicable in a much broader context.
We demonstrate its flexibility by simulating numerous different types of biological experiments, and using these simulations to develop new benchmarking techniques for computational tools.

\begin{figure}
	\centering
	\includegraphics[width=\linewidth]{fig/showcase_3.png} 
	\caption{Showcase of dyngen functionality. \textbf{TODO: change to pdf.} Remove B?} % TODO: update label
	\label{fig:dyngen}
\end{figure}

\section{Results}
dyngen is a simulator for single cells that develop over time. Throughout this section, a simple simulation of a cell undergoing a cyclic process is used to illustrate key strengths of dyngen (Figure \ref{fig:simplecyclic}). This example only comprises of a single cell containing 5 genes, but dyngen can easily scale up to thousands of simulations containing thousands of genes.

\begin{figure}
	\centering
	\includegraphics[width=\linewidth]{fig/simplecyclic_edited} 
	\caption{Showcase of dyngen functionality. A time resolution of 0.1 was used, but this can be increased or decreased without effect on performance of the execution of the simulation. \textbf{TODO: perhaps it's better to replace Figure \ref{fig:simplecyclic} with one subfigure for each of the paragraphs in this text.}}
	\label{fig:simplecyclic}
\end{figure}

In dyngen, a cell consists of a set of molecules, the abundance of which are affected by a set of reactions: transcription, splicing, translation, and degradation (Figure \ref{fig:simplecyclic}A). These reactions are determined from a predefined set of gene regulatory interactions (Figure \ref{fig:simplecyclic}B), henceforth referred to as a gene regulatory network (GRN). The likelihood of a reaction occurring at any given point in time is defined by the GRN and by the abundance of molecules involved each reaction.

One of dyngen's main advantages is that through careful engineering of the GRN, different cellular developmental processes can be obtained. Different GRNs can result in branching, converging, cyclic, or even disconnected developmental topologies. Multiple simulations with slightly different GRNs can emulate rewiring events in disease or perturbation experiments. % could use figure; one with GRNs of different topologies, another with rewiring events
Multiple simulations with different initial molecule abundance levels can be used to replicate batch effects.

Another advantage is that dyngen returns many modalities throughout the whole simulation: molecular abundance, cellular state, number of reaction firings, reaction likelihoods, and regulation activations (Figure \ref{fig:simplecyclic}C--F). These modalities can serve both as input data and ground truth for benchmarking many types of computational approaches. For example, a network inference method could use mRNA abundance and cellular states as inputs, and its output could be benchmarked against the gold standard GRN.

The final main advantage is that by making alterations to the simulation pipeline, multiple types of experiments (sampling technique or profiling technique) can be simulated. By default, dyngen supports snapshot experiments (uniformly sampling from an asynchronous dynamic process) and time-series experiments (sampling cells from different intervals in the simulation). 
It is possible to implement other experimental protocols (which perhaps do not exist in real life), such as sampling the same cell at regular intervals.

\section{Discussion}
As is, dyngen's single cell simulations can be used to evaluate common single-cell omics computational methods such as clustering, batch correction, trajectory inference and network inference.
However, the combined effect of these advantages results in a framework that is flexible enough to adapt to a broad range of applications; several examples are given.

Adding batch effects to snapshot simulations of linear (or even branching) trajectories allows to evaluate trajectory alignment methods -- which attempt to map two or more trajectories onto each other. Adding perturbations to the GRN allows to evaluate the performance of differential network inference methods -- which predict differential regulatory interactions between two or more groups of profiles.  Sampling a cell at a certain time point and once more at a later time point allows to evaluate the performance of RNA velocity approaches -- which predict the future state of a cell by looking at differences in pre-mRNA and mRNA abundance levels.

dyngen ultimately also allows anticipating technological developments in single-cell multi-omics. In this way, it is possible to design and evaluate the performance and robustness of new types of computational analyses before experimental data becomes available.
Similarly, it could also be used to compare which experimental technique will likely produce the most accurate result. For example, is it possible to infer directionality of regulatory interactions from snapshot experiments only, or are time series or knock down experiments a necessity in order to infer high quality regulatory networks?

Currently, dyngen focuses on simulating cells as standalone entities.
Future developments include extending the framework to simulate multiple cells in a virtual environment. Allowing cells to receive and react to environmental and intercellular stimuli would enable simulating essential cellular processes such as cell division and migration.

\section{Methods}

\subsection{Simulating a snapshot experiment with dyngen}
\subsubsection{Define the module network and the state network}
\subsubsection{Generate the transcription factor network}
\subsubsection{Generate targets and housekeeping genes}
\subsubsection{Convert gene regulatory network to a set of SSA reactions}
\subsubsection{Simulate backbone}
\subsubsection{Run SSA simulations}
\subsubsection{Map SSA simulations to backbone}
\subsubsection{Simulate snapshot experiment}

\subsection{Extensions}
\subsubsection{Predefined backbones}
\subsubsection{Backbone lego}
\subsubsection{Time series experiment}
\subsubsection{Perturbation experiment}
\subsubsection{Batch effects}

\subsection{Example use cases}
\subsubsection{Trajectory alignment}
From discussion: Adding batch effects to snapshot simulations of linear (or even branching) trajectories allows to evaluate trajectory alignment methods -- which attempt to map two or more trajectories onto each other. 
\subsubsection{Differential network inference}
From discussion: Adding perturbations to the GRN allows to evaluate the performance of differential network inference methods -- which predict differential regulatory interactions between two or more groups of profiles.
\subsubsection{RNA velocity}
From discussion: Sampling a cell at a certain time point and once more at a later time point allows to evaluate the performance of RNA velocity approaches -- which predict the future state of a cell by looking at differences in pre-mRNA and mRNA abundance levels.

%dyntoy and PROSSTT are specifically developed to generate datasets
% containing single cells which develop along a certain trajectory.
% These generators start a certain topology in mind, 
% simulate changing gene expressions along each of the branches of the topology, and sample cells
% from random points in the trajectory. splatter is focused on well replicating the specific noise characteristics of
% scRNA-seq data, and simulates differentially expressed genes in order to replicate clusters of cells, batch effects
% and even trajectories. dyngen 1.0 borrows a page from GeneNetWeaver in that it also converts a GRN into an ODE, but
% the GRN is constructed in such a way that the repeated simulations resemble single cells following a regulatory program
% (e.g. differentiation into one of two celltypes). BoolODE uses the dyngen 1.0 GRNs but uses boolean models to convert 
% the network into an ODE, and is used to evaluate the performance of NI methods, not TI methods.

%\begin{table}
%	\caption{\textbf{Comparison of existing single-cell omics simulators.} TI: trajectory inference, NI: network inference, Cl: clustering, DR: dimensionality reduction.} \label{tab:simulators}
%	\centering\fontsize{9}{11}\selectfont
%	\begin{tabularx}{\linewidth}{p{1.6cm}p{2.5cm}p{3cm}X}
%		Generator & Used to evaluate & Data type(s) & Approach \\ \hline
%		dyngen 1.0 & TI & mRNA \& protein & GRN, ODE, Euler--Maramuya \\
%		dyntoy & TI & mRNA & Visualise tree in plane, generate expression in plane, sample cells from tree \\
%		PROSSTT & TI & mRNA & Random walks along tree, sample cells from tree \\
%		splatter & Cl \& TI & mRNA & Simulate DE genes, simulate scRNA-seq noise \\
%		BoolODE & NI & mRNA \& protein & Use dyngen GRN's, ODE, Euler--Maramuya \\
%		dyngen 2.0 & TI, NI, Cl, DR & pre-mRNA, mRNA, protein & Backbone, GRN, SSA \\
%		dyngen 3? & TI, NI, Cl, DR & pre-mRNA, mRNA, miRNA, protein, protein complex & Backbone, GRN, SSA
%	\end{tabularx}
%\end{table}

### 2019-08-26 manuscript.tex
initial commit

\documentclass[11pt,hidelinks,a4paper]{article}

%%%%%%%%%%%%%%%%%%%%%%%
%%% IMPORT PACKAGES %%%
%%%%%%%%%%%%%%%%%%%%%%%

% text packages
\usepackage{csquotes}
\usepackage[english]{babel}
\usepackage{setspace}

% scripting packages
\usepackage{blindtext}

% style packages
\usepackage{lmodern}
\usepackage{fancyhdr}
\usepackage{microtype} % Sublim­i­nal re­fine­ments to­wards ty­po­graph­i­cal per­fec­tion

% used for defining abbreviations, see nomenclature.tex
\usepackage{nomencl}

% figure packages
\usepackage{tikz}
\usetikzlibrary{calc,shapes,arrows,intersections,shadows}
\tikzset{>=latex}
\usepackage{graphicx}
\usepackage[indention=0.5cm,labelsep=colon,font={sf,small},labelfont={bf,sf}]{caption}
\usepackage[indention=0.5cm,font={sf,small},labelfont={bf,sf}]{subcaption}
\usepackage{float}
\usepackage{multirow}
\usepackage{import}
\usepackage{xcolor}
\usepackage{colortbl}

% pseudocode packages
\usepackage{algorithm}
\usepackage{algpseudocode}

% symbols packages
\usepackage{amssymb}
\usepackage{amsmath}

% table packages
\usepackage{tabularx}
\usepackage{tabulary}
\usepackage{booktabs}

% references packages
\usepackage[sorting=none]{biblatex}
\bibliography{library}
\usepackage{bibentry}
\nobibliography*

% link packages
\usepackage{url}
\usepackage{doi}

% add checkmark commands
\usepackage{pifont}
\newcommand{\cmark}{\ding{51}}
\newcommand{\xmark}{\ding{55}}

% fix named refs for starred sections
\makeatletter
\def\ttl@useclass#1#2{%
	\@ifstar
	{\ttl@labeltrue\@dblarg{#1{#2}}}% {\ttl@labelfalse#1{#2}[]}%
	{\ttl@labeltrue\@dblarg{#1{#2}}}}
\makeatother

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%%% DEFINE FIGURE HELPERS %%%
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%

\newcommand{\hugefigure}{\textwidth}
\newcommand{\LARGEfigure}{0.8\textwidth} % 1 per col
\newcommand{\LARgefigure}{0.6\textwidth} % 1 per col
\newcommand{\Largefigure}{0.5\textwidth} % 1 per col
\newcommand{\largefigure}{0.37\textwidth} % 2 per col
\newcommand{\normalfigure}{0.3\textwidth} % 3 per col
\newcommand{\smallfigure}{0.24\textwidth} % 3 per col
\newcommand{\Smallfigure}{0.19\textwidth} % 4 per col
\newcommand{\SMALLfigure}{0.15\textwidth} % 5 per col

%%%%%%%%%%%%%%%%%%%%%
%%% DEFINE STYLES %%%
%%%%%%%%%%%%%%%%%%%%%

% use Sonny chapter style
\usepackage[Sonny]{fncychap} %Sonny

% set font after chapter sonny style
\usepackage{fontspec}
\setmainfont [Path = fonts/,
UprightFont = *-300,
ItalicFont = *-300-Italic,
BoldFont = *-700,
BoldItalicFont = *-700-Italic
]{MuseoSans}

%% hyperlinks
% load hyperref after fncychap
\usepackage{hyperref}

\renewcommand{\appendixname}{}
\renewcommand{\baselinestretch}{1.25}

\begin{document}

\include{content}

\end{document}

### 2019-08-26 analysis/multiple_backbones.R

@@ -1,0 +1,92 @@
+library(tidyverse)
+library(dyngen)
+
+backs <- list_backbones()
+
+for (nb in names(backs)) {
+  cat("=============== SIMULATING ", nb, " ===============\n", sep = "")
+  set.seed(1)
+  back <- backs[[nb]]()
+  model <-
+    initialise_model(
+      num_tfs = nrow(back$module_info) * 1.5,
+      num_targets = 30,
+      num_hks = 15,
+      num_cells = 300,
+      backbone = back,
+      verbose = TRUE,
+      download_cache_dir = "~/.cache/dyngen",
+      num_cores = 8,
+      simulation_params = simulation_default(ssa_algorithm = GillespieSSA2::ssa_exact(), num_simulations = 16, census_interval = .01)
+    )
+
+  out <- generate_dataset(model, make_plots = FALSE)
+
+  # recalc dimred
+  gs_counts <- out$model$gold_standard$counts %>% as.matrix
+  sim_counts <- out$model$simulations$counts[,colnames(gs_counts)] %>% as.matrix
+  counts <- rbind(gs_counts, sim_counts)
+  space <- dyndimred::dimred_landmark_mds(counts)
+  gs_ix <- seq_len(nrow(gs_counts))
+  out$model$gold_standard$dimred <- space[gs_ix,]
+  out$model$simulations$dimred <- space[-gs_ix,]
+
+  # plot_gold_mappings(out$model, do_facet = FALSE)
+
+  # save output
+  out_dir <- paste0("dyngen/bb_", nb, "/")
+  if (!dir.exists(out_dir)) dir.create(out_dir)
+  write_rds(out, paste0(out_dir, "/out.rds"), compress = "gz")
+
+  # plot model
+  model <- out$model
+
+  g <- plot_backbone_statenet(model)
+  ggsave(paste0(out_dir, "statenet.pdf"), g, width = 6, height = 6)
+  g <- plot_backbone_modulenet(model)
+  ggsave(paste0(out_dir, "/modulenet.pdf"), g, width = 8, height = 6)
+
+  g <- plot_feature_network(model, show_hks = TRUE)
+  ggsave(paste0(out_dir, "/featnet.pdf"), g, width = 8, height = 6)
+  g <- plot_feature_network(model, show_targets = FALSE)
+  ggsave(paste0(out_dir, "/featnet_onlytfs.pdf"), g, width = 8, height = 6)
+
[…]

### 2019-08-26 analysis/script.R

@@ -1,0 +1,133 @@
+library(tidyverse)
+library(dyngen)
+
+set.seed(10)
+
+backbone <- bblego(
+  bblego_start("A", type = "simple", num_modules = 1),
+  bblego_linear("A", "B", type = "simple", num_modules = 3),
+  bblego_branching("B", c("C", "D")),
+  bblego_end("C", type = "simple", num_modules = 2),
+  bblego_end("D", type = "simple", num_modules = 2)
+)
+model <-
+  initialise_model(
+    num_tfs = 12,
+    num_targets = 30,
+    num_hks = 15,
+    num_cells = 150,
+    backbone = backbone,
+    verbose = TRUE,
+    download_cache_dir = "~/.cache/dyngen",
+    num_cores = 8,
+    distance_metric = "euclidean",
+    simulation_params = simulation_default(ssa_algorithm = GillespieSSA2::ssa_exact(), num_simulations = 8, census_interval = .01)
+  ) %>%
+  generate_tf_network() %>%
+  generate_feature_network() %>%
+  generate_kinetics()
+
+## grn
+plot_backbone_statenet(model) %>% ggsave(filename = "dyngen/statenet.pdf", width = 6, height = 6)
+plot_backbone_modulenet(model) %>% ggsave(filename = "dyngen/modulenet.pdf", width = 8, height = 6)
+
+plot_feature_network(model, show_targets = FALSE)
+
+plot_feature_network(model)
+plot_feature_network(model, show_hks = TRUE)
+plot_feature_network(model)
+plot_feature_network(model, show_hks = TRUE) %>% ggsave(filename = "dyngen/featnet.pdf", width = 8, height = 6)
+plot_feature_network(model, show_targets = FALSE) %>% ggsave(filename = "dyngen/featnet_onlytfs.pdf", width = 8, height = 6)
+
+## simulate
+model2 <- model %>%
+  generate_gold_standard() %>%
+  generate_cells() %>%
+  generate_experiment()
+
+# recalculate dimred
+space <- prcomp(model2$simulations$counts %>% as.matrix, rank. = 2)$x
+colnames(space) <- c("comp_1", "comp_2")
+model2$simulations$dimred <- space
+
+write_rds(model2, "dyngen/model.rds", compress = "gzip")
+
[…]

### 2019-08-26 analysis/simplecyclic.R

@@ -1,0 +1,253 @@
+library(tidyverse)
+library(dyngen)
+
+set.seed(1)
+
+backbone <- backbone(
+  module_info = tribble(
+    ~module_id, ~a0, ~burn,
+    "A", 1, TRUE,
+    "B", 1, TRUE,
+    "C", 1, TRUE,
+    "D", 0, TRUE,
+    "E", 0, TRUE
+  ),
+  module_network = tribble(
+    ~from, ~to, ~effect, ~strength, ~cooperativity,
+    "A", "B", -1, 4, 2,
+    "B", "C", -1, 4, 2,
+    "C", "D", 1, 1, 2,
+    "D", "E", 1, 1, 2,
+    "E", "A", -1, 4, 2
+  ),
+  expression_patterns = tribble(
+    ~from, ~to, ~module_progression, ~start, ~burn, ~time,
+    "Burn0", "S1", "+A,+B,+C", TRUE, TRUE, 3,
+    "S1", "S2", "+D,+E,-C", FALSE, FALSE, 4,
+    "S2", "S3", "+C,-A,-B", FALSE, FALSE, 4,
+    "S3", "S1", "+A,+B,-D,-E", FALSE, FALSE, 4
+  )
+)
+census_interval <- .1
+total_time <- 10
+model <-
+  initialise_model(
+    num_tfs = nrow(backbone$module_info) * 1,
+    num_targets = 0,
+    num_hks = 0,
+    num_cells = 300,
+    backbone = backbone,
+    verbose = TRUE,
+    download_cache_dir = "~/.cache/dyngen",
+    num_cores = 8,
+    kinetics_params = kinetics_default(
+      sample_wpr = function(n) 100,
+      sample_spl = function(n) 10,
+      sample_xdr = function(n) 5,
+      sample_ypr = function(n) 5,
+      sample_ydr = function(n) 5
+    ),
+    simulation_params = simulation_default(
+      burn_time = 2,
+      total_time = total_time,
+      ssa_algorithm = GillespieSSA2::ssa_exact(),
+      num_simulations = 1,
+      census_interval = census_interval,
+      store_reaction_firings = TRUE,
+      store_reaction_propensities = TRUE
+    )
+  )
+
+model <- model %>%
+  generate_tf_network() %>%
+  generate_feature_network()
+
+notf <- function(x) gsub("_TF1$", "", x)
+model$feature_info <- model$feature_info %>% mutate_at(vars(feature_id), notf)
[…]

### 2019-08-28 content.tex
update
Update content.tex

@@
 
-\begin{figure}
+\begin{figure}[hb]
 	\centering
@@
 As is, dyngen's single cell simulations can be used to evaluate common single-cell omics computational methods such as clustering, batch correction, trajectory inference and network inference.
-However, the combined effect of these advantages results in a framework that is flexible enough to adapt to a broad range of applications; several examples are given.
+However, the combined effect of these advantages results in a framework that is flexible enough to adapt to a broad range of applications, some of which may still be needed in the future.
+This may include methods that integrate clustering, network inference and trajectory inference.
+In this respect, dyngen may promote the development of new tools in the single-cell field similarly as other simulators have done in the past \textbf{10.1093/bioinformatics/btx631} \textbf{10.1093/bioinformatics/btr373} \textbf{10.1038/nmeth.3407}.
 
@@
 Allowing cells to receive and react to environmental and intercellular stimuli would enable simulating essential cellular processes such as cell division and migration.
-
+%¡also adding more types of molecules, e.g. protein complex, small rnas, PPI
 \section{Methods}

### 2019-08-28 content.tex
add content

--- 2019-08-26 content.tex
@@
 A few generators of scRNA-seq profiles have already been developed (e.g.
-splatter \cite{zappia_splattersimulationsinglecell_2017} and PROSSTT \cite{papadopoulos_prossttprobabilisticsimulation_2018}).
+splatter \cite{zappia_splattersimulationsinglecell_2017}, powsimR \cite{vieth_powsimrpoweranalysis_2017} and PROSSTT \cite{papadopoulos_prossttprobabilisticsimulation_2018}).
 % BoolODE and dyntoy are also examples.
@@
 
+%% TODO: show that dyngen output resembles real data using e.g. countsimQC?
+
 \section{Discussion}
 As is, dyngen's single cell simulations can be used to evaluate common single-cell omics computational methods such as clustering, batch correction, trajectory inference and network inference.
-However, the combined effect of these advantages results in a framework that is flexible enough to adapt to a broad range of applications; several examples are given.
+However, the combined effect of these advantages results in a framework that is flexible enough to adapt to a broad range of applications.
+This may include methods that integrate clustering, network inference and trajectory inference.
+In this respect, dyngen may promote the development of new tools in the single-cell field similarly as other simulators have done in the past \cite{schaffter_genenetweaversilicobenchmark_2011,ewing_combiningtumorgenome_2015}.
 
-Adding batch effects to snapshot simulations of linear (or even branching) trajectories allows to evaluate trajectory alignment methods -- which attempt to map two or more trajectories onto each other.
-Adding perturbations to the GRN allows to evaluate the performance of differential network inference methods -- which predict differential regulatory interactions between two or more groups of profiles.
-Sampling a cell at a certain time point and once more at a later time point allows to evaluate the performance of RNA velocity approaches -- which predict the future state of a cell by looking at differences in pre-mRNA and mRNA abundance levels.
+Adding batch effects to snapshot simulations of linear (or even branching) trajectories allows evaluating trajectory alignment methods -- which attempt to map two or more trajectories onto each other.
+Adding perturbations to the GRN allows evaluating the performance of differential network inference methods -- which predict differential regulatory interactions between two or more groups of profiles.
+Sampling a cell at a particular time point and once more at a later time point allows evaluating the performance of RNA velocity approaches -- which predict the future state of a cell by looking at differences in pre-mRNA and mRNA abundance levels.
 
@@
 Similarly, it could also be used to compare which experimental technique will likely produce the most accurate result.
-For example, is it possible to infer directionality of regulatory interactions from snapshot experiments only, or are time series or knock down experiments a necessity in order to infer high quality regulatory networks?
+For example, is it possible to infer directionality of regulatory interactions from snapshot experiments only, or are time series or knockdown experiments a necessity in order to infer high-quality regulatory networks?
 
@@
 Allowing cells to receive and react to environmental and intercellular stimuli would enable simulating essential cellular processes such as cell division and migration.
-
+%¡also adding more types of molecules, e.g. protein complex, small rnas, PPI
 \section{Methods}
 
-\subsection{Simulating a snapshot experiment with dyngen}
-\subsubsection{Define the module network and the state network}
+dyngen's approach was inspired by GeneNetWeaver's \cite{schaffter_genenetweaversilicobenchmark_2011} workflow to generate \textit{in silico} bulk profiles for evaluating network inference methods \cite{marbach_wisdomcrowdsrobust_2012}.
+The workflow to generate \textit{in silico} single cell data consists of six main steps: 1.
+define the module network, 2.
+generate gene regulatory network, 3.
+convert regulatory interactions to biochemical reactions, 4.
+simulate the developmental backbone, 5.
+simulate single cells, and 6.
+simulate experiment.
+
+\subsection{Defining the module network and backbone}
+One of the main processes involved in cellular dynamic processes is gene regulation, where regulatory cascades and feedback loops lead to progressive changes in expression and decision making.
+The exact way a cell chooses a certain path during its differentiation is still an active research field, although certain models have already emerged and been tested \textit{in vivo}.
+One driver of bifurcation seems to be mutual antagonism, where two genes \cite{xu_regulationbifurcatingcell_2015} strongly repress each other, forcing one of the two to become inactive \cite{graf_forcingcellschange_2009}.
+Such mutual antagonism can be modelled and simulated \cite{wang_quantifyingwaddingtonlandscape_2011, ferrell_bistabilitybifurcationswaddington_2012}.
+Although the two-gene model is simple and elegant, the reality is frequently more complex, with multiple genes (grouped into modules) repressing each other \cite{yosef_dynamicregulatorynetwork_2013}.
+
+To simulate certain trajectory topologies, we therefore designed module networks in which the cells follow a particular trajectory topology given certain parameters.
+\subsubsection{Predefined backbones}
+\subsubsection{Backbone lego}
+
+\subsection{Generate gene regulatory network}
 \subsubsection{Generate the transcription factor network}
 \subsubsection{Generate targets and housekeeping genes}
-\subsubsection{Convert gene regulatory network to a set of SSA reactions}
-\subsubsection{Simulate backbone}
-\subsubsection{Run SSA simulations}
+
+\subsection{Convert gene regulatory network to a set of reactions}
+
+\subsection{Simulate developmental backbone}
+
+\subsection{Simulate single cells}
 \subsubsection{Map SSA simulations to backbone}
-\subsubsection{Simulate snapshot experiment}
 
-\subsection{Extensions}
-\subsubsection{Predefined backbones}
-\subsubsection{Backbone lego}
+\subsection{Simulate experiment}
+\subsubsection{Snapshot experiment}
 \subsubsection{Time series experiment}
 \subsubsection{Perturbation experiment}
-\subsubsection{Batch effects}
 
@@
 \subsubsection{Trajectory alignment}
-From discussion: Adding batch effects to snapshot simulations of linear (or even branching) trajectories allows to evaluate trajectory alignment methods -- which attempt to map two or more trajectories onto each other.
+From discussion: Adding batch effects to snapshot simulations of linear (or even branching) trajectories allows evaluating trajectory alignment methods -- which attempt to map two or more trajectories onto each other.
 \subsubsection{Differential network inference}
-From discussion: Adding perturbations to the GRN allows to evaluate the performance of differential network inference methods -- which predict differential regulatory interactions between two or more groups of profiles.
+From discussion: Adding perturbations to the GRN allows evaluating the performance of differential network inference methods -- which predict differential regulatory interactions between two or more groups of profiles.
 \subsubsection{RNA velocity}
-From discussion: Sampling a cell at a certain time point and once more at a later time point allows to evaluate the performance of RNA velocity approaches -- which predict the future state of a cell by looking at differences in pre-mRNA and mRNA abundance levels.
+From discussion: Sampling a cell at a certain time point and once more at a later time point allows evaluating the performance of RNA velocity approaches -- which predict the future state of a cell by looking at differences in pre-mRNA and mRNA abundance levels.
 

### 2019-08-28 content.tex
update

@@
 
-\begin{figure}[hb]
+\begin{figure}[htb!]
 	\centering
@@
 dyngen's approach was inspired by GeneNetWeaver's \cite{schaffter_genenetweaversilicobenchmark_2011} workflow to generate \textit{in silico} bulk profiles for evaluating network inference methods \cite{marbach_wisdomcrowdsrobust_2012}.
-The workflow to generate \textit{in silico} single cell data consists of six main steps: 1.
-define the module network, 2.
-generate gene regulatory network, 3.
-convert regulatory interactions to biochemical reactions, 4.
-simulate the developmental backbone, 5.
-simulate single cells, and 6.
-simulate experiment.
+The workflow to generate \textit{in silico} single cell data consists of six main steps (Figure \ref{fig:explain_methods}).
+
+\begin{figure}[htb!]
+	\centering
+	\includegraphics[width=\hugefigure]{fig/explain_methods} 
+	\caption{The workflow of dyngen.}
+	\label{fig:explain_methods}
+\end{figure}
 
@@
 
-To simulate certain trajectory topologies, we therefore designed module networks in which the cells follow a particular trajectory topology given certain parameters.
+In dyngen, the behaviour of the dynamic process is determined in the way that modules are regulating each other, where each module is a set of coregulated genes.
+By carefully designing the module network, the desired dynamic behaviour can be achieved.
+A module may have basal expression, which means that pre-mRNA of the genes in this module will be transcribed without the presence of transcription factor molecules.
+Basal expression
+Burn phase
+
+Several examples of module networks are given (Figure \ref{fig:example_backbones}).
+A linear process can by achieved by simply chaining together several modules.
+A cyclic process is a chain of modules, where the last module represses the first module.
+A bifurcating process consists of two modules that repress each other, and some modules leading up to and resulting from the bifurcation.
+A converging
+Note that these examples represent the bare minimum in terms of number of modules used. Using more modules is typically desired, since some transitions will otherwise occur too quickly.
+
+\begin{figure}[htb!]
+	\centering
+	\includegraphics[width=\LARgefigure]{fig/example_backbones} 
+	\caption{Example module networks}
+	\label{fig:example_backbones}
+\end{figure}
+
+\subsubsection{Backbone lego}
+\blindtext
+
 \subsubsection{Predefined backbones}
-\subsubsection{Backbone lego}
+\blindtext
 
 \subsection{Generate gene regulatory network}
+\blindtext
 \subsubsection{Generate the transcription factor network}
+\blindtext
 \subsubsection{Generate targets and housekeeping genes}
+\blindtext
 
 \subsection{Convert gene regulatory network to a set of reactions}
+\blindtext
 
 \subsection{Simulate developmental backbone}
+\blindtext
 
 \subsection{Simulate single cells}
+\blindtext
+
 \subsubsection{Map SSA simulations to backbone}
+\blindtext
 
 \subsection{Simulate experiment}
+\blindtext
+
 \subsubsection{Snapshot experiment}
+\blindtext
+
 \subsubsection{Time series experiment}
+\blindtext
+
 \subsubsection{Perturbation experiment}
+\blindtext
 

### 2019-08-28 analysis/example_methods_overview.R

@@ -1,0 +1,146 @@
+library(tidyverse)
+library(dyngen)
+
+set.seed(3)
+
+backbone <- backbone(
+  module_info = tribble(
+    ~module_id, ~a0, ~burn,
+    "A", 1, TRUE,
+    "B", 0, FALSE,
+    "C", 0, FALSE,
+    "D", 0, FALSE,
+  ),
+  module_network = tribble(
+    ~from, ~to, ~effect, ~strength, ~cooperativity,
+    "A", "B", 1, 1, 2,
+    "B", "C", 1, 1, 2,
+    "B", "D", 1, 1, 2,
+    "C", "D", -1, 10, 2,
+    "D", "C", -1, 10, 2
+  ),
+  expression_patterns = tribble(
+    ~from, ~to, ~module_progression, ~start, ~burn, ~time,
+    "Burn0", "S1", "+A", TRUE, TRUE, 2,
+    "S1", "S2", "+B", FALSE, FALSE, 2,
+    "S2", "S3", "+C", FALSE, FALSE, 2,
+    "S2", "S4", "+D", FALSE, FALSE, 2,
+  )
+)
+model <-
+  initialise_model(
+    num_tfs = nrow(backbone$module_info),
+    num_targets = 10,
+    num_hks = 10,
+    num_cells = 0,
+    backbone = backbone,
+    verbose = TRUE,
+    download_cache_dir = "~/.cache/dyngen",
+    num_cores = 8,
+    distance_metric = "euclidean",
+    simulation_params = simulation_default(ssa_algorithm = GillespieSSA2::ssa_exact(), num_simulations = 8, census_interval = .01)
+  ) %>%
+  generate_tf_network() %>%
+  generate_feature_network() %>%
+  generate_kinetics()
+
+## grn
+plot_backbone_statenet(model) #%>% ggsave(filename = "dyngen/statenet.pdf", width = 6, height = 6)
+plot_backbone_modulenet(model) #%>% ggsave(filename = "dyngen/modulenet.pdf", width = 8, height = 6)
+
+plot_feature_network(model, show_targets = FALSE)
+
+plot_feature_network(model)
+plot_feature_network(model, show_hks = TRUE)
+plot_feature_network(model)
+plot_feature_network(model, show_hks = TRUE) #%>% ggsave(filename = "dyngen/featnet.pdf", width = 8, height = 6)
[…]

### 2019-08-29 content.tex
update

@@
 	\includegraphics[width=\hugefigure]{fig/explain_methods} 
-	\caption{The workflow of dyngen.}
+	\caption{\textbf{The workflow of dyngen is comprised of six main steps.} \textbf{A:} The user needs to specify the desired module network or use a predefined module network.
+\textbf{B:} Each gene in a module is is regulated by one or more transcription factors from the upstream module.
+Additional target genes are generated.
+\textbf{C:} Each gene regulatory interaction in the GRN is converted to a set of biochemical reactions.
+\textbf{D:} Along with the module network, the user also needs to specify the backbone structure of expected cell states.
+The average expression of each edge in the backbone is simulated by activating a restricted set of genes for each edge.
+\textbf{E:} Multiple Gillespie SSA simulations are run using the reactions defined in step C.
+The counts of each of the mocules at each time step are extracted.
+Each time step is mapped to a point in the backbone.
+\textbf{F:} Multiple cells are sampled from each simulation.
+Molecules are sampled from each cell.}
 	\label{fig:explain_methods}
@@
 \subsection{Defining the module network and backbone}
+
 One of the main processes involved in cellular dynamic processes is gene regulation, where regulatory cascades and feedback loops lead to progressive changes in expression and decision making.
@@
 
-In dyngen, the behaviour of the dynamic process is determined in the way that modules are regulating each other, where each module is a set of coregulated genes.
-By carefully designing the module network, the desired dynamic behaviour can be achieved.
+In dyngen, the user defines the behaviour of the simulation by defining how sets of genes, called modules, are regulating eachother.
 A module may have basal expression, which means that pre-mRNA of the genes in this module will be transcribed without the presence of transcription factor molecules.
-Basal expression
-Burn phase
+A module marked as "active during the burn phase" means that this module will be allowed to generate expression of its genes during an initial warm-up phase (See section \ref{sec:dyngen_ssa}).
+At the end of the dyngen process, cells will not be sampled from the burn phase simulations.
 
 Several examples of module networks are given (Figure \ref{fig:example_backbones}). 
-A linear process can by achieved by simply chaining together several modules.
-A cyclic process is a chain of modules, where the last module represses the first module.
-A bifurcating process consists of two modules that repress each other, and some modules leading up to and resulting from the bifurcation.
-A converging
-Note that these examples represent the bare minimum in terms of number of modules used. Using more modules is typically desired, since some transitions will otherwise occur too quickly.
+A simple chain of modules (where one module upregulates the next) results in a \emph{linear} process.
+By having the last module repress the first module, the process becomes \emph{cyclic}.
+Two modules repressing eachother is the basis of a \emph{bifurcating} process, though several chains of modules have to be attached in order to achieve progression before and after the bifurcation process.
+Finally, a \emph{converging} process has a bifurcation occurring during the burn phase, after which any differences in module regulation is removed.
+
+Note that these examples represent the bare minimum in terms of number of modules used.
+Using longer chains of modules is typically desired.
+In addition, the fate decisions made in this example of a bifurcation is reversible, meaning cells can be reprogrammed to go down a different differentiation path.
+If this effect is undesirable, more safeguards need to be put in place to prevent reprogramming from occurring (Section \ref{sec:dyngen_backbone_lego}).
 
@@
 	\centering
-	\includegraphics[width=\LARgefigure]{fig/example_backbones} 
+	\includegraphics[width=\LARGEfigure]{fig/example_backbones} 
 	\caption{Example module networks}
@@
 
-\subsubsection{Backbone lego}
-\blindtext
+In addition to the module network, the user also needs to define a network of cellular states called the "backbone".
+Before simulating any cells, each transition in the backbone is simulated separately to obtain the average changes in expression along that transition (Figure \ref{fig:explain_methods}D).
+As part of the backbone, the user needs to specify which modules are allowed to alter its expression from one state to another.
+For example, in order to transition from state S0 to S1 in the cyclic example, gene modules A, B and C are turned on and a simulation is allowed to run.
+To transition from S1 to S2, gene modules D and E are turned on, and expression of gene module C is kept constant.
+To transition from S2 to S3, C is turned on again and now A and B are fixed.
+Finally, to transition from S3 to S1 again, A and B are turned on again and D and E are fixed again.
+Demonstrations of the backbone will be explained in more detail in section \ref{sec:dyngen_sim_backbone}.
+
+\subsubsection{Backbone lego} \label{sec:dyngen_backbone_lego}
+
+\paragraph{Simple linear}
+
+\paragraph{Linear flip flop}
+
+\paragraph{Linear double repressing}
+
+\paragraph{Linear quick double repressing}
+
+\paragraph{Bifurcation (reversible)}
+
+\paragraph{Bifurcation (irreversible)}
 
 \subsubsection{Predefined backbones}
-\blindtext
+
+\paragraph{Linear}
+\paragraph{Bifurcating}
+\paragraph{Cycle}
+\paragraph{Branching} (and binary tree, consecutive bifurcating, trifurcating)
+\paragraph{Converging}
+\paragraph{Bifurcating converging}
+\paragraph{Bifurcating cycle}
+\paragraph{Bifurcating loop}
+\paragraph{Disconnected}
 
@@
 
-\subsection{Simulate developmental backbone}
+\subsection{Simulate developmental backbone} \label{sec:dyngen_sim_backbone}
 \blindtext
 
-\subsection{Simulate single cells}
+\subsection{Simulate single cells} \label{sec:dyngen_ssa}
 \blindtext

### 2019-08-29 analysis/example_methods_overview.R

@@ -1,7 +1,9 @@
 library(tidyverse)
 library(dyngen)
 
-set.seed(3)
+set.seed(4)
+
+dir.create("fig/example_methods", showWarnings = FALSE, recursive = TRUE)
 
 backbone <- backbone(
   module_info = tribble(
@@ -21,23 +23,25 @@ backbone <- backbone(
   ),
   expression_patterns = tribble(
     ~from, ~to, ~module_progression, ~start, ~burn, ~time,
-    "Burn0", "S1", "+A", TRUE, TRUE, 2,
-    "S1", "S2", "+B", FALSE, FALSE, 2,
-    "S2", "S3", "+C", FALSE, FALSE, 2,
-    "S2", "S4", "+D", FALSE, FALSE, 2,
+    "Burn0", "S1", "+A", TRUE, TRUE, .5,
+    "S1", "S2", "+B", FALSE, FALSE, .5,
+    "S2", "S3", "+C", FALSE, FALSE, 1,
+    "S2", "S4", "+D", FALSE, FALSE, 1,
   )
 )
+backbone$module_info$color <- c("#e41a1c", "#377eb8", "#4daf4a", "#984ea3")
 model <-
   initialise_model(
-    num_tfs = nrow(backbone$module_info),
+    num_tfs = 8,
     num_targets = 10,
-    num_hks = 10,
-    num_cells = 0,
+    num_hks = 0,
+    num_cells = 100,
     backbone = backbone,
     verbose = TRUE,
     download_cache_dir = "~/.cache/dyngen",
     num_cores = 8,
     distance_metric = "euclidean",
+    tf_network_params = tf_network_default(min_tfs_per_module = 2, sample_num_regulators = function() 1),
     simulation_params = simulation_default(ssa_algorithm = GillespieSSA2::ssa_exact(), num_simulations = 8, census_interval = .01)
   ) %>%
   generate_tf_network() %>%
@@ -45,16 +49,16 @@ model <-
   generate_kinetics()
 
 ## grn
-plot_backbone_statenet(model) #%>% ggsave(filename = "dyngen/statenet.pdf", width = 6, height = 6)
-plot_backbone_modulenet(model) #%>% ggsave(filename = "dyngen/modulenet.pdf", width = 8, height = 6)
+plot_backbone_statenet(model) %>% ggsave(filename = "fig/example_methods/statenet.pdf", width = 6, height = 6)
[…]

### 2019-08-30 content.tex
update

@@
 %% TODO: show that dyngen output resembles real data using e.g. countsimQC?
+
+%% TODO: add result figures pertaining use cases
 
@@
 \subsubsection{Backbone lego} \label{sec:dyngen_backbone_lego}
+The backbone can make use of one or more "backbone lego" (BBL) pieces (Figure \ref{fig:backbone_lego}).
+A BBL consists of one or more modules which regulate each other such that the output modules present a specific behaviour, depending on the input module (Figure \ref{fig:backbone_lego}A).
+Parameters allow determining the number of modules involved in the process and the number of outputs.
+Multiple BBLs can be chained together in order to intuitively create milestone networks and corresponding state networks (Figure \ref{fig:backbone_lego}B).
+Note that not all dynamic processes can be represented by a combination of BBLs, but they can serve as common building blocks to aid the construction of the backbone.
 
-\paragraph{Simple linear}
+\begin{figure}[htb!]
+	\centering
+	\includegraphics[width=\LARGEfigure]{fig/backbone_lego} 
+	\caption{Backbone lego}
+	\label{fig:backbone_lego}
+\end{figure}
 
-\paragraph{Linear flip flop}
+When the input node of a \textbf{linear BBL} (Figure \ref{fig:backbone_lego}C) is upregulated, the module the BBL is connected to will be upregulated.
+A \emph{simple chain} is a set of modules where a module upregulates the next.
+A \emph{chain with double repression} has an uneven number of modules forming a chain where each module downregulates the next but all modules (except the input) have basal expression.
+A \emph{grid with double repression} is similar; except that modules do not have basal expression but instead get upregulated by an upstream module in the chain.
+Finally, a \emph{flip flop} consists of a simple chain where first the modules (except the last) are upregulated.
+Once the second to last module is upregulated, that module upregulates itself and the first module is strongly repressed, causing all other modules to lose expression and finally the last module to be upregulated.
+The \emph{flip flop} retains this output state, even when the input changes.
 
-\paragraph{Linear double repressing}
+When the input node of a \textbf{branching BBL} (Figure \ref{fig:backbone_lego}D) is upregulated, a subset of its output modules will eventually be upregulated.
+A \emph{simple branching} uses reciprocal inhibition to drive the upregulation of one of the output modules.
+Due to its simplicity, however, multiple output modules might be upregulated simultaneously, and over long periods of simulation time it might be possible that the choice of upregulated module changes.
+A \emph{robust branching} improves upon the simple branching by preventing upregulation of output modules until an internal branching decision has been made, and by repressing the decision mechanism to avoid other output modules being upregulated other than the one that has been chosen.
 
-\paragraph{Linear quick double repressing}
-
-\paragraph{Bifurcation (reversible)}
-
-\paragraph{Bifurcation (irreversible)}
-
-\subsubsection{Predefined backbones}
-
-\paragraph{Linear}
-\paragraph{Bifurcating}
-\paragraph{Cycle}
-\paragraph{Branching} (and binary tree, consecutive bifurcating, trifurcating)
-\paragraph{Converging}
-\paragraph{Bifurcating converging}
-\paragraph{Bifurcating cycle}
-\paragraph{Bifurcating loop}
-\paragraph{Disconnected}
+A \textbf{leaf BBL} (Figure \ref{fig:backbone_lego}E) is a linear BBL that has either no inputs or no outputs.
+A \emph{start} BBL is a linear BBL where the first module has basal expression, and all modules in this module will be active during the burn-in phase of the simulation (Section \ref{sec:dyngen_sim_backbone}).
+An \emph{end} BBL is also a linear BBL with its output regulating one final module.
 
 \subsection{Generate gene regulatory network}
-\blindtext
+% TODO: write section in words instead of bullet point list
+
 \subsubsection{Generate the transcription factor network}
-\blindtext
-\subsubsection{Generate targets and housekeeping genes}
-\blindtext
+Parameters: 
+\begin{itemize}
+	\item Number of TFs to generate $num\_tfs$
+	\item Minimum TFs per module $min\_tfs$, default = 1
+	\item Number of regulatory interactions per module interaction $num\_tf\_ints$, default = 2
+\end{itemize}
+
+Procedure: 
+\begin{itemize}
+	\item Divide $num\_tfs$ TFs amongst modules in backbone such that each module has at least $min\_tfs$ TFs.
+	\item For every TF in a particular module $M$, connect that TF with $num\_tf\_ints$ TFs for each upstream module of $M$.
+\end{itemize}
+
+\subsubsection{Generate targets}
+Parameters: 
+\begin{itemize}
+	\item Number of target genes $num\_targets$
+	\item Name of a FANTOM5 GRN
+	\item Damping factor $damping$, default = 0.05
+	\item Target resampling $resamp$, default = $+\infty$
+	\item Maximum in-degree $max\_in\_degree$, default = 5
+\end{itemize}
+
+Procedure:
+\begin{itemize}
+	\item Download the FANTOM5 GRN if not already available
+	\item Randomly map TFs to the regulators in the GRN
+	\item Compute page rank from the selected regulators in the GRN with damping factor $damping$
+	\item Perform weighted sample of  $\min(num\_targets, resamp)$ targets weighted by the page rank
+	\item Select subgraph induced by the TFs and the sampled targets in the GRN, remove edges to the TFs
+	\item Add subgraph to current TF-target network
+	\item If less than $num\_targets$ have been sampled in this way, go back to step 2.
+	\item Remove regulatory interactions if a target has more than $max\_in\_degree$ edges.
+\end{itemize}
+
+\subsubsection{Generate housekeeping genes}
+Parameters: 
+\begin{itemize}
+	\item Number of housekeeping genes $num\_targets$
+	\item Name of a FANTOM5 GRN
+	\item Target resampling $resamp$, default = $+\infty$
+	\item Maximum in-degree $max\_in\_degree$, default = 5
+\end{itemize}
+
+Procedure:
+\begin{itemize}
+	\item Use same FANTOM5 GRN
+	\item Subsample GRN such that each gene has a maximum in-degree of $max\_in\_degree$
+	\item Perform breadth-first-search from a random gene in the GRN, select $\min(num\_targets, resamp)$ first genes encountered
+	\item Select subgraph induced by the sampled housekeeping genes in the GRN
+	\item Add subgraph to current TF-target network
+	\item If less than $num\_targets$ have been sampled in this way, go back to step 2.
+\end{itemize}
 
 \subsection{Convert gene regulatory network to a set of reactions}
-\blindtext
+% TODO: write section in words instead of bullet point list
+dyngen uses Gillespie's SSA to simulate dynamic processes. An SSA simulation is an iterative process where 
+For each gene $G$ a set of reactions is created.
 
@@
 \blindtext
+
+\subsection{Example runs of predefined backbones}
+
+\subsubsection{Linear}
+\subsubsection{Bifurcating}
+\subsubsection{Cycle}
+\subsubsection{Branching} (and binary tree, consecutive bifurcating, trifurcating)
+\subsubsection{Converging}
+\subsubsection{Bifurcating converging}
+\subsubsection{Bifurcating cycle}
+\subsubsection{Bifurcating loop}
+\subsubsection{Disconnected}
 

### 2019-08-31 content.tex
update

@@
 dyngen was initially developed as part of a comprehensive benchmark of TI methods \cite{saelens_comparisonsinglecelltrajectory_2019} but has since been extended to be applicable in a much broader context.
-We demonstrate its flexibility by simulating numerous different types of biological experiments, and using these simulations to develop new benchmarking techniques for computational tools.
+
+Inspiration was drawn from GeneNetWeaver's \cite{schaffter_genenetweaversilicobenchmark_2011} workflow to generate \textit{in silico} bulk profiles for evaluating network inference methods \cite{marbach_wisdomcrowdsrobust_2012}.
+GeneNetWeaver translates a known gene regulatory network (GRN) into sets of Ordinary Differential Equations (ODE) in order to simulate \ldots.
+% todo: describe GNW in this one sentance
+There are two significant differences in dyngen in comparison to GeneNetWeaver. 
+dyngen simulates a single-cell instead of a bulk profile.
+However, taking random walks through ODE systems in which key molecules occur with low abundance can result in inaccurate results.
+%rephrase and cite relevant work
+Gillespie's Stochastic Simulation Algorithm (SSA)  \cite{gillespie_exactstochasticsimulation_1977} allows to simulate low abundance systems.
+Running a simulation with SSA is an iterative process where at each iteration one reaction is allowed to take place.
+Each reaction consists of its propensity -- a formula to calculate the probability of occurring during an infinitesimal time interval -- and the effect -- how it will affect the current state if triggered.
+Secondly, the GRN is constructed such that it mimics a dynamic process of interest, such as cell differentiation into multiple cell types.
+
+We demonstrate dyngen's flexibility by simulating numerous different types of biological experiments, and using these simulations to develop new benchmarking techniques for computational tools.
 
@@
 \section{Methods}
+NOTE: The method section is really rough at this stage.
 
-dyngen's approach was inspired by GeneNetWeaver's \cite{schaffter_genenetweaversilicobenchmark_2011} workflow to generate \textit{in silico} bulk profiles for evaluating network inference methods \cite{marbach_wisdomcrowdsrobust_2012}.
 The workflow to generate \textit{in silico} single cell data consists of six main steps (Figure \ref{fig:explain_methods}).
@@
 
-\subsection{Defining the module network and backbone}
+\subsection{Defining the backbone: modules and states} \label{sec:backbone}
 
@@
 If this effect is undesirable, more safeguards need to be put in place to prevent reprogramming from occurring (Section \ref{sec:dyngen_backbone_lego}).
+
+TODO: mention strength and cooperativity.
 
@@
 	\item For every TF in a particular module $M$, connect that TF with $num\_tf\_ints$ TFs for each upstream module of $M$.
+	\item The strength and cooperativity of interactions created this way are defined by the module network.
 \end{itemize}
@@
 
-\subsection{Convert gene regulatory network to a set of reactions}
+\subsection{Convert gene regulatory network to a set of reactions} \label{sec:reactions}
 % TODO: write section in words instead of bullet point list
-dyngen uses Gillespie's SSA to simulate dynamic processes. An SSA simulation is an iterative process where 
-For each gene $G$ a set of reactions is created.
 
-\subsection{Simulate developmental backbone} \label{sec:dyngen_sim_backbone}
-\blindtext
+Each reaction consists of its propensity -- a formula to calculate the probability of occurring during an infinitesimal time interval -- and the effect -- how it will affect the current state if triggered.
+
+We define the abundance levels of pre-mRNA, mRNA and protein of gene $G$ as $w_G$ $x_G$ and $y_G$ respectively.
+Five reactions affect the abundance levels of these molecules: transcription, splicing, mRNA degradation, translation, and protein degradation.
+The effects and propensity functions of these reactions are defined in Table \ref{tab:reaction_def}.
+
+% see https://en.wikipedia.org/wiki/Hill_equation_(biochemistry)#Regulation_of_gene_transcription
+\begin{table}[h]
+	\caption{\textbf{Reactions affecting the abundance levels of pre-mRNA $w_G$, mRNA $x_G$ and proteins $y_G$ of gene $G$.} Define the set of regulators of $G$ as $R_G$, the set of upregulating regulators of $G$ as $R^+_G$, and the set of downregulating regulators of $G$ as $R^-_G$.
+Parameters used in the propensity formulae are defined in Table \ref{tab:reaction_params}.} \label{tab:reaction_def}
+	\centering
+	\begin{tabular}{|lcc|}
+		\hline
+		Reaction & Effect & Propensity \\ \hline \hline
+		Transcription & $\emptyset \rightarrow w_G$ & $wpr_G \times \frac{ba_G - 1 + \prod\limits_{H \in R^+_G}(1 + reg_{G,H})}{ba_G - 1 + \prod\limits_{H \in R_G}(1 + reg_{G,H})}$, \\
+%		& & with $reg_{G,H} = \left(str_{G,H} \times \frac{y_H}{0.5 \times maxy_H}\right) ^ {co_{G,H}}$ \\
+%		& & and $maxy_H = \frac{wpr_H \times ypr_H}{xdr_H \times ydr_H}$ \\
+		Splicing & $w_G \rightarrow x_G$ & $wsr_G \times w_G$ \\
+		mRNA degradation & $x_G \rightarrow \emptyset$ & $xdr_G \times x_G$ \\
+		Translation & $x_G \rightarrow x_G + y_G$ & $ypr_G \times x_G$ \\
+		Protein degradation & $y_G \rightarrow \emptyset$ & $ydr_G \times y_G$ \\ \hline
+	\end{tabular}
+\end{table}
+
+% TODO: explain derivation of transcription formula? 
+% regG,H represents a 'probability' that H is bound to G (though does not yet scale between [0,1]).
+y_H / maxy_H is more likely to be a probability that H is bound to G, though this does not take into account that H can be bound to other genes as well.
+% in any case, if it is, then it upregulates G by strG,H. It is assumed binding events
+% are independant and result in multiplicative effect.
+
+% defining P_H is the probability that H is bound to G, 
+% and the number of molecules produced C_H = ifelse(eff_G,H == 1, wpr_G × str_{G,H}, 0)
+% should the expented no. of reactions (propensity) not be defined as:
+% BA_G + Prod_H {C_H × P_H + (1 - P_H)} - Prod_H {1 - P_H} ?
+
+% according to https://en.wikipedia.org/wiki/Hill_equation_(biochemistry)#Proportion_of_ligand-bound_receptors:
+% P_H = (y_H ^ N) / (degradation_H / production_H + y_H ^ N)
+% ... 
+% ... 
+% might need to go back to things like:
+% https://kar.kent.ac.uk/24077/1/myzabetpaper.pdf
+
+% max protein doesn't work as expected (given any simulation in model2):
+%df <- data.frame(time = model2$simulations$meta$sim_time, as.matrix(model2$simulations$counts)) %>% 
+%  gather(mol, value, -time) %>% 
+%  filter(grepl("^y_", mol)) %>% 
+%  mutate(feature_id = gsub("^y_", "", mol)) %>% 
+%  left_join(model2$feature_info %>% transmute(feature_id, max_protein = wpr / xdr * ypr / ydr, module_id), by = "feature_id") %>% 
+%  mutate(tfname = gsub("^[^_]*_", "", feature_id))
+%ggplot(df) + geom_line(aes(time, value / max_protein, colour = tfname, group = feature_id)) + facet_wrap(~module_id)
+
+\begin{table}[h]
+	\caption{\textbf{Parameters defined for the calculation of reaction propensity functions.}} \label{tab:reaction_params}
+	\centering
+	\begin{tabular}{|lrl|}
+		\hline
+		Parameter & Symbol & Definition \\ \hline \hline
+		Transcription rate & $wpr_G$ & $\in \{. | . \in N(100, 20), . \geq 10\}$ \\
+		Splicing rate & $wsr_G$ & $\in \{. | . \in N(10, 2), . \geq 2\}$ \\
+		mRNA degradation rate & $xdr_G$ & $\in \{. | . \in N(5, 1), . \geq 2\}$ \\
+		Translation rate & $ypr_G$ & $\in \{. | . \in N(5, 1), . \geq 2\}$ \\
+		Protein degradation rate & $ydr_G$ & $\in \{. | . \in N(3, 0.5), . \geq 1\}$ \\
+		Interaction strength & $str_{G,H}$ & $\in \{10^. | . \in U(0, 2)\}$ * \\ 
+		Interaction cooperativity & $co_{G,H}$ & $\in U(0.5, 2)$ * \\\hline\hline
+		Max protein & $maxy_H$ & $= \frac{wpr_H \times ypr_H}{xdr_H \times ydr_H}$ \\ 
+		Regulation activity & $reg_{G,H}$ & $= \left(str_{G,H} \times \frac{y_H}{0.5 \times maxy_H}\right) ^ {co_{G,H}}$ \\
+		Basal expression & $ba_G$ & $= \begin{cases} 1 & \mbox{if } R^+_G = \emptyset \\ 0.0001 & \mbox{if } R^-_G = \emptyset \mbox{ and } R^+_G \neq \emptyset \\ 0.5 & \mbox{otherwise} \end{cases}$ * \\ \hline
+		\multicolumn{3}{l}{*: unless already defined when $G$ is a TF.}
+	\end{tabular}
+\end{table}
+
+% todo ask Wouter, why not simply the following?
+% ba_G = ifelse(R^+_G is not empty, 1, 0)
+
+\subsection{Compute average expression along backbone transitions} \label{sec:dyngen_sim_backbone}
+
+% → basically an ODE
+When simulating the developmental backbone, we go through the edges of the backbone state network defined in an earlier step (Section \ref{sec:backbone}), starting from the root state.
+It is assumed the root state has no modules active and has no expression of any molecules.
+To get to next state, we follow a transition starting from the root state, activate and deactivate the modules as indicated by the transition, and compute the average molecule abundance along the transition.
+To compute the average abundance, we perform small time steps $t = 0.001$ and let each reaction (Section \ref{sec:reactions}) occur $t$ times its propensity.
 
 \subsection{Simulate single cells} \label{sec:dyngen_ssa}
-\blindtext
+%TODO: STOCKS \cite{kierzek_stocksstochastickinetic_2002} uses GSSA to simulate something.
+% maybe we could have a look at that! 
+% or https://nyaspubs.onlinelibrary.wiley.com/doi/full/10.1111/j.1749-6632.2008.03756.x
+
+% or even https://academic.oup.com/bioinformatics/article/24/10/1318/178687 ('AND' and 'OR')
+% and https://academic.oup.com/bioinformatics/article/25/9/1205/203870 (regulation motifs, kinetics)
+% https://nyaspubs.onlinelibrary.wiley.com/doi/full/10.1111/j.1749-6632.2008.03756.x (modules of genes)
+
+%TODO: improve this section a lot. Basically, explain SSA and GillespieSSA2.
+
+dyngen uses Gillespie's Stochastic Simulation Algorithm (SSA) to simulate dynamic processes. An SSA simulation is an iterative process where at each iteration one reaction is triggered.
+
+Each reaction consists of its propensity -- a formula to calculate the probability of occurring during an infinitesimal time interval -- and the effect -- how it will affect the current state if triggered.
+Each time a reaction is triggered, the simulation time is incremented by $\tau = \frac{1}{\sum_j prop_j} \ln\left(\frac{1}{r}\right)$, with $r \in U(0, 1)$ and $prop_j$ the propensity value of the $j$th reaction for the current state of the simulation.
+
+SSA simulations are notoriously slow.
+We use GillespieSSA2 which contains many optimisations such as translating and compiling all the propensity functions to C++ and implementations of approximations of SSA which allows to trigger many reactions simultaneously at each iteration.
+
+The framework allows to store the abundance levels of molecules only after a specific interval has passed since the previous census.
+By setting the census interval to 0, the whole simulation's trajectory is retained but many of these time points will contain very similar information.
+In addition to the abundance levels, also the propensity values and the number of firings of each of the reactions at each of the time steps can be retained, as well as specific sub-calculations of the propensity values, such as the regulator activity level $reg_{G,H}$.
 
 \subsubsection{Map SSA simulations to backbone}
-\blindtext
+The cellular state of each timepoint in the SSA simulation is mapped to the state network of the backbone by calculating the 1NN between a state vector in the simulation and the average expression levels along transitions.
 
 \subsection{Simulate experiment}
-\blindtext
+From the SSA simulation we obtain the abundance levels of all the molecules at the different time points.
+We need to replicate technical effects introduced by experimental protocols in order to obtain data that is similar to real data.
+For this, the cells are sampled from the simulations, and molecules are sampled for each of the cells.
+Real datasets are used in order to achieve similar data characteristics.
 
-\subsubsection{Snapshot experiment}
-\blindtext
+\subsubsection{Sample cells}
+Cells can be sampled from an unsynchronised population of single cells (snapshot) or at multiple time points in a synchronised population (time series).
 
-\subsubsection{Time series experiment}
-\blindtext
+\paragraph{Snapshot} Cells are just sampled randomly from the different time points in the simulation.
 
-\subsubsection{Perturbation experiment}
-\blindtext
+\paragraph{Time series} The timeline of the simulations is cut up into chunks. From several of these chunks, cells are sampled. For each cell it is known at which time point it was sampled.
+
+\subsubsection{Sample molecules}
+
+\begin{itemize}
+	\item From real dataset, look at the number of transcripts that was captured per cell. Library size $ls_i$ of cell $i$ is samples from this distribution.
+	\item Capture rate of each molecule type $j$ is drawn from $cr_j \in N(1, 0.05)$
+	\item For each cell $i$, a multinomial distribution is used to draw $ls_i$ molecules from molecule type $j$ with probability $cr_j \times ab_{i,j}$ with $ab_{i,j}$ the molecule abundance level of molecule $j$ in cell $i$.
+\end{itemize}
 
@@
 From discussion: Sampling a cell at a certain time point and once more at a later time point allows evaluating the performance of RNA velocity approaches -- which predict the future state of a cell by looking at differences in pre-mRNA and mRNA abundance levels.
+
+\subsubsection{Perturbation experiment}
 

### 2019-09-03 content.tex
update

@@
 
-We define the abundance levels of pre-mRNA, mRNA and protein of gene $G$ as $w_G$ $x_G$ and $y_G$ respectively.
+\newcommand{\w}[1]{\text{w}_{#1}}
+\newcommand{\x}[1]{\text{x}_{#1}}
+\newcommand{\y}[1]{\text{y}_{#1}}
+
+\newcommand{\rs}[1]{\text{R}_{#1}}
+\newcommand{\rp}[1]{\text{R}^+_{#1}}
+\newcommand{\rn}[1]{\text{R}^-_{#1}}
+
+\newcommand{\wpr}[1]{\text{wpr}_{#1}}
+\newcommand{\wsr}[1]{\text{wsr}_{#1}}
+\newcommand{\xdr}[1]{\text{xdr}_{#1}}
+\newcommand{\ypr}[1]{\text{ypr}_{#1}}
+\newcommand{\ydr}[1]{\text{ydr}_{#1}}
+
+\newcommand{\str}[1]{\text{str}_{#1}}
+\newcommand{\co}[1]{\text{co}_{#1}}
+\newcommand{\ind}[1]{\text{ind}_{#1}}
+\newcommand{\hmy}[1]{\text{hmy}_{#1}}
+\newcommand{\reg}[1]{\text{reg}_{#1}}
+\newcommand{\ba}[1]{\text{ba}_{#1}}
+
+We define the abundance levels of pre-mRNA, mRNA and protein of gene $G$ as $\w G$, $\x G$ and $\y G$ respectively.
 Five reactions affect the abundance levels of these molecules: transcription, splicing, mRNA degradation, translation, and protein degradation.
@@
 \begin{table}[h]
-	\caption{\textbf{Reactions affecting the abundance levels of pre-mRNA $w_G$, mRNA $x_G$ and proteins $y_G$ of gene $G$.} Define the set of regulators of $G$ as $R_G$, the set of upregulating regulators of $G$ as $R^+_G$, and the set of downregulating regulators of $G$ as $R^-_G$.
+	\caption{\textbf{Reactions affecting the abundance levels of pre-mRNA $\w G$, mRNA $\x G$ and proteins $\y G$ of gene $G$.} Define the set of regulators of $G$ as $\rs{G}$, the set of upregulating regulators of $G$ as $\rp G$, and the set of downregulating regulators of $G$ as $\rn G$.
 Parameters used in the propensity formulae are defined in Table \ref{tab:reaction_params}.} \label{tab:reaction_def}
@@
 		Reaction & Effect & Propensity \\ \hline \hline
-		Transcription & $\emptyset \rightarrow w_G$ & $wpr_G \times \frac{ba_G - 1 + \prod\limits_{H \in R^+_G}(1 + reg_{G,H})}{ba_G - 1 + \prod\limits_{H \in R_G}(1 + reg_{G,H})}$, \\
-%		& & with $reg_{G,H} = \left(str_{G,H} \times \frac{y_H}{0.5 \times maxy_H}\right) ^ {co_{G,H}}$ \\
-%		& & and $maxy_H = \frac{wpr_H \times ypr_H}{xdr_H \times ydr_H}$ \\
-		Splicing & $w_G \rightarrow x_G$ & $wsr_G \times w_G$ \\
-		mRNA degradation & $x_G \rightarrow \emptyset$ & $xdr_G \times x_G$ \\
-		Translation & $x_G \rightarrow x_G + y_G$ & $ypr_G \times x_G$ \\
-		Protein degradation & $y_G \rightarrow \emptyset$ & $ydr_G \times y_G$ \\ \hline
+		Transcription & $\emptyset \rightarrow \w G$ & $\wpr G \times \frac{\ba G - \ind{G}^{|\rp{G}|} + \prod\limits_{H \in \rp{G}}(\ind G + \reg{G,H})}{\prod\limits_{H \in \rp{G}}(1 + \reg{G,H})}$ \\
+		Splicing & $\w G \rightarrow \x G$ & $\wsr G \times \w G$ \\
+		mRNA degradation & $\x G \rightarrow \emptyset$ & $\xdr G \times \x G$ \\
+		Translation & $\x G \rightarrow \w G + \y G$ & $\ypr G \times \x G$ \\
+		Protein degradation & $\y G \rightarrow \emptyset$ & $\ydr G \times \y G$ \\ \hline
 	\end{tabular}
@@
 % in any case, if it is, then it upregulates G by strG,H. It is assumed binding events
-% are independant and result in multiplicative effect.
-
-% defining P_H is the probability that H is bound to G, 
-% and the number of molecules produced C_H = ifelse(eff_G,H == 1, wpr_G × str_{G,H}, 0)
-% should the expented no. of reactions (propensity) not be defined as:
-% BA_G + Prod_H {C_H × P_H + (1 - P_H)} - Prod_H {1 - P_H} ?
+% are independent and result in multiplicative effect.
 
@@
 
-% max protein doesn't work as expected (given any simulation in model2):
-%df <- data.frame(time = model2$simulations$meta$sim_time, as.matrix(model2$simulations$counts)) %>% 
-%  gather(mol, value, -time) %>% 
-%  filter(grepl("^y_", mol)) %>% 
-%  mutate(feature_id = gsub("^y_", "", mol)) %>% 
-%  left_join(model2$feature_info %>% transmute(feature_id, max_protein = wpr / xdr * ypr / ydr, module_id), by = "feature_id") %>% 
-%  mutate(tfname = gsub("^[^_]*_", "", feature_id))
-%ggplot(df) + geom_line(aes(time, value / max_protein, colour = tfname, group = feature_id)) + facet_wrap(~module_id)
-
 \begin{table}[h]
@@
 		Parameter & Symbol & Definition \\ \hline \hline
-		Transcription rate & $wpr_G$ & $\in \{. | . \in N(100, 20), . \geq 10\}$ \\
-		Splicing rate & $wsr_G$ & $\in \{. | . \in N(10, 2), . \geq 2\}$ \\
-		mRNA degradation rate & $xdr_G$ & $\in \{. | . \in N(5, 1), . \geq 2\}$ \\
-		Translation rate & $ypr_G$ & $\in \{. | . \in N(5, 1), . \geq 2\}$ \\
-		Protein degradation rate & $ydr_G$ & $\in \{. | . \in N(3, 0.5), . \geq 1\}$ \\
-		Interaction strength & $str_{G,H}$ & $\in \{10^. | . \in U(0, 2)\}$ * \\ 
-		Interaction cooperativity & $co_{G,H}$ & $\in U(0.5, 2)$ * \\\hline\hline
-		Max protein & $maxy_H$ & $= \frac{wpr_H \times ypr_H}{xdr_H \times ydr_H}$ \\ 
-		Regulation activity & $reg_{G,H}$ & $= \left(str_{G,H} \times \frac{y_H}{0.5 \times maxy_H}\right) ^ {co_{G,H}}$ \\
-		Basal expression & $ba_G$ & $= \begin{cases} 1 & \mbox{if } R^+_G = \emptyset \\ 0.0001 & \mbox{if } R^-_G = \emptyset \mbox{ and } R^+_G \neq \emptyset \\ 0.5 & \mbox{otherwise} \end{cases}$ * \\ \hline
+		Transcription rate & $\wpr{G}$ & $\in \{. | . \in N(100, 20), . \geq 10\}$ \\
+		Splicing rate & $\wsr G$ & $\in \{. | . \in N(10, 2), . \geq 2\}$ \\
+		mRNA degradation rate & $\xdr{G}$ & $\in \{. | . \in N(5, 1), . \geq 2\}$ \\
+		Translation rate & $\ypr{G}$ & $\in \{. | . \in N(5, 1), . \geq 2\}$ \\
+		Protein degradation rate & $\ydr G$ & $\in \{. | . \in N(3, 0.5), . \geq 1\}$ \\
+		Interaction strength & $\str{G,H}$ & $\in \{10^. | . \in U(0, 2)\}$ * \\ 
+		Interaction cooperativity & $\co{G,H}$ & $\in U(0.5, 2)$ * \\
+    Independence factor & $\ind G$ & $\in [0, 1]$ \\ \hline\hline
+		TF concentration at half-maximal binding & $\hmy H$ & $= 0.5 \times \frac{\wpr H \times \ypr H}{\xdr H \times \ydr H}$ \\ 
+		Regulation activity & $\reg{G,H}$ & $= \left(\str{G,H} \times \frac{\y H}{\hmy H}\right) ^ {\co{G,H}}$ \\
+		Basal expression & $\ba G$ & $= \begin{cases} 1 & \mbox{if } \rp{G} = \emptyset \\ 0.0001 & \mbox{if } \rn{G} = \emptyset \mbox{ and } \rp{G} \neq \emptyset \\ 0.5 & \mbox{otherwise} \end{cases}$ * \\ \hline
 		\multicolumn{3}{l}{*: unless already defined when $G$ is a TF.}

### 2019-09-03 content.tex
update

@@
 		Parameter & Symbol & Definition \\ \hline \hline
-		Transcription rate & $\wpr{G}$ & $\in \{. | . \in N(100, 20), . \geq 10\}$ \\
-		Splicing rate & $\wsr G$ & $\in \{. | . \in N(10, 2), . \geq 2\}$ \\
-		mRNA degradation rate & $\xdr{G}$ & $\in \{. | . \in N(5, 1), . \geq 2\}$ \\
-		Translation rate & $\ypr{G}$ & $\in \{. | . \in N(5, 1), . \geq 2\}$ \\
-		Protein degradation rate & $\ydr G$ & $\in \{. | . \in N(3, 0.5), . \geq 1\}$ \\
-		Interaction strength & $\str{G,H}$ & $\in \{10^. | . \in U(0, 2)\}$ * \\ 
+		Transcription rate & $\wpr{G}$ & $\in N(100, 20),\ \geq 10$ \\
+		Splicing rate & $\wsr G$ & $\in N(10, 2),\ \geq 2$ \\
+		mRNA degradation rate & $\xdr{G}$ & $\in N(5, 1),\ \geq 2$ \\
+		Translation rate & $\ypr{G}$ & $\in N(5, 1),\ \geq 2$ \\
+		Protein degradation rate & $\ydr G$ & $\in N(3, 0.5),\ \geq 1$ \\
+		Interaction strength & $\str{G,H}$ & $\in 10^{U(0, 2)}$ * \\ 
 		Interaction cooperativity & $\co{G,H}$ & $\in U(0.5, 2)$ * \\
-    Independence factor & $\ind G$ & $\in [0, 1]$ \\ \hline\hline
+    Independence factor & $\ind G$ & $\in [0, 1]$ * \\ \hline\hline
 		TF concentration at half-maximal binding & $\hmy H$ & $= 0.5 \times \frac{\wpr H \times \ypr H}{\xdr H \times \ydr H}$ \\ 

### 2019-09-06 content.tex
some changes

@@
 
-A few generators of scRNA-seq profiles have already been developed (e.g.
-splatter \cite{zappia_splattersimulationsinglecell_2017}, powsimR \cite{vieth_powsimrpoweranalysis_2017} and PROSSTT \cite{papadopoulos_prossttprobabilisticsimulation_2018}).
+A few simulators of scRNA-seq profiles have already been developed (e.g.
+splatter \cite{zappia_splattersimulationsinglecell_2017}, PROSSTT \cite{papadopoulos_prossttprobabilisticsimulation_2018} and SymSim \cite{zhang_simulatingmultiplefaceted_2019}).
 % BoolODE and dyntoy are also examples.
@@
 dyngen was initially developed as part of a comprehensive benchmark of TI methods \cite{saelens_comparisonsinglecelltrajectory_2019} but has since been extended to be applicable in a much broader context.
+We demonstrate its flexibility by simulating three different types of biological experiments, and using these simulations to develop new benchmarking techniques for computational tools.
 
-Inspiration was drawn from GeneNetWeaver's \cite{schaffter_genenetweaversilicobenchmark_2011} workflow to generate \textit{in silico} bulk profiles for evaluating network inference methods \cite{marbach_wisdomcrowdsrobust_2012}.
-GeneNetWeaver translates a known gene regulatory network (GRN) into sets of Ordinary Differential Equations (ODE) in order to simulate \ldots.
-% todo: describe GNW in this one sentance
-There are two significant differences in dyngen in comparison to GeneNetWeaver. 
-dyngen simulates a single-cell instead of a bulk profile.
-However, taking random walks through ODE systems in which key molecules occur with low abundance can result in inaccurate results.
-%rephrase and cite relevant work
-Gillespie's Stochastic Simulation Algorithm (SSA)  \cite{gillespie_exactstochasticsimulation_1977} allows to simulate low abundance systems.
-Running a simulation with SSA is an iterative process where at each iteration one reaction is allowed to take place.
-Each reaction consists of its propensity -- a formula to calculate the probability of occurring during an infinitesimal time interval -- and the effect -- how it will affect the current state if triggered.
-Secondly, the GRN is constructed such that it mimics a dynamic process of interest, such as cell differentiation into multiple cell types.
-
-We demonstrate dyngen's flexibility by simulating numerous different types of biological experiments, and using these simulations to develop new benchmarking techniques for computational tools.
-
-\begin{figure}[htb!]
+\begin{figure}[hb]
 	\centering
 	\includegraphics[width=\linewidth]{fig/showcase_3.png} 
-	\caption{Showcase of dyngen functionality. \textbf{TODO: change to pdf.} Remove B?} % TODO: update label
+	\caption{Showcase of dyngen functionality. \textbf{TODO: change to pdf.} Remove B? \textit{Yes please}} % TODO: update label
 	\label{fig:dyngen}
@@
 \section{Results}
-dyngen is a simulator for single cells that develop over time.
+dyngen simulates the transcriptomic changes of a cell over time using a model of gene regulation.
 Throughout this section, a simple simulation of a cell undergoing a cyclic process is used to illustrate key strengths of dyngen (Figure \ref{fig:simplecyclic}).
@@
 
-\begin{figure}[htb!]
+\begin{figure}[hb]
 	\centering
@@
 
-One of dyngen's main advantages is that through careful engineering of the GRN, different cellular developmental processes can be obtained.
+One of dyngen's main advantages is that through careful engineering of the GRN, different cellular developmental processes can be simulated.
 Different GRNs can result in branching, converging, cyclic, or even disconnected developmental topologies.
@@
 % could use figure; one with GRNs of different topologies, another with rewiring events
-Multiple simulations with different initial molecule abundance levels can be used to replicate batch effects.
+Multiple simulations with some slight perturbations can be used to replicate batch effects.
 
@@
 By default, dyngen supports snapshot experiments (uniformly sampling from an asynchronous dynamic process) and time-series experiments (sampling cells from different intervals in the simulation).
-It is possible to implement other experimental protocols (which perhaps do not exist in real life), such as sampling the same cell at regular intervals.
-
-%% TODO: show that dyngen output resembles real data using e.g. countsimQC?
-
-%% TODO: add result figures pertaining use cases
+It is possible to implement other experimental protocols, such as sampling the same cell at regular intervals.
 
@@
 As is, dyngen's single cell simulations can be used to evaluate common single-cell omics computational methods such as clustering, batch correction, trajectory inference and network inference.
-However, the combined effect of these advantages results in a framework that is flexible enough to adapt to a broad range of applications.
+However, the combined effect of these advantages results in a framework that is flexible enough to adapt to a broad range of applications, some of which may still be needed in the future.
 This may include methods that integrate clustering, network inference and trajectory inference.
-In this respect, dyngen may promote the development of new tools in the single-cell field similarly as other simulators have done in the past \cite{schaffter_genenetweaversilicobenchmark_2011,ewing_combiningtumorgenome_2015}.
+In this respect, dyngen may promote the development of new tools in the single-cell field similarly as other simulators have done in the past \textbf{10.1093/bioinformatics/btx631} \textbf{10.1093/bioinformatics/btr373} \textbf{10.1038/nmeth.3407}.
 
-Adding batch effects to snapshot simulations of linear (or even branching) trajectories allows evaluating trajectory alignment methods -- which attempt to map two or more trajectories onto each other.
-Adding perturbations to the GRN allows evaluating the performance of differential network inference methods -- which predict differential regulatory interactions between two or more groups of profiles.
-Sampling a cell at a particular time point and once more at a later time point allows evaluating the performance of RNA velocity approaches -- which predict the future state of a cell by looking at differences in pre-mRNA and mRNA abundance levels.
+Adding batch effects to snapshot simulations of linear (or even branching) trajectories allows to evaluate trajectory alignment methods -- which attempt to map two or more trajectories onto each other.
+Adding perturbations to the GRN allows to evaluate the performance of differential network inference methods -- which predict differential regulatory interactions between two or more groups of profiles.
+Sampling a cell at a certain time point and once more at a later time point allows to evaluate the performance of RNA velocity approaches -- which predict the future state of a cell by looking at differences in pre-mRNA and mRNA abundance levels.
 
@@
 Similarly, it could also be used to compare which experimental technique will likely produce the most accurate result.
-For example, is it possible to infer directionality of regulatory interactions from snapshot experiments only, or are time series or knockdown experiments a necessity in order to infer high-quality regulatory networks?
+For example, is it possible to infer directionality of regulatory interactions from snapshot experiments only, or are time series or knock down experiments a necessity in order to infer high quality regulatory networks?
 
@@
 \section{Methods}
-NOTE: The method section is really rough at this stage.
 
-The workflow to generate \textit{in silico} single cell data consists of six main steps (Figure \ref{fig:explain_methods}).
+\subsection{Simulating a snapshot experiment with dyngen}
+\subsubsection{Define the module network and the state network}
+\subsubsection{Generate the transcription factor network}
+\subsubsection{Generate targets and housekeeping genes}
+\subsubsection{Convert gene regulatory network to a set of SSA reactions}
+\subsubsection{Simulate backbone}
+\subsubsection{Run SSA simulations}
+\subsubsection{Map SSA simulations to backbone}
+\subsubsection{Simulate snapshot experiment}
 
-\begin{figure}[htb!]
-	\centering
-	\includegraphics[width=\hugefigure]{fig/explain_methods} 
-	\caption{\textbf{The workflow of dyngen is comprised of six main steps.} \textbf{A:} The user needs to specify the desired module network or use a predefined module network.
-\textbf{B:} Each gene in a module is is regulated by one or more transcription factors from the upstream module.
-Additional target genes are generated.
-\textbf{C:} Each gene regulatory interaction in the GRN is converted to a set of biochemical reactions.
-\textbf{D:} Along with the module network, the user also needs to specify the backbone structure of expected cell states.
-The average expression of each edge in the backbone is simulated by activating a restricted set of genes for each edge.
-\textbf{E:} Multiple Gillespie SSA simulations are run using the reactions defined in step C.
-The counts of each of the mocules at each time step are extracted.
-Each time step is mapped to a point in the backbone.
-\textbf{F:} Multiple cells are sampled from each simulation.
-Molecules are sampled from each cell.}
-	\label{fig:explain_methods}
-\end{figure}
-
-\subsection{Defining the backbone: modules and states} \label{sec:backbone}
-
-One of the main processes involved in cellular dynamic processes is gene regulation, where regulatory cascades and feedback loops lead to progressive changes in expression and decision making.
-The exact way a cell chooses a certain path during its differentiation is still an active research field, although certain models have already emerged and been tested \textit{in vivo}.
-One driver of bifurcation seems to be mutual antagonism, where two genes \cite{xu_regulationbifurcatingcell_2015} strongly repress each other, forcing one of the two to become inactive \cite{graf_forcingcellschange_2009}.
-Such mutual antagonism can be modelled and simulated \cite{wang_quantifyingwaddingtonlandscape_2011, ferrell_bistabilitybifurcationswaddington_2012}.
-Although the two-gene model is simple and elegant, the reality is frequently more complex, with multiple genes (grouped into modules) repressing each other \cite{yosef_dynamicregulatorynetwork_2013}.
-
-In dyngen, the user defines the behaviour of the simulation by defining how sets of genes, called modules, are regulating eachother.
-A module may have basal expression, which means that pre-mRNA of the genes in this module will be transcribed without the presence of transcription factor molecules.
-A module marked as "active during the burn phase" means that this module will be allowed to generate expression of its genes during an initial warm-up phase (See section \ref{sec:dyngen_ssa}).
-At the end of the dyngen process, cells will not be sampled from the burn phase simulations.
-
-Several examples of module networks are given (Figure \ref{fig:example_backbones}). 
-A simple chain of modules (where one module upregulates the next) results in a \emph{linear} process.
-By having the last module repress the first module, the process becomes \emph{cyclic}.
-Two modules repressing eachother is the basis of a \emph{bifurcating} process, though several chains of modules have to be attached in order to achieve progression before and after the bifurcation process.
-Finally, a \emph{converging} process has a bifurcation occurring during the burn phase, after which any differences in module regulation is removed.
-
-Note that these examples represent the bare minimum in terms of number of modules used.
-Using longer chains of modules is typically desired.
-In addition, the fate decisions made in this example of a bifurcation is reversible, meaning cells can be reprogrammed to go down a different differentiation path.
-If this effect is undesirable, more safeguards need to be put in place to prevent reprogramming from occurring (Section \ref{sec:dyngen_backbone_lego}).
-
-TODO: mention strength and cooperativity.
-
-\begin{figure}[htb!]
-	\centering
-	\includegraphics[width=\LARGEfigure]{fig/example_backbones} 
-	\caption{Example module networks}
-	\label{fig:example_backbones}
-\end{figure}
-
-In addition to the module network, the user also needs to define a network of cellular states called the "backbone".
-Before simulating any cells, each transition in the backbone is simulated separately to obtain the average changes in expression along that transition (Figure \ref{fig:explain_methods}D).
-As part of the backbone, the user needs to specify which modules are allowed to alter its expression from one state to another.
-For example, in order to transition from state S0 to S1 in the cyclic example, gene modules A, B and C are turned on and a simulation is allowed to run.
-To transition from S1 to S2, gene modules D and E are turned on, and expression of gene module C is kept constant.
-To transition from S2 to S3, C is turned on again and now A and B are fixed.
-Finally, to transition from S3 to S1 again, A and B are turned on again and D and E are fixed again.
-Demonstrations of the backbone will be explained in more detail in section \ref{sec:dyngen_sim_backbone}.
-
-\subsubsection{Backbone lego} \label{sec:dyngen_backbone_lego}
-The backbone can make use of one or more "backbone lego" (BBL) pieces (Figure \ref{fig:backbone_lego}).
-A BBL consists of one or more modules which regulate each other such that the output modules present a specific behaviour, depending on the input module (Figure \ref{fig:backbone_lego}A).
-Parameters allow determining the number of modules involved in the process and the number of outputs.
-Multiple BBLs can be chained together in order to intuitively create milestone networks and corresponding state networks (Figure \ref{fig:backbone_lego}B).
-Note that not all dynamic processes can be represented by a combination of BBLs, but they can serve as common building blocks to aid the construction of the backbone.
-
-\begin{figure}[htb!]
-	\centering
-	\includegraphics[width=\LARGEfigure]{fig/backbone_lego} 
-	\caption{Backbone lego}
-	\label{fig:backbone_lego}
-\end{figure}
-
-When the input node of a \textbf{linear BBL} (Figure \ref{fig:backbone_lego}C) is upregulated, the module the BBL is connected to will be upregulated.
-A \emph{simple chain} is a set of modules where a module upregulates the next.
-A \emph{chain with double repression} has an uneven number of modules forming a chain where each module downregulates the next but all modules (except the input) have basal expression.
-A \emph{grid with double repression} is similar; except that modules do not have basal expression but instead get upregulated by an upstream module in the chain.
-Finally, a \emph{flip flop} consists of a simple chain where first the modules (except the last) are upregulated.
-Once the second to last module is upregulated, that module upregulates itself and the first module is strongly repressed, causing all other modules to lose expression and finally the last module to be upregulated.
-The \emph{flip flop} retains this output state, even when the input changes.
-
-When the input node of a \textbf{branching BBL} (Figure \ref{fig:backbone_lego}D) is upregulated, a subset of its output modules will eventually be upregulated.
-A \emph{simple branching} uses reciprocal inhibition to drive the upregulation of one of the output modules.
-Due to its simplicity, however, multiple output modules might be upregulated simultaneously, and over long periods of simulation time it might be possible that the choice of upregulated module changes.
-A \emph{robust branching} improves upon the simple branching by preventing upregulation of output modules until an internal branching decision has been made, and by repressing the decision mechanism to avoid other output modules being upregulated other than the one that has been chosen.
-
-A \textbf{leaf BBL} (Figure \ref{fig:backbone_lego}E) is a linear BBL that has either no inputs or no outputs.
-A \emph{start} BBL is a linear BBL where the first module has basal expression, and all modules in this module will be active during the burn-in phase of the simulation (Section \ref{sec:dyngen_sim_backbone}).
-An \emph{end} BBL is also a linear BBL with its output regulating one final module.
-
-\subsection{Generate gene regulatory network}
-% TODO: write section in words instead of bullet point list
-
-\subsubsection{Generate the transcription factor network}
-Parameters: 
-\begin{itemize}
-	\item Number of TFs to generate $num\_tfs$
-	\item Minimum TFs per module $min\_tfs$, default = 1
-	\item Number of regulatory interactions per module interaction $num\_tf\_ints$, default = 2
-\end{itemize}
-
-Procedure: 
-\begin{itemize}
-	\item Divide $num\_tfs$ TFs amongst modules in backbone such that each module has at least $min\_tfs$ TFs.
-	\item For every TF in a particular module $M$, connect that TF with $num\_tf\_ints$ TFs for each upstream module of $M$.
-	\item The strength and cooperativity of interactions created this way are defined by the module network.
-\end{itemize}
-
-\subsubsection{Generate targets}
-Parameters: 
-\begin{itemize}
-	\item Number of target genes $num\_targets$
-	\item Name of a FANTOM5 GRN
-	\item Damping factor $damping$, default = 0.05
-	\item Target resampling $resamp$, default = $+\infty$
-	\item Maximum in-degree $max\_in\_degree$, default = 5
-\end{itemize}
-
-Procedure:
-\begin{itemize}
-	\item Download the FANTOM5 GRN if not already available
-	\item Randomly map TFs to the regulators in the GRN
-	\item Compute page rank from the selected regulators in the GRN with damping factor $damping$
-	\item Perform weighted sample of  $\min(num\_targets, resamp)$ targets weighted by the page rank
-	\item Select subgraph induced by the TFs and the sampled targets in the GRN, remove edges to the TFs
-	\item Add subgraph to current TF-target network
-	\item If less than $num\_targets$ have been sampled in this way, go back to step 2.
-	\item Remove regulatory interactions if a target has more than $max\_in\_degree$ edges.
-\end{itemize}
-
-\subsubsection{Generate housekeeping genes}
-Parameters: 
-\begin{itemize}
-	\item Number of housekeeping genes $num\_targets$
-	\item Name of a FANTOM5 GRN
-	\item Target resampling $resamp$, default = $+\infty$
-	\item Maximum in-degree $max\_in\_degree$, default = 5
-\end{itemize}
-
-Procedure:
-\begin{itemize}
-	\item Use same FANTOM5 GRN
-	\item Subsample GRN such that each gene has a maximum in-degree of $max\_in\_degree$
-	\item Perform breadth-first-search from a random gene in the GRN, select $\min(num\_targets, resamp)$ first genes encountered
-	\item Select subgraph induced by the sampled housekeeping genes in the GRN
-	\item Add subgraph to current TF-target network
-	\item If less than $num\_targets$ have been sampled in this way, go back to step 2.
-\end{itemize}
-
-\subsection{Convert gene regulatory network to a set of reactions} \label{sec:reactions}
-% TODO: write section in words instead of bullet point list
-
-Each reaction consists of its propensity -- a formula to calculate the probability of occurring during an infinitesimal time interval -- and the effect -- how it will affect the current state if triggered.
-
-\newcommand{\w}[1]{\text{w}_{#1}}
-\newcommand{\x}[1]{\text{x}_{#1}}
-\newcommand{\y}[1]{\text{y}_{#1}}
-
-\newcommand{\rs}[1]{\text{R}_{#1}}
-\newcommand{\rp}[1]{\text{R}^+_{#1}}
-\newcommand{\rn}[1]{\text{R}^-_{#1}}
-
-\newcommand{\wpr}[1]{\text{wpr}_{#1}}
-\newcommand{\wsr}[1]{\text{wsr}_{#1}}
-\newcommand{\xdr}[1]{\text{xdr}_{#1}}
-\newcommand{\ypr}[1]{\text{ypr}_{#1}}
-\newcommand{\ydr}[1]{\text{ydr}_{#1}}
-
-\newcommand{\str}[1]{\text{str}_{#1}}
-\newcommand{\co}[1]{\text{co}_{#1}}
-\newcommand{\ind}[1]{\text{ind}_{#1}}
-\newcommand{\hmy}[1]{\text{hmy}_{#1}}
-\newcommand{\reg}[1]{\text{reg}_{#1}}
-\newcommand{\ba}[1]{\text{ba}_{#1}}
-
-We define the abundance levels of pre-mRNA, mRNA and protein of gene $G$ as $\w G$, $\x G$ and $\y G$ respectively.
-Five reactions affect the abundance levels of these molecules: transcription, splicing, mRNA degradation, translation, and protein degradation.
-The effects and propensity functions of these reactions are defined in Table \ref{tab:reaction_def}.
-
-% see https://en.wikipedia.org/wiki/Hill_equation_(biochemistry)#Regulation_of_gene_transcription
-\begin{table}[h]
-	\caption{\textbf{Reactions affecting the abundance levels of pre-mRNA $\w G$, mRNA $\x G$ and proteins $\y G$ of gene $G$.} Define the set of regulators of $G$ as $\rs{G}$, the set of upregulating regulators of $G$ as $\rp G$, and the set of downregulating regulators of $G$ as $\rn G$.
-Parameters used in the propensity formulae are defined in Table \ref{tab:reaction_params}.} \label{tab:reaction_def}
-	\centering
-	\begin{tabular}{|lcc|}
-		\hline
-		Reaction & Effect & Propensity \\ \hline \hline
-		Transcription & $\emptyset \rightarrow \w G$ & $\wpr G \times \frac{\ba G - \ind{G}^{|\rp{G}|} + \prod\limits_{H \in \rp{G}}(\ind G + \reg{G,H})}{\prod\limits_{H \in \rp{G}}(1 + \reg{G,H})}$ \\
-		Splicing & $\w G \rightarrow \x G$ & $\wsr G \times \w G$ \\
-		mRNA degradation & $\x G \rightarrow \emptyset$ & $\xdr G \times \x G$ \\
-		Translation & $\x G \rightarrow \w G + \y G$ & $\ypr G \times \x G$ \\
-		Protein degradation & $\y G \rightarrow \emptyset$ & $\ydr G \times \y G$ \\ \hline
-	\end{tabular}
-\end{table}
-
-% TODO: explain derivation of transcription formula? 
-% regG,H represents a 'probability' that H is bound to G (though does not yet scale between [0,1]).
-y_H / maxy_H is more likely to be a probability that H is bound to G, though this does not take into account that H can be bound to other genes as well.
-% in any case, if it is, then it upregulates G by strG,H. It is assumed binding events
-% are independent and result in multiplicative effect.
-
-% according to https://en.wikipedia.org/wiki/Hill_equation_(biochemistry)#Proportion_of_ligand-bound_receptors:
-% P_H = (y_H ^ N) / (degradation_H / production_H + y_H ^ N)
-% ... 
-% ... 
-% might need to go back to things like:
-% https://kar.kent.ac.uk/24077/1/myzabetpaper.pdf
-
-\begin{table}[h]
-	\caption{\textbf{Parameters defined for the calculation of reaction propensity functions.}} \label{tab:reaction_params}
-	\centering
-	\begin{tabular}{|lrl|}
-		\hline
-		Parameter & Symbol & Definition \\ \hline \hline
-		Transcription rate & $\wpr{G}$ & $\in N(100, 20),\ \geq 10$ \\
-		Splicing rate & $\wsr G$ & $\in N(10, 2),\ \geq 2$ \\
-		mRNA degradation rate & $\xdr{G}$ & $\in N(5, 1),\ \geq 2$ \\
-		Translation rate & $\ypr{G}$ & $\in N(5, 1),\ \geq 2$ \\
-		Protein degradation rate & $\ydr G$ & $\in N(3, 0.5),\ \geq 1$ \\
-		Interaction strength & $\str{G,H}$ & $\in 10^{U(0, 2)}$ * \\ 
-		Interaction cooperativity & $\co{G,H}$ & $\in U(0.5, 2)$ * \\
-    Independence factor & $\ind G$ & $\in [0, 1]$ * \\ \hline\hline
-		TF concentration at half-maximal binding & $\hmy H$ & $= 0.5 \times \frac{\wpr H \times \ypr H}{\xdr H \times \ydr H}$ \\ 
-		Regulation activity & $\reg{G,H}$ & $= \left(\str{G,H} \times \frac{\y H}{\hmy H}\right) ^ {\co{G,H}}$ \\
-		Basal expression & $\ba G$ & $= \begin{cases} 1 & \mbox{if } \rp{G} = \emptyset \\ 0.0001 & \mbox{if } \rn{G} = \emptyset \mbox{ and } \rp{G} \neq \emptyset \\ 0.5 & \mbox{otherwise} \end{cases}$ * \\ \hline
-		\multicolumn{3}{l}{*: unless already defined when $G$ is a TF.}
-	\end{tabular}
-\end{table}
-
-% todo ask Wouter, why not simply the following?
-% ba_G = ifelse(R^+_G is not empty, 1, 0)
-
-\subsection{Compute average expression along backbone transitions} \label{sec:dyngen_sim_backbone}
-
-% → basically an ODE
-When simulating the developmental backbone, we go through the edges of the backbone state network defined in an earlier step (Section \ref{sec:backbone}), starting from the root state.
-It is assumed the root state has no modules active and has no expression of any molecules.
-To get to next state, we follow a transition starting from the root state, activate and deactivate the modules as indicated by the transition, and compute the average molecule abundance along the transition.
-To compute the average abundance, we perform small time steps $t = 0.001$ and let each reaction (Section \ref{sec:reactions}) occur $t$ times its propensity.
-
-\subsection{Simulate single cells} \label{sec:dyngen_ssa}
-%TODO: STOCKS \cite{kierzek_stocksstochastickinetic_2002} uses GSSA to simulate something.
-% maybe we could have a look at that! 
-% or https://nyaspubs.onlinelibrary.wiley.com/doi/full/10.1111/j.1749-6632.2008.03756.x
-
-% or even https://academic.oup.com/bioinformatics/article/24/10/1318/178687 ('AND' and 'OR')
-% and https://academic.oup.com/bioinformatics/article/25/9/1205/203870 (regulation motifs, kinetics)
-% https://nyaspubs.onlinelibrary.wiley.com/doi/full/10.1111/j.1749-6632.2008.03756.x (modules of genes)
-
-%TODO: improve this section a lot. Basically, explain SSA and GillespieSSA2.
-
-dyngen uses Gillespie's Stochastic Simulation Algorithm (SSA) to simulate dynamic processes. An SSA simulation is an iterative process where at each iteration one reaction is triggered.
-
-Each reaction consists of its propensity -- a formula to calculate the probability of occurring during an infinitesimal time interval -- and the effect -- how it will affect the current state if triggered.
-Each time a reaction is triggered, the simulation time is incremented by $\tau = \frac{1}{\sum_j prop_j} \ln\left(\frac{1}{r}\right)$, with $r \in U(0, 1)$ and $prop_j$ the propensity value of the $j$th reaction for the current state of the simulation.
-
-SSA simulations are notoriously slow.
-We use GillespieSSA2 which contains many optimisations such as translating and compiling all the propensity functions to C++ and implementations of approximations of SSA which allows to trigger many reactions simultaneously at each iteration.
-
-The framework allows to store the abundance levels of molecules only after a specific interval has passed since the previous census.
-By setting the census interval to 0, the whole simulation's trajectory is retained but many of these time points will contain very similar information.
-In addition to the abundance levels, also the propensity values and the number of firings of each of the reactions at each of the time steps can be retained, as well as specific sub-calculations of the propensity values, such as the regulator activity level $reg_{G,H}$.
-
-\subsubsection{Map SSA simulations to backbone}
-The cellular state of each timepoint in the SSA simulation is mapped to the state network of the backbone by calculating the 1NN between a state vector in the simulation and the average expression levels along transitions.
-
-\subsection{Simulate experiment}
-From the SSA simulation we obtain the abundance levels of all the molecules at the different time points.
-We need to replicate technical effects introduced by experimental protocols in order to obtain data that is similar to real data.
-For this, the cells are sampled from the simulations, and molecules are sampled for each of the cells.
-Real datasets are used in order to achieve similar data characteristics.
-
-\subsubsection{Sample cells}
-Cells can be sampled from an unsynchronised population of single cells (snapshot) or at multiple time points in a synchronised population (time series).
-
-\paragraph{Snapshot} Cells are just sampled randomly from the different time points in the simulation.
-
-\paragraph{Time series} The timeline of the simulations is cut up into chunks. From several of these chunks, cells are sampled. For each cell it is known at which time point it was sampled.
-
-\subsubsection{Sample molecules}
-
-\begin{itemize}
-	\item From real dataset, look at the number of transcripts that was captured per cell. Library size $ls_i$ of cell $i$ is samples from this distribution.
-	\item Capture rate of each molecule type $j$ is drawn from $cr_j \in N(1, 0.05)$
-	\item For each cell $i$, a multinomial distribution is used to draw $ls_i$ molecules from molecule type $j$ with probability $cr_j \times ab_{i,j}$ with $ab_{i,j}$ the molecule abundance level of molecule $j$ in cell $i$.
-\end{itemize}
-
-\subsection{Example runs of predefined backbones}
-
-\subsubsection{Linear}
-\subsubsection{Bifurcating}
-\subsubsection{Cycle}
-\subsubsection{Branching} (and binary tree, consecutive bifurcating, trifurcating)
-\subsubsection{Converging}
-\subsubsection{Bifurcating converging}
-\subsubsection{Bifurcating cycle}
-\subsubsection{Bifurcating loop}
-\subsubsection{Disconnected}
+\subsection{Extensions}
+\subsubsection{Predefined backbones}
+\subsubsection{Backbone lego}
+\subsubsection{Time series experiment}
+\subsubsection{Perturbation experiment}
+\subsubsection{Batch effects}
 
@@
 \subsubsection{Trajectory alignment}
-From discussion: Adding batch effects to snapshot simulations of linear (or even branching) trajectories allows evaluating trajectory alignment methods -- which attempt to map two or more trajectories onto each other.
+From discussion: Adding batch effects to snapshot simulations of linear (or even branching) trajectories allows to evaluate trajectory alignment methods -- which attempt to map two or more trajectories onto each other.
 \subsubsection{Differential network inference}
-From discussion: Adding perturbations to the GRN allows evaluating the performance of differential network inference methods -- which predict differential regulatory interactions between two or more groups of profiles.
+From discussion: Adding perturbations to the GRN allows to evaluate the performance of differential network inference methods -- which predict differential regulatory interactions between two or more groups of profiles.
 \subsubsection{RNA velocity}
-From discussion: Sampling a cell at a certain time point and once more at a later time point allows evaluating the performance of RNA velocity approaches -- which predict the future state of a cell by looking at differences in pre-mRNA and mRNA abundance levels.
-
-\subsubsection{Perturbation experiment}
+From discussion: Sampling a cell at a certain time point and once more at a later time point allows to evaluate the performance of RNA velocity approaches -- which predict the future state of a cell by looking at differences in pre-mRNA and mRNA abundance levels.
 

### 2019-09-09 content.tex

--- 2019-09-03 content.tex
@@
 A few generators of scRNA-seq profiles have already been developed (e.g.
-splatter \cite{zappia_splattersimulationsinglecell_2017}, powsimR \cite{vieth_powsimrpoweranalysis_2017} and PROSSTT \cite{papadopoulos_prossttprobabilisticsimulation_2018}).
-% BoolODE and dyntoy are also examples.
+splatter \cite{zappia_splattersimulationsinglecell_2017}, powsimR \cite{vieth_powsimrpoweranalysis_2017}, PROSSTT \cite{papadopoulos_prossttprobabilisticsimulation_2018} and SymSim \cite{zhang_simulatingmultiplefaceted_2019}).
 These can be used to evaluate the performance of computational tools, and to explore their strengths and weaknesses.
@@
 
-We introduce dyngen, a multi-modality simulator of single cells (Figure \ref{fig:dyngen}).
+We introduce dyngen, a multi-modality simulator of single cells and their dynamics (Figure \ref{fig:dyngen}).
 dyngen was initially developed as part of a comprehensive benchmark of TI methods \cite{saelens_comparisonsinglecelltrajectory_2019} but has since been extended to be applicable in a much broader context.
+We demonstrate its flexibility by simulating three different types of biological experiments, and using these simulations to develop new benchmarking techniques for computational tools.
 
-Inspiration was drawn from GeneNetWeaver's \cite{schaffter_genenetweaversilicobenchmark_2011} workflow to generate \textit{in silico} bulk profiles for evaluating network inference methods \cite{marbach_wisdomcrowdsrobust_2012}.
-GeneNetWeaver translates a known gene regulatory network (GRN) into sets of Ordinary Differential Equations (ODE) in order to simulate \ldots.
-% todo: describe GNW in this one sentance
-There are two significant differences in dyngen in comparison to GeneNetWeaver. 
-dyngen simulates a single-cell instead of a bulk profile.
-However, taking random walks through ODE systems in which key molecules occur with low abundance can result in inaccurate results.
-%rephrase and cite relevant work
-Gillespie's Stochastic Simulation Algorithm (SSA)  \cite{gillespie_exactstochasticsimulation_1977} allows to simulate low abundance systems.
-Running a simulation with SSA is an iterative process where at each iteration one reaction is allowed to take place.
-Each reaction consists of its propensity -- a formula to calculate the probability of occurring during an infinitesimal time interval -- and the effect -- how it will affect the current state if triggered.
-Secondly, the GRN is constructed such that it mimics a dynamic process of interest, such as cell differentiation into multiple cell types.
+Our simulator draws inspiration from a simulator for bulk transcriptomics data, GeneNetWeaver \cite{schaffter_genenetweaversilicobenchmark_2011,marbach_wisdomcrowdsrobust_2012}, but with two key improvements to make it work for single-cell data.
+First, instead of simulating continuous systems using stochastic differential equations (SDE), we simulate individual molecules and their reactions using derivatives of Gillespie's Stochastic Simulation Algorithm (SSA) \cite{gillespie_exactstochasticsimulation_1977}.
+Such a simulation better captures the stochasticity when low number of molecules are present  \cite{gillespie_exactstochasticsimulation_1977}, as is the case in single-cells \cite{zenklusen_singlernacountingreveals_2008a}.
+Processes such as transcriptional bursting may be difficult to represent in a continuous model, but develop naturally in single-molecule simulations \cite{schuh_genenetworkstranscriptional_2019}.
+A second improvement is that we include several ways to construct a GRN such that it mimics a dynamic process of interest, such as cell differentiation into multiple cell types.
 
@@
 	\includegraphics[width=\linewidth]{fig/showcase_3.png} 
-	\caption{Showcase of dyngen functionality. \textbf{TODO: change to pdf.} Remove B?} % TODO: update label
+	\caption{Showcase of dyngen functionality. \textbf{TODO: change to pdf.} Remove B? \textit{Yes please}} % TODO: update label
 	\label{fig:dyngen}
@@
 By default, dyngen supports snapshot experiments (uniformly sampling from an asynchronous dynamic process) and time-series experiments (sampling cells from different intervals in the simulation).
-It is possible to implement other experimental protocols (which perhaps do not exist in real life), such as sampling the same cell at regular intervals.
+It is possible to implement other experimental protocols, such as sampling the same cell at regular intervals.
 

### 2019-09-09 content.tex
convert showcase
fix references

@@
 First, instead of simulating continuous systems using stochastic differential equations (SDE), we simulate individual molecules and their reactions using derivatives of Gillespie's Stochastic Simulation Algorithm (SSA) \cite{gillespie_exactstochasticsimulation_1977}.
-Such a simulation better captures the stochasticity when low number of molecules are present  \cite{gillespie_exactstochasticsimulation_1977}, as is the case in single-cells \cite{zenklusen_singlernacountingreveals_2008a}.
+Such a simulation better captures the stochasticity when low number of molecules are present  \cite{gillespie_exactstochasticsimulation_1977}, as is the case in single-cells \cite{zenklusen_singlernacountingreveals_2008}.
 Processes such as transcriptional bursting may be difficult to represent in a continuous model, but develop naturally in single-molecule simulations \cite{schuh_genenetworkstranscriptional_2019}.
@@
 	\centering
-	\includegraphics[width=\linewidth]{fig/showcase_3.png} 
-	\caption{Showcase of dyngen functionality. \textbf{TODO: change to pdf.} Remove B? \textit{Yes please}} % TODO: update label
+	\includegraphics[width=\linewidth]{fig/showcase_4} 
+	\caption{Showcase of dyngen functionality. Remove B? \textit{Yes please}} % TODO: update label
 	\label{fig:dyngen}

### 2019-09-18 content.tex
update

@@
 
+\comment{GRN is defined nowhere}
+
 In dyngen, a cell consists of a set of molecules, the abundance of which are affected by a set of reactions: transcription, splicing, translation, and degradation (Figure \ref{fig:simplecyclic}A).
@@
 Multiple simulations with slightly different GRNs can emulate rewiring events in disease or perturbation experiments.
-% could use figure; one with GRNs of different topologies, another with rewiring events
 Multiple simulations with some slight perturbations can be used to replicate batch effects.
+
+\comment{could use figure; one with GRNs of different topologies, another with rewiring events}
 
@@
 
+\comment{explain what is meant by 'sampling' and 'profiling technique' a little better}
+
 The final main advantage is that by making alterations to the simulation pipeline, multiple types of experiments (sampling technique or profiling technique) can be simulated.
@@
 
-%% TODO: show that dyngen output resembles real data using e.g. countsimQC?
+\comment{show that dyngen output resembles real data using e.g. countsimQC?}
 
-%% TODO: add result figures pertaining use cases
+\comment{add result figures pertaining use cases}
 
@@
 Allowing cells to receive and react to environmental and intercellular stimuli would enable simulating essential cellular processes such as cell division and migration.
-%¡also adding more types of molecules, e.g. protein complex, small rnas, PPI
+
+\comment{also adding more types of molecules, e.g. protein complex, small rnas, PPI}
 \section{Methods}
-NOTE: The method section is really rough at this stage.
+\comment{The method section is REALLY rough at this stage.}
 
@@
 
-TODO: mention strength and cooperativity.
+\comment{mention strength and cooperativity.}
 
@@
 \subsection{Generate gene regulatory network}
-% TODO: write section in words instead of bullet point list
+\comment{Welcome to the beginning of the end}
 
@@
 \subsection{Convert gene regulatory network to a set of reactions} \label{sec:reactions}
-% TODO: write section in words instead of bullet point list
-
-Each reaction consists of its propensity -- a formula to calculate the probability of occurring during an infinitesimal time interval -- and the effect -- how it will affect the current state if triggered.
+Each reaction consists of its propensity -- a formula to calculate
+the probability of the reaction occurring during an infinitesimal
+time interval -- and the effect -- how it will
+ affect the current state if triggered.
 
@@
 
-% TODO: explain derivation of transcription formula? 
+\comment{TODO: add step-by-step  derivation of transcription formula}
 % regG,H represents a 'probability' that H is bound to G (though does not yet scale between [0,1]).
@@
 
-% todo ask Wouter, why not simply the following?
-% ba_G = ifelse(R^+_G is not empty, 1, 0)
-
 \subsection{Compute average expression along backbone transitions} \label{sec:dyngen_sim_backbone}
 
-% → basically an ODE
+\comment{→ basically an ODE}
 When simulating the developmental backbone, we go through the edges of the backbone state network defined in an earlier step (Section \ref{sec:backbone}), starting from the root state.
@@
 
-Each reaction consists of its propensity -- a formula to calculate the probability of occurring during an infinitesimal time interval -- and the effect -- how it will affect the current state if triggered.
+Each reaction consists of its propensity -- a formula to calculate the probability of the reaction occurring during an infinitesimal time interval -- and the effect -- how it will affect the current state if triggered.
 Each time a reaction is triggered, the simulation time is incremented by $\tau = \frac{1}{\sum_j prop_j} \ln\left(\frac{1}{r}\right)$, with $r \in U(0, 1)$ and $prop_j$ the propensity value of the $j$th reaction for the current state of the simulation.

### 2019-09-19 content.tex
update

@@
+% DYNGEN CONTENT
+
 \begin{quote}
@@
 
-\comment{GRN is defined nowhere}
+\mycomment{GRN is defined nowhere}
 
@@
 
-\comment{could use figure; one with GRNs of different topologies, another with rewiring events}
+\mycomment{could use figure; one with GRNs of different topologies, another with rewiring events}
 
@@
 
-\comment{explain what is meant by 'sampling' and 'profiling technique' a little better}
+\mycomment{explain what is meant by 'sampling' and 'profiling technique' a little better}
 
@@
 
-\comment{show that dyngen output resembles real data using e.g. countsimQC?}
+\mycomment{show that dyngen output resembles real data using e.g. countsimQC?}
 
-\comment{add result figures pertaining use cases}
+\mycomment{add result figures pertaining use cases}
 
@@
 
-\comment{also adding more types of molecules, e.g. protein complex, small rnas, PPI}
+\mycomment{also adding more types of molecules, e.g. protein complex, small rnas, PPI}
 \section{Methods}
-\comment{The method section is REALLY rough at this stage.}
+\mycomment{The method section is REALLY rough at this stage.}
 
@@
 
-\comment{mention strength and cooperativity.}
+\mycomment{mention strength and cooperativity.}
 
@@
 \subsection{Generate gene regulatory network}
-\comment{Welcome to the beginning of the end}
+\mycomment{Welcome to the beginning of the end}
 
@@
 
-\comment{TODO: add step-by-step  derivation of transcription formula}
+\mycomment{TODO: add step-by-step  derivation of transcription formula}
 % regG,H represents a 'probability' that H is bound to G (though does not yet scale between [0,1]).
@@
 
-\comment{→ basically an ODE}
+\mycomment{→ basically an ODE}
 When simulating the developmental backbone, we go through the edges of the backbone state network defined in an earlier step (Section \ref{sec:backbone}), starting from the root state.

### 2019-09-21 content.tex
update pdf

@@
 
-\mycomment{→ basically an ODE}
+\mycomment{this is basically an ODE, though we never reference ODEs anywhere. Maybe this is ok, this simplifies the section a bit?}
 When simulating the developmental backbone, we go through the edges of the backbone state network defined in an earlier step (Section \ref{sec:backbone}), starting from the root state.

### 2019-09-21 manuscript.tex
update
update pdf

@@
-\documentclass[11pt,hidelinks,a4paper]{article}
+\documentclass[11pt,a4paper]{article}
+
+% dyngen
 
@@
 %%%%%%%%%%%%%%%%%%%%%%%
+
+\usepackage[margin=2.5cm,a4paper]{geometry}
 
@@
 \usepackage{microtype} % Sublim­i­nal re­fine­ments to­wards ty­po­graph­i­cal per­fec­tion
-
-% used for defining abbreviations, see nomenclature.tex
-\usepackage{nomencl}
 
@@
 \usepackage{bibentry}
-\nobibliography*
 
@@
 
-% fix named refs for starred sections
-\makeatletter
-\def\ttl@useclass#1#2{%
-	\@ifstar
-	{\ttl@labeltrue\@dblarg{#1{#2}}}% {\ttl@labelfalse#1{#2}[]}%
-	{\ttl@labeltrue\@dblarg{#1{#2}}}}
-\makeatother
+% authors and affiliations
+\usepackage{authblk}
+
+% command for adding comments
+\newcommand{\mycomment}[1]{\noindent\fbox{\parbox{\textwidth}{\color{red}#1}}}
 
@@
 
-% use Sonny chapter style
-\usepackage[Sonny]{fncychap} %Sonny
-
-% set font after chapter sonny style
 \usepackage{fontspec}
@@
 %% hyperlinks
-% load hyperref after fncychap
 \usepackage{hyperref}
 
-\renewcommand{\appendixname}{}
 \renewcommand{\baselinestretch}{1.25}
@@
 
-\include{content}
+\title{dyngen: simulating developing single cells}
+
+\newcommand{\afdambi}{1}
+\newcommand{\afappmat}{2}
+\newcommand{\afcmgg}{3}
+\newcommand{\afcrig}{4}
+
+\author[\afdambi,\afappmat,\afcmgg,\afcrig]{Robrecht~Cannoodt}
+\author[\afdambi,\afappmat]{Wouter~Saelens}
+\author[\afdambi,\afappmat,*]{Yvan~Saeys}
+
+\affil[\afdambi]{{\footnotesize Data Mining and Modelling for Biomedicine group, VIB Center for Inflammation Research, Ghent, Belgium}}
+\affil[\afappmat]{{\footnotesize Department of Applied Mathematics, Computer Science and Statistics, Ghent University, Ghent, Belgium}}
+\affil[\afcmgg]{{\footnotesize Center of Medical Genetics, Ghent University Hospital, Ghent, Belgium}}
+\affil[\afcrig]{{\footnotesize Cancer Research Institute Ghent (CRIG), Ghent, Belgium}}
+\affil[*]{{\footnotesize Corresponding author, \href{mailto:yvan.saeys@ugent.be}{yvan.saeys@ugent.be}}}
+\date{}
+
+\maketitle
+
+\input{content}
+
+\appendix
+
+\printbibliography
 

### 2019-10-08 content.tex
add bib
update

--- 2019-09-19 content.tex
@@
 \section{Introduction}
-Continuous technological advancements to high-throughput profiling of single cells
-are having profound effects on how researchers can validate biological hypotheses. 
-For example, single-cell RNA sequencing (scRNA-seq) directly resulted in the development
-of a new type of computational method called trajectory inference (TI). By profiling
-the transcriptomics profiles of developing cells, TI methods attempt to reconstruct 
-and characterise the underlying dynamic processes \cite{cannoodt_computationalmethodstrajectory_2016}.
-While early experimental technologies allowed to profile one single modality (e.g. DNA sequence, 
-RNA or protein expression), recent developments permit profiling multiple modalities simultaneously.
+Continuous technological advancements to single-cells omics are having profound 
+effects on how researchers can validate biological hypotheses. 
+Early experimental technologies typically only allowed profiling a single modality (e.g. DNA sequence, 
+RNA or protein expression). However, recent developments permit profiling multiple modalities simultaneously,
+and every modality added allows for new types of analyses that can be performed.
+
+This presents method developers with a problem.
+The majority of the 250+ peer-reviewed computational tools for analysing single cell omics data were published without a quantitative assessment of the accuracy of the tool.
+This is partially due to low availability of suitable benchmarking datasets; even if there are sufficient suitable input datasets available, these are often not accompanied by the necessary metadata to serve as ground-truth for a benchmark.
+
+Here, synthetic data plays a crucial role in asserting minimum performance requirements for novel tools in anticipation of adequate real data.
+Generators of scRNA-seq data (e.g.
+splatter \cite{zappia_splattersimulationsinglecell_2017}, powsimR \cite{vieth_powsimrpoweranalysis_2017}, PROSSTT \cite{papadopoulos_prossttprobabilisticsimulation_2018} and SymSim \cite{zhang_simulatingmultiplefaceted_2019}) have already been widely used to explore the strengths and weakness of computational tools, both by method developers\cite{street_slingshotcelllineage_2018,parra_reconstructingcomplexlineage_2018,lummertzdarocha_reconstructioncomplexsinglecell_2018,lin_scclassifyhierarchicalclassification_2019} and independent benchmarkers\cite{duo_systematicperformanceevaluation_2018,saelens_comparisonsinglecelltrajectory_2019,soneson_biasrobustnessscalability_2018}.
+However, a limitation of scRNA-seq profiles generators is that they would require significant methodological alterations to add additional modalities.
 
@@
 molecular states, spatial positions and environmental interactions \cite{stuart_integrativesinglecellanalysis_2019}.
-While this falls outside the reach of current experimental technologies, \textit{in silico} simulations
-of single cells would allow developing the next wave of computational techniques
-in anticipation of new experimental technologies.
+While this falls outside the reach of current experimental technologies, generating synthetic data in anticipation of new experimental technologies would allow already developing the next wave of computational tools.
 
-A few generators of scRNA-seq profiles have already been developed (e.g.
-splatter \cite{zappia_splattersimulationsinglecell_2017}, powsimR \cite{vieth_powsimrpoweranalysis_2017}, PROSSTT \cite{papadopoulos_prossttprobabilisticsimulation_2018} and SymSim \cite{zhang_simulatingmultiplefaceted_2019}).
-These can be used to evaluate the performance of computational tools, and to explore their strengths and weaknesses.
-A limitation of directly simulating a scRNA-seq profile (instead of a single cell) is that extending the simulation to other aspects of the cell -- such as tracking the full history of molecular states -- becomes difficult.
+We introduce \texttt{dyngen}, a multi-modality simulator of single cells and their dynamics (Figure \ref{fig:dyngen}A).
+A cell is simulated using Gillespie's Stochastic Simulation Algorithm (SSA) \cite{gillespie_exactstochasticsimulation_1977} where a cell consists of a set of molecules.
+Throughout a simulation, reactions (e.g.
+transcription, splicing) modify the abundance of these molecules, and the likelihood of a reaction occurring is again dependent on molecule abundance.
 
-We introduce dyngen, a multi-modality simulator of single cells and their dynamics (Figure \ref{fig:dyngen}).
-dyngen was initially developed as part of a comprehensive benchmark of TI methods \cite{saelens_comparisonsinglecelltrajectory_2019} but has since been extended to be applicable in a much broader context.
-We demonstrate its flexibility by simulating three different types of biological experiments, and using these simulations to develop new benchmarking techniques for computational tools.
-
-Our simulator draws inspiration from a simulator for bulk transcriptomics data, GeneNetWeaver \cite{schaffter_genenetweaversilicobenchmark_2011,marbach_wisdomcrowdsrobust_2012}, but with two key improvements to make it work for single-cell data.
-First, instead of simulating continuous systems using stochastic differential equations (SDE), we simulate individual molecules and their reactions using derivatives of Gillespie's Stochastic Simulation Algorithm (SSA) \cite{gillespie_exactstochasticsimulation_1977}.
-Such a simulation better captures the stochasticity when low number of molecules are present  \cite{gillespie_exactstochasticsimulation_1977}, as is the case in single-cells \cite{zenklusen_singlernacountingreveals_2008}.
-Processes such as transcriptional bursting may be difficult to represent in a continuous model, but develop naturally in single-molecule simulations \cite{schuh_genenetworkstranscriptional_2019}.
-A second improvement is that we include several ways to construct a GRN such that it mimics a dynamic process of interest, such as cell differentiation into multiple cell types.
-
-We demonstrate dyngen's flexibility by simulating numerous different types of biological experiments, and using these simulations to develop new benchmarking techniques for computational tools.
+By simulating a cell over time in terms of its molecular state and the reactions that are allowed to occur, the simulator is more extendible to new modalities or experimental procedures (Figure~\ref{fig:dyngen}B).
+We demonstrate \texttt{dyngen}'s flexibility by simulating three different types of biological experiments, and using these simulations to develop benchmarking methodologies for recently developed computational tools.
 
@@
 	\includegraphics[width=\linewidth]{fig/showcase_4} 
-	\caption{Showcase of dyngen functionality. Remove B? \textit{Yes please}} % TODO: update label
+	\caption{
+		\textbf{Showcase of dyngen functionality.} 
+		\textbf{A:} 
+		\textbf{B:} Evaluating different types of computational tools requires simulating different types of experiments and extracting different layers of information from the simulation.
+	}
 	\label{fig:dyngen}
@@
 \section{Results}
-dyngen simulates the transcriptomic changes of a cell over time using a model of gene regulation.
-Throughout this section, a simple simulation of a cell undergoing a cyclic process is used to illustrate key strengths of dyngen (Figure \ref{fig:simplecyclic}).
-This example only comprises of a single cell containing 5 genes, but dyngen can easily scale up to thousands of simulations containing thousands of genes.
+
+A cell consists of a set of molecules, the abundance of which are affected by a set of reactions: transcription, splicing, translation, and degradation (Figure \ref{fig:simplecyclic}A). 
+A gene regulatory network (GRN) define the reactions that are allowed to occur (Figure \ref{fig:simplecyclic}B), and is constructed such that a cell will develop over time (Figure \ref{fig:simplecyclic}C--D).
+The likelihood of a reaction occurring is a function of the abundance of key molecules involved in each reaction (Figure \ref{fig:simplecyclic}E).
 
@@
 	\includegraphics[width=\linewidth]{fig/simplecyclic_edited} 
-	\caption{Showcase of dyngen functionality.
-A time resolution of 0.1 was used, but this can be increased or decreased without effect on performance of the execution of the simulation.
-\textbf{TODO: perhaps it's better to replace Figure \ref{fig:simplecyclic} with one subfigure for each of the paragraphs in this text.}}
+	\caption{
+		\textbf{Key features of \texttt{dyngen} are illustrated using a cyclic toy example.}
+		While this example comprises only of a single cell and 5 genes, \texttt{dyngen} is able to simulate thousands of cells with GRNs containing thousands of genes.
+	}
 	\label{fig:simplecyclic}
@@
 
-\mycomment{GRN is defined nowhere}
-
-In dyngen, a cell consists of a set of molecules, the abundance of which are affected by a set of reactions: transcription, splicing, translation, and degradation (Figure \ref{fig:simplecyclic}A).
-These reactions are determined from a predefined set of gene regulatory interactions (Figure \ref{fig:simplecyclic}B), henceforth referred to as a gene regulatory network (GRN).
-The likelihood of a reaction occurring at any given point in time is defined by the GRN and by the abundance of molecules involved each reaction.
-
-One of dyngen's main advantages is that through careful engineering of the GRN, different cellular developmental processes can be simulated.
+One of dyngen's main advantages is that depending on how the GRN is designed, different cellular developmental processes can be simulated.
 Different GRNs can result in branching, converging, cyclic, or even disconnected developmental topologies.
-Multiple simulations with slightly different GRNs can emulate rewiring events in disease or perturbation experiments.
-Multiple simulations with some slight perturbations can be used to replicate batch effects.
+Multiple simulations with slightly different GRNs can emulate rewiring events in disease or batch effects.
 
@@
 
-\mycomment{explain what is meant by 'sampling' and 'profiling technique' a little better}
-
-The final main advantage is that by making alterations to the simulation pipeline, multiple types of experiments (sampling technique or profiling technique) can be simulated.
+The final main advantage is that \texttt{dyngen} can be used to simulate different experimental conditions.
 By default, dyngen supports snapshot experiments (uniformly sampling from an asynchronous dynamic process) and time-series experiments (sampling cells from different intervals in the simulation).
-It is possible to implement other experimental protocols, such as sampling the same cell at regular intervals.
-
-\mycomment{show that dyngen output resembles real data using e.g. countsimQC?}
+It is possible to implement other experimental protocols, such as sampling the same cell at regular intervals or performing perturbation experiments.
 
@@
 
-\mycomment{also adding more types of molecules, e.g. protein complex, small rnas, PPI}
+%\mycomment{also adding more types of molecules, e.g. protein complex, small rnas, PPI}
+
 \section{Methods}
-\mycomment{The method section is REALLY rough at this stage.}
-
 The workflow to generate \textit{in silico} single cell data consists of six main steps (Figure \ref{fig:explain_methods}).
@@
 
-\mycomment{mention strength and cooperativity.}
+\mycomment{mention strength, cooperativity, effect and independence factor.}
 
@@
 
-\subsection{Generate gene regulatory network}
-\mycomment{Welcome to the beginning of the end}
+\subsection{Generating the gene regulatory network}
+The GRN is generated based on the given backbome in four main steps (Figure~\ref{fig:gen_feature_network}).
 
-\subsubsection{Generate the transcription factor network}
-Parameters: 
-\begin{itemize}
-	\item Number of TFs to generate $num\_tfs$
-	\item Minimum TFs per module $min\_tfs$, default = 1
-	\item Number of regulatory interactions per module interaction $num\_tf\_ints$, default = 2
-\end{itemize}
+\begin{figure}[htb!]
+	\centering
+	\includegraphics[width=\linewidth]{fig/gen_feature_network} 
+	\caption{
+		\textbf{Generating the feature network from a backbone consists of four main steps.}
+	}
+	\label{fig:gen_feature_network}
+\end{figure}
 
-Procedure: 
-\begin{itemize}
-	\item Divide $num\_tfs$ TFs amongst modules in backbone such that each module has at least $min\_tfs$ TFs.
-	\item For every TF in a particular module $M$, connect that TF with $num\_tf\_ints$ TFs for each upstream module of $M$.
-	\item The strength and cooperativity of interactions created this way are defined by the module network.
-\end{itemize}
+\paragraph{Step 1, sampling the transcription factors (TF).} The TFs are the main drivers of the molecular changes in the simulation.
+The user provides a backbone and the number of TFs to generate.
+Each TF is assigned to a module such that each module has at least $x$ parameters (default $x=1$).
+A TF inherits the 'burn' and 'basal expression' from the module in belongs to.
 
-\subsubsection{Generate targets}
-Parameters: 
-\begin{itemize}
-	\item Number of target genes $num\_targets$
-	\item Name of a FANTOM5 GRN
-	\item Damping factor $damping$, default = 0.05
-	\item Target resampling $resamp$, default = $+\infty$
-	\item Maximum in-degree $max\_in\_degree$, default = 5
-\end{itemize}
+\paragraph{Step 2, generating the TF interactions.} Let each TF be regulated according to the interactions in the backbone.
+These interactions inherit the effect, strength, and cooperativity parameters from the interactions in the backbone.
+A TF can only be regulated by other TFs or itself.
 
-Procedure:
-\begin{itemize}
-	\item Download the FANTOM5 GRN if not already available
-	\item Randomly map TFs to the regulators in the GRN
-	\item Compute page rank from the selected regulators in the GRN with damping factor $damping$
-	\item Perform weighted sample of  $\min(num\_targets, resamp)$ targets weighted by the page rank
-	\item Select subgraph induced by the TFs and the sampled targets in the GRN, remove edges to the TFs
-	\item Add subgraph to current TF-target network
-	\item If less than $num\_targets$ have been sampled in this way, go back to step 2.
-	\item Remove regulatory interactions if a target has more than $max\_in\_degree$ edges.
-\end{itemize}
+\paragraph{Step 3, sampling the target subnetwork.} 
+A user-defined number of target genes are added to the GRN.
+Target genes regulated by a TF or another target gene, but is always downstream of at least one TF.
+To sample the interactions between target genes, one of the many FANTOM5\cite{lizio_gatewaysfantom5promoter_2015} GRNs is sampled.
+The currently existing TFs are mapped to regulators in the FANTOM5 GRN.
+The targets are drawn from the FANTOM5 GRN, weighted by their page rank value.
+For each target, at most $x$ regulators are sampled from the induced FANTOM5 GRN (default $x=5$).
+The interactions connecting a target gene and its regulators are added the GRN.
 
-\subsubsection{Generate housekeeping genes}
-Parameters: 
-\begin{itemize}
-	\item Number of housekeeping genes $num\_targets$
-	\item Name of a FANTOM5 GRN
-	\item Target resampling $resamp$, default = $+\infty$
-	\item Maximum in-degree $max\_in\_degree$, default = 5
-\end{itemize}
-
-Procedure:
-\begin{itemize}
-	\item Use same FANTOM5 GRN
-	\item Subsample GRN such that each gene has a maximum in-degree of $max\_in\_degree$
-	\item Perform breadth-first-search from a random gene in the GRN, select $\min(num\_targets, resamp)$ first genes encountered
-	\item Select subgraph induced by the sampled housekeeping genes in the GRN
-	\item Add subgraph to current TF-target network
-	\item If less than $num\_targets$ have been sampled in this way, go back to step 2.
-\end{itemize}
+\paragraph{Step 4, sampling the housekeeping subnetwork.}
+Housekeeping genes are completely separate from any TFs or target genes.
+A user-defined set of housekeeping genes are also sampled from the FANTOM5 GRN.
+The interactions of the FANTOM5 GRN are first subsampled such that the maximum in-degree of each gene is $x$ (default $x=5$).
+A random gene is sampled, and a breadth-first-search is performed to sample the desired number of housekeeping genes.
 
 \subsection{Convert gene regulatory network to a set of reactions} \label{sec:reactions}
-Each reaction consists of its propensity -- a formula to calculate
-the probability of the reaction occurring during an infinitesimal
-time interval -- and the effect -- how it will
- affect the current state if triggered.
-
 \newcommand{\w}[1]{\text{w}_{#1}}
@@
 
-We define the abundance levels of pre-mRNA, mRNA and protein of gene $G$ as $\w G$, $\x G$ and $\y G$ respectively.
-Five reactions affect the abundance levels of these molecules: transcription, splicing, mRNA degradation, translation, and protein degradation.
+For every gene $G$ in the GRN, the simulation will keep track of the abundance levels of three molecules, a pre-mRNA, a mature mRNA and a protein.
+These abundance levels are represented as $\w G$, $\x G$ and $\y G$ respectively.
+
+Throughout a simulation, the abundance levels of molecules are affected by five reactions, namely transcription, splicing, mRNA degradation, translation, and protein degradation.
+Each reaction consists of its propensity -- a formula to calculate the probability of the reaction occurring during an infinitesimal time interval -- and the effect -- how it will affect the current state if triggered.
 The effects and propensity functions of these reactions are defined in Table \ref{tab:reaction_def}.
 
-% see https://en.wikipedia.org/wiki/Hill_equation_(biochemistry)#Regulation_of_gene_transcription
-\begin{table}[h]
+\begin{table}[ht]
 	\caption{\textbf{Reactions affecting the abundance levels of pre-mRNA $\w G$, mRNA $\x G$ and proteins $\y G$ of gene $G$.} Define the set of regulators of $G$ as $\rs{G}$, the set of upregulating regulators of $G$ as $\rp G$, and the set of downregulating regulators of $G$ as $\rn G$.
@@
 		Reaction & Effect & Propensity \\ \hline \hline
-		Transcription & $\emptyset \rightarrow \w G$ & $\wpr G \times \frac{\ba G - \ind{G}^{|\rp{G}|} + \prod\limits_{H \in \rp{G}}(\ind G + \reg{G,H})}{\prod\limits_{H \in \rp{G}}(1 + \reg{G,H})}$ \\
+		Transcription & $\emptyset \rightarrow \w G$ & $\wpr G \times \frac{\ba G - \ind{G}^{|\rp{G}|} + \prod\limits_{H \in \rp{G}}(\ind G + \reg{G,H})}{\prod\limits_{H \in \rs{G}}(1 + \reg{G,H})}$ \\
 		Splicing & $\w G \rightarrow \x G$ & $\wsr G \times \w G$ \\
@@
 \end{table}
+%!!!!!!!!!!!!!!!!!!!!!!!!!!!!!!!!!!!!!!!!!!!!!!!!!!!!!!!!!!!!!!!
+% TODO: add step-by-step  derivation of transcription formula
+%!!!!!!!!!!!!!!!!!!!!!!!!!!!!!!!!!!!!!!!!!!!!!!!!!!!!!!!!!!!!!!!
 
-\mycomment{TODO: add step-by-step  derivation of transcription formula}
 % regG,H represents a 'probability' that H is bound to G (though does not yet scale between [0,1]).
@@
 
-\begin{table}[h]
+\begin{table}[ht]
 	\caption{\textbf{Parameters defined for the calculation of reaction propensity functions.}} \label{tab:reaction_params}
@@
 		Basal expression & $\ba G$ & $= \begin{cases} 1 & \mbox{if } \rp{G} = \emptyset \\ 0.0001 & \mbox{if } \rn{G} = \emptyset \mbox{ and } \rp{G} \neq \emptyset \\ 0.5 & \mbox{otherwise} \end{cases}$ * \\ \hline
-		\multicolumn{3}{l}{*: unless already defined when $G$ is a TF.}
+		\multicolumn{3}{l}{*: unless $G$ is a TF, then the value is determined by the backbone.}
 	\end{tabular}
@@
 \subsection{Compute average expression along backbone transitions} \label{sec:dyngen_sim_backbone}
-
-\mycomment{→ basically an ODE}
 When simulating the developmental backbone, we go through the edges of the backbone state network defined in an earlier step (Section \ref{sec:backbone}), starting from the root state.
@@
 
-dyngen uses Gillespie's Stochastic Simulation Algorithm (SSA) to simulate dynamic processes. An SSA simulation is an iterative process where at each iteration one reaction is triggered.
+\texttt{dyngen} uses Gillespie's Stochastic Simulation Algorithm (SSA) to simulate dynamic processes. An SSA simulation is an iterative process where at each iteration one reaction is triggered.
 
@@
 
-SSA simulations are notoriously slow.
-We use GillespieSSA2 which contains many optimisations such as translating and compiling all the propensity functions to C++ and implementations of approximations of SSA which allows to trigger many reactions simultaneously at each iteration.
-
-The framework allows to store the abundance levels of molecules only after a specific interval has passed since the previous census.
+\texttt{GillespieSSA2} is an optimised library for performing SSA simulations.
+The propensity functions are compiled to C++, and SSA approximations can be used which allow to trigger many reactions simultaneously at each iteration.
+The framework also allows to store the abundance levels of molecules only after a specific interval has passed since the previous census.
 By setting the census interval to 0, the whole simulation's trajectory is retained but many of these time points will contain very similar information.
@@
 
-\paragraph{Snapshot} Cells are just sampled randomly from the different time points in the simulation.
+\paragraph{Snapshot} Cells are sampled uniformly from the different time points in the simulation.
 
-\paragraph{Time series} The timeline of the simulations is cut up into chunks. From several of these chunks, cells are sampled. For each cell it is known at which time point it was sampled.
+\paragraph{Time series} The timeline of the simulations is divided into multiple chunks. From several of these chunks, cells are sampled. For each cell it is known at which time point it was sampled.
 
 \subsubsection{Sample molecules} 
-
-\begin{itemize}
-	\item From real dataset, look at the number of transcripts that was captured per cell. Library size $ls_i$ of cell $i$ is samples from this distribution.
-	\item Capture rate of each molecule type $j$ is drawn from $cr_j \in N(1, 0.05)$
-	\item For each cell $i$, a multinomial distribution is used to draw $ls_i$ molecules from molecule type $j$ with probability $cr_j \times ab_{i,j}$ with $ab_{i,j}$ the molecule abundance level of molecule $j$ in cell $i$.
-\end{itemize}
+Molecules are sampled from the simulation to replicate how molecules are experimentally sampled.
+A real dataset is downloaded from a repository of single-cell RNA-seq datasets\cite{cannoodt_singlecellomicsdatasets_2018}.
+For each \textit{in silico} cell $i$, draw its library size $ls_i$from the distribution of transcript counts per cell in the real dataset.
+The capture rate $cr_j$ of each \textit{in silico} molecule type $j$ is drawn from $N(1, 0.05)$.
+Finally, for each cell $i$, draw $ls_i$ molecules from the multinomial distribution with probabilities $cr_j \times ab_{i,j}$ with $ab_{i,j}$ the molecule abundance level of molecule $j$ in cell $i$.
 
@@
 \subsubsection{Linear}
+
+\begin{figure}[H]
+	\centering
+	\includegraphics[width=\linewidth]{fig/example/bb_linear/plot} 
+	\caption{}
+	\label{fig:example_linear}
+\end{figure}
+
 \subsubsection{Bifurcating}
+
+\begin{figure}[H]
+	\centering
+	\includegraphics[width=\linewidth]{fig/example/bb_bifurcating/plot} 
+	\caption{}
+	\label{fig:example_bifurcating}
+\end{figure}
+
 \subsubsection{Cycle}
-\subsubsection{Branching} (and binary tree, consecutive bifurcating, trifurcating)
+
+\begin{figure}[H]
+	\centering
+	\includegraphics[width=\linewidth]{fig/example/bb_cycle/plot} 
+	\caption{}
+	\label{fig:example_cycle}
+\end{figure}
+
+%\subsubsection{Branching}
+%
+%\begin{figure}[H]
+%	\centering
+%	\includegraphics[width=\linewidth]{fig/example/bb_binary_tree/plot} 
+%	\caption{}
+%	\label{fig:example_branching}
+%\end{figure}
+
 \subsubsection{Converging}
+
+\begin{figure}[H]
+	\centering
+	\includegraphics[width=\linewidth]{fig/example/bb_converging/plot} 
+	\caption{}
+	\label{fig:example_converging}
+\end{figure}
+
 \subsubsection{Bifurcating converging}
-\subsubsection{Bifurcating cycle}
+
+\begin{figure}[H]
+	\centering
+	\includegraphics[width=\linewidth]{fig/example/bb_bifurcating_converging/plot} 
+	\caption{}
+	\label{fig:example_bifurcating_converging}
+\end{figure}
+
+%\subsubsection{Bifurcating cycle}
+%
+%\begin{figure}[H]
+%	\centering
+%	\includegraphics[width=\linewidth]{fig/example/bb_bifurcating_cycle/plot} 
+%	\caption{}
+%	\label{fig:example_bifurcating_cycle}
+%\end{figure}
+
 \subsubsection{Bifurcating loop}
-\subsubsection{Disconnected}
+
+\begin{figure}[H]
+	\centering
+	\includegraphics[width=\linewidth]{fig/example/bb_bifurcating_loop/plot} 
+	\caption{}
+	\label{fig:example_bifurcating_loop}
+\end{figure}
+
+%\subsubsection{Disconnected}
+%
+%\begin{figure}[H]
+%	\centering
+%	\includegraphics[width=\linewidth]{fig/example/bb_disconnected/plot} 
+%	\caption{}
+%	\label{fig:example_disconnected}
+%\end{figure}
 
@@
 
-%dyntoy and PROSSTT are specifically developed to generate datasets
-% containing single cells which develop along a certain trajectory.
-% These generators start a certain topology in mind, 
-% simulate changing gene expressions along each of the branches of the topology, and sample cells
-% from random points in the trajectory. splatter is focused on well replicating the specific noise characteristics of
-% scRNA-seq data, and simulates differentially expressed genes in order to replicate clusters of cells, batch effects
-% and even trajectories. dyngen 1.0 borrows a page from GeneNetWeaver in that it also converts a GRN into an ODE, but
-% the GRN is constructed in such a way that the repeated simulations resemble single cells following a regulatory program
-% (e.g. differentiation into one of two celltypes). BoolODE uses the dyngen 1.0 GRNs but uses boolean models to convert 
-% the network into an ODE, and is used to evaluate the performance of NI methods, not TI methods.
-
-%\begin{table}
-%	\caption{\textbf{Comparison of existing single-cell omics simulators.} TI: trajectory inference, NI: network inference, Cl: clustering, DR: dimensionality reduction.} \label{tab:simulators}
-%	\centering\fontsize{9}{11}\selectfont
-%	\begin{tabularx}{\linewidth}{p{1.6cm}p{2.5cm}p{3cm}X}
-%		Generator & Used to evaluate & Data type(s) & Approach \\ \hline
-%		dyngen 1.0 & TI & mRNA \& protein & GRN, ODE, Euler--Maramuya \\
-%		dyntoy & TI & mRNA & Visualise tree in plane, generate expression in plane, sample cells from tree \\
-%		PROSSTT & TI & mRNA & Random walks along tree, sample cells from tree \\
-%		splatter & Cl \& TI & mRNA & Simulate DE genes, simulate scRNA-seq noise \\
-%		BoolODE & NI & mRNA \& protein & Use dyngen GRN's, ODE, Euler--Maramuya \\
-%		dyngen 2.0 & TI, NI, Cl, DR & pre-mRNA, mRNA, protein & Backbone, GRN, SSA \\
-%		dyngen 3? & TI, NI, Cl, DR & pre-mRNA, mRNA, miRNA, protein, protein complex & Backbone, GRN, SSA
-%	\end{tabularx}
-%\end{table}
+\clearpage
+\section{References}
+\printbibliography[heading=none]

### 2019-10-08 analysis/example_methods_overview.R

@@ -7,7 +7,7 @@ dir.create("fig/example_methods", showWarnings = FALSE, recursive = TRUE)
 
 backbone <- backbone(
   module_info = tribble(
-    ~module_id, ~a0, ~burn,
+    ~module_id, ~ba, ~burn,
     "A", 1, TRUE,
     "B", 0, FALSE,
     "C", 0, FALSE,
@@ -59,6 +59,7 @@ plot_feature_network(model, show_hks = TRUE)
 plot_feature_network(model)
 plot_feature_network(model, show_hks = TRUE) %>% ggsave(filename = "fig/example_methods/featnet.pdf", width = 8, height = 6)
 plot_feature_network(model, show_targets = FALSE) %>% ggsave(filename = "fig/example_methods/featnet_onlytfs.pdf", width = 8, height = 6)
+
 
 ## simulate
 model2 <- model %>%
@@ -195,5 +196,182 @@ g <- ggplot(as.data.frame(dimred)) + geom_point(aes(comp_1, comp_2)) + coord_equ
   theme(axis.text = element_blank(), axis.ticks = element_blank()) +
   labs(x = "Comp 1", y = "Comp 2")
 ggsave("fig/example_methods/dimred.pdf", g, width = 8, height = 6)
+
+
+
+
+
+set.seed(1)
+
+model2 <-
+  initialise_model(
+    num_tfs = 10,
+    num_targets = 10,
+    num_hks = 10,
+    num_cells = 100,
+    backbone = backbone,
+    verbose = TRUE,
+    download_cache_dir = "~/.cache/dyngen",
+    num_cores = 8,
+    distance_metric = "euclidean",
+    tf_network_params = tf_network_default(min_tfs_per_module = 2, sample_num_regulators = function() 1),
+    simulation_params = simulation_default(ssa_algorithm = GillespieSSA2::ssa_exact(), num_simulations = 8, census_interval = .01)
+  ) %>%
+  generate_tf_network() %>%
+  generate_feature_network() %>%
+  generate_kinetics()
+
+
+
+
+## custom plots for grn
+feature_info <-
+  model2$feature_info %>%
+  mutate(
+    color_by = case_when(
+      !is.na(module_id) ~ module_id,
+      is_tf ~ "TF",
+      !is_hk ~ "Target",
+      is_hk ~ "HK"
+    )
+  )
+
+color_legend <- c(
+  "TF" = "black",
[…]

### 2019-10-08 analysis/multiple_backbones.R

@@ -3,9 +3,17 @@ library(dyngen)
 
 backs <- list_backbones()
 
+nb <- "bifurcating_converging"
+nb <- "linear"
+nb <- "disconnected"
+
 for (nb in names(backs)) {
   cat("=============== SIMULATING ", nb, " ===============\n", sep = "")
   set.seed(1)
+
+  out_dir <- paste0("fig/example/bb_", nb, "/")
+  if (!dir.exists(out_dir)) dir.create(out_dir, recursive = TRUE, showWarnings = FALSE)
+
   back <- backs[[nb]]()
   model <-
     initialise_model(
@@ -20,9 +28,13 @@ for (nb in names(backs)) {
       simulation_params = simulation_default(ssa_algorithm = GillespieSSA2::ssa_exact(), num_simulations = 16, census_interval = .01)
     )
 
-  out <- generate_dataset(model, make_plots = FALSE)
+  out <- generate_dataset(model, output_dir = out_dir, make_plots = TRUE)
 
   # recalc dimred
+  out <- lst(
+    model = read_rds(paste0(out_dir, "model.rds")),
+    dataset = read_rds(paste0(out_dir, "dataset.rds"))
+  )
   gs_counts <- out$model$gold_standard$counts %>% as.matrix
   sim_counts <- out$model$simulations$counts[,colnames(gs_counts)] %>% as.matrix
   counts <- rbind(gs_counts, sim_counts)
@@ -34,8 +46,6 @@ for (nb in names(backs)) {
   # plot_gold_mappings(out$model, do_facet = FALSE)
 
   # save output
-  out_dir <- paste0("dyngen/bb_", nb, "/")
-  if (!dir.exists(out_dir)) dir.create(out_dir)
   write_rds(out, paste0(out_dir, "/out.rds"), compress = "gz")
 
   # plot model
@@ -71,7 +81,7 @@ for (nb in names(backs)) {
   dataset <- out$dataset
 
   library(dynplot)
-  g <- plot_dimred(dataset)
+  g <- plot_dimred(dataset, dimred = dyndimred::dimred_landmark_mds)
   ggsave(paste0(out_dir, "traj_dimred.pdf"), g, width = 8, height = 6)
   g <- plot_graph(dataset)
   ggsave(paste0(out_dir, "traj_graph.pdf"), g, width = 8, height = 6)

### 2019-10-10 analysis/multiple_backbones.R
update scripts

@@ -1,102 +1,191 @@
 library(tidyverse)
 library(dyngen)
+library(dynplot)
 
 backs <- list_backbones()
 
-nb <- "bifurcating_converging"
-nb <- "linear"
-nb <- "disconnected"
+oks <- c("linear", "bifurcating", "converging", "cycle", "bifurcating_loop", "consecutive_bifurcating")
+oks <- c(oks, "bifurcating_cycle", "bifurcating_converging", "binary_tree", "branching", "disconnected")
+oks <- c("binary_tree", "branching", "disconnected")
 
-for (nb in names(backs)) {
-  cat("=============== SIMULATING ", nb, " ===============\n", sep = "")
-  set.seed(1)
+for (seed in seq_len(10))
+for (nb in oks) {
+  cat("=============== SIMULATING ", nb, " seed ", seed, " ===============\n", sep = "")
+  set.seed(seed)
 
-  out_dir <- paste0("fig/example/bb_", nb, "/")
+  out_dir <- paste0("fig/example/bb_", nb, "_", seed, "/")
   if (!dir.exists(out_dir)) dir.create(out_dir, recursive = TRUE, showWarnings = FALSE)
 
-  back <- backs[[nb]]()
-  model <-
-    initialise_model(
-      num_tfs = nrow(back$module_info) * 1.5,
-      num_targets = 30,
-      num_hks = 15,
-      num_cells = 300,
-      backbone = back,
-      verbose = TRUE,
-      download_cache_dir = "~/.cache/dyngen",
-      num_cores = 8,
-      simulation_params = simulation_default(ssa_algorithm = GillespieSSA2::ssa_exact(), num_simulations = 16, census_interval = .01)
-    )
+  if (!file.exists(paste0(out_dir, "traj_dimred.pdf"))) {
+    back <- backs[[nb]]()
+    model <-
+      initialise_model(
+        num_tfs = nrow(back$module_info) * 2,
+        num_targets = 100,
+        num_hks = 20,
+        num_cells = 1000,
+        backbone = back,
+        verbose = TRUE,
+        download_cache_dir = "~/.cache/dyngen",
+        num_cores = 4,
+        gold_standard_params = gold_standard_default(
+          tau = .001,
+          census_interval = .02
+        ),
+        simulation_params = simulation_default(
[…]

### 2019-10-10 analysis/tcell.R

@@ -1,0 +1,133 @@
+library(tidyverse)
+library(dyngen)
+library(dynplot)
+
+set.seed(1)
+
+out_dir <- "fig/usecase/tcell/"
+if (!dir.exists(out_dir)) dir.create(out_dir, recursive = TRUE, showWarnings = FALSE)
+
+# Read network ------------------------------------------------------------
+feature_info <-
+  read_tsv("fig/usecase/tcell/feature_info.tsv", col_types = c(id = "c", name = "c")) %>%
+  transmute(
+    feature_id = paste0("Gene", id),
+    module_id = NA_character_,
+    ba = 0,
+    burn = TRUE,
+    ind = 1,
+    is_tf = TRUE,
+    is_hk = FALSE,
+    feature_name = name
+  )
+feature_network <-
+  read_tsv("fig/usecase/tcell/feature_network.tsv", col_types = c(from = "c", to = "c", effect = "c")) %>%
+  group_by(from, to) %>% # may have duplicates..
+  slice(1) %>%
+  ungroup() %>%
+  transmute(
+    from = paste0("Gene", from),
+    to = paste0("Gene", to),
+    effect = as.integer(ifelse(effect == "?", "1", effect)),
+    strength = ifelse(effect == 1, sample.int(4, n(), replace = TRUE), sample.int(2, n(), replace = TRUE)),
+    cooperativity = 2
+  )
+feature_network$from[!feature_network$from %in% feature_info$feature_id]
+feature_network$to[!feature_network$to %in% feature_info$feature_id]
+feature_info$feature_id[!feature_info$feature_id %in% c(feature_network$from, feature_network$to)]
+feature_info %>% filter(duplicated(feature_info))
+
+
+gr <- igraph::graph_from_data_frame(
+  feature_network,
+  vertices = feature_info,
+  directed = TRUE
+)
+# igraph::cluster_edge_betweenness(gr)
+cl <- igraph::cluster_infomap(gr, modularity = FALSE)
+
+module_ids <- paste0("Mod", LETTERS[sort(unique(cl$membership))])
+feature_info$module_id <- module_ids[cl$membership]
+# igraph::cluster_optimal(gr)
+# igraph::cluster_louvain(gr)
+# plot(gr, vertex.label = igraph::V(gr)$feature_name)
+
+
+# Create model ------------------------------------------------------------
+
+back <- backbone(
+  module_info = tibble(module_id = module_ids, ba = 0, burn = TRUE, ind = 1),
+  module_network =
+    left_join(
[…]

### 2019-10-10 analysis/usecase_differential_network.R

@@ -1,0 +1,218 @@
+library(tidyverse)
+library(dyngen)
+
+set.seed(2)
+
+out_dir <- "fig/usecase/differential_network/"
+dir.create(out_dir, showWarnings = FALSE, recursive = TRUE)
+
+# Generate base model -----------------------------------------------------
+backbone <- bblego(
+  bblego_start("A", type = "simple", num_modules = 4),
+  bblego_linear("A", "B", type = "doublerep1", num_modules = 6),
+  bblego_linear("B", "C", type = "doublerep2", num_modules = 6),
+  bblego_end("C", type = "simple", num_modules = 4)
+)
+model <-
+  initialise_model(
+    num_tfs = 30,
+    num_targets = 50,
+    num_hks = 50,
+    num_cells = 1000,
+    backbone = backbone,
+    verbose = TRUE,
+    download_cache_dir = "~/.cache/dyngen",
+    num_cores = 8,
+    distance_metric = "pearson",
+    tf_network_params = tf_network_default(min_tfs_per_module = 1, sample_num_regulators = function() 1),
+    simulation_params = simulation_default(num_simulations = 10, census_interval = .01)
+  ) %>%
+  generate_tf_network() %>%
+  generate_feature_network() %>%
+  generate_kinetics()
+
+
+# Show GRN ----------------------------------------------------------------
+plot_backbone_statenet(model)
+plot_backbone_modulenet(model)
+plot_feature_network(model, show_targets = FALSE)
+plot_feature_network(model)
+plot_feature_network(model, show_hks = TRUE)
+
+
+# Generate 2 variations with slightly randomised kinetics -----------------
+model1 <- model %>%
+  generate_gold_standard() %>%
+  generate_cells()
+
+model2 <- model
+# flip effect
+ix <- sample.int(nrow(model$feature_network), 5)
+model2$feature_network$effect[ix] <- 1 - model2$feature_network$effect[ix]
+
+model2$simulation_system$reactions <- dyngen:::.kinetics_generate_formulae(model2)
+generate_kinetics
+model2$simulation_system$reactions
+model2$simulation_system$initial_state
+model2 <- model2 %>%
+  generate_gold_standard() %>%
+  generate_cells()
+
+
[…]

### 2019-10-10 analysis/usecase_trajectory_alignment.R

@@ -1,0 +1,144 @@
+library(tidyverse)
+library(dyngen)
+
+set.seed(1)
+
+out_dir <- "fig/usecase/trajectory_alignment/"
+dir.create(out_dir, showWarnings = FALSE, recursive = TRUE)
+
+# Generate base model -----------------------------------------------------
+backbone <- bblego(
+  bblego_start("A", type = "simple", num_modules = 4),
+  bblego_linear("A", "B", type = "doublerep1", num_modules = 6),
+  bblego_linear("B", "C", type = "doublerep2", num_modules = 6),
+  bblego_end("C", type = "simple", num_modules = 4)
+)
+model <-
+  initialise_model(
+    num_tfs = 100,
+    num_targets = 200,
+    num_hks = 200,
+    num_cells = 1000,
+    backbone = backbone,
+    verbose = TRUE,
+    download_cache_dir = "~/.cache/dyngen",
+    num_cores = 8,
+    distance_metric = "pearson",
+    tf_network_params = tf_network_default(min_tfs_per_module = 2, sample_num_regulators = function() 1),
+    simulation_params = simulation_default(num_simulations = 10, census_interval = .01)
+  ) %>%
+  generate_tf_network() %>%
+  generate_feature_network()
+
+
+# Show GRN ----------------------------------------------------------------
+plot_backbone_statenet(model)
+plot_backbone_modulenet(model)
+plot_feature_network(model, show_targets = FALSE)
+plot_feature_network(model)
+plot_feature_network(model, show_hks = TRUE)
+
+
+# Generate 2 variations with slightly randomised kinetics -----------------
+model1 <- model
+model1$feature_info <- model1$feature_info %>%
+  mutate_at(c("ba", "ind"), ~ . * rnorm(length(.), mean = 1, sd = .01)) %>%
+  mutate_at(c("ba", "ind"), ~ pmin(., 1))
+model1$feature_network <- model1$feature_network %>%
+  mutate_at(c("strength", "cooperativity"), ~ . * rnorm(length(.), mean = 1, sd = .01))
+model1 <- model1 %>%
+  generate_kinetics() %>%
+  generate_gold_standard() %>%
+  generate_cells()
+
+model2 <- model
+# by_backbone <- model2$feature_network$from %>% grepl("_TF", .)
[…]

### 2019-10-10 content.tex
update manuscript

@@
 
-We introduce \texttt{dyngen}, a multi-modality simulator of single cells and their dynamics (Figure \ref{fig:dyngen}A).
+We introduce \texttt{dyngen}, a multi-modality simulator of single cells and their dynamics (Figure \ref{fig:showcase}A).
 A cell is simulated using Gillespie's Stochastic Simulation Algorithm (SSA) \cite{gillespie_exactstochasticsimulation_1977} where a cell consists of a set of molecules.
@@
 
-By simulating a cell over time in terms of its molecular state and the reactions that are allowed to occur, the simulator is more extendible to new modalities or experimental procedures (Figure~\ref{fig:dyngen}B).
+By simulating a cell over time in terms of its molecular state and the reactions that are allowed to occur, the simulator is more extendible to new modalities or experimental procedures (Figure~\ref{fig:showcase}B).
 We demonstrate \texttt{dyngen}'s flexibility by simulating three different types of biological experiments, and using these simulations to develop benchmarking methodologies for recently developed computational tools.
@@
 	\caption{
-		\textbf{Showcase of dyngen functionality.} 
+		\textbf{Showcase of \texttt{dyngen} functionality.} 
 		\textbf{A:} 
@@
 	}
-	\label{fig:dyngen}
+	\label{fig:showcase}
 \end{figure}
@@
 
-One of dyngen's main advantages is that depending on how the GRN is designed, different cellular developmental processes can be simulated.
-Different GRNs can result in branching, converging, cyclic, or even disconnected developmental topologies.
-Multiple simulations with slightly different GRNs can emulate rewiring events in disease or batch effects.
-
-\mycomment{could use figure; one with GRNs of different topologies, another with rewiring events}
-
-Another advantage is that dyngen returns many modalities throughout the whole simulation: molecular abundance, cellular state, number of reaction firings, reaction likelihoods, and regulation activations (Figure \ref{fig:simplecyclic}C--F).
+\texttt{dyngen} returns many modalities throughout the whole simulation: molecular abundance, cellular state, number of reaction firings, reaction likelihoods, and regulation activations (Figure \ref{fig:simplecyclic}C--F).
 These modalities can serve both as input data and ground truth for benchmarking many types of computational approaches.
@@
 
-The final main advantage is that \texttt{dyngen} can be used to simulate different experimental conditions.
-By default, dyngen supports snapshot experiments (uniformly sampling from an asynchronous dynamic process) and time-series experiments (sampling cells from different intervals in the simulation).
-It is possible to implement other experimental protocols, such as sampling the same cell at regular intervals or performing perturbation experiments.
+Depending on how the GRN is designed, different cellular developmental processes can be simulated.
+\texttt{dyngen} includes generators of GRNs which result in many different developmental topologies (Figure~\ref{fig:example_runs}), including branching, converging, cyclic and even disconnected.
 
-\mycomment{add result figures pertaining use cases}
+\begin{figure}[ht]
+	\centering
+	\includegraphics[width=\LARgefigure]{fig/example_runs_2} 
+	\caption{
+		\textbf{Multiple executions of \texttt{dyngen} with different predefined backbones.} From each simulation of about 200 genes, 1000 cells were sampled. 
+		\textbf{A:}~Linear. \textbf{B:}~Bifurcating. \textbf{C:}~Converging.
+		\textbf{D:}~Cyclic. \textbf{E:}~Bifurcating loop. \textbf{F:}~Bifurcating converging.
+		\textbf{G:}~Consecutive branching. \textbf{H:}~Binary tree. \textbf{I:}~Disconnected.
+	}
+	\label{fig:example_runs}
+\end{figure}
+
+Custom-defined GRNs offer more fine-grained control over the simulation.
+Aside from simulating topologies currently not supported by \texttt{dyngen}, this has several other important use-cases.
+Simulations of the same GRN with small perturbations allow to emulate batch effects or perturbation experiments (Figure~\ref{fig:usecases}).
+Simulating perturbed GRNs allow evaluating trajectory alignment methods -- which attempt to map two or more trajectories onto each other -- or differential network inference methods -- which infer differential regulatory interactions between two or more groups of profiles.
+
+\begin{figure}[ht]
+	\centering
+	\includegraphics[width=\LARgefigure]{fig/usecases} 
+	\caption{
+		\textbf{Examples of simulations with perturbed GRNs.}  
+		\textbf{A:} The cells in the top half were simulated with the same GRN as the cells on the bottom half, except that all parameters (e.g. strength of the interaction) were randomised.
+		\textbf{B:} Only 10 interactions in the GRN were randomised. In this example, the effect of the GRN perturbation is more subtle.
+	}
+	\label{fig:usecases}
+\end{figure}
+
+\texttt{dyngen} can be used to simulate different experimental conditions.
+By default, \texttt{dyngen} supports snapshot experiments (uniformly sampling from an asynchronous dynamic process) and time-series experiments (sampling cells from different intervals in the simulation).
+However, it is possible to imagine and implement other sampling strategies, such as sampling a cell at a certain time point and once more at a later time point.
+This would allow evaluating the performance of RNA velocity approaches -- which predict the future state of a cell by looking at differences in pre-mRNA and mRNA abundance levels.
+
+%\subsection{Example use cases}
+%\subsubsection{Trajectory alignment}
+%\subsubsection{Differential network inference}
+%\subsubsection{RNA velocity}
+%\subsubsection{Perturbation experiment}
 
 \section{Discussion}
-As is, dyngen's single cell simulations can be used to evaluate common single-cell omics computational methods such as clustering, batch correction, trajectory inference and network inference.
+As is, \texttt{dyngen}'s single cell simulations can be used to evaluate common single-cell omics computational methods such as clustering, batch correction, trajectory inference and network inference.
 However, the combined effect of these advantages results in a framework that is flexible enough to adapt to a broad range of applications.
 This may include methods that integrate clustering, network inference and trajectory inference.
-In this respect, dyngen may promote the development of new tools in the single-cell field similarly as other simulators have done in the past \cite{schaffter_genenetweaversilicobenchmark_2011,ewing_combiningtumorgenome_2015}.
+In this respect, \texttt{dyngen} may promote the development of new tools in the single-cell field similarly as other simulators have done in the past \cite{schaffter_genenetweaversilicobenchmark_2011,ewing_combiningtumorgenome_2015}.
 
-Adding batch effects to snapshot simulations of linear (or even branching) trajectories allows evaluating trajectory alignment methods -- which attempt to map two or more trajectories onto each other.
-Adding perturbations to the GRN allows evaluating the performance of differential network inference methods -- which predict differential regulatory interactions between two or more groups of profiles.
-Sampling a cell at a particular time point and once more at a later time point allows evaluating the performance of RNA velocity approaches -- which predict the future state of a cell by looking at differences in pre-mRNA and mRNA abundance levels.
+\texttt{dyngen} ultimately allows anticipating technological developments in single-cell multi-omics.
+In this way, it is possible to design and evaluate the performance and robustness of new types of computational analyses before experimental data becomes available.
+In addition, it could also be used to compare which experimental protocol is the most cost-effective in producing the qualitative and robust results in downstream analysis.
 
-dyngen ultimately also allows anticipating technological developments in single-cell multi-omics.
-In this way, it is possible to design and evaluate the performance and robustness of new types of computational analyses before experimental data becomes available.
-Similarly, it could also be used to compare which experimental technique will likely produce the most accurate result.
-For example, is it possible to infer directionality of regulatory interactions from snapshot experiments only, or are time series or knockdown experiments a necessity in order to infer high-quality regulatory networks?
-
-Currently, dyngen focuses on simulating cells as standalone entities.
-Future developments include extending the framework to simulate multiple cells in a virtual environment.
-Allowing cells to receive and react to environmental and intercellular stimuli would enable simulating essential cellular processes such as cell division and migration.
-
-%\mycomment{also adding more types of molecules, e.g. protein complex, small rnas, PPI}
+Currently, \texttt{dyngen} focuses on simulating cells as standalone entities that are well mixed.
+Splitting up the simulation space into separate subvolumes could pave the way to better study key cellular processes such as cell division, intercellular communication and migration\cite{smith_spatialstochasticintracellular_2019}.
 
@@
 	\includegraphics[width=\hugefigure]{fig/explain_methods} 
-	\caption{\textbf{The workflow of dyngen is comprised of six main steps.} \textbf{A:} The user needs to specify the desired module network or use a predefined module network.
+	\caption{\textbf{The workflow of \texttt{dyngen} is comprised of six main steps.} \textbf{A:} The user needs to specify the desired module network or use a predefined module network.
 \textbf{B:} Each gene in a module is is regulated by one or more transcription factors from the upstream module.
@@
 
-In dyngen, the user defines the behaviour of the simulation by defining how sets of genes, called modules, are regulating eachother.
+In \texttt{dyngen}, the user defines the behaviour of the simulation by defining how sets of genes, called modules, are regulating eachother.
 A module may have basal expression, which means that pre-mRNA of the genes in this module will be transcribed without the presence of transcription factor molecules.
 A module marked as "active during the burn phase" means that this module will be allowed to generate expression of its genes during an initial warm-up phase (See section \ref{sec:dyngen_ssa}).
-At the end of the dyngen process, cells will not be sampled from the burn phase simulations.
+At the end of the \texttt{dyngen} process, cells will not be sampled from the burn phase simulations.
+Interactions between modules have a strength (which is a positive integer) and an effect (+1 for upregulating, -1 for downregulating).
 
@@
 If this effect is undesirable, more safeguards need to be put in place to prevent reprogramming from occurring (Section \ref{sec:dyngen_backbone_lego}).
-
-\mycomment{mention strength, cooperativity, effect and independence factor.}
 
@@
 
+%!!!!!!!!!!!!!!!!!!!!!!!!!!!!!!!!!!!!!!!!!!!!!!!!!!!!!!!!!!!!!!!
+% TODO: add step-by-step  derivation of transcription formula
+%!!!!!!!!!!!!!!!!!!!!!!!!!!!!!!!!!!!!!!!!!!!!!!!!!!!!!!!!!!!!!!!
+
+% YSA: Graag wat meer uitleg in de tekst over al deze formules, definitie van notaties en zo.
+
 \begin{table}[ht]
@@
 \end{table}
-%!!!!!!!!!!!!!!!!!!!!!!!!!!!!!!!!!!!!!!!!!!!!!!!!!!!!!!!!!!!!!!!
-% TODO: add step-by-step  derivation of transcription formula
-%!!!!!!!!!!!!!!!!!!!!!!!!!!!!!!!!!!!!!!!!!!!!!!!!!!!!!!!!!!!!!!!
 
@@
 \begin{table}[ht]
-	\caption{\textbf{Parameters defined for the calculation of reaction propensity functions.}} \label{tab:reaction_params}
+	\caption{\textbf{Default parameters defined for the calculation of reaction propensity functions.}} \label{tab:reaction_params}
 	\centering
@@
 It is assumed the root state has no modules active and has no expression of any molecules.
-To get to next state, we follow a transition starting from the root state, activate and deactivate the modules as indicated by the transition, and compute the average molecule abundance along the transition.
+To get to the next state, we follow a transition starting from the root state, activate and deactivate the modules as indicated by the transition, and compute the average molecule abundance along the transition.
 To compute the average abundance, we perform small time steps $t = 0.001$ and let each reaction (Section \ref{sec:reactions}) occur $t$ times its propensity.
+
+% mention that this is basically an ode?
 
@@
 
-\texttt{dyngen} uses Gillespie's Stochastic Simulation Algorithm (SSA) to simulate dynamic processes. An SSA simulation is an iterative process where at each iteration one reaction is triggered.
+\texttt{dyngen} uses Gillespie's stochastic simulation algorithm (SSA)\cite{gillespie_exactstochasticsimulation_1977} to simulate dynamic processes.
+An SSA simulation is an iterative process where at each iteration one reaction is triggered.
 
@@
 \subsubsection{Map SSA simulations to backbone}
-The cellular state of each timepoint in the SSA simulation is mapped to the state network of the backbone by calculating the 1NN between a state vector in the simulation and the average expression levels along transitions.
+We compute the Pearson correlation between the state vectors in the simulation and the average expression levels along a transition in the backbone.
+Each timepoint in the SSA simulation is mapped to the point in the backbone that has the highest correlation value.
 
@@
 
-\subsection{Example runs of predefined backbones}
-
-\subsubsection{Linear}
-
-\begin{figure}[H]
-	\centering
-	\includegraphics[width=\linewidth]{fig/example/bb_linear/plot} 
-	\caption{}
-	\label{fig:example_linear}
-\end{figure}
-
-\subsubsection{Bifurcating}
-
-\begin{figure}[H]
-	\centering
-	\includegraphics[width=\linewidth]{fig/example/bb_bifurcating/plot} 
-	\caption{}
-	\label{fig:example_bifurcating}
-\end{figure}
-
-\subsubsection{Cycle}
-
-\begin{figure}[H]
-	\centering
-	\includegraphics[width=\linewidth]{fig/example/bb_cycle/plot} 
-	\caption{}
-	\label{fig:example_cycle}
-\end{figure}
-
-%\subsubsection{Branching}
-%
-%\begin{figure}[H]
-%	\centering
-%	\includegraphics[width=\linewidth]{fig/example/bb_binary_tree/plot} 
-%	\caption{}
-%	\label{fig:example_branching}
-%\end{figure}
-
-\subsubsection{Converging}
-
-\begin{figure}[H]
-	\centering
-	\includegraphics[width=\linewidth]{fig/example/bb_converging/plot} 
-	\caption{}
-	\label{fig:example_converging}
-\end{figure}
-
-\subsubsection{Bifurcating converging}
-
-\begin{figure}[H]
-	\centering
-	\includegraphics[width=\linewidth]{fig/example/bb_bifurcating_converging/plot} 
-	\caption{}
-	\label{fig:example_bifurcating_converging}
-\end{figure}
-
-%\subsubsection{Bifurcating cycle}
-%
-%\begin{figure}[H]
-%	\centering
-%	\includegraphics[width=\linewidth]{fig/example/bb_bifurcating_cycle/plot} 
-%	\caption{}
-%	\label{fig:example_bifurcating_cycle}
-%\end{figure}
-
-\subsubsection{Bifurcating loop}
-
-\begin{figure}[H]
-	\centering
-	\includegraphics[width=\linewidth]{fig/example/bb_bifurcating_loop/plot} 
-	\caption{}
-	\label{fig:example_bifurcating_loop}
-\end{figure}
-
-%\subsubsection{Disconnected}
-%
-%\begin{figure}[H]
-%	\centering
-%	\includegraphics[width=\linewidth]{fig/example/bb_disconnected/plot} 
-%	\caption{}
-%	\label{fig:example_disconnected}
-%\end{figure}
-
-\subsection{Example use cases}
-\subsubsection{Trajectory alignment}
-From discussion: Adding batch effects to snapshot simulations of linear (or even branching) trajectories allows evaluating trajectory alignment methods -- which attempt to map two or more trajectories onto each other.
-\subsubsection{Differential network inference}
-From discussion: Adding perturbations to the GRN allows evaluating the performance of differential network inference methods -- which predict differential regulatory interactions between two or more groups of profiles.
-\subsubsection{RNA velocity}
-From discussion: Sampling a cell at a certain time point and once more at a later time point allows evaluating the performance of RNA velocity approaches -- which predict the future state of a cell by looking at differences in pre-mRNA and mRNA abundance levels.
-
-\subsubsection{Perturbation experiment}
-
 \clearpage

### 2019-10-14 content.tex
add comment
update

@@
 
+% YSA: RNA velocity: Onduidelijk hier, mensen kennen dit niet.
+
 %\subsection{Example use cases}
@@
 Splitting up the simulation space into separate subvolumes could pave the way to better study key cellular processes such as cell division, intercellular communication and migration\cite{smith_spatialstochasticintracellular_2019}.
+
+% We developed \texttt{dyngen}, a simulator of single cells to benchmark and stress-test computational tools for single-cell omics.
+%It supports a selection of existing experimental protocols, but can also simulate experiments for permitted by current technologies, such as generating omics profiles from the same cell at multiple timepoints.
+%This data can help kick-start emerging domains with low data availability more safely by allowing method developers to test their method before numerous datasets become publicly available.
+%\texttt{dyngen} has already been used successfully to evaluate trajectory inference\cite{saelens_comparisonsinglecelltrajectory_2019}, trajectory alignment\cite{vandenberge_trajectorybaseddifferentialexpression_2019}, and network inference\cite{pratapa_benchmarkingalgorithmsgene_2019} methods.
+Furthermore, we showed that it can also be used to evaluate differential network inference methods, and trajectory alignment methods.
+% TODO: And RNA velocity?
 
@@
 The exact way a cell chooses a certain path during its differentiation is still an active research field, although certain models have already emerged and been tested \textit{in vivo}.
-One driver of bifurcation seems to be mutual antagonism, where two genes \cite{xu_regulationbifurcatingcell_2015} strongly repress each other, forcing one of the two to become inactive \cite{graf_forcingcellschange_2009}.
+One driver of bifurcation seems to be mutual antagonism, where two genes strongly repress each other\cite{rekhtman_directinteractionhematopoietic_1999,xu_regulationbifurcatingcell_2015}, forcing one of the two to become inactive \cite{graf_forcingcellschange_2009}.
 Such mutual antagonism can be modelled and simulated \cite{wang_quantifyingwaddingtonlandscape_2011, ferrell_bistabilitybifurcationswaddington_2012}.
@@
 \subsection{Convert gene regulatory network to a set of reactions} \label{sec:reactions}
+% paper on transcriptional bursting
+% doi: 10.1039/C7MB00154A
+% interesting to try to replicate in dyngen! :)
 \newcommand{\w}[1]{\text{w}_{#1}}

### 2019-10-15 content.tex
update

@@
 
-\begin{quote}
-	\textbf{Abstract:} 
-\end{quote}
+%\begin{quote}
+%	\textbf{Abstract:} 
+%\end{quote}
 
@@
 Generators of scRNA-seq data (e.g.
-splatter \cite{zappia_splattersimulationsinglecell_2017}, powsimR \cite{vieth_powsimrpoweranalysis_2017}, PROSSTT \cite{papadopoulos_prossttprobabilisticsimulation_2018} and SymSim \cite{zhang_simulatingmultiplefaceted_2019}) have already been widely used to explore the strengths and weakness of computational tools, both by method developers\cite{street_slingshotcelllineage_2018,parra_reconstructingcomplexlineage_2018,lummertzdarocha_reconstructioncomplexsinglecell_2018,lin_scclassifyhierarchicalclassification_2019} and independent benchmarkers\cite{duo_systematicperformanceevaluation_2018,saelens_comparisonsinglecelltrajectory_2019,soneson_biasrobustnessscalability_2018}.
+splatter\cite{zappia_splattersimulationsinglecell_2017}, powsimR\cite{vieth_powsimrpoweranalysis_2017}, PROSSTT\cite{papadopoulos_prossttprobabilisticsimulation_2018} and SymSim\cite{zhang_simulatingmultiplefaceted_2019}) have already been widely used to explore the strengths and weakness of computational tools, both by method developers\cite{street_slingshotcelllineage_2018,parra_reconstructingcomplexlineage_2018,lummertzdarocha_reconstructioncomplexsinglecell_2018,lin_scclassifyhierarchicalclassification_2019} and independent benchmarkers\cite{duo_systematicperformanceevaluation_2018,saelens_comparisonsinglecelltrajectory_2019,soneson_biasrobustnessscalability_2018}.
 However, a limitation of scRNA-seq profiles generators is that they would require significant methodological alterations to add additional modalities.

### 2019-10-16 content.tex
~

@@
 By simulating a cell over time in terms of its molecular state and the reactions that are allowed to occur, the simulator is more extendible to new modalities or experimental procedures (Figure~\ref{fig:showcase}B).
-We demonstrate \texttt{dyngen}'s flexibility by simulating three different types of biological experiments, and using these simulations to develop benchmarking methodologies for recently developed computational tools.
+We demonstrate \texttt{dyngen}'s flexibility by simulating snapshot, time-series and pertubation experiments, allowing to benchmark a wide variety of computational tools such as trajectory alignment and differential network inference methods.
 

### 2019-10-16 content.tex
change cites

@@
 Generators of scRNA-seq data (e.g.
-splatter\cite{zappia_splattersimulationsinglecell_2017}, powsimR\cite{vieth_powsimrpoweranalysis_2017}, PROSSTT\cite{papadopoulos_prossttprobabilisticsimulation_2018} and SymSim\cite{zhang_simulatingmultiplefaceted_2019}) have already been widely used to explore the strengths and weakness of computational tools, both by method developers\cite{street_slingshotcelllineage_2018,parra_reconstructingcomplexlineage_2018,lummertzdarocha_reconstructioncomplexsinglecell_2018,lin_scclassifyhierarchicalclassification_2019} and independent benchmarkers\cite{duo_systematicperformanceevaluation_2018,saelens_comparisonsinglecelltrajectory_2019,soneson_biasrobustnessscalability_2018}.
+splatter \cite{zappia_splattersimulationsinglecell_2017}, powsimR \cite{vieth_powsimrpoweranalysis_2017}, PROSSTT \cite{papadopoulos_prossttprobabilisticsimulation_2018} and SymSim \cite{zhang_simulatingmultiplefaceted_2019}) have already been widely used to explore the strengths and weakness of computational tools, both by method developers \cite{street_slingshotcelllineage_2018,parra_reconstructingcomplexlineage_2018,lummertzdarocha_reconstructioncomplexsinglecell_2018,lin_scclassifyhierarchicalclassification_2019} and independent benchmarkers \cite{duo_systematicperformanceevaluation_2018,saelens_comparisonsinglecelltrajectory_2019,soneson_biasrobustnessscalability_2018}.
 However, a limitation of scRNA-seq profiles generators is that they would require significant methodological alterations to add additional modalities.
@@
 Currently, \texttt{dyngen} focuses on simulating cells as standalone entities that are well mixed.
-Splitting up the simulation space into separate subvolumes could pave the way to better study key cellular processes such as cell division, intercellular communication and migration\cite{smith_spatialstochasticintracellular_2019}.
+Splitting up the simulation space into separate subvolumes could pave the way to better study key cellular processes such as cell division, intercellular communication and migration \cite{smith_spatialstochasticintracellular_2019}.
 
@@
 %This data can help kick-start emerging domains with low data availability more safely by allowing method developers to test their method before numerous datasets become publicly available.
-%\texttt{dyngen} has already been used successfully to evaluate trajectory inference\cite{saelens_comparisonsinglecelltrajectory_2019}, trajectory alignment\cite{vandenberge_trajectorybaseddifferentialexpression_2019}, and network inference\cite{pratapa_benchmarkingalgorithmsgene_2019} methods.
+%\texttt{dyngen} has already been used successfully to evaluate trajectory inference \cite{saelens_comparisonsinglecelltrajectory_2019}, trajectory alignment \cite{vandenberge_trajectorybaseddifferentialexpression_2019}, and network inference \cite{pratapa_benchmarkingalgorithmsgene_2019} methods.
 Furthermore, we showed that it can also be used to evaluate differential network inference methods, and trajectory alignment methods.
@@
 The exact way a cell chooses a certain path during its differentiation is still an active research field, although certain models have already emerged and been tested \textit{in vivo}.
-One driver of bifurcation seems to be mutual antagonism, where two genes strongly repress each other\cite{rekhtman_directinteractionhematopoietic_1999,xu_regulationbifurcatingcell_2015}, forcing one of the two to become inactive \cite{graf_forcingcellschange_2009}.
+One driver of bifurcation seems to be mutual antagonism, where two genes strongly repress each other \cite{rekhtman_directinteractionhematopoietic_1999,xu_regulationbifurcatingcell_2015}, forcing one of the two to become inactive \cite{graf_forcingcellschange_2009}.
 Such mutual antagonism can be modelled and simulated \cite{wang_quantifyingwaddingtonlandscape_2011, ferrell_bistabilitybifurcationswaddington_2012}.
@@
 Target genes regulated by a TF or another target gene, but is always downstream of at least one TF.
-To sample the interactions between target genes, one of the many FANTOM5\cite{lizio_gatewaysfantom5promoter_2015} GRNs is sampled.
+To sample the interactions between target genes, one of the many FANTOM5 \cite{lizio_gatewaysfantom5promoter_2015} GRNs is sampled.
 The currently existing TFs are mapped to regulators in the FANTOM5 GRN.
@@
 
-\texttt{dyngen} uses Gillespie's stochastic simulation algorithm (SSA)\cite{gillespie_exactstochasticsimulation_1977} to simulate dynamic processes.
+\texttt{dyngen} uses Gillespie's stochastic simulation algorithm (SSA) \cite{gillespie_exactstochasticsimulation_1977} to simulate dynamic processes.
 An SSA simulation is an iterative process where at each iteration one reaction is triggered.
@@
 Molecules are sampled from the simulation to replicate how molecules are experimentally sampled.
-A real dataset is downloaded from a repository of single-cell RNA-seq datasets\cite{cannoodt_singlecellomicsdatasets_2018}.
+A real dataset is downloaded from a repository of single-cell RNA-seq datasets \cite{cannoodt_singlecellomicsdatasets_2018}.
 For each \textit{in silico} cell $i$, draw its library size $ls_i$from the distribution of transcript counts per cell in the real dataset.

### 2019-10-30 manuscript.Rmd
update to rmd

--- 2019-10-15 content.tex
@@
-% DYNGEN CONTENT
+---
+title: "dyngen: benchmarking with in silico single cells"
+author:
+- Robrecht Cannoodt
+- Wouter Saelens
+- Yvan Saeys
+date: "`r format(Sys.time(), '%d %B %Y')`"
+output:
+  pdf_document:
+        keep_tex: yes
+        latex_engine: xelatex
+        citation_package: biblatex
+biblatexoptions: sorting=none
+classoption:
+- table
+- 10pt
+- a4paper
+header-includes: |
+  \usepackage{tcolorbox}
+  \usepackage{colortbl}
+  \usepackage{booktabs}
+  \usepackage{tabularx}
+  \usepackage{fontspec}
+  \setmainfont [Path = fonts/,
+        UprightFont = *-300,
+        ItalicFont = *-300-Italic,
+        BoldFont = *-700,
+        BoldItalicFont = *-700-Italic
+  ]{MuseoSans}
+bibliography: library.bib
+---
 
-%\begin{quote}
-%	\textbf{Abstract:} 
-%\end{quote}
+```{r setup, include=FALSE}
+knitr::opts_chunk$set(collapse=TRUE, results="hold")
+library(tidyverse)
+```
 
-\section{Introduction}
-Continuous technological advancements to single-cells omics are having profound 
-effects on how researchers can validate biological hypotheses. 
-Early experimental technologies typically only allowed profiling a single modality (e.g. DNA sequence, 
-RNA or protein expression). However, recent developments permit profiling multiple modalities simultaneously,
-and every modality added allows for new types of analyses that can be performed.
+**Purpose:** A common problem of pioneering computational tools is that during their development, there are rarely sufficient datasets available for adequately quantitatively assessing its performance.
+**Results:** We developed `dyngen`, a multi-modality simulator of single cells.
+In `dyngen`, the biomolecular state of an _in silico_ cell changes over time according to a set of reactions defined by the cell's gene regulatory network.
+By simulating single cells in terms of its biomolecular state and reactions, the simulator is easily extendible for adding new modalities or experimental procedures.
+We demonstrate `dyngen`'s flexibility by simulating snapshot, time-series and perturbation experiments.
+**Conclusion:** `dyngen` lays the foundations for benchmarking a wide variety of computational single-cell tools, and can be used to help kick-start the development of future types of analyses.
+
+# Introduction
+Continuous technological advancements to single-cells omics are having profound effects on how researchers can validate biological hypotheses.
+Early experimental technologies typically only allowed profiling a single modality (e.g.
+DNA sequence, RNA or protein expression).
+However, recent developments permit profiling multiple modalities simultaneously, and every modality added allows for new types of analyses that can be performed.
 
@@
 Generators of scRNA-seq data (e.g.
-splatter\cite{zappia_splattersimulationsinglecell_2017}, powsimR\cite{vieth_powsimrpoweranalysis_2017}, PROSSTT\cite{papadopoulos_prossttprobabilisticsimulation_2018} and SymSim\cite{zhang_simulatingmultiplefaceted_2019}) have already been widely used to explore the strengths and weakness of computational tools, both by method developers\cite{street_slingshotcelllineage_2018,parra_reconstructingcomplexlineage_2018,lummertzdarocha_reconstructioncomplexsinglecell_2018,lin_scclassifyhierarchicalclassification_2019} and independent benchmarkers\cite{duo_systematicperformanceevaluation_2018,saelens_comparisonsinglecelltrajectory_2019,soneson_biasrobustnessscalability_2018}.
+splatter [@zappia_splattersimulationsinglecell_2017], powsimR [@vieth_powsimrpoweranalysis_2017], PROSSTT [@papadopoulos_prossttprobabilisticsimulation_2018] and SymSim [@zhang_simulatingmultiplefaceted_2019]) have already been widely used to explore the strengths and weaknesses of computational tools, both by method developers [@street_slingshotcelllineage_2018; @parra_reconstructingcomplexlineage_2018; @lummertzdarocha_reconstructioncomplexsinglecell_2018; @lin_scclassifyhierarchicalclassification_2019] and independent benchmarkers [@duo_systematicperformanceevaluation_2018; @saelens_comparisonsinglecelltrajectory_2019; @soneson_biasrobustnessscalability_2018].
 However, a limitation of scRNA-seq profiles generators is that they would require significant methodological alterations to add additional modalities.
 
-An ideal experiment would be able to observe all aspects of a cell, including a full history of its 
-molecular states, spatial positions and environmental interactions \cite{stuart_integrativesinglecellanalysis_2019}.
+An ideal experiment would be able to observe all aspects of a cell, including a full history of its molecular states, spatial positions and environmental interactions [@stuart_integrativesinglecellanalysis_2019].
 While this falls outside the reach of current experimental technologies, generating synthetic data in anticipation of new experimental technologies would allow already developing the next wave of computational tools.
 
-We introduce \texttt{dyngen}, a multi-modality simulator of single cells and their dynamics (Figure \ref{fig:showcase}A).
-A cell is simulated using Gillespie's Stochastic Simulation Algorithm (SSA) \cite{gillespie_exactstochasticsimulation_1977} where a cell consists of a set of molecules.
+We introduce `dyngen`, a multi-modality simulator of single cells and their dynamics (Figure \ref{fig:showcase}A).
+A cell is simulated using Gillespie's Stochastic Simulation Algorithm (SSA) [@gillespie_exactstochasticsimulation_1977] where a cell consists of a set of molecules.
 Throughout a simulation, reactions (e.g.
@@
 
-By simulating a cell over time in terms of its molecular state and the reactions that are allowed to occur, the simulator is more extendible to new modalities or experimental procedures (Figure~\ref{fig:showcase}B).
-We demonstrate \texttt{dyngen}'s flexibility by simulating three different types of biological experiments, and using these simulations to develop benchmarking methodologies for recently developed computational tools.
+By simulating a cell over time in terms of its molecular state and the reactions that are allowed to occur, the simulator is more extendible to new modalities or experimental procedures (Figure \ref{fig:showcase}B).
+We demonstrate `dyngen`'s flexibility by simulating snapshot, time-series and perturbation experiments, allowing to benchmark a wide variety of computational tools such as trajectory alignment and differential network inference methods.
 
@@
 
-\section{Results}
+# Results
 
@@
 
-\texttt{dyngen} returns many modalities throughout the whole simulation: molecular abundance, cellular state, number of reaction firings, reaction likelihoods, and regulation activations (Figure \ref{fig:simplecyclic}C--F).
+`dyngen` returns many modalities throughout the whole simulation: molecular abundance, cellular state, number of reaction firings, reaction likelihoods, and regulation activations (Figure \ref{fig:simplecyclic}C--F).
 These modalities can serve both as input data and ground truth for benchmarking many types of computational approaches.
@@
 Depending on how the GRN is designed, different cellular developmental processes can be simulated.
-\texttt{dyngen} includes generators of GRNs which result in many different developmental topologies (Figure~\ref{fig:example_runs}), including branching, converging, cyclic and even disconnected.
+`dyngen` includes generators of GRNs which result in many different developmental topologies (Figure \ref{fig:example_runs}), including branching, converging, cyclic and even disconnected.
 
@@
     \centering
-	\includegraphics[width=\LARgefigure]{fig/example_runs_2} 
+    \includegraphics[width=0.6\textwidth]{fig/example_runs_2}
     \caption{
@@
 Custom-defined GRNs offer more fine-grained control over the simulation.
-Aside from simulating topologies currently not supported by \texttt{dyngen}, this has several other important use-cases.
-Simulations of the same GRN with small perturbations allow to emulate batch effects or perturbation experiments (Figure~\ref{fig:usecases}).
+Aside from simulating topologies currently not supported by `dyngen`, this has several other important use-cases.
+Simulations of the same GRN with small perturbations allow to emulate batch effects or perturbation experiments (Figure \ref{fig:usecases}).
 Simulating perturbed GRNs allow evaluating trajectory alignment methods -- which attempt to map two or more trajectories onto each other -- or differential network inference methods -- which infer differential regulatory interactions between two or more groups of profiles.
@@
     \centering
-	\includegraphics[width=\LARgefigure]{fig/usecases} 
+    \includegraphics[width=0.6\textwidth]{fig/usecases}
     \caption{
@@
 
-\texttt{dyngen} can be used to simulate different experimental conditions.
-By default, \texttt{dyngen} supports snapshot experiments (uniformly sampling from an asynchronous dynamic process) and time-series experiments (sampling cells from different intervals in the simulation).
+`dyngen` can be used to simulate different experimental conditions.
+By default, `dyngen` supports snapshot experiments (uniformly sampling from an asynchronous dynamic process) and time-series experiments (sampling cells from different intervals in the simulation).
 However, it is possible to imagine and implement other sampling strategies, such as sampling a cell at a certain time point and once more at a later time point.
@@
 
-% YSA: RNA velocity: Onduidelijk hier, mensen kennen dit niet.
+<!-- YSA: RNA velocity: Onduidelijk hier, mensen kennen dit niet. -->
 
-%\subsection{Example use cases}
-%\subsubsection{Trajectory alignment}
-%\subsubsection{Differential network inference}
-%\subsubsection{RNA velocity}
-%\subsubsection{Perturbation experiment}
+<!-- ## Example use cases -->
 
-\section{Discussion}
-As is, \texttt{dyngen}'s single cell simulations can be used to evaluate common single-cell omics computational methods such as clustering, batch correction, trajectory inference and network inference.
+<!-- ### Trajectory alignment -->
+
+<!-- ### Differential network inference -->
+
+<!-- ### RNA velocity -->
+
+<!-- ### Perturbation experiment -->
+
+# Discussion
+As is, `dyngen`'s single cell simulations can be used to evaluate common single-cell omics computational methods such as clustering, batch correction, trajectory inference and network inference.
 However, the combined effect of these advantages results in a framework that is flexible enough to adapt to a broad range of applications.
 This may include methods that integrate clustering, network inference and trajectory inference.
-In this respect, \texttt{dyngen} may promote the development of new tools in the single-cell field similarly as other simulators have done in the past \cite{schaffter_genenetweaversilicobenchmark_2011,ewing_combiningtumorgenome_2015}.
+In this respect, `dyngen` may promote the development of new tools in the single-cell field similarly as other simulators have done in the past [@schaffter_genenetweaversilicobenchmark_2011; @ewing_combiningtumorgenome_2015].
 
-\texttt{dyngen} ultimately allows anticipating technological developments in single-cell multi-omics.
+`dyngen` ultimately allows anticipating technological developments in single-cell multi-omics.
 In this way, it is possible to design and evaluate the performance and robustness of new types of computational analyses before experimental data becomes available.
@@
 
-Currently, \texttt{dyngen} focuses on simulating cells as standalone entities that are well mixed.
-Splitting up the simulation space into separate subvolumes could pave the way to better study key cellular processes such as cell division, intercellular communication and migration\cite{smith_spatialstochasticintracellular_2019}.
+Currently, `dyngen` focuses on simulating cells as standalone entities that are well mixed.
+Splitting up the simulation space into separate subvolumes could pave the way to better study key cellular processes such as cell division, intercellular communication and migration [@smith_spatialstochasticintracellular_2019].
 
-% We developed \texttt{dyngen}, a simulator of single cells to benchmark and stress-test computational tools for single-cell omics.
-%It supports a selection of existing experimental protocols, but can also simulate experiments for permitted by current technologies, such as generating omics profiles from the same cell at multiple timepoints.
-%This data can help kick-start emerging domains with low data availability more safely by allowing method developers to test their method before numerous datasets become publicly available.
-%\texttt{dyngen} has already been used successfully to evaluate trajectory inference\cite{saelens_comparisonsinglecelltrajectory_2019}, trajectory alignment\cite{vandenberge_trajectorybaseddifferentialexpression_2019}, and network inference\cite{pratapa_benchmarkingalgorithmsgene_2019} methods.
-Furthermore, we showed that it can also be used to evaluate differential network inference methods, and trajectory alignment methods.
-% TODO: And RNA velocity?
-
-\section{Methods}
-The workflow to generate \textit{in silico} single cell data consists of six main steps (Figure \ref{fig:explain_methods}).
+# Methods
+The workflow to generate _in silico_ single cell data consists of six main steps (Figure \ref{fig:explain_methods}).
 
@@
     \centering
-	\includegraphics[width=\hugefigure]{fig/explain_methods} 
+    \includegraphics[width=\textwidth]{fig/explain_methods}
     \caption{\textbf{The workflow of \texttt{dyngen} is comprised of six main steps.} \textbf{A:} The user needs to specify the desired module network or use a predefined module network.
@@
 
-\subsection{Defining the backbone: modules and states} \label{sec:backbone}
+## Defining the backbone: modules and states
 
 One of the main processes involved in cellular dynamic processes is gene regulation, where regulatory cascades and feedback loops lead to progressive changes in expression and decision making.
-The exact way a cell chooses a certain path during its differentiation is still an active research field, although certain models have already emerged and been tested \textit{in vivo}.
-One driver of bifurcation seems to be mutual antagonism, where two genes strongly repress each other\cite{rekhtman_directinteractionhematopoietic_1999,xu_regulationbifurcatingcell_2015}, forcing one of the two to become inactive \cite{graf_forcingcellschange_2009}.
-Such mutual antagonism can be modelled and simulated \cite{wang_quantifyingwaddingtonlandscape_2011, ferrell_bistabilitybifurcationswaddington_2012}.
-Although the two-gene model is simple and elegant, the reality is frequently more complex, with multiple genes (grouped into modules) repressing each other \cite{yosef_dynamicregulatorynetwork_2013}.
+The exact way a cell chooses a certain path during its differentiation is still an active research field, although certain models have already emerged and been tested _in vivo_.
+One driver of bifurcation seems to be mutual antagonism, where two genes strongly repress each other [@rekhtman_directinteractionhematopoietic_1999; @xu_regulationbifurcatingcell_2015], forcing one of the two to become inactive [@graf_forcingcellschange_2009].
+Such mutual antagonism can be modelled and simulated [@wang_quantifyingwaddingtonlandscape_2011; @ferrell_bistabilitybifurcationswaddington_2012].
+Although the two-gene model is simple and elegant, the reality is frequently more complex, with multiple genes (grouped into modules) repressing each other [@yosef_dynamicregulatorynetwork_2013].
 
-In \texttt{dyngen}, the user defines the behaviour of the simulation by defining how sets of genes, called modules, are regulating eachother.
+In `dyngen`, the user defines the behaviour of the simulation by defining how sets of genes, called modules, are regulating each other.
 A module may have basal expression, which means that pre-mRNA of the genes in this module will be transcribed without the presence of transcription factor molecules.
-A module marked as "active during the burn phase" means that this module will be allowed to generate expression of its genes during an initial warm-up phase (See section \ref{sec:dyngen_ssa}).
-At the end of the \texttt{dyngen} process, cells will not be sampled from the burn phase simulations.
+A module marked as "active during the burn phase" means that this module will be allowed to generate expression of its genes during an initial warm-up phase.
+At the end of the `dyngen` process, cells will not be sampled from the burn phase simulations.
 Interactions between modules have a strength (which is a positive integer) and an effect (+1 for upregulating, -1 for downregulating).
@@
 In addition, the fate decisions made in this example of a bifurcation is reversible, meaning cells can be reprogrammed to go down a different differentiation path.
-If this effect is undesirable, more safeguards need to be put in place to prevent reprogramming from occurring (Section \ref{sec:dyngen_backbone_lego}).
+If this effect is undesirable, more safeguards need to be put in place to prevent reprogramming from occurring.
 
@@
     \centering
-	\includegraphics[width=\LARGEfigure]{fig/example_backbones} 
+    \includegraphics[width=0.8\textwidth]{fig/example_backbones}
     \caption{Example module networks}
@@
 Finally, to transition from S3 to S1 again, A and B are turned on again and D and E are fixed again.
-Demonstrations of the backbone will be explained in more detail in section \ref{sec:dyngen_sim_backbone}.
+Demonstrations of the backbone will be explained in more detail in a later section.
 
-\subsubsection{Backbone lego} \label{sec:dyngen_backbone_lego}
+### Backbone lego
 The backbone can make use of one or more "backbone lego" (BBL) pieces (Figure \ref{fig:backbone_lego}).
@@
     \centering
-	\includegraphics[width=\LARGEfigure]{fig/backbone_lego} 
+    \includegraphics[width=0.8\textwidth]{fig/backbone_lego}
     \caption{Backbone lego}
@@
 
-When the input node of a \textbf{linear BBL} (Figure \ref{fig:backbone_lego}C) is upregulated, the module the BBL is connected to will be upregulated.
+When the input node of a **linear BBL** (Figure \ref{fig:backbone_lego}C) is upregulated, the module the BBL is connected to will be upregulated.
 A \emph{simple chain} is a set of modules where a module upregulates the next.
@@
 
-When the input node of a \textbf{branching BBL} (Figure \ref{fig:backbone_lego}D) is upregulated, a subset of its output modules will eventually be upregulated.
+When the input node of a **branching BBL** (Figure \ref{fig:backbone_lego}D) is upregulated, a subset of its output modules will eventually be upregulated.
 A \emph{simple branching} uses reciprocal inhibition to drive the upregulation of one of the output modules.
@@
 
-A \textbf{leaf BBL} (Figure \ref{fig:backbone_lego}E) is a linear BBL that has either no inputs or no outputs.
-A \emph{start} BBL is a linear BBL where the first module has basal expression, and all modules in this module will be active during the burn-in phase of the simulation (Section \ref{sec:dyngen_sim_backbone}).
-An \emph{end} BBL is also a linear BBL with its output regulating one final module.
+A **leaf BBL** (Figure \ref{fig:backbone_lego}E) is a linear BBL that has either no inputs or no outputs.
+A _start_ BBL is a linear BBL where the first module has basal expression, and all modules in this module will be active during the burn-in phase of the simulation.
+An _end_ BBL is also a linear BBL with its output regulating one final module.
 
-\subsection{Generating the gene regulatory network}
-The GRN is generated based on the given backbome in four main steps (Figure~\ref{fig:gen_feature_network}).
+## Generating the gene regulatory network
+The GRN is generated based on the given backbome in four main steps (Figure \ref{fig:gen_feature_network}).
 
@@
 
-\paragraph{Step 1, sampling the transcription factors (TF).} The TFs are the main drivers of the molecular changes in the simulation.
+**Step 1, sampling the transcription factors (TF).** The TFs are the main drivers of the molecular changes in the simulation.
 The user provides a backbone and the number of TFs to generate.
@@
 
-\paragraph{Step 2, generating the TF interactions.} Let each TF be regulated according to the interactions in the backbone.
+**Step 2, generating the TF interactions.** Let each TF be regulated according to the interactions in the backbone.
 These interactions inherit the effect, strength, and cooperativity parameters from the interactions in the backbone.
@@
 
-\paragraph{Step 3, sampling the target subnetwork.} 
+**Step 3, sampling the target subnetwork.**
 A user-defined number of target genes are added to the GRN.
 Target genes regulated by a TF or another target gene, but is always downstream of at least one TF.
-To sample the interactions between target genes, one of the many FANTOM5\cite{lizio_gatewaysfantom5promoter_2015} GRNs is sampled.
+To sample the interactions between target genes, one of the many FANTOM5 [@lizio_gatewaysfantom5promoter_2015] GRNs is sampled.
 The currently existing TFs are mapped to regulators in the FANTOM5 GRN.
@@
 
-\paragraph{Step 4, sampling the housekeeping subnetwork.}
+**Step 4, sampling the housekeeping subnetwork.**
 Housekeeping genes are completely separate from any TFs or target genes.
@@
 
-\subsection{Convert gene regulatory network to a set of reactions} \label{sec:reactions}
-% paper on transcriptional bursting
-% doi: 10.1039/C7MB00154A
-% interesting to try to replicate in dyngen! :)
+## Convert gene regulatory network to a set of reactions
+
+<!-- paper on transcriptional bursting -->
+
+<!-- doi: 10.1039/C7MB00154A -->
+
+<!-- interesting to try to replicate in dyngen! :) -->
+
 \newcommand{\w}[1]{\text{w}_{#1}}
@@
 
-For every gene $G$ in the GRN, the simulation will keep track of the abundance levels of three molecules, a pre-mRNA, a mature mRNA and a protein.
+For every gene $G$ in the GRN, the simulation will keep track of the abundance levels of three molecules, a pre-mRNA, a mature mRNA, and a protein.
 These abundance levels are represented as $\w G$, $\x G$ and $\y G$ respectively.
@@
 
-%!!!!!!!!!!!!!!!!!!!!!!!!!!!!!!!!!!!!!!!!!!!!!!!!!!!!!!!!!!!!!!!
-% TODO: add step-by-step  derivation of transcription formula
-%!!!!!!!!!!!!!!!!!!!!!!!!!!!!!!!!!!!!!!!!!!!!!!!!!!!!!!!!!!!!!!!
+<!-- TODO: add step-by-step  derivation of transcription formula -->
 
-% YSA: Graag wat meer uitleg in de tekst over al deze formules, definitie van notaties en zo.
+<!-- YSA: Graag wat meer uitleg in de tekst over al deze formules, definitie van notaties en zo. -->
 
 \begin{table}[ht]
-	\caption{\textbf{Reactions affecting the abundance levels of pre-mRNA $\w G$, mRNA $\x G$ and proteins $\y G$ of gene $G$.} Define the set of regulators of $G$ as $\rs{G}$, the set of upregulating regulators of $G$ as $\rp G$, and the set of downregulating regulators of $G$ as $\rn G$.
-Parameters used in the propensity formulae are defined in Table \ref{tab:reaction_params}.} \label{tab:reaction_def}
+    \caption{\textbf{Default parameters defined for the calculation of reaction propensity functions.}} \label{tab:reaction_def}
     \centering
-	\begin{tabular}{|lcc|}
+    \begin{tabular}{|lrl|}
             \hline
-		Reaction & Effect & Propensity \\ \hline \hline
-		Transcription & $\emptyset \rightarrow \w G$ & $\wpr G \times \frac{\ba G - \ind{G}^{|\rp{G}|} + \prod\limits_{H \in \rp{G}}(\ind G + \reg{G,H})}{\prod\limits_{H \in \rs{G}}(1 + \reg{G,H})}$ \\
-		Splicing & $\w G \rightarrow \x G$ & $\wsr G \times \w G$ \\
-		mRNA degradation & $\x G \rightarrow \emptyset$ & $\xdr G \times \x G$ \\
-		Translation & $\x G \rightarrow \w G + \y G$ & $\ypr G \times \x G$ \\
-		Protein degradation & $\y G \rightarrow \emptyset$ & $\ydr G \times \y G$ \\ \hline
+            Parameter & Symbol & Definition \\ \hline \hline
+            Transcription rate & $\wpr{G}$ & $\in N(100, 20),\ \geq 10$ \\
+            Splicing rate & $\wsr G$ & $\in N(10, 2),\ \geq 2$ \\
+            mRNA degradation rate & $\xdr{G}$ & $\in N(5, 1),\ \geq 2$ \\
+            Translation rate & $\ypr{G}$ & $\in N(5, 1),\ \geq 2$ \\
+            Protein degradation rate & $\ydr G$ & $\in N(3, 0.5),\ \geq 1$ \\
+            Interaction strength & $\str{G,H}$ & $\in 10^{U(0, 2)}$ * \\
+            Interaction cooperativity & $\co{G,H}$ & $\in U(0.5, 2)$ * \\
+        Independence factor & $\ind G$ & $\in [0, 1]$ * \\ \hline\hline
+            TF concentration at half-maximal binding & $\hmy H$ & $= 0.5 \times \frac{\wpr H \times \ypr H}{\xdr H \times \ydr H}$ \\
+            Regulation activity & $\reg{G,H}$ & $= \left(\str{G,H} \times \frac{\y H}{\hmy H}\right) ^ {\co{G,H}}$ \\
+            Basal expression & $\ba G$ & $= \begin{cases} 1 & \mbox{if } \rp{G} = \emptyset \\ 0.0001 & \mbox{if } \rn{G} = \emptyset \mbox{ and } \rp{G} \neq \emptyset \\ 0.5 & \mbox{otherwise} \end{cases}$ * \\ \hline
+            \multicolumn{3}{l}{*: unless $G$ is a TF, then the value is determined by the backbone.}
     \end{tabular}
@@
 
+<!-- 
 % regG,H represents a 'probability' that H is bound to G (though does not yet scale between [0,1]).
@@
 % https://kar.kent.ac.uk/24077/1/myzabetpaper.pdf
+-->
 
@@
 
-\subsection{Compute average expression along backbone transitions} \label{sec:dyngen_sim_backbone}
-When simulating the developmental backbone, we go through the edges of the backbone state network defined in an earlier step (Section \ref{sec:backbone}), starting from the root state.
+## Compute average expression along backbone transitions
+When simulating the developmental backbone, we go through the edges of the backbone state network defined in an earlier step, starting from the root state.
 It is assumed the root state has no modules active and has no expression of any molecules.
 To get to the next state, we follow a transition starting from the root state, activate and deactivate the modules as indicated by the transition, and compute the average molecule abundance along the transition.
-To compute the average abundance, we perform small time steps $t = 0.001$ and let each reaction (Section \ref{sec:reactions}) occur $t$ times its propensity.
+To compute the average abundance, we perform small time steps $t = 0.001$ and let each reaction occur $t$ times its propensity.
 
-% mention that this is basically an ode?
+<!--  mention that this is basically an ode? -->
 
-\subsection{Simulate single cells} \label{sec:dyngen_ssa}
-%TODO: STOCKS \cite{kierzek_stocksstochastickinetic_2002} uses GSSA to simulate something.
+## Simulate single cells
+
+<!-- 
+%TODO: STOCKS [@kierzek_stocksstochastickinetic_2002] uses GSSA to simulate something.
 % maybe we could have a look at that!
@@
 %TODO: improve this section a lot. Basically, explain SSA and GillespieSSA2.
+-->
 
-\texttt{dyngen} uses Gillespie's stochastic simulation algorithm (SSA)\cite{gillespie_exactstochasticsimulation_1977} to simulate dynamic processes.
+`dyngen` uses Gillespie's stochastic simulation algorithm (SSA) [@gillespie_exactstochasticsimulation_1977] to simulate dynamic processes.
 An SSA simulation is an iterative process where at each iteration one reaction is triggered.
@@
 
-\texttt{GillespieSSA2} is an optimised library for performing SSA simulations.
+`GillespieSSA2` is an optimised library for performing SSA simulations.
 The propensity functions are compiled to C++, and SSA approximations can be used which allow to trigger many reactions simultaneously at each iteration.
@@
 
-\subsubsection{Map SSA simulations to backbone}
+### Map SSA simulations to backbone
 We compute the Pearson correlation between the state vectors in the simulation and the average expression levels along a transition in the backbone.
@@
 
-\subsection{Simulate experiment}
+## Simulate experiment
 From the SSA simulation we obtain the abundance levels of all the molecules at the different time points.
@@
 
-\subsubsection{Sample cells}
+### Sample cells
 Cells can be sampled from an unsynchronised population of single cells (snapshot) or at multiple time points in a synchronised population (time series).
 
-\paragraph{Snapshot} Cells are sampled uniformly from the different time points in the simulation.
+**Snapshot:** Cells are sampled uniformly from the different time points in the simulation.
 
-\paragraph{Time series} The timeline of the simulations is divided into multiple chunks. From several of these chunks, cells are sampled. For each cell it is known at which time point it was sampled.
+**Time series:** The timeline of the simulations is divided into multiple chunks. From several of these chunks, cells are sampled. For each cell it is known at which time point it was sampled.
 
-\subsubsection{Sample molecules} 
+### Sample molecules
 Molecules are sampled from the simulation to replicate how molecules are experimentally sampled.
-A real dataset is downloaded from a repository of single-cell RNA-seq datasets\cite{cannoodt_singlecellomicsdatasets_2018}.
-For each \textit{in silico} cell $i$, draw its library size $ls_i$from the distribution of transcript counts per cell in the real dataset.
-The capture rate $cr_j$ of each \textit{in silico} molecule type $j$ is drawn from $N(1, 0.05)$.
+A real dataset is downloaded from a repository of single-cell RNA-seq datasets [@cannoodt_singlecellomicsdatasets_2018].
+For each _in silico_ cell $i$, draw its library size $ls_i$from the distribution of transcript counts per cell in the real dataset.
+The capture rate $cr_j$ of each _in silico_ molecule type $j$ is drawn from $N(1, 0.05)$.
 Finally, for each cell $i$, draw $ls_i$ molecules from the multinomial distribution with probabilities $cr_j \times ab_{i,j}$ with $ab_{i,j}$ the molecule abundance level of molecule $j$ in cell $i$.
 
-\clearpage
-\section{References}
-\printbibliography[heading=none]
+# Author contributions
+
+* R.C. and W.S. designed the study.
+* R.C. and W.S. performed the experiments and analysed the data.
+* R.C. and W.S. implemented the `dyngen` software package.
+* R.C. wrote the original manuscript.
+* R.C., W.S., and Y.S. reviewed and edited the manuscript.
+* Y.S. supervised the project.

### 2019-10-30 render_manuscript.R

@@ -1,0 +1,24 @@
+library(tidyverse)
+library(googledrive)
+library(rmarkdown)
+
+# copy zotero library if on rcannood's computer
+if (Sys.info()[["user"]] == "rcannood") {
+  file.copy("~/Workspace/library.bib", "library.bib", overwrite = TRUE)
+}
+
+# download Rmd from google drive
+httr::set_config(httr::config(http_version = 0)) # avoid http2 framing layer bug
+drive <- drive_download(as_id("1BOltsUJBQ8NhF5s9kU_Q470U0U00QblbG8HBgQwEzRE"), type="text/plain", overwrite=TRUE, path = tempfile())
+system(str_glue("sed -i '1s/^.//' {drive$local_path}")) # remove first character, because this is some strange unicode character added by google
+system(str_glue("sed -i 's/ *\\(\\[@[^\\]]*\\]\\)/\\1/' {drive$local_path}")) # remove spaces before citations
+system(str_glue("sed -i 's/^\\(\\#\\#* \\)/\\n\\1/' {drive$local_path}")) # add enters before headings
+system(str_glue("cat {drive$local_path} > manuscript.Rmd"))
+
+# render the manuscript
+render("manuscript.Rmd")
+
+# copy content to content.tex
+lines <- read_lines("manuscript.tex")
+content <- lines[seq(which(lines == "\\maketitle")+1, which(lines == "\\printbibliography")-1)]
+write_lines(content, "content.tex")

### 2019-10-30 manuscript.Rmd
update

@@
 Several examples of module networks are given (Figure \ref{fig:example_backbones}).
-A simple chain of modules (where one module upregulates the next) results in a \emph{linear} process.
-By having the last module repress the first module, the process becomes \emph{cyclic}.
-Two modules repressing eachother is the basis of a \emph{bifurcating} process, though several chains of modules have to be attached in order to achieve progression before and after the bifurcation process.
+A simple chain of modules (where one module upregulates the next) results in a _linear_ process.
+By having the last module repress the first module, the process becomes _cyclic_.
+Two modules repressing eachother is the basis of a _bifurcating_ process, though several chains of modules have to be attached in order to achieve progression before and after the bifurcation process.
 Finally, a \emph{converging} process has a bifurcation occurring during the burn phase, after which any differences in module regulation is removed.
@@
     \includegraphics[width=0.8\textwidth]{fig/example_backbones}
-    \caption{Example module networks}
+    \caption{\textbf{Example module networks.}}
     \label{fig:example_backbones}
@@
     \includegraphics[width=0.8\textwidth]{fig/backbone_lego}
-    \caption{Backbone lego}
+    \caption{\textbf{Backbone lego.}}
     \label{fig:backbone_lego}
@@
 When the input node of a **linear BBL** (Figure \ref{fig:backbone_lego}C) is upregulated, the module the BBL is connected to will be upregulated.
-A \emph{simple chain} is a set of modules where a module upregulates the next.
-A \emph{chain with double repression} has an uneven number of modules forming a chain where each module downregulates the next but all modules (except the input) have basal expression.
-A \emph{grid with double repression} is similar; except that modules do not have basal expression but instead get upregulated by an upstream module in the chain.
-Finally, a \emph{flip flop} consists of a simple chain where first the modules (except the last) are upregulated.
+A _simple chain_ is a set of modules where a module upregulates the next.
+A _chain with double repression_ has an uneven number of modules forming a chain where each module downregulates the next but all modules (except the input) have basal expression.
+A _grid with double repression_ is similar; except that modules do not have basal expression but instead get upregulated by an upstream module in the chain.
+Finally, a _flip flop_} consists of a simple chain where first the modules (except the last) are upregulated.
 Once the second to last module is upregulated, that module upregulates itself and the first module is strongly repressed, causing all other modules to lose expression and finally the last module to be upregulated.
-The \emph{flip flop} retains this output state, even when the input changes.
+The _flip flop_ retains this output state, even when the input changes.
 

### 2019-11-08 analysis/vignette_trajectory_retardation.Rmd
usecase trajectory alignment

---
title: "Alignment of slowed down trajectories"
author: "Louise Deconinck"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Put the title of your vignette here}
  %\VignetteEngine{knitr::rmarkdown}
  \usepackage[utf8]{inputenc}
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## Alignment of slowed trajectories

```{r collapse=TRUE}
library(gtools)
library(tidyverse)
library(dyngen)
library(dynplot)
```

### Simulating data

We can use dyngen to generate different kinds of trajectories and align these with a known trajectory alignment technique.

First, we generate a normal linear trajectory. We construct a very simple linear backbone consisting of 20 modules and 3 milestones. We then specifiy the model will generate 5000 cells, use 30 transcription factors and generate 500 housekeeping genes and targets. We're also going to perform 50 different simulations from which we'll sample cells.

```{r eval=FALSE}

backbone <- bblego(
  bblego_start("A", type = "simple", num_modules = 4),
  bblego_linear("A", "B", type = "simple", num_modules = 5),
  bblego_linear("B", "C", type = "simple", num_modules = 6),
  bblego_end("C", type = "simple", num_modules = 5)
)

model <- initialise_model(
  backbone = backbone,
  num_cells = 5000,
  num_tfs = 30,
  num_targets = 500,
  num_hks = 500,
  simulation_params = simulation_default(num_simulations=50)
)

```

We generate a single-cell dataset in the standard way.

```{r eval=FALSE}
model <- generate_tf_network(model)
model <- generate_feature_network(model)
model <- generate_kinetics(model)
model <- generate_gold_standard(model)
model <- generate_cells(model)
model <- generate_experiment(model)
dataset1 <- wrap_dataset(model)
```

Next, we generate a dataset using the same feature network, gene regulatory network and kinetics, but with the transcription rate and mrna degradation rate of a single transcription factor slowed down. The transcription rate of a cell is regulated by the "wpr" parameter, the mrna degradation by the "xdr" parameter. These are slowed in the following way:

```{r eval=FALSE}
# Target is the transcription factor targeted
# Rate is the amount by which the transcription factor will be slowed
change_speed <- function(model, target, rate) {
  param_id <- which(target == model$feature_info$feature_id)[[1]]

  model$feature_info$wpr[param_id] <- model$feature_info$wpr[param_id] * rate
  model$feature_info$xdr[param_id] <- model$feature_info$xdr[param_id] * rate

  param_wpr <- paste0("wpr_", target)
  param_xdr <- paste0("xdr_", target)

  model$simulation_system$parameters[param_wpr] <- model$simulation_system$parameters[param_wpr] * rate
  model$simulation_system$parameters[param_xdr] <- model$simulation_system$parameters[param_xdr] * rate

  model
}
```

We can then apply this function and regenerate the gold standard, the cells and the actual single cell experiment to obtain a second, slowed dataset.

```{r eval=FALSE}
model2 <- model
model2 <- change_speed(model1, "A5_TF1", 0.5)
model2 <- generate_gold_standard(model2)
model2 <- generate_cells(model2)
model2 <- generate_experiment(model2)
dataset2 <- wrap_dataset(model2)
```

These are pretty large models. Running will take significant time. We load the two accompanying datasets that were generated using the procedure above.

```{r}
model1 <- readRDS("data/8_orig_model")
model2 <- readRDS("data/8_retardation_model_1")
dataset1 <- wrap_dataset(model1)
dataset2 <- wrap_dataset(model2)
```

### Smooth gene expression

In order to align these two trajectories, we will first generate waypoints or pseudocells along the trajectory. We generate 100 waypoints, evenly spaced through the pseudotime.

Waypoints consist of cell_ids, the id's of these waypoints, the milestone they originate from and are travelling to, and the pseudotime percentage between those two milestones.
In order to distribute these waypoints evenly, we generate the amount of waypoints according to the length of the pseudotime trajectory part.
```
A -------> B --------------> EndB
    0.5            1
```

This means that if we have the above trajectory, with the segment between milestone A and B being half as long as the segment between B and EndB, there will be half as much waypoints between A and B compared to the number of waypoints between B and EndB.

```{r}
get_waypoint_progression <- function(dataset, nr){
  cell_id <- character()
  froms <- character()
  tos <- character()
  percentage <- numeric()

  length_pieces <- dataset$milestone_network$length/sum(dataset$milestone_network$length)
  idx <- 1
  for(piece in length_pieces){
    amount <- round(nr * piece)
    wps <- modify(list(1:amount), function(x){x/amount})[[1]]
    ids <- modify(list(1:amount), function(x){sprintf("w%d_%03d", idx, x)})[[1]]
    from <- rep(dataset$milestone_network$from[idx], amount)
    to <- rep(dataset$milestone_network$to[idx], amount)

    cell_id <- c(cell_id, ids)
    froms <- c(froms, from)
    tos <- c(tos, to)
    percentage <- c(percentage, wps)

    idx <- idx + 1

  }
  waypoint_progressions <- tibble(cell_id=cell_id, from=froms, to=tos, percentage=percentage)
  waypoint_progressions
}

wpp1 <- get_waypoint_progression(dataset1, 100)
wpp2 <- get_waypoint_progression(dataset2, 100)
```

Next, we calculate the geodesic distance between these waypoints and all other cells. The geodesic distance is the distance along the trajectory.
First, the progressions calculated in the previous step need to be converted to milestone percentages using the dynwrap function.
Then, we calculate the geodesic distances.

```{r}
get_geodesic_distances_from_progressions <- function(dataset, waypoint_progressions){
  waypoint_milestone_percentages <- dynwrap::convert_progressions_to_milestone_percentages(cell_id, dataset$milestone_ids, dataset$milestone_network, waypoint_progressions)
  waypoint_milestone_percentages <- rename(waypoint_milestone_percentages, waypoint_id = cell_id)

  geodesic_distances <- dynwrap::calculate_geodesic_distances(dataset, waypoint_milestone_percentages = waypoint_milestone_percentages, directed=FALSE)
  geodesic_distances
}

gd1 <- get_geodesic_distances_from_progressions(dataset1, wpp1)
gd2 <- get_geodesic_distances_from_progressions(dataset2, wpp2)
```

Now we smooth the expression of the genes over these waypoints using an Euclidean kernel. We interpolate the expression of the genes at the pseudotimes of the waypoints. The influence of a single cell is determined by the proximity (using the geodesic distances) to the waypoint.

```{r}
interpolate_expression <- function(dataset, wps, gds){
  weighted_expressions <- do.call('cbind', lapply(seq_along(wps), function(idx){
    # Get the geodesic distance of this waypoint to all other cells
    dist <- gds[idx,]
    # Apply Euclidean kernel
    weighted <- exp(-(dist^2)/(0.1)^2)
    # Normalize
    weighted <- weighted/sum(weighted)
    # Calculate expression
    weighted_exp <- t(as.matrix(dataset$counts)) %*% weighted
    weighted_exp
  }))
  weighted_expressions
}

wexpr1 <- interpolate_expression(dataset1, wpp1$percentage, gd1)
wexpr2 <- interpolate_expression(dataset2, wpp2$percentage, gd2)
```

### Align using DTW

Now that we have smoothed gene expression values for the waypoints, we can align the two trajectories using dynamic time warping.

```{r collapse=TRUE}
library(dtw)

d <- dist(as.matrix(t(wexpr1)), as.matrix(t(wexpr2)), dist.method="Euclidean")
d <- (d - min(d)) / (max(d) - min(d))

alignment <- dtw(d, step.pattern = symmetric2, keep.internals = TRUE)
dtwPlotDensity(alignment)
abline(coef = c(0,1))

```

### 2019-11-08 analysis/dtw_functions.R

@@ -1,0 +1,80 @@
+library(tidyverse)
+library(dyngen)
+library(dynplot)
+library(gtools)
+library(dtw)
+
+calculate_correct_pseudotime <- function(dataset){
+  length_pieces <- dataset$milestone_network$length/sum(dataset$milestone_network$length)
+  nr_piece <- 0
+  start <- "sA"
+  traj <- numeric()
+  names_traj <- list()
+  # TODO convert to for for each piece
+  while(start %in% dataset$milestone_network[["from"]]){
+    idx <- which(start == dataset$milestone_network[["from"]])[[1]]
+    end <- dataset$milestone_network[["to"]][idx]
+    to_add <- sum(length_pieces[0:nr_piece])
+
+    perc <-(dataset[["progressions"]] %>% filter(from == start & to == end))$percentage * length_pieces[nr_piece + 1] + to_add
+    nams <-(dataset[["progressions"]] %>% filter(from == start & to == end))$cell_id
+
+    traj <- c(traj, perc)
+    names_traj <- c(names_traj, nams)
+
+    start <- end
+    print(start)
+    nr_piece <- nr_piece + 1
+
+  }
+
+  names(traj) <- names_traj
+  traj <- traj[mixedorder(names(traj))]
+  traj
+}
+
+get_geodesic_distances_from_progressions <- function(dataset, waypoint_progressions){
+  waypoint_milestone_percentages <- dynwrap::convert_progressions_to_milestone_percentages(cell_id, dataset$milestone_ids, dataset$milestone_network, waypoint_progressions)
+  waypoint_milestone_percentages <- rename(waypoint_milestone_percentages, waypoint_id = cell_id)
+
+  geodesic_distances <- dynwrap::calculate_geodesic_distances(dataset, waypoint_milestone_percentages = waypoint_milestone_percentages, directed=FALSE)
+  geodesic_distances
+}
+
+get_waypoint_progression <- function(dataset, nr){
+  cell_id <- character()
+  froms <- character()
+  tos <- character()
+  percentage <- numeric()
+
+  length_pieces <- dataset$milestone_network$length/sum(dataset$milestone_network$length)
+  idx <- 1
+  for(piece in length_pieces){
+    amount <- round(nr * piece)
+    wps <- modify(list(1:amount), function(x){x/amount})[[1]]
[…]

### 2019-11-08 analysis/usecase_trajectory_retardation.R

@@ -1,0 +1,280 @@
+library(tidyverse)
+library(dyngen)
+library(dynplot)
+library(gtools)
+library(dtw)
+
+setwd("/home/louisedc/Work/dyngen_analysis/analysis")
+source("dtw_functions.R")
+
+# # Function that will change the speed of the targeted gene.
+# # TODO: what is wpr & xdr again?
+# change_speed <- function(model, target, rate) {
+#   param_id <- which(target == model$feature_info$feature_id)[[1]]
+#
+#   model$feature_info$wpr[param_id] <- model$feature_info$wpr[param_id] * rate
+#   model$feature_info$xdr[param_id] <- model$feature_info$xdr[param_id] * rate
+#
+#   param_wpr <- paste0("wpr_", target)
+#   param_xdr <- paste0("xdr_", target)
+#
+#   model$simulation_system$parameters[param_wpr] <- model$simulation_system$parameters[param_wpr] * rate
+#   model$simulation_system$parameters[param_xdr] <- model$simulation_system$parameters[param_xdr] * rate
+#
+#   model
+# }
+#
+# generate_retardation <- function(model, id){
+#   model1 <- model
+#   model1 <- change_speed(model1, "A4_TF1", 0.5)
+#   model1 <- generate_gold_standard(model1)
+#   model1 <- generate_cells(model1)
+#   model1 <- generate_experiment(model1)
+#   dataset2 <- wrap_dataset(model1)
+#
+#   # save model & dataset
+#   # saveRDS(dataset2, paste0("data/retardation_ds_", id))
+#   # saveRDS(model1, paste0("analysis/data/retardation_model_", id))
+#
+#   # for(i in 1:9){
+#   #   generate_replication(model1, i)
+#   # }
+#
+#   dataset2
+# }
+#
+# generate_replication <- function(model, id){
+#   model <- generate_cells(model)
+#   model <- generate_experiment(model)
+#   saveRDS(model, paste0("analysis/data/retardation_replication_", id))
+# }
+#
+# # Construct a synthetic model
+# # A very simple linear model
+# backbone <- bblego(
+#   bblego_start("A", type = "simple", num_modules = 4),
[…]

### 2019-11-18 analysis/vignette_network_inference.Rmd
commit derivative code for GRN NI

---
title: "Use-case: evaluation of network inference"
author: "Robrecht Cannoodt"
output: rmarkdown::html_vignette
vignette: >
  \usepackage[utf8]{inputenc}
  %\VignetteIndexEntry{Use-case: evaluation of network inference}
  %\VignetteEngine{knitr::rmarkdown}
editor_options: 
  chunk_output_type: console
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

```{r}
library(tidyverse)
library(dyngen)

set.seed(1)

backbone <- backbone_bifurcating_cycle()
num_kos <- 40
model <- 
  initialise_model(
    num_cells = 1000,
    num_tfs = 200,
    num_targets = 900,
    num_hks = 900,
    distance_metric = "pearson",
    backbone = backbone,
    tf_network_params = tf_network_default(min_tfs_per_module = 2, sample_num_regulators = function() 2),
    feature_network_params = feature_network_default(target_resampling = 5000),
    kinetics_params = kinetics_default(),
    gold_standard_params = gold_standard_default(),
    simulation_params = simulation_default(
      census_interval = 1,
      burn_time = simtime_from_backbone(backbone, TRUE),
      total_time = simtime_from_backbone(backbone),
      experiment_params = bind_rows(
        simulation_type_wild_type(num_simulations = 32),
        simulation_type_knockdown(num_simulations = num_kos, num_genes = sample(1:10, num_kos, replace = TRUE))
      ),
      store_reaction_propensities = TRUE,
      store_grn = TRUE
    ),
    experiment_params = experiment_snapshot(),
    verbose = TRUE,
    download_cache_dir = "~/.cache/dyngen",
    num_cores = 8
  )

model <- model %>% 
  generate_tf_network() %>% 
  generate_feature_network() %>% 
  generate_kinetics() %>% 
  generate_gold_standard() %>%
  generate_cells() %>% 
  generate_experiment()

traj <- 
  model %>% 
  wrap_dataset(store_grn = TRUE)

traj$expression

reactions <- model$simulation_system$reactions 
names(reactions) <- map_chr(reactions, "name")

comp_reac <- dyngen:::.generate_cells_precompile_reactions(model)
sim <- dyngen:::.generate_cells_simulate_cell(
  simulation_i = 1, model, comp_reac, debug = TRUE
)

feature_network <- model$feature_network

knockouts <- 
  bind_rows(pbapply::pblapply(
    seq_len(nrow(feature_network)), 
    function(i) {
      from <- feature_network$from[[i]]
      to <- feature_network$to[[i]]
      reac_ix <- match(paste0("transcription_", to), comp_reac$reaction_ids)
      from_ix <- match(paste0("y_", from), names(sim$initial_state))

      sim$state <- model$simulations$counts[traj$cell_info$step_ix[[1]],] %>% unname
      sim$calculate_propensity()

      orig_prop <- sim$propensity[[reac_ix]]

      sim$state[[from_ix]] <- 0
      sim$calculate_propensity()
      new_prop <- sim$propensity[[reac_ix]]

      tibble(
        i,
        from,
        to,
        orig_prop,
        new_prop,
        reg_effect = orig_prop - new_prop,
        buffer_val = traj$feature_network_sc[1,paste0("regulation_", from, "_", to)]
      )
    }
  ))
max_prot <- model$feature_info %>% select(feature_id, max_protein) %>% deframe()
wpr <- model$feature_info %>% select(feature_id, wpr) %>% deframe()
xdr <- model$feature_info %>% select(feature_id, xdr) %>% deframe()
ggplot(knockouts) +
  geom_point(aes(log10(buffer_val), reg_effect / (wpr[to] - xdr[to]), colour = feature_network$effect %>% as.character)) +
  theme_bw()

feature_network$derivative <- 
  pbapply::pblapply(seq_len(nrow(feature_network)), function(i) {
  from <- feature_network$from[[i]]
  to <- feature_network$to[[i]]

  reaction <- reactions[[paste0("transcription_", to)]]
  prop <- gsub("pow\\(([^,]*) *, *([^\\)]*)\\)", "(\\1)^(\\2)", reaction$propensity)
  props <- prop %>% strsplit(" *; *") %>% first()
  propone <- props %>% last()
  subs <- tibble(
    var = gsub(" *=.*", "", props[-length(props)]),
    val = gsub(".*= *", "", props[-length(props)])
  )
  for (j in seq_len(nrow(subs))) {
    propone <- gsub(subs$var[[j]], subs$val[[j]], propone, fixed = TRUE)
  }

  deriv <- D(parse(text = propone), paste0("y_", from))
})

state <- c(
  model$simulations$counts[traj$cell_info$step_ix[[1]],],
  model$simulation_system$parameters
)
deriv <- feature_network$derivative[[1]]
# list2env(as.list(state), .GlobalEnv)
# eval(deriv)
env <- new.env()
list2env(as.list(state), env)

derivs <- map_dbl(feature_network$derivative, ~ eval(., envir = env))
reac_props <- model$simulations$reaction_propensities[traj$cell_info$step_ix[[1]], paste0("transcription_", model$feature_network$to)]

qplot(
  log10(traj$feature_network_sc[1,]),
  derivs,
  colour = reac_props
  # colour = as.character(feature_network$effect)
) + labs(
  x = "log10 sc importance = log10((y/k)^c)",
  y = "deriv. propensity function", 
  colour = "effect"
) +
  theme_bw() +
  viridis::scale_colour_viridis()
ggsave("~/deriv.pdf", width = 10, height = 10)

qplot(
  log10(traj$feature_network_sc[1,]),
  knockouts$reg_effect,
  colour = as.character(feature_network$effect)
) + labs(
  x = "log10 sc importance = log10((y/k)^c)",
  y = "KO", 
  colour = "effect"
) +
  theme_bw()

qplot(
  reac_props,
  derivs
) +
  theme_bw()

qplot(
  reac_props,
  log10(traj$feature_network_sc[1,])
) +
  theme_bw()
```

### 2019-11-18 analysis/vignette_network_inference.Rmd
use knockout ground truths

@@
   ))
-max_prot <- model$feature_info %>% select(feature_id, max_protein) %>% deframe()
-wpr <- model$feature_info %>% select(feature_id, wpr) %>% deframe()
-xdr <- model$feature_info %>% select(feature_id, xdr) %>% deframe()
+fito <- model$feature_info %>% slice(match(knockouts$to, feature_id)) %>% 
+  mutate(label = ifelse(is.na(module_id), gsub("[0-9]*$", "", feature_id), gsub("[0-9]*$", "", module_id)))
 ggplot(knockouts) +
-  geom_point(aes(log10(buffer_val), reg_effect / (wpr[to] - xdr[to]), colour = feature_network$effect %>% as.character)) +
+  geom_point(aes(log10(buffer_val), reg_effect / fito$wpr, colour = fito$label)) +
   theme_bw()
 
-feature_network$derivative <- 
-  pbapply::pblapply(seq_len(nrow(feature_network)), function(i) {
-  from <- feature_network$from[[i]]
-  to <- feature_network$to[[i]]
-
-  reaction <- reactions[[paste0("transcription_", to)]]
-  prop <- gsub("pow\\(([^,]*) *, *([^\\)]*)\\)", "(\\1)^(\\2)", reaction$propensity)
-  props <- prop %>% strsplit(" *; *") %>% first()
-  propone <- props %>% last()
-  subs <- tibble(
-    var = gsub(" *=.*", "", props[-length(props)]),
-    val = gsub(".*= *", "", props[-length(props)])
-  )
-  for (j in seq_len(nrow(subs))) {
-    propone <- gsub(subs$var[[j]], subs$val[[j]], propone, fixed = TRUE)
-  }
-
-  deriv <- D(parse(text = propone), paste0("y_", from))
-})
-
-state <- c(
-  model$simulations$counts[traj$cell_info$step_ix[[1]],],
-  model$simulation_system$parameters
-)
-deriv <- feature_network$derivative[[1]]
-# list2env(as.list(state), .GlobalEnv)
-# eval(deriv)
-env <- new.env()
-list2env(as.list(state), env)
-
-derivs <- map_dbl(feature_network$derivative, ~ eval(., envir = env))
-reac_props <- model$simulations$reaction_propensities[traj$cell_info$step_ix[[1]], paste0("transcription_", model$feature_network$to)]
-
-qplot(
-  log10(traj$feature_network_sc[1,]),
-  derivs,
-  colour = reac_props
-  # colour = as.character(feature_network$effect)
-) + labs(
-  x = "log10 sc importance = log10((y/k)^c)",
-  y = "deriv. propensity function", 
-  colour = "effect"
-) +
+ggplot(knockouts) +
+  geom_point(aes((buffer_val), reg_effect / fito$wpr, colour = factor(fito$basal))) +
   theme_bw() +
-  viridis::scale_colour_viridis()
-ggsave("~/deriv.pdf", width = 10, height = 10)
-
-qplot(
-  log10(traj$feature_network_sc[1,]),
-  knockouts$reg_effect,
-  colour = as.character(feature_network$effect)
-) + labs(
-  x = "log10 sc importance = log10((y/k)^c)",
-  y = "KO", 
-  colour = "effect"
-) +
-  theme_bw()
-
-qplot(
-  reac_props,
-  derivs
-) +
-  theme_bw()
-
-qplot(
-  reac_props,
-  log10(traj$feature_network_sc[1,])
-) +
-  theme_bw()
+  scale_x_log10()
 ```

### 2019-11-21 manuscript.Rmd
update rmd and others

@@
 - Wouter Saelens
+- Louise Deconinck
 - Yvan Saeys
@@
 We demonstrate `dyngen`'s flexibility by simulating snapshot, time-series and perturbation experiments.
-**Conclusion:** `dyngen` lays the foundations for benchmarking a wide variety of computational single-cell tools, and can be used to help kick-start the development of future types of analyses.
+**Conclusion:** `dyngen` lays the foundations for benchmarking a wide variety of computational single-cell tools and can be used to help kick-start the development of future types of analyses.
 
-# Introduction
+# Introduction {#sec:dyngen-introduction}
 Continuous technological advancements to single-cells omics are having profound effects on how researchers can validate biological hypotheses.
@@
 Generators of scRNA-seq data (e.g.
-splatter [@zappia_splattersimulationsinglecell_2017], powsimR [@vieth_powsimrpoweranalysis_2017], PROSSTT [@papadopoulos_prossttprobabilisticsimulation_2018] and SymSim [@zhang_simulatingmultiplefaceted_2019]) have already been widely used to explore the strengths and weaknesses of computational tools, both by method developers [@street_slingshotcelllineage_2018; @parra_reconstructingcomplexlineage_2018; @lummertzdarocha_reconstructioncomplexsinglecell_2018; @lin_scclassifyhierarchicalclassification_2019] and independent benchmarkers [@duo_systematicperformanceevaluation_2018; @saelens_comparisonsinglecelltrajectory_2019; @soneson_biasrobustnessscalability_2018].
+splatter[@zappia_splattersimulationsinglecell_2017], powsimR[@vieth_powsimrpoweranalysis_2017], PROSSTT[@papadopoulos_prossttprobabilisticsimulation_2018] and SymSim[@zhang_simulatingmultiplefaceted_2019]) have already been widely used to explore the strengths and weaknesses of computational tools, both by method developers[@street_slingshotcelllineage_2018; @parra_reconstructingcomplexlineage_2018; @lummertzdarocha_reconstructioncomplexsinglecell_2018; @lin_scclassifyhierarchicalclassification_2019] and independent benchmarkers[@duo_systematicperformanceevaluation_2018; @saelens_comparisonsinglecelltrajectory_2019; @soneson_biasrobustnessscalability_2018].
 However, a limitation of scRNA-seq profiles generators is that they would require significant methodological alterations to add additional modalities.
 
-An ideal experiment would be able to observe all aspects of a cell, including a full history of its molecular states, spatial positions and environmental interactions [@stuart_integrativesinglecellanalysis_2019].
+An ideal experiment would be able to observe all aspects of a cell, including a full history of its molecular states, spatial positions and environmental interactions[@stuart_integrativesinglecellanalysis_2019].
 While this falls outside the reach of current experimental technologies, generating synthetic data in anticipation of new experimental technologies would allow already developing the next wave of computational tools.
 
-We introduce `dyngen`, a multi-modality simulator of single cells and their dynamics (Figure \ref{fig:showcase}A).
-A cell is simulated using Gillespie's Stochastic Simulation Algorithm (SSA) [@gillespie_exactstochasticsimulation_1977] where a cell consists of a set of molecules.
+We introduce `dyngen`, a multi-modality simulator of single cells and their dynamics (Figure\ \ref{fig:showcase}A).
+A cell is simulated using Gillespie's Stochastic Simulation Algorithm (SSA)[@gillespie_exactstochasticsimulation_1977] where a cell consists of a set of molecules.
 Throughout a simulation, reactions (e.g.
-transcription, splicing) modify the abundance of these molecules, and the likelihood of a reaction occurring is again dependent on molecule abundance.
+transcription, splicing) modify the abundance of these molecules.
+The likelihood of a reaction occurring is again dependent on molecule abundance.
 
-By simulating a cell over time in terms of its molecular state and the reactions that are allowed to occur, the simulator is more extendible to new modalities or experimental procedures (Figure \ref{fig:showcase}B).
+By simulating a cell over time in terms of its molecular state and the reactions that are allowed to occur, the simulator is more extendible to new modalities or experimental procedures (Figure\ \ref{fig:showcase}B).
 We demonstrate `dyngen`'s flexibility by simulating snapshot, time-series and perturbation experiments, allowing to benchmark a wide variety of computational tools such as trajectory alignment and differential network inference methods.
@@
 
-# Results
+# Results {#sec:dyngen-results}
 
-A cell consists of a set of molecules, the abundance of which are affected by a set of reactions: transcription, splicing, translation, and degradation (Figure \ref{fig:simplecyclic}A).
-A gene regulatory network (GRN) define the reactions that are allowed to occur (Figure \ref{fig:simplecyclic}B), and is constructed such that a cell will develop over time (Figure \ref{fig:simplecyclic}C--D).
-The likelihood of a reaction occurring is a function of the abundance of key molecules involved in each reaction (Figure \ref{fig:simplecyclic}E).
+A cell consists of a set of molecules, the abundance of which are affected by a set of reactions: transcription, splicing, translation, and degradation (Figure\ \ref{fig:simplecyclic}A).
+A gene regulatory network (GRN) defines the reactions that are allowed to occur (Figure\ \ref{fig:simplecyclic}B) and is constructed such that a cell will develop over time (Figure\ \ref{fig:simplecyclic}C--D).
+The likelihood of a reaction occurring is a function of the abundance of key molecules involved in each reaction (Figure\ \ref{fig:simplecyclic}E).
 
@@
 
-`dyngen` returns many modalities throughout the whole simulation: molecular abundance, cellular state, number of reaction firings, reaction likelihoods, and regulation activations (Figure \ref{fig:simplecyclic}C--F).
+`dyngen` returns many modalities throughout the whole simulation: molecular abundance, cellular state, number of reaction firings, reaction likelihoods, and regulation activations (Figure\ \ref{fig:simplecyclic}C--F).
 These modalities can serve both as input data and ground truth for benchmarking many types of computational approaches.
-For example, a network inference method could use mRNA abundance and cellular states as inputs, and its output could be benchmarked against the gold standard GRN.
+For example, a network inference method could use mRNA abundance and cellular states as inputs and its output could be benchmarked against the gold standard GRN.
 
 Depending on how the GRN is designed, different cellular developmental processes can be simulated.
-`dyngen` includes generators of GRNs which result in many different developmental topologies (Figure \ref{fig:example_runs}), including branching, converging, cyclic and even disconnected.
+`dyngen` includes generators of GRNs which result in many different developmental topologies (Figure\ \ref{fig:example_runs}), including branching, converging, cyclic and even disconnected.
 
@@
 Aside from simulating topologies currently not supported by `dyngen`, this has several other important use-cases.
-Simulations of the same GRN with small perturbations allow to emulate batch effects or perturbation experiments (Figure \ref{fig:usecases}).
+Simulations of the same GRN with small perturbations allow to emulate batch effects or perturbation experiments (Figure\ \ref{fig:usecases}).
 Simulating perturbed GRNs allow evaluating trajectory alignment methods -- which attempt to map two or more trajectories onto each other -- or differential network inference methods -- which infer differential regulatory interactions between two or more groups of profiles.
@@
 
-<!-- ## Example use cases -->
+## Example use cases {#sec:dyngen-example}
 
-<!-- ### Trajectory alignment -->
+### Trajectory alignment
 
-<!-- ### Differential network inference -->
+### Single-cell network inference
 
-<!-- ### RNA velocity -->
+### RNA velocity
 
-<!-- ### Perturbation experiment -->
-
-# Discussion
+# Discussion {#sec:dyngen-discussion}
 As is, `dyngen`'s single cell simulations can be used to evaluate common single-cell omics computational methods such as clustering, batch correction, trajectory inference and network inference.
@@
 This may include methods that integrate clustering, network inference and trajectory inference.
-In this respect, `dyngen` may promote the development of new tools in the single-cell field similarly as other simulators have done in the past [@schaffter_genenetweaversilicobenchmark_2011; @ewing_combiningtumorgenome_2015].
+In this respect, `dyngen` may promote the development of new tools in the single-cell field similarly as other simulators have done in the past[@schaffter_genenetweaversilicobenchmark_2011; @ewing_combiningtumorgenome_2015].
 
@@
 Currently, `dyngen` focuses on simulating cells as standalone entities that are well mixed.
-Splitting up the simulation space into separate subvolumes could pave the way to better study key cellular processes such as cell division, intercellular communication and migration [@smith_spatialstochasticintracellular_2019].
+Splitting up the simulation space into separate subvolumes could pave the way to better study key cellular processes such as cell division, intercellular communication and migration[@smith_spatialstochasticintracellular_2019].
 
-# Methods
-The workflow to generate _in silico_ single cell data consists of six main steps (Figure \ref{fig:explain_methods}).
+# Methods {#sec:dyngen-methods}
+The workflow to generate _in silico_ single cell data consists of six main steps (Figure\ \ref{fig:explain_methods}).
 
@@
     \includegraphics[width=\textwidth]{fig/explain_methods}
-    \caption{\textbf{The workflow of \texttt{dyngen} is comprised of six main steps.} \textbf{A:} The user needs to specify the desired module network or use a predefined module network.
-\textbf{B:} Each gene in a module is is regulated by one or more transcription factors from the upstream module.
-Additional target genes are generated.
+    \caption{\textbf{The workflow of \texttt{dyngen} is comprised of six main steps.} 
+    \textbf{A:} The user needs to specify the desired module network or use a predefined module network. 
+    \textbf{B:} Each gene in a module is is regulated by one or more transcription factors from the upstream module. Additional target genes are generated. 
     \textbf{C:} Each gene regulatory interaction in the GRN is converted to a set of biochemical reactions. 
@@
     \textbf{E:} Multiple Gillespie SSA simulations are run using the reactions defined in step C.
-The counts of each of the mocules at each time step are extracted.
+The counts of each of the molecules at each time step are extracted.
 Each time step is mapped to a point in the backbone.
-\textbf{F:} Multiple cells are sampled from each simulation.
-Molecules are sampled from each cell.}
+    \textbf{F:} Multiple cells are sampled from each simulation. Molecules are sampled from each cell.
+    }
     \label{fig:explain_methods}
@@
 
-## Defining the backbone: modules and states
+## Defining the backbone: modules and states {#sec:dyngen-backbone}
 
@@
 The exact way a cell chooses a certain path during its differentiation is still an active research field, although certain models have already emerged and been tested _in vivo_.
-One driver of bifurcation seems to be mutual antagonism, where two genes strongly repress each other [@rekhtman_directinteractionhematopoietic_1999; @xu_regulationbifurcatingcell_2015], forcing one of the two to become inactive [@graf_forcingcellschange_2009].
-Such mutual antagonism can be modelled and simulated [@wang_quantifyingwaddingtonlandscape_2011; @ferrell_bistabilitybifurcationswaddington_2012].
-Although the two-gene model is simple and elegant, the reality is frequently more complex, with multiple genes (grouped into modules) repressing each other [@yosef_dynamicregulatorynetwork_2013].
+One driver of bifurcation seems to be mutual antagonism, where two genes strongly repress each other[@rekhtman_directinteractionhematopoietic_1999; @xu_regulationbifurcatingcell_2015], forcing one of the two to become inactive[@graf_forcingcellschange_2009].
+Such mutual antagonism can be modelled and simulated[@wang_quantifyingwaddingtonlandscape_2011; @ferrell_bistabilitybifurcationswaddington_2012].
+Although the two-gene model is simple and elegant, the reality is frequently more complex, with multiple genes (grouped into modules) repressing each other[@yosef_dynamicregulatorynetwork_2013].
 
@@
 
-Several examples of module networks are given (Figure \ref{fig:example_backbones}).
+Several examples of module networks are given (Figure\ \ref{fig:example_backbones}).
 A simple chain of modules (where one module upregulates the next) results in a _linear_ process.
 By having the last module repress the first module, the process becomes _cyclic_.
-Two modules repressing eachother is the basis of a _bifurcating_ process, though several chains of modules have to be attached in order to achieve progression before and after the bifurcation process.
+Two modules repressing each other is the basis of a _bifurcating_ process, though several chains of modules have to be attached in order to achieve progression before and after the bifurcation process.
 Finally, a \emph{converging} process has a bifurcation occurring during the burn phase, after which any differences in module regulation is removed.
@@
     \includegraphics[width=0.8\textwidth]{fig/example_backbones}
-    \caption{\textbf{Example module networks.}}
+    \caption{
+      \textbf{Example module networks.}
+    }
     \label{fig:example_backbones}
@@
 In addition to the module network, the user also needs to define a network of cellular states called the "backbone".
-Before simulating any cells, each transition in the backbone is simulated separately to obtain the average changes in expression along that transition (Figure \ref{fig:explain_methods}D).
+Before simulating any cells, each transition in the backbone is simulated separately to obtain the average changes in expression along that transition (Figure\ \ref{fig:explain_methods}D).
 As part of the backbone, the user needs to specify which modules are allowed to alter its expression from one state to another.
@@
 Finally, to transition from S3 to S1 again, A and B are turned on again and D and E are fixed again.
-Demonstrations of the backbone will be explained in more detail in a later section.
+Demonstrations of the backbone will be explained in more detail in section \ref{sec:dyngen-simbackbone}.
 
 ### Backbone lego
-The backbone can make use of one or more "backbone lego" (BBL) pieces (Figure \ref{fig:backbone_lego}).
-A BBL consists of one or more modules which regulate each other such that the output modules present a specific behaviour, depending on the input module (Figure \ref{fig:backbone_lego}A).
+The backbone can make use of one or more "backbone lego" (BBL) pieces (Figure\ \ref{fig:backbone_lego}).
+A BBL consists of one or more modules which regulate each other such that the output modules present a specific behaviour, depending on the input module (Figure\ \ref{fig:backbone_lego}A).
 Parameters allow determining the number of modules involved in the process and the number of outputs.
-Multiple BBLs can be chained together in order to intuitively create milestone networks and corresponding state networks (Figure \ref{fig:backbone_lego}B).
+Multiple BBLs can be chained together in order to intuitively create milestone networks and corresponding state networks (Figure\ \ref{fig:backbone_lego}B).
 Note that not all dynamic processes can be represented by a combination of BBLs, but they can serve as common building blocks to aid the construction of the backbone.
@@
     \includegraphics[width=0.8\textwidth]{fig/backbone_lego}
-    \caption{\textbf{Backbone lego.}}
+    \caption{
+      \textbf{Backbone lego.}
+    }
     \label{fig:backbone_lego}
@@
 
-When the input node of a **linear BBL** (Figure \ref{fig:backbone_lego}C) is upregulated, the module the BBL is connected to will be upregulated.
+When the input node of a **linear BBL** (Figure\ \ref{fig:backbone_lego}C) is upregulated, the module the BBL is connected to will be upregulated.
 A _simple chain_ is a set of modules where a module upregulates the next.
@@
 
-When the input node of a **branching BBL** (Figure \ref{fig:backbone_lego}D) is upregulated, a subset of its output modules will eventually be upregulated.
+When the input node of a **branching BBL** (Figure\ \ref{fig:backbone_lego}D) is upregulated, a subset of its output modules will eventually be upregulated.
 A \emph{simple branching} uses reciprocal inhibition to drive the upregulation of one of the output modules.
-Due to its simplicity, however, multiple output modules might be upregulated simultaneously, and over long periods of simulation time it might be possible that the choice of upregulated module changes.
-A \emph{robust branching} improves upon the simple branching by preventing upregulation of output modules until an internal branching decision has been made, and by repressing the decision mechanism to avoid other output modules being upregulated other than the one that has been chosen.
+Due to its simplicity, however, multiple output modules might be upregulated simultaneously and over long periods of simulation time it might be possible that the choice of upregulated module changes.
+A _robust branching_ improves upon the simple branching by preventing upregulation of output modules until an internal branching decision has been made, and by repressing the decision mechanism to avoid other output modules being upregulated other than the one that has been chosen.
 
-A **leaf BBL** (Figure \ref{fig:backbone_lego}E) is a linear BBL that has either no inputs or no outputs.
-A _start_ BBL is a linear BBL where the first module has basal expression, and all modules in this module will be active during the burn-in phase of the simulation.
+A **leaf BBL** (Figure\ \ref{fig:backbone_lego}E) is a linear BBL that has either no inputs or no outputs.
+A _start_ BBL is a linear BBL where the first module has basal expression, and all modules in this module will be active during the burn-in phase of the simulation (Section \ref{sec:dyngen-simbackbone}).
 An _end_ BBL is also a linear BBL with its output regulating one final module.
 
-## Generating the gene regulatory network
-The GRN is generated based on the given backbome in four main steps (Figure \ref{fig:gen_feature_network}).
+## Generating the gene regulatory network {#sec:dyngen-grn}
+The GRN is generated based on the given backbone in four main steps (Figure\ \ref{fig:gen_feature_network}).
 
@@
 Each TF is assigned to a module such that each module has at least $x$ parameters (default $x=1$).
-A TF inherits the 'burn' and 'basal expression' from the module in belongs to.
+A TF inherits the 'burn' and 'basal expression' from the module it belongs to.
 
@@
 A user-defined number of target genes are added to the GRN.
-Target genes regulated by a TF or another target gene, but is always downstream of at least one TF.
-To sample the interactions between target genes, one of the many FANTOM5 [@lizio_gatewaysfantom5promoter_2015] GRNs is sampled.
+Target genes are regulated by a TF or another target gene, but is always downstream of at least one TF.
+To sample the interactions between target genes, one of the many FANTOM5 \cite{lizio_gatewaysfantom5promoter_2015} GRNs is sampled.
 The currently existing TFs are mapped to regulators in the FANTOM5 GRN.
@@
 The interactions of the FANTOM5 GRN are first subsampled such that the maximum in-degree of each gene is $x$ (default $x=5$).
-A random gene is sampled, and a breadth-first-search is performed to sample the desired number of housekeeping genes.
+A random gene is sampled and a breadth-first-search is performed to sample the desired number of housekeeping genes.
 
-## Convert gene regulatory network to a set of reactions
+## Convert gene regulatory network to a set of reactions {#sec:dyngen-reactions}
 
@@
 \newcommand{\str}[1]{\text{str}_{#1}}
-\newcommand{\co}[1]{\text{co}_{#1}}
-\newcommand{\ind}[1]{\text{ind}_{#1}}
-\newcommand{\hmy}[1]{\text{hmy}_{#1}}
-\newcommand{\reg}[1]{\text{reg}_{#1}}
+\newcommand{\hill}[1]{\text{hill}_{#1}}
+\newcommand{\coop}[1]{\text{coop}_{#1}}
+\newcommand{\dis}[1]{\text{dis}_{#1}}
+\newcommand{\buf}[1]{\chi_{#1}}
 \newcommand{\ba}[1]{\text{ba}_{#1}}
@@
 Each reaction consists of its propensity -- a formula to calculate the probability of the reaction occurring during an infinitesimal time interval -- and the effect -- how it will affect the current state if triggered.
-The effects and propensity functions of these reactions are defined in Table \ref{tab:reaction_def}.
+The effects and propensity functions of these reactions are defined in Table\ \ref{tab:reaction_def}.
 
@@
 \begin{table}[ht]
-    \caption{\textbf{Default parameters defined for the calculation of reaction propensity functions.}} \label{tab:reaction_def}
+    \caption{
+      \textbf{Default parameters defined for the calculation of reaction propensity functions.}
+    } \label{tab:reaction_def}
     \centering
@@
             Interaction strength & $\str{G,H}$ & $\in 10^{U(0, 2)}$ * \\
-            Interaction cooperativity & $\co{G,H}$ & $\in U(0.5, 2)$ * \\
-        Independence factor & $\ind G$ & $\in [0, 1]$ * \\ \hline\hline
-            TF concentration at half-maximal binding & $\hmy H$ & $= 0.5 \times \frac{\wpr H \times \ypr H}{\xdr H \times \ydr H}$ \\
-            Regulation activity & $\reg{G,H}$ & $= \left(\str{G,H} \times \frac{\y H}{\hmy H}\right) ^ {\co{G,H}}$ \\
+            Hill coefficient & $\hill{G,H}$ & $\in U(0.5, 2)$ * \\
+        Cooperativity factor & $\coop G$ & $\in [0, 1]$ * \\ \hline\hline
+            Dissociation constant & $\dis H$ & $= 0.5 \times \frac{\wpr H \times \ypr H}{\xdr H \times \ydr H}$ \\
+            Regulation activity & $\buf{G,H}$ & $= \left(\str{G,H} \times \frac{\y H}{\dis H}\right) ^ {\hill{G,H}}$ \\
             Basal expression & $\ba G$ & $= \begin{cases} 1 & \mbox{if } \rp{G} = \emptyset \\ 0.0001 & \mbox{if } \rn{G} = \emptyset \mbox{ and } \rp{G} \neq \emptyset \\ 0.5 & \mbox{otherwise} \end{cases}$ * \\ \hline
@@
 \begin{table}[ht]
-    \caption{\textbf{Default parameters defined for the calculation of reaction propensity functions.}} \label{tab:reaction_params}
+    \caption{
+      \textbf{Default parameters defined for the calculation of reaction propensity functions.}
+    } \label{tab:reaction_params}
     \centering
@@
 
-## Compute average expression along backbone transitions
-When simulating the developmental backbone, we go through the edges of the backbone state network defined in an earlier step, starting from the root state.
+## Compute average expression along backbone transitions {#sec:dyngen-simbackbone}
+When simulating the developmental backbone, we go through the edges of the backbone state network defined in an earlier step (Section \ref{sec:dyngen-backbone}), starting from the root state.
 It is assumed the root state has no modules active and has no expression of any molecules.
 To get to the next state, we follow a transition starting from the root state, activate and deactivate the modules as indicated by the transition, and compute the average molecule abundance along the transition.
-To compute the average abundance, we perform small time steps $t = 0.001$ and let each reaction occur $t$ times its propensity.
+To compute the average abundance, we perform small time steps $t = 0.001$ and let each reaction (Section \ref{sec:dyngen-reactions}) occur $t$ times its propensity.
 
 <!--  mention that this is basically an ode? -->
-
-## Simulate single cells
+## Simulate single cells {#sec:dyngen-simcell}
 
 <!-- 
-%TODO: STOCKS [@kierzek_stocksstochastickinetic_2002] uses GSSA to simulate something.
+%TODO: STOCKS[@kierzek_stocksstochastickinetic_2002] uses GSSA to simulate something.
 % maybe we could have a look at that!
@@
 
-`dyngen` uses Gillespie's stochastic simulation algorithm (SSA) [@gillespie_exactstochasticsimulation_1977] to simulate dynamic processes.
+`dyngen` uses Gillespie's stochastic simulation algorithm (SSA)[@gillespie_exactstochasticsimulation_1977] to simulate dynamic processes.
 An SSA simulation is an iterative process where at each iteration one reaction is triggered.
@@
 `GillespieSSA2` is an optimised library for performing SSA simulations.
-The propensity functions are compiled to C++, and SSA approximations can be used which allow to trigger many reactions simultaneously at each iteration.
+The propensity functions are compiled to C++ and SSA approximations can be used which allow to trigger many reactions simultaneously at each iteration.
 The framework also allows to store the abundance levels of molecules only after a specific interval has passed since the previous census.
@@
 
-## Simulate experiment
+## Simulate experiment {#sec:dyngen-experiment}
 From the SSA simulation we obtain the abundance levels of all the molecules at the different time points.
 We need to replicate technical effects introduced by experimental protocols in order to obtain data that is similar to real data.
-For this, the cells are sampled from the simulations, and molecules are sampled for each of the cells.
+For this, the cells are sampled from the simulations and molecules are sampled for each of the cells.
 Real datasets are used in order to achieve similar data characteristics.
@@
 Molecules are sampled from the simulation to replicate how molecules are experimentally sampled.
-A real dataset is downloaded from a repository of single-cell RNA-seq datasets [@cannoodt_singlecellomicsdatasets_2018].
+A real dataset is downloaded from a repository of single-cell RNA-seq datasets[@cannoodt_singlecellomicsdatasets_2018].
 For each _in silico_ cell $i$, draw its library size $ls_i$from the distribution of transcript counts per cell in the real dataset.
@@
 
-# Author contributions
+# Author contributions {#sec:dyngen-contributions}
 

### 2019-11-21 render_manuscript.R

@@ -10,15 +10,14 @@ if (Sys.info()[["user"]] == "rcannood") {
 # download Rmd from google drive
 httr::set_config(httr::config(http_version = 0)) # avoid http2 framing layer bug
 drive <- drive_download(as_id("1BOltsUJBQ8NhF5s9kU_Q470U0U00QblbG8HBgQwEzRE"), type="text/plain", overwrite=TRUE, path = tempfile())
-system(str_glue("sed -i '1s/^.//' {drive$local_path}")) # remove first character, because this is some strange unicode character added by google
-system(str_glue("sed -i 's/ *\\(\\[@[^\\]]*\\]\\)/\\1/' {drive$local_path}")) # remove spaces before citations
-system(str_glue("sed -i 's/^\\(\\#\\#* \\)/\\n\\1/' {drive$local_path}")) # add enters before headings
-system(str_glue("cat {drive$local_path} > manuscript.Rmd"))
+read_lines(drive$local_path) %>%
+  str_replace_all(" *(\\[@[^\\]]*\\])", "\\1") %>% # remove spaces before citations
+  write_lines("manuscript.Rmd")
 
 # render the manuscript
 render("manuscript.Rmd")
 
 # copy content to content.tex
 lines <- read_lines("manuscript.tex")
-content <- lines[seq(which(lines == "\\maketitle")+1, which(lines == "\\printbibliography")-1)]
+content <- lines[seq(which(lines == "\\hypertarget{sec:dyngen-introduction}{%"), which(lines == "\\printbibliography")-1)]
 write_lines(content, "content.tex")

### 2019-11-25 analysis/usecase_trajectory_alignment.Rmd
Write part of the vignette

---
title: "Usecase Trajectory Alignment"
author: "Louise Deconinck"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Usecase Trajectory Alignment}
  %\VignetteEngine{knitr::rmarkdown}
  \usepackage[utf8]{inputenc}
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

When developing a way to align developmental trajectories, dyngen can be used as a tool to benchmark certain aspects that such a method should ideally possess. In this vignette, we will compare a couple of different approaches to trajectory alignment.

Currently, dynamic time warping is most often used to compare trajectories with each other. Sometimes, *pseudocells* are introduced and gene expression at those cells is interpolated/smoothed. We will compare both the method that uses these pseudocells and a method that does not.

In order to test this, we will generate two experiments with dyngen, but with the same underlying topology, gene regulatory network and kinetics. We do sample different cells and generate a different experiment, which will introduce some relevant noise in the datasets.
If we compare both trajectories, we expect to discover the same trajectory, with a warping path that does not deviate substantially from the diagonal.

## Load packages

```{r, loadpackages, cache=TRUE, collapse=TRUE}
library(tidyverse)
library(dyngen)
library(dynplot)
library(dtw)

```

## Generating data

First, we generate some data. The backbone present in this process will be linear. We then initialize the model, generate the transcription factors and the feature network. Next, we generate the kinetics, gold standard, cells and experiment.

```{r, generatedata, collapse=TRUE, cache=TRUE}
backbone <- bblego(
  bblego_start("A", type = "simple", num_modules = 4),
  bblego_linear("A", "B", type = "doublerep1", num_modules = 6),
  bblego_linear("B", "C", type = "doublerep2", num_modules = 6),
  bblego_end("C", type = "simple", num_modules = 4)
)

nr_cells <- 1000

model_init <-
  initialise_model(
    num_tfs = 100,
    num_targets = 200,
    num_hks = 200,
    num_cells = nr_cells,
    backbone = backbone,
    verbose = TRUE,
    download_cache_dir = "~/.cache/dyngen",
    num_cores = 8,
    distance_metric = "pearson",
    tf_network_params = tf_network_default(min_tfs_per_module = 2, sample_num_regulators = function() 1),
    simulation_params = simulation_default(num_simulations = 10, census_interval = .01),
    experiment_params = experiment_snapshot(sample_capture_rate = function(n) rnorm(n, 1, 0.1))
  ) %>%
  generate_tf_network() %>%
  generate_feature_network()

model_unperturbed <- generate_kinetics(model_init) %>%
  generate_gold_standard() %>%
  generate_cells()

# We store this in a different variable because it is easier to plot the same 
# models in the same dimensionality reduction if the experiment is not yet 
# generated. This does not cause a significant difference with regard to 
# the actual experiments.
model_unperturbed_ <- generate_experiment(model_unperturbed)
dataset_unperturbed <- wrap_dataset(model_unperturbed_)
```

As explained in the introduction, we will generate new cells and a new experiment that should contain the same trajectory.

```{r, generate_repl_data, collapse=TRUE, cache=TRUE, dependson=c("generatedata")}
model_replicate <- generate_cells(model_unperturbed)

# Same reasons as for the unperturbed data
model_replicate_ <- generate_experiment(model_replicate)
dataset_replicate <- wrap_dataset(model_replicate_)
```

If we merge these two datasets and plot them, with the included trajectories, we can definitely see that the trajectories present in the data are the same.

```{r, merge_data, collapse=TRUE, cache=TRUE, dependson=c("generatedata", "generate_repl_data")}
# Function to merge two models. Specifics are not important.
merge_models <- function(model, model1, model2){
  model12 <- model
  model12$feature_info <- model$feature_info %>%
    mutate(
      w = paste0("w_", feature_id),
      x = paste0("x_", feature_id),
      y = paste0("y_", feature_id)
    )
  m1gs <- model1$gold_standard
  m2gs <- model2$gold_standard
  model12$gold_standard <- list(
    meta = bind_rows(
      m1gs$meta %>% mutate_at(c("from", "to", "from_", "to_"), ~paste0("left_", .)),
      m2gs$meta %>% mutate_at(c("from", "to", "from_", "to_"), ~paste0("right_", .))
    ),
    counts = rbind(
      m1gs$counts,
      m2gs$counts
    ),
    network = bind_rows(
      m1gs$network %>% mutate_at(c("from", "to"), ~paste0("left_", .)),
      m2gs$network %>% mutate_at(c("from", "to"), ~paste0("right_", .))
    )
  )
  m1sim <- model1$simulations
  m2sim <- model2$simulations
  num_m1sim <- max(m1sim$meta$simulation_i)
  model12$simulations <- list(
    meta = bind_rows(
      m1sim$meta %>% mutate_at(c("from", "to"), ~paste0("left_", .)),
      m2sim$meta %>% mutate_at(c("from", "to"), ~paste0("right_", .)) %>% mutate(simulation_i = simulation_i + num_m1sim)
    ),
    counts = rbind(
      m1sim$counts,
      m2sim$counts
    ),
    regulation = rbind(
      m1sim$regulation,
      m2sim$regulation
    ),
    reaction_firings = rbind(
      m1sim$reaction_firings,
      m2sim$reaction_firings
    ),
    reaction_propensities = rbind(
      m1sim$reaction_propensities,
      m2sim$reaction_propensities
    )
  )
  model12
}

model_merged <- merge_models(model_init, model_unperturbed, model_replicate)

model_merged <- model_merged %>%
  dyngen:::calculate_dimred(dimred_premrna = FALSE) %>%
  generate_experiment()

```

```{r, plot_simulations_dimred, collapse=TRUE, fig.height=7, fig.width=7, fig.align = "center", results='hold', cache=TRUE, dependson=c("merge_data")}
dataset_merged <- wrap_dataset(model_merged)
plot_gold_simulations(model_merged)
plot_dimred(dataset_merged)
```

## Apply dynamic time warping

Next, we will apply dynamic time warping to the two trajectories. Dynamic time warping is a method that searches the lowest cost necessary to align two time series. A dissimilarity matrix is calculated, that calculates the distance between all cells in the two trajectories. Then the path in the matrix that minimizes the overall distance of the alignment is calculated. The path is constrained be the allowed steps, which can be defined in multiple ways. We will mainly use the "symmetric2" pattern, which allows the path to proceed diagonally, vertically and horizontally for one step, but the cost for a diagonal step is double. (This is because the diagonal path is twice as short as the correspoinding vertical & horizontal path.)

### Smooth gene expression
```{r, dtw_functions, include=FALSE}

# Define methods to add waypoints/smooth etc

library(tidyverse)
library(dyngen)
library(dynplot)
library(gtools)
library(dtw)

calculate_correct_pseudotime <- function(dataset){
  length_pieces <- dataset$milestone_network$length/sum(dataset$milestone_network$length)
  nr_piece <- 0
  start <- "sA"
  traj <- numeric()
  names_traj <- list()
  # TODO convert to for for each piece
  while(start %in% dataset$milestone_network[["from"]]){
    idx <- which(start == dataset$milestone_network[["from"]])[[1]]
    end <- dataset$milestone_network[["to"]][idx]
    to_add <- sum(length_pieces[0:nr_piece])

    perc <-(dataset[["progressions"]] %>% filter(from == start & to == end))$percentage * length_pieces[nr_piece + 1] + to_add
    nams <-(dataset[["progressions"]] %>% filter(from == start & to == end))$cell_id

    traj <- c(traj, perc)
    names_traj <- c(names_traj, nams)

    start <- end
    print(start)
    nr_piece <- nr_piece + 1

  }

  names(traj) <- names_traj
  traj <- traj[mixedorder(names(traj))]
  traj
}

get_geodesic_distances_from_progressions <- function(dataset, waypoint_progressions){
  waypoint_milestone_percentages <- dynwrap::convert_progressions_to_milestone_percentages(cell_id, dataset$milestone_ids, dataset$milestone_network, waypoint_progressions)
  waypoint_milestone_percentages <- dplyr::rename(waypoint_milestone_percentages, waypoint_id = cell_id)

  geodesic_distances <- dynwrap::calculate_geodesic_distances(dataset, waypoint_milestone_percentages = waypoint_milestone_percentages, directed=FALSE)
  geodesic_distances
}

get_waypoint_progression <- function(dataset, nr){
  cell_id <- character()
  froms <- character()
  tos <- character()
  percentage <- numeric()

  length_pieces <- dataset$milestone_network$length/sum(dataset$milestone_network$length)
  idx <- 1
  for(piece in length_pieces){
    amount <- round(nr * piece)
    wps <- modify(list(1:amount), function(x){x/amount})[[1]]
    ids <- modify(list(1:amount), function(x){sprintf("w%d_%03d", idx, x)})[[1]]
    from <- rep(dataset$milestone_network$from[idx], amount)
    to <- rep(dataset$milestone_network$to[idx], amount)

    cell_id <- c(cell_id, ids)
    froms <- c(froms, from)
    tos <- c(tos, to)
    percentage <- c(percentage, wps)

    idx <- idx + 1

  }
  waypoint_progressions <- tibble(cell_id=cell_id, from=froms, to=tos, percentage=percentage)
  waypoint_progressions
}

interpolate_expression <- function(dataset, wps, gds){
  weighted_expr <- do.call('cbind', lapply(seq_along(wps), function(idx){
    dist <- gds[idx,]
    weighted <- exp(-(dist^2)/(0.1)^2)
    weighted <- weighted/sum(weighted)
    weighted_exp <- t(as.matrix(dataset$counts)) %*% weighted
    weighted_exp
  }))
  weighted_expr
}

```

In order to smooth the gene expression, we construct a number of *psuedopoints* or *waypoints*. We place them equidistantly according to the pseudotime values. If a certain branch of the trajectory is twice as long as another branch, twice as many waypoints will be placed on that branch compared to the other branch.

We then determine the distance from these waypoints to all other cells present in the corresponding dataset. Taking these geodesic distances (distance along the trajectory graph) into account, we interpolate the gene expression values of these waypoints.

Finally, we calculate the dynamic time warping between the two datasets.

```{r, smooth_dtw, collapse=TRUE, fig.height=7, fig.width=7, fig.align = "center", results='hold', cache=TRUE, dependson=c("generatedata", "generate_repl_data")}
nr_wps <- nr_cells

wpp1 <- get_waypoint_progression(dataset_unperturbed, nr_wps)
wpp2 <- get_waypoint_progression(dataset_replicate, nr_wps)

gd1 <- get_geodesic_distances_from_progressions(dataset_unperturbed, wpp1)
gd2 <- get_geodesic_distances_from_progressions(dataset_replicate, wpp2)

wexpr1 <- interpolate_expression(dataset_unperturbed, wpp1$percentage, gd1)
wexpr2 <- interpolate_expression(dataset_replicate, wpp2$percentage, gd2)

# wexpr1 <- wexpr1[1:100,]
# wexpr2 <- wexpr2[1:100,]

d2 <- dist(as.matrix(t(wexpr1)), as.matrix(t(wexpr2)), dist.method="correlation")
am2 <- dtw(d2, step.pattern = symmetric2, keep.internals = TRUE, open.end = TRUE)
```

### Don't smooth gene expression

Here we simply use the count matrices of both datasets to calculate the distances between the two trajectories.

```{r, no_smooth_dtw, collapse=TRUE, fig.height=7, fig.width=7, fig.align = "center", results='hold', cache=TRUE, dependson=c("generatedata", "generate_repl_data")}
nor1 <- names(sort(calculate_correct_pseudotime(dataset_unperturbed)))
volg1 <- sapply(nor1, function(i) strtoi(substr(i, 5, 7)))
cnt1 <- as.matrix(dataset_unperturbed$counts)
cnts1 <- cnt1[volg1,]

nor2 <- names(sort(calculate_correct_pseudotime(dataset_replicate)))
volg2 <- sapply(nor2, function(i) strtoi(substr(i, 5, 7)))
cnt2 <- as.matrix(dataset_replicate$counts)
cnts2 <- cnt2[volg2,]

# cnts1 <- cnts1[,1:100]
# cnts2 <- cnts2[,1:100]

d3 <- dist(cnts1, cnts2, dist.method="correlation")
am3 <- dtw(d3, step.pattern = symmetric2, keep.internals = TRUE, open.end = TRUE)
```

### Compare both

We plot the smoothed and unsmoothed dtw graphs of these two datasets. We can see an obvious valley where the warping path lies in the case of the smoothed dtw. The graph with the unsmoothed dtw does not really show such a valley. Nonetheless, a diagonal warping path is found.

```{r, fig_comparison, collapse=TRUE, fig.height=7, fig.width=7, fig.align = "center", results='hold', cache=TRUE, dependson=c("no_smooth_dtw", "smooth_dtw")}

dtwPlotDensity(am2)
abline(coef= c(0,1))
dtwPlotDensity(am3)
abline(coef = c(0,1))
```

We can also compare the two dtw's quantitatively. The signal we want to capture is a warping path that lies on the diagonal. We can calculate either the area between the diagonal and the warping path or the sum of the absolute euclidean distance between the diagonal and the warping path, per x value.

```{r, compare_dtw, cache=TRUE, dependson=c("no_smooth_dtw", "smooth_dtw")}
library(MESS)

get_y_indices <- function(nr_points, index1, index2){
  xymap <- rep(list(), length = nr_points)
  for(i in 1:length(index1)){
    x_index = index1[[i]]
    y_index = index2[[i]]
    xymap[[x_index]] <- c(xymap[[x_index]], y_index)
  }

  sapply(sequence(nr_points), function(i) max(xymap[[i]]))
}

eucl_dist <- function(x1, x2, y1, y2){
  sqrt((x1 - x2)^2 + (y1 - y2)^2)
}

nr_x <- length(unique(am2$index1))

y_indices_smooth <- get_y_indices(length(unique(am2$index1)), am2$index1, am2$index2)
y_indices_no_smooth <- get_y_indices(length(unique(am2$index1)), am3$index1, am3$index2)

area_smooth <- MESS::auc(sequence(nr_x), y_indices_smooth - sequence(nr_x), absolutearea = TRUE)
dist_smooth <- sum(eucl_dist(sequence(nr_x), sequence(nr_x), y_indices_smooth, sequence(nr_x)))

area_no_smooth <- MESS::auc(sequence(nr_x), y_indices_no_smooth - sequence(nr_x), absolutearea = TRUE)
dist_no_smooth <- sum(eucl_dist(sequence(nr_x), sequence(nr_x), y_indices_no_smooth, sequence(nr_x)))

res <- data.frame("Smooth" = c("yes", "no"), "Area" = c(area_smooth, area_no_smooth), "Eucl" = c(dist_smooth, dist_no_smooth))

print(res)

```

## Generate and capture a topology change

Using dyngen, we can also simulate a topology change in the trajectory: a certain transcript factor that does not come to expression. This will lead to a shortened trajectory.

```{r, generatedata2, cache=TRUE}
backbone <- bblego(
  bblego_start("A", type = "simple", num_modules = 4),
  bblego_linear("A", "B", type = "simple", num_modules = 6),
  bblego_linear("B", "C", type = "simple", num_modules = 6),
  bblego_end("C", type = "simple", num_modules = 4)
)
model1 <-
  initialise_model(
    num_tfs = 20,
    num_targets = 200,
    num_hks = 200,
    num_cells = 1000,
    backbone = backbone,
    verbose = TRUE,
    download_cache_dir = "~/.cache/dyngen",
    num_cores = 8,
    distance_metric = "pearson",
    tf_network_params = tf_network_default(min_tfs_per_module = 1),
    simulation_params = simulation_default(num_simulations = 32, census_interval = .01),
  ) %>%
  generate_tf_network() %>%
  generate_feature_network() %>%
  generate_kinetics() %>%
  generate_gold_standard() %>%
  generate_cells()

model1_ <- generate_experiment(model1)
dataset1 <- wrap_dataset(model1_)

```

We can plot what this network looks like.

```{r, plot_model, cache=TRUE, dependson=c("generatedata2"), collapse=TRUE, fig.height=7, fig.width=7, fig.align = "center"}
plot_backbone_statenet(model1)
plot_backbone_modulenet(model1)
plot_feature_network(model1, show_targets = FALSE)
plot_feature_network(model1)
plot_feature_network(model1, show_hks = TRUE)
```

Next, we will reduce the transcription rate and the mRNA degradation rate of a certain transcription factor to 0. Here, we choose the transcription factor C1_TF1, because it is the fist transcription factor of C1. This makes it easy to later calculate with which percentage of the original trajectory this new trajectory will overlap.

```{r, generate_topology_change, cache=TRUE, dependson=c("generatedata2")}
change_speed <- function(model, target, rate) {
  param_id <- which(target == model$feature_info$feature_id)[[1]]

  model$feature_info$wpr[param_id] <- model$feature_info$wpr[param_id] * rate
  model$feature_info$xdr[param_id] <- model$feature_info$xdr[param_id] * rate

  param_wpr <- paste0("wpr_", target)
  param_xdr <- paste0("xdr_", target)

  model$simulation_system$parameters[param_wpr] <- model$simulation_system$parameters[param_wpr] * rate
  model$simulation_system$parameters[param_xdr] <- model$simulation_system$parameters[param_xdr] * rate

  model
}

model2 <- change_speed(model1, "C1_TF1", 0)
model2 <- generate_gold_standard(model2)
model2 <- generate_cells(model2)

model2_ <- generate_experiment(model2)
dataset2 <- wrap_dataset(model2_)
```

We can merge the models again and plot them in the same dimensionality reduction.

```{r, plot_topology_change, cache=TRUE, dependson=c("generatedata2", "generate_topology_change")}
model12 <- merge_models(model1, model1, model2)
model12 <- model12 %>%
  dyngen:::calculate_dimred(dimred_premrna = FALSE) %>%
  generate_experiment()

dataset <- wrap_dataset(model12)
plot_gold_simulations(model12)
plot_dimred(dataset)
plot_graph(dataset)
```

We can again calculate the dynamic time warping, both on smoothed and unsmoothed data.

```{r, dtw_topology_smoothed, cache=TRUE, dependson=c("generatedata2", "generate_topology_change")}
nr_wps <- 1000

wpp1 <- get_waypoint_progression(dataset1, nr_wps)
wpp2 <- get_waypoint_progression(dataset2, nr_wps)

gd1 <- get_geodesic_distances_from_progressions(dataset1, wpp1)
gd2 <- get_geodesic_distances_from_progressions(dataset2, wpp2)

wexpr1 <- interpolate_expression(dataset1, wpp1$percentage, gd1)
wexpr2 <- interpolate_expression(dataset2, wpp2$percentage, gd2)

d2 <- dist(as.matrix(t(wexpr2)), as.matrix(t(wexpr1)), dist.method="pearson")
am2 <- dtw(as.matrix(t(wexpr2)), as.matrix(t(wexpr1)), step.pattern = symmetric2, keep.internals = TRUE, open.end = TRUE)

```

```{r, dtw_topology_unsmoothed, cache=TRUE, dependson=c("generatedata2", "generate_topology_change")}
nor1 <- names(sort(calculate_correct_pseudotime(dataset1)))
volg1 <- sapply(nor1, function(i) strtoi(substr(i, 5, 7)))
cnt1 <- as.matrix(dataset1$counts)
cnts1 <- cnt1[volg1,]

nor2 <- names(sort(calculate_correct_pseudotime(dataset2)))
volg2 <- sapply(nor2, function(i) strtoi(substr(i, 5, 7)))
cnt2 <- as.matrix(dataset2$counts)
cnts2 <- cnt2[volg2,]

d3 <- dist(cnts2, cnts1, dist.method="pearson")
am3 <- dtw(d3, step.pattern = symmetric2, keep.internals = TRUE, open.end = TRUE)
```

We can plot both dimensionality reductions. TODO: why slope low

```{r, plot_topology_dtw, cache=TRUE, dependson=c("dtw_topology_smoothed", "dtw_topology_unsmoothed"), collapse=TRUE, fig.height=7, fig.width=7, fig.align = "center"}

dtwPlotDensity(am2)
abline(coef= c(0, sum(dataset1$milestone_network$length[1:2]) / sum(dataset1$milestone_network$length)))
dtwPlotDensity(am3)
abline(coef = c(0, sum(dataset1$milestone_network$length[1:2]) / sum(dataset1$milestone_network$length)))
```

<!-- library(MESS) -->

<!-- get_y_indices <- function(nr_points, index1, index2){ -->
<!--   xymap <- rep(list(), length = nr_points) -->
<!--   for(i in 1:length(index1)){ -->
<!--     x_index = index1[[i]] -->
<!--     y_index = index2[[i]] -->
<!--     xymap[[x_index]] <- c(xymap[[x_index]], y_index) -->
<!--   } -->

<!--   sapply(sequence(nr_points), function(i) max(xymap[[i]])) -->
<!-- } -->

<!-- eucl_dist <- function(x1, x2, y1, y2){ -->
<!--   sqrt((x1 - x2)^2 + (y1 - y2)^2) -->
<!-- } -->

<!-- nr_x <- length(unique(am2$index1)) -->

<!-- y_indices_smooth <- get_y_indices(length(unique(am2$index1)), am2$index1, am2$index2) -->
<!-- y_indices_no_smooth <- get_y_indices(length(unique(am2$index1)), am3$index1, am3$index2) -->

<!-- area_smooth <- MESS::auc(sequence(nr_x), y_indices_smooth - sequence(nr_x), absolutearea = TRUE) -->
<!-- dist_smooth <- sum(eucl_dist(sequence(nr_x), sequence(nr_x), y_indices_smooth, sequence(nr_x))) -->

<!-- area_no_smooth <- MESS::auc(sequence(nr_x), y_indices_no_smooth - sequence(nr_x), absolutearea = TRUE) -->
<!-- dist_no_smooth <- sum(eucl_dist(sequence(nr_x), sequence(nr_x), y_indices_no_smooth, sequence(nr_x))) -->

<!-- res <- data.frame("Smooth" = c("yes", "no"), "Area" = c(area_smooth, area_no_smooth), "Eucl" = c(dist_smooth, dist_no_smooth)) -->

<!-- print(res) -->

### 2019-11-25 analysis/dtw_functions.R

@@ -35,7 +35,7 @@ calculate_correct_pseudotime <- function(dataset){
 
 get_geodesic_distances_from_progressions <- function(dataset, waypoint_progressions){
   waypoint_milestone_percentages <- dynwrap::convert_progressions_to_milestone_percentages(cell_id, dataset$milestone_ids, dataset$milestone_network, waypoint_progressions)
-  waypoint_milestone_percentages <- rename(waypoint_milestone_percentages, waypoint_id = cell_id)
+  waypoint_milestone_percentages <- dplyr::rename(waypoint_milestone_percentages, waypoint_id = cell_id)
 
   geodesic_distances <- dynwrap::calculate_geodesic_distances(dataset, waypoint_milestone_percentages = waypoint_milestone_percentages, directed=FALSE)
   geodesic_distances

### 2019-11-25 analysis/usecase_trajectory_alignment.R

@@ -1,10 +1,65 @@
 library(tidyverse)
 library(dyngen)
+library(dynplot)
 
 set.seed(1)
 
 out_dir <- "fig/usecase/trajectory_alignment/"
 dir.create(out_dir, showWarnings = FALSE, recursive = TRUE)
+
+# Merge two models, useful if we want to plot the models together
+merge_models <- function(model, model1, model2){
+  model12 <- model
+  model12$feature_info <- model$feature_info %>%
+    mutate(
+      w = paste0("w_", feature_id),
+      x = paste0("x_", feature_id),
+      y = paste0("y_", feature_id)
+    )
+  m1gs <- model1$gold_standard
+  m2gs <- model2$gold_standard
+  model12$gold_standard <- list(
+    meta = bind_rows(
+      m1gs$meta %>% mutate_at(c("from", "to", "from_", "to_"), ~paste0("left_", .)),
+      m2gs$meta %>% mutate_at(c("from", "to", "from_", "to_"), ~paste0("right_", .))
+    ),
+    counts = rbind(
+      m1gs$counts,
+      m2gs$counts
+    ),
+    network = bind_rows(
+      m1gs$network %>% mutate_at(c("from", "to"), ~paste0("left_", .)),
+      m2gs$network %>% mutate_at(c("from", "to"), ~paste0("right_", .))
+    )
+  )
+  m1sim <- model1$simulations
+  m2sim <- model2$simulations
+  num_m1sim <- max(m1sim$meta$simulation_i)
+  model12$simulations <- list(
+    meta = bind_rows(
+      m1sim$meta %>% mutate_at(c("from", "to"), ~paste0("left_", .)),
+      m2sim$meta %>% mutate_at(c("from", "to"), ~paste0("right_", .)) %>% mutate(simulation_i = simulation_i + num_m1sim)
+    ),
+    counts = rbind(
+      m1sim$counts,
+      m2sim$counts
+    ),
+    regulation = rbind(
+      m1sim$regulation,
+      m2sim$regulation
+    ),
+    reaction_firings = rbind(
+      m1sim$reaction_firings,
+      m2sim$reaction_firings
+    ),
+    reaction_propensities = rbind(
+      m1sim$reaction_propensities,
+      m2sim$reaction_propensities
+    )
+  )
+  model12
+}
+
 
 # Generate base model -----------------------------------------------------
 backbone <- bblego(
[…]

### 2019-11-25 analysis/usecase_trajectory_retardation.R

@@ -1,280 +1,0 @@
-library(tidyverse)
-library(dyngen)
-library(dynplot)
-library(gtools)
-library(dtw)
-
-setwd("/home/louisedc/Work/dyngen_analysis/analysis")
-source("dtw_functions.R")
-
-# # Function that will change the speed of the targeted gene.
-# # TODO: what is wpr & xdr again?
-# change_speed <- function(model, target, rate) {
-#   param_id <- which(target == model$feature_info$feature_id)[[1]]
-#
-#   model$feature_info$wpr[param_id] <- model$feature_info$wpr[param_id] * rate
-#   model$feature_info$xdr[param_id] <- model$feature_info$xdr[param_id] * rate
-#
-#   param_wpr <- paste0("wpr_", target)
-#   param_xdr <- paste0("xdr_", target)
-#
-#   model$simulation_system$parameters[param_wpr] <- model$simulation_system$parameters[param_wpr] * rate
-#   model$simulation_system$parameters[param_xdr] <- model$simulation_system$parameters[param_xdr] * rate
-#
-#   model
-# }
-#
-# generate_retardation <- function(model, id){
-#   model1 <- model
-#   model1 <- change_speed(model1, "A4_TF1", 0.5)
-#   model1 <- generate_gold_standard(model1)
-#   model1 <- generate_cells(model1)
-#   model1 <- generate_experiment(model1)
-#   dataset2 <- wrap_dataset(model1)
-#
-#   # save model & dataset
-#   # saveRDS(dataset2, paste0("data/retardation_ds_", id))
-#   # saveRDS(model1, paste0("analysis/data/retardation_model_", id))
-#
-#   # for(i in 1:9){
-#   #   generate_replication(model1, i)
-#   # }
-#
-#   dataset2
-# }
-#
-# generate_replication <- function(model, id){
-#   model <- generate_cells(model)
-#   model <- generate_experiment(model)
-#   saveRDS(model, paste0("analysis/data/retardation_replication_", id))
-# }
-#
-# # Construct a synthetic model
-# # A very simple linear model
-# backbone <- bblego(
-#   bblego_start("A", type = "simple", num_modules = 4),
[…]

### 2019-12-01 analysis/network_inference/1_generate_datasets.R
update NI scripts

@@ -8,6 +8,9 @@ if (!dir.exists(derived_folder)) dir.create(derived_folder, recursive = TRUE)
 
 backbones <- list_backbones()
 
+backbones$binary_tree <- NULL
+backbones$branching <- NULL
+
 walk(names(backbones), function(backbone_name) {
 
   folder <- paste0(derived_folder, "datasets/", backbone_name, "/")

### 2019-12-01 analysis/network_inference/2_run_pyscenic.R

@@ -203,11 +203,26 @@ outs <- map(seq_len(nrow(cross)), function(i) {
 
 
 comb <- bind_rows(
-  map_df(outs, "eval_pyscenic_static") %>% mutate(type = "one_static"),
-  map_df(outs, "eval_pyscenic_sample")
-)
+  map_df(outs, "eval_pyscenic_static") %>% mutate(type = "both_static", type_nice = "pySCENIC evaluated\nagainst static GRN"),
+  map_df(outs, "eval_pyscenic_sample") %>% arrange(desc(type)) %>% mutate(type_nice = c(static = "pySCENIC evaluated\nagainst casewise GRN", casewise = "pySCENIC+AUCell evaluated\nagainst casewise GRN")[type])
+) %>%
+  mutate(type_nice = forcats::fct_inorder(type_nice))
 
-ggplot(comb) + geom_point(aes(auroc, aupr, colour = method)) + facet_grid(dataset ~ type) + theme_bw()
+g <-
+  ggplot(comb %>% sample_n(n())) +
+  geom_point(aes(auroc, aupr, colour = method, size = type)) +
+  facet_grid(dataset ~ type_nice) +
+  theme_bw() +
+  scale_colour_brewer(palette = "Set3") +
+  scale_size_manual(values = c("both_static" = 2, static = .75, casewise = .75))
+ggsave("fig/network_inference/evaluation.pdf", g, width = 8, height = 6)
 
+g <- ggplot(comb %>% sample_n(n())) +
+  geom_point(aes(auroc, aupr, colour = type, size = type)) +
+  facet_wrap(dataset ~ method, ncol = 4, scales = "free") +
+  theme_bw() +
+  scale_colour_brewer(palette = "Set1") +
+  scale_size_manual(values = c("both_static" = 2, static = .75, casewise = .75))
+ggsave("fig/network_inference/evaluation_2.pdf", g, width = 10, height = 10)
 
 

### 2019-12-01 render_manuscript.R
fix render manuscript script

@@ -1,16 +1,19 @@
 library(tidyverse)
 library(googledrive)
 library(rmarkdown)
+library(textreadr)
 
 # copy zotero library if on rcannood's computer
 if (Sys.info()[["user"]] == "rcannood") {
   file.copy("~/Workspace/library.bib", "library.bib", overwrite = TRUE)
 }
 
-# download Rmd from google drive
+# download as docx instead of txt, because otherwise comments get pushed into the document
 httr::set_config(httr::config(http_version = 0)) # avoid http2 framing layer bug
-drive <- drive_download(as_id("1BOltsUJBQ8NhF5s9kU_Q470U0U00QblbG8HBgQwEzRE"), type="text/plain", overwrite=TRUE, path = tempfile())
-read_lines(drive$local_path) %>%
+drive <- drive_download(as_id("1BOltsUJBQ8NhF5s9kU_Q470U0U00QblbG8HBgQwEzRE"), type="docx", overwrite=TRUE, path = tempfile())
+
+# read docx
+textreadr::read_docx(drive$local_path, remove.empty = FALSE, trim = FALSE) %>%
   str_replace_all(" *(\\[@[^\\]]*\\])", "\\1") %>% # remove spaces before citations
   write_lines("manuscript.Rmd")
 

### 2019-12-01 analysis/simplecyclic.R
update simple cyclic

@@ -5,20 +5,20 @@ set.seed(1)
 
 backbone <- backbone(
   module_info = tribble(
-    ~module_id, ~a0, ~burn,
-    "A", 1, TRUE,
-    "B", 1, TRUE,
-    "C", 1, TRUE,
-    "D", 0, TRUE,
-    "E", 0, TRUE
+    ~module_id, ~basal, ~burn, ~independence,
+    "A", 1, TRUE, 1,
+    "B", 1, TRUE, 1,
+    "C", 1, TRUE, 1,
+    "D", 0, TRUE, 1,
+    "E", 0, TRUE, 1
   ),
   module_network = tribble(
     ~from, ~to, ~effect, ~strength, ~cooperativity,
-    "A", "B", -1, 4, 2,
-    "B", "C", -1, 4, 2,
-    "C", "D", 1, 1, 2,
-    "D", "E", 1, 1, 2,
-    "E", "A", -1, 4, 2
+    "A", "B", -1L, 4, 2,
+    "B", "C", -1L, 4, 2,
+    "C", "D", 1L, 1, 2,
+    "D", "E", 1L, 1, 2,
+    "E", "A", -1L, 4, 2
   ),
   expression_patterns = tribble(
     ~from, ~to, ~module_progression, ~start, ~burn, ~time,
@@ -28,8 +28,9 @@ backbone <- backbone(
     "S3", "S1", "+A,+B,-D,-E", FALSE, FALSE, 4
   )
 )
-census_interval <- .1
-total_time <- 10
+census_interval <- .2
+total_time <- 20
+time_breaks = c(0, 10, 20)
 model <-
   initialise_model(
     num_tfs = nrow(backbone$module_info) * 1,
@@ -41,20 +42,22 @@ model <-
     download_cache_dir = "~/.cache/dyngen",
     num_cores = 8,
     kinetics_params = kinetics_default(
-      sample_wpr = function(n) 100,
-      sample_spl = function(n) 10,
-      sample_xdr = function(n) 5,
-      sample_ypr = function(n) 5,
-      sample_ydr = function(n) 5
+      sample_wpr = function(n) 50,
+      sample_wsr = function(n) 5,
+      sample_ypr = function(n) 2.5,
+      sample_whl = function(n) .25,
+      sample_xhl = function(n) .25,
+      sample_yhl = function(n) .25
     ),
     simulation_params = simulation_default(
-      burn_time = 2,
+      burn_time = 8,
[…]

### 2019-12-01 analysis/script.R
update small bifurc

@@ -1,133 +1,0 @@
-library(tidyverse)
-library(dyngen)
-
-set.seed(10)
-
-backbone <- bblego(
-  bblego_start("A", type = "simple", num_modules = 1),
-  bblego_linear("A", "B", type = "simple", num_modules = 3),
-  bblego_branching("B", c("C", "D")),
-  bblego_end("C", type = "simple", num_modules = 2),
-  bblego_end("D", type = "simple", num_modules = 2)
-)
-model <-
-  initialise_model(
-    num_tfs = 12,
-    num_targets = 30,
-    num_hks = 15,
-    num_cells = 150,
-    backbone = backbone,
-    verbose = TRUE,
-    download_cache_dir = "~/.cache/dyngen",
-    num_cores = 8,
-    distance_metric = "euclidean",
-    simulation_params = simulation_default(ssa_algorithm = GillespieSSA2::ssa_exact(), num_simulations = 8, census_interval = .01)
-  ) %>%
-  generate_tf_network() %>%
-  generate_feature_network() %>%
-  generate_kinetics()
-
-## grn
-plot_backbone_statenet(model) %>% ggsave(filename = "dyngen/statenet.pdf", width = 6, height = 6)
-plot_backbone_modulenet(model) %>% ggsave(filename = "dyngen/modulenet.pdf", width = 8, height = 6)
-
-plot_feature_network(model, show_targets = FALSE)
-
-plot_feature_network(model)
-plot_feature_network(model, show_hks = TRUE)
-plot_feature_network(model)
-plot_feature_network(model, show_hks = TRUE) %>% ggsave(filename = "dyngen/featnet.pdf", width = 8, height = 6)
-plot_feature_network(model, show_targets = FALSE) %>% ggsave(filename = "dyngen/featnet_onlytfs.pdf", width = 8, height = 6)
-
-## simulate
-model2 <- model %>%
-  generate_gold_standard() %>%
-  generate_cells() %>%
-  generate_experiment()
-
-# recalculate dimred
-space <- prcomp(model2$simulations$counts %>% as.matrix, rank. = 2)$x
-colnames(space) <- c("comp_1", "comp_2")
-model2$simulations$dimred <- space
-
-write_rds(model2, "dyngen/model.rds", compress = "gzip")
-
[…]

### 2019-12-01 analysis/small_bifurc.R

@@ -1,0 +1,197 @@
+library(tidyverse)
+library(dyngen)
+
+set.seed(10)
+
+backbone <- bblego(
+  bblego_start("A", type = "simple", num_modules = 1),
+  bblego_linear("A", "B", type = "simple", num_modules = 3),
+  bblego_branching("B", c("C", "D")),
+  bblego_end("C", type = "simple", num_modules = 2),
+  bblego_end("D", type = "simple", num_modules = 2)
+)
+model <-
+  initialise_model(
+    num_tfs = 12,
+    num_targets = 30,
+    num_hks = 15,
+    num_cells = 150,
+    backbone = backbone,
+    verbose = TRUE,
+    download_cache_dir = "~/.cache/dyngen",
+    num_cores = 8,
+    distance_metric = "euclidean",
+    simulation_params = simulation_default(
+      kinetics_noise_function = kinetics_noise_none(),
+      ssa_algorithm = GillespieSSA2::ssa_exact(),
+      experiment_params = simulation_type_wild_type(num_simulations = 8),
+      census_interval = .01,
+      store_grn = TRUE
+    )
+  ) %>%
+  generate_tf_network() %>%
+  generate_feature_network() %>%
+  generate_kinetics()
+
+## grn
+plot_backbone_statenet(model) %>% ggsave(filename = "fig/small_bifurc/statenet.pdf", width = 6, height = 6)
+plot_backbone_modulenet(model) %>% ggsave(filename = "fig/small_bifurc/modulenet.pdf", width = 8, height = 6)
+
+plot_feature_network(model, show_targets = FALSE)
+
+plot_feature_network(model)
+plot_feature_network(model, show_hks = TRUE)
+plot_feature_network(model)
+plot_feature_network(model, show_hks = TRUE) %>% ggsave(filename = "fig/small_bifurc/featnet.pdf", width = 8, height = 6)
+plot_feature_network(model, show_targets = FALSE) %>% ggsave(filename = "fig/small_bifurc/featnet_onlytfs.pdf", width = 8, height = 6)
+
+## simulate
+model2 <- model %>%
+  generate_gold_standard()
+
+g <- plot_gold_expression(model2, what = "x") # mrna
+ggsave("fig/small_bifurc/gold_mrna.pdf", g, width = 8, height = 6)
[…]

### 2019-12-01 manuscript.Rmd
update manuscript

@@
 **Purpose:** A common problem of pioneering computational tools is that during their development, there are rarely sufficient datasets available for adequately quantitatively assessing its performance.
-**Results:** We developed `dyngen`, a multi-modality simulator of single cells.
-In `dyngen`, the biomolecular state of an _in silico_ cell changes over time according to a set of reactions defined by the cell's gene regulatory network.
+**Results:** We developed dyngen, a multi-modality simulator of single cells.
+In dyngen, the biomolecular state of an _in silico_ cell changes over time according to a set of reactions defined by the cell's gene regulatory network.
 By simulating single cells in terms of its biomolecular state and reactions, the simulator is easily extendible for adding new modalities or experimental procedures.
-We demonstrate `dyngen`'s flexibility by simulating snapshot, time-series and perturbation experiments.
-**Conclusion:** `dyngen` lays the foundations for benchmarking a wide variety of computational single-cell tools and can be used to help kick-start the development of future types of analyses.
+We demonstrate dyngen's flexibility by simulating snapshot, time-series and perturbation experiments.
+**Conclusion:** dyngen lays the foundations for benchmarking a wide variety of computational single-cell tools and can be used to help kick-start the development of future types of analyses.
+
+**Author contributions:**
+
+* R.C. and W.S. designed the study.
+* R.C., W.S., and L.D. performed the experiments and analysed the data.
+* R.C. and W.S. implemented the dyngen software package.
+* R.C. wrote the original manuscript.
+* R.C., W.S., L.D., and Y.S. reviewed and edited the manuscript.
+* Y.S. supervised the project.
 
@@
 
-We introduce `dyngen`, a multi-modality simulator of single cells and their dynamics (Figure\ \ref{fig:showcase}A).
+We introduce dyngen, a multi-modality simulator of single cells and their dynamics (Figure\ \ref{fig:showcase}A).
 A cell is simulated using Gillespie's Stochastic Simulation Algorithm (SSA)[@gillespie_exactstochasticsimulation_1977] where a cell consists of a set of molecules.
@@
 By simulating a cell over time in terms of its molecular state and the reactions that are allowed to occur, the simulator is more extendible to new modalities or experimental procedures (Figure\ \ref{fig:showcase}B).
-We demonstrate `dyngen`'s flexibility by simulating snapshot, time-series and perturbation experiments, allowing to benchmark a wide variety of computational tools such as trajectory alignment and differential network inference methods.
+We demonstrate dyngen's flexibility by simulating snapshot, time-series and perturbation experiments, allowing to benchmark a wide variety of computational tools such as trajectory alignment and differential network inference methods.
 
@@
     \caption{
-            \textbf{Showcase of \texttt{dyngen} functionality.}
+        \textbf{Showcase of dyngen functionality.}
         \textbf{A:}
@@
     \caption{
-            \textbf{Key features of \texttt{dyngen} are illustrated using a cyclic toy example.}
-            While this example comprises only of a single cell and 5 genes, \texttt{dyngen} is able to simulate thousands of cells with GRNs containing thousands of genes.
+        \textbf{Key features of dyngen are illustrated using a cyclic toy example.}
+        While this example comprises only of a single cell and 5 genes, dyngen is able to simulate thousands of cells with GRNs containing thousands of genes.
     }
@@
 
-`dyngen` returns many modalities throughout the whole simulation: molecular abundance, cellular state, number of reaction firings, reaction likelihoods, and regulation activations (Figure\ \ref{fig:simplecyclic}C--F).
+dyngen returns many modalities throughout the whole simulation: molecular abundance, cellular state, number of reaction firings, reaction likelihoods, and regulation activations (Figure\ \ref{fig:simplecyclic}C--F).
 These modalities can serve both as input data and ground truth for benchmarking many types of computational approaches.
@@
 Depending on how the GRN is designed, different cellular developmental processes can be simulated.
-`dyngen` includes generators of GRNs which result in many different developmental topologies (Figure\ \ref{fig:example_runs}), including branching, converging, cyclic and even disconnected.
+dyngen includes generators of GRNs which result in many different developmental topologies (Figure\ \ref{fig:example_runs}), including branching, converging, cyclic and even disconnected.
 
@@
     \caption{
-            \textbf{Multiple executions of \texttt{dyngen} with different predefined backbones.} From each simulation of about 200 genes, 1000 cells were sampled.
+        \textbf{Multiple executions of dyngen with different predefined backbones.} From each simulation of about 200 genes, 1000 cells were sampled.
         \textbf{A:}~Linear. \textbf{B:}~Bifurcating. \textbf{C:}~Converging.
@@
 Custom-defined GRNs offer more fine-grained control over the simulation.
-Aside from simulating topologies currently not supported by `dyngen`, this has several other important use-cases.
+Aside from simulating topologies currently not supported by dyngen, this has several other important use-cases.
 Simulations of the same GRN with small perturbations allow to emulate batch effects or perturbation experiments (Figure\ \ref{fig:usecases}).
@@
 
-`dyngen` can be used to simulate different experimental conditions.
-By default, `dyngen` supports snapshot experiments (uniformly sampling from an asynchronous dynamic process) and time-series experiments (sampling cells from different intervals in the simulation).
+dyngen can be used to simulate different experimental conditions.
+By default, dyngen supports snapshot experiments (uniformly sampling from an asynchronous dynamic process) and time-series experiments (sampling cells from different intervals in the simulation).
 However, it is possible to imagine and implement other sampling strategies, such as sampling a cell at a certain time point and once more at a later time point.
 This would allow evaluating the performance of RNA velocity approaches -- which predict the future state of a cell by looking at differences in pre-mRNA and mRNA abundance levels.
-
-<!-- YSA: RNA velocity: Onduidelijk hier, mensen kennen dit niet. -->
 
@@
 # Discussion {#sec:dyngen-discussion}
-As is, `dyngen`'s single cell simulations can be used to evaluate common single-cell omics computational methods such as clustering, batch correction, trajectory inference and network inference.
+As is, dyngen's single cell simulations can be used to evaluate common single-cell omics computational methods such as clustering, batch correction, trajectory inference and network inference.
 However, the combined effect of these advantages results in a framework that is flexible enough to adapt to a broad range of applications.
 This may include methods that integrate clustering, network inference and trajectory inference.
-In this respect, `dyngen` may promote the development of new tools in the single-cell field similarly as other simulators have done in the past[@schaffter_genenetweaversilicobenchmark_2011; @ewing_combiningtumorgenome_2015].
+In this respect, dyngen may promote the development of new tools in the single-cell field similarly as other simulators have done in the past[@schaffter_genenetweaversilicobenchmark_2011; @ewing_combiningtumorgenome_2015].
 
-`dyngen` ultimately allows anticipating technological developments in single-cell multi-omics.
+dyngen ultimately allows anticipating technological developments in single-cell multi-omics.
 In this way, it is possible to design and evaluate the performance and robustness of new types of computational analyses before experimental data becomes available.
@@
 
-Currently, `dyngen` focuses on simulating cells as standalone entities that are well mixed.
+Currently, dyngen focuses on simulating cells as standalone entities that are well mixed.
 Splitting up the simulation space into separate subvolumes could pave the way to better study key cellular processes such as cell division, intercellular communication and migration[@smith_spatialstochasticintracellular_2019].
@@
     \includegraphics[width=\textwidth]{fig/explain_methods}
-    \caption{\textbf{The workflow of \texttt{dyngen} is comprised of six main steps.} 
+    \caption{\textbf{The workflow of dyngen is comprised of six main steps.} 
     \textbf{A:} The user needs to specify the desired module network or use a predefined module network. 
@@
 Each time step is mapped to a point in the backbone.
-    \textbf{F:} Multiple cells are sampled from each simulation. Molecules are sampled from each cell.
+    \textbf{F:} The molecule levels of multiple simulations are shown over time (left).
+From each simulation, multiple cells are sampled (from left to middle).
+Technical noise from profiling is simulated by sampling molecules from the set of molecules inside each cell (from middle to right).
     }
@@
 
-In `dyngen`, the user defines the behaviour of the simulation by defining how sets of genes, called modules, are regulating each other.
-A module may have basal expression, which means that pre-mRNA of the genes in this module will be transcribed without the presence of transcription factor molecules.
-A module marked as "active during the burn phase" means that this module will be allowed to generate expression of its genes during an initial warm-up phase.
-At the end of the `dyngen` process, cells will not be sampled from the burn phase simulations.
+To start a dyngen simulation, the user needs to define a module network and a backbone.
+The module network defines how sets of co-regulated genes, called modules, regulate each other.
+The module network is what mainly determines which dynamic processes occur within the simulated cells.
+The backbone is a separate set of simulations in which the ground-truth topology of the dynamic processes are defined, as it is difficult to determine the topology of the dynamic processes from the module network itself.
+
+A module network consists of modules connected together by regulatory interactions. 
+A module may have basal expression, which means genes in this module will be transcribed without the presence of transcription factor molecules.
+A module marked as "active during the burn phase" means that this module will be allowed to generate expression of its genes during an initial warm-up phase (See section \ref{sec:dyngen-simcell}).
+At the end of the dyngen process, cells will not be sampled from the burn phase simulations.
 Interactions between modules have a strength (which is a positive integer) and an effect (+1 for upregulating, -1 for downregulating).
@@
 In addition, the fate decisions made in this example of a bifurcation is reversible, meaning cells can be reprogrammed to go down a different differentiation path.
-If this effect is undesirable, more safeguards need to be put in place to prevent reprogramming from occurring.
+If this effect is undesirable, more safeguards need to be put in place to prevent reprogramming from occurring (Section \ref{sec:dyngen-bbl}).
 
@@
 
-### Backbone lego
+### Backbone lego {#sec:dyngen-bbl}
 The backbone can make use of one or more "backbone lego" (BBL) pieces (Figure\ \ref{fig:backbone_lego}).
@@
 Parameters allow determining the number of modules involved in the process and the number of outputs.
-Multiple BBLs can be chained together in order to intuitively create milestone networks and corresponding state networks (Figure\ \ref{fig:backbone_lego}B).
+Multiple BBLs can be chained together in order to intuitively create module networks and corresponding state networks (Figure\ \ref{fig:backbone_lego}B).
 Note that not all dynamic processes can be represented by a combination of BBLs, but they can serve as common building blocks to aid the construction of the backbone.
@@
 
-<!-- paper on transcriptional bursting -->
-
-<!-- doi: 10.1039/C7MB00154A -->
-
-<!-- interesting to try to replicate in dyngen! :) -->
-
 \newcommand{\w}[1]{\text{w}_{#1}}
@@
 \newcommand{\wpr}[1]{\text{wpr}_{#1}}
+\newcommand{\whl}[1]{\text{whl}_{#1}}
+\newcommand{\wdr}[1]{\text{wdr}_{#1}}
 \newcommand{\wsr}[1]{\text{wsr}_{#1}}
+\newcommand{\xhl}[1]{\text{xhl}_{#1}}
 \newcommand{\xdr}[1]{\text{xdr}_{#1}}
 \newcommand{\ypr}[1]{\text{ypr}_{#1}}
+\newcommand{\yhl}[1]{\text{yhl}_{#1}}
 \newcommand{\ydr}[1]{\text{ydr}_{#1}}
@@
 \newcommand{\hill}[1]{\text{hill}_{#1}}
-\newcommand{\coop}[1]{\text{coop}_{#1}}
+\newcommand{\coop}[1]{\text{co}_{#1}}
 \newcommand{\dis}[1]{\text{dis}_{#1}}
@@
 Each reaction consists of its propensity -- a formula to calculate the probability of the reaction occurring during an infinitesimal time interval -- and the effect -- how it will affect the current state if triggered.
-The effects and propensity functions of these reactions are defined in Table\ \ref{tab:reaction_def}.
 
-<!-- TODO: add step-by-step  derivation of transcription formula -->
+The effects of each reaction mimic the respective biological processes (Table\ \ref{tab:reaction_def}, middle).
+Transcription of gene $G$ results in the creation of a single pre-mRNA molecule $\w G$.
+Splicing turns one pre-mRNA $\w G$ into a mature mRNA $\w G$.
+Translation uses a mature mRNA $\x G$ to produce a protein $\y G$.
+Pre-mRNA, mRNA and protein degradation results in the removal of a $\w G$, $\x G$, and $\y G$ molecule, respectively.
 
-<!-- YSA: Graag wat meer uitleg in de tekst over al deze formules, definitie van notaties en zo. -->
+The propensity of all reactions except transcription are straight-forward (Table\ \ref{tab:reaction_def}, right).
+The degradation and splicing propensity of pre-mRNAs of a gene $G$ are linear functions of the abundance $\w G$, namely $\wdr G \times \w G$ and $\wsr G \times \w G$ respectively.
+Similarly, the degradation and translation of mature mRNAs are linear functions of $\x G$, namely $\xdr G \times \x G$ and $\ypr G \times \x G$.
+Degradation of proteins is also a linear function of $\y G$, namely $\ydr G \times \y G$.
+Constant parameters $\wdr G$, $\wsr G$, $\xdr G$, $\ypr G$, and $\ydr G$ are from respective normal distributions (Table\ \ref{tab:reaction_params}).
+These formulae are based on those of GNW[@schaffter_genenetweaversilicobenchmark_2011], except that the production and decay of molecules have been separated into two separate formulae.
 
@@
     \caption{
-      \textbf{Default parameters defined for the calculation of reaction propensity functions.}
+      \textbf{Reactions affecting the abundance levels of pre-mRNA $\w G$, mature mRNA $\x G$ and proteins $\y G$ of gene $G$.} Define the set of regulators of $G$ as $\rs{G}$, the set of upregulating regulators of $G$ as $\rp G$, and the set of downregulating regulators of $G$ as $\rn G$.
+Parameters used in the propensity formulae are defined in Table \ref{tab:reaction_params}.
     } \label{tab:reaction_def}
     \centering
-    \begin{tabular}{|lrl|}
+    \begin{tabular}{|lcc|}
         \hline
-            Parameter & Symbol & Definition \\ \hline \hline
-            Transcription rate & $\wpr{G}$ & $\in N(100, 20),\ \geq 10$ \\
-            Splicing rate & $\wsr G$ & $\in N(10, 2),\ \geq 2$ \\
-            mRNA degradation rate & $\xdr{G}$ & $\in N(5, 1),\ \geq 2$ \\
-            Translation rate & $\ypr{G}$ & $\in N(5, 1),\ \geq 2$ \\
-            Protein degradation rate & $\ydr G$ & $\in N(3, 0.5),\ \geq 1$ \\
-            Interaction strength & $\str{G,H}$ & $\in 10^{U(0, 2)}$ * \\
-            Hill coefficient & $\hill{G,H}$ & $\in U(0.5, 2)$ * \\
-        Cooperativity factor & $\coop G$ & $\in [0, 1]$ * \\ \hline\hline
-            Dissociation constant & $\dis H$ & $= 0.5 \times \frac{\wpr H \times \ypr H}{\xdr H \times \ydr H}$ \\
-            Regulation activity & $\buf{G,H}$ & $= \left(\str{G,H} \times \frac{\y H}{\dis H}\right) ^ {\hill{G,H}}$ \\
-            Basal expression & $\ba G$ & $= \begin{cases} 1 & \mbox{if } \rp{G} = \emptyset \\ 0.0001 & \mbox{if } \rn{G} = \emptyset \mbox{ and } \rp{G} \neq \emptyset \\ 0.5 & \mbox{otherwise} \end{cases}$ * \\ \hline
-            \multicolumn{3}{l}{*: unless $G$ is a TF, then the value is determined by the backbone.}
+        Reaction & Effect & Propensity \\ \hline \hline
+        Transcription & $\emptyset \rightarrow \w G$ & $\wpr G \times \frac{\ba G - \coop{G}^{|\rp{G}|} + \prod\limits_{H \in \rp{G}}(\coop G + \buf{G,H})}{\prod\limits_{H \in \rs{G}}(1 + \buf{G,H})}$ \\
+        Pre-mRNA degradation & $\w G \rightarrow \emptyset$ & $\wdr G \times \w G$ \\
+        Splicing & $\w G \rightarrow \x G$ & $\wsr G \times \w G$ \\
+        Mature mRNA degradation & $\x G \rightarrow \emptyset$ & $\xdr G \times \x G$ \\
+        Translation & $\x G \rightarrow \x G + \y G$ & $\ypr G \times \x G$ \\
+        Protein degradation & $\y G \rightarrow \emptyset$ & $\ydr G \times \y G$ \\ \hline
     \end{tabular}
 \end{table}
-
-<!-- 
-% regG,H represents a 'probability' that H is bound to G (though does not yet scale between [0,1]).
-y_H / maxy_H is more likely to be a probability that H is bound to G, though this does not take into account that H can be bound to other genes as well.
-% in any case, if it is, then it upregulates G by strG,H. It is assumed binding events
-% are independent and result in multiplicative effect.
-
-% according to https://en.wikipedia.org/wiki/Hill_equation_(biochemistry)#Proportion_of_ligand-bound_receptors:
-% P_H = (y_H ^ N) / (degradation_H / production_H + y_H ^ N)
-% ...
-% ...
-% might need to go back to things like:
-% https://kar.kent.ac.uk/24077/1/myzabetpaper.pdf
--->
 
@@
         Parameter & Symbol & Definition \\ \hline \hline
-            Transcription rate & $\wpr{G}$ & $\in N(100, 20),\ \geq 10$ \\
-            Splicing rate & $\wsr G$ & $\in N(10, 2),\ \geq 2$ \\
-            mRNA degradation rate & $\xdr{G}$ & $\in N(5, 1),\ \geq 2$ \\
-            Translation rate & $\ypr{G}$ & $\in N(5, 1),\ \geq 2$ \\
-            Protein degradation rate & $\ydr G$ & $\in N(3, 0.5),\ \geq 1$ \\
+        Transcription rate & $\wpr{G}$ & $\in N(50, 10),\ \geq 10$ \\
+        Splicing rate & $\wsr G$ & $\in N(5, 1),\ \geq 1$ \\
+        Translation rate & $\ypr{G}$ & $\in N(5, 1),\ \geq 1$ \\
+        Pre-mRNA half-life & $\whl{G}$ & $\in N(0.15, 0.03),\ \geq 0.05$ \\
+        Mature mRNA half-life & $\xhl{G}$ & $\in N(0.15, 0.03),\ \geq 0.05$ \\
+        Protein half-life rate & $\yhl G$ & $\in N(0.25, 0.05),\ \geq 0.1$ \\
         Interaction strength & $\str{G,H}$ & $\in 10^{U(0, 2)}$ * \\
@@
         Cooperativity factor & $\coop G$ & $\in [0, 1]$ * \\ \hline\hline
-            Dissociation constant & $\dis H$ & $= 0.5 \times \frac{\wpr H \times \ypr H}{\xdr H \times \ydr H}$ \\
-            Regulation activity & $\buf{G,H}$ & $= \left(\str{G,H} \times \frac{\y H}{\dis H}\right) ^ {\hill{G,H}}$ \\
+        Pre-mRNA degradation rate & $\wdr G$ & $= \ln(2)\ /\ \whl G$ \\
+        Mature mRNA degradation rate & $\xdr G$ & $= \ln(2)\ /\ \xhl G$ \\
+        Protein degradation rate & $\ydr G$ & $= \ln(2)\ /\ \yhl G$ \\
+        Dissociation constant & $\dis H$ & $= 0.5 \times \frac{\wpr H \times \ypr H}{\wdr H \times \xdr H \times \ydr H}$ \\
+        Regulatory strength & $\buf{G,H}$ & $= \left(\str{G,H} \times \y H\ /\ \dis H\right) ^ {\hill{G,H}}$ \\
         Basal expression & $\ba G$ & $= \begin{cases} 1 & \mbox{if } \rp{G} = \emptyset \\ 0.0001 & \mbox{if } \rn{G} = \emptyset \mbox{ and } \rp{G} \neq \emptyset \\ 0.5 & \mbox{otherwise} \end{cases}$ * \\ \hline
@@
 
-<!--  mention that this is basically an ode? -->
 ## Simulate single cells {#sec:dyngen-simcell}
-
-<!-- 
-%TODO: STOCKS[@kierzek_stocksstochastickinetic_2002] uses GSSA to simulate something.
-% maybe we could have a look at that!
-% or https://nyaspubs.onlinelibrary.wiley.com/doi/full/10.1111/j.1749-6632.2008.03756.x
-
-% or even https://academic.oup.com/bioinformatics/article/24/10/1318/178687 ('AND' and 'OR')
-% and https://academic.oup.com/bioinformatics/article/25/9/1205/203870 (regulation motifs, kinetics)
-% https://nyaspubs.onlinelibrary.wiley.com/doi/full/10.1111/j.1749-6632.2008.03756.x (modules of genes)
-
-%TODO: improve this section a lot. Basically, explain SSA and GillespieSSA2.
--->
-
-`dyngen` uses Gillespie's stochastic simulation algorithm (SSA)[@gillespie_exactstochasticsimulation_1977] to simulate dynamic processes.
+dyngen uses Gillespie's stochastic simulation algorithm (SSA)[@gillespie_exactstochasticsimulation_1977] to simulate dynamic processes.
 An SSA simulation is an iterative process where at each iteration one reaction is triggered.
@@
 
-`GillespieSSA2` is an optimised library for performing SSA simulations.
+GillespieSSA2 is an optimised library for performing SSA simulations.
 The propensity functions are compiled to C++ and SSA approximations can be used which allow to trigger many reactions simultaneously at each iteration.
@@
 ## Simulate experiment {#sec:dyngen-experiment}
-From the SSA simulation we obtain the abundance levels of all the molecules at the different time points.
+From the SSA simulation we obtain the abundance levels of all the molecules at every state.
 We need to replicate technical effects introduced by experimental protocols in order to obtain data that is similar to real data.
@@
 ### Sample cells
-Cells can be sampled from an unsynchronised population of single cells (snapshot) or at multiple time points in a synchronised population (time series).
+In this step, $N$ cells are sampled the simulations.
+Two approaches are implemented: sampling from an unsynchronised population of single cells (snapshot) or sampling at multiple time points in a synchronised population (time series).
 
-**Snapshot:** Cells are sampled uniformly from the different time points in the simulation.
+**Snapshot** The backbone consists of several states linked together by transition edges with length $L_i$, to which the different states in the different simulations have been mapped (Figure\ \ref{fig:sample_cells}A).
+From each transition, $N_i = N / \frac{L_i}{\sum L_i}$ cells are sampled uniformly, rounded such that $\sum N_i = N$.
 
-**Time series:** The timeline of the simulations is divided into multiple chunks. From several of these chunks, cells are sampled. For each cell it is known at which time point it was sampled.
+**Time series** Assuming that the final time of the simulations is $T$, the interval $[0, T]$ is divided into $k$ equal intervals of width $w$ separated by $k-1$ gaps of width $g$.
+$N_i = N / k$ cells are sampled uniformly from each interval (Figure\ \ref{fig:sample_cells}B), rounded such that $\sum N_i = N$.
+By default, $k = 8$ and $g = 0.75$.
+For usual dyngen simulations, $10 \leq T \leq 20$.
+For larger values of $T$, $k$ and $g$ should be increased accordingly.
+
+\begin{figure}[htb!]
+    \centering
+    \includegraphics[width=\linewidth]{fig/sample_cells.pdf}
+    \caption{
+        \textbf{Two approaches can be used to sample cells from simulations: snapshot and time-series.}
+    }
+    \label{fig:sample_cells}
+\end{figure}
 
@@
 
-# Author contributions {#sec:dyngen-contributions}
+## Extract state-specific regulatory effect {#sec:dyngen-extractgrn}
+When desired, the regulatory effect of a regulator $R$ on a target $T$ can be determined and returned (See Figure\ \ref{fig:explain_methods}).
+This information is useful, amongst others, for benchmarking casewise network inference methods.
+To calculate the regulatory effect of $R$ on $T$ at a particular state during a simulation,
 
-* R.C. and W.S. designed the study.
-* R.C. and W.S. performed the experiments and analysed the data.
-* R.C. and W.S. implemented the `dyngen` software package.
-* R.C. wrote the original manuscript.
-* R.C., W.S., and Y.S. reviewed and edited the manuscript.
-* Y.S. supervised the project.
+## Comparison of casewise network inference methods {#sec:dyngen:nicompare}

### 2019-12-01 analysis/small_bifurc.R
regen..

@@ -166,7 +166,13 @@ ann_colours <- list(
   `Regulator module` = gene_colours,
   `Target module` = gene_colours
 )
-color <- c("white", grDevices::colorRampPalette(RColorBrewer::brewer.pal(9, "Blues")[-1:-3])(99))
+# color <- c("white", grDevices::colorRampPalette(RColorBrewer::brewer.pal(9, "Blues")[-1:-3])(99))
+color <- c(
+  rev(grDevices::colorRampPalette(RColorBrewer::brewer.pal(7, "Reds"))(49)),
+  "#BBBBBB",
+  grDevices::colorRampPalette(RColorBrewer::brewer.pal(7, "Blues"))(49)
+)
+breaks <- seq(-1, 1, length.out = length(color) + 1)
 
 cell_df <- tibble(
   id = dataset$cell_ids,
@@ -176,7 +182,7 @@ cell_df <- tibble(
 ) %>%
   arrange(group_ord, pseudotime)
 pheatmap::pheatmap(
-  abs(t(as.matrix(dataset$feature_network_sc[cell_df$id,Matrix::colMeans(dataset$feature_network_sc != 0) > 0]))),
+  t(as.matrix(dataset$feature_network_sc[cell_df$id,Matrix::colMeans(dataset$feature_network_sc != 0) > 0])),
   cluster_cols = FALSE,
   show_rownames = FALSE,
   show_colnames = FALSE,
@@ -184,6 +190,7 @@ pheatmap::pheatmap(
   annotation_row = interaction_info,
   annotation_colors = ann_colours,
   color = color,
+  breaks = breaks,
   border = NA,
   filename = "fig/small_bifurc/scgrn.pdf",
   width = 8,
@@ -191,6 +198,23 @@ pheatmap::pheatmap(
   angle_col = 315,
   treeheight_col = 0,
   treeheight_row = 0,
+  gaps_col = which(diff(cell_df$group_ord) != 0),
+  clustering_distance_rows = "correlation"
+)
+
+pheatmap::pheatmap(
+  t(as.matrix(dataset$feature_network_sc[,Matrix::colMeans(dataset$feature_network_sc != 0) > 0])),
+  show_rownames = FALSE,
+  show_colnames = FALSE,
+  color = color,
+  breaks = breaks,
+  border = NA,
+  filename = "fig/small_bifurc/scgrn_noann.pdf",
+  width = 8,
+  height = 6,
+  angle_col = 315,
[…]

### 2019-12-01 analysis/network_inference/wip_other_methods.R
add script

@@ -1,0 +1,62 @@
+
+library(pandaR)
+# exprm <- as.matrix(expr)
+regulators <- unique(model$feature_network$from)
+targets <- colnames(exprm)
+samples <- rownames(exprm)
+
+# motif <- crossing(tf = regulators, gene = targets) %>% mutate(score = 1) %>% as.data.frame()
+#
+# out <- lioness(motif, t(exprm), progress = TRUE, hamming = .1)
+#
+#
+# data(pandaToyData)
+# linonessRes <- lioness(pandaToyData$motif,
+#                        pandaToyData$expression[,1:20],pandaToyData$ppi,hamming=.1,progress=FALSE)
+#
+
+base <- dynutils::calculate_similarity(
+  expr[,regulators],
+  expr[,targets],
+  method = "spearman",
+  margin = 2
+)
+
+sample_i <- 1
+sample_sp <- dynutils::calculate_similarity(
+  expr[-sample_i,regulators],
+  expr[-sample_i,targets],
+  method = "spearman",
+  margin = 2
+)
+
+pred <-
+  reshape2::melt(base - sample_sp, varnames = c("from", "to"), value.name = "importance") %>%
+  arrange(desc(importance)) %>%
+  head(10000)
+
+ground_truth_sample <-
+  with(
+    dataset$feature_network_sc[sample_i,,drop=FALSE] %>% Matrix::summary(),
+    tibble(
+      from = factor(dataset$feature_network$from[j], levels = regulators),
+      to = factor(dataset$feature_network$to[j], levels = targets),
+      gold_effect = dataset$feature_network$effect[j],
+      gold_value = x,
+      gold = 1
+    )
+  )
+
+
+pred2 <- pred %>%
+  left_join(ground_truth_sample, by = c("from", "to")) %>%
+  mutate(gold = gold %|% 0, gold_value = gold_value %|% 0, gold_effect = gold_effect %|% 0)
+
+
+eval_pyscenic_sample <- GENIE4::evaluate_ranking_direct(
+  values = pred2$importance,
+  are_true = pred2$gold,
+  num_positive_interactions = nrow(ground_truth_sample),
+  num_possible_interactions = length(regulators) * length(targets)
+)
+

### 2019-12-01 manuscript.Rmd
update manuscript

@@
 
-For every gene $G$ in the GRN, the simulation will keep track of the abundance levels of three molecules, a pre-mRNA, a mature mRNA, and a protein.
-These abundance levels are represented as $\w G$, $\x G$ and $\y G$ respectively.
+Simulating a cell's GRN makes use of a stochastic framework which tracks the abundance levels of molecules over time in a discrete quantity.
+For every gene $G$, the abundance levels of three molecules are tracked, namely of corresponding pre-mRNAs, mature mRNAs and proteins, which are represented by the terms $\w G$, $\x G$ and $\y G$ respectively.
+The GRN defines how a reaction affects the abundance levels of molecules and how likely it will occur.
+Gibson and Bruck[@gibson_probabilisticmodelprokaryotic_2000] provide a good introduction to modelling gene regulation with stochastic frameworks, on which many of the concepts below are based.
 
-Throughout a simulation, the abundance levels of molecules are affected by five reactions, namely transcription, splicing, mRNA degradation, translation, and protein degradation.
-Each reaction consists of its propensity -- a formula to calculate the probability of the reaction occurring during an infinitesimal time interval -- and the effect -- how it will affect the current state if triggered.
+For every gene in the GRN a set of reactions are defined, namely transcription, splicing, translation, and degradation.
+Each reaction consists of a propensity function -- a formula $f(.)$ to calculate the probability $f(.) \times \text{d}t$ of it occurring during a time interval $\text{d}t$ -- and the effect -- how it will affect the current state if triggered.
 
@@
 
-The propensity of all reactions except transcription are straight-forward (Table\ \ref{tab:reaction_def}, right).
-The degradation and splicing propensity of pre-mRNAs of a gene $G$ are linear functions of the abundance $\w G$, namely $\wdr G \times \w G$ and $\wsr G \times \w G$ respectively.
-Similarly, the degradation and translation of mature mRNAs are linear functions of $\x G$, namely $\xdr G \times \x G$ and $\ypr G \times \x G$.
-Degradation of proteins is also a linear function of $\y G$, namely $\ydr G \times \y G$.
-Constant parameters $\wdr G$, $\wsr G$, $\xdr G$, $\ypr G$, and $\ydr G$ are from respective normal distributions (Table\ \ref{tab:reaction_params}).
-These formulae are based on those of GNW[@schaffter_genenetweaversilicobenchmark_2011], except that the production and decay of molecules have been separated into two separate formulae.
+The propensity of all reactions except transcription are all linear functions (Table\ \ref{tab:reaction_def}, right) of the abundance level of some molecule multiplied by a constant drawn from a normal distribution (Table\ \ref{tab:reaction_params}).
+The propensity of transcription of a gene $G$ depends on the abundance levels of its TFs.
+
+In the most simple case, $G$ is regulated by a single TF $H$, which can either be bound (state $S_0$) or unbound (state $S_1$).
+The probability $P(S_1)$ that $H$ is bound to the promoter region of $G$ is given by the abundance levels of $H$.
 
@@
         Dissociation constant & $\dis H$ & $= 0.5 \times \frac{\wpr H \times \ypr H}{\wdr H \times \xdr H \times \ydr H}$ \\
-        Regulatory strength & $\buf{G,H}$ & $= \left(\str{G,H} \times \y H\ /\ \dis H\right) ^ {\hill{G,H}}$ \\
+        Binding & $\buf{G,H}$ & $= \left(\str{G,H} \times \y H\ /\ \dis H\right) ^ {\hill{G,H}}$ \\
         Basal expression & $\ba G$ & $= \begin{cases} 1 & \mbox{if } \rp{G} = \emptyset \\ 0.0001 & \mbox{if } \rn{G} = \emptyset \mbox{ and } \rp{G} \neq \emptyset \\ 0.5 & \mbox{otherwise} \end{cases}$ * \\ \hline
@@
     \centering
-    \includegraphics[width=\linewidth]{fig/sample_cells.pdf}
+    \includegraphics[width=.6\linewidth]{fig/sample_cells.pdf}
     \caption{
@@
 
-## Extract state-specific regulatory effect {#sec:dyngen-extractgrn}
-When desired, the regulatory effect of a regulator $R$ on a target $T$ can be determined and returned (See Figure\ \ref{fig:explain_methods}).
+## Determining the casewise ground-truth regulatory network {#sec:dyngen-extractgrn}
+Calculating the regulatory effect of a regulator $R$ on a target $T$ (Figure\ \ref{fig:explain_methods}F) requires determining the contribution of $R$ in the propensity function of the transcription of $T$ (section\ \ref{sec:dyngen-reactions}) with respect to other regulators.
 This information is useful, amongst others, for benchmarking casewise network inference methods.
-To calculate the regulatory effect of $R$ on $T$ at a particular state during a simulation,
 
-## Comparison of casewise network inference methods {#sec:dyngen:nicompare}
+The regulatory effect of $R$ on $T$ at a particular state $S$ is defined as the change in the propensity of transcription when $R$ is set to zero, scaled by the inverse of the pre-mRNA production rate of $T$.
+More formally:
+
+\begin{eqnarray*}
+  \text{regeffect}_G & = \frac{\text{proptrans}_G(S) - \text{proptrans}_G(S[\y T \leftarrow 0])}{\wpr G}
+\end{eqnarray*}
+
+Determining the regulatory effect for all interactions and cells in the dataset yields the complete casewise ground-truth GRN (Figure\ \ref{fig:casewise_grn}).
+The regulatory effect lie between $[-1, 1]$, where -1 represents complete inhibition of $T$ by $R$, 1 represents maximal activation of $T$ by $R$, and 0 represents inactivity of the regulatory interaction between $R$ and $T$.
+
+\begin{figure}[htb!]
+    \centering
+    \includegraphics[width=.8\linewidth]{fig/small_bifurc/scgrn.pdf}
+    \caption{
+        \textbf{The casewise regulatory effects of all interactions, computed on cells part of a bifurcation trajectory.} Negative values correspond to inhibitory interactions, positive values to activating interactions, and zero values correspond to inactive interactions.
+    }
+    \label{fig:casewise_grn}
+\end{figure}
+
+## Comparison of casewise network inference methods {#sec:dyngen-nicompare}
+Several datasets were generated using the different predefined backbones.
+For every cell in the dataset, the transcriptomics profile and the corresponding casewise ground-truth regulatory network was determined (Section\ \ref{sec:dyngen-extractgrn}).
+
+Several casewise NI methods were considered for comparison: SCENIC[@aibar_scenicsinglecellregulatory_2017], LIONESS[@kuijjer_estimatingsamplespecificregulatory_2015; @kuijjer_estimatingsamplespecificregulatory_2019], and SSN[@liu_personalizedcharacterizationdiseases_2016].
+
+LIONESS[@kuijjer_estimatingsamplespecificregulatory_2015; @kuijjer_estimatingsamplespecificregulatory_2019] computes the Pearson correlation between a regulator and one of its targets for all cells.
+The possible targets for a given regulator is predefined according to prior motif analysis.
+To derive casewise GRNs for a particular cell $C$, the Pearson correlation is recomputed for all cells excluding $C$.
+The difference in Pearson correlations results in a ranking of regulatory interactions for $C$.
+
+SSN[@liu_personalizedcharacterizationdiseases_2016] TODO: wip.
+
+SCENIC consists of four main steps: classical network inference with arboreto (similar to GENIE3), selection of 'regulons' from the GRN, filtering of the regulons using motif analysis, and casewise scoring of each regulon for each cell using AUCell.

### 2019-12-02 manuscript.Rmd
update

@@
 
-In the most simple case, $G$ is regulated by a single TF $H$, which can either be bound (state $S_0$) or unbound (state $S_1$).
-The probability $P(S_1)$ that $H$ is bound to the promoter region of $G$ is given by the abundance levels of $H$.
-
 \begin{table}[ht]
@@
 \end{table}
+
+\newcommand{\proptran}{f}
+\newcommand{\ai}[2]{$S_{#1} = S_{#2b}$}
+\newcommand{\yk}[1]{\frac{y_#1}{k_#1}^{c_#1}}
+\newcommand{\wi}[1]{\nu_#1}
+
+The propensity of the transcription of a gene $G$ is inspired by thermodynamic models of gene regulation [needcitation], in which the promoter of $G$ can be bound or unbound by a set of $N$ transcription factors $H_i$.
+Let $\proptran(\y 1, \y 2, \ldots, \y N)$ denote the propensity function of $G$, in function of the abundance levels of the transcription factors.
+
+### Propensity of transcription when $N=1$
+In the simplest case when $N=1$, the promoter can be in one of two states.
+In state $S_0$, the promoter is not bound by any transcription factors, and in state $S_1$ the promoter is bound by $H_1$.
+Each state $S_j$ is linked with a relative activation $\alpha_j$, a number between 0 and 1 representing the activity of the promoter at this particular state.
+The propensity function is thus equal to the expected value of the activity of the promoter multiplied by the pre-mRNA production rate of $G$.
+
+\begin{align}
+  f(y_1, y_2, \ldots, y_N) & = \text{wpr} \cdot \sum_{j = 0}^{2^N - 1} \alpha_j \cdot P(S_j) \label{eqn:activ0} \\
+\end{align}
+
+For $N=1$, $P(S_1)$ is equal to the Hill equation, where $k_i$ represents the concentration of $H_i$ at half-occupation and $n_i$ represents the Hill coefficient.
+
+\begin{align}
+  P(S_1) & = \frac{y_1^{n_1}}{k_1^{n_1} + y_1^{n_1}} \\
+     & = \frac{(y_1/k_1)^{n_1}}{1 + (y_1/k_1)^{n_1}}
+\end{align}
+
+The Hill equation can be simplified by letting $\nu_i = \left(\frac{y_i}{k_i}\right)^{n_i}$.
+
+\begin{align}
+P(S_1) & = \frac{\nu_1}{1 + \nu_1} \label{eqn:hillsimp}
+\end{align}
+
+Since $P(S_0) = 1 - P(S_1)$, the activation function is formulated and simplified as follows.
+
+\begin{align}
+f(y_1) & = \text{wpr} \cdot \left(\alpha_0 \cdot P(S_0) + \alpha_1 \cdot P(S_1)\right) \\
+   & = \text{wpr} \cdot \left(\alpha_0 \cdot \frac{1}{1 + \nu_1} + \alpha_1 \cdot \frac{\nu_1}{1 + \nu_1}\right) \\
+   & = \text{wpr} \cdot \frac{\alpha_0 + \alpha_1 \cdot \nu_1}{1 + \nu_1} \\
+\end{align}
+
+### Propensity of transcription when $N=2$
+
+When $N=2$, there are four states $S_j$.
+The relative activations $\alpha_j$ can be defined such that $H_1$ and $H_2$ are independent (additive) or synergistic (multiplicative).
+In order to define the propensity of transcription $f(.)$, the Hill equation $P(S_j)$ is extended for two transcription factors.
+
+Let $w_j$ be the numerator of $P(S_j)$, defined as the product of all transcription factors bound in that state:
+
+\begin{align}
+w_0 & = 1 \\
+w_1 & = \nu_1 \\
+w_2 & = \nu_2 \\
+w_3 & = \nu_1 \cdot \nu_2
+\end{align}
+
+The denominator of $P(S_j)$ is then equal to the sum of all $w_j$. The probability of state $S_j$ is thus defined as:
+
+\begin{align}
+    P(S_j) & = \frac{w_j}{\sum_{j=0}^{j < 2^N} w_j} \\
+       & = \frac{w_j}{1 + \nu_1 + \nu_2 + \nu_1 \cdot \nu_2} \\
+       & = \frac{w_j}{\prod_{i=1}^{i \leq N} (\nu_i + 1)}
+\end{align}
+
+Substituting $P(S_j)$ and $w_j$ into $f(.)$ results in the following equation:
+
+\begin{align}
+f(y_1, y_2) & = \text{wpr} \cdot \sum_{j = 0}^{2^N - 1} \alpha_j \cdot P(S_j) \\
+ & = \text{wpr} \cdot \frac{\sum_{j = 0}^{2^N - 1} \alpha_j \cdot w_j}{\prod_{i=1}^{i \leq N} (\nu_i + 1)} \\
+ & = \text{wpr} \cdot \frac{\alpha_0 + \alpha_1 \cdot \nu_1 + \alpha_2 \cdot \nu_2 + \alpha_3 \cdot \nu_1 \cdot \nu_2}{(\nu_1 + 1) \cdot (\nu_2 + 1)} \\
+\end{align}
+
+### Propensity of transcription for an arbitrary $N$
+For an arbitrary $N$, there are $2^N$ states $S_j$.
+The relative activations $\alpha_j$ can be defined such that $H_1$ and $H_2$ are independent (additive) or synergistic (multiplicative).
+In order to define the propensity of transcription $f(.)$, the Hill equation $P(S_j)$ is extended for $N$ transcription factors.
+
+Let $w_j$ be the numerator of $P(S_j)$, defined as the product of all transcription factors bound in that state:
+
+\begin{align}
+  w_j & = \prod_{i=1}^{i \leq N} (j \text{ mod } i) = 1 \text{ ? } \nu_i \text{ : } 1
+\end{align}
+
+The denominator of $P(S_j)$ is then equal to the sum of all $w_j$. The probability of state $S_j$ is thus defined as:
+
+\begin{align}
+P(S_j) & = \frac{w_j}{\sum_{j=0}^{j < 2^N} w_j} \\
+& = \frac{w_j}{\prod_{i=1}^{i \leq N} (\nu_i + 1)}
+\end{align}
+
+Substituting $P(S_j)$ into $f(.)$ yields:
+
+\begin{align}
+f(y_1, y_2, \ldots, y_N) & = \text{wpr} \cdot \sum_{j = 0}^{2^N - 1} \alpha_j \cdot P(S_j) \\
+& = \text{wpr} \cdot \frac{\sum_{j = 0}^{2^N - 1} \alpha_j \cdot w_j}{\prod_{i=1}^{i \leq N} (\nu_i + 1)} \label{eqn:prop2n}
+\end{align}
+
+### Propensity of transcription for a large $N$
+For large values of $N$, computing $f(.)$ is practically infeasible as it requires performing $2^N$ summations.
+In order to greatly simplify $f(.)$, $\alpha_j$ could be defined as 0 when one of the regulators inhibits transcription and 1 otherwise.
+
+\begin{equation}
+\alpha_j = \begin{cases}
+ 0 & \text{ if } \exists i : j \text{ mod } i = 1 \text{ and } H_i \text{ represses } G \\
+ 1 & \text{otherwise}
+\end{cases} \label{eqn:assalpha}
+\end{equation}
+
+Substituting equation \ref{eqn:assalpha} into equation \ref{eqn:prop2n} and defining $R = \{1, 2, \ldots, N\}$ and $R^+ = \{i | H_i \text{ activates } G\}$ yields the simplified propensity function:
+
+\begin{align}
+f(y_1, y_2, \ldots, y_N) & = \text{wpr} \cdot \frac{\prod_{i \in R^+} (\nu_i + 1)}{\prod_{i \in R} (\nu_i + 1)}
+\end{align}
+
+### Independence, synergism and basal expression
+The definition of $\alpha_j$ as in equation \ref{eqn:assalpha} presents two main limitations.
+Firstly, since $\alpha_0 = 1$, it is impossible to tweak the propensity of transcription when no transcription factors are bound.
+Secondly, it is not possible to tweak the independence and synergism of multiple regulators.
+
+Let $\text{ba} \in [0,1]$ denote the basal expression strength $G$ (i.e.
+how much will $G$ be expressed when no transcription factors are bound), and $\text{sy} \in [0,1]$ denote the synergism of regulators $H_i$ of $G$, the transcription propensity becomes:
+
+\begin{align}
+f(y_1, y_2, \ldots, y_N) & = \text{wpr} \cdot \frac{\text{ba} - \text{sy}^{|R^+|} + \prod_{i \in R^+} (\nu_i + \text{sy})}{\prod_{i \in R} (\nu_i + 1)}
+\end{align}
 

### 2019-12-02 manuscript.Rmd
update

@@
 Let $\proptran(\y 1, \y 2, \ldots, \y N)$ denote the propensity function of $G$, in function of the abundance levels of the transcription factors.
+The following subsections explain and define the propensity function when $N=1$, $N=2$, and finally for an arbitrary $N$.
 
@@
 
-LIONESS[@kuijjer_estimatingsamplespecificregulatory_2015; @kuijjer_estimatingsamplespecificregulatory_2019] computes the Pearson correlation between a regulator and one of its targets for all cells.
+LIONESS[@kuijjer_estimatingsamplespecificregulatory_2015; @kuijjer_estimatingsamplespecificregulatory_2019] 
+first infers a GRN by computing the Pearson correlation between regulators and targets for all samples and then post-processing this GRN.
 The possible targets for a given regulator is predefined according to prior motif analysis.
@@
 The difference in Pearson correlations results in a ranking of regulatory interactions for $C$.
+LIONESS was excluded from this comparison as it requires motif data to be available.
 
-SSN[@liu_personalizedcharacterizationdiseases_2016] TODO: wip.
+SSN[@liu_personalizedcharacterizationdiseases_2016] is similar to LIONESS in methodology.
+To infer a casewise GRN for one sample, also the difference between the Pearson correlation between all (regulator, target) pairs is calculated for all samples and the Pearson correlation between all (regulator, target) pairs is calculated for all samples except one.
+SNN was excluded from this comparison as no software was provided by the authors.
 
-SCENIC consists of four main steps: classical network inference with arboreto (similar to GENIE3), selection of 'regulons' from the GRN, filtering of the regulons using motif analysis, and casewise scoring of each regulon for each cell using AUCell.
+SCENIC[@aibar_scenicsinglecellregulatory_2017] is a pipeline that consists of four main steps.
+Step 1: classical network inference is performed with arboreto, which is similar to GENIE3[@huynh-thu_inferringregulatorynetworks_2010].
+Step 2: setting various cutoffs on the ranked interactions to form groups called 'regulons', each containing one regulator and multiple targets.
+Step 3: filter the regulons using motif analysis.
+Step 4: for each cell, determine an activity score of each regulon using AUCell.
+For this analysis, the Python implementation of SCENIC was used, namely pySCENIC.
+Since dyngen does not generate motif data, step 3 in this analysis is skipped.
+
+The AUROC and AUPR metrics are common metrics for evaluating a predicted GRN with a ground-truth GRN.
+To compare a predicted casewise GRN with the ground-truth casewise GRN, the top 10'000 interactions per cell were retained.
+For each cell-specific network, the AUROC and AUPR were calculated.

### 2019-12-05 manuscript.Rmd
pushy pushy

@@
 
-### Single-cell network inference
+### Casewise network inference
+
+Using a collection of transcriptomic profiles, network inference (NI) methods predict which genes are the regulators of a target gene.
+The output of an NI method is called a gene regulatory network (GRN, Figure\ \ref{fig:scgrn}A, top left), in which each predicted interaction can be assigned a predicted strength or effect (activating or inhibitory).
+
+However, cells are very molecularly homogeneous, and only a fraction of all possible gene regulatory interactions is active at any point in time.
+The advent of single-cell omics have enabled so-called casewise network inference (CNI) methods\footnote{Other names for CNI are single-cell network inference or sample-specific network inference.
+Single-cell network inference (SCNI) is often used ambiguously, as classical network inference methods adapted to deal with single-cell omics data are also referred to as such.
+Sample-specific network inference is also ambiguous in the context of single-cell omics, as a sample is from which single cells are extracted.} to try tackling a more complex variant of the same problem, namely "Which regulators are regulating which targets in which profiles?" The end result of such methods is a single GRN for every profile.
+
+A few CNI methods have already been developed, such as SCENIC[@aibar_scenicsinglecellregulatory_2017], LIONESS[@kuijjer_estimatingsamplespecificregulatory_2019], and SNN[@liu_personalizedcharacterizationdiseases_2016].
+However, a quantitative assessment of the performance of such methods is hitherto lacking, mainly because no large-scale information on cell-specific regulatory interactions with which to evaluate CNI methods exists.
+
+By knocking down the expression of a regulator, it is possible to determine its regulatory effect on targets by comparing the change in transcription rates with and without knock-down.
+Knocking down every regulator once results in a cell-specific GRN (Figure\ \ref{fig:scgrn}A) than can be used as a ground-truth for evaluating CNI methods (Figure\ \ref{fig:scgrn}B).
+
+\begin{figure}[htb!]
+    \centering
+    \includegraphics[width=.8\linewidth]{fig/network_inference/cni.pdf}
+    \caption{
+    \textbf{}
+    }
+    \label{fig:scgrn}
+\end{figure}
 
@@
         Protein degradation rate & $\ydr G$ & $= \ln(2)\ /\ \yhl G$ \\
-        Dissociation constant & $\dis H$ & $= 0.5 \times \frac{\wpr H \times \ypr H}{\wdr H \times \xdr H \times \ydr H}$ \\
+        Dissociation constant & $\dis H$ & $= 0.5 \times \frac{\wpr H \times \wsr H \times \ypr H}{(\wdr H + \wsr H) \times \xdr H \times \ydr H}$ \\
         Binding & $\buf{G,H}$ & $= \left(\str{G,H} \times \y H\ /\ \dis H\right) ^ {\hill{G,H}}$ \\
@@
 
-For $N=1$, $P(S_1)$ is equal to the Hill equation, where $k_i$ represents the concentration of $H_i$ at half-occupation and $n_i$ represents the Hill coefficient.
+For $N=1$, $P(S_1)$ is equal to the Hill equation, where $k_i$ represents the concentration of $H_i$ at half-occupation and $n_i$ represents the Hill coefficient. Typically, $n_i$ is between [1,10]
 
@@
 Step 4: for each cell, determine an activity score of each regulon using AUCell.
+As a post-processing of this output, all regulon and the corresponding activity scores are combined back into a casewise GRN consisting of (cell, regulator, target, score) pairs.
+If multiple regulons contain the same (regulator, target) interaction, the maximum activity score is retained.
 For this analysis, the Python implementation of SCENIC was used, namely pySCENIC.

### 2019-12-05 analysis/network_inference/2_evaluate_methods.R

@@ -19,43 +19,56 @@ datasets <- list_as_tibble(map(dataset_files, function(dataset_file) {
 
 methods <- list(
   "bred" = cni_bred(num_trees = 1000L),
-  "deltacor" = cni_deltacor(),
+  "SNN*" = cni_deltacor(),
   "pySCENIC GBM" = cni_pyscenic_sgbm(subsample = 1, n_estimators = 500L),
   "pySCENIC SGBM" = cni_pyscenic_sgbm(subsample = .9, n_estimators = 5000L)
 )
 
-evals <- map(methods, function(method) {
-  evaluate_ti_method(
-    dataset = datasets,
-    method = method,
-    metrics = list(auc = cni_auc),
-    parameters = NULL
-  )
-})
-
-aucs <- map_df(seq_along(evals), function(i) {
-  summ <- evals[[i]]$summary
-  mapdf_dfr(summ, function(li) {
-    li$evals %>% mutate(
-      cni_method_id = names(methods)[[i]],
-      dataset_id = li$dataset_id
+eval_file <- "derived_files/network_inference/evaluation.rds"
+if (!file.exists(eval_file)) {
+  evals <- map(methods, function(method) {
+    evaluate_ti_method(
+      dataset = datasets,
+      method = method,
+      metrics = list(auc = cni_auc),
+      parameters = NULL
     )
   })
-})
-int_ns <- mapdf_dfr(datasets, function(dat) dat$regulatory_network_sc %>% group_by(cell_id) %>% summarise(nints = n()) %>% mutate(dataset_id = dat$id))
-aucs <- aucs %>% left_join(int_ns, c("cell_id", "dataset_id"))
 
+  aucs <- map_df(seq_along(evals), function(i) {
+    summ <- evals[[i]]$summary
+    mapdf_dfr(summ, function(li) {
+      li$evals %>% mutate(
+        cni_method_id = names(methods)[[i]],
+        dataset_id = li$dataset_id
+      )
+    })
+  })
+  int_ns <- mapdf_dfr(datasets, function(dat) dat$regulatory_network_sc %>% group_by(cell_id) %>% summarise(nints = n()) %>% mutate(dataset_id = dat$id))
+  aucs <- aucs %>% left_join(int_ns, c("cell_id", "dataset_id"))
+
+  write_rds(lst(evals, aucs), eval_file, compress = "gz")
+}
+
+list2env(read_rds(eval_file))
+aucs <-
+  aucs %>%
[…]

### 2019-12-05 analysis/network_inference/3_plot_dataset_example.R

@@ -122,7 +122,7 @@ g <- patchwork::wrap_plots(
 ) &
   theme(
     plot.margin = margin(0, 0, 0, 0, "cm"),
-    plot.title = element_text(size = 10)
+    plot.title = element_text(size = 10, hjust = .5)
   )
 ggsave("fig/network_inference/casewise_grn.pdf", width = 8, height = 6)
 

### 2019-12-05 analysis/network_inference/4_compose_plot.R
update results

@@ -7,7 +7,7 @@ grnh <- 5
 bch <- 2.5
 g <- patchwork::wrap_plots(
   explots + labs(tag = "A"),
-  summplots[[2]] +
+  summplots$casewise_casewise +
     scale_x_continuous(limits = c(.4, .7)) +
     scale_y_continuous(limits = c(0, .06)) +
     labs(tag = "B", title = NULL) +

### 2019-12-05 manuscript.Rmd
update

@@
 We introduce dyngen, a multi-modality simulator of single cells and their dynamics (Figure\ \ref{fig:showcase}A).
-A cell is simulated using Gillespie's Stochastic Simulation Algorithm (SSA)[@gillespie_exactstochasticsimulation_1977] where a cell consists of a set of molecules.
-Throughout a simulation, reactions (e.g.
-transcription, splicing) modify the abundance of these molecules.
-The likelihood of a reaction occurring is again dependent on molecule abundance.
+dyngen uses Gillespie's Stochastic Simulation Algorithm (SSA)[@gillespie_exactstochasticsimulation_1977] to simulate cells at a single-molecule and single-reaction levels.
+This means at every time step in the simulation, a single reaction is executed, thereby modifying the abundance of the molecule counts within a cell.
 
@@
         \textbf{Showcase of dyngen functionality.}
-        \textbf{A:}
+        \textbf{A:} The typical process of generating a dataset with dyngen.
         \textbf{B:} Evaluating different types of computational tools requires simulating different types of experiments and extracting different layers of information from the simulation.
@@
 # Results {#sec:dyngen-results}
+dyngen offers multiple significant qualities, each of which we will discuss below.
+it's methodology allows keeping track of many modalities of the cell at a high resolution level
+simulate different various types of dynamic processes, such as cyclic, branching, disconnected
+perform time-series, perturbation, or knockdown experiments.
 
 A cell consists of a set of molecules, the abundance of which are affected by a set of reactions: transcription, splicing, translation, and degradation (Figure\ \ref{fig:simplecyclic}A).
-A gene regulatory network (GRN) defines the reactions that are allowed to occur (Figure\ \ref{fig:simplecyclic}B) and is constructed such that a cell will develop over time (Figure\ \ref{fig:simplecyclic}C--D).
-The likelihood of a reaction occurring is a function of the abundance of key molecules involved in each reaction (Figure\ \ref{fig:simplecyclic}E).
+A gene regulatory network (GRN) defines the reactions that are allowed to occur (Figure\ \ref{fig:simplecyclic}B), which is constructed in such a way that cells slowly develop over time (Figure\ \ref{fig:simplecyclic}C,D).
+With every time step $\text{d}t$ in the simulation, the probability of a reaction occurring is computed (not shown).
+From the probabilities are sampled which reactions occur during this time step $\text{d}t$ (Figure\ \ref{fig:simplecyclic}E).
 
@@
     \caption{
-        \textbf{Key features of dyngen are illustrated using a cyclic toy example.}
-        While this example comprises only of a single cell and 5 genes, dyngen is able to simulate thousands of cells with GRNs containing thousands of genes.
+        \textbf{dyngen models reactions at a single-molecule level and keeps track of multiple levels of information throughput a simulation.} 
+        \textbf{A:} Changes in abundance levels are driven strictly these gene regulatory reactions.
+        \textbf{B:} The input GRN is defined such that it models a dynamic process of interest.
+        \textbf{C:} The reactions define how abundance levels of molecules change at any particular timepoint.
+        \textbf{D:} Firing many reactions can significantly alter the cellular state over time.
+        \textbf{E:} dyngen keeps track of the reactions that were fired during small intervals of time.
+        \textbf{F:} Similarly, dyngen can also keep track of the regulatory activity of every interaction.
     }
@@
 dyngen includes generators of GRNs which result in many different developmental topologies (Figure\ \ref{fig:example_runs}), including branching, converging, cyclic and even disconnected.
+Custom-defined GRNs offer more fine-grained control over the simulation.
 
@@
 
-Custom-defined GRNs offer more fine-grained control over the simulation.
-Aside from simulating topologies currently not supported by dyngen, this has several other important use-cases.
-Simulations of the same GRN with small perturbations allow to emulate batch effects or perturbation experiments (Figure\ \ref{fig:usecases}).
-Simulating perturbed GRNs allow evaluating trajectory alignment methods -- which attempt to map two or more trajectories onto each other -- or differential network inference methods -- which infer differential regulatory interactions between two or more groups of profiles.
-
-\begin{figure}[ht]
-    \centering
-    \includegraphics[width=0.6\textwidth]{fig/usecases}
-    \caption{
-        \textbf{Examples of simulations with perturbed GRNs.}  
-        \textbf{A:} The cells in the top half were simulated with the same GRN as the cells on the bottom half, except that all parameters (e.g. strength of the interaction) were randomised.
-        \textbf{B:} Only 10 interactions in the GRN were randomised. In this example, the effect of the GRN perturbation is more subtle.
-    }
-    \label{fig:usecases}
-\end{figure}
-
-dyngen can be used to simulate different experimental conditions.
-By default, dyngen supports snapshot experiments (uniformly sampling from an asynchronous dynamic process) and time-series experiments (sampling cells from different intervals in the simulation).
-However, it is possible to imagine and implement other sampling strategies, such as sampling a cell at a certain time point and once more at a later time point.
-This would allow evaluating the performance of RNA velocity approaches -- which predict the future state of a cell by looking at differences in pre-mRNA and mRNA abundance levels.
+Together, these qualities allow it to be applicable in benchmarking a broad range of use-cases.
+To demonstrate, we apply dyngen on several novel computational approaches which have recently had a major impact on how single-cell analyses are performed but for which quantitative assessment of the performance was hitherto lacking.
 
@@
 However, cells are very molecularly homogeneous, and only a fraction of all possible gene regulatory interactions is active at any point in time.
-The advent of single-cell omics have enabled so-called casewise network inference (CNI) methods\footnote{Other names for CNI are single-cell network inference or sample-specific network inference.
-Single-cell network inference (SCNI) is often used ambiguously, as classical network inference methods adapted to deal with single-cell omics data are also referred to as such.
-Sample-specific network inference is also ambiguous in the context of single-cell omics, as a sample is from which single cells are extracted.} to try tackling a more complex variant of the same problem, namely "Which regulators are regulating which targets in which profiles?" The end result of such methods is a single GRN for every profile.
+Casewise network inference (CNI) methods\footnote{Other terms are commonly used when dealing with data from a particular source.
+For example, single-cell NI when applied to single-cell transcriptomics data; sample-specific NI when applied to bulk transcriptomics; patient-derived NI when applied to bulk profiles of patients.} tackle a more complex variant of NI, namely "Which regulators are regulating which targets in which case?" The end result of such methods is a single GRN for every case, be it a cell, tissue, patient or other sample.
 
-A few CNI methods have already been developed, such as SCENIC[@aibar_scenicsinglecellregulatory_2017], LIONESS[@kuijjer_estimatingsamplespecificregulatory_2019], and SNN[@liu_personalizedcharacterizationdiseases_2016].
-However, a quantitative assessment of the performance of such methods is hitherto lacking, mainly because no large-scale information on cell-specific regulatory interactions with which to evaluate CNI methods exists.
+While a few pioneering CNI approaches have already been developed[@aibar_scenicsinglecellregulatory_2017; @kuijjer_estimatingsamplespecificregulatory_2019; @liu_personalizedcharacterizationdiseases_2016], a quantitative assessment of the performance is hitherto lacking.
+This is not surprising, as neither real nor in silico datasets of cell-specific or even cell-type-specific interactions exists that is large enough so that it can be used as a ground-truth for evaluating CNI methods.
 
-By knocking down the expression of a regulator, it is possible to determine its regulatory effect on targets by comparing the change in transcription rates with and without knock-down.
-Knocking down every regulator once results in a cell-specific GRN (Figure\ \ref{fig:scgrn}A) than can be used as a ground-truth for evaluating CNI methods (Figure\ \ref{fig:scgrn}B).
+dyngen allows deriving the ground-truth regulatory effect of a regulator on a target in a specific cell state.
+This information is derived by knocking down the expression of a regulator and observing the change in transcription rate as a result.
+Performing in silico knockdowns for every regulator in every cell results in the ground-truth casewise GRN (Figure\ \ref{fig:scgrn}A).
+
+We used this ground-truth to compare the performance of three CNI methods  (Figure\ \ref{fig:scgrn}B).
+We calculated the AUROC and AUPR score -- which are common metrics for NI benchmarking -- for each cell individually.
+Computing the mean AUROC and AUPR per dataset showed that pySCENIC significantly outperforms LIONESS and SSN.
+
+This comparison could be extended to include analyses on the scalability of execution time w.r.t dataset size, the stability of the results in function of noise, and the usability toward end users.
 
@@
     \caption{
-    \textbf{}
+    \textbf{dyngen allows benchmarking Casewise Network Inference (CNI) methods.} 
+    \textbf{A:} A cell is simulated using the global GRN (top left). However, at any particular state in the simulation, only a fraction of the gene regulatory interactions are active.
+    \textbf{B:} CNI methods were executed to predict the regulatory interactions that are active in each cell specifically.
+Using the ground-truth casewise GRN, the performance of each method was quantified on 14 dyngen datasets.
     }
@@
 ### RNA velocity
+
+In eukaryotes, a gene is first transcribed to a pre-mRNA, and subsequently spliced into a mature mRNA.
+Both unspliced reads, which map to introns, and spliced reads, mapping to exons are often observed in RNA sequencing experiments, both in bulk[@zeisel_coupledpremrnamrna_2011] and single-cell.
 
@@
 
-LIONESS[@kuijjer_estimatingsamplespecificregulatory_2015; @kuijjer_estimatingsamplespecificregulatory_2019] 
-first infers a GRN by computing the Pearson correlation between regulators and targets for all samples and then post-processing this GRN.
-The possible targets for a given regulator is predefined according to prior motif analysis.
-To derive casewise GRNs for a particular cell $C$, the Pearson correlation is recomputed for all cells excluding $C$.
-The difference in Pearson correlations results in a ranking of regulatory interactions for $C$.
-LIONESS was excluded from this comparison as it requires motif data to be available.
+LIONESS[@@kuijjer_estimatingsamplespecificregulatory_2019; kuijjer_lionessrsinglesample_2019] uses the Pearson correlation to infer casewise GRNs.
+To do so, first the Pearson correlation is calculated between regulators and targets for all samples.
+Next, the Pearson correlation is again calculated for all samples excluding one sample.
+The difference between the two correlation matrices is considered a casewise GRN for that particular profile.
+This process is repeated for all profiles, resulting in a casewise GRN.
 
-SSN[@liu_personalizedcharacterizationdiseases_2016] is similar to LIONESS in methodology.
-To infer a casewise GRN for one sample, also the difference between the Pearson correlation between all (regulator, target) pairs is calculated for all samples and the Pearson correlation between all (regulator, target) pairs is calculated for all samples except one.
-SNN was excluded from this comparison as no software was provided by the authors.
+SSN[@liu_personalizedcharacterizationdiseases_2016] has, in essence, the exact same methodology as LIONESS.
+It is worth noting that the LIONESS preprint was released before the publication of SSN.
+Since no implementation was provided by the authors, we implemented SSN in R using basic R and tidyverse functions[@wickham_welcometidyverse_2019] and marked results from this implementation as "SSN*".
 
@@
 Step 1: classical network inference is performed with arboreto, which is similar to GENIE3[@huynh-thu_inferringregulatorynetworks_2010].
-Step 2: setting various cutoffs on the ranked interactions to form groups called 'regulons', each containing one regulator and multiple targets.
-Step 3: filter the regulons using motif analysis.
-Step 4: for each cell, determine an activity score of each regulon using AUCell.
-As a post-processing of this output, all regulon and the corresponding activity scores are combined back into a casewise GRN consisting of (cell, regulator, target, score) pairs.
-If multiple regulons contain the same (regulator, target) interaction, the maximum activity score is retained.
+Step 2: select the top 10 regulators per target.
+Interactions are grouped together in 'modules'; each module contains one regulator and all of its targets.
+Step 3: filter the modules using motif analysis.
+Step 4: for each cell, determine an activity score of each module using AUCell.
+As a post-processing of this output, all modules and the corresponding activity scores are combined back into a casewise GRN consisting of (cell, regulator, target, score) pairs.
 For this analysis, the Python implementation of SCENIC was used, namely pySCENIC.

### 2019-12-06 analysis/usecase_trajectory_alignment.Rmd
new benchmarking method: compare distances between a couple of trajectories

@@
 d2 <- dist(as.matrix(t(wexpr2)), as.matrix(t(wexpr1)), dist.method="pearson")
-am2 <- dtw(as.matrix(t(wexpr2)), as.matrix(t(wexpr1)), step.pattern = symmetric2, keep.internals = TRUE, open.end = TRUE)
+am2 <- dtw(as.matrix(t(wexpr2)), as.matrix(t(wexpr1)), step.pattern = symmetric2, keep.internals = TRUE, open.end = FALSE)
 
@@
 d3 <- dist(cnts2, cnts1, dist.method="pearson")
-am3 <- dtw(d3, step.pattern = symmetric2, keep.internals = TRUE, open.end = TRUE)
+am3 <- dtw(d3, step.pattern = symmetric2, keep.internals = TRUE, open.end = FALSE)
 ```
 
-We can plot both dimensionality reductions. TODO: why slope low
+We can plot both dynamic time warpings.
+The indicated slope is the trajectory of the warping path we expect.
+Because we surpressed the transcription rate and the mrna degradation rate of C1_TF1, we don't expect there to be any expression past that point.
+We know how long the different parts of the trajectory are, and we also know that the amount of milestones is linked to the length of the trajectory parts.
+
+This means that, if the length of the milestone network between A and C is 66% of the complete length, we expect the whole of the shortened trajectory to match to the first 66% of the complete trajectory.
 
@@
 abline(coef = c(0, sum(dataset1$milestone_network$length[1:2]) / sum(dataset1$milestone_network$length)))
+```
+
+```{r}
+# Calculate euclidean distance metric
+
 ```

### 2019-12-06 analysis/usecase_trajectory_alignment.R

@@ -80,7 +80,7 @@ model <-
     num_cores = 8,
     distance_metric = "pearson",
     tf_network_params = tf_network_default(min_tfs_per_module = 2, sample_num_regulators = function() 1),
-    simulation_params = simulation_default(num_simulations = 10, census_interval = .01)
+    simulation_params = simulation_default(census_interval = .01)
   ) %>%
   generate_tf_network() %>%
   generate_feature_network()
@@ -158,6 +158,8 @@ library(dtw)
 setwd("/home/louisedc/Work/dyngen_analysis/analysis")
 source("dtw_functions.R")
 
+traj_order <- calculate_correct_pseudotime(dataset1)
+
 wpp1 <- get_waypoint_progression(dataset1, 200)
 wpp2 <- get_waypoint_progression(dataset2, 200)
 
@@ -232,6 +234,8 @@ dataset_merged <- wrap_dataset(merged_model)
 
 library(dynplot)
 plot_dimred(dataset_merged)
+
+traj_order <- calculate_correct_pseudotime(dataset_unperturbed)
 
 # Apply dtw to these
 

### 2019-12-06 dyngen_traj_alignment_usecase.R

@@ -1,0 +1,214 @@
+library(tidyverse)
+library(dyngen)
+library(dynplot)
+library(dtw)
+
+set.seed(276)
+
+setwd("/home/louisedc/Work/dyngen_analysis/analysis")
+source("dtw_functions.R")
+
+change_speed <- function(model, target, rate) {
+  param_id <- which(target == model$feature_info$feature_id)[[1]]
+
+  model$feature_info$wpr[param_id] <- model$feature_info$wpr[param_id] * rate
+  model$feature_info$xdr[param_id] <- model$feature_info$xdr[param_id] * rate
+
+  param_wpr <- paste0("wpr_", target)
+  param_xdr <- paste0("xdr_", target)
+
+  model$simulation_system$parameters[param_wpr] <- model$simulation_system$parameters[param_wpr] * rate
+  model$simulation_system$parameters[param_xdr] <- model$simulation_system$parameters[param_xdr] * rate
+
+  model
+}
+
+get_wexpr <- function(dataset, amount){
+  wpp1 <- get_waypoint_progression(dataset, amount)
+  gd1 <- get_geodesic_distances_from_progressions(dataset, wpp1)
+  wexpr1 <- interpolate_expression(dataset, wpp1$percentage, gd1)
+  wexpr1
+}
+
+surpress_module <- function(model, module){
+  fts_to_surpress <- which(module == model$feature_network$to_module)
+  for(index in fts_to_surpress){
+    ft <- model$feature_network$from[[index]]
+    model <- change_speed(model, ft, 0)
+  }
+  model <- model %>% generate_gold_standard() %>% generate_cells() %>% generate_experiment()
+  model
+}
+
+backbone <- bblego(
+  bblego_start("A", type = "simple", num_modules = 4),
+  bblego_linear("A", "B", type = "simple", num_modules = 6),
+  bblego_linear("B", "C", type = "simple", num_modules = 6),
+  bblego_end("C", type = "simple", num_modules = 6)
+)
+
+runmodel <- function(){
+  model <- initialise_model(
+    num_tfs = 50,
+    num_targets = 200,
+    num_hks = 200,
+    num_cells = 1000,
+    backbone = backbone,
+    verbose = TRUE,
[…]

### 2019-12-09 analysis/velocity/0_design.R
update velocity testcase

@@ -3,11 +3,14 @@ library(tidyverse)
 dataset_design <- tibble(
   seed = 1:3
 ) %>%
-  mutate(id = as.character(seed))
+  crossing(backbone = c("linear", "linear_simple", "bifurcating", "cycle", "disconnected")) %>%
+  mutate(id = paste0(backbone, "_", seed))
 
 
 design_velocity <- tribble(
   ~method_id, ~params, ~params_id,
+  "velocyto", list(assumption = "constant_velocity"), "constant_velocity",
+  "velocyto", list(assumption = "constant_unspliced"), "constant_unspliced",
   "scvelo", list(mode = "deterministic"), "deterministic",
   "scvelo", list(mode = "dynamical"), "dynamical",
   "scvelo", list(mode = "stochastic"), "stochastic",

### 2019-12-09 analysis/velocity/1_datasets.R

@@ -3,7 +3,7 @@ library(tidyverse)
 source("analysis/velocity/1_datasets_functions.R")
 library(dyngen)
 
-pwalk(dataset_design[1, ], create_dataset)
+pwalk(dataset_design, create_dataset)
 
 
 
@@ -14,8 +14,8 @@ pwalk(dataset_design[1, ], create_dataset)
 library(dynwrap)
 library(dynplot2)
 
-dataset <- load_dataset("1")
-model <- load_model("1")
+dataset <- load_dataset("linear_simple_1")
+model <- load_model("linear_simple_1")
 
 dataset <- dataset %>% add_dimred(dyndimred::dimred_landmark_mds, pair_with_velocity = F)
 
@@ -24,16 +24,16 @@ dynplot_dimred(dataset) +
   geom_cell_point()
 
 # features
-feature_ids <- model$feature_info %>% filter(!burn) %>% group_by(module_id) %>% slice(1) %>% pull(feature_id)
+feature_ids <- model$feature_info %>% group_by(module_id) %>% slice(1) %>% pull(feature_id)
 
 dynplot_dimred(dataset) +
   geom_cell_point(aes(color = select_feature_expression(feature_id, .data))) +
   facet_wrap_data(feature_id = feature_ids) +
   scale_expression_fillcolour()
 
-
-# one feature spliced vs unspliced
-feature_id <- "B3_TF1"
+# Plot the spliced vs unspliced changes
+# feature_id <- model$feature_info$feature_id[[2]]
+feature_id <- model$feature_info %>% filter(!burn) %>% pull(feature_id) %>% first()
 meta <- model$simulations$meta %>%
   mutate(step_ix = row_number()) %>%
   # filter(simulation_i == 1)
@@ -57,11 +57,11 @@ plotdata <- bind_rows(
 )
 plotdata %>%
   pivot_wider(names_from = "molecule", values_from = "expression") %>%
-  ggplot(aes(spliced, unspliced)) + geom_path(aes(color = sim_time))
+  ggplot(aes(spliced, unspliced)) + geom_path(aes(color = sim_time)) +
+  facet_wrap(~simulation_i)
 
 
 
 plotdata %>%
   filter(simulation_i == 1) %>%
   ggplot(aes(step_ix, expression)) + geom_line(aes(color = molecule))
-

### 2019-12-09 analysis/velocity/1_datasets_functions.R

@@ -1,4 +1,4 @@
-create_dataset <- function(id = "test", seed = 10) {
+create_dataset <- function(id = "test", seed = 10, backbone = "linear_simple") {
   requireNamespace("dyngen")
 
   assertthat::assert_that(is.function(dataset_file))
@@ -8,8 +8,8 @@ create_dataset <- function(id = "test", seed = 10) {
   reread(file_dataset, function() {
     set.seed(seed)
     wsr_multiplier <- 5
-    backbone <- backbone_linear_simple()
-    backbone$expression_patterns$time <- backbone$expression_patterns$time * wsr_multiplier
+    backbone <- dyngen::list_backbones()[[backbone]]()
+    # backbone$expression_patterns$time <- backbone$expression_patterns$time * wsr_multiplier
     model <-
       initialise_model(
         num_tfs = 20,
@@ -20,18 +20,18 @@ create_dataset <- function(id = "test", seed = 10) {
         num_cells = 1000,
         download_cache_dir = "~/.cache/dyngen",
         kinetics_params = kinetics_default(
-          sample_wsr = function(n) rep(1, n)#rnorm_bounded(n, 2, 1, min = 1)
+          # sample_wsr = function(n) rep(1, n)#rnorm_bounded(n, 2, 1, min = 1)
         ),
         gold_standard_params = gold_standard_default(
-          tau = .001 * wsr_multiplier, census_interval = 0.01 * wsr_multiplier
+          # tau = .001 * wsr_multiplier, census_interval = 0.01 * wsr_multiplier
         ),
         simulation_params = simulation_default(
-          ssa_algorithm = GillespieSSA2::ssa_etl(tau = 0.001 * wsr_multiplier),
-          census_interval = .01 * wsr_multiplier,
+          # ssa_algorithm = GillespieSSA2::ssa_etl(tau = 0.001 * wsr_multiplier),
+          # census_interval = .01 * wsr_multiplier,
           burn_time = simtime_from_backbone(backbone, burn = TRUE),
           total_time = simtime_from_backbone(backbone),
           perform_dimred = F,
-          experiment_params = simulation_type_wild_type(num_simulations = 10),
+          # experiment_params = simulation_type_wild_type(num_simulations = 10),
           store_reaction_propensities = TRUE
         )
       )
@@ -64,43 +64,6 @@ load_model <- function(id) {
 
 
 
-
-extract_groundtruth_velocity <- function(model) {
[…]

### 2019-12-09 analysis/velocity/2_velocity.R

@@ -2,48 +2,22 @@ library(tidyverse)
 
 source("analysis/velocity/2_velocity_functions.R")
 
+# load either velocyto or scvelo
+source("analysis/velocity/2_velocity_velocyto.R")
 devtools::load_all('~/thesis/projects/dynverse/libraries/scvelo/')
 
-pmap(design_velocity, run_velocity)
+# run either velocyto or scvelo
+pmap(design_velocity %>% filter(method_id == "velocyto"), run_velocity)
+# pmap(design_velocity %>% filter(method_id == "scvelo"), run_velocity)
 
-list2env(dynutils::extract_row_to_list(design_velocity, 1), .GlobalEnv)
+list2env(dynutils::extract_row_to_list(design_velocity %>% filter(method_id == "velocyto"), 1), .GlobalEnv)
 
 dataset <- load_dataset(dataset_id)
 model <- load_model(dataset_id)
-velociy <- load_velocity(dataset_id, method_id, params_id)
 
-# Plot the spliced vs unspliced changes
-# feature_id <- model$feature_info$feature_id[[2]]
-feature_id <- "B3_TF1"
-meta <- model$simulations$meta %>%
-  mutate(step_ix = row_number()) %>%
-  # filter(simulation_i == 1)
-  identity()
-counts <- model$simulations$counts[meta$step_ix, ]
-plotdata <- bind_rows(
-  tibble(
-    expression =  counts[,paste0("w_", feature_id)],
-    step_ix = meta$step_ix,
-    simulation_i = meta$simulation_i,
-    sim_time = meta$sim_time,
-    molecule = "unspliced"
-  ),
-  tibble(
-    expression =  counts[,paste0("x_", feature_id)],
-    step_ix = meta$step_ix,
-    simulation_i = meta$simulation_i,
-    sim_time = meta$sim_time,
-    molecule = "spliced"
-  )
-)
-plotdata %>%
-  pivot_wider(names_from = "molecule", values_from = "expression") %>%
-  ggplot(aes(spliced, unspliced)) + geom_path(aes(color = sim_time))
+run_velocity(dataset_id, method_id, params, params_id)
+velocity <- load_velocity(dataset_id, method_id, params_id)
 
 
-
-plotdata %>%
-  filter(simulation_i == 1) %>%
-  ggplot(aes(step_ix, expression)) + geom_line(aes(color = molecule))
 
 

### 2019-12-09 analysis/velocity/2_velocity_functions.R

@@ -13,8 +13,8 @@ run_velocity <- function(
     function() {
       if(method_id == "scvelo") {
         velocity <- scvelo::get_velocity(dataset$expression, dataset$expression_unspliced, mode = params$mode %||% "deterministic")
-      } else if (params$method == "velocyto") {
-        stop()
+      } else if (method_id == "velocyto") {
+        velocity <- get_velocity_velocyto(dataset$expression, dataset$expression_unspliced, assumption = params$assumption %||% "constant_velocity")
       }
 
       write_rds(velocity, experiment_file("velocity.rds"))

### 2019-12-09 analysis/velocity/3_evaluate.R

@@ -4,6 +4,7 @@ source("analysis/velocity/3_evaluate_functions.R")
 
 list2env(design_velocity[1, ], .GlobalEnv)
 
+# Calculate scores -----
 scores <- design_velocity %>% pmap(function(method_id, params_id, dataset_id, ...) {
   model <- load_model(dataset_id)
   dataset <- load_dataset_velocity(dataset_id, method_id, params_id)
@@ -13,13 +14,13 @@ scores <- design_velocity %>% pmap(function(method_id, params_id, dataset_id, ..
   velocity_differences[velocity_differences == 0] <- NA
   velocity_differences[is.na(velocity_differences)] <- runif(sum(is.na(velocity_differences)), -1e-10, 1e-10)
 
-  scores <- pairwise_correlations(
+  pairwise_correlations <- pairwise_correlations(
     Matrix::t(velocity_differences),
     t(groundtruth_velocity)
   )
   scores <- tibble(
     feature_id = colnames(velocity_differences),
-    score = scores
+    score = pairwise_correlations
   )
 
   # scores <- pairwise_correlations(
@@ -32,33 +33,105 @@ scores <- design_velocity %>% pmap(function(method_id, params_id, dataset_id, ..
   bind_cols(design_velocity)
 
 
-scores %>%
+plot_scores <- scores %>%
   select(-params) %>%
   unnest(c(scores)) %>%
-  # left_join(model$feature_info) %>%
-  ggplot(aes(paste0(method_id, " ", params_id), score)) +
-    ggbeeswarm::geom_beeswarm(aes(color = dataset_id)) +
-    facet_wrap(~dataset_id)
+  group_by(dataset_id, method_id, params_id) %>%
+  summarise(score = mean(score)) %>%
+  left_join(dataset_design, c("dataset_id" = "id")) %>%
+  ggplot(aes(paste0(method_id, "\n", params_id), score)) +
+    ggbeeswarm::geom_quasirandom(aes(color = backbone))  +
+    theme_bw() +
+    scale_x_discrete("") +
+    scale_y_continuous("Correlation (higher is better)") +
+  theme_common()
+
+plot_scores
 
 
 
 
+dataset_id <- "bifurcating_2"
+
+transform_groundtruth_velocity <- function(x) {
+  log2(scales::squish(x, c(0.1, 10), only.finite = FALSE))
+}
+
+devtools::load_all('~/thesis/projects/dynverse/dynplot2/')
+plot_velocity_gs <- function(
[…]

### 2019-12-09 analysis/velocity/3_evaluate_functions.R

@@ -35,3 +35,42 @@ run_metric <- function(
 metric_file <- function(dataset_id, method_id, params_id, metric_id) {
   dynamic_file(derived_file("metrics", dataset_id, method_id, params_id, metric_id))
 }
+
+
+
+extract_groundtruth_velocity <- function(model) {
+  if(is.null(model$simulations$reaction_propensities)) {
+    stop("model has to contain reaction_propensities for estimation of velocity ground truth")
+  }
+
+  # extract for each spliced mRNA its relevant reactions
+  feature_ids <- model$feature_info$feature_id
+  relevant_molecules <- paste0("x_", feature_ids)
+
+  relevant_reactions <- map2_dfr(seq_along(model$simulation_system$reactions), model$simulation_system$reactions, function(reaction_ix, reaction) {
+    if(names(reaction$effect)[1] %in% relevant_molecules) {
+      tibble(reaction_ix = reaction_ix, feature_id = names(reaction$effect)[1], effect = reaction$effect[1], name = reaction$name)
+    }
+  })
+
+  propensities <- model$simulations$reaction_propensities[model$experiment$cell_info$step_ix, relevant_reactions$reaction_ix]
+  rownames(propensities) <- model$experiment$cell_info$cell_id
+  colnames(propensities) <- relevant_reactions$name
+
+  formulas <- relevant_reactions %>%
+    group_by(feature_id) %>%
+    mutate(effect = ifelse(effect == 1, "production", "degradation")) %>%
+    pivot_wider("feature_id", names_from = "effect", values_from = "name")
+
+  propensity_ratios <- formulas %>%
+    mutate(data = map2(production, degradation, function(production, degradation) {
+      propensities[,production] / propensities[,degradation]
+    })
+    ) %>%
+    pull(data) %>%
+    do.call(rbind, .)
+  propensity_ratios[is.na(propensity_ratios)] <- runif(sum(is.na(propensity_ratios)), 1-1e-10, 1+1e-10)
+
+  rownames(propensity_ratios) <- str_replace_all(formulas$feature_id, "x_", "")
+  t(propensity_ratios)
+}

### 2019-12-09 manuscript.Rmd

@@
 
-* R.C. and W.S. designed the study.
+* W.S. and R.C. designed the study.
 * R.C., W.S., and L.D. performed the experiments and analysed the data.
 * R.C. and W.S. implemented the dyngen software package.
-* R.C. wrote the original manuscript.
+* R.C. and W.S. wrote the original manuscript.
+* L.D. wrote the section on trajectory alignment.
 * R.C., W.S., L.D., and Y.S. reviewed and edited the manuscript.
@@
 splatter[@zappia_splattersimulationsinglecell_2017], powsimR[@vieth_powsimrpoweranalysis_2017], PROSSTT[@papadopoulos_prossttprobabilisticsimulation_2018] and SymSim[@zhang_simulatingmultiplefaceted_2019]) have already been widely used to explore the strengths and weaknesses of computational tools, both by method developers[@street_slingshotcelllineage_2018; @parra_reconstructingcomplexlineage_2018; @lummertzdarocha_reconstructioncomplexsinglecell_2018; @lin_scclassifyhierarchicalclassification_2019] and independent benchmarkers[@duo_systematicperformanceevaluation_2018; @saelens_comparisonsinglecelltrajectory_2019; @soneson_biasrobustnessscalability_2018].
-However, a limitation of scRNA-seq profiles generators is that they would require significant methodological alterations to add additional modalities.
+However, a limitation of scRNA-seq profiles generators is that they would require significant methodological alterations to add additional modalities or experimental conditions.
 
@@
 
-We introduce dyngen, a multi-modality simulator of single cells and their dynamics (Figure\ \ref{fig:showcase}A).
-dyngen uses Gillespie's Stochastic Simulation Algorithm (SSA)[@gillespie_exactstochasticsimulation_1977] to simulate cells at a single-molecule and single-reaction levels.
-This means at every time step in the simulation, a single reaction is executed, thereby modifying the abundance of the molecule counts within a cell.
-
-By simulating a cell over time in terms of its molecular state and the reactions that are allowed to occur, the simulator is more extendible to new modalities or experimental procedures (Figure\ \ref{fig:showcase}B).
-We demonstrate dyngen's flexibility by simulating snapshot, time-series and perturbation experiments, allowing to benchmark a wide variety of computational tools such as trajectory alignment and differential network inference methods.
+We developed a multi-modality simulator of single cells called dyngen (Figure\ \ref{fig:showcase}A).
+dyngen uses Gillespie's stochastic simulation algorithm[@gillespie_exactstochasticsimulation_1977]  to simulate gene regulatory interactions at a single-molecule level.
+Its methodology allows tracking of many layers of information throughout the simulation, including the abundance of any molecule in the cell, progression of the cell along a dynamic process, and the activation strength of individual regulatory interactions.
+dyngen can simulate a large variety of dynamic processes (e.g.
+cyclic, branching, disconnected) as well as a broad range of experimental conditions (e.g.
+batch effects and time-series, perturbation and knockdown experiments).
+The fine-grained controls over simulation parameters allow dyngen to be applicable to a broad range of use-cases.
+We demonstrate this by performing first quantitative evaluations of three types of novel computational approaches: RNA velocity, casewise network inference and trajectory alignment methods.
 
@@
 # Results {#sec:dyngen-results}
-dyngen offers multiple significant qualities, each of which we will discuss below.
-it's methodology allows keeping track of many modalities of the cell at a high resolution level
-simulate different various types of dynamic processes, such as cyclic, branching, disconnected
-perform time-series, perturbation, or knockdown experiments.
 
@@
 
-Using a collection of transcriptomic profiles, network inference (NI) methods predict which genes are the regulators of a target gene.
-The output of an NI method is called a gene regulatory network (GRN, Figure\ \ref{fig:scgrn}A, top left), in which each predicted interaction can be assigned a predicted strength or effect (activating or inhibitory).
-
-However, cells are very molecularly homogeneous, and only a fraction of all possible gene regulatory interactions is active at any point in time.
 Casewise network inference (CNI) methods\footnote{Other terms are commonly used when dealing with data from a particular source.
-For example, single-cell NI when applied to single-cell transcriptomics data; sample-specific NI when applied to bulk transcriptomics; patient-derived NI when applied to bulk profiles of patients.} tackle a more complex variant of NI, namely "Which regulators are regulating which targets in which case?" The end result of such methods is a single GRN for every case, be it a cell, tissue, patient or other sample.
+For example, single-cell NI when applied to single-cell transcriptomics data; sample-specific NI when applied to bulk transcriptomics; patient-derived NI when applied to bulk profiles of patients.} predict not only which transcription factors regulate which target genes (Figure\ \ref{fig:scgrn}A, top left), but also how active each interaction is in every case (Figure\ \ref{fig:scgrn}A).
+In CNI, a 'case' might be a cell, but it might also refer to a bulk sample.
 
@@
 
-dyngen allows deriving the ground-truth regulatory effect of a regulator on a target in a specific cell state.
-This information is derived by knocking down the expression of a regulator and observing the change in transcription rate as a result.
-Performing in silico knockdowns for every regulator in every cell results in the ground-truth casewise GRN (Figure\ \ref{fig:scgrn}A).
+Since dyngen computes tracks the probability of transcription, temporarily 'knocking down' the expression of a regulator and observing the change in transcription probability.
+This is a much more accurate ground-truth for regulatory interaction between a regulator and target in comparison to observing the change in transcript abundance levels when knocking down a regulator, as the regulation of the target could be indirect.
 
@@
     \textbf{dyngen allows benchmarking Casewise Network Inference (CNI) methods.} 
-    \textbf{A:} A cell is simulated using the global GRN (top left). However, at any particular state in the simulation, only a fraction of the gene regulatory interactions are active.
+    \textbf{A:} A cell is simulated using the global gene regulatory network (GRN, top left).
+However, at any particular state in the simulation, only a fraction of the gene regulatory interactions are active.
     \textbf{B:} CNI methods were executed to predict the regulatory interactions that are active in each cell specifically.
@@
 
-In eukaryotes, a gene is first transcribed to a pre-mRNA, and subsequently spliced into a mature mRNA.
-Both unspliced reads, which map to introns, and spliced reads, mapping to exons are often observed in RNA sequencing experiments, both in bulk[@zeisel_coupledpremrnamrna_2011] and single-cell.
+In eukaryotes, a gene is first transcribed to a pre-mRNA, and subsequently spliced into mature mRNA.
+Because reads coming from both unspliced and spliced transcripts are observed in expression data, the relative ratio between the two can tell us something about which genes are increasing, decreasing or remaining the same[@zeisel_coupledpremrnamrna_2011;@lamanno_rnavelocitysingle_2018].
+To determine this, some parameters have to be estimated to determine which fraction of unspliced and spliced mRNAs corresponds to an increase or decrease.
+The estimation of these parameters makes some assumptions, and can be handled in different ways in the two main algorithms that are now available for RNA velocity estimation: velocyto[@lamanno_rnavelocitysingle_2018] and scvelo[@bergen_generalizingrnavelocity_2019].
+It can be difficult to obtain ground truth data to benchmark these algorithms, given that it would require continuous data of transcriptional dynamics in individual cells.
+On the other hand, the ground truth velocity is rapidly extracted from the dyngen model, by looking whether each transcript is currently increasing or decreasing in expression.
+
+We tested scvelo and velocyto on 8 datasets containing linear, bifurcating, disconnected and cyclic trajectories, and varied the main parameter settings in which they estimate the velocity.
+We found that the original velocyto implementation, which assumes that the velocity remains constant in some cells, performed the best across all datasets.
+The dynamical estimation of velocyto, as implemented in scvelo, performed the worst of all parameter settings.
+This was mainly due to scvelo overestimating the dynamics of a gene, especially towards upregulation, while velocyto correctly estimated not only when a gene changes, but also when it remained in a steady state.
+
+\begin{figure}[htb!]
+    \centering
+    \includegraphics[width=.8\linewidth]{results/velocity/usecase.pdf}
+    \caption{
+    \textbf{dyngen allows benchmarking of RNA velocity methods.} 
+    \textbf{A:} We tested five different methods and parameter settings for the estimation of RNA velocity on datasets with varying backbones (colors).
+Overall, the velocyto method with the constant velocity assumption performed the best overall.
+    \textbf{B:} An example bifurcating dataset, with as illustration the expression and ground truth velocity of a gene that goes up and down in the trajectory. 
+\textbf{C:} The RNA velocity estimates of the different methods.
+\textbf{D:} The embedded RNA velocity of the different methods.
+    }
+    \label{fig:velocity}
+\end{figure}
 
@@
 For each cell-specific network, the AUROC and AUPR were calculated.
+## Comparison of RNA velocity methods {#sec:dyngen-velcompare}
+
+15 datasets were generated with 5 different backbones: linear, linear simple, bifurcating, cyclic, and disconnected.
+We extracted a ground truth RNA velocity by subtracting for each mRNA molecule the propensity of its production by the propensity of its degradation.
+If the expression of an mRNA will increase in the future, this value is positive, while it is negative if it is going to decrease.
+For each gene, we compared the ground truth velocity with the observed velocity by calculating the Spearman rank correlation.
+
+We compared two RNA velocity methods.
+The velocyto method[@lamanno_rnavelocitysingle_2018], as implemented in the velocyto.py package, in which we varied the "assumption" parameter between "constant_unspliced" and "constant_velocity".
+The scvelo method[@bergen_generalizingrnavelocity_2019], as implemented in the python scvelo package (http://scvelo.de), in which we varied the "mode" parameter between "deterministic", "stochastic" and "dynamical".
+For both methods, we used the same normalized data as provided by dyngen, with no extra cell or feature filtering.
+We also matched the parameters between both methods as best as possible, i.e.
+the k parameter for smoothing was set to 20 for both methods.
+
+To visualize the velocity on an embedding, we used the "velocity_embedding" function, implemented in the scvelo python package.

### 2019-12-09 analysis/usecase_velocity.R
update old usecase velocity

@@ -1,0 +1,300 @@
+library(tidyverse)
+devtools::load_all('~/projects/dynverse/dynwrap/')
+library(dyngen)
+devtools::load_all('~/projects/dynverse/dynplot2/')
+devtools::load_all('~/projects/dynverse/libraries/scvelo/')
+
+set.seed(10)
+wsr_multiplier <- 1
+backbone <- backbone_linear_simple()
+backbone$expression_patterns$time <- backbone$expression_patterns$time * wsr_multiplier
+model <-
+  initialise_model(
+    num_tfs = 12,
+    num_targets = 30,
+    num_hks = 15,
+    backbone = backbone,
+    verbose = TRUE,
+    num_cells = 1000,
+    download_cache_dir = "~/.cache/dyngen",
+    # kinetics_params = kinetics_default(sample_wsr = function(n) rnorm(n, 2, 1) %>% pmax(0.5)),
+    gold_standard_params = gold_standard_default(
+      # tau = .001 * wsr_multiplier, census_interval = 0.01 * wsr_multiplier
+    ),
+    simulation_params = simulation_default(
+      # ssa_algorithm = GillespieSSA2::ssa_etl(tau = 0.001 * wsr_multiplier),
+      census_interval = .01 * wsr_multiplier,
+      burn_time = simtime_from_backbone(backbone, burn = TRUE),
+      total_time = simtime_from_backbone(backbone),
+      perform_dimred = F,
+      experiment_params = simulation_type_wild_type(num_simulations = 1)
+    )
+  )
+model <- generate_tf_network(model)
+model <- generate_feature_network(model)
+model <- generate_kinetics(model)
+model <- generate_gold_standard(model)
+model <- generate_cells(model)
+model <- generate_experiment(model)
+dataset <- wrap_dataset(model)
+
+
+
+# plot_gold_simulations(model)
+# plot_simulations(model)
+
+plot_backbone_modulenet(model)
+
+dataset <- dataset %>% add_velocity(mode = "dynamical")
+
+# Plot the spliced vs unspliced changes
+# feature_id <- model$feature_info$feature_id[[2]]
+feature_id <- names(sort(apply(dataset$expression, 2, sd), decreasing = T)) %>% .[11]
+meta <- model$simulations$meta %>%
+  mutate(step_ix = row_number()) %>%
+  filter(simulation_i == 1) %>%
+  identity()
+counts <- model$simulations$counts[meta$step_ix, ]
+plotdata <- bind_rows(
+  tibble(
[…]

### 2019-12-09 analysis/usecase_velocity.R
:wave: use case velocity

@@ -1,300 +1,0 @@
-library(tidyverse)
-devtools::load_all('~/projects/dynverse/dynwrap/')
-library(dyngen)
-devtools::load_all('~/projects/dynverse/dynplot2/')
-devtools::load_all('~/projects/dynverse/libraries/scvelo/')
-
-set.seed(10)
-wsr_multiplier <- 1
-backbone <- backbone_linear_simple()
-backbone$expression_patterns$time <- backbone$expression_patterns$time * wsr_multiplier
-model <-
-  initialise_model(
-    num_tfs = 12,
-    num_targets = 30,
-    num_hks = 15,
-    backbone = backbone,
-    verbose = TRUE,
-    num_cells = 1000,
-    download_cache_dir = "~/.cache/dyngen",
-    # kinetics_params = kinetics_default(sample_wsr = function(n) rnorm(n, 2, 1) %>% pmax(0.5)),
-    gold_standard_params = gold_standard_default(
-      # tau = .001 * wsr_multiplier, census_interval = 0.01 * wsr_multiplier
-    ),
-    simulation_params = simulation_default(
-      # ssa_algorithm = GillespieSSA2::ssa_etl(tau = 0.001 * wsr_multiplier),
-      census_interval = .01 * wsr_multiplier,
-      burn_time = simtime_from_backbone(backbone, burn = TRUE),
-      total_time = simtime_from_backbone(backbone),
-      perform_dimred = F,
-      experiment_params = simulation_type_wild_type(num_simulations = 1)
-    )
-  )
-model <- generate_tf_network(model)
-model <- generate_feature_network(model)
-model <- generate_kinetics(model)
-model <- generate_gold_standard(model)
-model <- generate_cells(model)
-model <- generate_experiment(model)
-dataset <- wrap_dataset(model)
-
-
-
-# plot_gold_simulations(model)
-# plot_simulations(model)
-
-plot_backbone_modulenet(model)
-
-dataset <- dataset %>% add_velocity(mode = "dynamical")
-
-# Plot the spliced vs unspliced changes
-# feature_id <- model$feature_info$feature_id[[2]]
-feature_id <- names(sort(apply(dataset$expression, 2, sd), decreasing = T)) %>% .[11]
-meta <- model$simulations$meta %>%
-  mutate(step_ix = row_number()) %>%
-  filter(simulation_i == 1) %>%
-  identity()
-counts <- model$simulations$counts[meta$step_ix, ]
-plotdata <- bind_rows(
-  tibble(
[…]

### 2019-12-09 analysis/velocity/3_evaluate.R
update evlauate

@@ -84,10 +84,12 @@ plot_velocity_gs <- function(
     )
   }
 
+  labels <- setNames(LETTERS[seq_along(dataset$milestone_ids)], dataset$milestone_ids)
   plot_trajectory <- dynplot_dimred(dataset) +
     geom_cell_point(aes(color = milestone_percentages)) +
     scale_milestones_fillcolour() +
     geom_trajectory_segments(size = 2, color = "#333333") +
+    geom_milestone_label(aes(fill = milestone_id, label = labels[label]), color = "black") +
     ggtitle(method_id) +
     theme_common(legend.position = "none") +
     ggtitle("Trajectory")
@@ -216,6 +218,10 @@ tag_first <- function(x, tag) {
   x
 }
 
+plot_scores <- plot_scores + labs(tag = "A")
+plot_velocity_gs[[1]][[1]] <- plot_velocity_gs[[1]][[1]] + labs(tag = "B")
+plot_velocity_gs[[2]][[1]] <- plot_velocity_gs[[2]][[1]] + labs(tag = "C")
+plot_velocity_arrows[[1]] <- plot_velocity_arrows[[1]] + labs(tag = "D")
 
 pdf("results/velocity/usecase.pdf", height = 14, width = 10, useDingbats = F)
 patchwork::wrap_plots(
@@ -224,7 +230,7 @@ patchwork::wrap_plots(
   (plot_velocity_arrows & theme(plot.title = element_blank(), plot.subtitle = element_blank())),
   ncol = 1,
   heights = c(1, 2, 1)
-) + patchwork::plot_annotation(tag_levels = c("A"))
+)
 graphics.off()
 
 magick::image_read_pdf("results/velocity/usecase.pdf") %>%
@@ -232,131 +238,3 @@ magick::image_read_pdf("results/velocity/usecase.pdf") %>%
   magick::image_write("results/velocity/usecase.png") %>%
   invisible()
 
-
-
-
-
-
-
-
-
-
-compare_velocity_correlation <- function(model, dataset, velocity, ...) {
-  model <- load_model(dataset_id)
-  dataset <- load_dataset_velocity(dataset_id, method_id, params_id)
-  groundtruth_velocity <- extract_groundtruth_velocity(model)
-
-  velocity_differences <- dataset$expression_projected - dataset$expression
-  velocity_differences[velocity_differences == 0] <- NA
[…]

### 2019-12-09 manuscript.Rmd
update

@@
   \usepackage{fontspec}
+  \usepackage{pifont}
   \setmainfont [Path = fonts/,
@@
 splatter[@zappia_splattersimulationsinglecell_2017], powsimR[@vieth_powsimrpoweranalysis_2017], PROSSTT[@papadopoulos_prossttprobabilisticsimulation_2018] and SymSim[@zhang_simulatingmultiplefaceted_2019]) have already been widely used to explore the strengths and weaknesses of computational tools, both by method developers[@street_slingshotcelllineage_2018; @parra_reconstructingcomplexlineage_2018; @lummertzdarocha_reconstructioncomplexsinglecell_2018; @lin_scclassifyhierarchicalclassification_2019] and independent benchmarkers[@duo_systematicperformanceevaluation_2018; @saelens_comparisonsinglecelltrajectory_2019; @soneson_biasrobustnessscalability_2018].
-However, a limitation of scRNA-seq profiles generators is that they would require significant methodological alterations to add additional modalities or experimental conditions.
+However, a limitation of scRNA-seq profiles generators is that they would require significant methodological alterations to add additional modalities or experimental conditions, thereby limiting the applicability of each generator.
+
+```{r generators, echo=FALSE, results='asis', warning=FALSE, message=FALSE}
+modalities <- tribble(
+  ~id, ~label,
+  "premrna", "pre-mRNA expression",
+  "mrna", "mRNA expression",
+  "prot", "Protein expression",
+  "promotor", "Promotor activity",
+  "lineage", "Position in lineage tree",
+  "reaction", "Reaction frequency"
+) %>% mutate(type = "Modality")
+applications <- tribble(
+  ~id, ~label,
+  "cluster", "Clustering",
+  "diffexp", "Differential expression",
+  "batch", "Batch effect correction",
+  "linti", "Linear trajectory inference",
+  "brati", "Branching trajectory inference",
+  "disti", "Disconnected trajectory inference",
+  "traja", "Trajectory alignment",
+  "cni", "Casewise network inference",
+  "rnav", "RNA velocity"
+) %>% mutate(type = "Applications")
+labels <- bind_rows(modalities, applications)
+
+methods <- tribble(
+  ~package, ~citation, ~modality, ~application, ~model,
+  "splatter", "zappia_splattersimulationsinglecell_2017",
+  c("mrna"), c("cluster", "linti", "batch", "diffexp"),
+  "Several distributional models with parameters that allow the generation of datasets by sampling from distributions
+Cell types, linear trajectories and batch effects can be simulated by changing the parameters of certain genes",
+
+  "powsimR", "vieth_powsimrpoweranalysis_2017",
+  c("mrna"), c("diffexp", "cluster"),
+  "Negative binomial or zero-inflated negative binomial. Cell types can be simulated by changing the parameters of this distribution.",
+
+  "PROSSTT", "papadopoulos_prossttprobabilisticsimulation_2018",
+  c("mrna"), c("linti", "brati"),
+  "(1) Generation of a random tree.
+(2) The expression of each gene is seen as a sum of contributions from different expression programs.
+The activity of each program depends on pseudotime.
+(3) Generation of cells and expression by random sampling along the trajectory, and from a negative binomial distribution.",
+
+  "SymSim", "zhang_simulatingmultiplefaceted_2019",
+  c("mrna", "promotor"), c("cluster", "linti", "brati", "disti"),
+  "Kinetic model of each gene, in which it can be on or off, and where the mRNA can be degraded.
+The parameters of this model depend on a gene effect and cell effect, each of these effects represents a vector in an n-dimensional space (e.g.
+10).
+The cell effect vectors are sampled in such a way that cells that have to be more similar (according to a tree provided by the user), will also have a more similar vector.",
+
+  "dyngen", "",
+  c("premrna", "mrna", "prot", "promotor", "lineage", "reaction"),
+  c("cluster", "linti", "brati", "disti", "batch", "diffexp", "cni", "rnav", "traja"),
+  "Simulation of a thermodynamic model of gene regulation, in which a set of transcription factors can bind the promoter and can have varying effects on the activity of the promoter.
+Clusters or trajectories are simulated by specifying a regulatory network that will induce a certain diversification of expression.
+Expression is simulated using a single-molecule model, after which individual UMIs are simulated using a multinomial distribution.
+Apart from gene regulation, splicing and protein production are also simulated."
+) %>%
+  mutate(
+citation = ifelse(citation == "", "", paste0("\\cite{", citation, "}")),
+plabel = paste0(package, " ", citation)
+  )
+
+tab <-
+  full_join(
+labels %>% crossing(methods) %>% select(-modality, -application),
+methods %>% rowwise() %>% mutate(id = list(c(modality, application))) %>% ungroup() %>% unnest(id) %>% transmute(package, id, check = "\\ding{51}"),
+by = c("id", "package")
+  ) %>%
+  mutate(
+check = ifelse(is.na(check), "", check),
+package = factor(package, levels = methods$package),
+plabel = factor(plabel, levels = methods$plabel),
+id = factor(id, labels$id)
+  ) %>%
+  arrange(match(package, methods$package), type, match(id, labels$id))
+
+library(kableExtra)
+tab %>%
+  select(-id, -type, -model, -citation, -package) %>%
+  spread(plabel, check) %>%
+  knitr::kable(escape = FALSE) %>%
+  kableExtra::row_spec(0, angle = 0)
+```
 
@@
 We developed a multi-modality simulator of single cells called dyngen (Figure\ \ref{fig:showcase}A).
-dyngen uses Gillespie's stochastic simulation algorithm[@gillespie_exactstochasticsimulation_1977]  to simulate gene regulatory interactions at a single-molecule level.
+dyngen uses Gillespie's stochastic simulation algorithm[@gillespie_exactstochasticsimulation_1977]  to simulate gene regulation, splicing and translation at a single-molecule level.
 Its methodology allows tracking of many layers of information throughout the simulation, including the abundance of any molecule in the cell, progression of the cell along a dynamic process, and the activation strength of individual regulatory interactions.
@@
 Together, these qualities allow it to be applicable in benchmarking a broad range of use-cases.
-To demonstrate, we apply dyngen on several novel computational approaches which have recently had a major impact on how single-cell analyses are performed but for which quantitative assessment of the performance was hitherto lacking.
+In practice, dyngen has already successfully been used to evaluate trajectory inference[@saelens_comparisonsinglecelltrajectory_2019], trajectory-based differential expression[@vandenberge_trajectorybaseddifferentialexpression_2019], and network inference[@pratapa_benchmarkingalgorithmsgene_2019] methods.
+To demonstrate this point even further, we apply dyngen on several promising novel computational approaches which have recently had a major impact on how single-cell analyses are performed but for which quantitative assessment of the performance was hitherto lacking.
 
-## Example use cases {#sec:dyngen-example}
+## Trajectory alignment
 
-### Trajectory alignment
+Trajectory alignment is a technique used to discover the similarities and differences between the developmental trajectories of cells. 
+This technique has been used to compare gene expression kinetics resulting from different biological processes[@cacchiarelli_aligningsinglecelldevelopmental_2018], to compare human, chimpanzee and macaque neuronal development[@kanton_organoidsinglecellgenomic_2019], to find differences in gene regulation in the presence of certain growth factors[@mcfaline-figueroa_pooledsinglecellgenetic_2019], and to compare human and mouse embryogenesis[@alpert_alignmentsinglecelltrajectories_2018].
 
-### Casewise network inference
+Unfortunately, not many methods to align trajectories exist.
+DTW (Dynamic Time Warping)[@giorgino_computingvisualizingdynamic_2009] is most commonly used to align linear trajectories.
+DTW is a technique originating in the field of speech recognition and aligns temporal sequences by creating a warping path between two sequences that indicates which sequence must be dilated or contracted to best match the other one.
+
+dyngen can be used to evaluate trajectory alignment techniques.
+We generated two datasets and permuted each of those datasets in five ways: we knocked down the transcription factors belonging to the end of the second milestone, we set 10% or 20% of the count matrix to zero, and we performed the combination of those techniques.
+This results in 12 datasets, each containing a slightly different trajectory.
+We compared DTW when using smoothed pseudocells (with a gaussian kernel) and using just all cells.
+For each combination of trajectories, we calculated the DTW alignment.
+In figure x, the ratio between one trajectory with a certain permutation, and the other trajectory with the same permutation is shown.
+We expect this ratio to be low, as we want the alignment to be able to distinguish between different trajectories.
+We especially expect the smoothed versions to be able to overcome the extra 10% or 20% noise.
+We performed this experiment 10 times.
+
+In Figure \ref{fig:traj_align}, we can see that DTW with gaussian smoothing cannot distinguish between trajectory 1 and trajectory 2.
+When using the cells themselves, we find that they can clearly distinguish between trajectory 1 and 2 when noise is added, but not when the trajectory is limited to milestone C.
+
+<!--
+\begin{figure}[htb!]
+    \centering
+    \includegraphics[width=.8\linewidth]{fig/traj_alignment/tares.pdf}
+    \caption{
+    \textbf{dyngen allows benchmarking of trajectory alignment methods.} 
+    \label{fig:traj_align}
+\end{figure}
+-->
+
+## Casewise network inference
 
@@
 
-### RNA velocity
+## RNA velocity
 

### 2019-12-19 manuscript.Rmd
update manuscript

@@
 
-**Purpose:** A common problem of pioneering computational tools is that during their development, there are rarely sufficient datasets available for adequately quantitatively assessing its performance.
+**Purpose:** When developing new types of tools for single-cell analyses, there is often a lack of datasets on which to quantitatively assess the performance.A common problem of pioneering computational tools is that during their development, there are rarely sufficient datasets available for adequately quantitatively assessing its performance.
 **Results:** We developed dyngen, a multi-modality simulator of single cells.
-In dyngen, the biomolecular state of an _in silico_ cell changes over time according to a set of reactions defined by the cell's gene regulatory network.
-By simulating single cells in terms of its biomolecular state and reactions, the simulator is easily extendible for adding new modalities or experimental procedures.
-We demonstrate dyngen's flexibility by simulating snapshot, time-series and perturbation experiments.
+In dyngen, the biomolecular state of an _in silico_ cell changes over time according to a predefined gene regulatory network.
+We used dyngen to benchmark three upcoming methods for single-cell analyses: RNA velocity, casewise network inference and trajectory alignment methods.
+The simulator is easily extendible for adding new modalities or experimental procedures.
+We demonstrate dyngen's flexibility by applying it to benchmark RNA velocity, casewise network inference and trajectory alignment methods.
 **Conclusion:** dyngen lays the foundations for benchmarking a wide variety of computational single-cell tools and can be used to help kick-start the development of future types of analyses.
@@
 splatter[@zappia_splattersimulationsinglecell_2017], powsimR[@vieth_powsimrpoweranalysis_2017], PROSSTT[@papadopoulos_prossttprobabilisticsimulation_2018] and SymSim[@zhang_simulatingmultiplefaceted_2019]) have already been widely used to explore the strengths and weaknesses of computational tools, both by method developers[@street_slingshotcelllineage_2018; @parra_reconstructingcomplexlineage_2018; @lummertzdarocha_reconstructioncomplexsinglecell_2018; @lin_scclassifyhierarchicalclassification_2019] and independent benchmarkers[@duo_systematicperformanceevaluation_2018; @saelens_comparisonsinglecelltrajectory_2019; @soneson_biasrobustnessscalability_2018].
-However, a limitation of scRNA-seq profiles generators is that they would require significant methodological alterations to add additional modalities or experimental conditions, thereby limiting the applicability of each generator.
+However, a limitation of existing scRNA-seq profiles generators is that they would require significant methodological alterations to add additional modalities or experimental conditions., thereby limiting the applicability of each generator.
 
@@
 ```
+method
+model
+modalities
+(possible) applications
+splatter[@zappia_splattersimulationsinglecell_2017]
+Several distributional models with parameters that allow the generation of datasets by sampling from distributions
+Cell types, linear trajectories and batch effects can be simulated by changing the parameters of certain genes
+mRNA expression
+Clustering
+Linear trajectory inference
+Batch effect correction
+Differential expression
+powsimR[@vieth_powsimrpoweranalysis_2017]
+Negative binomial or zero-inflated negative binomial.
+Cell types can be simulated by changing the parameters of this distribution
+mRNA expression
+Differential expression
+Clustering
+PROSSTT[@papadopoulos_prossttprobabilisticsimulation_2018]
+(1) Generation of a random tree
+(2) The expression of each gene is seen as a sum of contributions from different expression programs. The activity of each program depends on pseudotime.
+(3) Generation of cells and expression by random sampling along the trajectory, and from a negative binomial distribution
+mRNA expression
+Linear and branching trajectory inference
+SymSim[@zhang_simulatingmultiplefaceted_2019]
+Kinetic model of each gene, in which it can be on or off, and where the mRNA can be degraded.
+The parameters of this model depend on a gene effect and cell effect, each of these effects represents a vector in an n-dimensional space (e.g.
+10).
+The cell effect vectors are sampled in such a way that cells that have to be more similar (according to a tree provided by the user), will also have a more similar vector.
+mRNA expression
+promoter activity
+Clustering
+Linear, branching and disconnected trajectory inference
+
+dyngen
+Simulation of a thermodynamic model of gene regulation, in which a set of transcription factors can bind the promoter and can have varying effects on the activity of the promoter.
+Clusters or trajectories are simulated by specifying a regulatory network that will induce a certain diversification of expression.
+Expression is simulated using a single-molecule model, after which individual UMIs are simulated using a multinomial distribution.
+Apart from gene regulation, splicing and protein production are also simulated.
+Protein expression
+mRNA expression
+pre-mRNA expression
+promoter activity
+lineage state
+reaction activity
+Clustering
+Linear, branching and disconnected trajectory inference
+Cellwise network inference
+Batch effect correction
+RNA velocity
+Trajectory alignment
 
@@
 In practice, dyngen has already successfully been used to evaluate trajectory inference[@saelens_comparisonsinglecelltrajectory_2019], trajectory-based differential expression[@vandenberge_trajectorybaseddifferentialexpression_2019], and network inference[@pratapa_benchmarkingalgorithmsgene_2019] methods.
-To demonstrate this point even further, we apply dyngen on several promising novel computational approaches which have recently had a major impact on how single-cell analyses are performed but for which quantitative assessment of the performance was hitherto lacking.
+To demonstrate this point even further, we apply dyngen on several promising novel computational approaches for which quantitative assessment of the performance was hitherto lacking.
 
@@
 
-Trajectory alignment is a technique used to discover the similarities and differences between the developmental trajectories of cells. 
-This technique has been used to compare gene expression kinetics resulting from different biological processes[@cacchiarelli_aligningsinglecelldevelopmental_2018], to compare human, chimpanzee and macaque neuronal development[@kanton_organoidsinglecellgenomic_2019], to find differences in gene regulation in the presence of certain growth factors[@mcfaline-figueroa_pooledsinglecellgenetic_2019], and to compare human and mouse embryogenesis[@alpert_alignmentsinglecelltrajectories_2018].
+Trajectory alignment allows studying the differences between multiple trajectories that are mostly similar.
+For example, the cell developmental process of a patient could be compared to that of a healthy control to detect the transcriptomic differences of a particular lineage branch.
 
-Unfortunately, not many methods to align trajectories exist.
-DTW (Dynamic Time Warping)[@giorgino_computingvisualizingdynamic_2009] is most commonly used to align linear trajectories.
-DTW is a technique originating in the field of speech recognition and aligns temporal sequences by creating a warping path between two sequences that indicates which sequence must be dilated or contracted to best match the other one.
+However, aligning trajectories becomes exponentially more difficult as the complexity of compared trajectories increases.
+For this reason, trajectory alignment remains a mostly unexplored territory within single-cell omics.
+
+Dynamic Time Warping (DTW)[@giorgino_computingvisualizingdynamic_2009] is most commonly used to align linear trajectories.
+DTW is a technique originating in the field of speech recognition and aligns temporal sequences by creating a warping path between the sequences that indicates which sequence must be dilated or contracted to best match the other one.
+DTW can also be used in combination with smoothed pseudocells[@alpert_alignmentsinglecelltrajectories_2018].
+
+Trajectory alignment, using DTW, has been used to compare gene expression kinetics resulting from different biological processes[@cacchiarelli_aligningsinglecelldevelopmental_2018], to compare human, chimpanzee and macaque neuronal development[@kanton_organoidsinglecellgenomic_2019], to find differences in gene regulation in the presence of certain growth factors[@mcfaline-figueroa_pooledsinglecellgenetic_2019], and to compare human and mouse embryogenesis[@alpert_alignmentsinglecelltrajectories_2018].
 
 dyngen can be used to evaluate trajectory alignment techniques.
-We generated two datasets and permuted each of those datasets in five ways: we knocked down the transcription factors belonging to the end of the second milestone, we set 10% or 20% of the count matrix to zero, and we performed the combination of those techniques.
-This results in 12 datasets, each containing a slightly different trajectory.
-We compared DTW when using smoothed pseudocells (with a gaussian kernel) and using just all cells.
-For each combination of trajectories, we calculated the DTW alignment.
-In figure x, the ratio between one trajectory with a certain permutation, and the other trajectory with the same permutation is shown.
-We expect this ratio to be low, as we want the alignment to be able to distinguish between different trajectories.
-We especially expect the smoothed versions to be able to overcome the extra 10% or 20% noise.
-We performed this experiment 10 times.
+We look how well DTW can continue to distinguish between different trajectories when an increasing amount of noise is added.
+In Figure \ref{fig:traj_align} we can see that when noise increases, DTW can no longer distinguish trajectories clearly, and that in this case, there is no difference whether smoothed pseudocells are used or when actual cells and their gene expression are used.
 
-In Figure \ref{fig:traj_align}, we can see that DTW with gaussian smoothing cannot distinguish between trajectory 1 and trajectory 2.
-When using the cells themselves, we find that they can clearly distinguish between trajectory 1 and 2 when noise is added, but not when the trajectory is limited to milestone C.
-
-<!--
 \begin{figure}[htb!]
@@
     \textbf{dyngen allows benchmarking of trajectory alignment methods.} 
+  \textbf{A: Example of trajectory alignment using DTW. The segments between the two trajectories show which cells are aligned.}
+\textbf{B: Shows the discrimination power of DTW decreasing as the amount of noise added increases.}
     \label{fig:traj_align}
 \end{figure}
--->
 
@@
     \caption{
-      \textbf{Backbone lego.}
+      \textbf{Reusable regulatory building blocks that can be used to construct GRNs that result in dynamic processes following an expected pattern.}
+      \textbf{A:} Each building block contains genes which serve as inputs that can be regulated by external transcription factors.
+The connections inside the building block transform the input signal in a particular way and has one or more output interactions of which the targets are not yet determined.
+      \textbf{B:} By connecting multiple building blocks together, a backbone of regulatory modules can be formed.
+      \textbf{C:} dyngen offers several variations of linear building blocks which allow for differing regulatory mechanisms.
+      \textbf{D:} A branching module will upregulate the expression of just one of the output interactions.
+Two variations are given of which the simple variant is more prone to reprogramming events in comparison to the robust branching module.
+      \textbf{E,F:} These components allow to specify the beginning (no inputs) and end (no outputs) of the backbone.
     }
@@
 **Step 2, generating the TF interactions.** Let each TF be regulated according to the interactions in the backbone.
-These interactions inherit the effect, strength, and cooperativity parameters from the interactions in the backbone.
+These interactions inherit the effect, strength, and independence parameters from the interactions in the backbone.
 A TF can only be regulated by other TFs or itself.
@@
 \newcommand{\hill}[1]{\text{hill}_{#1}}
-\newcommand{\coop}[1]{\text{co}_{#1}}
+\newcommand{\ind}[1]{\text{ind}_{#1}}
 \newcommand{\dis}[1]{\text{dis}_{#1}}
 \newcommand{\buf}[1]{\chi_{#1}}
-\newcommand{\ba}[1]{\text{ba}_{#1}}
+\newcommand{\ba}[1]{\text{bas}_{#1}}
 
@@
         Reaction & Effect & Propensity \\ \hline \hline
-        Transcription & $\emptyset \rightarrow \w G$ & $\wpr G \times \frac{\ba G - \coop{G}^{|\rp{G}|} + \prod\limits_{H \in \rp{G}}(\coop G + \buf{G,H})}{\prod\limits_{H \in \rs{G}}(1 + \buf{G,H})}$ \\
+        Transcription & $\emptyset \rightarrow \w G$ & $\wpr G \times \frac{\ba G - \ind{G}^{|\rp{G}|} + \prod\limits_{H \in \rp{G}}(\ind G + \buf{G,H})}{\prod\limits_{H \in \rs{G}}(1 + \buf{G,H})}$ \\
         Pre-mRNA degradation & $\w G \rightarrow \emptyset$ & $\wdr G \times \w G$ \\
@@
         Hill coefficient & $\hill{G,H}$ & $\in U(0.5, 2)$ * \\
-        Cooperativity factor & $\coop G$ & $\in [0, 1]$ * \\ \hline\hline
+        Independence factor & $\ind G$ & $\in [0, 1]$ * \\ \hline\hline
         Pre-mRNA degradation rate & $\wdr G$ & $= \ln(2)\ /\ \whl G$ \\
@@
 
-The propensity of the transcription of a gene $G$ is inspired by thermodynamic models of gene regulation [needcitation], in which the promoter of $G$ can be bound or unbound by a set of $N$ transcription factors $H_i$.
+The propensity of the transcription of a gene $G$ is inspired by thermodynamic models of gene regulation[@schilstra_biologicgeneexpression_2008needcitation], in which the promoter of $G$ can be bound or unbound by a set of $N$ transcription factors $H_i$.
 Let $\proptran(\y 1, \y 2, \ldots, \y N)$ denote the propensity function of $G$, in function of the abundance levels of the transcription factors.
@@
 
-LIONESS[@@kuijjer_estimatingsamplespecificregulatory_2019; kuijjer_lionessrsinglesample_2019] uses the Pearson correlation to infer casewise GRNs.
+LIONESS[@@kuijjer_estimatingsamplespecificregulatory_2019; @kuijjer_lionessrsinglesample_2019] uses the Pearson correlation to infer casewise GRNs.
 To do so, first the Pearson correlation is calculated between regulators and targets for all samples.
@@
 For each cell-specific network, the AUROC and AUPR were calculated.
+
 ## Comparison of RNA velocity methods {#sec:dyngen-velcompare}
@@
 To visualize the velocity on an embedding, we used the "velocity_embedding" function, implemented in the scvelo python package.
+
+## Comparison of trajectory alignment with added noise {#sec:dyngen-tacompare}
+
+2 datasets containing a trajectory were generated using dyngen.
+For each dataset, we shuffled 5\% to 75\% of the count matrix, with steps of 5\%.
+Next, we aligned (using DTW) all the base trajectories with all the noisy trajectories.
+
+Let the normalized alignment distance between base trajectory x and trajectory y with z\% noise be $dxy_z$.
+Then we looked at the discrimination power of DTW: how well DTW is able to distinguish between two trajectories with a certain amount of added noise.
+We looked at the following ratios to determine this:
+\begin{equation}
+    \frac{d12_{z}}{d11_{z}} \quad \forall z \in \{0.05, 0.1, 0.15, ..., 0.75 \}
+\end{equation}
+
+\begin{equation}
+    \frac{d21_{z}}{d22_{z}} \quad \forall z \in \{0.05, 0.1, 0.15, ..., 0.75 \}
+\end{equation}
+
+If this ratio is high, it means that DTW is able to distinguish well between the two trajectories. If this ratio is 1, DTW is not able to distinguish between the two trajectories at all.
+
+These datasets and ratios were generated 10 times to reduce possible stochastic effects.
+
+## Code availability {#sec:dyngen-code}
+
+dyngen is available as an open source R package at
+\url{https://github.com/dynverse/dyngen}. All code used
+in this study is made publicly available at
+\url{https://github.com/rcannood/dyngen_analysis}.

### 2020-02-06 render_manuscript.R
update script

@@ -14,14 +14,10 @@ drive <- drive_download(as_id("1BOltsUJBQ8NhF5s9kU_Q470U0U00QblbG8HBgQwEzRE"), t
 
 # read docx
 textreadr::read_docx(drive$local_path, remove.empty = FALSE, trim = FALSE) %>%
-  str_replace_all(" *(\\[@[^\\]]*\\])", "\\1") %>% # remove spaces before citations
+  str_replace_all("([^ ])(\\[@[^\\]]*\\])", "\\1 \\2") %>% # add spaces before citations
+  str_replace_all("^#", "\n#") %>%
   write_lines("manuscript.Rmd")
 
 # render the manuscript
 render("manuscript.Rmd")
-
-# copy content to content.tex
-lines <- read_lines("manuscript.tex")
-content <- lines[seq(which(lines == "\\hypertarget{sec:dyngen-introduction}{%"), which(lines == "\\printbibliography")-1)]
-write_lines(content, "content.tex")
 

### 2020-02-06 manuscript.Rmd
update manuscript

--- 2019-12-09 manuscript.Rmd
@@
 ---
-title: "dyngen: benchmarking with in silico single cells"
+title: "dyngen: a multi-modal simulator for spearheading new single-cell omics analyses"
 author:
-- Robrecht Cannoodt
-- Wouter Saelens
+- Robrecht Cannoodt*
+- Wouter Saelens*
 - Louise Deconinck
@@
 
-**Purpose:** A common problem of pioneering computational tools is that during their development, there are rarely sufficient datasets available for adequately quantitatively assessing its performance.
+# Abstract
+**Purpose:** When developing new types of tools for single-cell analyses, there is often a lack of datasets on which to quantitatively assess the performance.  
 **Results:** We developed dyngen, a multi-modality simulator of single cells.
-In dyngen, the biomolecular state of an _in silico_ cell changes over time according to a set of reactions defined by the cell's gene regulatory network.
-By simulating single cells in terms of its biomolecular state and reactions, the simulator is easily extendible for adding new modalities or experimental procedures.
-We demonstrate dyngen's flexibility by simulating snapshot, time-series and perturbation experiments.
+In dyngen, the biomolecular state of an _in silico_ cell changes over time according to a predefined gene regulatory network.
+We used dyngen to benchmark three emerging ways of analysing single-cell data: RNA velocity, cell-specific network inference and trajectory alignment methods.
 **Conclusion:** dyngen lays the foundations for benchmarking a wide variety of computational single-cell tools and can be used to help kick-start the development of future types of analyses.
-
-**Author contributions:**
-
-* W.S. and R.C. designed the study.
-* R.C., W.S., and L.D. performed the experiments and analysed the data.
-* R.C. and W.S. implemented the dyngen software package.
-* R.C. and W.S. wrote the original manuscript.
-* L.D. wrote the section on trajectory alignment.
-* R.C., W.S., L.D., and Y.S. reviewed and edited the manuscript.
-* Y.S. supervised the project.
 
@@
 
-This presents method developers with a problem.
-The majority of the 250+ peer-reviewed computational tools for analysing single cell omics data were published without a quantitative assessment of the accuracy of the tool.
+This confronts method developers with several challenges.
+The majority of the 250+ peer-reviewed computational tools for analysing single-cell omics data were published without a quantitative assessment of the accuracy of the tool.
 This is partially due to low availability of suitable benchmarking datasets; even if there are sufficient suitable input datasets available, these are often not accompanied by the necessary metadata to serve as ground-truth for a benchmark.
@@
 Generators of scRNA-seq data (e.g.
-splatter[@zappia_splattersimulationsinglecell_2017], powsimR[@vieth_powsimrpoweranalysis_2017], PROSSTT[@papadopoulos_prossttprobabilisticsimulation_2018] and SymSim[@zhang_simulatingmultiplefaceted_2019]) have already been widely used to explore the strengths and weaknesses of computational tools, both by method developers[@street_slingshotcelllineage_2018; @parra_reconstructingcomplexlineage_2018; @lummertzdarocha_reconstructioncomplexsinglecell_2018; @lin_scclassifyhierarchicalclassification_2019] and independent benchmarkers[@duo_systematicperformanceevaluation_2018; @saelens_comparisonsinglecelltrajectory_2019; @soneson_biasrobustnessscalability_2018].
-However, a limitation of scRNA-seq profiles generators is that they would require significant methodological alterations to add additional modalities or experimental conditions.
+splatter [@zappia_splattersimulationsinglecell_2017], powsimR [@vieth_powsimrpoweranalysis_2017], PROSSTT [@papadopoulos_prossttprobabilisticsimulation_2018] and SymSim [@zhang_simulatingmultiplefaceted_2019]) have already been widely used to explore the strengths and weaknesses of computational tools, both by method developers [@street_slingshotcelllineage_2018; @parra_reconstructingcomplexlineage_2018; @lummertzdarocha_reconstructioncomplexsinglecell_2018; @lin_scclassifyhierarchicalclassification_2019] and independent benchmarkers [@duo_systematicperformanceevaluation_2018; @saelens_comparisonsinglecelltrajectory_2019; @soneson_biasrobustnessscalability_2018].
+However, a limitation of existing scRNA-seq profiles generators is that they would require significant methodological alterations to add additional modalities or experimental conditions.
 
-An ideal experiment would be able to observe all aspects of a cell, including a full history of its molecular states, spatial positions and environmental interactions[@stuart_integrativesinglecellanalysis_2019].
+<!--
+\newcommand{\yes}{\ding{51}}
+\newcommand{\no}{}
+\newcommand{\grayline}{\arrayrulecolor{lightgray}\cline{3-7}}
+\newcommand{\blackline}{\arrayrulecolor{black}\cline{3-7}}
+\newcommand{\method}[1]{#1}
+
+\begin{table}[htb!]
+    \centering
+    \begin{tabular}{p{.25cm}l|*{5}{>{\centering\arraybackslash}p{1.25cm}|}}
+    \blackline
+    & \method{} & \method{splatter} & \method{powsimR} & \method{PROSSTT} & \method{SymSim} & \method{dyngen} \\
+    & & \cite{zappia_splattersimulationsinglecell_2017} & \cite{vieth_powsimrpoweranalysis_2017} & \cite{papadopoulos_prossttprobabilisticsimulation_2018} & \cite{zhang_simulatingmultiplefaceted_2019} & \\
+    \blackline
+    \multicolumn{7}{l}{\textbf{Available modality outputs}} \\
+    \blackline
+    - & mRNA expression & \yes & \yes & \yes & \yes & \yes \\ \grayline
+    - & Pre-mRNA expression & \no & \no & \no & \no & \yes \\ \grayline
+    - & Protein expression & \no & \no & \no & \no & \yes \\ \grayline
+    - & Promotor activity & \no & \no & \no & \yes & \yes \\ \grayline
+    - & Reaction activity & \no & \no & \no & \no & \yes \\
+    \blackline
+    \multicolumn{7}{l}{\textbf{Available ground-truth outputs}} \\
+    \blackline
+    - & True counts & \yes & \no & \no & \yes & \yes \\ \grayline
+    - & Cluster labels & \yes & \yes & \no & \yes & ? \\ \grayline
+    - & Trajectory & \yes & \no & \yes $ ^1$ & \yes & \yes \\ \grayline
+    - & Batch labels & \yes & \no & \no & \no & \yes $ ^1$ \\ \grayline
+    - & Differential expression & \no & \no & \no & \yes & \yes $^2$ \\ %\grayline
+    - & Knocked down regulators & \no & \no & \no & \no & \yes \\ \grayline
+- & Regulatory network & \no & \no & \no & \yes & \yes \\ \grayline
+- & Cell-specific regulatory network & \no & \no & \no & \no & \yes \\
+    \blackline
+    \multicolumn{7}{l}{\textbf{Emulate experimental effects}} \\
+    \blackline
+    - & Single-cell RNA sequencing & \yes & \yes & \yes & \yes & \yes \\ \grayline
+    - & Batch effects & \yes & \no & \no & \no & \yes $ ^1$ \\ \grayline
+    - & Knockdown experiment & \no & \no & \no & \no & \yes \\ \grayline
+    - & Time-series & ? & ? & ? & ? & \yes \\
+    - & Snapshot & ? & ? & ? & ? & \yes \\
+    \blackline
+    \multicolumn{7}{l}{\textbf{Possible applications}} \\
+    \blackline
+    - & Clustering & \yes & \yes & \no & \yes & ? \\ \grayline
+    - & Trajectory inference & \yes & \no & \yes & \yes & \yes \\ \grayline
+    - & Network inference & \no & \no & \no & \yes & \yes \\ \grayline
+    - & Cell-specific network inference & \no & \no & \no & \no & \yes \\ \grayline
+    - & Differential expression & \yes & \no & \no & \no & \yes $^2$ \\ \grayline
+    - & Batch effect correction & \yes & \no & \no & \no & \yes \\ \grayline
+    - & RNA Velocity & \no & \no & \no & \no & \yes \\ \grayline
+    - & Trajectory alignment & \no & \no & \no & \no & \yes \\
+    \blackline
+    \multicolumn{7}{l}{\textbf{Showcased dataset}} \\
+    \blackline
+    - & Number of cells &  &  &  &  & \\
+    - & Number of genes &  &  &  &  & \\
+    - & Execution time & & & & & \\
+    \blackline
+    \end{tabular}
+    \caption{
+    \textbf{Comparison of single-cell simulators.}
+    $ ^1$ Showcased by vignette. $^2$ Showcased by Van den Berge et al.  \cite{vandenberge_trajectorybaseddifferentialexpression_2019}.
+    }
+    \label{tab:compare}
+\end{table}
+-->
+
+<!--
+method
+model
+modalities
+(derived) ground truths
+(possible) applications
+splatter [@zappia_splattersimulationsinglecell_2017]
+Several distributional models with parameters that allow the generation of datasets by sampling from distributions
+Cell types, linear trajectories and batch effects can be simulated by changing the parameters of certain genes
+mRNA expression
+cluster labels
+trajectory
+batch labels
+differential expression
+true counts
+Clustering
+Linear trajectory inference
+Batch effect correction
+Differential expression
+powsimR [@vieth_powsimrpoweranalysis_2017]
+Negative binomial or zero-inflated negative binomial.
+Cell types can be simulated by changing the parameters of this distribution
+mRNA expression
+differential expression
+cluster labels
+Differential expression
+Clustering
+PROSSTT [@papadopoulos_prossttprobabilisticsimulation_2018]
+(1) Generation of a random tree
+(2) The expression of each gene is seen as a sum of contributions from different expression programs. The activity of each program depends on pseudotime.
+(3) Generation of cells and expression by random sampling along the trajectory, and from a negative binomial distribution
+mRNA expression
+trajectory (linear and branching)
+Linear and branching trajectory inference
+SymSim [@zhang_simulatingmultiplefaceted_2019]
+Kinetic model of each gene, in which it can be on or off, and where the mRNA can be degraded.
+The parameters of this model depend on a gene effect and cell effect, each of these effects represents a vector in an n-dimensional space (e.g.
+10).
+The cell effect vectors are sampled in such a way that cells that have to be more similar (according to a tree provided by the user), will also have a more similar vector.
+mRNA expression
+promoter activity
+cluster labels
+trajectory
+differential expression
+true counts
+Clustering
+Linear, branching and disconnected trajectory inference
+
+dyngen
+Simulation of a thermodynamic model of gene regulation, in which a set of transcription factors can bind the promoter and can have varying effects on the activity of the promoter.
+Clusters or trajectories are simulated by specifying a regulatory network that will induce a certain diversification of expression.
+Expression is simulated using a single-molecule model, after which individual UMIs are simulated using a multinomial distribution.
+Apart from gene regulation, splicing and protein production are also simulated.
+Protein expression
+mRNA expression
+pre-mRNA expression
+promoter activity
+reaction activity
+true counts
+trajectory
+batch labels
+(perturbed transcription factors)
+Clustering
+Linear, branching and disconnected trajectory inference
+Cell-specific network inference
+Batch effect correction
+RNA velocity
+Trajectory alignment
+-->
+
+An ideal experiment would be able to observe all aspects of a cell, including a full history of its molecular states, spatial positions and environmental interactions [@stuart_integrativesinglecellanalysis_2019].
 While this falls outside the reach of current experimental technologies, generating synthetic data in anticipation of new experimental technologies would allow already developing the next wave of computational tools.
@@
 We developed a multi-modality simulator of single cells called dyngen (Figure\ \ref{fig:showcase}A).
-dyngen uses Gillespie's stochastic simulation algorithm[@gillespie_exactstochasticsimulation_1977]  to simulate gene regulatory interactions at a single-molecule level.
-Its methodology allows tracking of many layers of information throughout the simulation, including the abundance of any molecule in the cell, progression of the cell along a dynamic process, and the activation strength of individual regulatory interactions.
+dyngen uses an optimized version of Gillespie's stochastic simulation algorithm [@gillespie_exactstochasticsimulation_1977]  to simulate gene regulation, splicing and translation at a single-molecule level.
+Its methodology allows tracking of many layers of information throughout the simulation, including the abundance of any molecule in the cell, the progression of the cell along a dynamic process, and the activation strength of individual regulatory interactions.
 dyngen can simulate a large variety of dynamic processes (e.g.
@@
 The fine-grained controls over simulation parameters allow dyngen to be applicable to a broad range of use-cases.
-We demonstrate this by performing first quantitative evaluations of three types of novel computational approaches: RNA velocity, casewise network inference and trajectory alignment methods.
+We demonstrate this by performing first quantitative evaluations of three types of novel computational approaches: RNA velocity, cell-specific network inference and trajectory alignment methods.
 
@@
         \textbf{B:} The input GRN is defined such that it models a dynamic process of interest.
-        \textbf{C:} The reactions define how abundance levels of molecules change at any particular timepoint.
+        \textbf{C:} The reactions define how abundance levels of molecules change at any particular time point.
         \textbf{D:} Firing many reactions can significantly alter the cellular state over time.
@@
 
-dyngen returns many modalities throughout the whole simulation: molecular abundance, cellular state, number of reaction firings, reaction likelihoods, and regulation activations (Figure\ \ref{fig:simplecyclic}C--F).
+dyngen returns many modalities throughout the whole simulation: molecular abundance, number of reaction firings, reaction likelihoods, and regulation activations (Figure\ \ref{fig:simplecyclic}C--F).
 These modalities can serve both as input data and ground truth for benchmarking many types of computational approaches.
-For example, a network inference method could use mRNA abundance and cellular states as inputs and its output could be benchmarked against the gold standard GRN.
+For example, a network inference method could use mRNA abundance and regulation activities as inputs and its output could be benchmarked against the gold standard GRN.
 
@@
 Together, these qualities allow it to be applicable in benchmarking a broad range of use-cases.
-To demonstrate, we apply dyngen on several novel computational approaches which have recently had a major impact on how single-cell analyses are performed but for which quantitative assessment of the performance was hitherto lacking.
+In practice, dyngen has already successfully been used to evaluate trajectory inference [@saelens_comparisonsinglecelltrajectory_2019], trajectory-based differential expression [@vandenberge_trajectorybaseddifferentialexpression_2019], and network inference [@pratapa_benchmarkingalgorithmsgene_2019] methods.
+To demonstrate this point even further, we apply dyngen on several promising novel computational approaches for which quantitative assessment of the performance was until now lacking.
 
-## Example use cases {#sec:dyngen-example}
+## RNA velocity
 
-### Trajectory alignment
-
-### Casewise network inference
-
-Casewise network inference (CNI) methods\footnote{Other terms are commonly used when dealing with data from a particular source.
-For example, single-cell NI when applied to single-cell transcriptomics data; sample-specific NI when applied to bulk transcriptomics; patient-derived NI when applied to bulk profiles of patients.} predict not only which transcription factors regulate which target genes (Figure\ \ref{fig:scgrn}A, top left), but also how active each interaction is in every case (Figure\ \ref{fig:scgrn}A).
-In CNI, a 'case' might be a cell, but it might also refer to a bulk sample.
-
-While a few pioneering CNI approaches have already been developed[@aibar_scenicsinglecellregulatory_2017; @kuijjer_estimatingsamplespecificregulatory_2019; @liu_personalizedcharacterizationdiseases_2016], a quantitative assessment of the performance is hitherto lacking.
-This is not surprising, as neither real nor in silico datasets of cell-specific or even cell-type-specific interactions exists that is large enough so that it can be used as a ground-truth for evaluating CNI methods.
-
-Since dyngen computes tracks the probability of transcription, temporarily 'knocking down' the expression of a regulator and observing the change in transcription probability.
-This is a much more accurate ground-truth for regulatory interaction between a regulator and target in comparison to observing the change in transcript abundance levels when knocking down a regulator, as the regulation of the target could be indirect.
-
-We used this ground-truth to compare the performance of three CNI methods  (Figure\ \ref{fig:scgrn}B).
-We calculated the AUROC and AUPR score -- which are common metrics for NI benchmarking -- for each cell individually.
-Computing the mean AUROC and AUPR per dataset showed that pySCENIC significantly outperforms LIONESS and SSN.
-
-This comparison could be extended to include analyses on the scalability of execution time w.r.t dataset size, the stability of the results in function of noise, and the usability toward end users.
-
-\begin{figure}[htb!]
-    \centering
-    \includegraphics[width=.8\linewidth]{fig/network_inference/cni.pdf}
-    \caption{
-    \textbf{dyngen allows benchmarking Casewise Network Inference (CNI) methods.} 
-    \textbf{A:} A cell is simulated using the global gene regulatory network (GRN, top left).
-However, at any particular state in the simulation, only a fraction of the gene regulatory interactions are active.
-    \textbf{B:} CNI methods were executed to predict the regulatory interactions that are active in each cell specifically.
-Using the ground-truth casewise GRN, the performance of each method was quantified on 14 dyngen datasets.
-    }
-    \label{fig:scgrn}
-\end{figure}
-
-### RNA velocity
-
-In eukaryotes, a gene is first transcribed to a pre-mRNA, and subsequently spliced into mature mRNA.
-Because reads coming from both unspliced and spliced transcripts are observed in expression data, the relative ratio between the two can tell us something about which genes are increasing, decreasing or remaining the same[@zeisel_coupledpremrnamrna_2011;@lamanno_rnavelocitysingle_2018].
-To determine this, some parameters have to be estimated to determine which fraction of unspliced and spliced mRNAs corresponds to an increase or decrease.
-The estimation of these parameters makes some assumptions, and can be handled in different ways in the two main algorithms that are now available for RNA velocity estimation: velocyto[@lamanno_rnavelocitysingle_2018] and scvelo[@bergen_generalizingrnavelocity_2019].
+In eukaryotes, a gene is first transcribed to a pre-mRNA and subsequently spliced into mature mRNA.
+Because reads coming from both unspliced and spliced transcripts are observed in expression data, the relative ratio between the two can tell us something about which genes are increasing, decreasing or remaining the same [@zeisel_coupledpremrnamrna_2011;@lamanno_rnavelocitysingle_2018].
+To determine this, some parameters have to be estimated to determine which fraction of unspliced and spliced mRNAs correspond to an increase or decrease.
+The estimation of these parameters makes some assumptions and can be handled in different ways in the two main algorithms that are now available for RNA velocity estimation: velocyto [@lamanno_rnavelocitysingle_2018] and scvelo [@bergen_generalizingrnavelocity_2019].
 It can be difficult to obtain ground truth data to benchmark these algorithms, given that it would require continuous data of transcriptional dynamics in individual cells.
-On the other hand, the ground truth velocity is rapidly extracted from the dyngen model, by looking whether each transcript is currently increasing or decreasing in expression.
+On the other hand, the ground truth velocity is rapidly extracted from the dyngen model, by looking at whether each transcript is currently increasing or decreasing in expression.
 
@@
     \textbf{dyngen allows benchmarking of RNA velocity methods.} 
-    \textbf{A:} We tested five different methods and parameter settings for the estimation of RNA velocity on datasets with varying backbones (colors).
+    \textbf{A:} We tested five different methods and parameter settings for the estimation of RNA velocity on datasets with varying backbones (colours).
 Overall, the velocyto method with the constant velocity assumption performed the best overall.
@@
 
+## Cell-specific network inference
+
+Cell-specific network inference (CSNI) methods\footnote{Different terms are commonly used when dealing with data from a particular source.
+For example, single-cell NI when applied to single-cell transcriptomics data; sample-specific NI when applied to bulk transcriptomics; patient-derived NI when applied to bulk profiles of patients.
+A more generaised variant of CSNI is casewise NI, which does not specify the type of data which is being analysed.} predict not only which transcription factors regulate which target genes (Figure\ \ref{fig:scgrn}A, top left), but also how active each interaction is in every case (Figure\ \ref{fig:scgrn}A).
+
+While a few pioneering CSNI approaches have already been developed [@aibar_scenicsinglecellregulatory_2017; @kuijjer_estimatingsamplespecificregulatory_2019; @liu_personalizedcharacterizationdiseases_2016], a quantitative assessment of the performance is until now lacking.
+This is not surprising, as neither real nor in silico datasets of cell-specific or even cell-type-specific interactions exists that is large enough so that it can be used as a ground-truth for evaluating CSNI methods.
+Extracting the ground-truth dynamic network in dyngen is straightforward though, given that we can calculate how target gene expression would change without the regulator being present.
+
+We used this ground-truth to compare the performance of three CSNI methods (Figure\ \ref{fig:scgrn}B).
+We calculated the AUROC and AUPR score -- which are common metrics for NI benchmarking -- for each cell individually.
+Computing the mean AUROC and AUPR per dataset showed that pySCENIC significantly outperforms LIONESS and SSN.
+
+\begin{figure}[htb!]
+    \centering
+    \includegraphics[width=.8\linewidth]{fig/network_inference/cni.pdf}
+    \caption{
+    \textbf{dyngen allows benchmarking Cell-specific Network Inference (CSNI) methods.} 
+    \textbf{A:} A cell is simulated using the global gene regulatory network (GRN, top left).
+However, at any particular state in the simulation, only a fraction of the gene regulatory interactions are active.
+    \textbf{B:} CSNI methods were executed to predict the regulatory interactions that are active in each cell specifically.
+Using the ground-truth cell-specific GRN, the performance of each method was quantified on 14 dyngen datasets.
+    }
+    \label{fig:scgrn}
+\end{figure}
+
+## Trajectory alignment
+
+Trajectory alignment allows studying the differences between the same trajectories from different samples.
+For example, the cell developmental process of a patient could be compared to that of a healthy control to detect the transcriptomic differences of a particular lineage.
+
+Dynamic Time Warping (DTW) [@giorgino_computingvisualizingdynamic_2009] is most commonly used to align linear trajectories.
+DTW is a technique originating in the field of speech recognition and aligns temporal sequences by creating a warping path between the sequences that indicates which sequence must be dilated or contracted to best match the other one.
+DTW can also be used in combination with smoothed pseudocells [@alpert_alignmentsinglecelltrajectories_2018].
+
+Trajectory alignment, using DTW, has been used to compare gene expression kinetics resulting from different biological processes [@cacchiarelli_aligningsinglecelldevelopmental_2018], to compare human, chimpanzee and macaque neuronal development [@kanton_organoidsinglecellgenomic_2019], to find differences in gene regulation in the presence of certain growth factors [@mcfaline-figueroa_pooledsinglecellgenetic_2019], and to compare human and mouse embryogenesis [@alpert_alignmentsinglecelltrajectories_2018].
+
+dyngen can be used to evaluate trajectory alignment techniques.
+We analyse how well DTW distinguishes between different trajectories for increasing amounts of noise.
+When noise increases, DTW can no longer distinguish trajectories clearly (Figure\ \ref{fig:traj_align}).
+In this case, there is no difference between when smoothed pseudocells are used or when actual cells and their gene expression are used.
+
+\begin{figure}[htb!]
+    \centering
+    \includegraphics[width=.8\linewidth]{fig/traj_alignment/tares.pdf}
+    \caption{
+    \textbf{dyngen allows benchmarking of trajectory alignment methods.} 
+  \textbf{A:} Example of trajectory alignment using DTW. The segments between the two trajectories show which cells are aligned.
+\textbf{B:} Shows the discrimination power of DTW decreasing as the amount of noise added increases.
+    }
+    \label{fig:traj_align}
+\end{figure}
+
 # Discussion {#sec:dyngen-discussion}
-As is, dyngen's single cell simulations can be used to evaluate common single-cell omics computational methods such as clustering, batch correction, trajectory inference and network inference.
+As is, dyngen's single-cell simulations can be used to evaluate common single-cell omics computational methods such as clustering, batch correction, trajectory inference and network inference.
 However, the combined effect of these advantages results in a framework that is flexible enough to adapt to a broad range of applications.
 This may include methods that integrate clustering, network inference and trajectory inference.
-In this respect, dyngen may promote the development of new tools in the single-cell field similarly as other simulators have done in the past[@schaffter_genenetweaversilicobenchmark_2011; @ewing_combiningtumorgenome_2015].
+In this respect, dyngen may promote the development of new tools in the single-cell field similarly as other simulators have done in the past [@schaffter_genenetweaversilicobenchmark_2011; @ewing_combiningtumorgenome_2015].
 
@@
 In this way, it is possible to design and evaluate the performance and robustness of new types of computational analyses before experimental data becomes available.
-In addition, it could also be used to compare which experimental protocol is the most cost-effective in producing the qualitative and robust results in downstream analysis.
+In addition, it could also be used to compare which experimental protocol is the most cost-effective in producing qualitative and robust results in downstream analysis.
 
 Currently, dyngen focuses on simulating cells as standalone entities that are well mixed.
-Splitting up the simulation space into separate subvolumes could pave the way to better study key cellular processes such as cell division, intercellular communication and migration[@smith_spatialstochasticintracellular_2019].
+Splitting up the simulation space into separate subvolumes could pave the way to better study key cellular processes such as cell division, intercellular communication and migration [@smith_spatialstochasticintracellular_2019].
 
 # Methods {#sec:dyngen-methods}
-The workflow to generate _in silico_ single cell data consists of six main steps (Figure\ \ref{fig:explain_methods}).
+The workflow to generate _in silico_ single-cell data consists of six main steps (Figure\ \ref{fig:explain_methods}).
 
@@
 The exact way a cell chooses a certain path during its differentiation is still an active research field, although certain models have already emerged and been tested _in vivo_.
-One driver of bifurcation seems to be mutual antagonism, where two genes strongly repress each other[@rekhtman_directinteractionhematopoietic_1999; @xu_regulationbifurcatingcell_2015], forcing one of the two to become inactive[@graf_forcingcellschange_2009].
-Such mutual antagonism can be modelled and simulated[@wang_quantifyingwaddingtonlandscape_2011; @ferrell_bistabilitybifurcationswaddington_2012].
-Although the two-gene model is simple and elegant, the reality is frequently more complex, with multiple genes (grouped into modules) repressing each other[@yosef_dynamicregulatorynetwork_2013].
+One driver of bifurcation seems to be mutual antagonism, where two genes strongly repress each other [@rekhtman_directinteractionhematopoietic_1999; @xu_regulationbifurcatingcell_2015], forcing one of the two to become inactive [@graf_forcingcellschange_2009].
+Such mutual antagonism can be modelled and simulated [@wang_quantifyingwaddingtonlandscape_2011; @ferrell_bistabilitybifurcationswaddington_2012].
+Although the two-gene model is simple and elegant, the reality is frequently more complex, with multiple genes (grouped into modules) repressing each other [@yosef_dynamicregulatorynetwork_2013].
 
@@
 The module network is what mainly determines which dynamic processes occur within the simulated cells.
-The backbone is a separate set of simulations in which the ground-truth topology of the dynamic processes are defined, as it is difficult to determine the topology of the dynamic processes from the module network itself.
 
@@
 Two modules repressing each other is the basis of a _bifurcating_ process, though several chains of modules have to be attached in order to achieve progression before and after the bifurcation process.
-Finally, a \emph{converging} process has a bifurcation occurring during the burn phase, after which any differences in module regulation is removed.
+Finally, a _converging_ process has a bifurcation occurring during the burn phase, after which any differences in module regulation is removed.
 
-Note that these examples represent the bare minimum in terms of number of modules used.
+Note that these examples represent the bare minimum in terms of the number of modules used.
 Using longer chains of modules is typically desired.
@@
 
-In addition to the module network, the user also needs to define a network of cellular states called the "backbone".
-Before simulating any cells, each transition in the backbone is simulated separately to obtain the average changes in expression along that transition (Figure\ \ref{fig:explain_methods}D).
-As part of the backbone, the user needs to specify which modules are allowed to alter its expression from one state to another.
-For example, in order to transition from state S0 to S1 in the cyclic example, gene modules A, B and C are turned on and a simulation is allowed to run.
-To transition from S1 to S2, gene modules D and E are turned on, and expression of gene module C is kept constant.
-To transition from S2 to S3, C is turned on again and now A and B are fixed.
-Finally, to transition from S3 to S1 again, A and B are turned on again and D and E are fixed again.
-Demonstrations of the backbone will be explained in more detail in section \ref{sec:dyngen-simbackbone}.
-
 ### Backbone lego {#sec:dyngen-bbl}
@@
 A **leaf BBL** (Figure\ \ref{fig:backbone_lego}E) is a linear BBL that has either no inputs or no outputs.
-A _start_ BBL is a linear BBL where the first module has basal expression, and all modules in this module will be active during the burn-in phase of the simulation (Section \ref{sec:dyngen-simbackbone}).
+A _start_ BBL is a linear BBL where the first module has basal expression, and all modules in this module will be active during the burn-in phase of the simulation.
 An _end_ BBL is also a linear BBL with its output regulating one final module.
@@
 The GRN defines how a reaction affects the abundance levels of molecules and how likely it will occur.
-Gibson and Bruck[@gibson_probabilisticmodelprokaryotic_2000] provide a good introduction to modelling gene regulation with stochastic frameworks, on which many of the concepts below are based.
+Gibson and Bruck [@gibson_probabilisticmodelprokaryotic_2000] provide a good introduction to modelling gene regulation with stochastic frameworks, on which many of the concepts below are based.
 
@@
         Dissociation constant & $\dis H$ & $= 0.5 \times \frac{\wpr H \times \wsr H \times \ypr H}{(\wdr H + \wsr H) \times \xdr H \times \ydr H}$ \\
-        Binding & $\buf{G,H}$ & $= \left(\str{G,H} \times \y H\ /\ \dis H\right) ^ {\hill{G,H}}$ \\
+        Binding & $\buf{G,H}$ & $= \str{G,H} \times \left(\y H\ /\ \dis H\right) ^ {\hill{G,H}}$ \\
         Basal expression & $\ba G$ & $= \begin{cases} 1 & \mbox{if } \rp{G} = \emptyset \\ 0.0001 & \mbox{if } \rn{G} = \emptyset \mbox{ and } \rp{G} \neq \emptyset \\ 0.5 & \mbox{otherwise} \end{cases}$ * \\ \hline
@@
 
-The propensity of the transcription of a gene $G$ is inspired by thermodynamic models of gene regulation [needcitation], in which the promoter of $G$ can be bound or unbound by a set of $N$ transcription factors $H_i$.
+The propensity of the transcription of a gene $G$ is inspired by thermodynamic models of gene regulation [@schilstra_biologicgeneexpression_2008], in which the promoter of $G$ can be bound or unbound by a set of $N$ transcription factors $H_i$.
 Let $\proptran(\y 1, \y 2, \ldots, \y N)$ denote the propensity function of $G$, in function of the abundance levels of the transcription factors.
@@
 
-## Compute average expression along backbone transitions {#sec:dyngen-simbackbone}
-When simulating the developmental backbone, we go through the edges of the backbone state network defined in an earlier step (Section \ref{sec:dyngen-backbone}), starting from the root state.
-It is assumed the root state has no modules active and has no expression of any molecules.
-To get to the next state, we follow a transition starting from the root state, activate and deactivate the modules as indicated by the transition, and compute the average molecule abundance along the transition.
-To compute the average abundance, we perform small time steps $t = 0.001$ and let each reaction (Section \ref{sec:dyngen-reactions}) occur $t$ times its propensity.
-
 ## Simulate single cells {#sec:dyngen-simcell}
-dyngen uses Gillespie's stochastic simulation algorithm (SSA)[@gillespie_exactstochasticsimulation_1977] to simulate dynamic processes.
+dyngen uses Gillespie's stochastic simulation algorithm (SSA) [@gillespie_exactstochasticsimulation_1977] to simulate dynamic processes.
 An SSA simulation is an iterative process where at each iteration one reaction is triggered.
@@
 GillespieSSA2 is an optimised library for performing SSA simulations.
-The propensity functions are compiled to C++ and SSA approximations can be used which allow to trigger many reactions simultaneously at each iteration.
-The framework also allows to store the abundance levels of molecules only after a specific interval has passed since the previous census.
+The propensity functions are compiled to C++ and SSA approximations can be used which allow triggering many reactions simultaneously at each iteration.
+The framework also allows storing the abundance levels of molecules only after a specific interval has passed since the previous census.
 By setting the census interval to 0, the whole simulation's trajectory is retained but many of these time points will contain very similar information.
 In addition to the abundance levels, also the propensity values and the number of firings of each of the reactions at each of the time steps can be retained, as well as specific sub-calculations of the propensity values, such as the regulator activity level $reg_{G,H}$.
-
-### Map SSA simulations to backbone
-We compute the Pearson correlation between the state vectors in the simulation and the average expression levels along a transition in the backbone.
-Each timepoint in the SSA simulation is mapped to the point in the backbone that has the highest correlation value.
 
@@
 Molecules are sampled from the simulation to replicate how molecules are experimentally sampled.
-A real dataset is downloaded from a repository of single-cell RNA-seq datasets[@cannoodt_singlecellomicsdatasets_2018].
-For each _in silico_ cell $i$, draw its library size $ls_i$from the distribution of transcript counts per cell in the real dataset.
+A real dataset is downloaded from a repository of single-cell RNA-seq datasets [@cannoodt_singlecellomicsdatasets_2018].
+For each _in silico_ cell $i$, draw its library size $ls_i$ from the distribution of transcript counts per cell in the real dataset.
 The capture rate $cr_j$ of each _in silico_ molecule type $j$ is drawn from $N(1, 0.05)$.
@@
 
-## Determining the casewise ground-truth regulatory network {#sec:dyngen-extractgrn}
+## Determining the ground-truth trajectory {#sec:dyngen-groundtruth}
+To determine the state of a cell at any point in the simulation, we simulate the expression of the ground-truth trajectory.
+Subsequently, map the expression of the cell to the closest resembling state in the ground-truth trajectory by finding the point in the ground-truth trajectory with the highest Pearson correlation value w.r.t.
+the expression values.
+
+To simulate the expression of the ground-truth expression, we iterate over the edges in the backbone state network starting from the root state.
+We assume the root state has no modules active and has no expression of any molecules.
+To transition to a next state, we activate or deactivate certain gene modules as defined by the backbone (Section\ \ref{sec:dyngen-backbone}) and perform a noiseless ODE simulation.
+For each time interval $t=0.01$ along the trajectory, the expression values are retained.
+
+## Determining the cell-specific ground-truth regulatory network {#sec:dyngen-extractgrn}
 Calculating the regulatory effect of a regulator $R$ on a target $T$ (Figure\ \ref{fig:explain_methods}F) requires determining the contribution of $R$ in the propensity function of the transcription of $T$ (section\ \ref{sec:dyngen-reactions}) with respect to other regulators.
-This information is useful, amongst others, for benchmarking casewise network inference methods.
+This information is useful, amongst others, for benchmarking cell-specific network inference methods.
 
@@
 
-Determining the regulatory effect for all interactions and cells in the dataset yields the complete casewise ground-truth GRN (Figure\ \ref{fig:casewise_grn}).
+Determining the regulatory effect for all interactions and cells in the dataset yields the complete cell-specific ground-truth GRN (Figure\ \ref{fig:cell-specific_grn}).
 The regulatory effect lie between $[-1, 1]$, where -1 represents complete inhibition of $T$ by $R$, 1 represents maximal activation of $T$ by $R$, and 0 represents inactivity of the regulatory interaction between $R$ and $T$.
@@
     \caption{
-        \textbf{The casewise regulatory effects of all interactions, computed on cells part of a bifurcation trajectory.} Negative values correspond to inhibitory interactions, positive values to activating interactions, and zero values correspond to inactive interactions.
+        \textbf{The cell-specific regulatory effects of all interactions, computed on cells part of a bifurcation trajectory.} Negative values correspond to inhibitory interactions, positive values to activating interactions, and zero values correspond to inactive interactions.
     }
-    \label{fig:casewise_grn}
+    \label{fig:cell-specific_grn}
 \end{figure}
 
-## Comparison of casewise network inference methods {#sec:dyngen-nicompare}
+## Comparison of cell-specific network inference methods {#sec:dyngen-nicompare}
 Several datasets were generated using the different predefined backbones.
-For every cell in the dataset, the transcriptomics profile and the corresponding casewise ground-truth regulatory network was determined (Section\ \ref{sec:dyngen-extractgrn}).
+For every cell in the dataset, the transcriptomics profile and the corresponding cell-specific ground-truth regulatory network was determined (Section\ \ref{sec:dyngen-extractgrn}).
 
-Several casewise NI methods were considered for comparison: SCENIC[@aibar_scenicsinglecellregulatory_2017], LIONESS[@kuijjer_estimatingsamplespecificregulatory_2015; @kuijjer_estimatingsamplespecificregulatory_2019], and SSN[@liu_personalizedcharacterizationdiseases_2016].
+Several cell-specific NI methods were considered for comparison: SCENIC [@aibar_scenicsinglecellregulatory_2017], LIONESS [@kuijjer_estimatingsamplespecificregulatory_2015; @kuijjer_estimatingsamplespecificregulatory_2019], and SSN [@liu_personalizedcharacterizationdiseases_2016].
 
-LIONESS[@@kuijjer_estimatingsamplespecificregulatory_2019; kuijjer_lionessrsinglesample_2019] uses the Pearson correlation to infer casewise GRNs.
+LIONESS [@@kuijjer_estimatingsamplespecificregulatory_2019; @kuijjer_lionessrsinglesample_2019] uses the Pearson correlation to infer cell-specific GRNs.
 To do so, first the Pearson correlation is calculated between regulators and targets for all samples.
 Next, the Pearson correlation is again calculated for all samples excluding one sample.
-The difference between the two correlation matrices is considered a casewise GRN for that particular profile.
-This process is repeated for all profiles, resulting in a casewise GRN.
+The difference between the two correlation matrices is considered a cell-specific GRN for that particular profile.
+This process is repeated for all profiles, resulting in a cell-specific GRN.
 
-SSN[@liu_personalizedcharacterizationdiseases_2016] has, in essence, the exact same methodology as LIONESS.
+SSN [@liu_personalizedcharacterizationdiseases_2016] has, in essence, the exact same methodology as LIONESS.
 It is worth noting that the LIONESS preprint was released before the publication of SSN.
-Since no implementation was provided by the authors, we implemented SSN in R using basic R and tidyverse functions[@wickham_welcometidyverse_2019] and marked results from this implementation as "SSN*".
+Since no implementation was provided by the authors, we implemented SSN in R using basic R and tidyverse functions [@wickham_welcometidyverse_2019] and marked results from this implementation as "SSN*".
 
-SCENIC[@aibar_scenicsinglecellregulatory_2017] is a pipeline that consists of four main steps.
-Step 1: classical network inference is performed with arboreto, which is similar to GENIE3[@huynh-thu_inferringregulatorynetworks_2010].
+SCENIC [@aibar_scenicsinglecellregulatory_2017] is a pipeline that consists of four main steps.
+Step 1: classical network inference is performed with arboreto, which is similar to GENIE3 [@huynh-thu_inferringregulatorynetworks_2010].
 Step 2: select the top 10 regulators per target.
@@
 Step 4: for each cell, determine an activity score of each module using AUCell.
-As a post-processing of this output, all modules and the corresponding activity scores are combined back into a casewise GRN consisting of (cell, regulator, target, score) pairs.
+As a post-processing of this output, all modules and the corresponding activity scores are combined back into a cell-specific GRN consisting of (cell, regulator, target, score) pairs.
 For this analysis, the Python implementation of SCENIC was used, namely pySCENIC.
@@
 The AUROC and AUPR metrics are common metrics for evaluating a predicted GRN with a ground-truth GRN.
-To compare a predicted casewise GRN with the ground-truth casewise GRN, the top 10'000 interactions per cell were retained.
+To compare a predicted cell-specific GRN with the ground-truth cell-specific GRN, the top 10'000 interactions per cell were retained.
 For each cell-specific network, the AUROC and AUPR were calculated.
+
 ## Comparison of RNA velocity methods {#sec:dyngen-velcompare}
@@
 We compared two RNA velocity methods.
-The velocyto method[@lamanno_rnavelocitysingle_2018], as implemented in the velocyto.py package, in which we varied the "assumption" parameter between "constant_unspliced" and "constant_velocity".
-The scvelo method[@bergen_generalizingrnavelocity_2019], as implemented in the python scvelo package (http://scvelo.de), in which we varied the "mode" parameter between "deterministic", "stochastic" and "dynamical".
+The velocyto method [@lamanno_rnavelocitysingle_2018], as implemented in the velocyto.py package, in which we varied the "assumption" parameter between "constant_unspliced" and "constant_velocity".
+The scvelo method [@bergen_generalizingrnavelocity_2019], as implemented in the python scvelo package [scvelo.de](http://scvelo.de), in which we varied the "mode" parameter between "deterministic", "stochastic" and "dynamical".
 For both methods, we used the same normalized data as provided by dyngen, with no extra cell or feature filtering.
@@
 To visualize the velocity on an embedding, we used the "velocity_embedding" function, implemented in the scvelo python package.
+
+## Comparison of trajectory alignment with added noise {#sec:dyngen-tacompare}
+
+2 datasets containing a trajectory were generated using dyngen.
+For each dataset, we shuffled 5\% to 75\% of the count matrix, with steps of 5\%.
+Next, we aligned (using DTW) all the base trajectories with all the noisy trajectories.
+
+Let the normalized alignment distance between base trajectory x and trajectory y with z\% noise be $dxy_z$.
+To calculate how well an alignment algorithm can distinguish the two trajectories, we define the discrimination power of DTW.
+
+\begin{equation}
+    \frac{d12_{z}}{d11_{z}} \quad \forall z \in \{0.05, 0.1, 0.15, ..., 0.75 \}
+\end{equation}
+
+\begin{equation}
+    \frac{d21_{z}}{d22_{z}} \quad \forall z \in \{0.05, 0.1, 0.15, ..., 0.75 \}
+\end{equation}
+
+If this ratio is high, it means that DTW is able to distinguish well between the two trajectories. If this ratio is 1, DTW is not able to distinguish between the two trajectories at all.
+
+These datasets and ratios were generated 10 times.
+
+# Availability
+dyngen is available as an R package on GitHub [github.com/dynverse/dyngen](https://github.com/dynverse/dyngen).
+The dyngen codebase is still under development and is thus likely subject to change, including the adding or removing of functionality, renaming functions or parameters, and changing default parameter values.
+The analyses performed in this manuscript are available on GitHub [github.com/dynverse/dyngen_manuscript](https://github.com/dynverse/dyngen_manuscript).
+
+# Author contributions
+
+* W.S. and R.C. designed the study.
+* R.C., W.S., and L.D. performed the experiments and analysed the data.
+* R.C. and W.S. implemented the dyngen software package.
+* R.C. and W.S. wrote the original manuscript.
+* L.D. wrote the section on trajectory alignment.
+* R.C., W.S., L.D., and Y.S. reviewed and edited the manuscript.
+* Y.S. supervised the project.

### 2020-02-06 manuscript.Rmd
change date to equal contribution

@@
 - Yvan Saeys
-date: "`r format(Sys.time(), '%d %B %Y')`"
+date: "* Equal contribution"
 output:

### 2020-05-18 analysis/usecase_trajectory_alignment/1_datasets.R
small improvements

@@ -14,7 +14,7 @@ backbone <- bblego(
 )
 
 noise_levels <- seq(from = 0.1, to = 1, by = 0.1)
-only_noise <- T
+only_noise <- TRUE
 
 # setup dataset design
 design_datasets <- exp$result("design_datasets.rds") %cache% {
@@ -23,10 +23,12 @@ design_datasets <- exp$result("design_datasets.rds") %cache% {
     backbone_name = "linear",
     noise = noise_levels
   ) %>%
-    mutate(base_id1 = paste0(backbone_name, seed, "_1")) %>%
-    mutate(base_id2 = paste0(backbone_name, seed, "_2")) %>%
-    mutate(id1 = paste0(base_id1, "_", noise)) %>%
-    mutate(id2 = paste0(base_id2, "_", noise))
+    mutate(
+      base_id1 = paste0(backbone_name, seed, "_1"),
+      base_id2 = paste0(backbone_name, seed, "_2"),
+      id1 = paste0(base_id1, "_", noise),
+      id2 = paste0(base_id2, "_", noise)
+    )
 }
 
 # Generates only the base models. The noise is added onto these base models
@@ -60,15 +62,15 @@ pwalk(design_datasets, function(base_id1, base_id2, id1, id2, seed, backbone_nam
       make_plots = TRUE
     )
 
-    dataset1 <- readRDS(paste0(exp$dataset_folder(base_id1), "dataset.rds"))
-    model1 <- readRDS(paste0(exp$dataset_folder(base_id1), "model.rds"))
+    dataset1 <- read_rds(paste0(exp$dataset_folder(base_id1), "dataset.rds"))
+    model1 <- read_rds(paste0(exp$dataset_folder(base_id1), "model.rds"))
 
     # Generate the second trajectory of the pair
     model2 <- model1 %>% generate_cells() %>% generate_experiment()
     dataset2 <- wrap_dataset(model2)
 
-    saveRDS(model2, file = paste0(exp$dataset_folder(base_id2), "model.rds"))
-    saveRDS(dataset2, file = paste0(exp$dataset_folder(base_id2), "dataset.rds"))
+    write_rds(model2, paste0(exp$dataset_folder(base_id2), "model.rds"), compress = "gz")
+    write_rds(dataset2, paste0(exp$dataset_folder(base_id2), "dataset.rds"), compress = "gz")
 
[…]

### 2020-05-18 analysis/usecase_trajectory_alignment/2_smoothing.R

@@ -6,67 +6,52 @@ library(viridis)
 
 exp <- start_analysis("usecase_trajectory_alignment")
 
-# For each pair, use 3 different kind of processing: smoothing, subsampling, original cells
+# For each pair, use 3 different kind of processing
 design_smoothing <- exp$result("design_smoothing.rds") %cache% {
-  read_rds(exp$result("design_datasets.rds")) %>%
-    select("base_id1","base_id2", "id1", "id2", "noise") %>%
-    mutate("smooth" = c(rep("smoothed", 20), rep("subsampled", 40), rep("original cells", 40))) %>%
-    expand(nesting(base_id1, base_id2, id1, id2, noise), smooth)
+  crossing(
+    read_rds(exp$result("design_datasets.rds")),
+    smooth = c("smoothed", "subsampled", "original cells")
+  )
 }
 
+#' design_smoothing %>% mutate(rn = row_number()) %>% dynutils::extract_row_to_list(1) %>% list2env(.GlobalEnv)
+result_smoothing <- exp$result("result_smoothing.rds") %cache% pmap_dfr(
+  design_smoothing %>% mutate(rn = row_number()),
+  function(base_id1, base_id2, id1, id2, smooth, rn, noise, ...) {
 
-alignment_results <- pmap(design_smoothing %>% mutate(rn = row_number()),
-      function(base_id1, base_id2, id1, id2, smooth, rn, noise) {
+    dataset1 <- read_rds(exp$dataset_file(id1))
+    dataset2 <- read_rds(exp$dataset_file(id2))
 
-        dataset1 <- read_rds(exp$dataset_file(id1))
-        dataset2 <- read_rds(exp$dataset_file(id2))
+    if (smooth == "smoothed") {
+      res1 <- get_waypoint_expression(dataset1, 100)
+      res2 <- get_waypoint_expression(dataset2, 100)
+    } else {
+      res1 <- get_cell_expression(dataset1)
+      res2 <- get_cell_expression(dataset2)
+    }
 
-        if(smooth == "smoothed"){
-          res1 <- get_waypoint_expression(dataset1, 100)
-          res2 <- get_waypoint_expression(dataset2, 100)
-        } else {
-          res1 <- get_cell_expression(dataset1)
-          res2 <- get_cell_expression(dataset2)
-        }
[…]

### 2020-05-18 analysis/usecase_trajectory_alignment/3_plots.R

@@ -1,36 +1,29 @@
+library(tidyverse)
 library(dyngen.manuscript)
-library(ggplot2)
-library(viridis)
-library(dtw)
-library(reshape2)
 library(patchwork)
-library(readr)
-library(dplyr)
 
 exp <- start_analysis("usecase_trajectory_alignment")
 
-result_smoothing <- read_rds(exp$result("result_smoothing.rds"))
+result_smoothing <- read_rds(exp$result("result_smoothing.rds")) %>%
+  mutate(
+    noise = ordered(as.factor(noise)),
+    smooth = factor(smooth, levels = c("smoothed", "original cells", "subsampled")),
+    score_scaled = ifelse(smooth == "original cells", score / 10, score)
+  )
 
-result_smoothing$noise <- ordered(as.factor(result_smoothing$noise))
-result_smoothing$smooth <- factor(result_smoothing$smooth, levels = c("smoothed", "original cells", "subsampled"))
-result_smoothing$result_scaled <-ifelse(result_smoothing$smooth == "original cells", result_smoothing$result / 10,  result_smoothing$result)
-
-g <- ggplot(data = result_smoothing, aes(x = ordered(as.factor(noise)), y = result_scaled, fill = smooth)) +
+g <-
+  ggplot(data = result_smoothing, aes(noise, score_scaled, fill = smooth)) +
   geom_boxplot(width = 0.5, size=0.45,outlier.size=0.5) +
-  # scale_fill_viridis_d(direction=-1) +
-  # scale_color_viridis(discrete = TRUE, alpha=0.8) +
-  labs(fill = "Processing method") +
-  xlab("Amount of added noise") +
-  ylab("Distance (lower is better)") +
-  theme_linedraw() +
-  theme(legend.position = "bottom")
+  theme_bw() +
+  theme_common() +
+  labs(x = "Amount of added noise", y = "Distance (lower is better)", fill = "Processing method")
 
 # Plot the three different alignments of a single pair of trajectories
 plot_density <- function(a1, title = "Distance matrix + warping path", subtitle = ""){
   x <- as_tibble(a1$index1)
   x$Y <- a1$index2
 
-  gga_1 <- melt(a1$costMatrix)
+  gga_1 <- reshape2::melt(a1$costMatrix)
   p_heat1 <- ggplot(gga_1, aes(Var1, Var2, fill= value)) +
[…]

### 2020-05-18 analysis/usecase_rna_velocity/1_datasets.R
update

@@ -122,3 +122,13 @@ pwalk(design_datasets, function(id, seed, backbone_name, tr_rate_multiplier, ...
 #   gc()
 # })
 
+
+# # write counts
+# pwalk(design_datasets, function(id, ...) {
+#   cat(id, "\n", sep = "")
+#   dataset <- read_rds(exp$dataset_file(id))
+#   dataset$counts %>% as.matrix %>% as.data.frame() %>% rownames_to_column("cell_id") %>% write_tsv(exp$temporary(paste0("counts/", id, "_spliced.tsv")))
+#   dataset$counts_unspliced %>% as.matrix %>% as.data.frame() %>% rownames_to_column("cell_id") %>% write_tsv(exp$temporary(paste0("counts/", id, "_unspliced.tsv")))
+#   gc()
+# })
+

### 2020-05-18 analysis/usecase_rna_velocity/4_plot_usecases.R

@@ -46,13 +46,49 @@ plot_part_A <-
   ggbeeswarm::geom_quasirandom(aes(color = difficulty), size = 2)  +
   theme_bw() +
   theme_common() +
-  labs(x = "", y = "Correlation (higher is better)", colour = "Difficulty") +
+  labs(x = "", y = "Score", colour = "Difficulty") +
   scale_y_continuous(limits = c(0, 1)) +
   guides(colour = guide_legend(nrow = 3), shape = guide_legend(nrow = 3)) +
   facet_wrap(~metric_label, ncol = 2) +
+  # facet_wrap(~metric_label, ncol = 2, scales = "free_y") +
   scale_colour_brewer(palette = "Dark2")
 
+# methods_info <- scores$summ %>% select(method_id, params_id) %>% unique() %>% mutate(id = paste0(method_id, "_", params_id))
+# stat_df <-
+#   crossing(
+#     id_x = methods_info$id,
+#     id_y = methods_info$id
+#   ) %>%
+#   filter(id_x != id_y) %>%
+#   inner_join(
+#     scores$summ %>%
+#       gather(metric, score, cor, mean_cosine) %>%
+#       transmute(id_x = paste0(method_id, "_", params_id), dataset_id, metric, score_x = score),
+#     by = c("id_x")
+#   ) %>%
+#   inner_join(
+#     scores$summ %>%
+#       gather(metric, score, cor, mean_cosine) %>%
+#       transmute(id_y = paste0(method_id, "_", params_id), dataset_id, metric, score_y = score),
+#     by = c("id_y", "dataset_id", "metric")
+#   ) %>%
+#   group_by(id_x, id_y, metric) %>%
+#   summarise(
+#     test = list(wilcox.test(score_x, score_y, paired = TRUE, alternative = "greater")),
+#     p.value = test[[1]]$p.value
+#   ) %>%
+#   ungroup() %>%
+#   mutate(p.value = p.adjust(p.value, "holm"))
+# stat_df %>% select(-test) %>% mutate(p.value = round(p.value, 2)) %>% spread(metric, p.value) %>% print(n = 30)
+# stat_df %>% select(-test) %>% mutate(p.value = round(p.value, 2)) %>% spread(metric, p.value) %>% reshape2::acast(id_x~id_y, value.var = "cor")
[…]

### 2020-05-18 manuscript.Rmd
update

--- 2020-02-06 manuscript.Rmd
@@
 
-<!--
-method
-model
-modalities
-(derived) ground truths
-(possible) applications
-splatter [@zappia_splattersimulationsinglecell_2017]
-Several distributional models with parameters that allow the generation of datasets by sampling from distributions
-Cell types, linear trajectories and batch effects can be simulated by changing the parameters of certain genes
-mRNA expression
-cluster labels
-trajectory
-batch labels
-differential expression
-true counts
-Clustering
-Linear trajectory inference
-Batch effect correction
-Differential expression
-powsimR [@vieth_powsimrpoweranalysis_2017]
-Negative binomial or zero-inflated negative binomial.
-Cell types can be simulated by changing the parameters of this distribution
-mRNA expression
-differential expression
-cluster labels
-Differential expression
-Clustering
-PROSSTT [@papadopoulos_prossttprobabilisticsimulation_2018]
-(1) Generation of a random tree
-(2) The expression of each gene is seen as a sum of contributions from different expression programs. The activity of each program depends on pseudotime.
-(3) Generation of cells and expression by random sampling along the trajectory, and from a negative binomial distribution
-mRNA expression
-trajectory (linear and branching)
-Linear and branching trajectory inference
-SymSim [@zhang_simulatingmultiplefaceted_2019]
-Kinetic model of each gene, in which it can be on or off, and where the mRNA can be degraded.
-The parameters of this model depend on a gene effect and cell effect, each of these effects represents a vector in an n-dimensional space (e.g.
-10).
-The cell effect vectors are sampled in such a way that cells that have to be more similar (according to a tree provided by the user), will also have a more similar vector.
-mRNA expression
-promoter activity
-cluster labels
-trajectory
-differential expression
-true counts
-Clustering
-Linear, branching and disconnected trajectory inference
-
-dyngen
-Simulation of a thermodynamic model of gene regulation, in which a set of transcription factors can bind the promoter and can have varying effects on the activity of the promoter.
-Clusters or trajectories are simulated by specifying a regulatory network that will induce a certain diversification of expression.
-Expression is simulated using a single-molecule model, after which individual UMIs are simulated using a multinomial distribution.
-Apart from gene regulation, splicing and protein production are also simulated.
-Protein expression
-mRNA expression
-pre-mRNA expression
-promoter activity
-reaction activity
-true counts
-trajectory
-batch labels
-(perturbed transcription factors)
-Clustering
-Linear, branching and disconnected trajectory inference
-Cell-specific network inference
-Batch effect correction
-RNA velocity
-Trajectory alignment
--->
-
 An ideal experiment would be able to observe all aspects of a cell, including a full history of its molecular states, spatial positions and environmental interactions [@stuart_integrativesinglecellanalysis_2019].
-While this falls outside the reach of current experimental technologies, generating synthetic data in anticipation of new experimental technologies would allow already developing the next wave of computational tools.
+While this falls outside the reach of current experimental technologies, generating synthetic data in anticipation of new experimental technologies would allow developing the next wave of computational tools.
 
 We developed a multi-modality simulator of single cells called dyngen (Figure\ \ref{fig:showcase}A).
-dyngen uses an optimized version of Gillespie's stochastic simulation algorithm [@gillespie_exactstochasticsimulation_1977]  to simulate gene regulation, splicing and translation at a single-molecule level.
+dyngen uses an optimized version of Gillespie's stochastic simulation algorithm [@gillespie_exactstochasticsimulation_1977]  to simulate gene regulation, splicing, and translation at a single-molecule level.
 Its methodology allows tracking of many layers of information throughout the simulation, including the abundance of any molecule in the cell, the progression of the cell along a dynamic process, and the activation strength of individual regulatory interactions.
@@
 The fine-grained controls over simulation parameters allow dyngen to be applicable to a broad range of use-cases.
-We demonstrate this by performing first quantitative evaluations of three types of novel computational approaches: RNA velocity, cell-specific network inference and trajectory alignment methods.
+We demonstrate this by performing the first quantitative evaluations of three types of novel computational approaches: RNA velocity, cell-specific network inference, and trajectory alignment methods.
 
@@
     \centering
-    \includegraphics[width=\linewidth]{fig/showcase_4}
+    \includegraphics[width=\linewidth]{result_files/fig1_showcase_functionality/overview_1.pdf}
     \caption{
@@
 
-A cell consists of a set of molecules, the abundance of which are affected by a set of reactions: transcription, splicing, translation, and degradation (Figure\ \ref{fig:simplecyclic}A).
+A cell consists of a set of molecules, the abundance of which is affected by a set of reactions: transcription, splicing, translation, and degradation (Figure\ \ref{fig:simplecyclic}A).
 A gene regulatory network (GRN) defines the reactions that are allowed to occur (Figure\ \ref{fig:simplecyclic}B), which is constructed in such a way that cells slowly develop over time (Figure\ \ref{fig:simplecyclic}C,D).
@@
     \centering
-    \includegraphics[width=\linewidth]{fig/simplecyclic_edited}
+    \includegraphics[width=\linewidth]{result_files/fig2_showcase_data/overview.pdf}
     \caption{
-        \textbf{dyngen models reactions at a single-molecule level and keeps track of multiple levels of information throughput a simulation.} 
-        \textbf{A:} Changes in abundance levels are driven strictly these gene regulatory reactions.
+        \textbf{dyngen models reactions at a single-molecule level and keeps track of multiple levels of information throughout a simulation.} 
+        \textbf{A:} Changes in abundance levels are driven strictly by gene regulatory reactions.
         \textbf{B:} The input GRN is defined such that it models a dynamic process of interest.
@@
     \centering
-    \includegraphics[width=0.6\textwidth]{fig/example_runs_2}
+    \includegraphics[width=0.6\textwidth]{result_files/fig3_showcase_backbones/overview_2.pdf}
     \caption{
@@
 
-\begin{figure}[htb!]
+\begin{figure}[ht]
     \centering
-    \includegraphics[width=.8\linewidth]{results/velocity/usecase.pdf}
+    \includegraphics[width=\linewidth]{result_files/usecase_rna_velocity/usecase.pdf}
     \caption{
     \textbf{dyngen allows benchmarking of RNA velocity methods.} 
-    \textbf{A:} We tested five different methods and parameter settings for the estimation of RNA velocity on datasets with varying backbones (colours).
-Overall, the velocyto method with the constant velocity assumption performed the best overall.
-    \textbf{B:} An example bifurcating dataset, with as illustration the expression and ground truth velocity of a gene that goes up and down in the trajectory. 
-\textbf{C:} The RNA velocity estimates of the different methods.
-\textbf{D:} The embedded RNA velocity of the different methods.
+    \textbf{A:} An example bifurcating cycle dataset, with as illustration the expression and ground truth velocity of a gene D1\_TF1 that goes up and down in one branch of the trajectory. 
+\textbf{B:} The RNA velocity estimates of gene D1\_TF1 by the different methods.
+\textbf{C:} The velocity stream plots produced from the predictions of each method, as generated by scvelo.
+\textbf{D:} TODO: discuss new benchmark results
     }
@@
 For example, single-cell NI when applied to single-cell transcriptomics data; sample-specific NI when applied to bulk transcriptomics; patient-derived NI when applied to bulk profiles of patients.
-A more generaised variant of CSNI is casewise NI, which does not specify the type of data which is being analysed.} predict not only which transcription factors regulate which target genes (Figure\ \ref{fig:scgrn}A, top left), but also how active each interaction is in every case (Figure\ \ref{fig:scgrn}A).
+A more generalised variant of CSNI is casewise NI, which does not specify the type of data which is being analysed.} predict not only which transcription factors regulate which target genes (Figure\ \ref{fig:scgrn}A, top left), but also how active each interaction is in every case (Figure\ \ref{fig:scgrn}A).
 
 While a few pioneering CSNI approaches have already been developed [@aibar_scenicsinglecellregulatory_2017; @kuijjer_estimatingsamplespecificregulatory_2019; @liu_personalizedcharacterizationdiseases_2016], a quantitative assessment of the performance is until now lacking.
-This is not surprising, as neither real nor in silico datasets of cell-specific or even cell-type-specific interactions exists that is large enough so that it can be used as a ground-truth for evaluating CSNI methods.
+This is not surprising, as neither real nor in silico datasets of cell-specific or even cell-type-specific interactions exists that are large enough so that it can be used as a ground-truth for evaluating CSNI methods.
 Extracting the ground-truth dynamic network in dyngen is straightforward though, given that we can calculate how target gene expression would change without the regulator being present.
@@
     \centering
-    \includegraphics[width=.8\linewidth]{fig/network_inference/cni.pdf}
+    \includegraphics[width=.8\linewidth]{result_files/usecase_network_inference/cni.pdf}
     \caption{
@@
 
-Trajectory alignment allows studying the differences between the same trajectories from different samples.
+Trajectory alignment allows studying the differences between the same trajectories from different samples, as shown in Figure \ref{fig:traj_align} A.
 For example, the cell developmental process of a patient could be compared to that of a healthy control to detect the transcriptomic differences of a particular lineage.
@@
 Dynamic Time Warping (DTW) [@giorgino_computingvisualizingdynamic_2009] is most commonly used to align linear trajectories.
-DTW is a technique originating in the field of speech recognition and aligns temporal sequences by creating a warping path between the sequences that indicates which sequence must be dilated or contracted to best match the other one.
-DTW can also be used in combination with smoothed pseudocells [@alpert_alignmentsinglecelltrajectories_2018].
+DTW is a technique originating in the field of speech recognition and aligns temporal sequences by creating a warping path between the sequences that indicate which sequence must be dilated or contracted to best match the other one.
+DTW can also be used in combination with smoothed pseudo-cells [@alpert_alignmentsinglecelltrajectories_2018].
 
@@
 
-dyngen can be used to evaluate trajectory alignment techniques.
-We analyse how well DTW distinguishes between different trajectories for increasing amounts of noise.
-When noise increases, DTW can no longer distinguish trajectories clearly (Figure\ \ref{fig:traj_align}).
-In this case, there is no difference between when smoothed pseudocells are used or when actual cells and their gene expression are used.
+The pseudotime values produced by dyngen are comparable across experiments, which allows us to evaluate the accuracy of an alignment technique.
+We use DTW, but process the trajectories in 3 different ways: we either used the original cells, we used pseudocells with a smoothed gene expression, or we used every tenth cell in the trajectory.
+An example of the different results obtained is shown in \ref{fig:traj_align} B, C and D.
+We can see that the smoothed way of processing recovers the signal in the data best, it finds a diagonal warping path, which indicates that the two trajectories develop in a similar fashion.
+We test this on 10 pairs of two datasets, where an increasing amount of noise was added to the count matrix. As shown in Figure \ref{fig:traj_align} E, the smoothed pseudocells perform best.
 
@@
     \centering
-    \includegraphics[width=.8\linewidth]{fig/traj_alignment/tares.pdf}
+    \includegraphics[width=.8\linewidth]{result_files/usecase_trajectory_alignment/usecase.pdf}
     \caption{
     \textbf{dyngen allows benchmarking of trajectory alignment methods.} 
-  \textbf{A:} Example of trajectory alignment using DTW. The segments between the two trajectories show which cells are aligned.
-\textbf{B:} Shows the discrimination power of DTW decreasing as the amount of noise added increases.
+ \textbf{A}: Example of a trajectory alignment process. The pseudotimes of individual inferred trajectories can be aligned to indicate differences between the trajectories. 
+  \textbf{B, C, D}: Shows the accumulated distance matrices obtained after using DTW on two trajectories where noise (noise level of 0.4) was added to the count matrix.
+In B the complete count matrices were used to perform the alignment.
+In C, each 10th cell was used to perform the alignment.
+In D, smoothed pseudocells were used.
+  \textbf{E:} Shows the influence of added noise to the different processing methods. We can see that smoothing performs best in noisy circumstances.
     }
@@
 # Discussion {#sec:dyngen-discussion}
-As is, dyngen's single-cell simulations can be used to evaluate common single-cell omics computational methods such as clustering, batch correction, trajectory inference and network inference.
+As is, dyngen's single-cell simulations can be used to evaluate common single-cell omics computational methods such as clustering, batch correction, trajectory inference, and network inference.
 However, the combined effect of these advantages results in a framework that is flexible enough to adapt to a broad range of applications.
-This may include methods that integrate clustering, network inference and trajectory inference.
+This may include methods that integrate clustering, network inference, and trajectory inference.
 In this respect, dyngen may promote the development of new tools in the single-cell field similarly as other simulators have done in the past [@schaffter_genenetweaversilicobenchmark_2011; @ewing_combiningtumorgenome_2015].
@@
 Currently, dyngen focuses on simulating cells as standalone entities that are well mixed.
-Splitting up the simulation space into separate subvolumes could pave the way to better study key cellular processes such as cell division, intercellular communication and migration [@smith_spatialstochasticintracellular_2019].
+Splitting up the simulation space into separate subvolumes could pave the way to better study key cellular processes such as cell division, intercellular communication, and migration [@smith_spatialstochasticintracellular_2019].
 
@@
     \centering
-    \includegraphics[width=\textwidth]{fig/explain_methods}
-    \caption{\textbf{The workflow of dyngen is comprised of six main steps.} 
-    \textbf{A:} The user needs to specify the desired module network or use a predefined module network. 
-    \textbf{B:} Each gene in a module is is regulated by one or more transcription factors from the upstream module. Additional target genes are generated. 
+    \includegraphics[width=\textwidth]{result_files/explain_methods}
+    \caption{\textbf{The workflow of dyngen consists of six main steps.} 
+    \textbf{A:} The user needs to specify the desired module network or use a predefined module network. The module network is what determines the dynamic behaviour of simulated cells.
+    \textbf{B:} The number of desired transcription factors (which drive the desired dynamic process) are amongst the given modules and adds regulatory interactions according to the module network.
+Additional target genes (which do not influence the dynamic process) are added by sampling interactions from GRN interaction databases.
     \textbf{C:} Each gene regulatory interaction in the GRN is converted to a set of biochemical reactions. 
@@
 
-## Defining the backbone: modules and states {#sec:dyngen-backbone}
+## Defining the module network {#sec:dyngen-modules}
 
@@
 The exact way a cell chooses a certain path during its differentiation is still an active research field, although certain models have already emerged and been tested _in vivo_.
-One driver of bifurcation seems to be mutual antagonism, where two genes strongly repress each other [@rekhtman_directinteractionhematopoietic_1999; @xu_regulationbifurcatingcell_2015], forcing one of the two to become inactive [@graf_forcingcellschange_2009].
+One driver of bifurcation is mutual antagonism, where two genes strongly repress each other [@rekhtman_directinteractionhematopoietic_1999; @xu_regulationbifurcatingcell_2015], forcing one of the two to become inactive [@graf_forcingcellschange_2009].
 Such mutual antagonism can be modelled and simulated [@wang_quantifyingwaddingtonlandscape_2011; @ferrell_bistabilitybifurcationswaddington_2012].
@@
 
-To start a dyngen simulation, the user needs to define a module network and a backbone.
-The module network defines how sets of co-regulated genes, called modules, regulate each other.
-The module network is what mainly determines which dynamic processes occur within the simulated cells.
+To start a dyngen simulation, the user needs to define a module network.
+The module network describes how sets of genes regulate each other and is what mainly determines which dynamic processes occur within the simulated cells.
 
-A module network consists of modules connected together by regulatory interactions. 
+A module network consists of modules connected together by regulatory interactions, which can be either up- or down-regulating.
 A module may have basal expression, which means genes in this module will be transcribed without the presence of transcription factor molecules.
@@
 
-Several examples of module networks are given (Figure\ \ref{fig:example_backbones}).
+Several examples of module networks are given (Figure\ \ref{fig:example_backbones_onlymodules}).
 A simple chain of modules (where one module upregulates the next) results in a _linear_ process.
@@
     \centering
-    \includegraphics[width=0.8\textwidth]{fig/example_backbones}
+    \includegraphics[width=0.8\textwidth]{result_files/example_backbones_onlymodules}
     \caption{
-      \textbf{Example module networks.}
+      \textbf{The module network determines the type of dynamic process which simulated cells will undergo.} A module network describes the regulatory interactions between sets of transcription factors which drive the desired dynamic process.
     }
-    \label{fig:example_backbones}
+    \label{fig:example_backbones_onlymodules}
 \end{figure}
 
-### Backbone lego {#sec:dyngen-bbl}
-The backbone can make use of one or more "backbone lego" (BBL) pieces (Figure\ \ref{fig:backbone_lego}).
-A BBL consists of one or more modules which regulate each other such that the output modules present a specific behaviour, depending on the input module (Figure\ \ref{fig:backbone_lego}A).
-Parameters allow determining the number of modules involved in the process and the number of outputs.
-Multiple BBLs can be chained together in order to intuitively create module networks and corresponding state networks (Figure\ \ref{fig:backbone_lego}B).
-Note that not all dynamic processes can be represented by a combination of BBLs, but they can serve as common building blocks to aid the construction of the backbone.
+## Generating the gene regulatory network {#sec:dyngen-grn}
+The GRN is generated based on the given module network in four main steps (Figure\ \ref{fig:gen_feature_network}).
 
@@
     \centering
-    \includegraphics[width=0.8\textwidth]{fig/backbone_lego}
-    \caption{
-      \textbf{Reusable regulatory building blocks that can be used to construct GRNs that result in dynamic processes following an expected pattern.}
-      \textbf{A:} Each building block contains genes which serve as inputs that can be regulated by external transcription factors.
-The connections inside the building block transform the input signal in a particular way and has one or more output interactions of which the targets are not yet determined.
-      \textbf{B:} By connecting multiple building blocks together, a backbone of regulatory modules can be formed.
-      \textbf{C:} dyngen offers several variations of linear building blocks which allow for differing regulatory mechanisms.
-      \textbf{D:} A branching module will upregulate the expression of just one of the output interactions.
-Two variations are given of which the simple variant is more prone to reprogramming events in comparison to the robust branching module.
-      \textbf{E,F:} These components allow to specify the beginning (no inputs) and end (no outputs) of the backbone.
-    }
-    \label{fig:backbone_lego}
-\end{figure}
-
-When the input node of a **linear BBL** (Figure\ \ref{fig:backbone_lego}C) is upregulated, the module the BBL is connected to will be upregulated.
-A _simple chain_ is a set of modules where a module upregulates the next.
-A _chain with double repression_ has an uneven number of modules forming a chain where each module downregulates the next but all modules (except the input) have basal expression.
-A _grid with double repression_ is similar; except that modules do not have basal expression but instead get upregulated by an upstream module in the chain.
-Finally, a _flip flop_} consists of a simple chain where first the modules (except the last) are upregulated.
-Once the second to last module is upregulated, that module upregulates itself and the first module is strongly repressed, causing all other modules to lose expression and finally the last module to be upregulated.
-The _flip flop_ retains this output state, even when the input changes.
-
-When the input node of a **branching BBL** (Figure\ \ref{fig:backbone_lego}D) is upregulated, a subset of its output modules will eventually be upregulated.
-A \emph{simple branching} uses reciprocal inhibition to drive the upregulation of one of the output modules.
-Due to its simplicity, however, multiple output modules might be upregulated simultaneously and over long periods of simulation time it might be possible that the choice of upregulated module changes.
-A _robust branching_ improves upon the simple branching by preventing upregulation of output modules until an internal branching decision has been made, and by repressing the decision mechanism to avoid other output modules being upregulated other than the one that has been chosen.
-
-A **leaf BBL** (Figure\ \ref{fig:backbone_lego}E) is a linear BBL that has either no inputs or no outputs.
-A _start_ BBL is a linear BBL where the first module has basal expression, and all modules in this module will be active during the burn-in phase of the simulation.
-An _end_ BBL is also a linear BBL with its output regulating one final module.
-
-## Generating the gene regulatory network {#sec:dyngen-grn}
-The GRN is generated based on the given backbone in four main steps (Figure\ \ref{fig:gen_feature_network}).
-
-\begin{figure}[htb!]
-    \centering
-    \includegraphics[width=\linewidth]{fig/gen_feature_network}
+    \includegraphics[width=\linewidth]{result_files/gen_feature_network}
     \caption{
@@
 A user-defined number of target genes are added to the GRN.
-Target genes are regulated by a TF or another target gene, but is always downstream of at least one TF.
+Target genes are regulated by a TF or another target gene, but are always downstream of at least one TF.
 To sample the interactions between target genes, one of the many FANTOM5 \cite{lizio_gatewaysfantom5promoter_2015} GRNs is sampled.
@@
 For each target, at most $x$ regulators are sampled from the induced FANTOM5 GRN (default $x=5$).
-The interactions connecting a target gene and its regulators are added the GRN.
+The interactions connecting a target gene and its regulators are added to the GRN.
 
@@
 
-\newcommand{\w}[1]{\text{w}_{#1}}
 \newcommand{\x}[1]{\text{x}_{#1}}
 \newcommand{\y}[1]{\text{y}_{#1}}
+\newcommand{\z}[1]{\text{z}_{#1}}
 
@@
 
-\newcommand{\wpr}[1]{\text{wpr}_{#1}}
-\newcommand{\whl}[1]{\text{whl}_{#1}}
-\newcommand{\wdr}[1]{\text{wdr}_{#1}}
-\newcommand{\wsr}[1]{\text{wsr}_{#1}}
+\newcommand{\xpr}[1]{\text{xpr}_{#1}}
 \newcommand{\xhl}[1]{\text{xhl}_{#1}}
-\newcommand{\xdr}[1]{\text{xdr}_{#1}}
-\newcommand{\ypr}[1]{\text{ypr}_{#1}}
+\newcommand{\ysr}[1]{\text{ysr}_{#1}}
 \newcommand{\yhl}[1]{\text{yhl}_{#1}}
 \newcommand{\ydr}[1]{\text{ydr}_{#1}}
+\newcommand{\zpr}[1]{\text{zpr}_{#1}}
+\newcommand{\zhl}[1]{\text{zhl}_{#1}}
+\newcommand{\zdr}[1]{\text{zdr}_{#1}}
 
@@
 \newcommand{\dis}[1]{\text{dis}_{#1}}
-\newcommand{\buf}[1]{\chi_{#1}}
+\newcommand{\buf}[1]{\text{bind}_{#1}}
 \newcommand{\ba}[1]{\text{bas}_{#1}}
@@
 Simulating a cell's GRN makes use of a stochastic framework which tracks the abundance levels of molecules over time in a discrete quantity.
-For every gene $G$, the abundance levels of three molecules are tracked, namely of corresponding pre-mRNAs, mature mRNAs and proteins, which are represented by the terms $\w G$, $\x G$ and $\y G$ respectively.
+For every gene $G$, the abundance levels of three molecules are tracked, namely of corresponding pre-mRNAs, mature mRNAs and proteins, which are represented by the terms $\x G$, $\y G$ and $\z G$ respectively.
 The GRN defines how a reaction affects the abundance levels of molecules and how likely it will occur.
@@
 The effects of each reaction mimic the respective biological processes (Table\ \ref{tab:reaction_def}, middle).
-Transcription of gene $G$ results in the creation of a single pre-mRNA molecule $\w G$.
-Splicing turns one pre-mRNA $\w G$ into a mature mRNA $\w G$.
-Translation uses a mature mRNA $\x G$ to produce a protein $\y G$.
-Pre-mRNA, mRNA and protein degradation results in the removal of a $\w G$, $\x G$, and $\y G$ molecule, respectively.
+Transcription of gene $G$ results in the creation of a single pre-mRNA molecule $\x G$.
+Splicing turns one pre-mRNA $\x G$ into a mature mRNA $\x G$.
+Translation uses a mature mRNA $\y G$ to produce a protein $\z G$.
+Pre-mRNA, mRNA and protein degradation results in the removal of a $\x G$, $\y G$, and $\z G$ molecule, respectively.
 
-The propensity of all reactions except transcription are all linear functions (Table\ \ref{tab:reaction_def}, right) of the abundance level of some molecule multiplied by a constant drawn from a normal distribution (Table\ \ref{tab:reaction_params}).
+The propensity of all reactions except transcription are all linear functions (Table\ \ref{tab:reaction_def}, right) of the abundance level of some molecule multiplied by a per-gene constant (Table\ \ref{tab:reaction_params}).
 The propensity of transcription of a gene $G$ depends on the abundance levels of its TFs.
+The per-gene and per-interaction constants are based on the median reported production-rates and half-lives of molecules measured of 5000 mammalian genes [@schwanhausser_globalquantificationmammalian_2011], except that the transcription rate has been amplified by a factor of 10.
 
@@
     \caption{
-      \textbf{Reactions affecting the abundance levels of pre-mRNA $\w G$, mature mRNA $\x G$ and proteins $\y G$ of gene $G$.} Define the set of regulators of $G$ as $\rs{G}$, the set of upregulating regulators of $G$ as $\rp G$, and the set of downregulating regulators of $G$ as $\rn G$.
+      \textbf{Reactions affecting the abundance levels of pre-mRNA $\x G$, mature mRNA $\y G$ and proteins $\z G$ of gene $G$.} Define the set of regulators of $G$ as $\rs{G}$, the set of upregulating regulators of $G$ as $\rp G$, and the set of downregulating regulators of $G$ as $\rn G$.
 Parameters used in the propensity formulae are defined in Table \ref{tab:reaction_params}.
@@
         Reaction & Effect & Propensity \\ \hline \hline
-        Transcription & $\emptyset \rightarrow \w G$ & $\wpr G \times \frac{\ba G - \ind{G}^{|\rp{G}|} + \prod\limits_{H \in \rp{G}}(\ind G + \buf{G,H})}{\prod\limits_{H \in \rs{G}}(1 + \buf{G,H})}$ \\
-        Pre-mRNA degradation & $\w G \rightarrow \emptyset$ & $\wdr G \times \w G$ \\
-        Splicing & $\w G \rightarrow \x G$ & $\wsr G \times \w G$ \\
-        Mature mRNA degradation & $\x G \rightarrow \emptyset$ & $\xdr G \times \x G$ \\
-        Translation & $\x G \rightarrow \x G + \y G$ & $\ypr G \times \x G$ \\
-        Protein degradation & $\y G \rightarrow \emptyset$ & $\ydr G \times \y G$ \\ \hline
+        Transcription & $\emptyset \rightarrow \x G$ & $\xpr G \times \frac{\ba G - \ind{G}^{|\rp{G}|} + \prod\limits_{H \in \rp{G}}(\ind G + \buf{G,H})}{\prod\limits_{H \in \rs{G}}(1 + \buf{G,H})}$ \\
+        Splicing & $\x G \rightarrow \y G$ & $\ysr G \times \x G$ \\
+        Translation & $\y G \rightarrow \y G + \z G$ & $\zpr G \times \y G$ \\ \hline\hline
+        Pre-mRNA degradation & $\x G \rightarrow \emptyset$ & $\ydr G \times \x G$ \\
+        Mature mRNA degradation & $\y G \rightarrow \emptyset$ & $\ydr G \times \y G$ \\
+        Protein degradation & $\z G \rightarrow \emptyset$ & $\zdr G \times \z G$ \\ \hline
     \end{tabular}
@@
         Parameter & Symbol & Definition \\ \hline \hline
-        Transcription rate & $\wpr{G}$ & $\in N(50, 10),\ \geq 10$ \\
-        Splicing rate & $\wsr G$ & $\in N(5, 1),\ \geq 1$ \\
-        Translation rate & $\ypr{G}$ & $\in N(5, 1),\ \geq 1$ \\
-        Pre-mRNA half-life & $\whl{G}$ & $\in N(0.15, 0.03),\ \geq 0.05$ \\
-        Mature mRNA half-life & $\xhl{G}$ & $\in N(0.15, 0.03),\ \geq 0.05$ \\
-        Protein half-life rate & $\yhl G$ & $\in N(0.25, 0.05),\ \geq 0.1$ \\
+        Transcription rate & $\xpr{G}$ & $\in U(10, 20)$ \\
+        Splicing rate & $\ysr G$ & $= \ln(2)\ /\ 2$ \\
+        Translation rate & $\zpr{G}$ & $\in U(100, 150)$ \\
+        (Pre-)mRNA half-life & $\yhl{G}$ & $\in U(2.5, 5)$ \\
+        Protein half-life & $\zhl G$ & $\in U(5, 10)$ \\
         Interaction strength & $\str{G,H}$ & $\in 10^{U(0, 2)}$ * \\
         Hill coefficient & $\hill{G,H}$ & $\in U(0.5, 2)$ * \\
-        Independence factor & $\ind G$ & $\in [0, 1]$ * \\ \hline\hline
-        Pre-mRNA degradation rate & $\wdr G$ & $= \ln(2)\ /\ \whl G$ \\
-        Mature mRNA degradation rate & $\xdr G$ & $= \ln(2)\ /\ \xhl G$ \\
-        Protein degradation rate & $\ydr G$ & $= \ln(2)\ /\ \yhl G$ \\
-        Dissociation constant & $\dis H$ & $= 0.5 \times \frac{\wpr H \times \wsr H \times \ypr H}{(\wdr H + \wsr H) \times \xdr H \times \ydr H}$ \\
-        Binding & $\buf{G,H}$ & $= \str{G,H} \times \left(\y H\ /\ \dis H\right) ^ {\hill{G,H}}$ \\
+        Independence factor & $\ind G$ & $\in U(0, 1)$ * \\ \hline\hline
+        (Pre-)mRNA degradation rate & $\ydr G$ & $= \ln(2)\ /\ \yhl G$ \\
+        Protein degradation rate & $\zdr G$ & $= \ln(2)\ /\ \zhl G$ \\
+        Dissociation constant & $\dis H$ & $= 0.5 \times \frac{\xpr H \times \ysr H \times \zpr H}{(\ydr H + \ysr H) \times \ydr H \times \zdr H}$ \\
+        Binding strength & $\buf{G,H}$ & $= \str{G,H} \times \left(\z H\ /\ \dis H\right) ^ {\hill{G,H}}$ \\
         Basal expression & $\ba G$ & $= \begin{cases} 1 & \mbox{if } \rp{G} = \emptyset \\ 0.0001 & \mbox{if } \rn{G} = \emptyset \mbox{ and } \rp{G} \neq \emptyset \\ 0.5 & \mbox{otherwise} \end{cases}$ * \\ \hline
@@
 \newcommand{\ai}[2]{$S_{#1} = S_{#2b}$}
-\newcommand{\yk}[1]{\frac{y_#1}{k_#1}^{c_#1}}
+\newcommand{\zk}[1]{\frac{y_#1}{k_#1}^{c_#1}}
 \newcommand{\wi}[1]{\nu_#1}
@@
 The propensity of the transcription of a gene $G$ is inspired by thermodynamic models of gene regulation [@schilstra_biologicgeneexpression_2008], in which the promoter of $G$ can be bound or unbound by a set of $N$ transcription factors $H_i$.
-Let $\proptran(\y 1, \y 2, \ldots, \y N)$ denote the propensity function of $G$, in function of the abundance levels of the transcription factors.
+Let $\proptran(\z 1, \z 2, \ldots, \z N)$ denote the propensity function of $G$, in function of the abundance levels of the transcription factors.
 The following subsections explain and define the propensity function when $N=1$, $N=2$, and finally for an arbitrary $N$.
@@
 \begin{align}
-  f(y_1, y_2, \ldots, y_N) & = \text{wpr} \cdot \sum_{j = 0}^{2^N - 1} \alpha_j \cdot P(S_j) \label{eqn:activ0} \\
+  f(y_1, y_2, \ldots, y_N) & = \text{xpr} \cdot \sum_{j = 0}^{2^N - 1} \alpha_j \cdot P(S_j) \label{eqn:activ0} \\
 \end{align}
@@
 \begin{align}
-f(y_1) & = \text{wpr} \cdot \left(\alpha_0 \cdot P(S_0) + \alpha_1 \cdot P(S_1)\right) \\
-   & = \text{wpr} \cdot \left(\alpha_0 \cdot \frac{1}{1 + \nu_1} + \alpha_1 \cdot \frac{\nu_1}{1 + \nu_1}\right) \\
-   & = \text{wpr} \cdot \frac{\alpha_0 + \alpha_1 \cdot \nu_1}{1 + \nu_1} \\
+f(y_1) & = \text{xpr} \cdot \left(\alpha_0 \cdot P(S_0) + \alpha_1 \cdot P(S_1)\right) \\
+   & = \text{xpr} \cdot \left(\alpha_0 \cdot \frac{1}{1 + \nu_1} + \alpha_1 \cdot \frac{\nu_1}{1 + \nu_1}\right) \\
+   & = \text{xpr} \cdot \frac{\alpha_0 + \alpha_1 \cdot \nu_1}{1 + \nu_1} \\
 \end{align}
@@
 \begin{align}
-f(y_1, y_2) & = \text{wpr} \cdot \sum_{j = 0}^{2^N - 1} \alpha_j \cdot P(S_j) \\
- & = \text{wpr} \cdot \frac{\sum_{j = 0}^{2^N - 1} \alpha_j \cdot w_j}{\prod_{i=1}^{i \leq N} (\nu_i + 1)} \\
- & = \text{wpr} \cdot \frac{\alpha_0 + \alpha_1 \cdot \nu_1 + \alpha_2 \cdot \nu_2 + \alpha_3 \cdot \nu_1 \cdot \nu_2}{(\nu_1 + 1) \cdot (\nu_2 + 1)} \\
+f(y_1, y_2) & = \text{xpr} \cdot \sum_{j = 0}^{2^N - 1} \alpha_j \cdot P(S_j) \\
+ & = \text{xpr} \cdot \frac{\sum_{j = 0}^{2^N - 1} \alpha_j \cdot w_j}{\prod_{i=1}^{i \leq N} (\nu_i + 1)} \\
+ & = \text{xpr} \cdot \frac{\alpha_0 + \alpha_1 \cdot \nu_1 + \alpha_2 \cdot \nu_2 + \alpha_3 \cdot \nu_1 \cdot \nu_2}{(\nu_1 + 1) \cdot (\nu_2 + 1)} \\
 \end{align}
@@
 \begin{align}
-f(y_1, y_2, \ldots, y_N) & = \text{wpr} \cdot \sum_{j = 0}^{2^N - 1} \alpha_j \cdot P(S_j) \\
-& = \text{wpr} \cdot \frac{\sum_{j = 0}^{2^N - 1} \alpha_j \cdot w_j}{\prod_{i=1}^{i \leq N} (\nu_i + 1)} \label{eqn:prop2n}
+f(y_1, y_2, \ldots, y_N) & = \text{xpr} \cdot \sum_{j = 0}^{2^N - 1} \alpha_j \cdot P(S_j) \\
+& = \text{xpr} \cdot \frac{\sum_{j = 0}^{2^N - 1} \alpha_j \cdot w_j}{\prod_{i=1}^{i \leq N} (\nu_i + 1)} \label{eqn:prop2n}
 \end{align}
@@
 \begin{align}
-f(y_1, y_2, \ldots, y_N) & = \text{wpr} \cdot \frac{\prod_{i \in R^+} (\nu_i + 1)}{\prod_{i \in R} (\nu_i + 1)}
+f(y_1, y_2, \ldots, y_N) & = \text{xpr} \cdot \frac{\prod_{i \in R^+} (\nu_i + 1)}{\prod_{i \in R} (\nu_i + 1)}
 \end{align}
@@
 \begin{align}
-f(y_1, y_2, \ldots, y_N) & = \text{wpr} \cdot \frac{\text{ba} - \text{sy}^{|R^+|} + \prod_{i \in R^+} (\nu_i + \text{sy})}{\prod_{i \in R} (\nu_i + 1)}
+f(y_1, y_2, \ldots, y_N) & = \text{xpr} \cdot \frac{\text{ba} - \text{sy}^{|R^+|} + \prod_{i \in R^+} (\nu_i + \text{sy})}{\prod_{i \in R} (\nu_i + 1)}
 \end{align}
@@
     \centering
-    \includegraphics[width=.6\linewidth]{fig/sample_cells.pdf}
+    \includegraphics[width=.6\linewidth]{result_files/sample_cells.pdf}
     \caption{
@@
 We assume the root state has no modules active and has no expression of any molecules.
-To transition to a next state, we activate or deactivate certain gene modules as defined by the backbone (Section\ \ref{sec:dyngen-backbone}) and perform a noiseless ODE simulation.
+To transition to a next state, we activate or deactivate certain gene modules as defined by the backbone (Section\ \ref{sec:dyngen-modules}) and perform a noiseless ODE simulation.
 For each time interval $t=0.01$ along the trajectory, the expression values are retained.
+
+## Constructing backbones {#sec:dyngen-backbone}
+
+\begin{figure}[htb!]
+    \centering
+    \includegraphics[width=0.8\textwidth]{result_files/example_backbones}
+    \caption{
+      TODO
+    }
+    \label{fig:example_backbones}
+\end{figure}
+
+### Backbone lego {#sec:dyngen-bbl}
+The backbone can make use of one or more "backbone lego" (BBL) pieces (Figure\ \ref{fig:backbone_lego}).
+A BBL consists of one or more modules which regulate each other such that the output modules present a specific behaviour, depending on the input module (Figure\ \ref{fig:backbone_lego}A).
+Parameters allow determining the number of modules involved in the process and the number of outputs.
+Multiple BBLs can be chained together in order to intuitively create module networks and corresponding state networks (Figure\ \ref{fig:backbone_lego}B).
+Note that not all dynamic processes can be represented by a combination of BBLs, but they can serve as common building blocks to aid the construction of the backbone.
+
+\begin{figure}[htb!]
+    \centering
+    \includegraphics[width=0.8\textwidth]{result_files/backbone_lego}
+    \caption{
+      \textbf{Reusable regulatory building blocks that can be used to construct GRNs that result in dynamic processes following an expected pattern.}
+      \textbf{A:} Each building block contains genes which serve as inputs that can be regulated by external transcription factors.
+The connections inside the building block transform the input signal in a particular way and has one or more output interactions of which the targets are not yet determined.
+      \textbf{B:} By connecting multiple building blocks together, a backbone of regulatory modules can be formed.
+      \textbf{C:} dyngen offers several variations of linear building blocks which allow for differing regulatory mechanisms.
+      \textbf{D:} A branching module will upregulate the expression of just one of the output interactions.
+Two variations are given of which the simple variant is more prone to reprogramming events in comparison to the robust branching module.
+      \textbf{E,F:} These components allow to specify the beginning (no inputs) and end (no outputs) of the backbone.
+    }
+    \label{fig:backbone_lego}
+\end{figure}
+
+When the input node of a **linear BBL** (Figure\ \ref{fig:backbone_lego}C) is upregulated, the module the BBL is connected to will be upregulated.
+A _simple chain_ is a set of modules where a module upregulates the next.
+A _chain with double repression_ has an uneven number of modules forming a chain where each module downregulates the next but all modules (except the input) have basal expression.
+A _grid with double repression_ is similar; except that modules do not have basal expression but instead get upregulated by an upstream module in the chain.
+Finally, a _flip flop_} consists of a simple chain where first the modules (except the last) are upregulated.
+Once the second to last module is upregulated, that module upregulates itself and the first module is strongly repressed, causing all other modules to lose expression and finally the last module to be upregulated.
+The _flip flop_ retains this output state, even when the input changes.
+
+When the input node of a **branching BBL** (Figure\ \ref{fig:backbone_lego}D) is upregulated, a subset of its output modules will eventually be upregulated.
+A \emph{simple branching} uses reciprocal inhibition to drive the upregulation of one of the output modules.
+Due to its simplicity, however, multiple output modules might be upregulated simultaneously and over long periods of simulation time it might be possible that the choice of upregulated module changes.
+A _robust branching_ improves upon the simple branching by preventing upregulation of output modules until an internal branching decision has been made, and by repressing the decision mechanism to avoid other output modules being upregulated other than the one that has been chosen.
+
+A **leaf BBL** (Figure\ \ref{fig:backbone_lego}E) is a linear BBL that has either no inputs or no outputs.
+A _start_ BBL is a linear BBL where the first module has basal expression, and all modules in this module will be active during the burn-in phase of the simulation.
+An _end_ BBL is also a linear BBL with its output regulating one final module.
 
@@
 \begin{eqnarray*}
-  \text{regeffect}_G & = \frac{\text{proptrans}_G(S) - \text{proptrans}_G(S[\y T \leftarrow 0])}{\wpr G}
+  \text{regeffect}_G & = \frac{\text{proptrans}_G(S) - \text{proptrans}_G(S[\z T \leftarrow 0])}{\xpr G}
 \end{eqnarray*}
@@
     \centering
-    \includegraphics[width=.8\linewidth]{fig/small_bifurc/scgrn.pdf}
+    \includegraphics[width=.8\linewidth]{result_files/fig1_showcase_functionality/scgrn.pdf}
     \caption{
@@
 
-LIONESS [@@kuijjer_estimatingsamplespecificregulatory_2019; @kuijjer_lionessrsinglesample_2019] uses the Pearson correlation to infer cell-specific GRNs.
-To do so, first the Pearson correlation is calculated between regulators and targets for all samples.
-Next, the Pearson correlation is again calculated for all samples excluding one sample.
-The difference between the two correlation matrices is considered a cell-specific GRN for that particular profile.
+LIONESS [@kuijjer_estimatingsamplespecificregulatory_2019] runs a NI method multiple times to construct cell-specific GRNs.
+LIONESS first infers a GRN with all of the samples.
+A second GRN is inferred with all samples except one particular profile.
+The cell-specific GRN for that particular profile is defined as the difference between the two GRN matrices.
 This process is repeated for all profiles, resulting in a cell-specific GRN.
+By default, LIONESS uses PANDA [@glass_passingmessagesbiological_2013] to infer GRNs, but since dyngen does not produce motif data and motif data is required by PANDA, PANDA is inapplicable in this context.
+Instead, we used the lionessR [@kuijjer_lionessrsinglesample_2019] implementation of LIONESS, which uses by default the Pearson correlation as a NI method.
+We marked results from this implementation as "LIONESS + Pearson".
 
-SSN [@liu_personalizedcharacterizationdiseases_2016] has, in essence, the exact same methodology as LIONESS.
+SSN [@liu_personalizedcharacterizationdiseases_2016] follows, in essence, the exact same methodology as LIONESS except that it specifically only uses the Pearson correlation.
 It is worth noting that the LIONESS preprint was released before the publication of SSN.
@@
 
-2 datasets containing a trajectory were generated using dyngen.
-For each dataset, we shuffled 5\% to 75\% of the count matrix, with steps of 5\%.
-Next, we aligned (using DTW) all the base trajectories with all the noisy trajectories.
-
-Let the normalized alignment distance between base trajectory x and trajectory y with z\% noise be $dxy_z$.
-To calculate how well an alignment algorithm can distinguish the two trajectories, we define the discrimination power of DTW.
-
+We generated 10 base datasets that contain a linear trajectory using dyngen.
+For each dataset, we performed the cell generation and experiment generation twice.
+This results in 20 datasets, with 10 pairs of similar trajectories.
+For each of these pairs of datasets, we generated 10 progressively noisier pairs, in which we added noise to the complete count matrix.
+Let $q$ be the 75th quantile of the non-zero values in the count matrix. The noise added to the count matrix is calculated as follows, with $i$ being the noise parameter.
 \begin{equation}
-    \frac{d12_{z}}{d11_{z}} \quad \forall z \in \{0.05, 0.1, 0.15, ..., 0.75 \}
+ countmatrix + \frac{1}{\sqrt{2\pi}}qie^{\frac{x^2}{2(qi)^2}}  \quad \forall i \in \{0.1, 0.2, ... 1.0\}
 \end{equation}
 
-\begin{equation}
-    \frac{d21_{z}}{d22_{z}} \quad \forall z \in \{0.05, 0.1, 0.15, ..., 0.75 \}
-\end{equation}
+We aligned each of these 150 pairs of trajectories (the first trajectory in the pair with the second trajectory in the pair) in 3 different ways.
+Either we used the original count matrix, we used 1/10th of the cells (according to the ordering of the trajectory, so the 1st, 10th, 20th, ...
+cell of the trajectory) or we used the smoothed pseudocells.
+The pseudocells are generated at regular points across the pseudotime of the trajectory. The gene expression of each pseudocell is inferred using a Gaussian kernel with a window size of 0.05.
 
-If this ratio is high, it means that DTW is able to distinguish well between the two trajectories. If this ratio is 1, DTW is not able to distinguish between the two trajectories at all.
-
-These datasets and ratios were generated 10 times.
+Each of the cells or pseudocells received a pseudotime from dyngen.
+This time can be compared: cells from the first trajectory in the pair should be aligned closely to cells from the second trajectory with similar pseudotimes.
+We used the sum of differences between the pseudotimes of the aligned cells to determine the distance between the two trajectories, which should ideally be 0.
+In order to compare the alignments with 1000 cells per trajectory and the ones with 100 cells per trajectory, we divide the distance of those alignments by 10.
 

### 2020-05-18 render_manuscript.R

@@ -10,7 +10,7 @@ if (Sys.info()[["user"]] == "rcannood") {
 
 # download as docx instead of txt, because otherwise comments get pushed into the document
 httr::set_config(httr::config(http_version = 0)) # avoid http2 framing layer bug
-drive <- drive_download(as_id("1BOltsUJBQ8NhF5s9kU_Q470U0U00QblbG8HBgQwEzRE"), type="docx", overwrite=TRUE, path = tempfile())
+drive <- drive_download(as_id("15ZkzzB-XSYFfZutO4YPkUIw6iUvxJS92fSPfzqR1cLA"), type = "docx", overwrite = TRUE, path = tempfile())
 
 # read docx
 textreadr::read_docx(drive$local_path, remove.empty = FALSE, trim = FALSE) %>%

### 2020-06-01 analysis/summary/script.R
update all plots

@@ -1,0 +1,145 @@
+library(tidyverse)
+library(dyngen)
+library(dyngen.manuscript)
+library(dynplot)
+library(patchwork)
+
+set.seed(1)
+
+exp <- start_analysis("summary")
+
+
+# FIRST ROW: illustrate dyngen pipeline -----------------------------------
+model1 <- exp$result("model_branching.rds") %cache% {
+  backbone <- bblego(
+    bblego_start("A", type = "simple", num_modules = 1),
+    bblego_linear("A", "B", type = "simple", num_modules = 3),
+    bblego_branching("B", c("C", "D"), num_steps = 1),
+    bblego_end("C", type = "simple", num_modules = 2),
+    bblego_end("D", type = "simple", num_modules = 2)
+  )
+  model1 <-
+    initialise_model(
+      num_tfs = 12,
+      num_targets = 30,
+      num_hks = 15,
+      num_cells = 150,
+      backbone = backbone,
+      verbose = TRUE,
+      download_cache_dir = "~/.cache/dyngen",
+      num_cores = 8,
+      distance_metric = "euclidean",
+      simulation_params = simulation_default(
+        kinetics_noise_function = kinetics_noise_none(),
+        experiment_params = simulation_type_wild_type(num_simulations = 8),
+        compute_cellwise_grn = TRUE,
+        compute_rna_velocity = TRUE
+      )
+    ) %>%
+    generate_tf_network() %>%
+    generate_feature_network() %>%
+    generate_kinetics() %>%
+    generate_gold_standard() %>%
+    generate_cells() %>%
+    generate_experiment()
+
+  model1
+}
+
+theme_dr <- function() {
+  theme_bw() +
+    theme(
+      legend.position = "none",
+      axis.line = element_blank(),
+      axis.ticks = element_blank(),
+      axis.title = element_blank(),
+      axis.text = element_blank(),
+      panel.grid = element_blank(),
+      panel.border = element_blank()
+    )
+}
+
+# feature network
+g1a <-
+  plot_feature_network(model1, show_targets = TRUE) +
+  theme_dr()
+
+# simulation expression
+molecules <- model1$feature_info %>% filter(is_tf) %>% gather(mol, val, mol_mrna) %>% pull(val)
+df <- bind_cols(
+  model1$simulations$meta,
[…]

### 2020-06-01 analysis/usecase_network_inference/3_plot_dataset_example.R

@@ -95,3 +95,18 @@ g
 ggsave(exp$result("casewise_grn.pdf"), g, width = 8, height = 5)
 write_rds(g, exp$result("casewise_grn.rds"), compress = "gz")
 
+
+gtiny <- ggplot() +
+  geom_point(aes(x, y), colour = "gray", node_df) +
+  geom_segment(aes(x = x, y = y, xend = xend, yend = yend, colour = effect), edge_df %>% filter(effect >= 0, group == "GRN of cell 1"), arrow = arrow_up) +
+  geom_segment(aes(x = x, y = y, xend = xend, yend = yend, colour = effect), edge_df %>% filter(effect < 0, group == "GRN of cell 1"), arrow = arrow_down) +
+  theme_graph(base_family = "Helvetica") +
+  coord_equal() +
+  labs(colour = "Regulatory\nactivity") +
+  scale_colour_gradientn(colours = c(rev(RColorBrewer::brewer.pal(5, "Reds")), RColorBrewer::brewer.pal(5, "Greens"))) +
+  theme(
+    plot.margin = margin(0, 0, 0, 0, "cm")
+  )
+write_rds(gtiny, exp$result("cell1.rds"), compress = "gz")
+
+

### 2020-06-01 analysis/usecase_rna_velocity/4_plot_usecases.R

@@ -301,3 +301,4 @@ g <- patchwork::wrap_plots(
 )
 ggsave(exp$result("usecase.pdf"), g, height = 12, width = 12, useDingbats = FALSE)
 ggsave(exp$result("usecase.png"), g, height = 12, width = 12)
+write_rds(g[[3]][[1]] + labs(tag = NULL), exp$result("one_rna_velocity.rds"), compress = "gz")

### 2020-06-01 manuscript.Rmd
update

@@
 # Abstract
-**Purpose:** When developing new types of tools for single-cell analyses, there is often a lack of datasets on which to quantitatively assess the performance.  
-**Results:** We developed dyngen, a multi-modality simulator of single cells.
-In dyngen, the biomolecular state of an _in silico_ cell changes over time according to a predefined gene regulatory network.
-We used dyngen to benchmark three emerging ways of analysing single-cell data: RNA velocity, cell-specific network inference and trajectory alignment methods.
-**Conclusion:** dyngen lays the foundations for benchmarking a wide variety of computational single-cell tools and can be used to help kick-start the development of future types of analyses.
+**Purpose:** A recent breakthrough in single-cell omics is the ability to perform multi-modal measurements at single-cell level.
+However, in many cases insufficient metadata is available to accurately benchmark new types of algorithms.
+Existing generators of scRNA-seq are valuable in benchmarking novel computational tools more rigorously, but adding additional modalities or experimental conditions usually requires significant methodological alterations.
+**Results:** We introduce dyngen, a simulator of synthetic single cells by simulating gene regulation, splicing and translation at a single molecule level.
+From a simulation many layers of information can be extracted, including molecule abundance, progression along a dynamic process, activation strength of individual regulatory interactions, and a ground-truth RNA velocity.
+We demonstrate dyngen by showcasing its use on three novel computational approaches where no benchmarking of tools have been described for up to date: RNA velocity, cell-specific gene regulatory network inference and trajectory alignment methods.
+In itself, each of these three application cases constitutes a novel result that has not yet been described in the literature so far, and which we believe will spearhead research on development of novel computational tools for single-cell data.
+**Conclusion:** Ultimately, dyngen allows anticipating technological developments in single-cell multi-omics.
+It is thus possible to design and evaluate the performance and robustness of new types of computational analyses before experimental data becomes available.
+In addition, dyngen could also be used to compare which experimental protocol is the most cost-effective in producing qualitative and robust results in downstream analysis.
 
@@
 This confronts method developers with several challenges.
-The majority of the 250+ peer-reviewed computational tools for analysing single-cell omics data were published without a quantitative assessment of the accuracy of the tool.
+The majority of the >250 peer-reviewed computational tools for analysing single-cell omics data were published without a quantitative assessment of the accuracy of the tool.
 This is partially due to low availability of suitable benchmarking datasets; even if there are sufficient suitable input datasets available, these are often not accompanied by the necessary metadata to serve as ground-truth for a benchmark.
@@
 Generators of scRNA-seq data (e.g.
-splatter [@zappia_splattersimulationsinglecell_2017], powsimR [@vieth_powsimrpoweranalysis_2017], PROSSTT [@papadopoulos_prossttprobabilisticsimulation_2018] and SymSim [@zhang_simulatingmultiplefaceted_2019]) have already been widely used to explore the strengths and weaknesses of computational tools, both by method developers [@street_slingshotcelllineage_2018; @parra_reconstructingcomplexlineage_2018; @lummertzdarocha_reconstructioncomplexsinglecell_2018; @lin_scclassifyhierarchicalclassification_2019] and independent benchmarkers [@duo_systematicperformanceevaluation_2018; @saelens_comparisonsinglecelltrajectory_2019; @soneson_biasrobustnessscalability_2018].
+splatter [@zappia_splattersimulationsinglecell_2017], powsimR [@vieth_powsimrpoweranalysis_2017], PROSSTT [@papadopoulos_prossttprobabilisticsimulation_2018] and SymSim [@zhang_simulatingmultiplefaceted_2019]) have already been used extensively to explore the strengths and weaknesses of computational tools, both by method developers [@street_slingshotcelllineage_2018; @parra_reconstructingcomplexlineage_2018; @lummertzdarocha_reconstructioncomplexsinglecell_2018; @lin_scclassifyhierarchicalclassification_2019] and independent benchmarkers [@duo_systematicperformanceevaluation_2018; @saelens_comparisonsinglecelltrajectory_2019; @soneson_biasrobustnessscalability_2018].
 However, a limitation of existing scRNA-seq profiles generators is that they would require significant methodological alterations to add additional modalities or experimental conditions.
-
-<!--
-\newcommand{\yes}{\ding{51}}
-\newcommand{\no}{}
-\newcommand{\grayline}{\arrayrulecolor{lightgray}\cline{3-7}}
-\newcommand{\blackline}{\arrayrulecolor{black}\cline{3-7}}
-\newcommand{\method}[1]{#1}
-
-\begin{table}[htb!]
-    \centering
-    \begin{tabular}{p{.25cm}l|*{5}{>{\centering\arraybackslash}p{1.25cm}|}}
-    \blackline
-    & \method{} & \method{splatter} & \method{powsimR} & \method{PROSSTT} & \method{SymSim} & \method{dyngen} \\
-    & & \cite{zappia_splattersimulationsinglecell_2017} & \cite{vieth_powsimrpoweranalysis_2017} & \cite{papadopoulos_prossttprobabilisticsimulation_2018} & \cite{zhang_simulatingmultiplefaceted_2019} & \\
-    \blackline
-    \multicolumn{7}{l}{\textbf{Available modality outputs}} \\
-    \blackline
-    - & mRNA expression & \yes & \yes & \yes & \yes & \yes \\ \grayline
-    - & Pre-mRNA expression & \no & \no & \no & \no & \yes \\ \grayline
-    - & Protein expression & \no & \no & \no & \no & \yes \\ \grayline
-    - & Promotor activity & \no & \no & \no & \yes & \yes \\ \grayline
-    - & Reaction activity & \no & \no & \no & \no & \yes \\
-    \blackline
-    \multicolumn{7}{l}{\textbf{Available ground-truth outputs}} \\
-    \blackline
-    - & True counts & \yes & \no & \no & \yes & \yes \\ \grayline
-    - & Cluster labels & \yes & \yes & \no & \yes & ? \\ \grayline
-    - & Trajectory & \yes & \no & \yes $ ^1$ & \yes & \yes \\ \grayline
-    - & Batch labels & \yes & \no & \no & \no & \yes $ ^1$ \\ \grayline
-    - & Differential expression & \no & \no & \no & \yes & \yes $^2$ \\ %\grayline
-    - & Knocked down regulators & \no & \no & \no & \no & \yes \\ \grayline
-- & Regulatory network & \no & \no & \no & \yes & \yes \\ \grayline
-- & Cell-specific regulatory network & \no & \no & \no & \no & \yes \\
-    \blackline
-    \multicolumn{7}{l}{\textbf{Emulate experimental effects}} \\
-    \blackline
-    - & Single-cell RNA sequencing & \yes & \yes & \yes & \yes & \yes \\ \grayline
-    - & Batch effects & \yes & \no & \no & \no & \yes $ ^1$ \\ \grayline
-    - & Knockdown experiment & \no & \no & \no & \no & \yes \\ \grayline
-    - & Time-series & ? & ? & ? & ? & \yes \\
-    - & Snapshot & ? & ? & ? & ? & \yes \\
-    \blackline
-    \multicolumn{7}{l}{\textbf{Possible applications}} \\
-    \blackline
-    - & Clustering & \yes & \yes & \no & \yes & ? \\ \grayline
-    - & Trajectory inference & \yes & \no & \yes & \yes & \yes \\ \grayline
-    - & Network inference & \no & \no & \no & \yes & \yes \\ \grayline
-    - & Cell-specific network inference & \no & \no & \no & \no & \yes \\ \grayline
-    - & Differential expression & \yes & \no & \no & \no & \yes $^2$ \\ \grayline
-    - & Batch effect correction & \yes & \no & \no & \no & \yes \\ \grayline
-    - & RNA Velocity & \no & \no & \no & \no & \yes \\ \grayline
-    - & Trajectory alignment & \no & \no & \no & \no & \yes \\
-    \blackline
-    \multicolumn{7}{l}{\textbf{Showcased dataset}} \\
-    \blackline
-    - & Number of cells &  &  &  &  & \\
-    - & Number of genes &  &  &  &  & \\
-    - & Execution time & & & & & \\
-    \blackline
-    \end{tabular}
-    \caption{
-    \textbf{Comparison of single-cell simulators.}
-    $ ^1$ Showcased by vignette. $^2$ Showcased by Van den Berge et al.  \cite{vandenberge_trajectorybaseddifferentialexpression_2019}.
-    }
-    \label{tab:compare}
-\end{table}
--->
 
@@
 
-We developed a multi-modality simulator of single cells called dyngen (Figure\ \ref{fig:showcase}A).
-dyngen uses an optimized version of Gillespie's stochastic simulation algorithm [@gillespie_exactstochasticsimulation_1977]  to simulate gene regulation, splicing, and translation at a single-molecule level.
+We developed a multi-modality simulator of single cells called dyngen (Figure\ \ref{fig:showcase}).
+dyngen uses an optimised version of Gillespie's stochastic simulation algorithm [@gillespie_exactstochasticsimulation_1977]  to simulate gene regulation, splicing, and translation at a single-molecule level.
 Its methodology allows tracking of many layers of information throughout the simulation, including the abundance of any molecule in the cell, the progression of the cell along a dynamic process, and the activation strength of individual regulatory interactions.
@@
     \centering
-    \includegraphics[width=\linewidth]{result_files/fig1_showcase_functionality/overview_1.pdf}
+    \includegraphics[width=\linewidth]{result_files/summary/overview.pdf}
     \caption{
         \textbf{Showcase of dyngen functionality.}
-        \textbf{A:} The typical process of generating a dataset with dyngen.
-        \textbf{B:} Evaluating different types of computational tools requires simulating different types of experiments and extracting different layers of information from the simulation.
+        \textbf{A:} The typical process of generating a bifurcating dataset with dyngen.
+Step 1: generating a gene regulatory network.
+Step 2: simulating changes in expression over time.
+Step 3: combining multiple simulations and mapping to the ground-truth trajectory.
+Step 4: sampling cells and molecules.
+        \textbf{B:} dyngen includes different pre-defined trajectories, including converging, cyclic, bifurcating converging, and a binary tree.
+        \textbf{C:} From the different types of ground-truth information, dyngen allows -- amongst others -- suppleming quantitative benchmarks of trajectory alignment, RNA velocity, and cellwise network inference methods.
     }
@@
     \centering
-    \includegraphics[width=0.6\textwidth]{result_files/fig3_showcase_backbones/overview_2.pdf}
+    \includegraphics[width=0.6\textwidth]{result_files/fig3_showcase_backbones/overview.pdf}
     \caption{
@@
         \textbf{D:}~Cyclic. \textbf{E:}~Bifurcating loop. \textbf{F:}~Bifurcating converging.
-        \textbf{G:}~Consecutive branching. \textbf{H:}~Binary tree. \textbf{I:}~Disconnected.
+        \textbf{G:}~Bifurcating cycle. \textbf{H:}~Consecutive branching. \textbf{I:}~Disconnected.
     }
@@
 \textbf{C:} The velocity stream plots produced from the predictions of each method, as generated by scvelo.
-\textbf{D:} TODO: discuss new benchmark results
+\textbf{D:} The predictions scored by two different metrics, the velocity correlation and the vector cosine similarity.
+The velocity correlation is the correlation between the ground-truth velocity (A, right) and the predicted velocity (B).
+The vector cosine similarity is the cosine similarity between the direction of segments of the ground-truth trajectory (A, left) and the RNA velocity values calculated at those points (C).
     }
@@
 While a few pioneering CSNI approaches have already been developed [@aibar_scenicsinglecellregulatory_2017; @kuijjer_estimatingsamplespecificregulatory_2019; @liu_personalizedcharacterizationdiseases_2016], a quantitative assessment of the performance is until now lacking.
-This is not surprising, as neither real nor in silico datasets of cell-specific or even cell-type-specific interactions exists that are large enough so that it can be used as a ground-truth for evaluating CSNI methods.
+This is not surprising, as neither real nor in silico datasets of cell-specific or even cell-type-specific interactions exist that are large enough so that it can be used as a ground-truth for evaluating CSNI methods.
 Extracting the ground-truth dynamic network in dyngen is straightforward though, given that we can calculate how target gene expression would change without the regulator being present.
@@
 We calculated the AUROC and AUPR score -- which are common metrics for NI benchmarking -- for each cell individually.
-Computing the mean AUROC and AUPR per dataset showed that pySCENIC significantly outperforms LIONESS and SSN.
+Computing the mean AUROC and AUPR per dataset showed that pySCENIC significantly outperforms LIONESS + Pearson, which in turn outperforms SSN*.
 
@@
 In addition, the fate decisions made in this example of a bifurcation is reversible, meaning cells can be reprogrammed to go down a different differentiation path.
-If this effect is undesirable, more safeguards need to be put in place to prevent reprogramming from occurring (Section \ref{sec:dyngen-bbl}).
+If this effect is undesirable, more safeguards need to be put in place to prevent reprogramming from occurring.
 
@@
 ## Determining the ground-truth trajectory {#sec:dyngen-groundtruth}
-To determine the state of a cell at any point in the simulation, we simulate the expression of the ground-truth trajectory.
-Subsequently, map the expression of the cell to the closest resembling state in the ground-truth trajectory by finding the point in the ground-truth trajectory with the highest Pearson correlation value w.r.t.
-the expression values.
-
-To simulate the expression of the ground-truth expression, we iterate over the edges in the backbone state network starting from the root state.
-We assume the root state has no modules active and has no expression of any molecules.
-To transition to a next state, we activate or deactivate certain gene modules as defined by the backbone (Section\ \ref{sec:dyngen-modules}) and perform a noiseless ODE simulation.
-For each time interval $t=0.01$ along the trajectory, the expression values are retained.
-
-## Constructing backbones {#sec:dyngen-backbone}
+To construct the ground-truth trajectory, the user needs to provide the ground-truth state network alongside the initial module network (Figure~\ref{fig:example_backbones}).
+Each edge in the state network specifies which modules are allowed to change in expression in transitioning from one state to another.
+For each edge, a simulation is run using the end state of an upstream branch as the initial expression vector, and only allowing the modules as predefined by the attribute to change.
 
@@
     \includegraphics[width=0.8\textwidth]{result_files/example_backbones}
-    \caption{
-      TODO
-    }
+    \caption{Examples of the ground-truth state networks which need to be provided alongside the module network.}
     \label{fig:example_backbones}
@@
 
-### Backbone lego {#sec:dyngen-bbl}
-The backbone can make use of one or more "backbone lego" (BBL) pieces (Figure\ \ref{fig:backbone_lego}).
-A BBL consists of one or more modules which regulate each other such that the output modules present a specific behaviour, depending on the input module (Figure\ \ref{fig:backbone_lego}A).
-Parameters allow determining the number of modules involved in the process and the number of outputs.
-Multiple BBLs can be chained together in order to intuitively create module networks and corresponding state networks (Figure\ \ref{fig:backbone_lego}B).
-Note that not all dynamic processes can be represented by a combination of BBLs, but they can serve as common building blocks to aid the construction of the backbone.
+As an example, consider the cyclic trajectory shown in Figure~\ref{fig:example_backbones}.
+State S0 begins with an expression vector of all zero values.
+To simulate the transition from S0 to S1, regulation of the genes in modules A, B and C are turned on.
+After a predefined period of time, the end state of this transition is considered the expression vector of state S1.
+To simulate the transition from S1 to S2, regulation of the genes in modules D and E are turned on, while the regulation of genes in module C is turned off.
+During this simulation, the expression of genes in modules A, B, D, and E is thus allowed to change.
+The end state of the simulation is considered the expression vector of state S2.
 
-\begin{figure}[htb!]
-    \centering
-    \includegraphics[width=0.8\textwidth]{result_files/backbone_lego}
-    \caption{
-      \textbf{Reusable regulatory building blocks that can be used to construct GRNs that result in dynamic processes following an expected pattern.}
-      \textbf{A:} Each building block contains genes which serve as inputs that can be regulated by external transcription factors.
-The connections inside the building block transform the input signal in a particular way and has one or more output interactions of which the targets are not yet determined.
-      \textbf{B:} By connecting multiple building blocks together, a backbone of regulatory modules can be formed.
-      \textbf{C:} dyngen offers several variations of linear building blocks which allow for differing regulatory mechanisms.
-      \textbf{D:} A branching module will upregulate the expression of just one of the output interactions.
-Two variations are given of which the simple variant is more prone to reprogramming events in comparison to the robust branching module.
-      \textbf{E,F:} These components allow to specify the beginning (no inputs) and end (no outputs) of the backbone.
-    }
-    \label{fig:backbone_lego}
-\end{figure}
-
-When the input node of a **linear BBL** (Figure\ \ref{fig:backbone_lego}C) is upregulated, the module the BBL is connected to will be upregulated.
-A _simple chain_ is a set of modules where a module upregulates the next.
-A _chain with double repression_ has an uneven number of modules forming a chain where each module downregulates the next but all modules (except the input) have basal expression.
-A _grid with double repression_ is similar; except that modules do not have basal expression but instead get upregulated by an upstream module in the chain.
-Finally, a _flip flop_} consists of a simple chain where first the modules (except the last) are upregulated.
-Once the second to last module is upregulated, that module upregulates itself and the first module is strongly repressed, causing all other modules to lose expression and finally the last module to be upregulated.
-The _flip flop_ retains this output state, even when the input changes.
-
-When the input node of a **branching BBL** (Figure\ \ref{fig:backbone_lego}D) is upregulated, a subset of its output modules will eventually be upregulated.
-A \emph{simple branching} uses reciprocal inhibition to drive the upregulation of one of the output modules.
-Due to its simplicity, however, multiple output modules might be upregulated simultaneously and over long periods of simulation time it might be possible that the choice of upregulated module changes.
-A _robust branching_ improves upon the simple branching by preventing upregulation of output modules until an internal branching decision has been made, and by repressing the decision mechanism to avoid other output modules being upregulated other than the one that has been chosen.
-
-A **leaf BBL** (Figure\ \ref{fig:backbone_lego}E) is a linear BBL that has either no inputs or no outputs.
-A _start_ BBL is a linear BBL where the first module has basal expression, and all modules in this module will be active during the burn-in phase of the simulation.
-An _end_ BBL is also a linear BBL with its output regulating one final module.
+For each of the branches in the state network, an expression matrix and the corresponding progression time along that branch are retained.
+To map a simulated cell to the ground-truth, the correlation between its expression values and the expression matrix of the ground-truth trajectory is calculated, and the cell is mapped to the position in the ground-truth trajectory that has the highest correlation.
 
@@
 ## Comparison of cell-specific network inference methods {#sec:dyngen-nicompare}
-Several datasets were generated using the different predefined backbones.
+14 datasets were generated using the 14 different predefined backbones.
 For every cell in the dataset, the transcriptomics profile and the corresponding cell-specific ground-truth regulatory network was determined (Section\ \ref{sec:dyngen-extractgrn}).
@@
 
-15 datasets were generated with 5 different backbones: linear, linear simple, bifurcating, cyclic, and disconnected.
+For each of the 14 different predefined backbones, nine datasets were generated with three difficulty settings and three different seeds.
+The different difficulty settings were obtained by multiplying the transcription rate by a factor of 25 (easy), 5 (medium) and 1 (hard).
+After manual quality control, the easy and medium datasets for backbones "bifurcating_converging", "bifurcating_loop", "converging" and "disconnected" were removed, resulting in a final collection of 102 datasets.
+
 We extracted a ground truth RNA velocity by subtracting for each mRNA molecule the propensity of its production by the propensity of its degradation.
@@
 The velocyto method [@lamanno_rnavelocitysingle_2018], as implemented in the velocyto.py package, in which we varied the "assumption" parameter between "constant_unspliced" and "constant_velocity".
-The scvelo method [@bergen_generalizingrnavelocity_2019], as implemented in the python scvelo package [scvelo.de](http://scvelo.de), in which we varied the "mode" parameter between "deterministic", "stochastic" and "dynamical".
-For both methods, we used the same normalized data as provided by dyngen, with no extra cell or feature filtering.
-We also matched the parameters between both methods as best as possible, i.e.
-the k parameter for smoothing was set to 20 for both methods.
+The scvelo method [@bergen_generalizingrnavelocity_2019], as implemented in the python scvelo package [scvelo.de](http://scvelo.de), in which we varied the "mode" parameter between "deterministic", "stochastic", "dynamical", "dynamical_residuals".
+For both methods, we used the same normalized data as provided by dyngen, with no extra cell or feature filtering, but otherwise matched the parameters to their respective tutorial vignettes as well as possible.
 

### 2020-06-08 analysis/fig2_showcase_data/script.R
merge

@@ -338,10 +338,8 @@ g7a <- read_rds(exp7a$result("explanation_flat.rds"))[[4]]
 exp7b <- start_analysis("usecase_rna_velocity")
 g7b <- read_rds(exp7b$result("one_rna_velocity.rds"))
 
-<<<<<<< HEAD
 exp7c <- start_analysis("usecase_network_inference")
 g7c <- read_rds(exp7c$result("cell1.rds")) + theme(legend.position = "none")
-
 
 g <- wrap_plots(
   wrap_plots(
@@ -363,16 +361,6 @@ g <- wrap_plots(
     ncol = 1
   ),
   widths = c(7, 3)
-=======
-g <- patchwork::wrap_plots(
-  patchwork::wrap_plots(g0a, g0b, nrow = 1, widths = c(1, 1)),
-  g1,
-  g2,
-  g3,
-  g5,
-  heights = c(2, 3, 1, 3, 1.2),
-  ncol = 1
->>>>>>> 09111ea44aa265d28e2194659e0ec00d1918565f
 ) +
   plot_annotation(tag_levels = c('A'))
 ggsave(exp$result("overview.pdf"), g, width = 10, height = 10, device = cairo_pdf)

### 2020-06-08 analysis/fig2_showcase_data/script.R
update

@@ -345,7 +345,7 @@ g <- wrap_plots(
   wrap_plots(
     wrap_plots(g0a, g0b, nrow = 1, widths = c(1, 1)),
     g1 + theme(legend.position = "none"),
-    g2 + theme(legend.position = "bottom"),
+    g2 + theme(legend.position = "none"),
     g3 + theme(legend.position = "none"),
     g5 + theme(legend.position = "bottom"),
     heights = c(2, 3, 1, 3, 1.5),
@@ -364,8 +364,6 @@ g <- wrap_plots(
 ) +
   plot_annotation(tag_levels = c('A'))
 ggsave(exp$result("overview.pdf"), g, width = 10, height = 10, device = cairo_pdf)
-write_rds(g, exp$result("overview.rds"), compress = "gz")
-
 
 
 

### 2020-06-08 analysis/fig2_showcase_data/script.R
add dimred as well

@@ -287,7 +287,7 @@ g4 <-
     strip.text = element_blank(),
     legend.margin = margin()
   ) +
-  scale_x_continuous(breaks = time_breaks) +
+  scale_x_continuous(breaks = time_breaks, expand = c(0, 0)) +
   scale_y_continuous(breaks = scales::breaks_extended(n = 3)) +
   geom_text(aes(label = reaction_text), label_prop_df, size = 3, hjust = 0.5, vjust = 1)
 
@@ -306,7 +306,7 @@ g5 <- ggplot(reg_df) +
     axis.title.y = element_text(angle = 0, vjust = 0.5, hjust = 1),
     legend.position = "bottom"
   ) +
-  scale_x_continuous(breaks = time_breaks) +
+  scale_x_continuous(breaks = time_breaks, expand = c(0, 0)) +
   scale_y_continuous(breaks = c(0, .5, 1), limits = c(0, 1))
 g5
 
@@ -332,6 +332,14 @@ g6 <- ggplot(velocity, aes(time, value)) +
 #######
 # Applications
 #######
+
+
+exp7d <- start_analysis("fig3_showcase_backbones")
+dataset <- read_rds(exp7d$dataset_file("bb_consecutive_bifurcating_3"))
+dimred <- dyndimred::dimred_mds(dataset$expression, distance_method = "pearson")
+g7d <- dynplot::plot_dimred(dataset, dimred = dimred, size_milestones = 4, size_cells = 2)
+
+
 exp7a <- start_analysis("usecase_trajectory_alignment")
 g7a <- read_rds(exp7a$result("explanation_flat.rds"))[[4]]
 
@@ -353,7 +361,7 @@ g <- wrap_plots(
   ),
   wrap_plots(
     plot_spacer(),
-    plot_spacer(),
+    g7d,
     g7a,
     g7b,
     g7c,

### 2020-06-08 analysis/fig2_showcase_data/script.R
remove old plots

@@ -267,31 +267,6 @@ g3 <-
   geom_text(aes(label = reaction_text), label_fir_df, size = 3, hjust = 0.5, vjust = 1)
 
 
-
-#######
-# REACTION PROPENSITIES
-#######
-prop_df_A <- prop_df %>% filter(gene == "Gene A")
-label_prop_df <- prop_df_A %>% group_by(reaction) %>% summarise(time = mean(time), value = max(value)*1.2) %>% mutate(reaction_text = paste0("Gene A ", reaction))
-
-g4 <-
-  ggplot(prop_df_A, aes(time, value)) +
-  facet_wrap(~reaction, ncol = 1, scales = "free_y") +
-  geom_area(aes(fill = forcats::fct_rev(reaction)), position = "stack") +
-  theme_classic() +
-  scale_fill_brewer(palette = "Set2") +
-  labs(x = "Simulation time", y = "Propensity", fill = "Reaction type") +
-  theme(
-    text = element_text(family = "Helvetica"),
-    strip.background = element_blank(),
-    strip.text = element_blank(),
-    legend.margin = margin()
-  ) +
-  scale_x_continuous(breaks = time_breaks, expand = c(0, 0)) +
-  scale_y_continuous(breaks = scales::breaks_extended(n = 3)) +
-  geom_text(aes(label = reaction_text), label_prop_df, size = 3, hjust = 0.5, vjust = 1)
-
-
 #######
 # REGULATION
 #######
@@ -310,24 +285,6 @@ g5 <- ggplot(reg_df) +
   scale_y_continuous(breaks = c(0, .5, 1), limits = c(0, 1))
 g5
 
-#######
-# Prop diff
-#######
-g6 <- ggplot(velocity, aes(time, value)) +
-  geom_step(aes(colour = gene)) +
-  # facet_wrap(~gene, ncol = 1, scales = "free_y") +
-  theme_classic() +
-  labs(x = "Simulation time", y = "Velocity ground-truth", colour = "Molecule") +
-  scale_colour_brewer(palette = "Set1") +
-  # scale_colour_manual(values = c("pre-mRNA" = "#4daf4a", "mRNA" = "#377eb8", "protein" = "#e41a1c", "gene" = "black")) +
-  theme(
-    text = element_text(family = "Helvetica"),
-    strip.background = element_blank(),
-    strip.text = element_blank(),
-    legend.margin = margin(),
[…]

### 2020-06-08 analysis/fig2_showcase_data/script.R
update plots

@@ -201,7 +201,7 @@ g1 <- ggplot(expr_df, aes(time, value)) +
   # scale_colour_manual(values = c("pre-mRNA" = "#4daf4a", "mRNA" = "#377eb8", "protein" = "#e41a1c", "gene" = "black")) +
   theme(
     text = element_text(family = "Helvetica"),
-    axis.title.y = element_text(angle = 0, vjust = 0.5, hjust = 0),
+    # axis.title.y = element_text(angle = 0, vjust = 0.5, hjust = 0),
     strip.background = element_blank(),
     strip.text = element_blank(),
     legend.margin = margin(),
@@ -234,7 +234,7 @@ g2 <- ggplot(state_df) +
   theme(
     text = element_text(family = "Helvetica"),
     legend.margin = margin(),
-    axis.title.y = element_text(angle = 0, vjust = 0.5, hjust = 0),
+    # axis.title.y = element_text(angle = 0, vjust = 0.5, hjust = 0),
     legend.position = "none"
   ) +
   scale_x_continuous(breaks = time_breaks, expand = c(0, 0)) +
@@ -259,7 +259,7 @@ g3 <-
     strip.background = element_blank(),
     strip.text = element_blank(),
     legend.margin = margin(),
-    axis.title.y = element_text(angle = 0, vjust = 0.5, hjust = 1),
+    # axis.title.y = element_text(angle = 0, vjust = 0.5, hjust = 1),
     legend.position = "none"
   ) +
   scale_x_continuous(breaks = time_breaks, expand = c(0, 0)) +
@@ -278,12 +278,12 @@ g5 <- ggplot(reg_df) +
   theme(
     text = element_text(family = "Helvetica"),
     legend.margin = margin(),
-    axis.title.y = element_text(angle = 0, vjust = 0.5, hjust = 1),
-    legend.position = "bottom"
+    # axis.title.y = element_text(angle = 0, vjust = 0.5, hjust = 1)
   ) +
   scale_x_continuous(breaks = time_breaks, expand = c(0, 0)) +
-  scale_y_continuous(breaks = c(0, .5, 1), limits = c(0, 1))
-g5
+  scale_y_continuous(breaks = c(0, .5, 1), limits = c(0, 1.2)) +
+  theme(legend.position = c(.5, .95)) +
[…]

### 2020-06-08 analysis/usecase_trajectory_alignment/3_explanation_plot.R

@@ -106,12 +106,23 @@ comp2t <- comb_traj_less$comp_2
 segm_traj_test <- segm_traj %>% filter(comp_1 %in% comp1t & comp_2 %in% comp2t & comp_3 %in% comp1t & comp_4 %in% comp2t)
 
 cellmappings <- ggplot() +
+  geom_segment(data = segm_traj_test, mapping = aes(x = color2, y = comp_1, xend = color22, yend = comp_3), alpha = 0.25) +
   geom_point(data = comb_traj_less, mapping = aes(x = color2, y = comp_1, colour = as.factor(color)), size = 3.5, show.legend = F) +
   scale_colour_manual(values = c("#fd8d3c", "#6baed6"), label = "") +
-  geom_segment(data = segm_traj_test, mapping = aes(x = color2, y = comp_1, xend = color22, yend = comp_3), alpha = 0.25) +
-  geom_segment(aes(x = 0, y = -0.5, xend = 1, yend = -0.5), arrow = arrow(length = unit(0.25, "cm"))) +
-  annotate(geom = "text", color = "black", x = 0.5, y = -0.75, label = "Pseudotime", size = 4) +
-  theme_void()
+  scale_x_continuous(breaks = c(0, 1), labels = c("Start", "End")) +
+  theme_void() +
+  theme(
+    axis.title = element_text(),
+    axis.title.y = element_blank(),
+    axis.line = element_line(),
+    axis.line.y = element_blank(),
+    axis.ticks = element_line(),
+    axis.ticks.y = element_blank(),
+    axis.text = element_text(),
+    axis.text.y = element_blank()
+  ) +
+  labs(x = "Pseudotime")
+write_rds(lst(segm_traj_test, comb_traj_less), exp$result("explanation_plot_data.rds"), compress = "gz")
 cellmappings
 
 

### 2020-06-08 analysis/fig1_showcase_functionality/script.R
clean up scripts

@@ -1,235 +1,0 @@
-library(tidyverse)
-library(dyngen)
-library(dyngen.manuscript)
-
-set.seed(1)
-
-exp <- start_analysis("fig1_showcase_functionality")
-
-backbone <- bblego(
-  bblego_start("A", type = "simple", num_modules = 1),
-  bblego_linear("A", "B", type = "simple", num_modules = 3),
-  bblego_branching("B", c("C", "D"), num_steps = 1),
-  bblego_end("C", type = "simple", num_modules = 2),
-  bblego_end("D", type = "simple", num_modules = 2)
-)
-model <-
-  initialise_model(
-    num_tfs = 12,
-    num_targets = 30,
-    num_hks = 15,
-    num_cells = 150,
-    backbone = backbone,
-    verbose = TRUE,
-    download_cache_dir = "~/.cache/dyngen",
-    num_cores = 8,
-    distance_metric = "euclidean",
-    simulation_params = simulation_default(
-      kinetics_noise_function = kinetics_noise_none(),
-      experiment_params = simulation_type_wild_type(num_simulations = 8),
-      compute_cellwise_grn = TRUE,
-      compute_rna_velocity = TRUE
-    )
-  ) %>%
-  generate_tf_network() %>%
-  generate_feature_network() %>%
-  generate_kinetics()
-
-## grn
-plot_backbone_statenet(model) %>% ggsave(filename = exp$result("statenet.pdf"), width = 6, height = 6)
-plot_backbone_modulenet(model) %>% ggsave(filename = exp$result("modulenet.pdf"), width = 8, height = 6)
-
-plot_feature_network(model, show_targets = FALSE)
-
-plot_feature_network(model)
-plot_feature_network(model, show_hks = TRUE)
-plot_feature_network(model)
-plot_feature_network(model, show_hks = TRUE) %>% ggsave(filename = exp$result("featnet.pdf"), width = 8, height = 6)
-plot_feature_network(model, show_targets = FALSE) %>% ggsave(filename = exp$result("featnet_onlytfs.pdf"), width = 8, height = 6)
-
-## simulate
-model2 <- model %>%
-  generate_gold_standard()
-
-g <- plot_gold_expression(model2, what = "mol_mrna") # mrna
-ggsave(exp$result("gold_mrna.pdf"), g, width = 8, height = 6)
[…]

### 2020-06-08 analysis/fig2_showcase_data/script.R

@@ -1,351 +1,0 @@
-library(tidyverse)
-library(dyngen)
-library(dyngen.manuscript)
-library(grImport)
-library(patchwork)
-
-set.seed(1)
-
-exp <- start_analysis("fig2_showcase_data")
-
-# Read regulation svg from file
-fil <- tempfile()
-on.exit(file.remove(fil))
-PostScriptTrace(exp$result("gene_regulation.eps"), outfilename = fil)
-g0a <- pictureGrob(readPicture(fil))
-
-# generate cyclic trajectory
-census_interval <- 5
-total_time <- 500
-time_breaks <- c(0, 500)
-model <- exp$temporary("model.rds") %cache% {
-  backbone <- backbone(
-    module_info = tribble(
-      ~module_id, ~basal, ~burn, ~independence,
-      "A", 1, TRUE, 1,
-      "B", 1, TRUE, 1,
-      "C", 1, TRUE, 1,
-      "D", 0, TRUE, 1,
-      "E", 0, TRUE, 1
-    ),
-    module_network = tribble(
-      ~from, ~to, ~effect, ~strength, ~hill,
-      "A", "B", -1L, 4, 2,
-      "B", "C", -1L, 4, 2,
-      "C", "D", 1L, 1, 2,
-      "D", "E", 1L, 1, 2,
-      "E", "A", -1L, 4, 2
-    ),
-    expression_patterns = tribble(
-      ~from, ~to, ~module_progression, ~start, ~burn, ~time,
-      "Burn0", "S1", "+A,+B,+C", TRUE, TRUE, 150,
-      "S1", "S2", "+D,+E,-C", FALSE, FALSE, 120,
-      "S2", "S3", "+C,-A,-B", FALSE, FALSE, 120,
-      "S3", "S1", "+A,+B,-D,-E", FALSE, FALSE, 120
-    )
-  )
-  model <-
-    initialise_model(
-      num_tfs = nrow(backbone$module_info),
-      num_targets = 0,
-      num_hks = 0,
-      num_cells = 300,
-      backbone = backbone,
-      verbose = TRUE,
-      download_cache_dir = "~/.cache/dyngen",
-      num_cores = 8,
-      simulation_params = simulation_default(
-        burn_time = 240,
-        total_time = total_time,
-        experiment_params = simulation_type_wild_type(num_simulations = 1),
-        census_interval = census_interval,
-        store_reaction_firings = TRUE,
-        store_reaction_propensities = TRUE,
[…]

### 2020-06-08 analysis/summary/script.R

@@ -1,145 +1,351 @@
 library(tidyverse)
 library(dyngen)
 library(dyngen.manuscript)
-library(dynplot)
+library(grImport)
 library(patchwork)
 
 set.seed(1)
 
 exp <- start_analysis("summary")
 
+# Read regulation svg from file
+fil <- tempfile()
+on.exit(file.remove(fil))
+PostScriptTrace(exp$result("gene_regulation.eps"), outfilename = fil)
+g0a <- pictureGrob(readPicture(fil))
 
-# FIRST ROW: illustrate dyngen pipeline -----------------------------------
-model1 <- exp$result("model_branching.rds") %cache% {
-  backbone <- bblego(
-    bblego_start("A", type = "simple", num_modules = 1),
-    bblego_linear("A", "B", type = "simple", num_modules = 3),
-    bblego_branching("B", c("C", "D"), num_steps = 1),
-    bblego_end("C", type = "simple", num_modules = 2),
-    bblego_end("D", type = "simple", num_modules = 2)
+# generate cyclic trajectory
+census_interval <- 5
+total_time <- 500
+time_breaks <- c(0, 500)
+model <- exp$temporary("model.rds") %cache% {
+  backbone <- backbone(
+    module_info = tribble(
+      ~module_id, ~basal, ~burn, ~independence,
+      "A", 1, TRUE, 1,
+      "B", 1, TRUE, 1,
+      "C", 1, TRUE, 1,
+      "D", 0, TRUE, 1,
+      "E", 0, TRUE, 1
+    ),
+    module_network = tribble(
+      ~from, ~to, ~effect, ~strength, ~hill,
+      "A", "B", -1L, 4, 2,
+      "B", "C", -1L, 4, 2,
+      "C", "D", 1L, 1, 2,
+      "D", "E", 1L, 1, 2,
+      "E", "A", -1L, 4, 2
+    ),
+    expression_patterns = tribble(
+      ~from, ~to, ~module_progression, ~start, ~burn, ~time,
+      "Burn0", "S1", "+A,+B,+C", TRUE, TRUE, 150,
+      "S1", "S2", "+D,+E,-C", FALSE, FALSE, 120,
+      "S2", "S3", "+C,-A,-B", FALSE, FALSE, 120,
+      "S3", "S1", "+A,+B,-D,-E", FALSE, FALSE, 120
+    )
   )
-  model1 <-
+  model <-
     initialise_model(
-      num_tfs = 12,
[…]

### 2020-06-08 analysis/summary/script.R
update summary

@@ -182,11 +182,19 @@ g0b <-
   geom_node_point(aes(colour = name), size = 7) +
   geom_node_point(colour = "white", size = 6) +
   geom_node_text(aes(label = name)) +
-  theme_graph(base_family = 'Helvetica') +
   scale_edge_width_continuous(trans = "log10", range = c(.5, 3)) +
   scale_colour_brewer(palette = "Set1") +
-  theme(legend.position = "none") +
-  expand_limits(y = c(-.1, 1.1), x = c(-.05, 1.05))
+  expand_limits(y = c(-.2, 1.2), x = c(-.05, 1.05)) +
+  coord_cartesian() +
+  theme_classic() +
+  theme(
+    legend.position = "none",
+    text = element_text(family = "Helvetica"),
+    axis.title = element_blank(),
+    axis.line = element_blank(),
+    axis.ticks = element_blank(),
+    axis.text = element_blank()
+  )
 
 #######
 # ABUNDANCE LEVELS
@@ -198,16 +206,14 @@ g1 <- ggplot(expr_df, aes(time, value)) +
   labs(x = "Simulation time", y = "Molecule levels", colour = "Molecule") +
   scale_colour_brewer(palette = "Set1") +
   scale_y_continuous(n.breaks = 3, expand = c(0, 0)) +
-  # scale_colour_manual(values = c("pre-mRNA" = "#4daf4a", "mRNA" = "#377eb8", "protein" = "#e41a1c", "gene" = "black")) +
   theme(
     text = element_text(family = "Helvetica"),
-    # axis.title.y = element_text(angle = 0, vjust = 0.5, hjust = 0),
     strip.background = element_blank(),
     strip.text = element_blank(),
     legend.margin = margin(),
     legend.position = "none"
   ) +
-  scale_x_continuous(breaks = time_breaks, expand = c(0, 0)) +
+  scale_x_continuous(breaks = time_breaks, expand = c(0.02, 0.02)) +
   geom_text(aes(x = mean(time_breaks), y = max, label = molecule), expr_df %>% group_by(molecule) %>% summarise(max = max(value)), size = 3, hjust = 0.5, vjust = 1)
 
 #######
@@ -234,37 +240,35 @@ g2 <- ggplot(state_df) +
   theme(
     text = element_text(family = "Helvetica"),
     legend.margin = margin(),
[…]

### 2020-06-08 analysis/summary/script.R
update

@@ -359,7 +359,7 @@ g7c <- read_rds(exp7c$result("cell1.rds")) +
     axis.ticks = element_blank(),
     axis.text = element_blank()
   ) +
-  labs(title = "Cellwise network inference")
+  labs(title = "Cell-specific network inference")
 
 such_great_heights <- c(2.5, 3, 2, 4, 3)
 g <- wrap_plots(

### 2020-06-13 analysis/summary/figure_2.R
update

@@ -7,6 +7,21 @@ set.seed(1)
 
 exp <- start_analysis("summary")
 
+theme_title <- function() {
+  theme(
+    plot.title = element_text(hjust = .5),
+    axis.title = element_blank(),
+    axis.text = element_blank(),
+    line = element_blank(),
+    rect = element_blank()
+  )
+}
+plot_title <- function(value, angle = 0, size = 10) {
+  ggplot() + geom_text(aes(1, 1, label = value), angle = angle, size = size) + theme_title()
+  # ggplot() + geom_point(aes(1,1), alpha = 0) + labs(title = title, subtitle = subtitle, tag = tag) + theme_title()
+}
+
+
 #################
 # TRAJECTORY ALIGNMENT
 #################
@@ -30,7 +45,7 @@ ga2 <- plots_a$prediction +
   coord_cartesian() +
   theme_classic() +
   theme(
-    legend.position = "none",
+    legend.position = "right",
     text = element_text(family = "Helvetica"),
     axis.title = element_blank(),
     axis.line = element_blank(),
@@ -40,8 +55,7 @@ ga2 <- plots_a$prediction +
   labs(title = NULL, subtitle = NULL, tag = NULL)
 
 scores_a <-
-  read_rds(exp_a$result("results.rds")) %>%
-  mutate(noise = factor(noise))
+  read_rds(exp_a$result("results.rds"))
 
 plota_data <-
   scores_a %>%
@@ -56,16 +70,19 @@ plota_data <-
   ungroup()
 
 ga3 <-
-  ggplot(plota_data) +
+  ggplot(plota_data %>% filter(noise %in% c(.2, .4, .6, .8, 1))) +
   geom_boxplot(
-    aes(noise, ymin = min, lower = lower, middle = mean, upper = upper, max = max, fill = method, group = paste0(noise, method)),
+    aes(forcats::fct_rev(factor(noise)), ymin = min, lower = lower, middle = mean, upper = upper, max = max, fill = method, group = paste0(noise, method)),
     stat = "identity", width = 0.5, size = 0.45
   ) +
+  coord_flip() +
   theme_classic() +
   theme_common() +
-  labs(x = "Noise levels", y = "Distance (lower is better)", colour = "Method") +
-  scale_fill_brewer(palette = "Dark2") +
-  theme(legend.position = "right")
+  labs(x = "Noise levels", y = "Distance (lower is better)", fill = "Method") +
[…]

### 2020-06-13 analysis/usecase_trajectory_alignment/3_explanation_plot.R

@@ -65,45 +65,42 @@ write_rds(dataset2D, paste0(exp$dataset_folder("expplot"), "dataset2D.rds"), com
 milestone_colors <- c("#ff681c", "#ff681c", "#ff681c", "#ff681c", "#3abbba","#3abbba", "#3abbba", "#3abbba", "#ff681c", "#3abbba")
 together <- plot_trajectory_in_color(datasetD, milestone_colors, c("sample 1", "sample 2"), c("#3abbba", "#ff681c"), plot_sd = 0.11)
 
-t1_pt <- plot_pseudotime(dataset1, palette = "Oranges", trajectory_projection_sd = 0.11, plot_trajectory = T)
-t2_pt <- plot_pseudotime(dataset2D, palette = "Blues", trajectory_projection_sd = 0.11, plot_trajectory = T)
+t1_pt <- plot_pseudotime(dataset1, palette = "Oranges", trajectory_projection_sd = 0.11, plot_trajectory = TRUE)
+t2_pt <- plot_pseudotime(dataset2D, palette = "Blues", trajectory_projection_sd = 0.11, plot_trajectory = TRUE)
 
 expr1 <- get_cell_expression(dataset1, dataset1$milestone_network, "sA")$expression
 expr2 <- get_cell_expression(dataset2D, dataset2D$milestone_network, "sA")$expression
-a <- dtw(expr2, expr1, keep.internals = T, open.end = F)
+a <- dtw(expr2, expr1, keep.internals = TRUE, open.end = FALSE)
 dtwPlotAlignment(a)
 
-plot_dens <- plot_density(a, show_legend = TRUE) + labs(title = NULL, subtitle = NULL)
-plot_dens
+plot_dens <- plot_density(a, title = "Alignment between\ndiseased/healthy")
 
-
-e1 <- calculate_correct_pseudotime(dataset1, dataset1$milestone_network, "sA", normalized = F)
-e2 <- calculate_correct_pseudotime(dataset2D, dataset2D$milestone_network, "sA", normalized = F)
+e1 <- calculate_correct_pseudotime(dataset1, "sA", normalized = FALSE)
+e2 <- calculate_correct_pseudotime(dataset2D, "sA", normalized = FALSE)
+maxe12 <- max(c(e1, e2))
+e1 <- e1 / maxe12
+e2 <- e2 / maxe12
 
 # warped segmenten
-traj1 <- tibble(x = sort(e1), color = 1, y = 0)
[…]

### 2020-06-13 analysis/usecase_trajectory_alignment/4_plots.R

@@ -51,7 +51,7 @@ expr1 <- res1$expression
 expr2 <- res2$expression
 
 alignment_original <- dtw(expr2, expr1, step.pattern=symmetric2, keep.internals=T)
-ao <- plot_density(alignment_original) + labs(title = "Original cells", subtitle = NULL)
+ao <- plot_density(alignment_original, title = "DTW")
 
 smp1 <- seq(from = 1, to = 1000, by = 10) #sample(1000, size = 100, replace = FALSE)
 pt1_smp <- pt1[seq(from = 1, to = 1000, by = 10)]
@@ -59,19 +59,29 @@ pt2_smp <- pt2[seq(from = 1, to = 1000, by = 10)]
 expr1_smp <- expr1[names(pt1_smp),]
 expr2_smp <- expr2[names(pt2_smp),]
 
-alignment_subsample <- dtw(expr2_smp, expr1_smp, step.pattern=symmetric2, keep.internals=T)
-asubs <- plot_density(alignment_subsample) + labs(title = "Subsampled", subtitle = NULL)
-
 alignment_smooth <- dtw(res2_sm$expression, res1_sm$expression, step.pattern=symmetric2, keep.internals=T)
-a_sm <- plot_density(alignment_smooth) + labs(title = "Smoothed", subtitle = NULL)
+a_sm <- plot_density(alignment_smooth, title = "DTW+smoothing")
 
 
 # PART 1: Explanation plot ------------------------------------------------
+part1plots <- read_rds(exp$result("usecase_separateplots.rds"))
 
-part1 <- readRDS(exp$result("explanation_flat.rds"))
-part2 <- (ao + plot_spacer() + a_sm) + plot_layout(widths = c(0.35, 0.3, 0.35))
-all_plots <- part1 / part2 / g
-all_plots <- all_plots + plot_annotation(tag_levels = c('A')) + plot_layout(heights = c(0.5, 1.5, 2))
+all_plots <- wrap_plots(
+  wrap_plots(
+    part1plots$t2_pt + labs(tag = "A", colour = "Healthy\nprogression   "),
+    part1plots$t1_pt + labs(tag = "B", colour = "Diseased\nprogression   "),
+    part1plots$prediction + theme(legend.position = "bottom") + labs(tag = "C")
+  ),
+  wrap_plots(
+    part1plots$plot_dens + labs(tag = "D") + theme(legend.position = "none"),
[…]

### 2020-06-13 package/R/cell_expression.R

@@ -4,7 +4,7 @@
 #'
 #' @export
 get_cell_expression <- function(dataset, milestone_network, start){
-  pseudotime <- sort(calculate_correct_pseudotime(dataset, milestone_network, start))
+  pseudotime <- sort(calculate_correct_pseudotime(dataset, start))
   nor2 <- names(pseudotime)
   volg2 <- sapply(nor2, function(i) strtoi(substr(i, 5, nchar(i))))
   cnt2 <- as.matrix(dataset$counts)
@@ -25,44 +25,25 @@ get_cell_expression <- function(dataset, milestone_network, start){
 #' Get corrected pseudotime
 #'
 #' @param dataset A dyngen dataset
+#' @param start The start milestone
 #'
 #' @importFrom gtools mixedorder
 #'
 #' @export
-calculate_correct_pseudotime <- function(dataset, milestone_network, start, normalized = T){
-  if(normalized){
-    length_pieces <- milestone_network$length/sum(milestone_network$length)
-  } else {
-    length_pieces <- milestone_network$length
+calculate_correct_pseudotime <- function(dataset, start, normalized = TRUE){
+  if (normalized) {
+    dataset$milestone_network <-
+      dataset$milestone_network %>%
+      mutate(length = length / sum(length))
   }
-  nr_piece <- 0
-  traj <- numeric()
-  names_traj <- list()
-  # TODO convert to for for each piece
-  while(start %in% milestone_network[["from"]]){
-    idx <- which(start == milestone_network[["from"]])[[1]]
-    end <- milestone_network[["to"]][idx]
-    to_add <- sum(length_pieces[0:nr_piece])
-
-    if(normalized){
-      perc <-(dataset[["progressions"]] %>% filter(from == start & to == end))$percentage * length_pieces[nr_piece + 1] + to_add
-    } else {
-      perc <-(dataset[["progressions"]] %>% filter(from == start & to == end))$percentage * length_pieces[nr_piece + 1] + to_add
-
-    }
-    nams <-(dataset[["progressions"]] %>% filter(from == start & to == end))$cell_id
-
-    traj <- c(traj, perc)
-    names_traj <- c(names_traj, nams)
-
-    start <- end
-    nr_piece <- nr_piece + 1
-
-  }
-
-  names(traj) <- names_traj
-  traj <- traj[mixedorder(names(traj))]
-  traj
+  pct <- dynwrap::calculate_geodesic_distances(
+    dataset,
[…]

### 2020-06-13 package/R/plot_explanationplot.R

@@ -1,35 +1,34 @@
 #' @export
-plot_density <- function(a1, title = "Distance matrix +\n warping path", subtitle = "", show_legend = F){
-  x <- as_tibble(a1$index1)
-  x$Y <- a1$index2
-
-  gga_1 <- melt(a1$costMatrix)
-  p_heat1 <- ggplot(gga_1, aes(Var1, Var2, fill= value)) +
-    geom_raster(show.legend = show_legend) +
-    scale_fill_distiller(palette = "RdYlGn", breaks = range(gga_1$value), labels = c("min", "max"), name = "Accumulated\ndistance") +
-    scale_x_continuous(expand = c(0, 0)) + scale_y_continuous(expand = c(0, 0)) +
-    geom_line(data=x, aes(x=value, y=Y)) +
-    theme(panel.grid.major = element_blank(),
-          panel.grid.minor = element_blank()) +
-    theme(legend.position="bottom") +
-    labs(title = title, x = "Sample 2", y = "Sample 1", fill = "Distance")
-
-  p_heat1
+plot_density <- function(a1, title = "Distance matrix +\n warping path", subtitle = NULL){
+  linedf <- tibble(
+    x = a1$index1,
+    y = a1$index2
+  )
+  rasterdf <- melt(a1$costMatrix, varnames = c("x", "y"), value.name = "distance")
+  ggplot() +
+    geom_raster(aes(x, y, fill = distance), rasterdf) +
+    geom_line(aes(x, y), linedf) +
+    scale_fill_distiller(palette = "RdYlGn", breaks = range(rasterdf$distance), labels = c("min", "max")) +
+    scale_x_continuous(expand = c(0, 0), breaks = range(linedf$x), labels = c("start", "end")) +
+    scale_y_continuous(expand = c(0, 0), breaks = range(linedf$y), labels = c("start", "end")) +
+    theme(
+      panel.grid.major = element_blank(),
+      panel.grid.minor = element_blank(),
+      legend.position = "bottom"
+    ) +
+    labs(title = title, x = "Sample 2", y = "Sample 1", fill = "Accumulated\ndistance")
 }
 
 #' @export
 plot_pseudotime <- function(ds, palette = "Blues", ...){
-  p_traj1 <- plot_dimred(ds, color_cells = "pseudotime", size_cells = 3, ...)
-  p_traj1 <- p_traj1 +
[…]

### 2020-06-13 manuscript.Rmd
update

@@
     citation_package: biblatex
-biblatexoptions: sorting=none
+biblatexoptions: sorting=none,url=false
 classoption:
@@
   \usepackage{pifont}
+  \usepackage{float}
   \setmainfont [Path = fonts/,
@@
   ]{MuseoSans}
+
 bibliography: library.bib
@@
 ```
+<!-- link naar de oude tekst: https://docs.google.com/document/d/15ZkzzB-XSYFfZutO4YPkUIw6iUvxJS92fSPfzqR1cLA/edit -->
 
 # Abstract 
-**Purpose:** A recent breakthrough in single-cell omics is the ability to perform multi-modal measurements at single-cell level.
-However, in many cases insufficient metadata is available to accurately benchmark new types of algorithms.
-Existing generators of scRNA-seq are valuable in benchmarking novel computational tools more rigorously, but adding additional modalities or experimental conditions usually requires significant methodological alterations.
-**Results:** We introduce dyngen, a simulator of synthetic single cells by simulating gene regulation, splicing and translation at a single molecule level.
-From a simulation many layers of information can be extracted, including molecule abundance, progression along a dynamic process, activation strength of individual regulatory interactions, and a ground-truth RNA velocity.
-We demonstrate dyngen by showcasing its use on three novel computational approaches where no benchmarking of tools have been described for up to date: RNA velocity, cell-specific gene regulatory network inference and trajectory alignment methods.
-In itself, each of these three application cases constitutes a novel result that has not yet been described in the literature so far, and which we believe will spearhead research on development of novel computational tools for single-cell data.
-**Conclusion:** Ultimately, dyngen allows anticipating technological developments in single-cell multi-omics.
-It is thus possible to design and evaluate the performance and robustness of new types of computational analyses before experimental data becomes available.
-In addition, dyngen could also be used to compare which experimental protocol is the most cost-effective in producing qualitative and robust results in downstream analysis.
+<!-- format: max. 70 words -->
+We present dyngen, a novel, multi-modal simulation engine for studying dynamic cellular processes at single-cell resolution.
+dyngen is more flexible than current single-cell simulation engines, and allows better method development and benchmarking, thereby stimulating development and testing of novel computational methods.
+We demonstrate its potential for spearheading novel computational methods on three novel applications: aligning cell developmental trajectories, single-cell regulatory network inference and estimation of RNA velocity.
 
-# Introduction {#sec:dyngen-introduction}
-Continuous technological advancements to single-cells omics are having profound effects on how researchers can validate biological hypotheses.
-Early experimental technologies typically only allowed profiling a single modality (e.g.
-DNA sequence, RNA or protein expression).
-However, recent developments permit profiling multiple modalities simultaneously, and every modality added allows for new types of analyses that can be performed.
+# Main text
+<!-- format: 1500 Words, no headings, max. 2 figures, max. 20 references-->
+Single-cell simulation engines are becoming increasingly important for testing and benchmarking computational methods, a pressing need in the widely expanding field of single-cell biology.
+Complementary to real biological data, synthetic data provides a valuable alternative where the actual ground truth is completely known and thus can be compared to, in order to make quantitative evaluations of computational methods that aim to reconstruct this ground truth [@zappia_splattersimulationsinglecell_2017].
+In addition, simulation engines are more flexible when it comes to stress-testing computational methods, for example by varying the parameters of the simulation, such as the amount of noise, samples, and cells measured, allowing benchmarking of methods over a wide range of possible scenarios.
+In this way, they can even guide the design of real biological experiments, finding out the best conditions to be used as input for subsequent computational pipelines.
+Another, more experimental use of simulation engines is their important role in spearheading the development of novel computational methods, possibly even before real data is available.
+In this way, simulation engines can be used to assess the value of novel experimental protocols or treatments.
+Simulation engines are also increasingly important when it comes to finding alternatives to animal models, for example for drug testing and precision medicine.
+In such scenarios, cellular simulations can act as digital twins, offering unlimited experimentation in silico.
 
-This confronts method developers with several challenges.
-The majority of the >250 peer-reviewed computational tools for analysing single-cell omics data were published without a quantitative assessment of the accuracy of the tool.
-This is partially due to low availability of suitable benchmarking datasets; even if there are sufficient suitable input datasets available, these are often not accompanied by the necessary metadata to serve as ground-truth for a benchmark.
+Here, we introduce dyngen, a novel multi-modal simulator of dynamic biological processes at single-cell resolution (Figure\ \ref{fig:overview}).
+dyngen uses Gillespie's stochastic simulation algorithm [@gillespie_exactstochasticsimulation_1977] to simulate gene regulation, splicing, and translation at a single-molecule level.
+Other generators of scRNA-seq data (e.g.
+splatter [@zappia_splattersimulationsinglecell_2017], powsimR [@vieth_powsimrpoweranalysis_2017], PROSSTT [@papadopoulos_prossttprobabilisticsimulation_2018] and SymSim [@zhang_simulatingmultiplefaceted_2019]) have already been used extensively to explore the strengths and weaknesses of computational tools, both by method developers [@street_slingshotcelllineage_2018; @parra_reconstructingcomplexlineage_2018; @lummertzdarocha_reconstructioncomplexsinglecell_2018; @lin_scclassifyhierarchicalclassification_2019] and independent benchmarkers [@duo_systematicperformanceevaluation_2018; @saelens_comparisonsinglecelltrajectory_2019; @soneson_biasrobustnessscalability_2018].
+However, a limitation of these existing simulators is that they would require significant methodological alterations to add additional modalities or experimental conditions (Table\ \ref{tab:comparison}).
 
-Here, synthetic data plays a crucial role in asserting minimum performance requirements for novel tools in anticipation of adequate real data.
-Generators of scRNA-seq data (e.g.
-splatter [@zappia_splattersimulationsinglecell_2017], powsimR [@vieth_powsimrpoweranalysis_2017], PROSSTT [@papadopoulos_prossttprobabilisticsimulation_2018] and SymSim [@zhang_simulatingmultiplefaceted_2019]) have already been used extensively to explore the strengths and weaknesses of computational tools, both by method developers [@street_slingshotcelllineage_2018; @parra_reconstructingcomplexlineage_2018; @lummertzdarocha_reconstructioncomplexsinglecell_2018; @lin_scclassifyhierarchicalclassification_2019] and independent benchmarkers [@duo_systematicperformanceevaluation_2018; @saelens_comparisonsinglecelltrajectory_2019; @soneson_biasrobustnessscalability_2018].
-However, a limitation of existing scRNA-seq profiles generators is that they would require significant methodological alterations to add additional modalities or experimental conditions.
-
-An ideal experiment would be able to observe all aspects of a cell, including a full history of its molecular states, spatial positions and environmental interactions [@stuart_integrativesinglecellanalysis_2019].
-While this falls outside the reach of current experimental technologies, generating synthetic data in anticipation of new experimental technologies would allow developing the next wave of computational tools.
-
-We developed a multi-modality simulator of single cells called dyngen (Figure\ \ref{fig:showcase}).
-dyngen uses an optimised version of Gillespie's stochastic simulation algorithm [@gillespie_exactstochasticsimulation_1977]  to simulate gene regulation, splicing, and translation at a single-molecule level.
-Its methodology allows tracking of many layers of information throughout the simulation, including the abundance of any molecule in the cell, the progression of the cell along a dynamic process, and the activation strength of individual regulatory interactions.
-dyngen can simulate a large variety of dynamic processes (e.g.
-cyclic, branching, disconnected) as well as a broad range of experimental conditions (e.g.
-batch effects and time-series, perturbation and knockdown experiments).
-The fine-grained controls over simulation parameters allow dyngen to be applicable to a broad range of use-cases.
-We demonstrate this by performing the first quantitative evaluations of three types of novel computational approaches: RNA velocity, cell-specific network inference, and trajectory alignment methods.
-
-\begin{figure}[htb!]
+\begin{figure}[t!]
     \centering
-    \includegraphics[width=\linewidth]{result_files/summary/overview.pdf}
+    \includegraphics[width=\linewidth]{result_files/summary/figure_1_edited.pdf}
     \caption{
       \textbf{Showcase of dyngen functionality.}
-        \textbf{A:} The typical process of generating a bifurcating dataset with dyngen.
-Step 1: generating a gene regulatory network.
-Step 2: simulating changes in expression over time.
-Step 3: combining multiple simulations and mapping to the ground-truth trajectory.
-Step 4: sampling cells and molecules.
-        \textbf{B:} dyngen includes different pre-defined trajectories, including converging, cyclic, bifurcating converging, and a binary tree.
-        \textbf{C:} From the different types of ground-truth information, dyngen allows -- amongst others -- suppleming quantitative benchmarks of trajectory alignment, RNA velocity, and cellwise network inference methods.
-    }
-    \label{fig:showcase}
-\end{figure}
-
-# Results {#sec:dyngen-results}
-
-A cell consists of a set of molecules, the abundance of which is affected by a set of reactions: transcription, splicing, translation, and degradation (Figure\ \ref{fig:simplecyclic}A).
-A gene regulatory network (GRN) defines the reactions that are allowed to occur (Figure\ \ref{fig:simplecyclic}B), which is constructed in such a way that cells slowly develop over time (Figure\ \ref{fig:simplecyclic}C,D).
-With every time step $\text{d}t$ in the simulation, the probability of a reaction occurring is computed (not shown).
-From the probabilities are sampled which reactions occur during this time step $\text{d}t$ (Figure\ \ref{fig:simplecyclic}E).
-
-\begin{figure}[htb!]
-    \centering
-    \includegraphics[width=\linewidth]{result_files/fig2_showcase_data/overview.pdf}
-    \caption{
-        \textbf{dyngen models reactions at a single-molecule level and keeps track of multiple levels of information throughout a simulation.} 
       \textbf{A:} Changes in abundance levels are driven strictly by gene regulatory reactions.
@@
       \textbf{D:} Firing many reactions can significantly alter the cellular state over time.
-        \textbf{E:} dyngen keeps track of the reactions that were fired during small intervals of time.
+      \textbf{E:} dyngen keeps track of the likelihood of a reaction firing during small intervals of time, called the propensity, as well as the actual number of firings.
       \textbf{F:} Similarly, dyngen can also keep track of the regulatory activity of every interaction.
+      \textbf{G:} A benchmark of trajectory inference methods has already been performed using the cell state ground-truth.
+      \textbf{H:} The cell state ground-truth enables evaluating trajectory alignment methods \cite{saelens_comparisonsinglecelltrajectory_2019}.
+      \textbf{I:} The reaction propensity ground-truth enables evaluating RNA velocity methods.
+      \textbf{J:} The cellwise regulatory network ground-truth enables evaluating cell-specific gene regulatory network inference methods.
     }
-    \label{fig:simplecyclic}
+    \label{fig:overview}
+\end{figure}
+dyngen was designed to include all of these functionalities and more by design.
+Its methodology allows tracking many layers of information throughout the simulation, including the abundance of any molecule in the cell, the progression of the cell along a dynamic process, and the activation strength of individual regulatory interactions.
+dyngen can simulate a large variety of dynamic processes (e.g.
+cyclic, branching, disconnected) as well as a broad range of experimental conditions (e.g.
+batch effects and time-series, perturbation and knockdown experiments).
+The fine-grained controls over simulation parameters allow dyngen to be applicable to a broad range of use-cases to simulate dynamic biological processes.
+The original design of dyngen was motivated by the plethora of methods available for trajectory inference, where dyngen allowed the first large-scale benchmarking of such methods [@saelens_comparisonsinglecelltrajectory_2019; @vandenberge_trajectorybaseddifferentialexpression_2019].
+
+Here, we highlight the novel functionality of dyngen by evaluating three novel types of computational approaches for which no simulation engines exist yet: single-cell network inference, trajectory alignment and RNA velocity (Figure\ \ref{fig:applications}).
+We emphasize that our main aim here is to illustrate the potential of dyngen for these evaluations, rather than performing large-scale benchmarking, which would require assessing many more quantitative and qualitative aspects of each method [@weber_essentialguidelinescomputational_2019].
+
+\begin{figure}[t!]
+    \centering
+    \includegraphics[width=\linewidth]{result_files/summary/figure_2.pdf}
+    \caption{
+      \textbf{dyngen provides ground-truth data for a variety of applications (left), which can be used to quantitatively evaluate methods (right).}
+      \textbf{A:} Trajectory alignment aligns two trajectories between samples.
+dyngen can simulate different scenarios in which alignment is necessary, such as a premature stop as shown here.
+We compared two versions of dynamic time warping (DTW): normal DTW aligns all individual cells, while DTW+smoothing first smooths the expression data and then uses DTW to align the smoothed cells.
+      \textbf{B:} RNA velocity calculates for each cell the direction in which the expression of each gene is moving.
+We evaluated scVelo and velocyto by comparing these vectors with the known velocity vector (velocity correlation) and with the known direction of the cellular trajectory in a dimensionality reduction (velocity arrow cosine).
+      \textbf{C:} Cell-specific network inference (CSNI) predicts the regulatory network of every individual cell.
+We evaluate each cell-specific regulatory network with typical metrics for network inference: the Area Under the Receiver Operating Characteristics-curve (AUROC) and Area Under the Precision Recall-curve (AUPR).
+We evaluate three CSNI methods by computing the mean AUROC and AUPR across all cells.
+    }
+    \label{fig:applications}
 \end{figure}
 
-dyngen returns many modalities throughout the whole simulation: molecular abundance, number of reaction firings, reaction likelihoods, and regulation activations (Figure\ \ref{fig:simplecyclic}C--F).
-These modalities can serve both as input data and ground truth for benchmarking many types of computational approaches.
-For example, a network inference method could use mRNA abundance and regulation activities as inputs and its output could be benchmarked against the gold standard GRN.
+**Trajectory alignment** methods align trajectories from different samples and allow studying the differences between the different trajectories.
+For example, by comparing the transcriptomic profiles of cells from a diseased patient to a healthy control, it might be possible to detect transcriptomics differences (differential expression) of particular cells along a developmental process, or to detect an early stop of the trajectory of the diseased patient.
+Currently, trajectory alignment is limited to aligning linear trajectories, though other topologies of trajectory could be aligned as well.
+Dynamic Time Warping (DTW) [@giorgino_computingvisualizingdynamic_2009] is a method designed for aligning temporal sequences for speech recognition but has since been used to compare gene expression kinetics from many different biological processes [@cacchiarelli_aligningsinglecelldevelopmental_2018; @kanton_organoidsinglecellgenomic_2019; @mcfaline-figueroa_pooledsinglecellgenetic_2019; @alpert_alignmentsinglecelltrajectories_2018].
+We evaluate the performance of DTW by simulating two linear trajectories with slightly different simulation kinetics and assessing the accuracy of the alignment by DTW.
+We compare the performance of DTW against a version of DTW where gene expression is first smoothed, for varying degrees of noise added to the gene expression (Figure\ \ref{fig:traj_align}).
+We observe that DTW+smoother performs significantly better than the unsmoothed version across all levels of noise.
 
-Depending on how the GRN is designed, different cellular developmental processes can be simulated.
-dyngen includes generators of GRNs which result in many different developmental topologies (Figure\ \ref{fig:example_runs}), including branching, converging, cyclic and even disconnected.
-Custom-defined GRNs offer more fine-grained control over the simulation.
+**RNA velocity** methods use the relative ratio between pre-mRNA and mature mRNA reads to predict the velocity at which the RNA expression of genes is increasing or decreasing [@zeisel_coupledpremrnamrna_2011;@lamanno_rnavelocitysingle_2018].
+Already two algorithms are currently available for estimating the RNA velocity vector from spliced and unspliced counts: velocyto [@lamanno_rnavelocitysingle_2018] and scvelo [@bergen_generalizingrnavelocity_2019].
+Yet, to date, no quantitative assessment of their accuracy has been performed, mainly due to the difficulty in obtaining real ground-truth data to do so.
+In contrast, the ground-truth RNA velocity can be easily extracted from a dyngen, as it is possible to store the rate at which mRNA molecules are being transcribed and degraded (called the propensity) at any particular point in time.
+We executed scvelo and velocyto (with 6 different parameter settings in total) on 102 datasets with varying degrees of difficulty (easy, medium, hard) and a variety of backbones (including linear, bifurcating, cyclic, disconnected).
+We evaluated the predictions using two metrics (Figure\ \ref{fig:velocity}), one which directly compares the predicted RNA velocity of each gene with the ground-truth RNA velocity (called the "velocity correlation"), and one which compares the direction of the ground-truth trajectory embedded in a dimensionality reduction with the average RNA velocity of cells in that neighbourhood (called the "velocity arrow cosine").
+We found that in almost all cases, both velocyto and scvelo obtained high velocity arrow cosine, meaning that the overview obtained by embedding RNA velocity arrows in a 2D or 3D dimensionality reduction allows users to correctly identify the general progression of cells.
+However, depending on the difficulty of the dataset, the correlation between the predicted RNA velocity and the ground-truth RNA velocity varies between high (>0.75) to low (<0.25).
+For this particular metric, the dynamic estimation of velocyto performs significantly worse than any of the other prediction methods.
 
-\begin{figure}[ht]
-    \centering
-    \includegraphics[width=0.6\textwidth]{result_files/fig3_showcase_backbones/overview.pdf}
+**Cell-specific network inference** (CSNI) methods predict not only which transcription factors regulate which target genes, but also aim to identify how active each interaction is in each of the cells, since interactions can be turned off and on depending on the cellular state.
+While a few pioneering CSNI approaches have already been developed [@aibar_scenicsinglecellregulatory_2017; @kuijjer_estimatingsamplespecificregulatory_2019; @liu_personalizedcharacterizationdiseases_2016], a quantitative assessment of the performance is until now lacking.
+This is not surprising, as neither real nor in silico datasets of cell-specific or even cell-type-specific interactions exist that are large enough so that it can be used as a ground-truth for evaluating CSNI methods.
+Extracting the ground-truth dynamic network in dyngen is straightforward though, given that we can calculate how target gene expression would change without the regulator being present.
+We used this ground-truth to compare the performance of three CSNI methods: LIONESS [@kuijjer_estimatingsamplespecificregulatory_2019], SSN [@liu_personalizedcharacterizationdiseases_2016] and SCENIC [@aibar_scenicsinglecellregulatory_2017].
+We calculated the AUROC and AUPR score for each cell individually.
+Computing the mean AUROC and AUPR per dataset showed that pySCENIC significantly outperforms LIONESS + Pearson, which in turn outperforms SSN*.
+
+<!-- DISCUSSION -->
+As is, dyngen's single-cell simulations can be used to evaluate common single-cell omics computational methods such as clustering, batch correction, trajectory inference, and network inference.
+However, the framework is flexible enough to be adaptable to a broad range of applications, including methods that integrate clustering, network inference, and trajectory inference.
+In this respect, dyngen may promote the development of new tools in the single-cell field similarly as other simulators have done in the past [@schaffter_genenetweaversilicobenchmark_2011; @ewing_combiningtumorgenome_2015].
+
+dyngen ultimately allows anticipating technological developments in single-cell multi-omics.
+In this way, it is possible to design and evaluate the performance and robustness of new types of computational analyses before experimental data becomes available.
+In addition, it could also be used to compare which experimental protocol is the most cost-effective in producing qualitative and robust results in downstream analysis.
+
+Currently, dyngen focuses on simulating cells as standalone entities that are well mixed.
+Splitting up the simulation space into separate subvolumes could pave the way to better study key cellular processes such as cell division, intercellular communication, and migration [@smith_spatialstochasticintracellular_2019].
+
+# Availability
+dyngen is available as an R package on GitHub at [github.com/dynverse/dyngen](https://github.com/dynverse/dyngen).
+The analyses performed in this manuscript are available on GitHub at [github.com/dynverse/dyngen_manuscript](https://github.com/dynverse/dyngen_manuscript).
+
+# Author contributions
+
+* W.S. and R.C. designed the study.
+* R.C., W.S., and L.D. performed the experiments and analysed the data.
+* R.C. and W.S. implemented the dyngen software package.
+* R.C., W.S., L.D., and Y.S. wrote the manuscript.
+* Y.S. supervised the project.
+
+\printbibliography
+
+# Supplementary Figures
+
+\newrefsection
+\setcounter{table}{0}
+\renewcommand{\thetable}{S\arabic{table}}
+\setcounter{figure}{0}
+\renewcommand{\thefigure}{S\arabic{figure}}
+
+\newcommand{\yes}{\ding{51}}
+\newcommand{\no}{}
+\definecolor{light-gray}{rgb}{.85,.85,.85}
+\newcommand{\grayline}{\arrayrulecolor{light-gray}\cline{3-7}\arrayrulecolor{black}}
+\newcommand{\blackline}{\arrayrulecolor{black}\cline{3-7}}
+\newcommand{\method}[1]{#1}
+
+\begin{table}[H]
     \caption{
-        \textbf{Multiple executions of dyngen with different predefined backbones.} From each simulation of about 200 genes, 1000 cells were sampled.
-        \textbf{A:}~Linear. \textbf{B:}~Bifurcating. \textbf{C:}~Converging.
-        \textbf{D:}~Cyclic. \textbf{E:}~Bifurcating loop. \textbf{F:}~Bifurcating converging.
-        \textbf{G:}~Bifurcating cycle. \textbf{H:}~Consecutive branching. \textbf{I:}~Disconnected.
-    }
-    \label{fig:example_runs}
-\end{figure}
+      \textbf{Feature comparison of single-cell synthetic data generators and simulators.} $ ^1$ As showcased by vignette.
+$^2$ As showcased by Van den Berge et al.
+\cite{vandenberge_trajectorybaseddifferentialexpression_2019}.
+    } \label{tab:comparison}
+    \small
+    \begin{tabular}{p{.25cm}l|*{5}{>{\centering\arraybackslash}p{1.25cm}|}}
+    \blackline
+    & \method{} & \method{splatter} & \method{powsimR} & \method{PROSSTT} & \method{SymSim} & \method{dyngen} \\
+    \blackline
+    \multicolumn{7}{l}{\textbf{Available modality outputs}} \\
+    \blackline
+    - & mRNA expression & \yes & \yes & \yes & \yes & \yes \\ \grayline
+    - & Pre-mRNA expression & \no & \no & \no & \no & \yes \\ \grayline
+    - & Protein expression & \no & \no & \no & \no & \yes \\ \grayline
+    - & Promotor activity & \no & \no & \no & \yes & \yes \\ \grayline
+    - & Reaction activity & \no & \no & \no & \no & \yes \\
+    \blackline
+    \multicolumn{7}{l}{\textbf{Available ground-truth outputs}} \\
+    \blackline
+    - & True counts & \yes & \no & \no & \yes & \yes \\ \grayline
+    - & Cluster labels & \yes & \yes & \no & \yes & \yes \\ \grayline
+    - & Trajectory & \yes & \no & \yes $ ^1$ & \yes & \yes \\ \grayline
+    - & Batch labels & \yes & \no & \no & \no & \yes $ ^1$\\ \grayline
+    - & Differential expression & \no & \no & \no & \yes & \yes $ ^2$ \\ \grayline
+    - & Knocked down regulators & \no & \no & \no & \no & \yes \\ \grayline
+    - & Regulatory network & \no & \no & \no & \yes & \yes \\ \grayline
+    - & Cell-specific regulatory network & \no & \no & \no & \no & \yes \\
+    \blackline
+    \multicolumn{7}{l}{\textbf{Emulate experimental effects}} \\
+    \blackline
+    - & Single-cell RNA sequencing & \yes & \yes & \yes & \yes & \yes \\ \grayline
+    - & Batch effects & \yes & \no & \no & \no & \yes $ ^1$ \\ \grayline
+    - & Knockdown experiment & \no & \no & \no & \no & \yes \\ \grayline
+    - & Time-series & \no & \no & \no & \no & \yes \\
+    - & Snapshot & \no & \no & \no & \no & \yes \\
+    \blackline
+    \multicolumn{7}{l}{\textbf{Evaluation applications}} \\
+    \blackline
+    - & Clustering & \yes & \yes & \no & \yes & \no \\ \grayline
+    - & Trajectory inference & \yes & \no & \yes & \yes & \yes \\ \grayline
+    - & Network inference & \no & \no & \no & \yes & \yes \\ \grayline
+    - & Cell-specific network inference & \no & \no & \no & \no & \yes \\ \grayline
+    - & Differential expression & \yes & \no & \no & \no & \yes $ ^2$ \\ \grayline
+    - & Batch effect correction & \yes & \no & \no & \no & \yes \\ \grayline
+    - & RNA Velocity & \no & \no & \no & \no & \yes \\ \grayline
+    - & Trajectory alignment & \no & \no & \no & \no & \yes \\
+    \blackline
+    %\multicolumn{7}{l}{\textbf{Showcased dataset}} \\
+    %\blackline
+    %- & Number of cells &  &  &  &  & \\ \grayline
+    %- & Number of genes &  &  &  &  & \\ \grayline
+    %- & Execution time & & & & & \\
+    \blackline
 
-Together, these qualities allow it to be applicable in benchmarking a broad range of use-cases.
-In practice, dyngen has already successfully been used to evaluate trajectory inference [@saelens_comparisonsinglecelltrajectory_2019], trajectory-based differential expression [@vandenberge_trajectorybaseddifferentialexpression_2019], and network inference [@pratapa_benchmarkingalgorithmsgene_2019] methods.
-To demonstrate this point even further, we apply dyngen on several promising novel computational approaches for which quantitative assessment of the performance was until now lacking.
+\end{tabular}
+\end{table}
 
-## RNA velocity
-
-In eukaryotes, a gene is first transcribed to a pre-mRNA and subsequently spliced into mature mRNA.
-Because reads coming from both unspliced and spliced transcripts are observed in expression data, the relative ratio between the two can tell us something about which genes are increasing, decreasing or remaining the same [@zeisel_coupledpremrnamrna_2011;@lamanno_rnavelocitysingle_2018].
-To determine this, some parameters have to be estimated to determine which fraction of unspliced and spliced mRNAs correspond to an increase or decrease.
-The estimation of these parameters makes some assumptions and can be handled in different ways in the two main algorithms that are now available for RNA velocity estimation: velocyto [@lamanno_rnavelocitysingle_2018] and scvelo [@bergen_generalizingrnavelocity_2019].
-It can be difficult to obtain ground truth data to benchmark these algorithms, given that it would require continuous data of transcriptional dynamics in individual cells.
-On the other hand, the ground truth velocity is rapidly extracted from the dyngen model, by looking at whether each transcript is currently increasing or decreasing in expression.
-
-We tested scvelo and velocyto on 8 datasets containing linear, bifurcating, disconnected and cyclic trajectories, and varied the main parameter settings in which they estimate the velocity.
-We found that the original velocyto implementation, which assumes that the velocity remains constant in some cells, performed the best across all datasets.
-The dynamical estimation of velocyto, as implemented in scvelo, performed the worst of all parameter settings.
-This was mainly due to scvelo overestimating the dynamics of a gene, especially towards upregulation, while velocyto correctly estimated not only when a gene changes, but also when it remained in a steady state.
-
-\begin{figure}[ht]
+\begin{figure}[H]
     \centering
@@
 \textbf{C:} The velocity stream plots produced from the predictions of each method, as generated by scvelo.
-\textbf{D:} The predictions scored by two different metrics, the velocity correlation and the vector cosine similarity.
+\textbf{D:} The predictions scored by two different metrics, the velocity correlation and the velocity arrow cosine.
 The velocity correlation is the correlation between the ground-truth velocity (A, right) and the predicted velocity (B).
-The vector cosine similarity is the cosine similarity between the direction of segments of the ground-truth trajectory (A, left) and the RNA velocity values calculated at those points (C).
+The velocity arrow cosine is the cosine similarity between the direction of segments of the ground-truth trajectory (A, left) and the RNA velocity values calculated at those points (C).
     }
@@
 
-## Cell-specific network inference
-
-Cell-specific network inference (CSNI) methods\footnote{Different terms are commonly used when dealing with data from a particular source.
-For example, single-cell NI when applied to single-cell transcriptomics data; sample-specific NI when applied to bulk transcriptomics; patient-derived NI when applied to bulk profiles of patients.
-A more generalised variant of CSNI is casewise NI, which does not specify the type of data which is being analysed.} predict not only which transcription factors regulate which target genes (Figure\ \ref{fig:scgrn}A, top left), but also how active each interaction is in every case (Figure\ \ref{fig:scgrn}A).
-
-While a few pioneering CSNI approaches have already been developed [@aibar_scenicsinglecellregulatory_2017; @kuijjer_estimatingsamplespecificregulatory_2019; @liu_personalizedcharacterizationdiseases_2016], a quantitative assessment of the performance is until now lacking.
-This is not surprising, as neither real nor in silico datasets of cell-specific or even cell-type-specific interactions exist that are large enough so that it can be used as a ground-truth for evaluating CSNI methods.
-Extracting the ground-truth dynamic network in dyngen is straightforward though, given that we can calculate how target gene expression would change without the regulator being present.
-
-We used this ground-truth to compare the performance of three CSNI methods (Figure\ \ref{fig:scgrn}B).
-We calculated the AUROC and AUPR score -- which are common metrics for NI benchmarking -- for each cell individually.
-Computing the mean AUROC and AUPR per dataset showed that pySCENIC significantly outperforms LIONESS + Pearson, which in turn outperforms SSN*.
-
-\begin{figure}[htb!]
+\begin{figure}[H]
     \centering
@@
 
-## Trajectory alignment
-
-Trajectory alignment allows studying the differences between the same trajectories from different samples, as shown in Figure \ref{fig:traj_align} A.
-For example, the cell developmental process of a patient could be compared to that of a healthy control to detect the transcriptomic differences of a particular lineage.
-
-Dynamic Time Warping (DTW) [@giorgino_computingvisualizingdynamic_2009] is most commonly used to align linear trajectories.
-DTW is a technique originating in the field of speech recognition and aligns temporal sequences by creating a warping path between the sequences that indicate which sequence must be dilated or contracted to best match the other one.
-DTW can also be used in combination with smoothed pseudo-cells [@alpert_alignmentsinglecelltrajectories_2018].
-
-Trajectory alignment, using DTW, has been used to compare gene expression kinetics resulting from different biological processes [@cacchiarelli_aligningsinglecelldevelopmental_2018], to compare human, chimpanzee and macaque neuronal development [@kanton_organoidsinglecellgenomic_2019], to find differences in gene regulation in the presence of certain growth factors [@mcfaline-figueroa_pooledsinglecellgenetic_2019], and to compare human and mouse embryogenesis [@alpert_alignmentsinglecelltrajectories_2018].
-
-The pseudotime values produced by dyngen are comparable across experiments, which allows us to evaluate the accuracy of an alignment technique.
-We use DTW, but process the trajectories in 3 different ways: we either used the original cells, we used pseudocells with a smoothed gene expression, or we used every tenth cell in the trajectory.
-An example of the different results obtained is shown in \ref{fig:traj_align} B, C and D.
-We can see that the smoothed way of processing recovers the signal in the data best, it finds a diagonal warping path, which indicates that the two trajectories develop in a similar fashion.
-We test this on 10 pairs of two datasets, where an increasing amount of noise was added to the count matrix. As shown in Figure \ref{fig:traj_align} E, the smoothed pseudocells perform best.
-
-\begin{figure}[htb!]
+\begin{figure}[H]
     \centering
@@
     \textbf{dyngen allows benchmarking of trajectory alignment methods.} 
- \textbf{A}: Example of a trajectory alignment process. The pseudotimes of individual inferred trajectories can be aligned to indicate differences between the trajectories. 
-  \textbf{B, C, D}: Shows the accumulated distance matrices obtained after using DTW on two trajectories where noise (noise level of 0.4) was added to the count matrix.
-In B the complete count matrices were used to perform the alignment.
-In C, each 10th cell was used to perform the alignment.
-In D, smoothed pseudocells were used.
-  \textbf{E:} Shows the influence of added noise to the different processing methods. We can see that smoothing performs best in noisy circumstances.
+ \textbf{A, B}: Two samples containing a linear trajectory, generated by dyngen.
+  \textbf{C, D}: Result of the DTW alignment on the samples in \textbf{A} and \textbf{B}.
+In \textbf{C}, the individual mappings of the alignment between cells are shown.
+In \textbf{D}, the accumulated distance matrix between the two trajectories is shown, including the black warping path, corresponding to these cell to cell alignments in \textbf{C}.
+  \textbf{E, F}: Shows the accumulated distance matrices obtained after using DTW on two trajectories where noise (noise level of 0.4) was added to the count matrix.
+In E the complete count matrices were used to perform the alignment.
+In F, smoothed pseudocells were used.
+  \textbf{G:} Shows the influence of added noise to the different processing methods. We can see that DTW + smoothing performs best in noisy circumstances.
     }
@@
 \end{figure}
-
-# Discussion {#sec:dyngen-discussion}
-As is, dyngen's single-cell simulations can be used to evaluate common single-cell omics computational methods such as clustering, batch correction, trajectory inference, and network inference.
-However, the combined effect of these advantages results in a framework that is flexible enough to adapt to a broad range of applications.
-This may include methods that integrate clustering, network inference, and trajectory inference.
-In this respect, dyngen may promote the development of new tools in the single-cell field similarly as other simulators have done in the past [@schaffter_genenetweaversilicobenchmark_2011; @ewing_combiningtumorgenome_2015].
-
-dyngen ultimately allows anticipating technological developments in single-cell multi-omics.
-In this way, it is possible to design and evaluate the performance and robustness of new types of computational analyses before experimental data becomes available.
-In addition, it could also be used to compare which experimental protocol is the most cost-effective in producing qualitative and robust results in downstream analysis.
-
-Currently, dyngen focuses on simulating cells as standalone entities that are well mixed.
-Splitting up the simulation space into separate subvolumes could pave the way to better study key cellular processes such as cell division, intercellular communication, and migration [@smith_spatialstochasticintracellular_2019].
 
@@
 
-Several examples of module networks are given (Figure\ \ref{fig:example_backbones_onlymodules}).
+Several examples of module networks are given in Figure\ \ref{fig:example_backbones_onlymodules}.
 A simple chain of modules (where one module upregulates the next) results in a _linear_ process.
@@
 The currently existing TFs are mapped to regulators in the FANTOM5 GRN.
-The targets are drawn from the FANTOM5 GRN, weighted by their page rank value.
+The targets are drawn from the FANTOM5 GRN weighted by their page rank value, to create an induced GRN.
 For each target, at most $x$ regulators are sampled from the induced FANTOM5 GRN (default $x=5$).
@@
 Housekeeping genes are completely separate from any TFs or target genes.
-A user-defined set of housekeeping genes are also sampled from the FANTOM5 GRN.
+A user-defined set of housekeeping genes is also sampled from the FANTOM5 GRN.
 The interactions of the FANTOM5 GRN are first subsampled such that the maximum in-degree of each gene is $x$ (default $x=5$).
@@
     \caption{
-      \textbf{Reactions affecting the abundance levels of pre-mRNA $\x G$, mature mRNA $\y G$ and proteins $\z G$ of gene $G$.} Define the set of regulators of $G$ as $\rs{G}$, the set of upregulating regulators of $G$ as $\rp G$, and the set of downregulating regulators of $G$ as $\rn G$.
+      \textbf{Reactions affecting the abundance levels of pre-mRNA $\x G$, mature mRNA $\y G$ and proteins $\z G$ of gene $G$.} Define the set of regulators of $G$ as $\rs{G}$, the set of upregulating regulators of $G$ as $\rp G$, and the set of down-regulating regulators of $G$ as $\rn G$.
 Parameters used in the propensity formulae are defined in Table \ref{tab:reaction_params}.
@@
 For this, the cells are sampled from the simulations and molecules are sampled for each of the cells.
-Real datasets are used in order to achieve similar data characteristics.
+Gene capture rates and library sizes are empirically derived from real datasets as to match real technical variation.
 
@@
 
+## Simulating batch effects {#sec:dyngen-batcheffect}
+Simulating batch effects can be performed in multiple ways.
+One such way is to perform the first two steps of the creation of a dyngen model (defining the module network and generating the GRN).
+For each desired batch, create a separate model for which random kinetics are generated and perform all subsequent dyngen steps (convert to reactions, simulate gold standard, simulate single cells, simulate experiment).
+Since each separate model has different underlying kinetics, the combined output will resemble having batch effects.
+
 ## Determining the ground-truth trajectory {#sec:dyngen-groundtruth}
@@
     \includegraphics[width=0.8\textwidth]{result_files/example_backbones}
-    \caption{Examples of the ground-truth state networks which need to be provided alongside the module network.}
+    \caption{\textbf{Examples of the ground-truth state networks which need to be provided alongside the module network.}}
     \label{fig:example_backbones}
@@
 
-Determining the regulatory effect for all interactions and cells in the dataset yields the complete cell-specific ground-truth GRN (Figure\ \ref{fig:cell-specific_grn}).
-The regulatory effect lie between $[-1, 1]$, where -1 represents complete inhibition of $T$ by $R$, 1 represents maximal activation of $T$ by $R$, and 0 represents inactivity of the regulatory interaction between $R$ and $T$.
-
+Determining the regulatory effect for all interactions and cells in the dataset yields the complete cell-specific ground-truth GRN <!--(Figure\ \ref{fig:cell-specific_grn})-->.
+The regulatory effect lies between $[-1, 1]$, where -1 represents complete inhibition of $T$ by $R$, 1 represents maximal activation of $T$ by $R$, and 0 represents inactivity of the regulatory interaction between $R$ and $T$.
+<!--
 \begin{figure}[htb!]
@@
 \end{figure}
+-->
 
@@
 
-The AUROC and AUPR metrics are common metrics for evaluating a predicted GRN with a ground-truth GRN.
-To compare a predicted cell-specific GRN with the ground-truth cell-specific GRN, the top 10'000 interactions per cell were retained.
-For each cell-specific network, the AUROC and AUPR were calculated.
+Cell-specific network inference (CSNI) predicts the regulatory network of every individual cell.
+We evaluate each cell-specific regulatory network with typical metrics for network inference: the Area Under the Receiver Operating Characteristics-curve (AUROC) and Area Under the Precision Recall-curve (AUPR).
+We evaluate three CSNI methods by computing the mean AUROC and AUPR across all cells.
+
+The Area Under the Receiver Operating Characteristic-curve (AUROC) and Area Under the Precision-Recall curve (AUPR) metrics are common metrics for evaluating a predicted GRN with a ground-truth GRN [@marbach_generatingrealisticsilico_2009].
+To compare a predicted cell-specific GRN with the ground-truth cell-specific GRN, the top 10'000 interactions per cell is retained, and the mean AUROC and AUPR scores are calculated across all cells.
 
@@
 
-We extracted a ground truth RNA velocity by subtracting for each mRNA molecule the propensity of its production by the propensity of its degradation.
+We calculated two evaluation metrics: the velocity correlation and the velocity arrow cosine.
+For the velocity correlation, we extracted a ground truth RNA velocity by subtracting for each mRNA molecule the propensity of its production by the propensity of its degradation.
 If the expression of an mRNA will increase in the future, this value is positive, while it is negative if it is going to decrease.
-For each gene, we compared the ground truth velocity with the observed velocity by calculating the Spearman rank correlation.
+For each gene, we determined its velocity correlation by calculating the Spearman rank correlation between the ground truth velocity with the observed velocity.
+For the velocity arrow cosine, we determined a set of 100 trajectory waypoints uniformly spread on the trajectory.
+For each waypoint, we weighted each cell based on a Gaussian kernel on its geodesic distance from the waypoint.
+These weights were used to calculate a weighted average velocity vector of each waypoint.
+We then calculated for each waypoint the cosine similarity between this velocity vector and the known direction of the trajectory.
 
@@
 
-## Comparison of trajectory alignment with added noise {#sec:dyngen-tacompare}
+## Comparison of trajectory alignment with increasing levels of noise {#sec:dyngen-tacompare}
 
-We generated 10 base datasets that contain a linear trajectory using dyngen.
-For each dataset, we performed the cell generation and experiment generation twice.
-This results in 20 datasets, with 10 pairs of similar trajectories.
+10 base datasets, containing a linear trajectory, were generated using dyngen.
+For each dataset, cells and experiments were generated twice, resulting in 20 paired datasets containing a similar trajectory with slightly different kinetics.
 For each of these pairs of datasets, we generated 10 progressively noisier pairs, in which we added noise to the complete count matrix.
@@
 \begin{equation}
- countmatrix + \frac{1}{\sqrt{2\pi}}qie^{\frac{x^2}{2(qi)^2}}  \quad \forall i \in \{0.1, 0.2, ... 1.0\}
+ \textrm{countmatrix} + \frac{1}{\sqrt{2\pi}}qie^{\frac{x^2}{2(qi)^2}}  \quad \forall i \in \{0.1, 0.2, ... 1.0\}
 \end{equation}
 
-We aligned each of these 150 pairs of trajectories (the first trajectory in the pair with the second trajectory in the pair) in 3 different ways.
-Either we used the original count matrix, we used 1/10th of the cells (according to the ordering of the trajectory, so the 1st, 10th, 20th, ...
-cell of the trajectory) or we used the smoothed pseudocells.
-The pseudocells are generated at regular points across the pseudotime of the trajectory. The gene expression of each pseudocell is inferred using a Gaussian kernel with a window size of 0.05.
+We aligned each of these 100 pairs of trajectories (the first trajectory in the pair with the second trajectory in the pair) using Dynamic Time Warping (DTW) [@giorgino_computingvisualizingdynamic_2009].
+DTW is designed to align temporal sequences by dilating or contracting the sequences to best match each other.
+We compared the alignment of DTW with the alignment of DTW after first smoothing the expression along the trajectory using a Gaussian kernel with window size 0.05.
 
-Each of the cells or pseudocells received a pseudotime from dyngen.
-This time can be compared: cells from the first trajectory in the pair should be aligned closely to cells from the second trajectory with similar pseudotimes.
-We used the sum of differences between the pseudotimes of the aligned cells to determine the distance between the two trajectories, which should ideally be 0.
-In order to compare the alignments with 1000 cells per trajectory and the ones with 100 cells per trajectory, we divide the distance of those alignments by 10.
-
-# Availability
-dyngen is available as an R package on GitHub [github.com/dynverse/dyngen](https://github.com/dynverse/dyngen).
-The dyngen codebase is still under development and is thus likely subject to change, including the adding or removing of functionality, renaming functions or parameters, and changing default parameter values.
-The analyses performed in this manuscript are available on GitHub [github.com/dynverse/dyngen_manuscript](https://github.com/dynverse/dyngen_manuscript).
-
-# Author contributions
-
-* W.S. and R.C. designed the study.
-* R.C., W.S., and L.D. performed the experiments and analysed the data.
-* R.C. and W.S. implemented the dyngen software package.
-* R.C. and W.S. wrote the original manuscript.
-* L.D. wrote the section on trajectory alignment.
-* R.C., W.S., L.D., and Y.S. reviewed and edited the manuscript.
-* Y.S. supervised the project.
+To evaluate a trajectory alignment method on a paired dataset we computed the geodesic distances of each cell from the start of the trajectory.
+The geodesic distances between the pair of datasets are directly comparable to each other, and thus cells from the first trajectory should align relatively closely to cells from the second trajectory.
+We after aligning the cells, we calculate the mean absolute difference between the geodesic distances of the pairs of cells, which should ideally be 0.

### 2020-06-13 analysis/summary/figure_1.R

@@ -311,15 +311,11 @@ g7d <- dynplot::plot_dimred(dataset, dimred = dimred, size_milestones = 3, size_
   labs(title = "Trajectory inference")
 
 exp7a <- start_analysis("usecase_trajectory_alignment")
-li7a <- read_rds(exp7a$result("cell_mappings_data.rds"))
-
-g7a <- ggplot() +
-  geom_segment(data = li7a$leftover_combinatie, mapping = aes(x = x, y = y, xend = x2, yend = y2), alpha = 1) +
-  geom_point(data = li7a$alles, mapping = aes(x = x1, y = y, colour = as.factor(color)), size = 3.5, show.legend = F) +
-  scale_colour_manual(values = c("#fd8d3c", "#6baed6"), label = "") +
-  scale_x_continuous(breaks = c(0, 1), labels = c("Start", "End")) +
-  theme_void() +
+g7a <- read_rds(exp7a$result("usecase_separateplots.rds"))$prediction +
+  theme_classic() +
   theme(
+    legend.position = "none",
+    text = element_text(family = "Helvetica"),
     axis.title = element_text(),
     axis.title.y = element_blank(),
     axis.line = element_line(),
@@ -331,7 +327,6 @@ g7a <- ggplot() +
   ) +
   labs(x = "Simulation time", title = "Trajectory alignment") +
   coord_cartesian()
-g7a
 
 exp7b <- start_analysis("usecase_rna_velocity")
 g7b <- read_rds(exp7b$result("usecase_separateplots.rds"))$prediction +

### 2020-06-13 manuscript.Rmd
update

@@
 ```
-<!-- link naar de oude tekst: https://docs.google.com/document/d/15ZkzzB-XSYFfZutO4YPkUIw6iUvxJS92fSPfzqR1cLA/edit -->
 
@@
 <!-- DISCUSSION -->
-As is, dyngen's single-cell simulations can be used to evaluate common single-cell omics computational methods such as clustering, batch correction, trajectory inference, and network inference.
+In summary, dyngen's single-cell simulations can be used to evaluate common single-cell omics computational methods such as clustering, batch correction, trajectory inference, and network inference.
 However, the framework is flexible enough to be adaptable to a broad range of applications, including methods that integrate clustering, network inference, and trajectory inference.
 In this respect, dyngen may promote the development of new tools in the single-cell field similarly as other simulators have done in the past [@schaffter_genenetweaversilicobenchmark_2011; @ewing_combiningtumorgenome_2015].
-
-dyngen ultimately allows anticipating technological developments in single-cell multi-omics.
-In this way, it is possible to design and evaluate the performance and robustness of new types of computational analyses before experimental data becomes available.
-In addition, it could also be used to compare which experimental protocol is the most cost-effective in producing qualitative and robust results in downstream analysis.
-
-Currently, dyngen focuses on simulating cells as standalone entities that are well mixed.
+Additionally, one could anticipate technological developments in single-cell multi-omics.
+In this way, dyngen allows designing and evaluating the performance and robustness of new types of computational analyses before experimental data becomes available, comparing which experimental protocol is the most cost-effective in producing qualitative and robust results in downstream analysis.
+One major assumption of dyngen is that cells are regarded as standalone entities that are well mixed.
 Splitting up the simulation space into separate subvolumes could pave the way to better study key cellular processes such as cell division, intercellular communication, and migration [@smith_spatialstochasticintracellular_2019].

### 2020-06-13 manuscript.Rmd
fix supp fig ordering

@@
 Extracting the ground-truth dynamic network in dyngen is straightforward though, given that we can calculate how target gene expression would change without the regulator being present.
-We used this ground-truth to compare the performance of three CSNI methods: LIONESS [@kuijjer_estimatingsamplespecificregulatory_2019], SSN [@liu_personalizedcharacterizationdiseases_2016] and SCENIC [@aibar_scenicsinglecellregulatory_2017].
+We used this ground-truth to compare the performance of three CSNI methods (Figure\ \ref{fig:scgrn}): LIONESS [@kuijjer_estimatingsamplespecificregulatory_2019], SSN [@liu_personalizedcharacterizationdiseases_2016] and SCENIC [@aibar_scenicsinglecellregulatory_2017].
 We calculated the AUROC and AUPR score for each cell individually.
@@
     \centering
+    \includegraphics[width=.8\linewidth]{result_files/usecase_trajectory_alignment/usecase.pdf}
+    \caption{
+    \textbf{dyngen allows benchmarking of trajectory alignment methods.} 
+ \textbf{A, B}: Two samples containing a linear trajectory, generated by dyngen.
+  \textbf{C, D}: Result of the DTW alignment on the samples in \textbf{A} and \textbf{B}.
+In \textbf{C}, the individual mappings of the alignment between cells are shown.
+In \textbf{D}, the accumulated distance matrix between the two trajectories is shown, including the black warping path, corresponding to these cell to cell alignments in \textbf{C}.
+  \textbf{E, F}: Shows the accumulated distance matrices obtained after using DTW on two trajectories where noise (noise level of 0.4) was added to the count matrix.
+In E the complete count matrices were used to perform the alignment.
+In F, smoothed pseudocells were used.
+  \textbf{G:} Shows the influence of added noise to the different processing methods. We can see that DTW + smoothing performs best in noisy circumstances.
+    }
+    \label{fig:traj_align}
+\end{figure}
+
+\begin{figure}[H]
+    \centering
     \includegraphics[width=\linewidth]{result_files/usecase_rna_velocity/usecase.pdf}
@@
     \label{fig:scgrn}
-\end{figure}
-
-\begin{figure}[H]
-    \centering
-    \includegraphics[width=.8\linewidth]{result_files/usecase_trajectory_alignment/usecase.pdf}
-    \caption{
-    \textbf{dyngen allows benchmarking of trajectory alignment methods.} 
- \textbf{A, B}: Two samples containing a linear trajectory, generated by dyngen.
-  \textbf{C, D}: Result of the DTW alignment on the samples in \textbf{A} and \textbf{B}.
-In \textbf{C}, the individual mappings of the alignment between cells are shown.
-In \textbf{D}, the accumulated distance matrix between the two trajectories is shown, including the black warping path, corresponding to these cell to cell alignments in \textbf{C}.
-  \textbf{E, F}: Shows the accumulated distance matrices obtained after using DTW on two trajectories where noise (noise level of 0.4) was added to the count matrix.
-In E the complete count matrices were used to perform the alignment.
-In F, smoothed pseudocells were used.
-  \textbf{G:} Shows the influence of added noise to the different processing methods. We can see that DTW + smoothing performs best in noisy circumstances.
-    }
-    \label{fig:traj_align}
 \end{figure}

### 2020-06-13 manuscript.Rmd
update

@@
 Simulation engines are also increasingly important when it comes to finding alternatives to animal models, for example for drug testing and precision medicine.
-In such scenarios, cellular simulations can act as digital twins, offering unlimited experimentation in silico.
+In such scenarios, cellular simulations can act as digital twins, offering unlimited experimentation _in silico_ [@bjornsson_digitaltwinspersonalize_2019].
 
@@
       \textbf{A:} Changes in abundance levels are driven strictly by gene regulatory reactions.
-      \textbf{B:} The input GRN is defined such that it models a dynamic process of interest.
+      \textbf{B:} The input Gene Regulatory Network (GRN) is defined such that it models a dynamic process of interest.
       \textbf{C:} The reactions define how abundance levels of molecules change at any particular time point.

### 2020-06-13 render_manuscript.R

@@ -3,11 +3,6 @@ library(googledrive)
 library(rmarkdown)
 library(textreadr)
 
-# copy zotero library if on rcannood's computer
-if (Sys.info()[["user"]] == "rcannood") {
-  file.copy("~/Workspace/library.bib", "library.bib", overwrite = TRUE)
-}
-
 # download as docx instead of txt, because otherwise comments get pushed into the document
 httr::set_config(httr::config(http_version = 0)) # avoid http2 framing layer bug
 drive <- drive_download(as_id("15ZkzzB-XSYFfZutO4YPkUIw6iUvxJS92fSPfzqR1cLA"), type = "docx", overwrite = TRUE, path = tempfile())

### 2020-06-13 manuscript.Rmd
update

@@
   \usepackage{float}
+  \usepackage{caption}
+  \captionsetup[table]{position=bottom}
   \setmainfont [Path = fonts/,
@@
 \printbibliography
-
-# Supplementary Figures
-
-\newrefsection
-\setcounter{table}{0}
-\renewcommand{\thetable}{S\arabic{table}}
-\setcounter{figure}{0}
-\renewcommand{\thefigure}{S\arabic{figure}}
-
-\newcommand{\yes}{\ding{51}}
-\newcommand{\no}{}
-\definecolor{light-gray}{rgb}{.85,.85,.85}
-\newcommand{\grayline}{\arrayrulecolor{light-gray}\cline{3-7}\arrayrulecolor{black}}
-\newcommand{\blackline}{\arrayrulecolor{black}\cline{3-7}}
-\newcommand{\method}[1]{#1}
-
-\begin{table}[H]
-    \caption{
-      \textbf{Feature comparison of single-cell synthetic data generators and simulators.} $ ^1$ As showcased by vignette.
-$^2$ As showcased by Van den Berge et al.
-\cite{vandenberge_trajectorybaseddifferentialexpression_2019}.
-    } \label{tab:comparison}
-    \small
-    \begin{tabular}{p{.25cm}l|*{5}{>{\centering\arraybackslash}p{1.25cm}|}}
-    \blackline
-    & \method{} & \method{splatter} & \method{powsimR} & \method{PROSSTT} & \method{SymSim} & \method{dyngen} \\
-    \blackline
-    \multicolumn{7}{l}{\textbf{Available modality outputs}} \\
-    \blackline
-    - & mRNA expression & \yes & \yes & \yes & \yes & \yes \\ \grayline
-    - & Pre-mRNA expression & \no & \no & \no & \no & \yes \\ \grayline
-    - & Protein expression & \no & \no & \no & \no & \yes \\ \grayline
-    - & Promotor activity & \no & \no & \no & \yes & \yes \\ \grayline
-    - & Reaction activity & \no & \no & \no & \no & \yes \\
-    \blackline
-    \multicolumn{7}{l}{\textbf{Available ground-truth outputs}} \\
-    \blackline
-    - & True counts & \yes & \no & \no & \yes & \yes \\ \grayline
-    - & Cluster labels & \yes & \yes & \no & \yes & \yes \\ \grayline
-    - & Trajectory & \yes & \no & \yes $ ^1$ & \yes & \yes \\ \grayline
-    - & Batch labels & \yes & \no & \no & \no & \yes $ ^1$\\ \grayline
-    - & Differential expression & \no & \no & \no & \yes & \yes $ ^2$ \\ \grayline
-    - & Knocked down regulators & \no & \no & \no & \no & \yes \\ \grayline
-    - & Regulatory network & \no & \no & \no & \yes & \yes \\ \grayline
-    - & Cell-specific regulatory network & \no & \no & \no & \no & \yes \\
-    \blackline
-    \multicolumn{7}{l}{\textbf{Emulate experimental effects}} \\
-    \blackline
-    - & Single-cell RNA sequencing & \yes & \yes & \yes & \yes & \yes \\ \grayline
-    - & Batch effects & \yes & \no & \no & \no & \yes $ ^1$ \\ \grayline
-    - & Knockdown experiment & \no & \no & \no & \no & \yes \\ \grayline
-    - & Time-series & \no & \no & \no & \no & \yes \\
-    - & Snapshot & \no & \no & \no & \no & \yes \\
-    \blackline
-    \multicolumn{7}{l}{\textbf{Evaluation applications}} \\
-    \blackline
-    - & Clustering & \yes & \yes & \no & \yes & \no \\ \grayline
-    - & Trajectory inference & \yes & \no & \yes & \yes & \yes \\ \grayline
-    - & Network inference & \no & \no & \no & \yes & \yes \\ \grayline
-    - & Cell-specific network inference & \no & \no & \no & \no & \yes \\ \grayline
-    - & Differential expression & \yes & \no & \no & \no & \yes $ ^2$ \\ \grayline
-    - & Batch effect correction & \yes & \no & \no & \no & \yes \\ \grayline
-    - & RNA Velocity & \no & \no & \no & \no & \yes \\ \grayline
-    - & Trajectory alignment & \no & \no & \no & \no & \yes \\
-    \blackline
-    %\multicolumn{7}{l}{\textbf{Showcased dataset}} \\
-    %\blackline
-    %- & Number of cells &  &  &  &  & \\ \grayline
-    %- & Number of genes &  &  &  &  & \\ \grayline
-    %- & Execution time & & & & & \\
-    \blackline
-
-\end{tabular}
-\end{table}
-
-\begin{figure}[H]
-    \centering
-    \includegraphics[width=.8\linewidth]{result_files/usecase_trajectory_alignment/usecase.pdf}
-    \caption{
-    \textbf{dyngen allows benchmarking of trajectory alignment methods.} 
- \textbf{A, B}: Two samples containing a linear trajectory, generated by dyngen.
-  \textbf{C, D}: Result of the DTW alignment on the samples in \textbf{A} and \textbf{B}.
-In \textbf{C}, the individual mappings of the alignment between cells are shown.
-In \textbf{D}, the accumulated distance matrix between the two trajectories is shown, including the black warping path, corresponding to these cell to cell alignments in \textbf{C}.
-  \textbf{E, F}: Shows the accumulated distance matrices obtained after using DTW on two trajectories where noise (noise level of 0.4) was added to the count matrix.
-In E the complete count matrices were used to perform the alignment.
-In F, smoothed pseudocells were used.
-  \textbf{G:} Shows the influence of added noise to the different processing methods. We can see that DTW + smoothing performs best in noisy circumstances.
-    }
-    \label{fig:traj_align}
-\end{figure}
-
-\begin{figure}[H]
-    \centering
-    \includegraphics[width=\linewidth]{result_files/usecase_rna_velocity/usecase.pdf}
-    \caption{
-    \textbf{dyngen allows benchmarking of RNA velocity methods.} 
-    \textbf{A:} An example bifurcating cycle dataset, with as illustration the expression and ground truth velocity of a gene D1\_TF1 that goes up and down in one branch of the trajectory. 
-\textbf{B:} The RNA velocity estimates of gene D1\_TF1 by the different methods.
-\textbf{C:} The velocity stream plots produced from the predictions of each method, as generated by scvelo.
-\textbf{D:} The predictions scored by two different metrics, the velocity correlation and the velocity arrow cosine.
-The velocity correlation is the correlation between the ground-truth velocity (A, right) and the predicted velocity (B).
-The velocity arrow cosine is the cosine similarity between the direction of segments of the ground-truth trajectory (A, left) and the RNA velocity values calculated at those points (C).
-    }
-    \label{fig:velocity}
-\end{figure}
-
-\begin{figure}[H]
-    \centering
-    \includegraphics[width=.8\linewidth]{result_files/usecase_network_inference/cni.pdf}
-    \caption{
-    \textbf{dyngen allows benchmarking Cell-specific Network Inference (CSNI) methods.} 
-    \textbf{A:} A cell is simulated using the global gene regulatory network (GRN, top left).
-However, at any particular state in the simulation, only a fraction of the gene regulatory interactions are active.
-    \textbf{B:} CSNI methods were executed to predict the regulatory interactions that are active in each cell specifically.
-Using the ground-truth cell-specific GRN, the performance of each method was quantified on 14 dyngen datasets.
-    }
-    \label{fig:scgrn}
-\end{figure}
+\newpage
 
@@
 The workflow to generate _in silico_ single-cell data consists of six main steps (Figure\ \ref{fig:explain_methods}).
-
-\begin{figure}[htb!]
-    \centering
-    \includegraphics[width=\textwidth]{result_files/explain_methods}
-    \caption{\textbf{The workflow of dyngen consists of six main steps.} 
-    \textbf{A:} The user needs to specify the desired module network or use a predefined module network. The module network is what determines the dynamic behaviour of simulated cells.
-    \textbf{B:} The number of desired transcription factors (which drive the desired dynamic process) are amongst the given modules and adds regulatory interactions according to the module network.
-Additional target genes (which do not influence the dynamic process) are added by sampling interactions from GRN interaction databases.
-    \textbf{C:} Each gene regulatory interaction in the GRN is converted to a set of biochemical reactions. 
-    \textbf{D:} Along with the module network, the user also needs to specify the backbone structure of expected cell states.
-The average expression of each edge in the backbone is simulated by activating a restricted set of genes for each edge.
-    \textbf{E:} Multiple Gillespie SSA simulations are run using the reactions defined in step C.
-The counts of each of the molecules at each time step are extracted.
-Each time step is mapped to a point in the backbone.
-    \textbf{F:} The molecule levels of multiple simulations are shown over time (left).
-From each simulation, multiple cells are sampled (from left to middle).
-Technical noise from profiling is simulated by sampling molecules from the set of molecules inside each cell (from middle to right).
-    }
-    \label{fig:explain_methods}
-\end{figure}
 
@@
 
-\begin{figure}[htb!]
-    \centering
-    \includegraphics[width=0.8\textwidth]{result_files/example_backbones_onlymodules}
-    \caption{
-      \textbf{The module network determines the type of dynamic process which simulated cells will undergo.} A module network describes the regulatory interactions between sets of transcription factors which drive the desired dynamic process.
-    }
-    \label{fig:example_backbones_onlymodules}
-\end{figure}
-
 ## Generating the gene regulatory network {#sec:dyngen-grn}
 The GRN is generated based on the given module network in four main steps (Figure\ \ref{fig:gen_feature_network}).
-
-\begin{figure}[htb!]
-    \centering
-    \includegraphics[width=\linewidth]{result_files/gen_feature_network}
-    \caption{
-        \textbf{Generating the feature network from a backbone consists of four main steps.}
-    }
-    \label{fig:gen_feature_network}
-\end{figure}
 
@@
 The per-gene and per-interaction constants are based on the median reported production-rates and half-lives of molecules measured of 5000 mammalian genes [@schwanhausser_globalquantificationmammalian_2011], except that the transcription rate has been amplified by a factor of 10.
-
-\begin{table}[ht]
-    \caption{
-      \textbf{Reactions affecting the abundance levels of pre-mRNA $\x G$, mature mRNA $\y G$ and proteins $\z G$ of gene $G$.} Define the set of regulators of $G$ as $\rs{G}$, the set of upregulating regulators of $G$ as $\rp G$, and the set of down-regulating regulators of $G$ as $\rn G$.
-Parameters used in the propensity formulae are defined in Table \ref{tab:reaction_params}.
-    } \label{tab:reaction_def}
-    \centering
-    \begin{tabular}{|lcc|}
-        \hline
-        Reaction & Effect & Propensity \\ \hline \hline
-        Transcription & $\emptyset \rightarrow \x G$ & $\xpr G \times \frac{\ba G - \ind{G}^{|\rp{G}|} + \prod\limits_{H \in \rp{G}}(\ind G + \buf{G,H})}{\prod\limits_{H \in \rs{G}}(1 + \buf{G,H})}$ \\
-        Splicing & $\x G \rightarrow \y G$ & $\ysr G \times \x G$ \\
-        Translation & $\y G \rightarrow \y G + \z G$ & $\zpr G \times \y G$ \\ \hline\hline
-        Pre-mRNA degradation & $\x G \rightarrow \emptyset$ & $\ydr G \times \x G$ \\
-        Mature mRNA degradation & $\y G \rightarrow \emptyset$ & $\ydr G \times \y G$ \\
-        Protein degradation & $\z G \rightarrow \emptyset$ & $\zdr G \times \z G$ \\ \hline
-    \end{tabular}
-\end{table}
-
-\begin{table}[ht]
-    \caption{
-      \textbf{Default parameters defined for the calculation of reaction propensity functions.}
-    } \label{tab:reaction_params}
-    \centering
-    \begin{tabular}{|lrl|}
-        \hline
-        Parameter & Symbol & Definition \\ \hline \hline
-        Transcription rate & $\xpr{G}$ & $\in U(10, 20)$ \\
-        Splicing rate & $\ysr G$ & $= \ln(2)\ /\ 2$ \\
-        Translation rate & $\zpr{G}$ & $\in U(100, 150)$ \\
-        (Pre-)mRNA half-life & $\yhl{G}$ & $\in U(2.5, 5)$ \\
-        Protein half-life & $\zhl G$ & $\in U(5, 10)$ \\
-        Interaction strength & $\str{G,H}$ & $\in 10^{U(0, 2)}$ * \\
-        Hill coefficient & $\hill{G,H}$ & $\in U(0.5, 2)$ * \\
-        Independence factor & $\ind G$ & $\in U(0, 1)$ * \\ \hline\hline
-        (Pre-)mRNA degradation rate & $\ydr G$ & $= \ln(2)\ /\ \yhl G$ \\
-        Protein degradation rate & $\zdr G$ & $= \ln(2)\ /\ \zhl G$ \\
-        Dissociation constant & $\dis H$ & $= 0.5 \times \frac{\xpr H \times \ysr H \times \zpr H}{(\ydr H + \ysr H) \times \ydr H \times \zdr H}$ \\
-        Binding strength & $\buf{G,H}$ & $= \str{G,H} \times \left(\z H\ /\ \dis H\right) ^ {\hill{G,H}}$ \\
-        Basal expression & $\ba G$ & $= \begin{cases} 1 & \mbox{if } \rp{G} = \emptyset \\ 0.0001 & \mbox{if } \rn{G} = \emptyset \mbox{ and } \rp{G} \neq \emptyset \\ 0.5 & \mbox{otherwise} \end{cases}$ * \\ \hline
-        \multicolumn{3}{l}{*: unless $G$ is a TF, then the value is determined by the backbone.}
-    \end{tabular}
-\end{table}
 
@@
 
-\begin{figure}[htb!]
-    \centering
-    \includegraphics[width=.6\linewidth]{result_files/sample_cells.pdf}
-    \caption{
-        \textbf{Two approaches can be used to sample cells from simulations: snapshot and time-series.}
-    }
-    \label{fig:sample_cells}
-\end{figure}
-
 ### Sample molecules
@@
 For each edge, a simulation is run using the end state of an upstream branch as the initial expression vector, and only allowing the modules as predefined by the attribute to change.
-
-\begin{figure}[htb!]
-    \centering
-    \includegraphics[width=0.8\textwidth]{result_files/example_backbones}
-    \caption{\textbf{Examples of the ground-truth state networks which need to be provided alongside the module network.}}
-    \label{fig:example_backbones}
-\end{figure}
 
@@
 
-Determining the regulatory effect for all interactions and cells in the dataset yields the complete cell-specific ground-truth GRN <!--(Figure\ \ref{fig:cell-specific_grn})-->.
+Determining the regulatory effect for all interactions and cells in the dataset yields the complete cell-specific ground-truth GRN.
 The regulatory effect lies between $[-1, 1]$, where -1 represents complete inhibition of $T$ by $R$, 1 represents maximal activation of $T$ by $R$, and 0 represents inactivity of the regulatory interaction between $R$ and $T$.
-<!--
-\begin{figure}[htb!]
-    \centering
-    \includegraphics[width=.8\linewidth]{result_files/fig1_showcase_functionality/scgrn.pdf}
-    \caption{
-        \textbf{The cell-specific regulatory effects of all interactions, computed on cells part of a bifurcation trajectory.} Negative values correspond to inhibitory interactions, positive values to activating interactions, and zero values correspond to inactive interactions.
-    }
-    \label{fig:cell-specific_grn}
-\end{figure}
--->
 
@@
 We after aligning the cells, we calculate the mean absolute difference between the geodesic distances of the pairs of cells, which should ideally be 0.
+
+\printbibliography
+\newpage
+
+# Supplementary Figures
+\setcounter{page}{1}
+\newrefsection
+\setcounter{table}{0}
+\renewcommand{\thetable}{S\arabic{table}}
+\setcounter{figure}{0}
+\renewcommand{\thefigure}{S\arabic{figure}}
+
+\newcommand{\yes}{\ding{51}}
+\newcommand{\no}{}
+\definecolor{light-gray}{rgb}{.85,.85,.85}
+\newcommand{\grayline}{\arrayrulecolor{light-gray}\cline{3-7}\arrayrulecolor{black}}
+\newcommand{\blackline}{\arrayrulecolor{black}\cline{3-7}}
+\newcommand{\method}[1]{#1}
+
+\begin{table}[H]
+    \caption{
+      \textbf{Feature comparison of single-cell synthetic data generators and simulators.} $ ^1$ As showcased by vignette.
+$^2$ As showcased by Van den Berge et al.
+\cite{vandenberge_trajectorybaseddifferentialexpression_2019}.
+    } \label{tab:comparison}
+    \small
+    \begin{tabular}{p{.25cm}l|*{5}{>{\centering\arraybackslash}p{1.25cm}|}}
+    \blackline
+    & \method{} & \method{splatter} & \method{powsimR} & \method{PROSSTT} & \method{SymSim} & \method{dyngen} \\
+    \blackline
+    \multicolumn{7}{l}{\textbf{Available modality outputs}} \\
+    \blackline
+    - & mRNA expression & \yes & \yes & \yes & \yes & \yes \\ \grayline
+    - & Pre-mRNA expression & \no & \no & \no & \no & \yes \\ \grayline
+    - & Protein expression & \no & \no & \no & \no & \yes \\ \grayline
+    - & Promotor activity & \no & \no & \no & \yes & \yes \\ \grayline
+    - & Reaction activity & \no & \no & \no & \no & \yes \\
+    \blackline
+    \multicolumn{7}{l}{\textbf{Available ground-truth outputs}} \\
+    \blackline
+    - & True counts & \yes & \no & \no & \yes & \yes \\ \grayline
+    - & Cluster labels & \yes & \yes & \no & \yes & \yes \\ \grayline
+    - & Trajectory & \yes & \no & \yes $ ^1$ & \yes & \yes \\ \grayline
+    - & Batch labels & \yes & \no & \no & \no & \yes $ ^1$\\ \grayline
+    - & Differential expression & \no & \no & \no & \yes & \yes $ ^2$ \\ \grayline
+    - & Knocked down regulators & \no & \no & \no & \no & \yes \\ \grayline
+    - & Regulatory network & \no & \no & \no & \yes & \yes \\ \grayline
+    - & Cell-specific regulatory network & \no & \no & \no & \no & \yes \\
+    \blackline
+    \multicolumn{7}{l}{\textbf{Emulate experimental effects}} \\
+    \blackline
+    - & Single-cell RNA sequencing & \yes & \yes & \yes & \yes & \yes \\ \grayline
+    - & Batch effects & \yes & \no & \no & \no & \yes $ ^1$ \\ \grayline
+    - & Knockdown experiment & \no & \no & \no & \no & \yes \\ \grayline
+    - & Time-series & \no & \no & \no & \no & \yes \\
+    - & Snapshot & \no & \no & \no & \no & \yes \\
+    \blackline
+    \multicolumn{7}{l}{\textbf{Evaluation applications}} \\
+    \blackline
+    - & Clustering & \yes & \yes & \no & \yes & \no \\ \grayline
+    - & Trajectory inference & \yes & \no & \yes & \yes & \yes \\ \grayline
+    - & Network inference & \no & \no & \no & \yes & \yes \\ \grayline
+    - & Cell-specific network inference & \no & \no & \no & \no & \yes \\ \grayline
+    - & Differential expression & \yes & \no & \no & \no & \yes $ ^2$ \\ \grayline
+    - & Batch effect correction & \yes & \no & \no & \no & \yes \\ \grayline
+    - & RNA Velocity & \no & \no & \no & \no & \yes \\ \grayline
+    - & Trajectory alignment & \no & \no & \no & \no & \yes \\
+    \blackline
+    %\multicolumn{7}{l}{\textbf{Showcased dataset}} \\
+    %\blackline
+    %- & Number of cells &  &  &  &  & \\ \grayline
+    %- & Number of genes &  &  &  &  & \\ \grayline
+    %- & Execution time & & & & & \\
+    \blackline
+
+\end{tabular}
+\end{table}
+
+\begin{figure}[H]
+    \centering
+    \includegraphics[width=\linewidth]{result_files/usecase_trajectory_alignment/usecase.pdf}
+    \caption{
+    \textbf{dyngen allows benchmarking of trajectory alignment methods.} 
+ \textbf{A, B}: Two samples containing a linear trajectory, generated by dyngen.
+  \textbf{C, D}: Result of the DTW alignment on the samples in \textbf{A} and \textbf{B}.
+In \textbf{C}, the individual mappings of the alignment between cells are shown.
+In \textbf{D}, the accumulated distance matrix between the two trajectories is shown, including the black warping path, corresponding to these cell to cell alignments in \textbf{C}.
+  \textbf{E, F}: Shows the accumulated distance matrices obtained after using DTW on two trajectories where noise (noise level of 0.4) was added to the count matrix.
+In E the complete count matrices were used to perform the alignment.
+In F, smoothed pseudocells were used.
+  \textbf{G:} Shows the influence of added noise to the different processing methods. We can see that DTW + smoothing performs best in noisy circumstances.
+    }
+    \label{fig:traj_align}
+\end{figure}
+
+\begin{figure}[H]
+    \centering
+    \includegraphics[width=\linewidth]{result_files/usecase_rna_velocity/usecase.pdf}
+    \caption{
+    \textbf{dyngen allows benchmarking of RNA velocity methods.} 
+    \textbf{A:} An example bifurcating cycle dataset, with as illustration the expression and ground truth velocity of a gene D1\_TF1 that goes up and down in one branch of the trajectory. 
+\textbf{B:} The RNA velocity estimates of gene D1\_TF1 by the different methods.
+\textbf{C:} The velocity stream plots produced from the predictions of each method, as generated by scvelo.
+\textbf{D:} The predictions scored by two different metrics, the velocity correlation and the velocity arrow cosine.
+The velocity correlation is the correlation between the ground-truth velocity (A, right) and the predicted velocity (B).
+The velocity arrow cosine is the cosine similarity between the direction of segments of the ground-truth trajectory (A, left) and the RNA velocity values calculated at those points (C).
+    }
+    \label{fig:velocity}
+\end{figure}
+
+\begin{figure}[H]
+    \centering
+    \includegraphics[width=.8\linewidth]{result_files/usecase_network_inference/cni.pdf}
+    \caption{
+    \textbf{dyngen allows benchmarking Cell-specific Network Inference (CSNI) methods.} 
+    \textbf{A:} A cell is simulated using the global gene regulatory network (GRN, top left).
+However, at any particular state in the simulation, only a fraction of the gene regulatory interactions are active.
+    \textbf{B:} CSNI methods were executed to predict the regulatory interactions that are active in each cell specifically.
+Using the ground-truth cell-specific GRN, the performance of each method was quantified on 14 dyngen datasets.
+    }
+    \label{fig:scgrn}
+\end{figure}
+
+\begin{figure}[H]
+    \centering
+    \includegraphics[width=\textwidth]{result_files/explain_methods}
+    \caption{\textbf{The workflow of dyngen consists of six main steps.} 
+    \textbf{A:} The user needs to specify the desired module network or use a predefined module network. The module network is what determines the dynamic behaviour of simulated cells.
+    \textbf{B:} The number of desired transcription factors (which drive the desired dynamic process) are amongst the given modules and adds regulatory interactions according to the module network.
+Additional target genes (which do not influence the dynamic process) are added by sampling interactions from GRN interaction databases.
+    \textbf{C:} Each gene regulatory interaction in the GRN is converted to a set of biochemical reactions. 
+    \textbf{D:} Along with the module network, the user also needs to specify the backbone structure of expected cell states.
+The average expression of each edge in the backbone is simulated by activating a restricted set of genes for each edge.
+    \textbf{E:} Multiple Gillespie SSA simulations are run using the reactions defined in step C.
+The counts of each of the molecules at each time step are extracted.
+Each time step is mapped to a point in the backbone.
+    \textbf{F:} The molecule levels of multiple simulations are shown over time (left).
+From each simulation, multiple cells are sampled (from left to middle).
+Technical noise from profiling is simulated by sampling molecules from the set of molecules inside each cell (from middle to right).
+    }
+    \label{fig:explain_methods}
+\end{figure}
+
+\begin{figure}[H]
+    \centering
+    \includegraphics[width=0.8\textwidth]{result_files/example_backbones_onlymodules}
+    \caption{
+      \textbf{The module network determines the type of dynamic process which simulated cells will undergo.} A module network describes the regulatory interactions between sets of transcription factors which drive the desired dynamic process.
+    }
+    \label{fig:example_backbones_onlymodules}
+\end{figure}
+
+\begin{figure}[H]
+    \centering
+    \includegraphics[width=\linewidth]{result_files/gen_feature_network}
+    \caption{
+        \textbf{Generating the feature network from a backbone consists of four main steps.}
+    }
+    \label{fig:gen_feature_network}
+\end{figure}
+
+\begin{table}[H]
+    \caption{
+      \textbf{Reactions affecting the abundance levels of pre-mRNA $\x G$, mature mRNA $\y G$ and proteins $\z G$ of gene $G$.} Define the set of regulators of $G$ as $\rs{G}$, the set of upregulating regulators of $G$ as $\rp G$, and the set of down-regulating regulators of $G$ as $\rn G$.
+Parameters used in the propensity formulae are defined in Table \ref{tab:reaction_params}.
+    } \label{tab:reaction_def}
+    \centering
+    \begin{tabular}{|lcc|}
+        \hline
+        Reaction & Effect & Propensity \\ \hline \hline
+        Transcription & $\emptyset \rightarrow \x G$ & $\xpr G \times \frac{\ba G - \ind{G}^{|\rp{G}|} + \prod\limits_{H \in \rp{G}}(\ind G + \buf{G,H})}{\prod\limits_{H \in \rs{G}}(1 + \buf{G,H})}$ \\
+        Splicing & $\x G \rightarrow \y G$ & $\ysr G \times \x G$ \\
+        Translation & $\y G \rightarrow \y G + \z G$ & $\zpr G \times \y G$ \\ \hline\hline
+        Pre-mRNA degradation & $\x G \rightarrow \emptyset$ & $\ydr G \times \x G$ \\
+        Mature mRNA degradation & $\y G \rightarrow \emptyset$ & $\ydr G \times \y G$ \\
+        Protein degradation & $\z G \rightarrow \emptyset$ & $\zdr G \times \z G$ \\ \hline
+    \end{tabular}
+\end{table}
+
+\begin{table}[H]
+    \caption{
+      \textbf{Default parameters defined for the calculation of reaction propensity functions.}
+    } \label{tab:reaction_params}
+    \centering
+    \begin{tabular}{|lrl|}
+        \hline
+        Parameter & Symbol & Definition \\ \hline \hline
+        Transcription rate & $\xpr{G}$ & $\in U(10, 20)$ \\
+        Splicing rate & $\ysr G$ & $= \ln(2)\ /\ 2$ \\
+        Translation rate & $\zpr{G}$ & $\in U(100, 150)$ \\
+        (Pre-)mRNA half-life & $\yhl{G}$ & $\in U(2.5, 5)$ \\
+        Protein half-life & $\zhl G$ & $\in U(5, 10)$ \\
+        Interaction strength & $\str{G,H}$ & $\in 10^{U(0, 2)}$ * \\
+        Hill coefficient & $\hill{G,H}$ & $\in U(0.5, 2)$ * \\
+        Independence factor & $\ind G$ & $\in U(0, 1)$ * \\ \hline\hline
+        (Pre-)mRNA degradation rate & $\ydr G$ & $= \ln(2)\ /\ \yhl G$ \\
+        Protein degradation rate & $\zdr G$ & $= \ln(2)\ /\ \zhl G$ \\
+        Dissociation constant & $\dis H$ & $= 0.5 \times \frac{\xpr H \times \ysr H \times \zpr H}{(\ydr H + \ysr H) \times \ydr H \times \zdr H}$ \\
+        Binding strength & $\buf{G,H}$ & $= \str{G,H} \times \left(\z H\ /\ \dis H\right) ^ {\hill{G,H}}$ \\
+        Basal expression & $\ba G$ & $= \begin{cases} 1 & \mbox{if } \rp{G} = \emptyset \\ 0.0001 & \mbox{if } \rn{G} = \emptyset \mbox{ and } \rp{G} \neq \emptyset \\ 0.5 & \mbox{otherwise} \end{cases}$ * \\ \hline
+        \multicolumn{3}{l}{*: unless $G$ is a TF, then the value is determined by the backbone.}
+    \end{tabular}
+\end{table}
+
+\begin{figure}[H]
+    \centering
+    \includegraphics[width=.6\linewidth]{result_files/sample_cells.pdf}
+    \caption{
+        \textbf{Two approaches can be used to sample cells from simulations: snapshot and time-series.}
+    }
+    \label{fig:sample_cells}
+\end{figure}
+
+\begin{figure}[H]
+    \centering
+    \includegraphics[width=0.8\textwidth]{result_files/example_backbones}
+    \caption{\textbf{Examples of the ground-truth state networks which need to be provided alongside the module network.}}
+    \label{fig:example_backbones}
+\end{figure}

### 2020-06-13 render_manuscript.R

@@ -16,3 +16,6 @@ textreadr::read_docx(drive$local_path, remove.empty = FALSE, trim = FALSE) %>%
 # render the manuscript
 render("manuscript.Rmd")
 
+system("pdftk manuscript.pdf cat 1-17 output manu.pdf")
+system("pdftk manuscript.pdf cat 18-25 output supp.pdf")
+

### 2020-06-13 render.Rmd
update

--- 2020-06-13 manuscript.Rmd
@@
 Other generators of scRNA-seq data (e.g.
-splatter [@zappia_splattersimulationsinglecell_2017], powsimR [@vieth_powsimrpoweranalysis_2017], PROSSTT [@papadopoulos_prossttprobabilisticsimulation_2018] and SymSim [@zhang_simulatingmultiplefaceted_2019]) have already been used extensively to explore the strengths and weaknesses of computational tools, both by method developers [@street_slingshotcelllineage_2018; @parra_reconstructingcomplexlineage_2018; @lummertzdarocha_reconstructioncomplexsinglecell_2018; @lin_scclassifyhierarchicalclassification_2019] and independent benchmarkers [@duo_systematicperformanceevaluation_2018; @saelens_comparisonsinglecelltrajectory_2019; @soneson_biasrobustnessscalability_2018].
+splatter [@zappia_splattersimulationsinglecell_2017], powsimR [@vieth_powsimrpoweranalysis_2017], PROSSTT [@papadopoulos_prossttprobabilisticsimulation_2019] and SymSim [@zhang_simulatingmultiplefaceted_2019]) have already been used extensively to explore the strengths and weaknesses of computational tools, both by method developers [@street_slingshotcelllineage_2018; @parra_reconstructingcomplexlineage_2019; @lummertzdarocha_reconstructioncomplexsinglecell_2018; @lin_scclassifyhierarchicalclassification_2019] and independent benchmarkers [@duo_systematicperformanceevaluation_2018; @saelens_comparisonsinglecelltrajectory_2019; @soneson_biasrobustnessscalability_2018].
 However, a limitation of these existing simulators is that they would require significant methodological alterations to add additional modalities or experimental conditions (Table\ \ref{tab:comparison}).
@@
 The fine-grained controls over simulation parameters allow dyngen to be applicable to a broad range of use-cases to simulate dynamic biological processes.
-The original design of dyngen was motivated by the plethora of methods available for trajectory inference, where dyngen allowed the first large-scale benchmarking of such methods [@saelens_comparisonsinglecelltrajectory_2019; @vandenberge_trajectorybaseddifferentialexpression_2019].
+The original design of dyngen was motivated by the plethora of methods available for trajectory inference, where dyngen allowed the first large-scale benchmarking of such methods [@saelens_comparisonsinglecelltrajectory_2019; @vandenberge_trajectorybaseddifferentialexpression_2020].
 
@@
 $^2$ As showcased by Van den Berge et al.
-\cite{vandenberge_trajectorybaseddifferentialexpression_2019}.
+\cite{vandenberge_trajectorybaseddifferentialexpression_2020}.
     } \label{tab:comparison}

### 2020-06-13 render_manuscript.R

@@ -11,11 +11,12 @@ drive <- drive_download(as_id("15ZkzzB-XSYFfZutO4YPkUIw6iUvxJS92fSPfzqR1cLA"), t
 textreadr::read_docx(drive$local_path, remove.empty = FALSE, trim = FALSE) %>%
   str_replace_all("([^ ])(\\[@[^\\]]*\\])", "\\1 \\2") %>% # add spaces before citations
   str_replace_all("^#", "\n#") %>%
-  write_lines("manuscript.Rmd")
+  write_lines("render.Rmd")
 
 # render the manuscript
-render("manuscript.Rmd")
+render("render.Rmd")
 
-system("pdftk manuscript.pdf cat 1-17 output manu.pdf")
-system("pdftk manuscript.pdf cat 18-25 output supp.pdf")
+system("pdftk render.pdf cat 1-17 output manuscript.pdf")
+system("pdftk render.pdf cat 18-25 output supplementary_files.pdf")
 
+file.remove("render.pdf")

### 2020-07-14 analysis/example_methods_overview/script.R
refactor script

@@ -1,9 +1,10 @@
 library(tidyverse)
 library(dyngen)
+library(dyngen.manuscript)
+
+exp <- start_analysis("example_methods")
 
 set.seed(4)
-
-dir.create("fig/example_methods", showWarnings = FALSE, recursive = TRUE)
 
 backbone <- backbone(
   module_info = tribble(
@@ -49,16 +50,16 @@ model <-
   generate_kinetics()
 
 ## grn
-plot_backbone_statenet(model) %>% ggsave(filename = "fig/example_methods/statenet.pdf", width = 6, height = 6)
-plot_backbone_modulenet(model) %>% ggsave(filename = "fig/example_methods/modulenet.pdf", width = 8, height = 6)
+plot_backbone_statenet(model) %>% ggsave(filename = exp$result("statenet.pdf"), width = 6, height = 6)
+plot_backbone_modulenet(model) %>% ggsave(filename = exp$result("modulenet.pdf"), width = 8, height = 6)
 
 plot_feature_network(model, show_targets = FALSE)
 
 plot_feature_network(model)
 plot_feature_network(model, show_hks = TRUE)
 plot_feature_network(model)
-plot_feature_network(model, show_hks = TRUE) %>% ggsave(filename = "fig/example_methods/featnet.pdf", width = 8, height = 6)
-plot_feature_network(model, show_targets = FALSE) %>% ggsave(filename = "fig/example_methods/featnet_onlytfs.pdf", width = 8, height = 6)
+plot_feature_network(model, show_hks = TRUE) %>% ggsave(filename = exp$result("featnet.pdf"), width = 8, height = 6)
+plot_feature_network(model, show_targets = FALSE) %>% ggsave(filename = exp$result("featnet_onlytfs.pdf"), width = 8, height = 6)
 
 
 ## simulate
@@ -68,13 +69,13 @@ model2 <- model %>%
   generate_experiment()
 
 model2 <- dyngen:::calculate_dimred(model2, dimred_premrna = FALSE)
-write_rds(model2, "fig/example_methods/model.rds", compress = "gz")
+write_rds(model2, exp$result("model.rds"), compress = "gz")
 
[…]

### 2020-07-14 manuscript/render_manuscript.R
move files around

@@ -1,0 +1,23 @@
+library(tidyverse)
+library(googledrive)
+library(rmarkdown)
+library(textreadr)
+
+# download as docx instead of txt, because otherwise comments get pushed into the document
+httr::set_config(httr::config(http_version = 0)) # avoid http2 framing layer bug
+drive <- drive_download(as_id("15ZkzzB-XSYFfZutO4YPkUIw6iUvxJS92fSPfzqR1cLA"), type = "text", overwrite = TRUE, path = tempfile())
+
+# read docx
+
+readr::read_lines(drive$local_path) %>%
+  str_replace_all("([^ ])(\\[@[^\\]]*\\])", "\\1 \\2") %>% # add spaces before citations
+  str_replace_all("^#", "\n#") %>%
+  write_lines("manuscript/render.Rmd")
+
+# render the manuscript
+render("manuscript/render.Rmd", output_dir = "manuscript/")
+
+system("pdftk manuscript/render.pdf cat 1-17 output manuscript/manuscript.pdf")
+system("pdftk manuscript/render.pdf cat 18-25 output manuscript/supplementary_files.pdf")
+
+file.remove("manuscript/render.pdf")

### 2020-07-14 render_manuscript.R

@@ -1,22 +1,0 @@
-library(tidyverse)
-library(googledrive)
-library(rmarkdown)
-library(textreadr)
-
-# download as docx instead of txt, because otherwise comments get pushed into the document
-httr::set_config(httr::config(http_version = 0)) # avoid http2 framing layer bug
-drive <- drive_download(as_id("15ZkzzB-XSYFfZutO4YPkUIw6iUvxJS92fSPfzqR1cLA"), type = "docx", overwrite = TRUE, path = tempfile())
-
-# read docx
-textreadr::read_docx(drive$local_path, remove.empty = FALSE, trim = FALSE) %>%
-  str_replace_all("([^ ])(\\[@[^\\]]*\\])", "\\1 \\2") %>% # add spaces before citations
-  str_replace_all("^#", "\n#") %>%
-  write_lines("render.Rmd")
-
-# render the manuscript
-render("render.Rmd")
-
-system("pdftk render.pdf cat 1-17 output manuscript.pdf")
-system("pdftk render.pdf cat 18-25 output supplementary_files.pdf")
-
-file.remove("render.pdf")

### 2020-07-14 manuscript/render_manuscript.R
add check

@@ -3,16 +3,16 @@ library(googledrive)
 library(rmarkdown)
 library(textreadr)
 
-# download as docx instead of txt, because otherwise comments get pushed into the document
-httr::set_config(httr::config(http_version = 0)) # avoid http2 framing layer bug
-drive <- drive_download(as_id("15ZkzzB-XSYFfZutO4YPkUIw6iUvxJS92fSPfzqR1cLA"), type = "text", overwrite = TRUE, path = tempfile())
+if (Sys.info()[["user"]] == "rcannood") {
+  httr::set_config(httr::config(http_version = 0)) # avoid http2 framing layer bug
+  drive <- drive_download(as_id("15ZkzzB-XSYFfZutO4YPkUIw6iUvxJS92fSPfzqR1cLA"), type = "text", overwrite = TRUE, path = tempfile())
 
-# read docx
-
-readr::read_lines(drive$local_path) %>%
-  str_replace_all("([^ ])(\\[@[^\\]]*\\])", "\\1 \\2") %>% # add spaces before citations
-  str_replace_all("^#", "\n#") %>%
-  write_lines("manuscript/render.Rmd")
+  # read docx
+  readr::read_lines(drive$local_path) %>%
+    str_replace_all("([^ ])(\\[@[^\\]]*\\])", "\\1 \\2") %>% # add spaces before citations
+    str_replace_all("^#", "\n#") %>%
+    write_lines("manuscript/render.Rmd")
+}
 
 # render the manuscript
 render("manuscript/render.Rmd", output_dir = "manuscript/")

### 2020-09-15 analysis/usecase_trajectory_alignment/3_explanation_plot.R
make combine_models more generic

@@ -48,7 +48,7 @@ model2_diseased <-
   ) %>% generate_cells()
 
 combined_model <-
-  combine_models(model1_control, model2_diseased) %>%
+  combine_models(list("control" = model1_control, "diseased" = model2_diseased)) %>%
   generate_experiment()
 
 datasetD <- wrap_dataset(combined_model)

### 2020-09-15 package/R/combine_models.R

@@ -1,0 +1,125 @@
+
+#' Combine multiple dyngen models
+#'
+#' Assume the given models ran up to `generate_cells()`, and
+#' will run `generate_experiment()` on the combined models.
+#'
+#' @param models A list of models.
+#'
+#' @export
+#'
+#' @examples
+#' \donttest{
+#' # create the config object
+#' init_config <- initialise_model(
+#'   backbone = backbone_linear(),
+#'   num_cells = 500,
+#'   num_targets = 250,
+#'   num_hks = 50,
+#'   gold_standard_params = gold_standard_default(census_interval = 1, tau = 0.05),
+#'   simulation_params = simulation_default(
+#'     # burn_time = 10,
+#'     # total_time = 10,
+#'     census_interval = 1,
+#'     ssa_algorithm = ssa_etl(tau = 0.05),
+#'     experiment_params = simulation_type_wild_type(num_simulations = 40)
+#'   ),
+#'   # set this to the number of cores in your system
+#'   num_cores = 8,
+#'   # set this to a directory where dyngen can cache some files
+#'   download_cache_dir = "~/.cache/dyngen"
+#' )
+#'
+#' # generate the genes and their kinetics
+#' model_common <-
+#'   init_config %>%
+#'   generate_tf_network() %>%
+#'   generate_feature_network()
+#' plot_feature_network(model_common)
+#'
+#' # run the simulation once
+#' model_a <- model_common %>%
+#'   generate_kinetics() %>%
+#'   generate_gold_standard() %>%
+#'   generate_cells()
+#'
+#' # run the simulation once more
+#' model_b <- model_common %>%
+#'   generate_kinetics() %>%
+#'   generate_gold_standard() %>%
+#'   generate_cells()
+#'
+#' # combine models, do experiment afterwards
+#' model_ab <- combine_models(list("left" = model_a, "right" = model_b) %>%
+#'   generate_experiment()
+#'
+#' # show a dimensionality reduction
+#' plot_simulations(model_ab)
+#' plot_gold_mappings(model_ab, do_facet = FALSE)
+#'
+#' # create a dynwrap dataset
+#' dataset <- wrap_dataset(model_ab)
+#' }
+combine_models <- function(models) {
+  assert_that(
+    is.list(models),
+    length(models) >= 1,
+    !is.null(names(models))
+  )
+
+  model_combined <- models[[1]]
+  model_combined$gold_standard <- list()
+  model_combined$simulations <- list()
+
+  for (i in seq_along(models)) {
[…]

### 2020-09-15 package/R/perturb_dataset.R

@@ -28,57 +28,3 @@ generate_diff_abundance <- function(model, froms, tos, ratios) {
 
   model
 }
-
-#' Combine two dyngen models
-#'
-#' @param model1 the first model
-#' @param model2 the second model
-#'
-#' @export
-combine_models <- function(model1, model2){
-  model12 <- model1
-
-  # combine gold standard
-  m1gs <- model1$gold_standard
-  m2gs <- model2$gold_standard
-  model12$gold_standard <- list(
-    mod_changes = bind_rows(
-      m1gs$mod_changes %>% mutate_at(c("from", "to", "from_", "to_"), ~paste0("left_", .)),
-      m2gs$mod_changes %>% mutate_at(c("from", "to", "from_", "to_"), ~paste0("right_", .))
-    ),
-    meta = bind_rows(
-      m1gs$meta %>% mutate_at(c("from", "to", "from_", "to_"), ~paste0("left_", .)),
-      m2gs$meta %>% mutate_at(c("from", "to", "from_", "to_"), ~paste0("right_", .))
-    ),
-    counts = rbind(
-      m1gs$counts,
-      m2gs$counts
-    ),
-    network = bind_rows(
-      m1gs$network %>% mutate_at(c("from", "to"), ~paste0("left_", .)),
-      m2gs$network %>% mutate_at(c("from", "to"), ~paste0("right_", .))
-    )
-  )
-
-  # combine simulations
-  m1sim <- model1$simulations
-  m2sim <- model2$simulations
-  num_m1sim <- max(m1sim$meta$simulation_i)
-  model12$simulations <- list(
-    meta = bind_rows(
-      m1sim$meta %>% mutate_at(c("from", "to"), ~paste0("left_", .)),
-      m2sim$meta %>% mutate_at(c("from", "to"), ~paste0("right_", .)) %>% mutate(simulation_i = simulation_i + num_m1sim)
-    ),
-    counts = rbind(
-      m1sim$counts,
-      m2sim$counts
-    )
-    # could also propensities, rna velocity, etc
-  )
-
-  # recalculate the dimred
-  model12 <- model12 %>%
-    dyngen:::calculate_dimred()
-
-  model12
-}

### 2020-09-15 manuscript/render.Rmd
move files around

--- 2020-06-13 render.Rmd
@@
 \end{figure}
+
 dyngen was designed to include all of these functionalities and more by design.
@@
 # Availability
-dyngen is available as an R package on GitHub at [github.com/dynverse/dyngen](https://github.com/dynverse/dyngen).
+dyngen is available as an open-source R package on CRAN at [cran.r-project.org/package=dyngen](https://cran.r-project.org/package=dyngen).
 The analyses performed in this manuscript are available on GitHub at [github.com/dynverse/dyngen_manuscript](https://github.com/dynverse/dyngen_manuscript).
@@
 The geodesic distances between the pair of datasets are directly comparable to each other, and thus cells from the first trajectory should align relatively closely to cells from the second trajectory.
-We after aligning the cells, we calculate the mean absolute difference between the geodesic distances of the pairs of cells, which should ideally be 0.
+After aligning the cells, we calculate the mean absolute difference between the geodesic distances of the pairs of cells, which should ideally be 0.
 
@@
             \blackline
-
 \end{tabular}

### 2021-03-25 analysis/usecase_trajectory_alignment/4_plots.R
preliminary figures

@@ -39,7 +39,7 @@ g <-
 g
 
 # normalized minimum global distance computed
-results <- ggstatsplot::ggwithinstats(
+res_plot <- ggstatsplot::ggwithinstats(
   results,
   x = method,
   y = distance,
@@ -47,7 +47,7 @@ results <- ggstatsplot::ggwithinstats(
   pairwise.comparisons = TRUE,
   pairwise.display = TRUE
 )
-results
+res_plot
 
 # results %>%
 #   ggplot() +
@@ -108,10 +108,11 @@ dtw_dens
 #
 # alignment_smooth <- dtw::dtw(res2_sm$expression, res1_sm$expression, step.pattern = dtw::symmetric2, keep.internals = TRUE)
 # a_sm <- plot_density(alignment_smooth, title = "DTW+smoothing")
-t1 <- (ground_truth + prediction) / (dtw_dens + cal_dens + results)
+t1 <- (plot_spacer() + ground_truth + plot_spacer() + prediction + plot_spacer() + plot_layout(nrow = 1, widths = c(1, 2, 1, 2, 1))) / (dtw_dens + cal_dens + results)
+t1
 t2 <- (ground_truth + prediction) / (dtw_dens + results)
 t2
-ggsave(exp$result("usecase1.pdf"), t1, height = 9, width = 9)
+ggsave(exp$result("usecase1.pdf"), t1, height = 7, width = 9)
 
 # PART 1: Explanation plot ------------------------------------------------
 part1plots <- read_rds(exp$result("usecase_separateplots.rds"))

### 2021-03-25 manuscript/render.Rmd

--- 2020-06-13 render.Rmd
@@
 dyngen is more flexible than current single-cell simulation engines, and allows better method development and benchmarking, thereby stimulating development and testing of novel computational methods.
-We demonstrate its potential for spearheading novel computational methods on three novel applications: aligning cell developmental trajectories, single-cell regulatory network inference and estimation of RNA velocity.
+We demonstrate its potential for spearheading novel computational methods on three novel applications: aligning cell developmental trajectories, cell-specific regulatory network inference and estimation of RNA velocity.
 
@@
 In this way, they can even guide the design of real biological experiments, finding out the best conditions to be used as input for subsequent computational pipelines.
+
 Another, more experimental use of simulation engines is their important role in spearheading the development of novel computational methods, possibly even before real data is available.
@@
 
-Here, we introduce dyngen, a novel multi-modal simulator of dynamic biological processes at single-cell resolution (Figure\ \ref{fig:overview}).
-dyngen uses Gillespie's stochastic simulation algorithm [@gillespie_exactstochasticsimulation_1977] to simulate gene regulation, splicing, and translation at a single-molecule level.
-Other generators of scRNA-seq data (e.g.
-splatter [@zappia_splattersimulationsinglecell_2017], powsimR [@vieth_powsimrpoweranalysis_2017], PROSSTT [@papadopoulos_prossttprobabilisticsimulation_2019] and SymSim [@zhang_simulatingmultiplefaceted_2019]) have already been used extensively to explore the strengths and weaknesses of computational tools, both by method developers [@street_slingshotcelllineage_2018; @parra_reconstructingcomplexlineage_2019; @lummertzdarocha_reconstructioncomplexsinglecell_2018; @lin_scclassifyhierarchicalclassification_2019] and independent benchmarkers [@duo_systematicperformanceevaluation_2018; @saelens_comparisonsinglecelltrajectory_2019; @soneson_biasrobustnessscalability_2018].
-However, a limitation of these existing simulators is that they would require significant methodological alterations to add additional modalities or experimental conditions (Table\ \ref{tab:comparison}).
+Simulating realistic data requires that the underlying biology is recapitulated as best as possible, and in the case of transcriptomics data this typically involves modelling the underlying gene regulatory networks.
+Simulators of "bulk" microarray or RNA-sequencing profiles simulate biological processes (e.g.
+transcription, translation) by translating a database of known regulatory interactions into a set of ordinary differential equations (ODE) [@roy_systemgeneratingtranscription_2008; @hache_gengesystematicgeneration_2009; @schaffter_genenetweaversilicobenchmark_2011; @vandenbulcke_syntrengeneratorsynthetic_2006].
+These methods have been instrumental in performing benchmarking studies [@prill_rigorousassessmentsystems_2010; @marbach_revealingstrengthsweaknesses_2010; @marbach_wisdomcrowdsrobust_2012].
+However, the advent of single-cell omics introduced several new types of analyses (e.g.
+trajectory inference, RNA velocity, cell-specific network inference) which exploit the higher resolution of single-cell versus bulk omics [@luecken_currentbestpractices_2019].
+In addition, the data characteristics of single-cell omics are vastly different from bulk omics, typically having much lower library sizes and a higher dropout rate, but also a high number of profiles [@vallejos_normalizingsinglecellrna_2017].
+The low library sizes, in particular, are problematic as ODEs are ill-suited for performing low-molecule simulations [@gillespie_exactstochasticsimulation_1977].
+This necessitates the development of new single-cell simulators.
+
+To this end, single-cell omics simulators emulate the technical procedures from single-cell omics protocols.
+Simulators such as Splatter [@zappia_splattersimulationsinglecell_2017], powsimR [@vieth_powsimrpoweranalysis_2017], PROSSTT [@papadopoulos_prossttprobabilisticsimulation_2019] and SymSim [@zhang_simulatingmultiplefaceted_2019]) have already been widely used to compare single-cell methods [@street_slingshotcelllineage_2018; @parra_reconstructingcomplexlineage_2019; @lummertzdarocha_reconstructioncomplexsinglecell_2018; @lin_scclassifysamplesize_2020] and perform independent benchmarks [@duo_systematicperformanceevaluation_2018; @saelens_comparisonsinglecelltrajectory_2019; @soneson_biasrobustnessscalability_2018].
+However, by focusing more on simulating the single-cell omics protocol (e.g.
+RNA capture, amplification, sequencing) and less on the underlying biology (e.g.
+transcription, splicing, translation), their applicability and reusability is limited towards the specific application for which they were designed (e.g.
+benchmarking clustering or differential expression methods), and extending these tools to include additional modalities or experimental conditions is challenging.
+
+We introduce dyngen, a method for simulating cellular dynamics at a single-cell, single-transcript resolution (Figure\ \ref{fig:overview}).
+This problem is tackled in three fully-configurable main steps.
+First, biological processes are mimicked by translating a gene regulatory network in a set of reactions (regulation, transcription, splicing, translation).
+Second, individual cells are simulated using Gillespie's stochastic simulation algorithm [@gillespie_exactstochasticsimulation_1977], which is designed to work well in low-molecule simulations.
+Finally, (real) reference datasets are used to emulate single-cell omics profiling protocols.
+
+Throughout a simulation, dyngen tracks many layers of information, including the abundance of any molecule in the cell, the progression of the cell along a dynamic process, and the activation strength of individual regulatory interactions.
+In addition, dyngen can simulate a large variety of dynamic processes (e.g.
+cyclic, branching, disconnected) as well as a broad range of experimental conditions (e.g.
+batch effects and time-series, perturbation and single-cell knockdown experiments).
+For these reasons, dyngen can cater to a wide range of benchmarking applications, including trajectory inference, trajectory alignment, and trajectory differential expression (Table\ \ref{tab:comparison}).
 
@@
 \end{figure}
-dyngen was designed to include all of these functionalities and more by design.
-Its methodology allows tracking many layers of information throughout the simulation, including the abundance of any molecule in the cell, the progression of the cell along a dynamic process, and the activation strength of individual regulatory interactions.
-dyngen can simulate a large variety of dynamic processes (e.g.
-cyclic, branching, disconnected) as well as a broad range of experimental conditions (e.g.
-batch effects and time-series, perturbation and knockdown experiments).
-The fine-grained controls over simulation parameters allow dyngen to be applicable to a broad range of use-cases to simulate dynamic biological processes.
-The original design of dyngen was motivated by the plethora of methods available for trajectory inference, where dyngen allowed the first large-scale benchmarking of such methods [@saelens_comparisonsinglecelltrajectory_2019; @vandenberge_trajectorybaseddifferentialexpression_2020].
 
-Here, we highlight the novel functionality of dyngen by evaluating three novel types of computational approaches for which no simulation engines exist yet: single-cell network inference, trajectory alignment and RNA velocity (Figure\ \ref{fig:applications}).
+<!-- RESULTS -->
+We demonstrate dyngen's broad applicability by evaluating three novel types of computational approaches for which no simulation engines exist yet: cell-specific network inference, trajectory alignment and RNA velocity (Figure\ \ref{fig:applications}).
 We emphasize that our main aim here is to illustrate the potential of dyngen for these evaluations, rather than performing large-scale benchmarking, which would require assessing many more quantitative and qualitative aspects of each method [@weber_essentialguidelinescomputational_2019].
@@
       \textbf{A:} Trajectory alignment aligns two trajectories between samples.
-dyngen can simulate different scenarios in which alignment is necessary, such as a premature stop as shown here.
-We compared two versions of dynamic time warping (DTW): normal DTW aligns all individual cells, while DTW+smoothing first smooths the expression data and then uses DTW to align the smoothed cells.
+We use dyngen to evaluate dynamic time warping (DTW) and cellAlign when aligning two linear trajectories with different kinetic parameters.
       \textbf{B:} RNA velocity calculates for each cell the direction in which the expression of each gene is moving.
@@
 We observe that DTW+smoother performs significantly better than the unsmoothed version across all levels of noise.
+Note that, in this comparison, only linear trajectory alignment is performed.
+While dyngen can generate non-linear trajectories (e.g.
+cyclic or branching), constructing a quantitative accuracy metric for non-linear trajectory alignment is not trivial and an avenue for future work.[a]
 
-**RNA velocity** methods use the relative ratio between pre-mRNA and mature mRNA reads to predict the velocity at which the RNA expression of genes is increasing or decreasing [@zeisel_coupledpremrnamrna_2011;@lamanno_rnavelocitysingle_2018].
-Already two algorithms are currently available for estimating the RNA velocity vector from spliced and unspliced counts: velocyto [@lamanno_rnavelocitysingle_2018] and scvelo [@bergen_generalizingrnavelocity_2019].
+**RNA velocity** methods use the relative ratio between pre-mRNA and mature mRNA reads to predict the rate of increase/decrease of RNA molecule abundance, as this can be used to predict the directionality of single cell differentiation in trajectories [@zeisel_coupledpremrnamrna_2011;@lamanno_rnavelocitysingle_2018].
+Already two algorithms are currently available for estimating the RNA velocity vector from spliced and unspliced counts: velocyto [@lamanno_rnavelocitysingle_2018] and scvelo [@bergen_generalizingrnavelocity_2020].
 Yet, to date, no quantitative assessment of their accuracy has been performed, mainly due to the difficulty in obtaining real ground-truth data to do so.
-In contrast, the ground-truth RNA velocity can be easily extracted from a dyngen, as it is possible to store the rate at which mRNA molecules are being transcribed and degraded (called the propensity) at any particular point in time.
-We executed scvelo and velocyto (with 6 different parameter settings in total) on 102 datasets with varying degrees of difficulty (easy, medium, hard) and a variety of backbones (including linear, bifurcating, cyclic, disconnected).
+In contrast, the ground-truth RNA velocity can be easily extracted from a dyngen simulation, as it is possible to store the rate at which mRNA molecules are being transcribed and degraded at any particular point in time.
+We executed velocyto and scvelo (with 2 different parameter settings, stochastic and dynamical) on 42 datasets with a variety of backbones (including linear, bifurcating, cyclic, disconnected).
 We evaluated the predictions using two metrics (Figure\ \ref{fig:velocity}), one which directly compares the predicted RNA velocity of each gene with the ground-truth RNA velocity (called the "velocity correlation"), and one which compares the direction of the ground-truth trajectory embedded in a dimensionality reduction with the average RNA velocity of cells in that neighbourhood (called the "velocity arrow cosine").
-We found that in almost all cases, both velocyto and scvelo obtained high velocity arrow cosine, meaning that the overview obtained by embedding RNA velocity arrows in a 2D or 3D dimensionality reduction allows users to correctly identify the general progression of cells.
-However, depending on the difficulty of the dataset, the correlation between the predicted RNA velocity and the ground-truth RNA velocity varies between high (>0.75) to low (<0.25).
-For this particular metric, the dynamic estimation of velocyto performs significantly worse than any of the other prediction methods.
+While both velocyto and scvelo obtained high scores for the velocity arrow cosine metric (overall 25th percentile = 0.606), the velocity correlation is rather low (overall 75th percentile = 0.156).
+This means that predicting the RNA velocity (i.e.
+transcription rate minus the decay rate) for any particular gene is challenging, but the combined information is very informative in determining the directionality of cell progression in the trajectory.
+In terms of velocity correlation, no method performed significantly better than the other, whereas scvelo stochastic performed slightly worse than scvelo dynamical and velocyto in terms of velocity arrow cosine score.
 
 **Cell-specific network inference** (CSNI) methods predict not only which transcription factors regulate which target genes, but also aim to identify how active each interaction is in each of the cells, since interactions can be turned off and on depending on the cellular state.
-While a few pioneering CSNI approaches have already been developed [@aibar_scenicsinglecellregulatory_2017; @kuijjer_estimatingsamplespecificregulatory_2019; @liu_personalizedcharacterizationdiseases_2016], a quantitative assessment of the performance is until now lacking.
+While a few pioneering CSNI approaches have already been developed [@aibar_scenicsinglecellregulatory_2017; @kuijjer_estimatingsamplespecificregulatory_2019; @liu_personalizedcharacterizationdiseases_2016], a quantitative assessment of their performance is until now lacking.
 This is not surprising, as neither real nor in silico datasets of cell-specific or even cell-type-specific interactions exist that are large enough so that it can be used as a ground-truth for evaluating CSNI methods.
@@
 We used this ground-truth to compare the performance of three CSNI methods (Figure\ \ref{fig:scgrn}): LIONESS [@kuijjer_estimatingsamplespecificregulatory_2019], SSN [@liu_personalizedcharacterizationdiseases_2016] and SCENIC [@aibar_scenicsinglecellregulatory_2017].
-We calculated the AUROC and AUPR score for each cell individually.
-Computing the mean AUROC and AUPR per dataset showed that pySCENIC significantly outperforms LIONESS + Pearson, which in turn outperforms SSN*.
+For each dataset, we computed the mean AUROC and AUPR scores of the individual cells.
+Comparing the mean AUROC and AUPR showed that pySCENIC significantly outperforms both LIONESS and SSN, and in turn that LIONESS significantly outperforms SSN.
+The poor performance of SSN is expected, as its methodology for predicting a cell-specific is simply computing the difference in Pearson correlation values applied to the whole dataset and the whole dataset minus one sample.
+This strategy performs poorly in large datasets where cell correlations are high, as the removal of one cell will not yield large differences in correlation values and will result in mostly noise.
+Overall, pySCENIC almost always performs better than LIONESS, except for a few datasets where LIONESS does manage to obtain a higher AUROC score.
+However, by using a different internal network inference (e.g.
+GENIE3 [@huynh-thu_inferringregulatorynetworks_2010] or pySCENIC's GRNBoost2 [@moerman_grnboost2arboretoefficient_2019]) could significantly increase the performance obtained by LIONESS.
 
@@
 In this way, dyngen allows designing and evaluating the performance and robustness of new types of computational analyses before experimental data becomes available, comparing which experimental protocol is the most cost-effective in producing qualitative and robust results in downstream analysis.
-One major assumption of dyngen is that cells are regarded as standalone entities that are well mixed.
-Splitting up the simulation space into separate subvolumes could pave the way to better study key cellular processes such as cell division, intercellular communication, and migration [@smith_spatialstochasticintracellular_2019].
+One major assumption of dyngen is that cells are assumed to be well-mixed and independent from each other.
+Subdividing a cell into multiple 2D or 3D subvolumes or allowing cells to exchange molecules, respectively, could pave the way to better study key cellular processes such as cell division, intercellular communication, and migration [@smith_spatialstochasticintracellular_2019].
 
-# Availability
-dyngen is available as an R package on GitHub at [github.com/dynverse/dyngen](https://github.com/dynverse/dyngen).
+# Data Availability[b]
+
+# Code Availability
+dyngen is available as an open-source R package on CRAN at [cran.r-project.org/package=dyngen](https://cran.r-project.org/package=dyngen).
 The analyses performed in this manuscript are available on GitHub at [github.com/dynverse/dyngen_manuscript](https://github.com/dynverse/dyngen_manuscript).
@@
 # Author contributions
-
 * W.S. and R.C. designed the study.
@@
 
-\printbibliography
 \newpage
@@
 ### Sample cells
-In this step, $N$ cells are sampled the simulations.
+In this step, $N$ cells are sampled from the simulations.
 Two approaches are implemented: sampling from an unsynchronised population of single cells (snapshot) or sampling at multiple time points in a synchronised population (time series).
@@
 
-**Time series** Assuming that the final time of the simulations is $T$, the interval $[0, T]$ is divided into $k$ equal intervals of width $w$ separated by $k-1$ gaps of width $g$.
+**Time series** Assuming that the final time of the simulation is $T$, the interval $[0, T]$ is divided into $k$ equal intervals of width $w$ separated by $k-1$ gaps of width $g$.
 $N_i = N / k$ cells are sampled uniformly from each interval (Figure\ \ref{fig:sample_cells}B), rounded such that $\sum N_i = N$.
@@
 ## Determining the ground-truth trajectory {#sec:dyngen-groundtruth}
-To construct the ground-truth trajectory, the user needs to provide the ground-truth state network alongside the initial module network (Figure~\ref{fig:example_backbones}).
+To construct the ground-truth trajectory, the user needs to provide the ground-truth state network alongside the initial module network (Figure\ \ref{fig:example_backbones}).
 Each edge in the state network specifies which modules are allowed to change in expression in transitioning from one state to another.
@@
 
-As an example, consider the cyclic trajectory shown in Figure~\ref{fig:example_backbones}.
+As an example, consider the cyclic trajectory shown in Figure\ \ref{fig:example_backbones}.
 State S0 begins with an expression vector of all zero values.
@@
 ## Comparison of cell-specific network inference methods {#sec:dyngen-nicompare}
-14 datasets were generated using the 14 different predefined backbones.
+42 datasets were generated using the 14 different predefined backbones and three different seeds.
 For every cell in the dataset, the transcriptomics profile and the corresponding cell-specific ground-truth regulatory network was determined (Section\ \ref{sec:dyngen-extractgrn}).
 
-Several cell-specific NI methods were considered for comparison: SCENIC [@aibar_scenicsinglecellregulatory_2017], LIONESS [@kuijjer_estimatingsamplespecificregulatory_2015; @kuijjer_estimatingsamplespecificregulatory_2019], and SSN [@liu_personalizedcharacterizationdiseases_2016].
+We selected three cell-specific NI methods: SCENIC [@aibar_scenicsinglecellregulatory_2017], LIONESS [@kuijjer_estimatingsamplespecificregulatory_2015; @kuijjer_estimatingsamplespecificregulatory_2019], and SSN [@liu_personalizedcharacterizationdiseases_2016].
 
@@
 
-SCENIC [@aibar_scenicsinglecellregulatory_2017] is a pipeline that consists of four main steps.
-Step 1: classical network inference is performed with arboreto, which is similar to GENIE3 [@huynh-thu_inferringregulatorynetworks_2010].
-Step 2: select the top 10 regulators per target.
+SCENIC [@aibar_scenicsinglecellregulatory_2017] consists of four main steps.
+First, classical network inference is performed with stochastic gradient boosting machines using `arboreto` [@moerman_grnboost2arboretoefficient_2019].
+Second, the top 10 regulators of every target gene are selected.
 Interactions are grouped together in 'modules'; each module contains one regulator and all of its targets.
-Step 3: filter the modules using motif analysis.
-Step 4: for each cell, determine an activity score of each module using AUCell.
+Next, the modules are filtered using motif analysis.
+Finally, for each module and each cell, an activity score is calculated using AUCell.
 As a post-processing of this output, all modules and the corresponding activity scores are combined back into a cell-specific GRN consisting of (cell, regulator, target, score) pairs.
-For this analysis, the Python implementation of SCENIC was used, namely pySCENIC.
+For this analysis, the Python implementation of SCENIC was used, namely pySCENIC [@vandesande_scalablescenicworkflow_2020].
 Since dyngen does not generate motif data, step 3 in this analysis is skipped.
-
-Cell-specific network inference (CSNI) predicts the regulatory network of every individual cell.
-We evaluate each cell-specific regulatory network with typical metrics for network inference: the Area Under the Receiver Operating Characteristics-curve (AUROC) and Area Under the Precision Recall-curve (AUPR).
-We evaluate three CSNI methods by computing the mean AUROC and AUPR across all cells.
 
@@
 
+We compared the mean AUROC and AUPR scores obtained by the three CSNI methods across all datasets by performing pairwise non-parametric paired Durbin-Conover tests [@conover_multiplecomparisonsprocedures_1979] using `pairwiseComparisons` [@patil_pairwisecomparisonsmultiplepairwise_2019].
+Reported p-values are adjusted for multiple testing using Holm correction [@holm_simplesequentiallyrejective_1979].
+
 ## Comparison of RNA velocity methods {#sec:dyngen-velcompare}
 
-For each of the 14 different predefined backbones, nine datasets were generated with three difficulty settings and three different seeds.
-The different difficulty settings were obtained by multiplying the transcription rate by a factor of 25 (easy), 5 (medium) and 1 (hard).
-After manual quality control, the easy and medium datasets for backbones "bifurcating_converging", "bifurcating_loop", "converging" and "disconnected" were removed, resulting in a final collection of 102 datasets.
+Three datasets were generated for each of the 14 different predefined backbones, resulting in a collection of 42 datasets.
+Throughout each of the simulation, the propensity of the transcription and mRNA decay is collected, as the RNA velocity of a gene at any point in the simulation is the difference between the transcription propensity and the mRNA decay propensity.
 
-We calculated two evaluation metrics: the velocity correlation and the velocity arrow cosine.
+We applied two RNA velocity methods: velocyto [@lamanno_rnavelocitysingle_2018], as implemented in the `velocyto.py` package, and scvelo method [@bergen_generalizingrnavelocity_2020], as implemented in the `scvelo` package.
+For scvelo, we chose two parameter settings for "mode", namely "stochastic" and "dynamical".
+For both methods, we used the same normalized data as provided by dyngen, with no extra cell or feature filtering, but otherwise matched the parameters to their respective tutorial vignettes as well as possible.
+
+We compared each RNA velocity prediction to the ground-truth using two metrics: the velocity correlation and the velocity arrow cosine.
 For the velocity correlation, we extracted a ground truth RNA velocity by subtracting for each mRNA molecule the propensity of its production by the propensity of its degradation.
@@
 
-We compared two RNA velocity methods.
-The velocyto method [@lamanno_rnavelocitysingle_2018], as implemented in the velocyto.py package, in which we varied the "assumption" parameter between "constant_unspliced" and "constant_velocity".
-The scvelo method [@bergen_generalizingrnavelocity_2019], as implemented in the python scvelo package [scvelo.de](http://scvelo.de), in which we varied the "mode" parameter between "deterministic", "stochastic", "dynamical", "dynamical_residuals".
-For both methods, we used the same normalized data as provided by dyngen, with no extra cell or feature filtering, but otherwise matched the parameters to their respective tutorial vignettes as well as possible.
+We compared the velocity correlation and velocity arrow cosine scores obtained by velocyto and scvelo across all datasets by performing pairwise non-parametric paired Durbin-Conover tests [@conover_multiplecomparisonsprocedures_1979] using `pairwiseComparisons` [@patil_pairwisecomparisonsmultiplepairwise_2019].
+Reported p-values are adjusted for multiple testing using Holm correction [@holm_simplesequentiallyrejective_1979].
 
-To visualize the velocity on an embedding, we used the "velocity_embedding" function, implemented in the scvelo python package.
+## Comparison of trajectory alignment {#sec:dyngen-tacompare}[c]
 
-## Comparison of trajectory alignment with increasing levels of noise {#sec:dyngen-tacompare}
+Four custom linear backbones of varying sizes were constructed.
+For each of these backbones, 10 datasets were generated with 10 different seeds, resulting in a total of 40 datasets.
+Every dataset is generated in three main steps.
+First, the GRN is generated based on the given backbone.
+Next, generating the kinetics, gold standard, and cells is performed twice, resulting in two sub-datasets.
+Finally, the two sub-datasets are combined and cells are sampled from the combined dataset.
+Since the two sub-datasets were simulated with different kinetic parameters, the combined dataset will contain two trajectories.
 
-10 base datasets, containing a linear trajectory, were generated using dyngen.
-For each dataset, cells and experiments were generated twice, resulting in 20 paired datasets containing a similar trajectory with slightly different kinetics.
-For each of these pairs of datasets, we generated 10 progressively noisier pairs, in which we added noise to the complete count matrix.
-Let $q$ be the 75th quantile of the non-zero values in the count matrix. The noise added to the count matrix is calculated as follows, with $i$ being the noise parameter.
+On each combined dataset we applied two trajectory alignment methods, Dynamic Time Warping (DTW) [@giorgino_computingvisualizingdynamic_2009] and cellAlign [@alpert_alignmentsinglecelltrajectories_2018].
+DTW is designed to align temporal sequences by dilating or contracting the sequences to best match each other.
+cellAlign …[d] .
+
+To evaluate a trajectory alignment method on a paired dataset we computed the geodesic distances of each cell from the start of the trajectory, also called the _pseudotime_.
+For each dataset, the pseudotime values are rescaled between 0 and 1 to allow for easier comparison.
+A trajectory alignment produces a sequence of index pairs $[(i_0, j_0), (i_1, j_1), \ldots, (i_N, j_N)]$, where $i_0$ and $j_0$ are equal to 0 (the first position in both pseudotime series), $i_N$ and $j_N$ are equal to the respective last positions in the pair of pseudotime series, and $[i_0, i_1, \ldots, i_N]$ and $[j_0, j_1, \ldots, j_N]$ are in ascending order and can contain duplicates values.
+The 'score'[e] metric is defined as follows, where a score of 0 represents the worst possible ordering and 1 represents the best possible ordering.
+
 \begin{equation}
- \textrm{countmatrix} + \frac{1}{\sqrt{2\pi}}qie^{\frac{x^2}{2(qi)^2}}  \quad \forall i \in \{0.1, 0.2, ... 1.0\}
+  \textrm{score} = 1 - \textrm{area\_under\_curve}(pt_1[i_0 .. i_N] + pt_2[j_0 .. j_N], abs(pt_1[i_0 .. i_N] - pt_2[j_0 .. j_N]))
 \end{equation}
-
-We aligned each of these 100 pairs of trajectories (the first trajectory in the pair with the second trajectory in the pair) using Dynamic Time Warping (DTW) [@giorgino_computingvisualizingdynamic_2009].
-DTW is designed to align temporal sequences by dilating or contracting the sequences to best match each other.
-We compared the alignment of DTW with the alignment of DTW after first smoothing the expression along the trajectory using a Gaussian kernel with window size 0.05.
-
-To evaluate a trajectory alignment method on a paired dataset we computed the geodesic distances of each cell from the start of the trajectory.
-The geodesic distances between the pair of datasets are directly comparable to each other, and thus cells from the first trajectory should align relatively closely to cells from the second trajectory.
-We after aligning the cells, we calculate the mean absolute difference between the geodesic distances of the pairs of cells, which should ideally be 0.
 
@@
             - & Knockdown experiment & \no & \no & \no & \no & \yes \\ \grayline
-    - & Time-series & \no & \no & \no & \no & \yes \\
+            - & Time-series & \no & \no & \no & \no & \yes \\ \grayline
             - & Snapshot & \no & \no & \no & \no & \yes \\
@@
             - & Cell-specific network inference & \no & \no & \no & \no & \yes \\ \grayline
-    - & Differential expression & \yes & \no & \no & \no & \yes $ ^2$ \\ \grayline
+            - & Differential expression & \yes & \no & \no & \no & \no \\ \grayline
+            - & Trajectory differential expression & \no & \no & \no & \no & \yes $ ^2$ \\ \grayline
             - & Batch effect correction & \yes & \no & \no & \no & \yes \\ \grayline
@@
     \centering
-    \includegraphics[width=\linewidth]{result_files/usecase_trajectory_alignment/usecase.pdf}
+    \includegraphics[width=\linewidth]{result_files/usecase_trajectory_alignment/supp_fig.pdf}
     \caption{
@@
   \textbf{G:} Shows the influence of added noise to the different processing methods. We can see that DTW + smoothing performs best in noisy circumstances.
-    }
+    }[f]
     \label{fig:traj_align}
@@
     \centering
-    \includegraphics[width=\linewidth]{result_files/usecase_rna_velocity/usecase.pdf}
+    \includegraphics[width=\linewidth]{result_files/usecase_rna_velocity/supp_fig.pdf}
     \caption{
             \textbf{dyngen allows benchmarking of RNA velocity methods.} 
-    \textbf{A:} An example bifurcating cycle dataset, with as illustration the expression and ground truth velocity of a gene D1\_TF1 that goes up and down in one branch of the trajectory. 
-\textbf{B:} The RNA velocity estimates of gene D1\_TF1 by the different methods.
+            \textbf{A:} The ground-truth information of a bifurcating dataset: ground-truth trajectory (left), gene expression of a gene B5\_TF1 (middle), and the RNA velocity of B5\_TF1 (right).
+\textbf{B:} The RNA velocity estimates of gene B5\_TF1 by the different methods.
 \textbf{C:} The velocity stream plots produced from the predictions of each method, as generated by scvelo.
@@
     \centering
-    \includegraphics[width=.8\linewidth]{result_files/usecase_network_inference/cni.pdf}
+    \includegraphics[width=\linewidth]{result_files/usecase_network_inference/supp_fig.pdf}
     \caption{
@@
             \textbf{B:} CSNI methods were executed to predict the regulatory interactions that are active in each cell specifically.
-Using the ground-truth cell-specific GRN, the performance of each method was quantified on 14 dyngen datasets.
+Using the ground-truth cell-specific GRN, the performance of each method was quantified on 42 dyngen datasets.
     }
@@
 \end{figure}
+
+[a]todo: update
+[b]// todo: upload temporary data to zenodo?
+[c]todo: update this section
+[d]@louise.deconinck@gmail.com could you add a description for cellAlign?
+_Assigned to Louise_
+[e]TODO: rename
+[f]TODO: update text

### 2021-03-25 analysis/usecase_trajectory_alignment/4_plots.R
Fix merge

@@ -55,7 +55,6 @@ cal_dens
 dtw_dens <- plot_density(dtw_alignment, title = NULL)
 dtw_dens
 
-<<<<<<< HEAD
 # PART 2: dtw heatmaps ----------------------------------------------------
 # d1 <- readRDS(exp$dataset_file("linear1_1_0.5"))
 # d2 <- readRDS(exp$dataset_file("linear1_2_0.5"))
@@ -87,8 +86,7 @@ t1
 t2 <- (ground_truth + prediction) / (dtw_dens + results)
 t2
 ggsave(exp$result("usecase1.pdf"), t1, height = 7, width = 9)
-=======
->>>>>>> dd81428b6598c8e698684bb87ce30442b59030b8
+
 
 # PART 2bis: naupa explanation --------------------------------------------
 

### 2021-03-25 analysis/summary/source_data.R
fix r pitfall

@@ -9,7 +9,7 @@ out <- list()
 exp_a <- start_analysis("usecase_trajectory_alignment")
 scores_a <- read_rds(exp_a$result("results.rds"))
 
-out$trajectory_alignment_summary <- scores_a %>% select(-distance, -aupt) %>% rename(dataset_id = id, method_id = method)
+out$trajectory_alignment_summary <- scores_a %>% select(-mean_distance) %>% rename(dataset_id = id, method_id = method)
 
 # RNA velocity
 exp_b <- start_analysis("usecase_rna_velocity")

### 2021-03-25 analysis/usecase_trajectory_alignment/2_evaluate_methods.R

@@ -7,7 +7,7 @@ exp <- start_analysis("usecase_trajectory_alignment")
 design_smoothing <-
   crossing(
     read_rds(exp$result("design_datasets.rds")),
-    method = forcats::fct_inorder(c("DTW", "cellAlign"))
+    method = c("DTW", "cellAlign")
   )
 
 #' design_smoothing %>% mutate(rn = row_number()) %>% dynutils::extract_row_to_list(1) %>% list2env(.GlobalEnv)
@@ -24,22 +24,16 @@ results <- exp$result("results.rds") %cache% pmap_dfr(
       dataset2 <- dataset %>% filter_cells(model == "right")
 
       out <- method_fun(dataset1, dataset2)
-      pt1_aligned <- out$pt1_aligned
-      pt2_aligned <- out$pt2_aligned
+      pt1 <- out$pt1_aligned
+      pt2 <- out$pt2_aligned
 
-      distance <- mean(abs(pt1_aligned - pt2_aligned))
+      distance <- mean(abs(pt1 - pt2))
 
-      aupt <- pracma::trapz(
-        pt1_aligned + pt2_aligned,
-        abs(pt1_aligned - pt2_aligned)
-      )
-
-      score <- 1 - aupt
+      abwap <- ta_abwap(pt1, pt2)
 
       design_smoothing[rn, ] %>% mutate(
-        distance,
-        aupt,
-        score
+        mean_distance = distance,
+        abwap
       )
     }
   }
@@ -48,16 +42,7 @@ results <- exp$result("results.rds") %cache% pmap_dfr(
 ggstatsplot::ggwithinstats(
   results,
   x = method,
-  y = score,
+  y = abwap,
   type = "np",
   pairwise.display = "all"
 )
-
-z <- ggstatsplot::ggwithinstats(
-  results,
-  x = method,
-  y = distance,
-  type = "np",
-  pairwise.display = "all"
-)
-z$labels$subtitle

### 2021-03-25 analysis/usecase_trajectory_alignment/3_plots.R
rework TA supp fig plotting

@@ -1,0 +1,408 @@
+library(tidyverse)
+library(dyngen.manuscript)
+library(patchwork)
+library(dynplot2)
+
+exp <- start_analysis("usecase_trajectory_alignment")
+
+
+# PART 0: Load example dataset -------------------------------------------
+# read dataset
+design_dataset <- read_rds(exp$result("design_datasets.rds"))
+file <- exp$dataset_file(design_dataset$id[[1]])
+dataset <- read_rds(file) %>% dynwrap::add_dimred(dimred = dyndimred::dimred_landmark_mds)
+
+# split up in two
+dataset1 <- dataset %>% filter_cells(model == "left", keep_dimred = TRUE)
+dataset2 <- dataset %>% filter_cells(model == "right", keep_dimred = TRUE)
+
+# add healthy/diseased grouping
+dataset$cell_info <- dataset$cell_info %>% mutate(
+  group = c("left" = "Healthy", "right" = "Diseased")[model]
+)
+group_palette <- c(Diseased = "#fd8d3c", Healthy = "#6baed6")
+
+# run methods
+dtw_out <- ta_methods$DTW(dataset1, dataset2)
+# cal_out <- ta_methods$cellAlign(dataset1, dataset2)
+
+
+# PART 1: create explanation plots ----------------------------------------
+# construct gold standard alignment
+step_progressions <- crossing(
+  dataset$milestone_network %>% select(from, to),
+  percentage = seq(0, 1, by = .2)
+) %>%
+  mutate(cell_id = paste0("step", row_number()))
+step_waypoints <- list(
+  milestone_percentages = dynwrap::convert_progressions_to_milestone_percentages(
+    cell_id = step_progressions$waypoint_id,
+    milestone_ids = dataset$milestone_ids,
+    milestone_network = dataset$milestone_network,
+    progressions = step_progressions
+  ) %>% rename(waypoint_id = cell_id)
+)
+step_proj <- dynwrap::project_waypoints(trajectory = dataset, space = dataset$dimred, waypoints = step_waypoints)
+
+ds <- data.frame(step_progressions, step_proj[step_progressions$cell_id, ]) %>%
+  as_tibble() %>%
[…]

### 2021-03-25 analysis/usecase_trajectory_alignment/4_plots.R

@@ -1,252 +1,0 @@
-library(tidyverse)
-library(dyngen.manuscript)
-library(patchwork)
-library(dynplot)
-
-exp <- start_analysis("usecase_trajectory_alignment")
-
-# PART 1: Explanation plot ------------------------------------------------
-part1plots <- read_rds(exp$result("usecase_separateplots.rds"))
-
-# PART 2: heatmaps --------------------------------------------------------
-
-# normalized minimum global distance computed
-res_plot <- ggstatsplot::ggwithinstats(
-  results,
-  x = method,
-  y = distance,
-  type = "np",
-  pairwise.comparisons = TRUE,
-  pairwise.display = TRUE
-)
-res_plot
-
-# results %>%
-#   ggplot() +
-#   geom_path(
-#     aes(noise, score, colour = method, group = paste0(method, "_", seed))
-#   ) +
-#   theme_bw() +
-#   theme_common() +
-#   labs(x = "Amount of added noise", y = "Distance (lower is better)", fill = "Processing method") +
-#   scale_colour_brewer(palette = "Set2")
-#
-# results %>% spread(method, score) %>%
-#   ggplot() + geom_point(aes(DTW, `DTW+smoothing`)) + geom_abline(intercept = 0, slope = 1)
-# results %>% spread(method, score) %>%
-#   ggplot() + geom_point(aes(`DTW+smoothing`, `cellAlign`)) + geom_abline(intercept = 0, slope = 1)
-
-# PART 2: dtw heatmaps
-
-# read dataset
-design_dataset <- read_rds(exp$result("design_datasets.rds"))
-file <- exp$dataset_file(design_dataset$id[[1]])
-dataset <- read_rds(file) %>% dynwrap::add_dimred(dimred = dyndimred::dimred_landmark_mds)
-
-# split up in two
-dataPastel2 <- dataset %>% filter_cells(model == "left", keep_dimred = TRUE)
-dataset2 <- dataset %>% filter_cells(model == "right", keep_dimred = TRUE)
-
-cal_alignment <- ta_methods$cellAlign(dataPastel2, dataset2)
-dtw_alignment <- ta_methods$DTW(dataPastel2, dataset2)
-
-cal_dens <- plot_density(cal_alignment, title = NULL)
-cal_dens
-dtw_dens <- plot_density(dtw_alignment, title = NULL)
-dtw_dens
-
[…]

### 2021-03-25 analysis/summary/figure_2.R
fix TA results

@@ -81,7 +81,8 @@ ga3 <-
   theme_common() +
   labs(x = NULL, y = "ABWAP score") +
   scale_fill_brewer(palette = "Pastel2") +
-  theme(legend.position = "none")
+  theme(legend.position = "none") +
+  expand_limits(y = c(0, 1))
 
 
 #################

### 2021-03-25 analysis/usecase_trajectory_alignment/3_plots.R

@@ -145,23 +145,6 @@ write_rds(plots, exp$result("usecase_separateplots.rds"), compress = "xz")
 
 # PART 2: heatmaps --------------------------------------------------------
 
-# best_pt <- tibble(
-#   pt = sort(c(res1$pseudotime, res2$pseudotime)),
-#   in1 = names(pt) %in% names(res1$pseudotime),
-#   in2 = names(pt) %in% names(res2$pseudotime),
-#   index2 = cumsum(in1),
-#   index1 = cumsum(in2)
-# )
-#
-# heatmap_paths <- bind_rows(
-#   # tibble(pt1 = cal_out$pt1, pt2 = cal_out$pt2, method = "cellAlign"),
-#   tibble(pt1 = dtw_out$pt1, pt2 = dtw_out$pt2, method = "Prediction", index1 = dtw_out$index1, index2 = dtw_out$index2),
-#   best_pt %>% select(-pt, -in1, -in2) %>% mutate(method = "Best possible\nalignment")
-#   # tibble(pt1 = c(0, 1, 1), pt2 = c(0, 0, 1), index1 = pt1 * 1000, index2 = pt2 * 1000, method = "Worst"),
-#   # tibble(pt1 = c(0, 1), pt2 = c(0, 1), index1 = pt1 * 1000, index2 = pt2 * 1000, method = "Best possible\nalignment")
-# ) %>%
-#   mutate(method = factor(method, levels = c("Prediction", "Best possible\nalignment", "Worst")))
-
 heatmap_paths <- dtw_out[c("index1", "index2")] %>% as_tibble()
 
 heatmap_raster <-
@@ -176,9 +159,9 @@ dtw_dens <-
   ggplot(mapping = aes(index1, index2)) +
   geom_raster(aes(fill = distance), heatmap_raster, alpha = .9) +
   geom_path(data = heatmap_paths) +
-  scale_fill_distiller(palette = "RdYlGn", breaks = range(rasterdf$distance), labels = c("min", "max")) +
-  scale_x_continuous(expand = c(0, 0), breaks = quantile(linedf$x, c(0, .5, 1)), labels = c("0.0", "0.5", "1.0")) +
-  scale_y_continuous(expand = c(0, 0), breaks = quantile(linedf$y, c(0, .5, 1)), labels = c("0.0", "0.5", "1.0")) +
+  scale_fill_distiller(palette = "RdYlGn", breaks = range(heatmap_raster$distance), labels = c("min", "max")) +
[…]

### 2021-03-25 package/R/ta_methods.R

@@ -22,8 +22,8 @@ ta_methods <- list(
     )
 
     alignment <- cellAlign::globalAlign(
-      interp1$interpolatedVals,
-      interp2$interpolatedVals,
+      cellAlign::scaleInterpolate(interp1)$scaledData,
+      cellAlign::scaleInterpolate(interp2)$scaledData,
       scores = list(query = interp1$traj, ref = interp2$traj),
       sigCalc = FALSE
     )

### 2021-03-25 manuscript/render.Rmd
update manuscript

@@
 ---
-title: "dyngen: a multi-modal simulator for spearheading new single-cell omics analyses"
-author:
-- Robrecht Cannoodt*
-- Wouter Saelens*
-- Louise Deconinck
-- Yvan Saeys
-date: "`r format(Sys.time(), '%d %B %Y')`"
+title: "Response to reviewers' comments"
 output:
@@
 ---
+<!-- Notes (remove this before submitting)
+dyngen manuscript v4  
+github.com/dynverse/dyngen  
+github.com/dynverse/dyngen_manuscript
+dyngen.dynverse.org
 
-```{r setup, include=FALSE}
-knitr::opts_chunk$set(collapse=TRUE, results="hold")
-library(tidyverse)
-```
+Antwoord naar editor: wordt gebruikt in dynverse, tradeseq, cellrank, BEELINE
+-->
 
-# Abstract
-<!-- format: max. 70 words -->
-We present dyngen, a novel, multi-modal simulation engine for studying dynamic cellular processes at single-cell resolution.
-dyngen is more flexible than current single-cell simulation engines, and allows better method development and benchmarking, thereby stimulating development and testing of novel computational methods.
-We demonstrate its potential for spearheading novel computational methods on three novel applications: aligning cell developmental trajectories, cell-specific regulatory network inference and estimation of RNA velocity.
+# dyngen rebuttal
 
-# Main text
-<!-- format: 1500 Words, no headings, max. 2 figures, max. 20 references-->
-Single-cell simulation engines are becoming increasingly important for testing and benchmarking computational methods, a pressing need in the widely expanding field of single-cell biology.
-Complementary to real biological data, synthetic data provides a valuable alternative where the actual ground truth is completely known and thus can be compared to, in order to make quantitative evaluations of computational methods that aim to reconstruct this ground truth [@zappia_splattersimulationsinglecell_2017].
-In addition, simulation engines are more flexible when it comes to stress-testing computational methods, for example by varying the parameters of the simulation, such as the amount of noise, samples, and cells measured, allowing benchmarking of methods over a wide range of possible scenarios.
-In this way, they can even guide the design of real biological experiments, finding out the best conditions to be used as input for subsequent computational pipelines.
+We would like to thank the reviewer for the constructive comments, which has motivated us to greatly improve the experiments performed in this work and the documentation provided for the tool.
+Notable changes are:
 
-Another, more experimental use of simulation engines is their important role in spearheading the development of novel computational methods, possibly even before real data is available.
-In this way, simulation engines can be used to assess the value of novel experimental protocols or treatments.
-Simulation engines are also increasingly important when it comes to finding alternatives to animal models, for example for drug testing and precision medicine.
-In such scenarios, cellular simulations can act as digital twins, offering unlimited experimentation _in silico_ [@bjornsson_digitaltwinspersonalize_2019].
+ * We created a documentation website available at [dyngen.dynverse.org](https://dyngen.dynverse.org). [a]Several insights 
+ * We simplified and reran all use-case experiments, which allowed us to add some interpretation of the results for each of the use cases.
+ * We performed new experiments assessing the scalability of dyngen and the similarity in output in comparison to a reference (real) dataset. The results are included in the supplementary material.
+ As we think that several of the remarks raised by the reviewer are also interesting for the target audience of the paper, we not only include our responses to these remarks here, but in addition have created several tutorials that can be used to perform these experiments with dyngen.
+These tutorials now also allow users to reproduce some of the additional analyses requested by the reviewer, and they have been added to a new website that we created for our method, with a lot more documentation website ([dyngen.dynverse.org](https://dyngen.dynverse.org)).
+* … todo, add more ...[b]
 
-Simulating realistic data requires that the underlying biology is recapitulated as best as possible, and in the case of transcriptomics data this typically involves modelling the underlying gene regulatory networks.
+## Comment 1
+> Bioinformatics has been looking into GRNs, their simulation and estimation for more than a decade (in bulk at the time), and I miss some important references and insights from then.
+Such as alternate definitions of what a GRN is - eg does logic play a role, should you use boolean networks (see Albert&Othma, JTB 2003) -, do old-school bulk simulators such as GeneNetWeaver (Bioinformatics 2011) be used, can one use transformation methods (eg Wittmann et al, BMC Sysbio 2009) and how is this different from what has been introduced here (what is novelty here).
+
+Certainly simulating expression datasets using GRNs has been around for a long time and dyngen is inspired heavily by such software (GeneNetWeaver in particular).
+Nevertheless, there are great differences between bulk and single-cell simulators:  the data characteristics of single-cell omics are vastly different from bulk omics, typically having much lower library sizes and a higher dropout rate, but also a high number of profiles.
+The low library sizes, in particular, are problematic as ODEs are ill-suited for performing low-molecule simulations.
+This necessitates the development of new single-cell simulators.
+
+We have added paragraph §3 to the introduction to discuss the relevance of bulk simulators and why they are not suitable for simulating single-cell expression data:
+
+> Simulating realistic data requires that the underlying biology is recapitulated as best as possible, and in the case of transcriptomics data this typically involves modelling the underlying gene regulatory networks.
 Simulators of "bulk" microarray or RNA-sequencing profiles simulate biological processes (e.g.
@@
 
-To this end, single-cell omics simulators emulate the technical procedures from single-cell omics protocols.
+## Comment 2
+> Similarly, the single cell genomics field has used simulations also for dynamics for some time as evaluation at least implicitely.
+Please discuss advances and differences or similarities to those, eg.
+to simulations from (Ocone, Bioinf 2015) or (Pratapa, Nat Meth 2020).
+
+We expanded upon paragraph §4 and made a more direct comparison to the methods listed in Supplemental Table 1. Paragraph §4:
+
+> To this end, single-cell omics simulators emulate the technical procedures from single-cell omics protocols.
 Simulators such as Splatter [@zappia_splattersimulationsinglecell_2017], powsimR [@vieth_powsimrpoweranalysis_2017], PROSSTT [@papadopoulos_prossttprobabilisticsimulation_2019] and SymSim [@zhang_simulatingmultiplefaceted_2019]) have already been widely used to compare single-cell methods [@street_slingshotcelllineage_2018; @parra_reconstructingcomplexlineage_2019; @lummertzdarocha_reconstructioncomplexsinglecell_2018; @lin_scclassifysamplesize_2020] and perform independent benchmarks [@duo_systematicperformanceevaluation_2018; @saelens_comparisonsinglecelltrajectory_2019; @soneson_biasrobustnessscalability_2018].
@@
 
-We introduce dyngen, a method for simulating cellular dynamics at a single-cell, single-transcript resolution (Figure\ \ref{fig:overview}).
+We did not include the reference to Ocone Bioinf 2015 as the advances proposed by this research were mainly related to performing Network Inference.
+In terms of generating synthetic data, the simulator used by Ocone et al.
+offered little advantage over aforementioned bulk simulators.
+We also did not include a reference to Pratapa et al., as their methodology is borrowed heavily from dyngen itself and offers little advantage over it.
+
+## Comment 3
+> One of the main aims of any simulation framework should be to capture features of real biological data.
+I’d be great if the authors could demonstrate the fidelity of the generated data by comparing it to some real datasets, e.g.
+distribution of mean gene expression and number of zero counts per gene.
+How does it perform compared to other simulation frameworks?
+This would help me as well as community to evaluate how realistic the induced gene correlations/dependencies are, and what may be missing.
+
+A vignette has been added to the package and supplementary material which uses countsimQC [@soneson_unifiedqualityverification_2018] to explore the characteristics of an dyngen simulation and the reference dataset used by it ([dyngen.dynverse.org/articles/advanced_topics/comparison_reference.html](https://dyngen.dynverse.org/articles/advanced_topics/comparison_reference.html)).
+The distribution plots in this vignette, while not perfect, nicely demonstrates that the output generated by dyngen mimics the given reference dataset for the evaluated metrics.
+
+## Comment 4
+> Simulated data for the described key applications of dyngen (network inference, trajectory alignment, and RNA velocity) has already been generated within the respective works of lioness (Fig.
+S1+S4), cellAlign (Fig.
+S6), velocyto (Fig.
+S5), scvelo (Fig.
+S3), dynamo (Fig.
+1), many of which also use Gillespie.
+What does dyngen add conceptually?
+
+We modified paragraphs §5 and §6 to better highlight the scientific contributions dyngen makes:
+
+> We introduce dyngen, a method for simulating cellular dynamics at a single-cell, single-transcript resolution (Figure\ \ref{fig:overview}).
 This problem is tackled in three fully-configurable main steps.
@@
 Second, individual cells are simulated using Gillespie's stochastic simulation algorithm [@gillespie_exactstochasticsimulation_1977], which is designed to work well in low-molecule simulations.
-Finally, (real) reference datasets are used to emulate single-cell omics profiling protocols.
-
-Throughout a simulation, dyngen tracks many layers of information, including the abundance of any molecule in the cell, the progression of the cell along a dynamic process, and the activation strength of individual regulatory interactions.
+Finally, real reference datasets are used to emulate single-cell omics profiling protocols.
+> 
+> Throughout a simulation, dyngen tracks many layers of information, including the abundance of any molecule in the cell, the progression of the cell along a dynamic process, and the activation strength of individual regulatory interactions.
 In addition, dyngen can simulate a large variety of dynamic processes (e.g.
@@
 
-\begin{figure}[t!]
-    \centering
-    \includegraphics[width=\linewidth]{result_files/summary/figure_1_edited.pdf}
-    \caption{
-      \textbf{Showcase of dyngen functionality.}
-      \textbf{A:} Changes in abundance levels are driven strictly by gene regulatory reactions.
-      \textbf{B:} The input Gene Regulatory Network (GRN) is defined such that it models a dynamic process of interest.
-      \textbf{C:} The reactions define how abundance levels of molecules change at any particular time point.
-      \textbf{D:} Firing many reactions can significantly alter the cellular state over time.
-      \textbf{E:} dyngen keeps track of the likelihood of a reaction firing during small intervals of time, called the propensity, as well as the actual number of firings.
-      \textbf{F:} Similarly, dyngen can also keep track of the regulatory activity of every interaction.
-      \textbf{G:} A benchmark of trajectory inference methods has already been performed using the cell state ground-truth.
-      \textbf{H:} The cell state ground-truth enables evaluating trajectory alignment methods \cite{saelens_comparisonsinglecelltrajectory_2019}.
-      \textbf{I:} The reaction propensity ground-truth enables evaluating RNA velocity methods.
-      \textbf{J:} The cellwise regulatory network ground-truth enables evaluating cell-specific gene regulatory network inference methods.
+To the best of our knowledge, the works listed above use very different approaches to benchmarking their tool:
+
+* LIONESS, network inference method: Uses randomly generated networks & boolean networks to generate synthetic data.
+The networks used to generate datasets are very synthetic and completely devoid of any biological relevance.
+By using Gillespie's SSA algorithm instead of boolean networks, we can use dyngen to track many layers of information (e.g.
+the number of times a biological reaction has been triggered during a given time interval).
+This allows us to benchmark methods which require this information (e.g.
+RNA velocity).
+* cellAlign, trajectory alignment: Perturbs real datasets to assess the robustness of cellAlign.
+When benchmarking a tool, assessing its robustness is almost as important as assessing its accuracy.
+Robustness analysis can be performed without knowing the gold standard, whereas the gold standard is needed for assessing the accuracy of a tool.
+* velocyto, RNA velocity: Uses ODEs to simulate splicing mechanics.
+The code for reproducing these datasets couldn't be found.
+Based on the description, these simulations were limited to simulating only the splicing mechanics under a particular set of conditions, thereby limiting its applicability outside of this particular context.
+* scvelo, RNA velocity: Same as velocyto.
+* dynamo, RNA velocity: Uses Gillespie's SSA to simulate splicing reactions.
+However, this system is very limited, including only two genes.
+This application of Gillespie's SSA seems very tailored towards this particular use-case and would require major redesigns to be made applicable elsewhere.
+
+It should be noted that from the list above, only LIONESS really compares its accuracy performance with other methods using a quantitative metric.
+We would thus argue that these cases further highlight the need for broadly applicable synthetic data generators such as dyngen.
+
+## Comment 5
+> The authors evaluate several computational tools, including the afore-mentioned.
+Even if the authors mention not to aim for a large-scale benchmarking, they still should provide some insights into why some tools perform better than others.
+Having an interpretable simulation engine in hand, it is unfortunate that it yet lacks any explanation and the reader is left alone with performance statements that are not well-founded.
+Some intuition would be very desirable.
+Why does SCENIC outperform the other methods?
+Why does smoothing improve trajectory alignment?
+And what causes differences in velocity estimation?
+Could you, for instance, generate an example where the differences in velocity estimates become more obvious, where one estimation procedure fails to fully capture the known direction?
+
+Performing in-depth benchmarks on each of these use-cases can be considered separate studies on their own and is beyond the scope of this work.
+However, we did reanalyse the experiments and added some interpretation to the respective sections.
+
+ * Trajectory alignment: We removed DTW+smoothing (as it was a custom method developed for the sake of this analysis) and added cellAlign to the comparison, as cellAlign also uses smoothing and DTW to get to the same objective, and has been developed specifically with single cell data in mind.
+ * RNA velocity: We removed some parameter settings which are designed to be very similar and thus have no added benefit in the visualisation.
+We removed the 'difficulty' settings for the different datasets, as the 'easy' and 'medium' datasets were unrealistic; only the 'hard' settings were retained.
+While there are no large difference in performance between the different methods, we added interpretation on why methods obtain low scores on one metric and high scores on the other.
+ * Cell-specific network inference: We added information on why certain methods significantly outperform others and provide possible solutions to problems, where possible.
+
+## Minor comment 1
+> Could the authors please discuss runtime and scalability, which obviously can easily become a weakness for such a complex model.
+
+A vignette has been added to the package and supplementary material explores the execution time of dyngen for an increasing number of cells and genes ([dyngen.dynverse.org/articles/advanced_topics/scalability_and_runtime.html](https://dyngen.dynverse.org/articles/advanced_topics/scalability_and_runtime.html)).
+These figures show that the execution time of dyngen scales linearly w.r.t.
+the number of cells and also linearly w.r.t.
+the number of genes.
+
+## Minor comment 2
+> Please state how many genes have been used in the examples.
+It seems like the number of genes used in the examples is rather small (< 100).
+Wouldn’t it be better to use a more realistic number of features (~10,000)?
+
+Unfortunately, the examples shown in the package documentation need to be very small to abide by CRAN package submission policies, as the complete testing of the software package (including but not limited to running all examples) can strictly take 10 minutes on CRAN's server using only one thread.
+
+However, the scalability vignette mentioned above also shows the execution time of each of the steps taken by dyngen in simulating a 10'000 by 10'000 dataset, which takes about 20 minutes to complete.
+
+## Minor comment 3
+> Congratulations for making the clearly documented package available.
+Installation and going through the vignette works fine.
+At several points during the simulation files are downloaded without asking permission.
+It may be better practice to ask the user for permission, or download into a temporary directory, or read the files directly without downloading, or include in the package etc.
+
+By default, required files are downloaded to a temporary directory.
+Unfortunately, these files cannot be included inside the package as they are reference datasets which are much larger than what is allowed by CRAN policies.
+
+Based on the reviewers suggestion, we modified the installation instructions such that it better reflects that files will be downloaded when using dyngen, and provides instructions on how to configure the system such that it will cache these files in a user-specified directory ([dyngen.dynverse.org/articles/installation.html](https://dyngen.dynverse.org/articles/installation.html)).
+
+## Minor comment 4
+> Further, it might be very useful to output the simulated data to a standard object, such as Seurat or SingleCellExperiment, or to provide instructions on how to convert it.
+
+Thank you for this suggestion.
+We agree that conversion functions from the typical dyngen output to the typical object types familiar to single-cell omics programmers would be very useful.
+Therefore, we included such functions in the latest version of dyngen to convert its output to the following formats: anndata, dyno, SingleCellExperiment, Seurat.
+
+[a]containing what?
+[b]todo
+
+# Supplementary Files
+
+## Vignette: Comparison to reference dataset
+
+In this vignette, we will take a look at characteristic features of
+dyngen versus the reference dataset it uses. To this end, we’ll be using
+[`countsimQC`](https://www.bioconductor.org/packages/release/bioc/html/countsimQC.html)
+[@soneson_unifiedqualityverification_2018] to calculate key statistics of both datasets
+and create comparative visualisations.
+
+### Run dyngen simulation
+
+We use an internal function from the dyngen package to download and
+cache one of the reference datasets.
+
+``` r
+library(tidyverse)
+library(dyngen)
+
+set.seed(1)
+
+data("realcounts", package = "dyngen")
+name_realcounts <- "zenodo_1443566_real_silver_bone-marrow-mesenchyme-erythrocyte-differentiation_mca"
+url_realcounts <- realcounts %>% filter(name == name_realcounts) %>% pull(url)
+realcount <- dyngen:::.download_cacheable_file(url_realcounts, getOption("dyngen_download_cache_dir"), verbose = FALSE)
+```
+
+We run a simple dyngen dataset as follows, where the number of cells and
+genes are determined by the size of the reference dataset.
+
+``` r
+backbone <- backbone_bifurcating_loop()
+
+num_cells <- nrow(realcount)
+num_feats <- ncol(realcount)
+num_tfs <- nrow(backbone$module_info)
+num_tar <- round((num_feats - num_tfs) / 2)
+num_hks <- num_feats - num_tfs - num_tar
+
+config <-
+  initialise_model(
+    backbone = backbone,
+    num_cells = num_cells,
+    num_tfs = num_tfs,
+    num_targets = num_tar,
+    num_hks = num_hks,
+    gold_standard_params = gold_standard_default(),
+    simulation_params = simulation_default(
+      total_time = 1000,
+      experiment_params = simulation_type_wild_type(num_simulations = 100)
+    ),
+    experiment_params = experiment_snapshot(
+      realcount = realcount
+    ),
+    verbose = FALSE
+  )
+```
+
+``` r
+
+# the simulation is being sped up because rendering all vignettes with one core
+
+# for pkgdown can otherwise take a very long time
+set.seed(1)
+
+config <-
+  initialise_model(
+    backbone = backbone,
+    num_cells = num_cells,
+    num_tfs = num_tfs,
+    num_targets = num_tar,
+    num_hks = num_hks,
+    verbose = interactive(),
+    download_cache_dir = tools::R_user_dir("dyngen", "data"),
+    simulation_params = simulation_default(
+      total_time = 1000,
+      census_interval = 2, 
+      ssa_algorithm = ssa_etl(tau = 300/3600),
+      experiment_params = simulation_type_wild_type(num_simulations = 10)
+    ),
+    experiment_params = experiment_snapshot(
+      realcount = realcount
+    )
+  )
+```
+
+``` r
+out <- generate_dataset(config, make_plots = TRUE)
+```
+
+    ## Generating TF network
+    ## Sampling feature network from real network
+    ## Generating kinetics for 3025 features
+    ## Generating formulae
+    ## Generating gold standard mod changes
+    ## Precompiling reactions for gold standard
+    ## Running gold simulations
+    ##   |                                                  | 0 % elapsed=00s     |========                                          | 14% elapsed=00s, remaining~01s  |===============                                   | 29% elapsed=00s, remaining~00s  |======================                            | 43% elapsed=00s, remaining~00s  |=============================                     | 57% elapsed=00s, remaining~00s  |====================================              | 71% elapsed=00s, remaining~00s  |===========================================       | 86% elapsed=00s, remaining~00s  |==================================================| 100% elapsed=01s, remaining~00s
+    ## Precompiling reactions for simulations
+    ## Running 10 simulations
+    ## Mapping simulations to gold standard
+    ## Performing dimred
+    ## Simulating experiment
+    ## Wrapping dataset
+    ## Making plots
+
+``` r
+out$plot
+```
+
+![](comparison_reference_files/figure-gfm/dyngen_generate-1.png)<!-- -->
+
+Both datasets are stored in a list for easy usage by countsimQC.
+
+``` r
+datasets <- list(
+  real = t(as.matrix(realcount)),
+  dyngen = t(as.matrix(out$dataset$counts))
+)
+
+ddsList <- lapply(datasets, function(ds) {
+  DESeq2::DESeqDataSetFromMatrix(
+    countData = round(as.matrix(ds)), 
+    colData = data.frame(sample = seq_len(ncol(ds))), 
+    design = ~1
+  )
+})
+```
+
+### Run countsimQC computations
+
+Below are some computations countsimQC makes. Normally these are not
+visible to the user, but for the sake of transparency these are included
+in the vignette.
+
+``` r
+library(countsimQC)
+
+### Define helper objects
+nDatasets <- length(ddsList)
+colRow <- c(2, 1)
+panelSize <- 4
+thm <- 
+  theme_bw() + 
+  theme(
+    axis.text = element_text(size = 15),
+    axis.title = element_text(size = 14),
+    strip.text = element_text(size = 15)
+  )
+```
+
+Compute key characteristics
+
+``` r
+obj <- countsimQC:::calculateDispersionsddsList(ddsList = ddsList, maxNForDisp = Inf)
+
+sampleCorrDF <- countsimQC:::calculateSampleCorrs(ddsList = obj, maxNForCorr = 500)
+
+featureCorrDF <- countsimQC:::calculateFeatureCorrs(ddsList = obj, maxNForCorr = 500)
+```
+
+Summarize sample characteristics
+
+``` r
+sampleDF <- map2_df(obj, names(obj), function(x, dataset_name) {
+  tibble(
+    dataset = dataset_name,
+    Libsize = colSums(x$dge$counts),
+    Fraczero = colMeans(x$dge$counts == 0),
+    TMM = x$dge$samples$norm.factors,
+    EffLibsize = Libsize * TMM
+  )
+})
+```
+
+Summarize feature characteristics
+
+``` r
+featureDF <- map2_df(obj, names(obj), function(x, dataset_name) {
+  rd <- SummarizedExperiment::rowData(x$dds)
+  tibble(
+    dataset = dataset_name,
+    Tagwise = sqrt(x$dge$tagwise.dispersion),
+    Common = sqrt(x$dge$common.dispersion),
+    Trend = sqrt(x$dge$trended.dispersion),
+    AveLogCPM = x$dge$AveLogCPM,
+    AveLogCPMDisp = x$dge$AveLogCPMDisp, 
+    average_log2_cpm = apply(edgeR::cpm(x$dge, prior.count = 2, log = TRUE), 1, mean), 
+    variance_log2_cpm = apply(edgeR::cpm(x$dge, prior.count = 2, log = TRUE), 1, var),
+    Fraczero = rowMeans(x$dge$counts == 0),
+    dispGeneEst = rd$dispGeneEst,
+    dispFit = rd$dispFit,
+    dispFinal = rd$dispersion,
+    baseMeanDisp = rd$baseMeanDisp,
+    baseMean = rd$baseMean
+  )
+})
+```
+
+Summarize data set characteristics
+
+``` r
+datasetDF <- map2_df(obj, names(obj), function(x, dataset_name) {
+  tibble(
+    dataset = dataset_name,
+    prior_df = paste0("prior.df = ", round(x$dge$prior.df, 2)),
+    nVars = nrow(x$dge$counts),
+    nSamples = ncol(x$dge$counts),
+    AveLogCPMDisp = 0.8 * max(featureDF$AveLogCPMDisp),
+    Tagwise = 0.9 * max(featureDF$Tagwise)
+  )
+})
+```
+
+### Data set dimensions
+
+These bar plots show the number of samples (columns) and features (rows)
+in each data set.
+
+Number of samples (columns)
+
+``` r
+ggplot(datasetDF, aes(x = dataset, y = nSamples, fill = dataset)) + 
+  geom_bar(stat = "identity", alpha = 0.5) + 
+  xlab("") + ylab("Number of samples (columns)") + 
+  thm + theme(axis.text.x = element_text(angle = 90, hjust = 1, vjust = 0.5))
+```
+
+![](comparison_reference_files/figure-gfm/nSamples-1.png)<!-- -->
+
+Number of features (rows)
+
+``` r
+ggplot(datasetDF, aes(x = dataset, y = nVars, fill = dataset)) + 
+  geom_bar(stat = "identity", alpha = 0.5) + 
+  xlab("") + ylab("Number of features (rows)") + 
+  thm + theme(axis.text.x = element_text(angle = 90, hjust = 1, vjust = 0.5))
+```
+
+![](comparison_reference_files/figure-gfm/nVariables-1.png)<!-- -->
+
+### Dispersion/BCV plots
+
+Disperson/BCV plots show the association between the average abundance
+and the dispersion or “biological coefficient of variation”
+(sqrt(dispersion)), as calculated by
+[`edgeR`](https://bioconductor.org/packages/release/bioc/html/edgeR.html)
+[@robinson_edgerbioconductorpackage_2010] and
+[`DESeq2`](http://bioconductor.org/packages/release/bioc/html/DESeq2.html)
+[@love_moderatedestimationfold_2014]. In the `edgeR` plot, the estimate of the
+prior degrees of freedom is indicated.
+
+#### edgeR
+
+The black dots represent the tagwise dispersion estimates, the red line
+the common dispersion and the blue curve represents the trended
+dispersion estimates. For further information about the dispersion
+estimation in `edgeR`, see Chen et al. [@chen_differentialexpressionanalysis_2014].
+
+``` r
+ggplot(featureDF %>% dplyr::arrange(AveLogCPMDisp), 
+       aes(x = AveLogCPMDisp, y = Tagwise)) + 
+  geom_point(size = 0.25, alpha = 0.5) + 
+  facet_wrap(~dataset, nrow = colRow[2]) + 
+  geom_line(aes(y = Trend), color = "blue", size = 1.5) + 
+  geom_line(aes(y = Common), color = "red", size = 1.5) +
+  geom_text(data = datasetDF, aes(label = prior_df)) + 
+  xlab("Average log CPM") + ylab("Biological coefficient of variation") + 
+  thm
+```
+
+![](comparison_reference_files/figure-gfm/BCVedgeR-1.png)<!-- -->
+
+#### DESeq2
+
+The black dots are the gene-wise dispersion estimates, the red curve the
+fitted mean-dispersion relationship and the blue circles represent the
+final dispersion estimates.For further information about the dispersion
+estimation in `DESeq2`, see Love et al. [@love_moderatedestimationfold_2014].
+
+``` r
+ggplot(featureDF %>% dplyr::arrange(baseMeanDisp), 
+       aes(x = baseMeanDisp, y = dispGeneEst)) + 
+  geom_point(size = 0.25, alpha = 0.5) + 
+  facet_wrap(~dataset, nrow = colRow[2]) + scale_x_log10() + scale_y_log10() +  
+  geom_point(aes(y = dispFinal), color = "lightblue", shape = 21) + 
+  geom_line(aes(y = dispFit), color = "red", size = 1.5) + 
+  xlab("Base mean") + ylab("Dispersion") + 
+  thm
+```
+
+![](comparison_reference_files/figure-gfm/dispersionDESeq2-1.png)<!-- -->
+
+### Mean-variance plots
+
+This scatter plot shows the relation between the empirical mean and
+variance of the features. The difference between these mean-variance
+plots and the mean-dispersion plots above is that the plots in this
+section do not take the information about the experimental design and
+sample grouping into account, but simply display the mean and variance
+of log2(CPM) estimates across all samples, calculated using the `cpm`
+function from
+[`edgeR`](https://bioconductor.org/packages/release/bioc/html/edgeR.html)
+[@robinson_edgerbioconductorpackage_2010], with a prior count of 2.
+
+``` r
+ggplot(featureDF, aes(x = average_log2_cpm, y = variance_log2_cpm)) + 
+  geom_point(size = 0.75, alpha = 0.5) + 
+  facet_wrap(~dataset, nrow = colRow[2]) + 
+  xlab("Mean of log2(CPM)") + ylab("Variance of log2(CPM)") + 
+  thm
+```
+
+![](comparison_reference_files/figure-gfm/meanVarSepScatter-1.png)<!-- -->
+
+### Library sizes
+
+This plot shows a histogram of the total read count per sample, i.e.,
+the column sums of the respective data matrices.
+
+``` r
+ggplot(sampleDF, aes(x = Libsize)) + geom_histogram(bins = 30) + 
+  facet_wrap(~dataset, nrow = colRow[2]) +
+  xlab("Library size") + thm
+```
+
+![](comparison_reference_files/figure-gfm/libsizeSepHist-1.png)<!-- -->
+
+### TMM normalization factors
+
+This plot shows a histogram of the TMM normalization factors [@robinson_scalingnormalizationmethod_2010], intended to adjust for differences in RNA
+composition, as calculated by
+[`edgeR`](https://bioconductor.org/packages/release/bioc/html/edgeR.html)
+[@robinson_edgerbioconductorpackage_2010].
+
+``` r
+ggplot(sampleDF, aes(x = TMM)) + geom_histogram(bins = 30) + 
+  facet_wrap(~dataset, nrow = colRow[2]) +
+  xlab("TMM normalization factor") + thm
+```
+
+![](comparison_reference_files/figure-gfm/tmmSepHist-1.png)<!-- -->
+
+### Effective library sizes
+
+This plot shows a histogram of the “effective library sizes,” defined as
+the total count per sample multiplied by the corresponding TMM
+normalization factor.
+
+``` r
+ggplot(sampleDF, aes(x = EffLibsize)) + geom_histogram(bins = 30) + 
+  facet_wrap(~dataset, nrow = colRow[2]) +
+  xlab("Effective library size") + thm
+```
+
+![](comparison_reference_files/figure-gfm/effLibsizeSepHist-1.png)<!-- -->
+
+### Expression distributions (average log CPM)
+
+This plot shows the distribution of average abundance values for the
+features. The abundances are log CPM values calculated by `edgeR`.
+
+``` r
+ggplot(featureDF, aes(x = AveLogCPM)) + geom_histogram(bins = 30) + 
+  facet_wrap(~dataset, nrow = colRow[2]) +
+  xlab("Average log CPM") + thm
+```
+
+![](comparison_reference_files/figure-gfm/logCPMSepHist-1.png)<!-- -->
+
+### Fraction zeros per sample
+
+This plot shows the distribution of the fraction of zeros observed per
+sample (column) in the count matrices.
+
+``` r
+ggplot(sampleDF, aes(x = Fraczero)) + geom_histogram(bins = 30) + 
+  facet_wrap(~dataset, nrow = colRow[2]) +
+  xlab("Fraction zeros per sample") + thm
+```
+
+![](comparison_reference_files/figure-gfm/fraczeroSampleSepHist-1.png)<!-- -->
+
+### Fraction zeros per feature
+
+This plot illustrates the distribution of the fraction of zeros observed
+per feature (row) in the count matrices.
+
+``` r
+ggplot(featureDF, aes(x = Fraczero)) + geom_histogram(bins = 30) + 
+  facet_wrap(~dataset, nrow = colRow[2]) +
+  xlab("Fraction zeros per feature") + thm
+```
+
+![](comparison_reference_files/figure-gfm/fraczeroFeatureSepHist-1.png)<!-- -->
+
+### Sample-sample correlations
+
+The plot below shows the distribution of Spearman correlation
+coefficients for pairs of samples, calculated from the log(CPM) values
+obtained via the `cpm` function from `edgeR`, with a prior.count of 2.
+
+``` r
+ggplot(sampleCorrDF, aes(x = Correlation)) + geom_histogram(bins = 30) + 
+  facet_wrap(~dataset, nrow = colRow[2]) +
+  xlab("Sample-sample correlation") + thm
+```
+
+![](comparison_reference_files/figure-gfm/sampleCorrSepHist-1.png)<!-- -->
+
+### Feature-feature correlations
+
+This plot illustrates the distribution of Spearman correlation
+coefficients for pairs of features, calculated from the log(CPM) values
+obtained via the `cpm` function from `edgeR`, with a prior.count of 2.
+Only non-constant features are considered.
+
+``` r
+ggplot(featureCorrDF, aes(x = Correlation)) + geom_histogram(bins = 30) + 
+  facet_wrap(~dataset, nrow = colRow[2]) +
+  xlab("Feature-feature correlation") + thm
+```
+
+![](comparison_reference_files/figure-gfm/featureCorrSepHist-1.png)<!-- -->
+
+### Library size vs fraction zeros
+
+This scatter plot shows the association between the total count (column
+sums) and the fraction of zeros observed per sample.
+
+``` r
+ggplot(sampleDF, aes(x = Libsize, y = Fraczero)) + 
+  geom_point(size = 1, alpha = 0.5) + 
+  facet_wrap(~dataset, nrow = colRow[2]) + 
+  xlab("Library size") + ylab("Fraction zeros") + thm
+```
+
+![](comparison_reference_files/figure-gfm/libsizeFraczeroSepScatter-1.png)<!-- -->
+
+### Mean expression vs fraction zeros
+
+This scatter plot shows the association between the average abundance
+and the fraction of zeros observed per feature. The abundance is defined
+as the log(CPM) values as calculated by `edgeR`.
+
+``` r
+ggplot(featureDF, aes(x = AveLogCPM, y = Fraczero)) + 
+  geom_point(size = 0.75, alpha = 0.5) + 
+  facet_wrap(~dataset, nrow = colRow[2]) + 
+  xlab("Average log CPM") + ylab("Fraction zeros") + thm
+```
+
+![](comparison_reference_files/figure-gfm/logCPMFraczeroSepScatter-1.png)<!-- -->
+
+## Vignette: On scalability and runtime
+
+In this vignette, we will take a look at the runtime of dyngen as the
+number of genes and the number of cells sampled is increased. We’ll be
+using the bifurcating cycle backbone which is well known for its
+beautiful 3D butterfly shape!
+
+``` r
+library(dyngen)
+library(tidyverse)
+
+set.seed(1)
+
+save_dir <- "scalability_and_runtime_runs"
+if (!dir.exists(save_dir)) dir.create(save_dir, recursive = TRUE)
+
+backbone <- backbone_bifurcating_cycle()
+```
+
+### Initial run
+
+We’ll be running this simulation a few times, with different values for
+`num_cells` and `num_features` to assess the scalability of dyngen. An
+example of a resulting dyngen model is shown here.
+
+``` r
+num_cells <- 100
+num_features <- 100
+num_tfs <- nrow(backbone$module_info)
+num_targets <- round((num_features - num_tfs) / 2)
+num_hks <- num_features - num_targets - num_tfs
+
+out <- 
+  initialise_model(
+    backbone = backbone,
+    num_tfs = num_tfs,
+    num_targets = num_targets,
+    num_hks = num_hks,
+    num_cells = num_cells,
+    gold_standard_params = gold_standard_default(
+      census_interval = 1,
+      tau = 100/3600
+    ),
+    simulation_params = simulation_default(
+      census_interval = 10,
+      ssa_algorithm = ssa_etl(tau = 300/3600),
+      experiment_params = simulation_type_wild_type(
+        num_simulations = num_cells / 10
+      )
+    ),
+    verbose = FALSE
+  ) %>% 
+  generate_dataset(make_plots = TRUE)
+```
+
+``` r
+out$plot
+```
+
+![](scalability_and_runtime_files/figure-gfm/example-1.png)<!-- -->
+
+We tweaked some of the parameters by running this particular backbone
+once with `num_cells = 100` and `num_features = 100` and verifying that
+the new parameters still yield the desired outcome. The parameters we
+tweaked are:
+
+-   On average, 10 cells are sampled per simulation
+    (e.g. `num_simulations = 100` and `num_cells = 1000`). You could
+    increase this ratio to get a better cell count yield from a given
+    set of simulations, but cells from the same simulation that are
+    temporally close will have highly correlated expression profiles.
+-   Increased time steps `tau`. This will make the Gillespie algorithm
+    slightly faster but might result in unexpected artifacts in the
+    simulated data.
+-   `census_interval` increased from 4 to 10. This will cause dyngen to
+    store an expression profile only every 10 time units. Since the
+    total simulation time is xxx, each simulation will result in yyy
+    data points. Note that on average only 10 data points are sampled
+    per simulation.
+
+For more information on parameter tuning, see the vignette ‘Advanced:
+tuning the simulation parameters’.
+
+### Timing experiments
+
+The simulations are run once with a large `num_features` and
+`num_cells`, a few times with varying `num_cells` and then once more
+with varying `num_features`. Every run is repeated three times in order
+to get a bit more stable time measurements. Since some of the
+simulations can take over 10 minutes, the timings results of the
+simulations are cached in the ‘scalability\_and\_runtime\_runs’
+folder.\`
+
+``` r
+settings <- bind_rows(
+  tibble(num_cells = 10000, num_features = 10000, rep = 1), #, rep = seq_len(3)),
+  crossing(
+    num_cells = seq(1000, 10000, by = 1000),
+    num_features = 100,
+    rep = seq_len(3)
+  ),
+  crossing(
+    num_cells = 100,
+    num_features = seq(1000, 10000, by = 1000),
+    rep = seq_len(3)
+  )
+) %>% 
+  mutate(filename = paste0(save_dir, "/cells", num_cells, "_feats", num_features, "_rep", rep, ".rds"))
+
+timings <- pmap_dfr(settings, function(num_cells, num_features, rep, filename) {
+  if (!file.exists(filename)) {
+    set.seed(rep)
+
+    cat("Running num_cells: ", num_cells, ", num_features: ", num_features, ", rep: ", rep, "\n", sep = "")
+    num_tfs <- nrow(backbone$module_info)
+    num_targets <- round((num_features - num_tfs) / 2)
+    num_hks <- num_features - num_targets - num_tfs
+
+    out <- 
+      initialise_model(
+        backbone = backbone,
+        num_tfs = num_tfs,
+        num_targets = num_targets,
+        num_hks = num_hks,
+        num_cells = num_cells,
+        gold_standard_params = gold_standard_default(
+          census_interval = 1,
+          tau = 100/3600
+        ),
+        simulation_params = simulation_default(
+          census_interval = 10,
+          ssa_algorithm = ssa_etl(tau = 300/3600),
+          experiment_params = simulation_type_wild_type(
+            num_simulations = num_cells / 10
+          )
+        ),
+        verbose = FALSE
+      ) %>% 
+      generate_dataset()
+
+    tim <- 
+      get_timings(out$model) %>% 
+      mutate(rep, num_cells, num_features)
+
+    write_rds(tim, filename, compress = "gz")
   }
-    \label{fig:overview}
-\end{figure}
 
-<!-- RESULTS -->
-We demonstrate dyngen's broad applicability by evaluating three novel types of computational approaches for which no simulation engines exist yet: cell-specific network inference, trajectory alignment and RNA velocity (Figure\ \ref{fig:applications}).
-We emphasize that our main aim here is to illustrate the potential of dyngen for these evaluations, rather than performing large-scale benchmarking, which would require assessing many more quantitative and qualitative aspects of each method [@weber_essentialguidelinescomputational_2019].
+  read_rds(filename)
+})
 
-\begin{figure}[t!]
-    \centering
-    \includegraphics[width=\linewidth]{result_files/summary/figure_2.pdf}
-    \caption{
-      \textbf{dyngen provides ground-truth data for a variety of applications (left), which can be used to quantitatively evaluate methods (right).}
-      \textbf{A:} Trajectory alignment aligns two trajectories between samples.
-We use dyngen to evaluate dynamic time warping (DTW) and cellAlign when aligning two linear trajectories with different kinetic parameters.
-      \textbf{B:} RNA velocity calculates for each cell the direction in which the expression of each gene is moving.
-We evaluated scVelo and velocyto by comparing these vectors with the known velocity vector (velocity correlation) and with the known direction of the cellular trajectory in a dimensionality reduction (velocity arrow cosine).
-      \textbf{C:} Cell-specific network inference (CSNI) predicts the regulatory network of every individual cell.
-We evaluate each cell-specific regulatory network with typical metrics for network inference: the Area Under the Receiver Operating Characteristics-curve (AUROC) and Area Under the Precision Recall-curve (AUPR).
-We evaluate three CSNI methods by computing the mean AUROC and AUPR across all cells.
-    }
-    \label{fig:applications}
-\end{figure}
+timings_gr <- 
+  timings %>% 
+  group_by(group, task, num_cells, num_features) %>% 
+  summarise(time_elapsed = mean(time_elapsed), .groups = "drop")
 
-**Trajectory alignment** methods align trajectories from different samples and allow studying the differences between the different trajectories.
-For example, by comparing the transcriptomic profiles of cells from a diseased patient to a healthy control, it might be possible to detect transcriptomics differences (differential expression) of particular cells along a developmental process, or to detect an early stop of the trajectory of the diseased patient.
-Currently, trajectory alignment is limited to aligning linear trajectories, though other topologies of trajectory could be aligned as well.
-Dynamic Time Warping (DTW) [@giorgino_computingvisualizingdynamic_2009] is a method designed for aligning temporal sequences for speech recognition but has since been used to compare gene expression kinetics from many different biological processes [@cacchiarelli_aligningsinglecelldevelopmental_2018; @kanton_organoidsinglecellgenomic_2019; @mcfaline-figueroa_pooledsinglecellgenetic_2019; @alpert_alignmentsinglecelltrajectories_2018].
-We evaluate the performance of DTW by simulating two linear trajectories with slightly different simulation kinetics and assessing the accuracy of the alignment by DTW.
-We compare the performance of DTW against a version of DTW where gene expression is first smoothed, for varying degrees of noise added to the gene expression (Figure\ \ref{fig:traj_align}).
-We observe that DTW+smoother performs significantly better than the unsmoothed version across all levels of noise.
-Note that, in this comparison, only linear trajectory alignment is performed.
-While dyngen can generate non-linear trajectories (e.g.
-cyclic or branching), constructing a quantitative accuracy metric for non-linear trajectory alignment is not trivial and an avenue for future work.[a]
+timings_sum <-
+  timings %>% 
+  group_by(num_cells, num_features, rep) %>%
+  summarise(time_elapsed = sum(time_elapsed), .groups = "drop")
+```
 
-**RNA velocity** methods use the relative ratio between pre-mRNA and mature mRNA reads to predict the rate of increase/decrease of RNA molecule abundance, as this can be used to predict the directionality of single cell differentiation in trajectories [@zeisel_coupledpremrnamrna_2011;@lamanno_rnavelocitysingle_2018].
-Already two algorithms are currently available for estimating the RNA velocity vector from spliced and unspliced counts: velocyto [@lamanno_rnavelocitysingle_2018] and scvelo [@bergen_generalizingrnavelocity_2020].
-Yet, to date, no quantitative assessment of their accuracy has been performed, mainly due to the difficulty in obtaining real ground-truth data to do so.
-In contrast, the ground-truth RNA velocity can be easily extracted from a dyngen simulation, as it is possible to store the rate at which mRNA molecules are being transcribed and degraded at any particular point in time.
-We executed velocyto and scvelo (with 2 different parameter settings, stochastic and dynamical) on 42 datasets with a variety of backbones (including linear, bifurcating, cyclic, disconnected).
-We evaluated the predictions using two metrics (Figure\ \ref{fig:velocity}), one which directly compares the predicted RNA velocity of each gene with the ground-truth RNA velocity (called the "velocity correlation"), and one which compares the direction of the ground-truth trajectory embedded in a dimensionality reduction with the average RNA velocity of cells in that neighbourhood (called the "velocity arrow cosine").
-While both velocyto and scvelo obtained high scores for the velocity arrow cosine metric (overall 25th percentile = 0.606), the velocity correlation is rather low (overall 75th percentile = 0.156).
-This means that predicting the RNA velocity (i.e.
-transcription rate minus the decay rate) for any particular gene is challenging, but the combined information is very informative in determining the directionality of cell progression in the trajectory.
-In terms of velocity correlation, no method performed significantly better than the other, whereas scvelo stochastic performed slightly worse than scvelo dynamical and velocyto in terms of velocity arrow cosine score.
+### Simulate a large dataset (10k × 10k)
 
-**Cell-specific network inference** (CSNI) methods predict not only which transcription factors regulate which target genes, but also aim to identify how active each interaction is in each of the cells, since interactions can be turned off and on depending on the cellular state.
-While a few pioneering CSNI approaches have already been developed [@aibar_scenicsinglecellregulatory_2017; @kuijjer_estimatingsamplespecificregulatory_2019; @liu_personalizedcharacterizationdiseases_2016], a quantitative assessment of their performance is until now lacking.
-This is not surprising, as neither real nor in silico datasets of cell-specific or even cell-type-specific interactions exist that are large enough so that it can be used as a ground-truth for evaluating CSNI methods.
-Extracting the ground-truth dynamic network in dyngen is straightforward though, given that we can calculate how target gene expression would change without the regulator being present.
-We used this ground-truth to compare the performance of three CSNI methods (Figure\ \ref{fig:scgrn}): LIONESS [@kuijjer_estimatingsamplespecificregulatory_2019], SSN [@liu_personalizedcharacterizationdiseases_2016] and SCENIC [@aibar_scenicsinglecellregulatory_2017].
-For each dataset, we computed the mean AUROC and AUPR scores of the individual cells.
-Comparing the mean AUROC and AUPR showed that pySCENIC significantly outperforms both LIONESS and SSN, and in turn that LIONESS significantly outperforms SSN.
-The poor performance of SSN is expected, as its methodology for predicting a cell-specific is simply computing the difference in Pearson correlation values applied to the whole dataset and the whole dataset minus one sample.
-This strategy performs poorly in large datasets where cell correlations are high, as the removal of one cell will not yield large differences in correlation values and will result in mostly noise.
-Overall, pySCENIC almost always performs better than LIONESS, except for a few datasets where LIONESS does manage to obtain a higher AUROC score.
-However, by using a different internal network inference (e.g.
-GENIE3 [@huynh-thu_inferringregulatorynetworks_2010] or pySCENIC's GRNBoost2 [@moerman_grnboost2arboretoefficient_2019]) could significantly increase the performance obtained by LIONESS.
+Below is shown the timings of each of the steps in simulating a dyngen
+dataset containing 10’000 genes and 10’000 features. The total
+simulation time required is 1147 seconds, most of which is spent
+performing the simulations itself.
 
-<!-- DISCUSSION -->
-In summary, dyngen's single-cell simulations can be used to evaluate common single-cell omics computational methods such as clustering, batch correction, trajectory inference, and network inference.
-However, the framework is flexible enough to be adaptable to a broad range of applications, including methods that integrate clustering, network inference, and trajectory inference.
-In this respect, dyngen may promote the development of new tools in the single-cell field similarly as other simulators have done in the past [@schaffter_genenetweaversilicobenchmark_2011; @ewing_combiningtumorgenome_2015].
-Additionally, one could anticipate technological developments in single-cell multi-omics.
-In this way, dyngen allows designing and evaluating the performance and robustness of new types of computational analyses before experimental data becomes available, comparing which experimental protocol is the most cost-effective in producing qualitative and robust results in downstream analysis.
-One major assumption of dyngen is that cells are assumed to be well-mixed and independent from each other.
-Subdividing a cell into multiple 2D or 3D subvolumes or allowing cells to exchange molecules, respectively, could pave the way to better study key cellular processes such as cell division, intercellular communication, and migration [@smith_spatialstochasticintracellular_2019].
+``` r
+timings0 <- 
+  timings_gr %>% 
+  filter(num_cells == 10000, num_features == 10000) %>% 
+  mutate(name = forcats::fct_rev(forcats::fct_inorder(paste0(group, ": ", task))))
 
-# Data Availability[b]
+ggplot(timings0) + 
+  geom_bar(aes(x = name, y = time_elapsed, fill = group), stat = "identity") +
+  scale_fill_brewer(palette = "Dark2") + 
+  theme_classic() +
+  theme(legend.position = "none") +
+  coord_flip() + 
+  labs(x = NULL, y = "Time (s)", fill = "dyngen stage")
+```
 
-# Code Availability
-dyngen is available as an open-source R package on CRAN at [cran.r-project.org/package=dyngen](https://cran.r-project.org/package=dyngen).
-The analyses performed in this manuscript are available on GitHub at [github.com/dynverse/dyngen_manuscript](https://github.com/dynverse/dyngen_manuscript).
+![](scalability_and_runtime_files/figure-gfm/bblego-1.png)<!-- -->
 
-# Author contributions
-* W.S. and R.C. designed the study.
-* R.C., W.S., and L.D. performed the experiments and analysed the data.
-* R.C. and W.S. implemented the dyngen software package.
-* R.C., W.S., L.D., and Y.S. wrote the manuscript.
-* Y.S. supervised the project.
+### Increasing the number of cells
 
-\newpage
+By increasing the number of cells from 1000 to 10’000 whilst keeping the
+number of features fixed, we can get an idea of how the simulation time
+scales w.r.t. the number of cells.
 
-# Methods {#sec:dyngen-methods}
-The workflow to generate _in silico_ single-cell data consists of six main steps (Figure\ \ref{fig:explain_methods}).
+``` r
+timings1 <- 
+  timings_gr %>% 
+  filter(num_features == 100) %>% 
+  group_by(num_cells, num_features, group) %>%
+  summarise(time_elapsed = sum(time_elapsed), .groups = "drop")
 
-## Defining the module network {#sec:dyngen-modules}
+ggplot(timings1) + 
+  geom_bar(aes(x = forcats::fct_inorder(as.character(num_cells)), y = time_elapsed, fill = forcats::fct_inorder(group)), stat = "identity") +
+  theme_classic() +
+  scale_fill_brewer(palette = "Dark2") +
+  labs(x = "Number of cells", y = "Average time (s)", fill = "dyngen step")
+```
 
-One of the main processes involved in cellular dynamic processes is gene regulation, where regulatory cascades and feedback loops lead to progressive changes in expression and decision making.
-The exact way a cell chooses a certain path during its differentiation is still an active research field, although certain models have already emerged and been tested _in vivo_.
-One driver of bifurcation is mutual antagonism, where two genes strongly repress each other [@rekhtman_directinteractionhematopoietic_1999; @xu_regulationbifurcatingcell_2015], forcing one of the two to become inactive [@graf_forcingcellschange_2009].
-Such mutual antagonism can be modelled and simulated [@wang_quantifyingwaddingtonlandscape_2011; @ferrell_bistabilitybifurcationswaddington_2012].
-Although the two-gene model is simple and elegant, the reality is frequently more complex, with multiple genes (grouped into modules) repressing each other [@yosef_dynamicregulatorynetwork_2013].
+![](scalability_and_runtime_files/figure-gfm/figure1-1.png)<!-- -->
 
-To start a dyngen simulation, the user needs to define a module network.
-The module network describes how sets of genes regulate each other and is what mainly determines which dynamic processes occur within the simulated cells.
+It seems the execution time scales linearly w.r.t. the number of cells.
+This makes sense, because as the number of cells are increased, so do we
+increase the number of simulations made (which is not necessarily
+mandatory). Since the simulations are independent of each other and take
+up the most time, the execution time will scale linearly.
 
-A module network consists of modules connected together by regulatory interactions, which can be either up- or down-regulating.
-A module may have basal expression, which means genes in this module will be transcribed without the presence of transcription factor molecules.
-A module marked as "active during the burn phase" means that this module will be allowed to generate expression of its genes during an initial warm-up phase (See section \ref{sec:dyngen-simcell}).
-At the end of the dyngen process, cells will not be sampled from the burn phase simulations.
-Interactions between modules have a strength (which is a positive integer) and an effect (+1 for upregulating, -1 for downregulating).
+``` r
+ggplot(timings_sum %>% filter(num_features == 100)) + 
+  theme_bw() +
+  geom_point(aes(num_cells, time_elapsed)) +
+  scale_x_continuous(limits = c(0, 10000)) +
+  scale_y_continuous(limits = c(0, 300)) +
+  geom_abline(intercept = 22.097, slope = 0.0252) +
+  labs(x = "Number of cells", y = "Execution time (s)")
+```
 
-Several examples of module networks are given in Figure\ \ref{fig:example_backbones_onlymodules}.
-A simple chain of modules (where one module upregulates the next) results in a _linear_ process.
-By having the last module repress the first module, the process becomes _cyclic_.
-Two modules repressing each other is the basis of a _bifurcating_ process, though several chains of modules have to be attached in order to achieve progression before and after the bifurcation process.
-Finally, a _converging_ process has a bifurcation occurring during the burn phase, after which any differences in module regulation is removed.
+![](scalability_and_runtime_files/figure-gfm/plot_timings_cell-1.png)<!-- -->
 
-Note that these examples represent the bare minimum in terms of the number of modules used.
-Using longer chains of modules is typically desired.
-In addition, the fate decisions made in this example of a bifurcation is reversible, meaning cells can be reprogrammed to go down a different differentiation path.
-If this effect is undesirable, more safeguards need to be put in place to prevent reprogramming from occurring.
+### Increasing the number of features
 
-## Generating the gene regulatory network {#sec:dyngen-grn}
-The GRN is generated based on the given module network in four main steps (Figure\ \ref{fig:gen_feature_network}).
+By increasing the number of features from 1000 to 10’000 whilst keeping
+the number of cells fixed, we can get an idea of how the simulation time
+scales w.r.t. the number of features
 
-**Step 1, sampling the transcription factors (TF).** The TFs are the main drivers of the molecular changes in the simulation.
-The user provides a backbone and the number of TFs to generate.
-Each TF is assigned to a module such that each module has at least $x$ parameters (default $x=1$).
-A TF inherits the 'burn' and 'basal expression' from the module it belongs to.
+``` r
+timings2 <- 
+  timings_gr %>% 
+  filter(num_cells == 100) %>% 
+  group_by(num_cells, num_features, group) %>% 
+  summarise(time_elapsed = sum(time_elapsed), .groups = "drop")
 
-**Step 2, generating the TF interactions.** Let each TF be regulated according to the interactions in the backbone.
-These interactions inherit the effect, strength, and independence parameters from the interactions in the backbone.
-A TF can only be regulated by other TFs or itself.
+ggplot(timings2) + 
+  geom_bar(aes(x = forcats::fct_inorder(as.character(num_features)), y = time_elapsed, fill = forcats::fct_inorder(group)), stat = "identity") +
+  theme_classic() +
+  scale_fill_brewer(palette = "Dark2") +
+  labs(x = "Number of features", y = "Average time (s)", fill = "dyngen step")
+```
 
-**Step 3, sampling the target subnetwork.**
-A user-defined number of target genes are added to the GRN.
-Target genes are regulated by a TF or another target gene, but are always downstream of at least one TF.
-To sample the interactions between target genes, one of the many FANTOM5 \cite{lizio_gatewaysfantom5promoter_2015} GRNs is sampled.
-The currently existing TFs are mapped to regulators in the FANTOM5 GRN.
-The targets are drawn from the FANTOM5 GRN weighted by their page rank value, to create an induced GRN.
-For each target, at most $x$ regulators are sampled from the induced FANTOM5 GRN (default $x=5$).
-The interactions connecting a target gene and its regulators are added to the GRN.
+![](scalability_and_runtime_files/figure-gfm/figure2-1.png)<!-- -->
 
-**Step 4, sampling the housekeeping subnetwork.**
-Housekeeping genes are completely separate from any TFs or target genes.
-A user-defined set of housekeeping genes is also sampled from the FANTOM5 GRN.
-The interactions of the FANTOM5 GRN are first subsampled such that the maximum in-degree of each gene is $x$ (default $x=5$).
-A random gene is sampled and a breadth-first-search is performed to sample the desired number of housekeeping genes.
+It seems the execution time also scales linearly w.r.t. the number of
+features. As more genes are added to the underlying gene regulatory
+network, the density of the graph doesn’t change, so it makes sense that
+the execution time also scales linearly w.r.t. the number of features.
 
-## Convert gene regulatory network to a set of reactions {#sec:dyngen-reactions}
+``` r
+ggplot(timings_sum %>% filter(num_cells == 100)) + 
+  theme_bw() +
+  geom_point(aes(num_features, time_elapsed)) +
+  scale_x_continuous(limits = c(0, 10000)) +
+  scale_y_continuous(limits = c(0, 850)) +
+  geom_abline(intercept = 0.5481, slope = 0.07988) +
+  labs(x = "Number of features", y = "Execution time (s)")
+```
 
-\newcommand{\x}[1]{\text{x}_{#1}}
-\newcommand{\y}[1]{\text{y}_{#1}}
-\newcommand{\z}[1]{\text{z}_{#1}}
+![](scalability_and_runtime_files/figure-gfm/plot_timings_feats-1.png)<!-- -->
 
-\newcommand{\rs}[1]{\text{R}_{#1}}
-\newcommand{\rp}[1]{\text{R}^+_{#1}}
-\newcommand{\rn}[1]{\text{R}^-_{#1}}
+### Execution platform
 
-\newcommand{\xpr}[1]{\text{xpr}_{#1}}
-\newcommand{\xhl}[1]{\text{xhl}_{#1}}
-\newcommand{\ysr}[1]{\text{ysr}_{#1}}
-\newcommand{\yhl}[1]{\text{yhl}_{#1}}
-\newcommand{\ydr}[1]{\text{ydr}_{#1}}
-\newcommand{\zpr}[1]{\text{zpr}_{#1}}
-\newcommand{\zhl}[1]{\text{zhl}_{#1}}
-\newcommand{\zdr}[1]{\text{zdr}_{#1}}
-
-\newcommand{\str}[1]{\text{str}_{#1}}
-\newcommand{\hill}[1]{\text{hill}_{#1}}
-\newcommand{\ind}[1]{\text{ind}_{#1}}
-\newcommand{\dis}[1]{\text{dis}_{#1}}
-\newcommand{\buf}[1]{\text{bind}_{#1}}
-\newcommand{\ba}[1]{\text{bas}_{#1}}
-
-Simulating a cell's GRN makes use of a stochastic framework which tracks the abundance levels of molecules over time in a discrete quantity.
-For every gene $G$, the abundance levels of three molecules are tracked, namely of corresponding pre-mRNAs, mature mRNAs and proteins, which are represented by the terms $\x G$, $\y G$ and $\z G$ respectively.
-The GRN defines how a reaction affects the abundance levels of molecules and how likely it will occur.
-Gibson and Bruck [@gibson_probabilisticmodelprokaryotic_2000] provide a good introduction to modelling gene regulation with stochastic frameworks, on which many of the concepts below are based.
-
-For every gene in the GRN a set of reactions are defined, namely transcription, splicing, translation, and degradation.
-Each reaction consists of a propensity function -- a formula $f(.)$ to calculate the probability $f(.) \times \text{d}t$ of it occurring during a time interval $\text{d}t$ -- and the effect -- how it will affect the current state if triggered.
-
-The effects of each reaction mimic the respective biological processes (Table\ \ref{tab:reaction_def}, middle).
-Transcription of gene $G$ results in the creation of a single pre-mRNA molecule $\x G$.
-Splicing turns one pre-mRNA $\x G$ into a mature mRNA $\x G$.
-Translation uses a mature mRNA $\y G$ to produce a protein $\z G$.
-Pre-mRNA, mRNA and protein degradation results in the removal of a $\x G$, $\y G$, and $\z G$ molecule, respectively.
-
-The propensity of all reactions except transcription are all linear functions (Table\ \ref{tab:reaction_def}, right) of the abundance level of some molecule multiplied by a per-gene constant (Table\ \ref{tab:reaction_params}).
-The propensity of transcription of a gene $G$ depends on the abundance levels of its TFs.
-The per-gene and per-interaction constants are based on the median reported production-rates and half-lives of molecules measured of 5000 mammalian genes [@schwanhausser_globalquantificationmammalian_2011], except that the transcription rate has been amplified by a factor of 10.
-
-\newcommand{\proptran}{f}
-\newcommand{\ai}[2]{$S_{#1} = S_{#2b}$}
-\newcommand{\zk}[1]{\frac{y_#1}{k_#1}^{c_#1}}
-\newcommand{\wi}[1]{\nu_#1}
-
-The propensity of the transcription of a gene $G$ is inspired by thermodynamic models of gene regulation [@schilstra_biologicgeneexpression_2008], in which the promoter of $G$ can be bound or unbound by a set of $N$ transcription factors $H_i$.
-Let $\proptran(\z 1, \z 2, \ldots, \z N)$ denote the propensity function of $G$, in function of the abundance levels of the transcription factors.
-The following subsections explain and define the propensity function when $N=1$, $N=2$, and finally for an arbitrary $N$.
-
-### Propensity of transcription when $N=1$
-In the simplest case when $N=1$, the promoter can be in one of two states.
-In state $S_0$, the promoter is not bound by any transcription factors, and in state $S_1$ the promoter is bound by $H_1$.
-Each state $S_j$ is linked with a relative activation $\alpha_j$, a number between 0 and 1 representing the activity of the promoter at this particular state.
-The propensity function is thus equal to the expected value of the activity of the promoter multiplied by the pre-mRNA production rate of $G$.
-
-\begin{align}
-  f(y_1, y_2, \ldots, y_N) & = \text{xpr} \cdot \sum_{j = 0}^{2^N - 1} \alpha_j \cdot P(S_j) \label{eqn:activ0} \\
-\end{align}
-
-For $N=1$, $P(S_1)$ is equal to the Hill equation, where $k_i$ represents the concentration of $H_i$ at half-occupation and $n_i$ represents the Hill coefficient. Typically, $n_i$ is between [1,10]
-
-\begin{align}
-  P(S_1) & = \frac{y_1^{n_1}}{k_1^{n_1} + y_1^{n_1}} \\
-             & = \frac{(y_1/k_1)^{n_1}}{1 + (y_1/k_1)^{n_1}}
-\end{align}
-
-The Hill equation can be simplified by letting $\nu_i = \left(\frac{y_i}{k_i}\right)^{n_i}$.
-
-\begin{align}
-P(S_1) & = \frac{\nu_1}{1 + \nu_1} \label{eqn:hillsimp}
-\end{align}
-
-Since $P(S_0) = 1 - P(S_1)$, the activation function is formulated and simplified as follows.
-
-\begin{align}
-f(y_1) & = \text{xpr} \cdot \left(\alpha_0 \cdot P(S_0) + \alpha_1 \cdot P(S_1)\right) \\
-           & = \text{xpr} \cdot \left(\alpha_0 \cdot \frac{1}{1 + \nu_1} + \alpha_1 \cdot \frac{\nu_1}{1 + \nu_1}\right) \\
-           & = \text{xpr} \cdot \frac{\alpha_0 + \alpha_1 \cdot \nu_1}{1 + \nu_1} \\
-\end{align}
-
-### Propensity of transcription when $N=2$
-
-When $N=2$, there are four states $S_j$.
-The relative activations $\alpha_j$ can be defined such that $H_1$ and $H_2$ are independent (additive) or synergistic (multiplicative).
-In order to define the propensity of transcription $f(.)$, the Hill equation $P(S_j)$ is extended for two transcription factors.
-
-Let $w_j$ be the numerator of $P(S_j)$, defined as the product of all transcription factors bound in that state:
-
-\begin{align}
-w_0 & = 1 \\
-w_1 & = \nu_1 \\
-w_2 & = \nu_2 \\
-w_3 & = \nu_1 \cdot \nu_2
-\end{align}
-
-The denominator of $P(S_j)$ is then equal to the sum of all $w_j$. The probability of state $S_j$ is thus defined as:
-
-\begin{align}
-    P(S_j) & = \frac{w_j}{\sum_{j=0}^{j < 2^N} w_j} \\
-               & = \frac{w_j}{1 + \nu_1 + \nu_2 + \nu_1 \cdot \nu_2} \\
-               & = \frac{w_j}{\prod_{i=1}^{i \leq N} (\nu_i + 1)}
-\end{align}
-
-Substituting $P(S_j)$ and $w_j$ into $f(.)$ results in the following equation:
-
-\begin{align}
-f(y_1, y_2) & = \text{xpr} \cdot \sum_{j = 0}^{2^N - 1} \alpha_j \cdot P(S_j) \\
- & = \text{xpr} \cdot \frac{\sum_{j = 0}^{2^N - 1} \alpha_j \cdot w_j}{\prod_{i=1}^{i \leq N} (\nu_i + 1)} \\
- & = \text{xpr} \cdot \frac{\alpha_0 + \alpha_1 \cdot \nu_1 + \alpha_2 \cdot \nu_2 + \alpha_3 \cdot \nu_1 \cdot \nu_2}{(\nu_1 + 1) \cdot (\nu_2 + 1)} \\
-\end{align}
-
-### Propensity of transcription for an arbitrary $N$
-For an arbitrary $N$, there are $2^N$ states $S_j$.
-The relative activations $\alpha_j$ can be defined such that $H_1$ and $H_2$ are independent (additive) or synergistic (multiplicative).
-In order to define the propensity of transcription $f(.)$, the Hill equation $P(S_j)$ is extended for $N$ transcription factors.
-
-Let $w_j$ be the numerator of $P(S_j)$, defined as the product of all transcription factors bound in that state:
-
-\begin{align}
-  w_j & = \prod_{i=1}^{i \leq N} (j \text{ mod } i) = 1 \text{ ? } \nu_i \text{ : } 1
-\end{align}
-
-The denominator of $P(S_j)$ is then equal to the sum of all $w_j$. The probability of state $S_j$ is thus defined as:
-
-\begin{align}
-P(S_j) & = \frac{w_j}{\sum_{j=0}^{j < 2^N} w_j} \\
-& = \frac{w_j}{\prod_{i=1}^{i \leq N} (\nu_i + 1)}
-\end{align}
-
-Substituting $P(S_j)$ into $f(.)$ yields:
-
-\begin{align}
-f(y_1, y_2, \ldots, y_N) & = \text{xpr} \cdot \sum_{j = 0}^{2^N - 1} \alpha_j \cdot P(S_j) \\
-& = \text{xpr} \cdot \frac{\sum_{j = 0}^{2^N - 1} \alpha_j \cdot w_j}{\prod_{i=1}^{i \leq N} (\nu_i + 1)} \label{eqn:prop2n}
-\end{align}
-
-### Propensity of transcription for a large $N$
-For large values of $N$, computing $f(.)$ is practically infeasible as it requires performing $2^N$ summations.
-In order to greatly simplify $f(.)$, $\alpha_j$ could be defined as 0 when one of the regulators inhibits transcription and 1 otherwise.
-
-\begin{equation}
-\alpha_j = \begin{cases}
- 0 & \text{ if } \exists i : j \text{ mod } i = 1 \text{ and } H_i \text{ represses } G \\
- 1 & \text{otherwise}
-\end{cases} \label{eqn:assalpha}
-\end{equation}
-
-Substituting equation \ref{eqn:assalpha} into equation \ref{eqn:prop2n} and defining $R = \{1, 2, \ldots, N\}$ and $R^+ = \{i | H_i \text{ activates } G\}$ yields the simplified propensity function:
-
-\begin{align}
-f(y_1, y_2, \ldots, y_N) & = \text{xpr} \cdot \frac{\prod_{i \in R^+} (\nu_i + 1)}{\prod_{i \in R} (\nu_i + 1)}
-\end{align}
-
-### Independence, synergism and basal expression
-The definition of $\alpha_j$ as in equation \ref{eqn:assalpha} presents two main limitations.
-Firstly, since $\alpha_0 = 1$, it is impossible to tweak the propensity of transcription when no transcription factors are bound.
-Secondly, it is not possible to tweak the independence and synergism of multiple regulators.
-
-Let $\text{ba} \in [0,1]$ denote the basal expression strength $G$ (i.e.
-how much will $G$ be expressed when no transcription factors are bound), and $\text{sy} \in [0,1]$ denote the synergism of regulators $H_i$ of $G$, the transcription propensity becomes:
-
-\begin{align}
-f(y_1, y_2, \ldots, y_N) & = \text{xpr} \cdot \frac{\text{ba} - \text{sy}^{|R^+|} + \prod_{i \in R^+} (\nu_i + \text{sy})}{\prod_{i \in R} (\nu_i + 1)}
-\end{align}
-
-## Simulate single cells {#sec:dyngen-simcell}
-dyngen uses Gillespie's stochastic simulation algorithm (SSA) [@gillespie_exactstochasticsimulation_1977] to simulate dynamic processes.
-An SSA simulation is an iterative process where at each iteration one reaction is triggered.
-
-Each reaction consists of its propensity -- a formula to calculate the probability of the reaction occurring during an infinitesimal time interval -- and the effect -- how it will affect the current state if triggered.
-Each time a reaction is triggered, the simulation time is incremented by $\tau = \frac{1}{\sum_j prop_j} \ln\left(\frac{1}{r}\right)$, with $r \in U(0, 1)$ and $prop_j$ the propensity value of the $j$th reaction for the current state of the simulation.
-
-GillespieSSA2 is an optimised library for performing SSA simulations.
-The propensity functions are compiled to C++ and SSA approximations can be used which allow triggering many reactions simultaneously at each iteration.
-The framework also allows storing the abundance levels of molecules only after a specific interval has passed since the previous census.
-By setting the census interval to 0, the whole simulation's trajectory is retained but many of these time points will contain very similar information.
-In addition to the abundance levels, also the propensity values and the number of firings of each of the reactions at each of the time steps can be retained, as well as specific sub-calculations of the propensity values, such as the regulator activity level $reg_{G,H}$.
-
-## Simulate experiment {#sec:dyngen-experiment}
-From the SSA simulation we obtain the abundance levels of all the molecules at every state.
-We need to replicate technical effects introduced by experimental protocols in order to obtain data that is similar to real data.
-For this, the cells are sampled from the simulations and molecules are sampled for each of the cells.
-Gene capture rates and library sizes are empirically derived from real datasets as to match real technical variation.
-
-### Sample cells
-In this step, $N$ cells are sampled from the simulations.
-Two approaches are implemented: sampling from an unsynchronised population of single cells (snapshot) or sampling at multiple time points in a synchronised population (time series).
-
-**Snapshot** The backbone consists of several states linked together by transition edges with length $L_i$, to which the different states in the different simulations have been mapped (Figure\ \ref{fig:sample_cells}A).
-From each transition, $N_i = N / \frac{L_i}{\sum L_i}$ cells are sampled uniformly, rounded such that $\sum N_i = N$.
-
-**Time series** Assuming that the final time of the simulation is $T$, the interval $[0, T]$ is divided into $k$ equal intervals of width $w$ separated by $k-1$ gaps of width $g$.
-$N_i = N / k$ cells are sampled uniformly from each interval (Figure\ \ref{fig:sample_cells}B), rounded such that $\sum N_i = N$.
-By default, $k = 8$ and $g = 0.75$.
-For usual dyngen simulations, $10 \leq T \leq 20$.
-For larger values of $T$, $k$ and $g$ should be increased accordingly.
-
-### Sample molecules
-Molecules are sampled from the simulation to replicate how molecules are experimentally sampled.
-A real dataset is downloaded from a repository of single-cell RNA-seq datasets [@cannoodt_singlecellomicsdatasets_2018].
-For each _in silico_ cell $i$, draw its library size $ls_i$ from the distribution of transcript counts per cell in the real dataset.
-The capture rate $cr_j$ of each _in silico_ molecule type $j$ is drawn from $N(1, 0.05)$.
-Finally, for each cell $i$, draw $ls_i$ molecules from the multinomial distribution with probabilities $cr_j \times ab_{i,j}$ with $ab_{i,j}$ the molecule abundance level of molecule $j$ in cell $i$.
-
-## Simulating batch effects {#sec:dyngen-batcheffect}
-Simulating batch effects can be performed in multiple ways.
-One such way is to perform the first two steps of the creation of a dyngen model (defining the module network and generating the GRN).
-For each desired batch, create a separate model for which random kinetics are generated and perform all subsequent dyngen steps (convert to reactions, simulate gold standard, simulate single cells, simulate experiment).
-Since each separate model has different underlying kinetics, the combined output will resemble having batch effects.
-
-## Determining the ground-truth trajectory {#sec:dyngen-groundtruth}
-To construct the ground-truth trajectory, the user needs to provide the ground-truth state network alongside the initial module network (Figure\ \ref{fig:example_backbones}).
-Each edge in the state network specifies which modules are allowed to change in expression in transitioning from one state to another.
-For each edge, a simulation is run using the end state of an upstream branch as the initial expression vector, and only allowing the modules as predefined by the attribute to change.
-
-As an example, consider the cyclic trajectory shown in Figure\ \ref{fig:example_backbones}.
-State S0 begins with an expression vector of all zero values.
-To simulate the transition from S0 to S1, regulation of the genes in modules A, B and C are turned on.
-After a predefined period of time, the end state of this transition is considered the expression vector of state S1.
-To simulate the transition from S1 to S2, regulation of the genes in modules D and E are turned on, while the regulation of genes in module C is turned off.
-During this simulation, the expression of genes in modules A, B, D, and E is thus allowed to change.
-The end state of the simulation is considered the expression vector of state S2.
-
-For each of the branches in the state network, an expression matrix and the corresponding progression time along that branch are retained.
-To map a simulated cell to the ground-truth, the correlation between its expression values and the expression matrix of the ground-truth trajectory is calculated, and the cell is mapped to the position in the ground-truth trajectory that has the highest correlation.
-
-## Determining the cell-specific ground-truth regulatory network {#sec:dyngen-extractgrn}
-Calculating the regulatory effect of a regulator $R$ on a target $T$ (Figure\ \ref{fig:explain_methods}F) requires determining the contribution of $R$ in the propensity function of the transcription of $T$ (section\ \ref{sec:dyngen-reactions}) with respect to other regulators.
-This information is useful, amongst others, for benchmarking cell-specific network inference methods.
-
-The regulatory effect of $R$ on $T$ at a particular state $S$ is defined as the change in the propensity of transcription when $R$ is set to zero, scaled by the inverse of the pre-mRNA production rate of $T$.
-More formally:
-
-\begin{eqnarray*}
-  \text{regeffect}_G & = \frac{\text{proptrans}_G(S) - \text{proptrans}_G(S[\z T \leftarrow 0])}{\xpr G}
-\end{eqnarray*}
-
-Determining the regulatory effect for all interactions and cells in the dataset yields the complete cell-specific ground-truth GRN.
-The regulatory effect lies between $[-1, 1]$, where -1 represents complete inhibition of $T$ by $R$, 1 represents maximal activation of $T$ by $R$, and 0 represents inactivity of the regulatory interaction between $R$ and $T$.
-
-## Comparison of cell-specific network inference methods {#sec:dyngen-nicompare}
-42 datasets were generated using the 14 different predefined backbones and three different seeds.
-For every cell in the dataset, the transcriptomics profile and the corresponding cell-specific ground-truth regulatory network was determined (Section\ \ref{sec:dyngen-extractgrn}).
-
-We selected three cell-specific NI methods: SCENIC [@aibar_scenicsinglecellregulatory_2017], LIONESS [@kuijjer_estimatingsamplespecificregulatory_2015; @kuijjer_estimatingsamplespecificregulatory_2019], and SSN [@liu_personalizedcharacterizationdiseases_2016].
-
-LIONESS [@kuijjer_estimatingsamplespecificregulatory_2019] runs a NI method multiple times to construct cell-specific GRNs.
-LIONESS first infers a GRN with all of the samples.
-A second GRN is inferred with all samples except one particular profile.
-The cell-specific GRN for that particular profile is defined as the difference between the two GRN matrices.
-This process is repeated for all profiles, resulting in a cell-specific GRN.
-By default, LIONESS uses PANDA [@glass_passingmessagesbiological_2013] to infer GRNs, but since dyngen does not produce motif data and motif data is required by PANDA, PANDA is inapplicable in this context.
-Instead, we used the lionessR [@kuijjer_lionessrsinglesample_2019] implementation of LIONESS, which uses by default the Pearson correlation as a NI method.
-We marked results from this implementation as "LIONESS + Pearson".
-
-SSN [@liu_personalizedcharacterizationdiseases_2016] follows, in essence, the exact same methodology as LIONESS except that it specifically only uses the Pearson correlation.
-It is worth noting that the LIONESS preprint was released before the publication of SSN.
-Since no implementation was provided by the authors, we implemented SSN in R using basic R and tidyverse functions [@wickham_welcometidyverse_2019] and marked results from this implementation as "SSN*".
-
-SCENIC [@aibar_scenicsinglecellregulatory_2017] consists of four main steps.
-First, classical network inference is performed with stochastic gradient boosting machines using `arboreto` [@moerman_grnboost2arboretoefficient_2019].
-Second, the top 10 regulators of every target gene are selected.
-Interactions are grouped together in 'modules'; each module contains one regulator and all of its targets.
-Next, the modules are filtered using motif analysis.
-Finally, for each module and each cell, an activity score is calculated using AUCell.
-As a post-processing of this output, all modules and the corresponding activity scores are combined back into a cell-specific GRN consisting of (cell, regulator, target, score) pairs.
-For this analysis, the Python implementation of SCENIC was used, namely pySCENIC [@vandesande_scalablescenicworkflow_2020].
-Since dyngen does not generate motif data, step 3 in this analysis is skipped.
-
-The Area Under the Receiver Operating Characteristic-curve (AUROC) and Area Under the Precision-Recall curve (AUPR) metrics are common metrics for evaluating a predicted GRN with a ground-truth GRN [@marbach_generatingrealisticsilico_2009].
-To compare a predicted cell-specific GRN with the ground-truth cell-specific GRN, the top 10'000 interactions per cell is retained, and the mean AUROC and AUPR scores are calculated across all cells.
-
-We compared the mean AUROC and AUPR scores obtained by the three CSNI methods across all datasets by performing pairwise non-parametric paired Durbin-Conover tests [@conover_multiplecomparisonsprocedures_1979] using `pairwiseComparisons` [@patil_pairwisecomparisonsmultiplepairwise_2019].
-Reported p-values are adjusted for multiple testing using Holm correction [@holm_simplesequentiallyrejective_1979].
-
-## Comparison of RNA velocity methods {#sec:dyngen-velcompare}
-
-Three datasets were generated for each of the 14 different predefined backbones, resulting in a collection of 42 datasets.
-Throughout each of the simulation, the propensity of the transcription and mRNA decay is collected, as the RNA velocity of a gene at any point in the simulation is the difference between the transcription propensity and the mRNA decay propensity.
-
-We applied two RNA velocity methods: velocyto [@lamanno_rnavelocitysingle_2018], as implemented in the `velocyto.py` package, and scvelo method [@bergen_generalizingrnavelocity_2020], as implemented in the `scvelo` package.
-For scvelo, we chose two parameter settings for "mode", namely "stochastic" and "dynamical".
-For both methods, we used the same normalized data as provided by dyngen, with no extra cell or feature filtering, but otherwise matched the parameters to their respective tutorial vignettes as well as possible.
-
-We compared each RNA velocity prediction to the ground-truth using two metrics: the velocity correlation and the velocity arrow cosine.
-For the velocity correlation, we extracted a ground truth RNA velocity by subtracting for each mRNA molecule the propensity of its production by the propensity of its degradation.
-If the expression of an mRNA will increase in the future, this value is positive, while it is negative if it is going to decrease.
-For each gene, we determined its velocity correlation by calculating the Spearman rank correlation between the ground truth velocity with the observed velocity.
-For the velocity arrow cosine, we determined a set of 100 trajectory waypoints uniformly spread on the trajectory.
-For each waypoint, we weighted each cell based on a Gaussian kernel on its geodesic distance from the waypoint.
-These weights were used to calculate a weighted average velocity vector of each waypoint.
-We then calculated for each waypoint the cosine similarity between this velocity vector and the known direction of the trajectory.
-
-We compared the velocity correlation and velocity arrow cosine scores obtained by velocyto and scvelo across all datasets by performing pairwise non-parametric paired Durbin-Conover tests [@conover_multiplecomparisonsprocedures_1979] using `pairwiseComparisons` [@patil_pairwisecomparisonsmultiplepairwise_2019].
-Reported p-values are adjusted for multiple testing using Holm correction [@holm_simplesequentiallyrejective_1979].
-
-## Comparison of trajectory alignment {#sec:dyngen-tacompare}[c]
-
-Four custom linear backbones of varying sizes were constructed.
-For each of these backbones, 10 datasets were generated with 10 different seeds, resulting in a total of 40 datasets.
-Every dataset is generated in three main steps.
-First, the GRN is generated based on the given backbone.
-Next, generating the kinetics, gold standard, and cells is performed twice, resulting in two sub-datasets.
-Finally, the two sub-datasets are combined and cells are sampled from the combined dataset.
-Since the two sub-datasets were simulated with different kinetic parameters, the combined dataset will contain two trajectories.
-
-On each combined dataset we applied two trajectory alignment methods, Dynamic Time Warping (DTW) [@giorgino_computingvisualizingdynamic_2009] and cellAlign [@alpert_alignmentsinglecelltrajectories_2018].
-DTW is designed to align temporal sequences by dilating or contracting the sequences to best match each other.
-cellAlign …[d] .
-
-To evaluate a trajectory alignment method on a paired dataset we computed the geodesic distances of each cell from the start of the trajectory, also called the _pseudotime_.
-For each dataset, the pseudotime values are rescaled between 0 and 1 to allow for easier comparison.
-A trajectory alignment produces a sequence of index pairs $[(i_0, j_0), (i_1, j_1), \ldots, (i_N, j_N)]$, where $i_0$ and $j_0$ are equal to 0 (the first position in both pseudotime series), $i_N$ and $j_N$ are equal to the respective last positions in the pair of pseudotime series, and $[i_0, i_1, \ldots, i_N]$ and $[j_0, j_1, \ldots, j_N]$ are in ascending order and can contain duplicates values.
-The 'score'[e] metric is defined as follows, where a score of 0 represents the worst possible ordering and 1 represents the best possible ordering.
-
-\begin{equation}
-  \textrm{score} = 1 - \textrm{area\_under\_curve}(pt_1[i_0 .. i_N] + pt_2[j_0 .. j_N], abs(pt_1[i_0 .. i_N] - pt_2[j_0 .. j_N]))
-\end{equation}
-
-\printbibliography
-\newpage
-
-# Supplementary Figures
-\setcounter{page}{1}
-\newrefsection
-\setcounter{table}{0}
-\renewcommand{\thetable}{S\arabic{table}}
-\setcounter{figure}{0}
-\renewcommand{\thefigure}{S\arabic{figure}}
-
-\newcommand{\yes}{\ding{51}}
-\newcommand{\no}{}
-\definecolor{light-gray}{rgb}{.85,.85,.85}
-\newcommand{\grayline}{\arrayrulecolor{light-gray}\cline{3-7}\arrayrulecolor{black}}
-\newcommand{\blackline}{\arrayrulecolor{black}\cline{3-7}}
-\newcommand{\method}[1]{#1}
-
-\begin{table}[H]
-    \caption{
-      \textbf{Feature comparison of single-cell synthetic data generators and simulators.} $ ^1$ As showcased by vignette.
-$^2$ As showcased by Van den Berge et al.
-\cite{vandenberge_trajectorybaseddifferentialexpression_2020}.
-    } \label{tab:comparison}
-    \small
-    \begin{tabular}{p{.25cm}l|*{5}{>{\centering\arraybackslash}p{1.25cm}|}}
-            \blackline
-            & \method{} & \method{splatter} & \method{powsimR} & \method{PROSSTT} & \method{SymSim} & \method{dyngen} \\
-            \blackline
-            \multicolumn{7}{l}{\textbf{Available modality outputs}} \\
-            \blackline
-            - & mRNA expression & \yes & \yes & \yes & \yes & \yes \\ \grayline
-            - & Pre-mRNA expression & \no & \no & \no & \no & \yes \\ \grayline
-            - & Protein expression & \no & \no & \no & \no & \yes \\ \grayline
-            - & Promotor activity & \no & \no & \no & \yes & \yes \\ \grayline
-            - & Reaction activity & \no & \no & \no & \no & \yes \\
-            \blackline
-            \multicolumn{7}{l}{\textbf{Available ground-truth outputs}} \\
-            \blackline
-            - & True counts & \yes & \no & \no & \yes & \yes \\ \grayline
-            - & Cluster labels & \yes & \yes & \no & \yes & \yes \\ \grayline
-            - & Trajectory & \yes & \no & \yes $ ^1$ & \yes & \yes \\ \grayline
-            - & Batch labels & \yes & \no & \no & \no & \yes $ ^1$\\ \grayline
-            - & Differential expression & \no & \no & \no & \yes & \yes $ ^2$ \\ \grayline
-            - & Knocked down regulators & \no & \no & \no & \no & \yes \\ \grayline
-            - & Regulatory network & \no & \no & \no & \yes & \yes \\ \grayline
-            - & Cell-specific regulatory network & \no & \no & \no & \no & \yes \\
-            \blackline
-            \multicolumn{7}{l}{\textbf{Emulate experimental effects}} \\
-            \blackline
-            - & Single-cell RNA sequencing & \yes & \yes & \yes & \yes & \yes \\ \grayline
-            - & Batch effects & \yes & \no & \no & \no & \yes $ ^1$ \\ \grayline
-            - & Knockdown experiment & \no & \no & \no & \no & \yes \\ \grayline
-            - & Time-series & \no & \no & \no & \no & \yes \\ \grayline
-            - & Snapshot & \no & \no & \no & \no & \yes \\
-            \blackline
-            \multicolumn{7}{l}{\textbf{Evaluation applications}} \\
-            \blackline
-            - & Clustering & \yes & \yes & \no & \yes & \no \\ \grayline
-            - & Trajectory inference & \yes & \no & \yes & \yes & \yes \\ \grayline
-            - & Network inference & \no & \no & \no & \yes & \yes \\ \grayline
-            - & Cell-specific network inference & \no & \no & \no & \no & \yes \\ \grayline
-            - & Differential expression & \yes & \no & \no & \no & \no \\ \grayline
-            - & Trajectory differential expression & \no & \no & \no & \no & \yes $ ^2$ \\ \grayline
-            - & Batch effect correction & \yes & \no & \no & \no & \yes \\ \grayline
-            - & RNA Velocity & \no & \no & \no & \no & \yes \\ \grayline
-            - & Trajectory alignment & \no & \no & \no & \no & \yes \\
-            \blackline
-            %\multicolumn{7}{l}{\textbf{Showcased dataset}} \\
-            %\blackline
-            %- & Number of cells &  &  &  &  & \\ \grayline
-            %- & Number of genes &  &  &  &  & \\ \grayline
-            %- & Execution time & & & & & \\
-            \blackline
-\end{tabular}
-\end{table}
-
-\begin{figure}[H]
-    \centering
-    \includegraphics[width=\linewidth]{result_files/usecase_trajectory_alignment/supp_fig.pdf}
-    \caption{
-            \textbf{dyngen allows benchmarking of trajectory alignment methods.} 
- \textbf{A, B}: Two samples containing a linear trajectory, generated by dyngen.
-  \textbf{C, D}: Result of the DTW alignment on the samples in \textbf{A} and \textbf{B}.
-In \textbf{C}, the individual mappings of the alignment between cells are shown.
-In \textbf{D}, the accumulated distance matrix between the two trajectories is shown, including the black warping path, corresponding to these cell to cell alignments in \textbf{C}.
-  \textbf{E, F}: Shows the accumulated distance matrices obtained after using DTW on two trajectories where noise (noise level of 0.4) was added to the count matrix.
-In E the complete count matrices were used to perform the alignment.
-In F, smoothed pseudocells were used.
-  \textbf{G:} Shows the influence of added noise to the different processing methods. We can see that DTW + smoothing performs best in noisy circumstances.
-    }[f]
-    \label{fig:traj_align}
-\end{figure}
-
-\begin{figure}[H]
-    \centering
-    \includegraphics[width=\linewidth]{result_files/usecase_rna_velocity/supp_fig.pdf}
-    \caption{
-            \textbf{dyngen allows benchmarking of RNA velocity methods.} 
-            \textbf{A:} The ground-truth information of a bifurcating dataset: ground-truth trajectory (left), gene expression of a gene B5\_TF1 (middle), and the RNA velocity of B5\_TF1 (right).
-\textbf{B:} The RNA velocity estimates of gene B5\_TF1 by the different methods.
-\textbf{C:} The velocity stream plots produced from the predictions of each method, as generated by scvelo.
-\textbf{D:} The predictions scored by two different metrics, the velocity correlation and the velocity arrow cosine.
-The velocity correlation is the correlation between the ground-truth velocity (A, right) and the predicted velocity (B).
-The velocity arrow cosine is the cosine similarity between the direction of segments of the ground-truth trajectory (A, left) and the RNA velocity values calculated at those points (C).
-    }
-    \label{fig:velocity}
-\end{figure}
-
-\begin{figure}[H]
-    \centering
-    \includegraphics[width=\linewidth]{result_files/usecase_network_inference/supp_fig.pdf}
-    \caption{
-            \textbf{dyngen allows benchmarking Cell-specific Network Inference (CSNI) methods.} 
-            \textbf{A:} A cell is simulated using the global gene regulatory network (GRN, top left).
-However, at any particular state in the simulation, only a fraction of the gene regulatory interactions are active.
-            \textbf{B:} CSNI methods were executed to predict the regulatory interactions that are active in each cell specifically.
-Using the ground-truth cell-specific GRN, the performance of each method was quantified on 42 dyngen datasets.
-    }
-    \label{fig:scgrn}
-\end{figure}
-
-\begin{figure}[H]
-    \centering
-    \includegraphics[width=\textwidth]{result_files/explain_methods}
-    \caption{\textbf{The workflow of dyngen consists of six main steps.} 
-    \textbf{A:} The user needs to specify the desired module network or use a predefined module network. The module network is what determines the dynamic behaviour of simulated cells.
-    \textbf{B:} The number of desired transcription factors (which drive the desired dynamic process) are amongst the given modules and adds regulatory interactions according to the module network.
-Additional target genes (which do not influence the dynamic process) are added by sampling interactions from GRN interaction databases.
-    \textbf{C:} Each gene regulatory interaction in the GRN is converted to a set of biochemical reactions. 
-    \textbf{D:} Along with the module network, the user also needs to specify the backbone structure of expected cell states.
-The average expression of each edge in the backbone is simulated by activating a restricted set of genes for each edge.
-    \textbf{E:} Multiple Gillespie SSA simulations are run using the reactions defined in step C.
-The counts of each of the molecules at each time step are extracted.
-Each time step is mapped to a point in the backbone.
-    \textbf{F:} The molecule levels of multiple simulations are shown over time (left).
-From each simulation, multiple cells are sampled (from left to middle).
-Technical noise from profiling is simulated by sampling molecules from the set of molecules inside each cell (from middle to right).
-    }
-    \label{fig:explain_methods}
-\end{figure}
-
-\begin{figure}[H]
-    \centering
-    \includegraphics[width=0.8\textwidth]{result_files/example_backbones_onlymodules}
-    \caption{
-      \textbf{The module network determines the type of dynamic process which simulated cells will undergo.} A module network describes the regulatory interactions between sets of transcription factors which drive the desired dynamic process.
-    }
-    \label{fig:example_backbones_onlymodules}
-\end{figure}
-
-\begin{figure}[H]
-    \centering
-    \includegraphics[width=\linewidth]{result_files/gen_feature_network}
-    \caption{
-        \textbf{Generating the feature network from a backbone consists of four main steps.}
-    }
-    \label{fig:gen_feature_network}
-\end{figure}
-
-\begin{table}[H]
-    \caption{
-      \textbf{Reactions affecting the abundance levels of pre-mRNA $\x G$, mature mRNA $\y G$ and proteins $\z G$ of gene $G$.} Define the set of regulators of $G$ as $\rs{G}$, the set of upregulating regulators of $G$ as $\rp G$, and the set of down-regulating regulators of $G$ as $\rn G$.
-Parameters used in the propensity formulae are defined in Table \ref{tab:reaction_params}.
-    } \label{tab:reaction_def}
-    \centering
-    \begin{tabular}{|lcc|}
-        \hline
-        Reaction & Effect & Propensity \\ \hline \hline
-        Transcription & $\emptyset \rightarrow \x G$ & $\xpr G \times \frac{\ba G - \ind{G}^{|\rp{G}|} + \prod\limits_{H \in \rp{G}}(\ind G + \buf{G,H})}{\prod\limits_{H \in \rs{G}}(1 + \buf{G,H})}$ \\
-        Splicing & $\x G \rightarrow \y G$ & $\ysr G \times \x G$ \\
-        Translation & $\y G \rightarrow \y G + \z G$ & $\zpr G \times \y G$ \\ \hline\hline
-        Pre-mRNA degradation & $\x G \rightarrow \emptyset$ & $\ydr G \times \x G$ \\
-        Mature mRNA degradation & $\y G \rightarrow \emptyset$ & $\ydr G \times \y G$ \\
-        Protein degradation & $\z G \rightarrow \emptyset$ & $\zdr G \times \z G$ \\ \hline
-    \end{tabular}
-\end{table}
-
-\begin{table}[H]
-    \caption{
-      \textbf{Default parameters defined for the calculation of reaction propensity functions.}
-    } \label{tab:reaction_params}
-    \centering
-    \begin{tabular}{|lrl|}
-        \hline
-        Parameter & Symbol & Definition \\ \hline \hline
-        Transcription rate & $\xpr{G}$ & $\in U(10, 20)$ \\
-        Splicing rate & $\ysr G$ & $= \ln(2)\ /\ 2$ \\
-        Translation rate & $\zpr{G}$ & $\in U(100, 150)$ \\
-        (Pre-)mRNA half-life & $\yhl{G}$ & $\in U(2.5, 5)$ \\
-        Protein half-life & $\zhl G$ & $\in U(5, 10)$ \\
-        Interaction strength & $\str{G,H}$ & $\in 10^{U(0, 2)}$ * \\
-        Hill coefficient & $\hill{G,H}$ & $\in U(0.5, 2)$ * \\
-        Independence factor & $\ind G$ & $\in U(0, 1)$ * \\ \hline\hline
-        (Pre-)mRNA degradation rate & $\ydr G$ & $= \ln(2)\ /\ \yhl G$ \\
-        Protein degradation rate & $\zdr G$ & $= \ln(2)\ /\ \zhl G$ \\
-        Dissociation constant & $\dis H$ & $= 0.5 \times \frac{\xpr H \times \ysr H \times \zpr H}{(\ydr H + \ysr H) \times \ydr H \times \zdr H}$ \\
-        Binding strength & $\buf{G,H}$ & $= \str{G,H} \times \left(\z H\ /\ \dis H\right) ^ {\hill{G,H}}$ \\
-        Basal expression & $\ba G$ & $= \begin{cases} 1 & \mbox{if } \rp{G} = \emptyset \\ 0.0001 & \mbox{if } \rn{G} = \emptyset \mbox{ and } \rp{G} \neq \emptyset \\ 0.5 & \mbox{otherwise} \end{cases}$ * \\ \hline
-        \multicolumn{3}{l}{*: unless $G$ is a TF, then the value is determined by the backbone.}
-    \end{tabular}
-\end{table}
-
-\begin{figure}[H]
-    \centering
-    \includegraphics[width=.6\linewidth]{result_files/sample_cells.pdf}
-    \caption{
-        \textbf{Two approaches can be used to sample cells from simulations: snapshot and time-series.}
-    }
-    \label{fig:sample_cells}
-\end{figure}
-
-\begin{figure}[H]
-    \centering
-    \includegraphics[width=0.8\textwidth]{result_files/example_backbones}
-    \caption{\textbf{Examples of the ground-truth state networks which need to be provided alongside the module network.}}
-    \label{fig:example_backbones}
-\end{figure}
-
-[a]todo: update
-[b]// todo: upload temporary data to zenodo?
-[c]todo: update this section
-[d]@louise.deconinck@gmail.com could you add a description for cellAlign?
-_Assigned to Louise_
-[e]TODO: rename
-[f]TODO: update text
+These timings were measured using 30 (out of 32) threads using a AMD
+Ryzen 9 5950X clocked at 3.4GHz.

### 2021-03-25 manuscript/render_manuscript.R

@@ -8,7 +8,7 @@ if (Sys.info()[["user"]] == "rcannood") {
   drive <- drive_download(as_id("1gU6--vq988dQS3qYQRlIqC1QO96A4XM81s54K6YJKIk"), type = "text", overwrite = TRUE, path = tempfile())
 
   # read docx
-  readr::read_lines(drive$local_path) %>%
+  c(readr::read_lines(drive$local_path), readr::read_lines("manuscript/vignettes.Rmd")) %>%
     str_replace_all("([^ ])(\\[@[^\\]]*\\])", "\\1 \\2") %>% # add spaces before citations
     str_replace_all("^#", "\n#") %>%
     write_lines("manuscript/render.Rmd")
@@ -17,7 +17,7 @@ if (Sys.info()[["user"]] == "rcannood") {
 # render the manuscript
 render("manuscript/render.Rmd")
 
-system("pdftk manuscript/render.pdf cat 1-19 output manuscript/manuscript.pdf")
-system("pdftk manuscript/render.pdf cat 20-27 output manuscript/supplementary_files.pdf")
+system("pdftk manuscript/render.pdf cat 1-16 output manuscript/manuscript.pdf")
+system("pdftk manuscript/render.pdf cat 17-45 output manuscript/supplementary_files.pdf")
 
-file.remove("manuscript/render.pdf")
+# file.remove("manuscript/render.pdf")

### 2021-03-29 manuscript/render.Rmd
add submission files

--- 2021-03-25 manuscript/render.Rmd
@@
 This problem is tackled in three fully-configurable main steps.
-First, biological processes are mimicked by translating a gene regulatory network in a set of reactions (regulation, transcription, splicing, translation).
+First, biological processes are mimicked by translating a gene regulatory network into a set of reactions (regulation, transcription, splicing, translation).
 Second, individual cells are simulated using Gillespie's stochastic simulation algorithm [@gillespie_exactstochasticsimulation_1977], which is designed to work well in low-molecule simulations.
-Finally, (real) reference datasets are used to emulate single-cell omics profiling protocols.
+Finally, real reference datasets are used to emulate single-cell omics profiling protocols.
 
@@
       \textbf{F:} Similarly, dyngen can also keep track of the regulatory activity of every interaction.
-      \textbf{G:} A benchmark of trajectory inference methods has already been performed using the cell state ground-truth.
-      \textbf{H:} The cell state ground-truth enables evaluating trajectory alignment methods \cite{saelens_comparisonsinglecelltrajectory_2019}.
+      \textbf{G:} A benchmark of trajectory inference methods has already been performed using the cell state ground-truth \cite{saelens_comparisonsinglecelltrajectory_2019}.
+      \textbf{H:} The cell state ground-truth enables evaluating trajectory alignment methods.
       \textbf{I:} The reaction propensity ground-truth enables evaluating RNA velocity methods.
@@
       \textbf{A:} Trajectory alignment aligns two trajectories between samples.
-We use dyngen to evaluate dynamic time warping (DTW) and cellAlign when aligning two linear trajectories with different kinetic parameters.
+We evaluate dynamic time warping (DTW) and cellAlign when aligning two linear trajectories with different kinetic parameters based on the area differences between the worst possible alignment and the predicted alignment (Area Between Worst And Prediction, or ABWAP).
       \textbf{B:} RNA velocity calculates for each cell the direction in which the expression of each gene is moving.
@@
 For example, by comparing the transcriptomic profiles of cells from a diseased patient to a healthy control, it might be possible to detect transcriptomics differences (differential expression) of particular cells along a developmental process, or to detect an early stop of the trajectory of the diseased patient.
-Currently, trajectory alignment is limited to aligning linear trajectories, though other topologies of trajectory could be aligned as well.
+Currently, trajectory alignment is limited to aligning linear trajectories, though other topologies of a trajectory could be aligned as well.
 Dynamic Time Warping (DTW) [@giorgino_computingvisualizingdynamic_2009] is a method designed for aligning temporal sequences for speech recognition but has since been used to compare gene expression kinetics from many different biological processes [@cacchiarelli_aligningsinglecelldevelopmental_2018; @kanton_organoidsinglecellgenomic_2019; @mcfaline-figueroa_pooledsinglecellgenetic_2019; @alpert_alignmentsinglecelltrajectories_2018].
-We evaluate the performance of DTW by simulating two linear trajectories with slightly different simulation kinetics and assessing the accuracy of the alignment by DTW.
-We compare the performance of DTW against a version of DTW where gene expression is first smoothed, for varying degrees of noise added to the gene expression (Figure\ \ref{fig:traj_align}).
-We observe that DTW+smoother performs significantly better than the unsmoothed version across all levels of noise.
+cellAlign [@alpert_alignmentsinglecelltrajectories_2018] uses DTW to perform trajectory alignment, but also includes interpolation and scaling of the single cell data as a preprocessing step.
+We evaluate the performance of DTW and cellAlign by simulating 40 datasets, each containing two linear trajectories generated with the same gene regulatory network but with slightly different simulation kinetics.
+We assess the accuracy of the obtained alignments by comparing the generated alignment path with the worst possible alignment that could be performed (ABWAP).
+A visual guide for this metric can be found in Figure \ \ref{fig:traj_align}D.
+Overall, cellAlign performs significantly better than DTW (Figure\ \ref{fig:traj_align}, which is likely due to the interpolation and scaling steps provided by cellAlign, reducing noise in the data and improving the comparability of the trajectories.
 Note that, in this comparison, only linear trajectory alignment is performed.
 While dyngen can generate non-linear trajectories (e.g.
-cyclic or branching), constructing a quantitative accuracy metric for non-linear trajectory alignment is not trivial and an avenue for future work.[a]
+cyclic or branching), both aligning non-linear trajectories and constructing a quantitative accuracy metric for non-linear trajectory alignment is not trivial and an avenue for future work.
 
@@
 transcription rate minus the decay rate) for any particular gene is challenging, but the combined information is very informative in determining the directionality of cell progression in the trajectory.
-In terms of velocity correlation, no method performed significantly better than the other, whereas scvelo stochastic performed slightly worse than scvelo dynamical and velocyto in terms of velocity arrow cosine score.
+In terms of velocity correlation, no method performed significantly better than the other, whereas "scvelo stochastic" performed slightly worse than "scvelo dynamical" and velocyto in terms of velocity arrow cosine score.
 
@@
 
-# Data Availability[b]
+# Data Availability
+Source data for box plots in Figures\ \ref{fig:applications}, \ref{fig:traj_align}, \ref{fig:velocity} and \ref{fig:scgrn} are provided with this paper.
+All code and data required to reproduce the analysis are available on GitHub at [github.com/dynverse/dyngen_manuscript](https://github.com/dynverse/dyngen_manuscript).
+The datasets generated for the different use cases are available on Zenodo with record number 4637926 (doi: [10.5281/zenodo.4637926](https://doi.org/10.5281/zenodo.4637926)).
 
@@
 
-We compared the mean AUROC and AUPR scores obtained by the three CSNI methods across all datasets by performing pairwise non-parametric paired Durbin-Conover tests [@conover_multiplecomparisonsprocedures_1979] using `pairwiseComparisons` [@patil_pairwisecomparisonsmultiplepairwise_2019].
+We compared the mean AUROC and AUPR scores obtained by the three CSNI methods across all datasets by performing pairwise non-parametric paired two-sided Durbin-Conover tests [@conover_multiplecomparisonsprocedures_1979] using `pairwiseComparisons` [@patil_pairwisecomparisonsmultiplepairwise_2019].
 Reported p-values are adjusted for multiple testing using Holm correction [@holm_simplesequentiallyrejective_1979].
@@
 
-We compared the velocity correlation and velocity arrow cosine scores obtained by velocyto and scvelo across all datasets by performing pairwise non-parametric paired Durbin-Conover tests [@conover_multiplecomparisonsprocedures_1979] using `pairwiseComparisons` [@patil_pairwisecomparisonsmultiplepairwise_2019].
+We compared the velocity correlation and velocity arrow cosine scores obtained by velocyto and scvelo across all datasets by performing pairwise non-parametric paired two-sided Durbin-Conover tests [@conover_multiplecomparisonsprocedures_1979] using `pairwiseComparisons` [@patil_pairwisecomparisonsmultiplepairwise_2019].
 Reported p-values are adjusted for multiple testing using Holm correction [@holm_simplesequentiallyrejective_1979].
 
-## Comparison of trajectory alignment {#sec:dyngen-tacompare}[c]
+## Comparison of trajectory alignment {#sec:dyngen-tacompare}
 
@@
 DTW is designed to align temporal sequences by dilating or contracting the sequences to best match each other.
-cellAlign …[d] .
-
-To evaluate a trajectory alignment method on a paired dataset we computed the geodesic distances of each cell from the start of the trajectory, also called the _pseudotime_.
+cellAlign uses DTW to perform this alignment, but first interpolates and rescales the input data in order to better cope with single-cell omics data.
+
+To evaluate a trajectory alignment method on a combined dataset we computed the geodesic distances of each cell from the start of the trajectory, also called the _pseudotime_.
 For each dataset, the pseudotime values are rescaled between 0 and 1 to allow for easier comparison.
 A trajectory alignment produces a sequence of index pairs $[(i_0, j_0), (i_1, j_1), \ldots, (i_N, j_N)]$, where $i_0$ and $j_0$ are equal to 0 (the first position in both pseudotime series), $i_N$ and $j_N$ are equal to the respective last positions in the pair of pseudotime series, and $[i_0, i_1, \ldots, i_N]$ and $[j_0, j_1, \ldots, j_N]$ are in ascending order and can contain duplicates values.
-The 'score'[e] metric is defined as follows, where a score of 0 represents the worst possible ordering and 1 represents the best possible ordering.
+The ABWAP metric is defined as follows, where $pt_1$ and $pt_2$ are the unit pseudotime vectors.
+See Figure\ \ref{fig:traj_align}D for a visual interpretation of this metric.
 
 \begin{equation}
-  \textrm{score} = 1 - \textrm{area\_under\_curve}(pt_1[i_0 .. i_N] + pt_2[j_0 .. j_N], abs(pt_1[i_0 .. i_N] - pt_2[j_0 .. j_N]))
+  \textrm{ABWAP} = 1 - \textrm{area\_under\_curve}(pt_1[i_0 .. i_N] + pt_2[j_0 .. j_N], abs(pt_1[i_0 .. i_N] - pt_2[j_0 .. j_N]))
 \end{equation}
+
+We compared the ABWAP scores obtained by DTW and cellAlign across all datasets by performing pairwise non-parametric paired two-sided Durbin-Conover tests [@conover_multiplecomparisonsprocedures_1979] using `pairwiseComparisons` [@patil_pairwisecomparisonsmultiplepairwise_2019].
+Reported p-values are adjusted for multiple testing using Holm correction [@holm_simplesequentiallyrejective_1979].
 
@@
             \textbf{dyngen allows benchmarking of trajectory alignment methods.} 
- \textbf{A, B}: Two samples containing a linear trajectory, generated by dyngen.
-  \textbf{C, D}: Result of the DTW alignment on the samples in \textbf{A} and \textbf{B}.
-In \textbf{C}, the individual mappings of the alignment between cells are shown.
-In \textbf{D}, the accumulated distance matrix between the two trajectories is shown, including the black warping path, corresponding to these cell to cell alignments in \textbf{C}.
-  \textbf{E, F}: Shows the accumulated distance matrices obtained after using DTW on two trajectories where noise (noise level of 0.4) was added to the count matrix.
-In E the complete count matrices were used to perform the alignment.
-In F, smoothed pseudocells were used.
-  \textbf{G:} Shows the influence of added noise to the different processing methods. We can see that DTW + smoothing performs best in noisy circumstances.
-    }[f]
+ \textbf{A}: An example linear dataset in need of trajectory alignment. Dashed lines represent the gold-standard alignment between the two trajectories according to the respective pseudotimes.
+  \textbf{B}: Result of the DTW alignment on the two trajectories. 
+ \textbf{C}: DTW calculates an accumulated distance matrix.
+In this matrix, a warping path (shown in black), following a valley in the matrix from the bottom left to the top right corner is found.
+This shows how the trajectories best match each other.
+  \textbf{D}: Illustration of the Area Between Worst and Prediction (ABWAP) metric.
+The warping path from subfigure C is mapped to the respective pseudotimes from both trajectories.
+The ABWAP score is equal to the area between the prediction and the worst possible prediction.
+  \textbf{E:} An evaluation of DTW versus cellAlign on 40 different linear trajectories, in which cellAlign significantly outperforms DTW.
+    }
     \label{fig:traj_align}
@@
 
-[a]todo: update
-[b]// todo: upload temporary data to zenodo?
-[c]todo: update this section
-[d]@louise.deconinck@gmail.com could you add a description for cellAlign?
-_Assigned to Louise_
-[e]TODO: rename
-[f]TODO: update text
+# Supplementary Files
+
+## Vignette: Comparison to reference dataset
+
+In this vignette, we will take a look at characteristic features of
+dyngen versus the reference dataset it uses. To this end, we’ll be using
+[`countsimQC`](https://www.bioconductor.org/packages/release/bioc/html/countsimQC.html)
+[@soneson_unifiedqualityverification_2018] to calculate key statistics of both datasets
+and create comparative visualisations.
+
+### Run dyngen simulation
+
+We use an internal function from the dyngen package to download and
+cache one of the reference datasets.
+
+``` r
+library(tidyverse)
+library(dyngen)
+
+set.seed(1)
+
+data("realcounts", package = "dyngen")
+name_realcounts <- "zenodo_1443566_real_silver_bone-marrow-mesenchyme-erythrocyte-differentiation_mca"
+url_realcounts <- realcounts %>% filter(name == name_realcounts) %>% pull(url)
+realcount <- dyngen:::.download_cacheable_file(url_realcounts, getOption("dyngen_download_cache_dir"), verbose = FALSE)
+```
+
+We run a simple dyngen dataset as follows, where the number of cells and
+genes are determined by the size of the reference dataset.
+
+``` r
+backbone <- backbone_bifurcating_loop()
+
+num_cells <- nrow(realcount)
+num_feats <- ncol(realcount)
+num_tfs <- nrow(backbone$module_info)
+num_tar <- round((num_feats - num_tfs) / 2)
+num_hks <- num_feats - num_tfs - num_tar
+
+config <-
+  initialise_model(
+    backbone = backbone,
+    num_cells = num_cells,
+    num_tfs = num_tfs,
+    num_targets = num_tar,
+    num_hks = num_hks,
+    gold_standard_params = gold_standard_default(),
+    simulation_params = simulation_default(
+      total_time = 1000,
+      experiment_params = simulation_type_wild_type(num_simulations = 100)
+    ),
+    experiment_params = experiment_snapshot(
+      realcount = realcount
+    ),
+    verbose = FALSE
+  )
+```
+
+``` r
+
+# the simulation is being sped up because rendering all vignettes with one core
+
+# for pkgdown can otherwise take a very long time
+set.seed(1)
+
+config <-
+  initialise_model(
+    backbone = backbone,
+    num_cells = num_cells,
+    num_tfs = num_tfs,
+    num_targets = num_tar,
+    num_hks = num_hks,
+    verbose = interactive(),
+    download_cache_dir = tools::R_user_dir("dyngen", "data"),
+    simulation_params = simulation_default(
+      total_time = 1000,
+      census_interval = 2, 
+      ssa_algorithm = ssa_etl(tau = 300/3600),
+      experiment_params = simulation_type_wild_type(num_simulations = 10)
+    ),
+    experiment_params = experiment_snapshot(
+      realcount = realcount
+    )
+  )
+```
+
+``` r
+out <- generate_dataset(config, make_plots = TRUE)
+```
+
+    ## Generating TF network
+    ## Sampling feature network from real network
+    ## Generating kinetics for 3025 features
+    ## Generating formulae
+    ## Generating gold standard mod changes
+    ## Precompiling reactions for gold standard
+    ## Running gold simulations
+    ##   |                                                  | 0 % elapsed=00s     |========                                          | 14% elapsed=00s, remaining~01s  |===============                                   | 29% elapsed=00s, remaining~00s  |======================                            | 43% elapsed=00s, remaining~00s  |=============================                     | 57% elapsed=00s, remaining~00s  |====================================              | 71% elapsed=00s, remaining~00s  |===========================================       | 86% elapsed=00s, remaining~00s  |==================================================| 100% elapsed=01s, remaining~00s
+    ## Precompiling reactions for simulations
+    ## Running 10 simulations
+    ## Mapping simulations to gold standard
+    ## Performing dimred
+    ## Simulating experiment
+    ## Wrapping dataset
+    ## Making plots
+
+``` r
+out$plot
+```
+
+![](comparison_reference_files/figure-gfm/dyngen_generate-1.png)<!-- -->
+
+Both datasets are stored in a list for easy usage by countsimQC.
+
+``` r
+datasets <- list(
+  real = t(as.matrix(realcount)),
+  dyngen = t(as.matrix(out$dataset$counts))
+)
+
+ddsList <- lapply(datasets, function(ds) {
+  DESeq2::DESeqDataSetFromMatrix(
+    countData = round(as.matrix(ds)), 
+    colData = data.frame(sample = seq_len(ncol(ds))), 
+    design = ~1
+  )
+})
+```
+
+### Run countsimQC computations
+
+Below are some computations countsimQC makes. Normally these are not
+visible to the user, but for the sake of transparency these are included
+in the vignette.
+
+``` r
+library(countsimQC)
+
+### Define helper objects
+nDatasets <- length(ddsList)
+colRow <- c(2, 1)
+panelSize <- 4
+thm <- 
+  theme_bw() + 
+  theme(
+    axis.text = element_text(size = 15),
+    axis.title = element_text(size = 14),
+    strip.text = element_text(size = 15)
+  )
+```
+
+Compute key characteristics
+
+``` r
+obj <- countsimQC:::calculateDispersionsddsList(ddsList = ddsList, maxNForDisp = Inf)
+
+sampleCorrDF <- countsimQC:::calculateSampleCorrs(ddsList = obj, maxNForCorr = 500)
+
+featureCorrDF <- countsimQC:::calculateFeatureCorrs(ddsList = obj, maxNForCorr = 500)
+```
+
+Summarize sample characteristics
+
+``` r
+sampleDF <- map2_df(obj, names(obj), function(x, dataset_name) {
+  tibble(
+    dataset = dataset_name,
+    Libsize = colSums(x$dge$counts),
+    Fraczero = colMeans(x$dge$counts == 0),
+    TMM = x$dge$samples$norm.factors,
+    EffLibsize = Libsize * TMM
+  )
+})
+```
+
+Summarize feature characteristics
+
+``` r
+featureDF <- map2_df(obj, names(obj), function(x, dataset_name) {
+  rd <- SummarizedExperiment::rowData(x$dds)
+  tibble(
+    dataset = dataset_name,
+    Tagwise = sqrt(x$dge$tagwise.dispersion),
+    Common = sqrt(x$dge$common.dispersion),
+    Trend = sqrt(x$dge$trended.dispersion),
+    AveLogCPM = x$dge$AveLogCPM,
+    AveLogCPMDisp = x$dge$AveLogCPMDisp, 
+    average_log2_cpm = apply(edgeR::cpm(x$dge, prior.count = 2, log = TRUE), 1, mean), 
+    variance_log2_cpm = apply(edgeR::cpm(x$dge, prior.count = 2, log = TRUE), 1, var),
+    Fraczero = rowMeans(x$dge$counts == 0),
+    dispGeneEst = rd$dispGeneEst,
+    dispFit = rd$dispFit,
+    dispFinal = rd$dispersion,
+    baseMeanDisp = rd$baseMeanDisp,
+    baseMean = rd$baseMean
+  )
+})
+```
+
+Summarize data set characteristics
+
+``` r
+datasetDF <- map2_df(obj, names(obj), function(x, dataset_name) {
+  tibble(
+    dataset = dataset_name,
+    prior_df = paste0("prior.df = ", round(x$dge$prior.df, 2)),
+    nVars = nrow(x$dge$counts),
+    nSamples = ncol(x$dge$counts),
+    AveLogCPMDisp = 0.8 * max(featureDF$AveLogCPMDisp),
+    Tagwise = 0.9 * max(featureDF$Tagwise)
+  )
+})
+```
+
+### Data set dimensions
+
+These bar plots show the number of samples (columns) and features (rows)
+in each data set.
+
+Number of samples (columns)
+
+``` r
+ggplot(datasetDF, aes(x = dataset, y = nSamples, fill = dataset)) + 
+  geom_bar(stat = "identity", alpha = 0.5) + 
+  xlab("") + ylab("Number of samples (columns)") + 
+  thm + theme(axis.text.x = element_text(angle = 90, hjust = 1, vjust = 0.5))
+```
+
+![](comparison_reference_files/figure-gfm/nSamples-1.png)<!-- -->
+
+Number of features (rows)
+
+``` r
+ggplot(datasetDF, aes(x = dataset, y = nVars, fill = dataset)) + 
+  geom_bar(stat = "identity", alpha = 0.5) + 
+  xlab("") + ylab("Number of features (rows)") + 
+  thm + theme(axis.text.x = element_text(angle = 90, hjust = 1, vjust = 0.5))
+```
+
+![](comparison_reference_files/figure-gfm/nVariables-1.png)<!-- -->
+
+### Dispersion/BCV plots
+
+Disperson/BCV plots show the association between the average abundance
+and the dispersion or “biological coefficient of variation”
+(sqrt(dispersion)), as calculated by
+[`edgeR`](https://bioconductor.org/packages/release/bioc/html/edgeR.html)
+[@robinson_edgerbioconductorpackage_2010] and
+[`DESeq2`](http://bioconductor.org/packages/release/bioc/html/DESeq2.html)
+[@love_moderatedestimationfold_2014]. In the `edgeR` plot, the estimate of the
+prior degrees of freedom is indicated.
+
+#### edgeR
+
+The black dots represent the tagwise dispersion estimates, the red line
+the common dispersion and the blue curve represents the trended
+dispersion estimates. For further information about the dispersion
+estimation in `edgeR`, see Chen et al. [@chen_differentialexpressionanalysis_2014].
+
+``` r
+ggplot(featureDF %>% dplyr::arrange(AveLogCPMDisp), 
+       aes(x = AveLogCPMDisp, y = Tagwise)) + 
+  geom_point(size = 0.25, alpha = 0.5) + 
+  facet_wrap(~dataset, nrow = colRow[2]) + 
+  geom_line(aes(y = Trend), color = "blue", size = 1.5) + 
+  geom_line(aes(y = Common), color = "red", size = 1.5) +
+  geom_text(data = datasetDF, aes(label = prior_df)) + 
+  xlab("Average log CPM") + ylab("Biological coefficient of variation") + 
+  thm
+```
+
+![](comparison_reference_files/figure-gfm/BCVedgeR-1.png)<!-- -->
+
+#### DESeq2
+
+The black dots are the gene-wise dispersion estimates, the red curve the
+fitted mean-dispersion relationship and the blue circles represent the
+final dispersion estimates.For further information about the dispersion
+estimation in `DESeq2`, see Love et al. [@love_moderatedestimationfold_2014].
+
+``` r
+ggplot(featureDF %>% dplyr::arrange(baseMeanDisp), 
+       aes(x = baseMeanDisp, y = dispGeneEst)) + 
+  geom_point(size = 0.25, alpha = 0.5) + 
+  facet_wrap(~dataset, nrow = colRow[2]) + scale_x_log10() + scale_y_log10() +  
+  geom_point(aes(y = dispFinal), color = "lightblue", shape = 21) + 
+  geom_line(aes(y = dispFit), color = "red", size = 1.5) + 
+  xlab("Base mean") + ylab("Dispersion") + 
+  thm
+```
+
+![](comparison_reference_files/figure-gfm/dispersionDESeq2-1.png)<!-- -->
+
+### Mean-variance plots
+
+This scatter plot shows the relation between the empirical mean and
+variance of the features. The difference between these mean-variance
+plots and the mean-dispersion plots above is that the plots in this
+section do not take the information about the experimental design and
+sample grouping into account, but simply display the mean and variance
+of log2(CPM) estimates across all samples, calculated using the `cpm`
+function from
+[`edgeR`](https://bioconductor.org/packages/release/bioc/html/edgeR.html)
+[@robinson_edgerbioconductorpackage_2010], with a prior count of 2.
+
+``` r
+ggplot(featureDF, aes(x = average_log2_cpm, y = variance_log2_cpm)) + 
+  geom_point(size = 0.75, alpha = 0.5) + 
+  facet_wrap(~dataset, nrow = colRow[2]) + 
+  xlab("Mean of log2(CPM)") + ylab("Variance of log2(CPM)") + 
+  thm
+```
+
+![](comparison_reference_files/figure-gfm/meanVarSepScatter-1.png)<!-- -->
+
+### Library sizes
+
+This plot shows a histogram of the total read count per sample, i.e.,
+the column sums of the respective data matrices.
+
+``` r
+ggplot(sampleDF, aes(x = Libsize)) + geom_histogram(bins = 30) + 
+  facet_wrap(~dataset, nrow = colRow[2]) +
+  xlab("Library size") + thm
+```
+
+![](comparison_reference_files/figure-gfm/libsizeSepHist-1.png)<!-- -->
+
+### TMM normalization factors
+
+This plot shows a histogram of the TMM normalization factors [@robinson_scalingnormalizationmethod_2010], intended to adjust for differences in RNA
+composition, as calculated by
+[`edgeR`](https://bioconductor.org/packages/release/bioc/html/edgeR.html)
+[@robinson_edgerbioconductorpackage_2010].
+
+``` r
+ggplot(sampleDF, aes(x = TMM)) + geom_histogram(bins = 30) + 
+  facet_wrap(~dataset, nrow = colRow[2]) +
+  xlab("TMM normalization factor") + thm
+```
+
+![](comparison_reference_files/figure-gfm/tmmSepHist-1.png)<!-- -->
+
+### Effective library sizes
+
+This plot shows a histogram of the “effective library sizes,” defined as
+the total count per sample multiplied by the corresponding TMM
+normalization factor.
+
+``` r
+ggplot(sampleDF, aes(x = EffLibsize)) + geom_histogram(bins = 30) + 
+  facet_wrap(~dataset, nrow = colRow[2]) +
+  xlab("Effective library size") + thm
+```
+
+![](comparison_reference_files/figure-gfm/effLibsizeSepHist-1.png)<!-- -->
+
+### Expression distributions (average log CPM)
+
+This plot shows the distribution of average abundance values for the
+features. The abundances are log CPM values calculated by `edgeR`.
+
+``` r
+ggplot(featureDF, aes(x = AveLogCPM)) + geom_histogram(bins = 30) + 
+  facet_wrap(~dataset, nrow = colRow[2]) +
+  xlab("Average log CPM") + thm
+```
+
+![](comparison_reference_files/figure-gfm/logCPMSepHist-1.png)<!-- -->
+
+### Fraction zeros per sample
+
+This plot shows the distribution of the fraction of zeros observed per
+sample (column) in the count matrices.
+
+``` r
+ggplot(sampleDF, aes(x = Fraczero)) + geom_histogram(bins = 30) + 
+  facet_wrap(~dataset, nrow = colRow[2]) +
+  xlab("Fraction zeros per sample") + thm
+```
+
+![](comparison_reference_files/figure-gfm/fraczeroSampleSepHist-1.png)<!-- -->
+
+### Fraction zeros per feature
+
+This plot illustrates the distribution of the fraction of zeros observed
+per feature (row) in the count matrices.
+
+``` r
+ggplot(featureDF, aes(x = Fraczero)) + geom_histogram(bins = 30) + 
+  facet_wrap(~dataset, nrow = colRow[2]) +
+  xlab("Fraction zeros per feature") + thm
+```
+
+![](comparison_reference_files/figure-gfm/fraczeroFeatureSepHist-1.png)<!-- -->
+
+### Sample-sample correlations
+
+The plot below shows the distribution of Spearman correlation
+coefficients for pairs of samples, calculated from the log(CPM) values
+obtained via the `cpm` function from `edgeR`, with a prior.count of 2.
+
+``` r
+ggplot(sampleCorrDF, aes(x = Correlation)) + geom_histogram(bins = 30) + 
+  facet_wrap(~dataset, nrow = colRow[2]) +
+  xlab("Sample-sample correlation") + thm
+```
+
+![](comparison_reference_files/figure-gfm/sampleCorrSepHist-1.png)<!-- -->
+
+### Feature-feature correlations
+
+This plot illustrates the distribution of Spearman correlation
+coefficients for pairs of features, calculated from the log(CPM) values
+obtained via the `cpm` function from `edgeR`, with a prior.count of 2.
+Only non-constant features are considered.
+
+``` r
+ggplot(featureCorrDF, aes(x = Correlation)) + geom_histogram(bins = 30) + 
+  facet_wrap(~dataset, nrow = colRow[2]) +
+  xlab("Feature-feature correlation") + thm
+```
+
+![](comparison_reference_files/figure-gfm/featureCorrSepHist-1.png)<!-- -->
+
+### Library size vs fraction zeros
+
+This scatter plot shows the association between the total count (column
+sums) and the fraction of zeros observed per sample.
+
+``` r
+ggplot(sampleDF, aes(x = Libsize, y = Fraczero)) + 
+  geom_point(size = 1, alpha = 0.5) + 
+  facet_wrap(~dataset, nrow = colRow[2]) + 
+  xlab("Library size") + ylab("Fraction zeros") + thm
+```
+
+![](comparison_reference_files/figure-gfm/libsizeFraczeroSepScatter-1.png)<!-- -->
+
+### Mean expression vs fraction zeros
+
+This scatter plot shows the association between the average abundance
+and the fraction of zeros observed per feature. The abundance is defined
+as the log(CPM) values as calculated by `edgeR`.
+
+``` r
+ggplot(featureDF, aes(x = AveLogCPM, y = Fraczero)) + 
+  geom_point(size = 0.75, alpha = 0.5) + 
+  facet_wrap(~dataset, nrow = colRow[2]) + 
+  xlab("Average log CPM") + ylab("Fraction zeros") + thm
+```
+
+![](comparison_reference_files/figure-gfm/logCPMFraczeroSepScatter-1.png)<!-- -->
+
+## Vignette: On scalability and runtime
+
+In this vignette, we will take a look at the runtime of dyngen as the
+number of genes and the number of cells sampled is increased. We’ll be
+using the bifurcating cycle backbone which is well known for its
+beautiful 3D butterfly shape!
+
+``` r
+library(dyngen)
+library(tidyverse)
+
+set.seed(1)
+
+save_dir <- "scalability_and_runtime_runs"
+if (!dir.exists(save_dir)) dir.create(save_dir, recursive = TRUE)
+
+backbone <- backbone_bifurcating_cycle()
+```
+
+### Initial run
+
+We’ll be running this simulation a few times, with different values for
+`num_cells` and `num_features` to assess the scalability of dyngen. An
+example of a resulting dyngen model is shown here.
+
+``` r
+num_cells <- 100
+num_features <- 100
+num_tfs <- nrow(backbone$module_info)
+num_targets <- round((num_features - num_tfs) / 2)
+num_hks <- num_features - num_targets - num_tfs
+
+out <- 
+  initialise_model(
+    backbone = backbone,
+    num_tfs = num_tfs,
+    num_targets = num_targets,
+    num_hks = num_hks,
+    num_cells = num_cells,
+    gold_standard_params = gold_standard_default(
+      census_interval = 1,
+      tau = 100/3600
+    ),
+    simulation_params = simulation_default(
+      census_interval = 10,
+      ssa_algorithm = ssa_etl(tau = 300/3600),
+      experiment_params = simulation_type_wild_type(
+        num_simulations = num_cells / 10
+      )
+    ),
+    verbose = FALSE
+  ) %>% 
+  generate_dataset(make_plots = TRUE)
+```
+
+``` r
+out$plot
+```
+
+![](scalability_and_runtime_files/figure-gfm/example-1.png)<!-- -->
+
+We tweaked some of the parameters by running this particular backbone
+once with `num_cells = 100` and `num_features = 100` and verifying that
+the new parameters still yield the desired outcome. The parameters we
+tweaked are:
+
+-   On average, 10 cells are sampled per simulation
+    (e.g. `num_simulations = 100` and `num_cells = 1000`). You could
+    increase this ratio to get a better cell count yield from a given
+    set of simulations, but cells from the same simulation that are
+    temporally close will have highly correlated expression profiles.
+-   Increased time steps `tau`. This will make the Gillespie algorithm
+    slightly faster but might result in unexpected artifacts in the
+    simulated data.
+-   `census_interval` increased from 4 to 10. This will cause dyngen to
+    store an expression profile only every 10 time units. Since the
+    total simulation time is xxx, each simulation will result in yyy
+    data points. Note that on average only 10 data points are sampled
+    per simulation.
+
+For more information on parameter tuning, see the vignette ‘Advanced:
+tuning the simulation parameters’.
+
+### Timing experiments
+
+The simulations are run once with a large `num_features` and
+`num_cells`, a few times with varying `num_cells` and then once more
+with varying `num_features`. Every run is repeated three times in order
+to get a bit more stable time measurements. Since some of the
+simulations can take over 10 minutes, the timings results of the
+simulations are cached in the ‘scalability\_and\_runtime\_runs’
+folder.\`
+
+``` r
+settings <- bind_rows(
+  tibble(num_cells = 10000, num_features = 10000, rep = 1), #, rep = seq_len(3)),
+  crossing(
+    num_cells = seq(1000, 10000, by = 1000),
+    num_features = 100,
+    rep = seq_len(3)
+  ),
+  crossing(
+    num_cells = 100,
+    num_features = seq(1000, 10000, by = 1000),
+    rep = seq_len(3)
+  )
+) %>% 
+  mutate(filename = paste0(save_dir, "/cells", num_cells, "_feats", num_features, "_rep", rep, ".rds"))
+
+timings <- pmap_dfr(settings, function(num_cells, num_features, rep, filename) {
+  if (!file.exists(filename)) {
+    set.seed(rep)
+
+    cat("Running num_cells: ", num_cells, ", num_features: ", num_features, ", rep: ", rep, "\n", sep = "")
+    num_tfs <- nrow(backbone$module_info)
+    num_targets <- round((num_features - num_tfs) / 2)
+    num_hks <- num_features - num_targets - num_tfs
+
+    out <- 
+      initialise_model(
+        backbone = backbone,
+        num_tfs = num_tfs,
+        num_targets = num_targets,
+        num_hks = num_hks,
+        num_cells = num_cells,
+        gold_standard_params = gold_standard_default(
+          census_interval = 1,
+          tau = 100/3600
+        ),
+        simulation_params = simulation_default(
+          census_interval = 10,
+          ssa_algorithm = ssa_etl(tau = 300/3600),
+          experiment_params = simulation_type_wild_type(
+            num_simulations = num_cells / 10
+          )
+        ),
+        verbose = FALSE
+      ) %>% 
+      generate_dataset()
+
+    tim <- 
+      get_timings(out$model) %>% 
+      mutate(rep, num_cells, num_features)
+
+    write_rds(tim, filename, compress = "gz")
+  }
+
+  read_rds(filename)
+})
+
+timings_gr <- 
+  timings %>% 
+  group_by(group, task, num_cells, num_features) %>% 
+  summarise(time_elapsed = mean(time_elapsed), .groups = "drop")
+
+timings_sum <-
+  timings %>% 
+  group_by(num_cells, num_features, rep) %>%
+  summarise(time_elapsed = sum(time_elapsed), .groups = "drop")
+```
+
+### Simulate a large dataset (10k × 10k)
+
+Below is shown the timings of each of the steps in simulating a dyngen
+dataset containing 10’000 genes and 10’000 features. The total
+simulation time required is 1147 seconds, most of which is spent
+performing the simulations itself.
+
+``` r
+timings0 <- 
+  timings_gr %>% 
+  filter(num_cells == 10000, num_features == 10000) %>% 
+  mutate(name = forcats::fct_rev(forcats::fct_inorder(paste0(group, ": ", task))))
+
+ggplot(timings0) + 
+  geom_bar(aes(x = name, y = time_elapsed, fill = group), stat = "identity") +
+  scale_fill_brewer(palette = "Dark2") + 
+  theme_classic() +
+  theme(legend.position = "none") +
+  coord_flip() + 
+  labs(x = NULL, y = "Time (s)", fill = "dyngen stage")
+```
+
+![](scalability_and_runtime_files/figure-gfm/bblego-1.png)<!-- -->
+
+### Increasing the number of cells
+
+By increasing the number of cells from 1000 to 10’000 whilst keeping the
+number of features fixed, we can get an idea of how the simulation time
+scales w.r.t. the number of cells.
+
+``` r
+timings1 <- 
+  timings_gr %>% 
+  filter(num_features == 100) %>% 
+  group_by(num_cells, num_features, group) %>%
+  summarise(time_elapsed = sum(time_elapsed), .groups = "drop")
+
+ggplot(timings1) + 
+  geom_bar(aes(x = forcats::fct_inorder(as.character(num_cells)), y = time_elapsed, fill = forcats::fct_inorder(group)), stat = "identity") +
+  theme_classic() +
+  scale_fill_brewer(palette = "Dark2") +
+  labs(x = "Number of cells", y = "Average time (s)", fill = "dyngen step")
+```
+
+![](scalability_and_runtime_files/figure-gfm/figure1-1.png)<!-- -->
+
+It seems the execution time scales linearly w.r.t. the number of cells.
+This makes sense, because as the number of cells are increased, so do we
+increase the number of simulations made (which is not necessarily
+mandatory). Since the simulations are independent of each other and take
+up the most time, the execution time will scale linearly.
+
+``` r
+ggplot(timings_sum %>% filter(num_features == 100)) + 
+  theme_bw() +
+  geom_point(aes(num_cells, time_elapsed)) +
+  scale_x_continuous(limits = c(0, 10000)) +
+  scale_y_continuous(limits = c(0, 300)) +
+  geom_abline(intercept = 22.097, slope = 0.0252) +
+  labs(x = "Number of cells", y = "Execution time (s)")
+```
+
+![](scalability_and_runtime_files/figure-gfm/plot_timings_cell-1.png)<!-- -->
+
+### Increasing the number of features
+
+By increasing the number of features from 1000 to 10’000 whilst keeping
+the number of cells fixed, we can get an idea of how the simulation time
+scales w.r.t. the number of features
+
+``` r
+timings2 <- 
+  timings_gr %>% 
+  filter(num_cells == 100) %>% 
+  group_by(num_cells, num_features, group) %>% 
+  summarise(time_elapsed = sum(time_elapsed), .groups = "drop")
+
+ggplot(timings2) + 
+  geom_bar(aes(x = forcats::fct_inorder(as.character(num_features)), y = time_elapsed, fill = forcats::fct_inorder(group)), stat = "identity") +
+  theme_classic() +
+  scale_fill_brewer(palette = "Dark2") +
+  labs(x = "Number of features", y = "Average time (s)", fill = "dyngen step")
+```
+
+![](scalability_and_runtime_files/figure-gfm/figure2-1.png)<!-- -->
+
+It seems the execution time also scales linearly w.r.t. the number of
+features. As more genes are added to the underlying gene regulatory
+network, the density of the graph doesn’t change, so it makes sense that
+the execution time also scales linearly w.r.t. the number of features.
+
+``` r
+ggplot(timings_sum %>% filter(num_cells == 100)) + 
+  theme_bw() +
+  geom_point(aes(num_features, time_elapsed)) +
+  scale_x_continuous(limits = c(0, 10000)) +
+  scale_y_continuous(limits = c(0, 850)) +
+  geom_abline(intercept = 0.5481, slope = 0.07988) +
+  labs(x = "Number of features", y = "Execution time (s)")
+```
+
+![](scalability_and_runtime_files/figure-gfm/plot_timings_feats-1.png)<!-- -->
+
+### Execution platform
+
+These timings were measured using 30 (out of 32) threads using a AMD
+Ryzen 9 5950X clocked at 3.4GHz.

### 2021-03-29 manuscript/v3/render.Rmd
add submission files

--- 2020-06-13 render.Rmd
@@
 - Yvan Saeys
-date: "`r format(Sys.time(), '%d %B %Y')`"
+date: "14 September 2020"
 output:
@@
     \centering
-    \includegraphics[width=\linewidth]{result_files/usecase_rna_velocity/usecase.pdf}
+    \includegraphics[width=\linewidth]{result_files/usecase_rna_velocity/supp_fig.pdf}
     \caption{
@@
     \centering
-    \includegraphics[width=.8\linewidth]{result_files/usecase_network_inference/cni.pdf}
+    \includegraphics[width=.8\linewidth]{result_files/usecase_network_inference/supp_fig.pdf}
     \caption{

### 2021-03-29 manuscript/vignettes.Rmd
add submission files

--- 2021-03-25 manuscript/render.Rmd
@@
----
-title: "Response to reviewers' comments"
-output:
-  pdf_document:
-    keep_tex: yes
-    latex_engine: xelatex
-    citation_package: biblatex
-biblatexoptions: sorting=none,url=false
-classoption:
-- table
-- 10pt
-- a4paper
-header-includes: |
-  \usepackage{tcolorbox}
-  \usepackage{colortbl}
-  \usepackage{booktabs}
-  \usepackage{tabularx}
-  \usepackage{fontspec}
-  \usepackage{pifont}
-  \usepackage{float}
-  \usepackage{caption}
-  \captionsetup[table]{position=bottom}
-  \setmainfont [Path = fonts/,
-    UprightFont = *-300,
-    ItalicFont = *-300-Italic,
-    BoldFont = *-700,
-    BoldItalicFont = *-700-Italic
-  ]{MuseoSans}
-
-bibliography: library.bib
----
-<!-- Notes (remove this before submitting)
-dyngen manuscript v4  
-github.com/dynverse/dyngen  
-github.com/dynverse/dyngen_manuscript
-dyngen.dynverse.org
-
-Antwoord naar editor: wordt gebruikt in dynverse, tradeseq, cellrank, BEELINE
--->
-
-# dyngen rebuttal
-
-We would like to thank the reviewer for the constructive comments, which has motivated us to greatly improve the experiments performed in this work and the documentation provided for the tool.
-Notable changes are:
-
- * We created a documentation website available at [dyngen.dynverse.org](https://dyngen.dynverse.org). [a]Several insights 
- * We simplified and reran all use-case experiments, which allowed us to add some interpretation of the results for each of the use cases.
- * We performed new experiments assessing the scalability of dyngen and the similarity in output in comparison to a reference (real) dataset. The results are included in the supplementary material.
- As we think that several of the remarks raised by the reviewer are also interesting for the target audience of the paper, we not only include our responses to these remarks here, but in addition have created several tutorials that can be used to perform these experiments with dyngen.
-These tutorials now also allow users to reproduce some of the additional analyses requested by the reviewer, and they have been added to a new website that we created for our method, with a lot more documentation website ([dyngen.dynverse.org](https://dyngen.dynverse.org)).
-* … todo, add more ...[b]
-
-## Comment 1
-> Bioinformatics has been looking into GRNs, their simulation and estimation for more than a decade (in bulk at the time), and I miss some important references and insights from then.
-Such as alternate definitions of what a GRN is - eg does logic play a role, should you use boolean networks (see Albert&Othma, JTB 2003) -, do old-school bulk simulators such as GeneNetWeaver (Bioinformatics 2011) be used, can one use transformation methods (eg Wittmann et al, BMC Sysbio 2009) and how is this different from what has been introduced here (what is novelty here).
-
-Certainly simulating expression datasets using GRNs has been around for a long time and dyngen is inspired heavily by such software (GeneNetWeaver in particular).
-Nevertheless, there are great differences between bulk and single-cell simulators:  the data characteristics of single-cell omics are vastly different from bulk omics, typically having much lower library sizes and a higher dropout rate, but also a high number of profiles.
-The low library sizes, in particular, are problematic as ODEs are ill-suited for performing low-molecule simulations.
-This necessitates the development of new single-cell simulators.
-
-We have added paragraph §3 to the introduction to discuss the relevance of bulk simulators and why they are not suitable for simulating single-cell expression data:
-
-> Simulating realistic data requires that the underlying biology is recapitulated as best as possible, and in the case of transcriptomics data this typically involves modelling the underlying gene regulatory networks.
-Simulators of "bulk" microarray or RNA-sequencing profiles simulate biological processes (e.g.
-transcription, translation) by translating a database of known regulatory interactions into a set of ordinary differential equations (ODE) [@roy_systemgeneratingtranscription_2008; @hache_gengesystematicgeneration_2009; @schaffter_genenetweaversilicobenchmark_2011; @vandenbulcke_syntrengeneratorsynthetic_2006].
-These methods have been instrumental in performing benchmarking studies [@prill_rigorousassessmentsystems_2010; @marbach_revealingstrengthsweaknesses_2010; @marbach_wisdomcrowdsrobust_2012].
-However, the advent of single-cell omics introduced several new types of analyses (e.g.
-trajectory inference, RNA velocity, cell-specific network inference) which exploit the higher resolution of single-cell versus bulk omics [@luecken_currentbestpractices_2019].
-In addition, the data characteristics of single-cell omics are vastly different from bulk omics, typically having much lower library sizes and a higher dropout rate, but also a high number of profiles [@vallejos_normalizingsinglecellrna_2017].
-The low library sizes, in particular, are problematic as ODEs are ill-suited for performing low-molecule simulations [@gillespie_exactstochasticsimulation_1977].
-This necessitates the development of new single-cell simulators.
-
-## Comment 2
-> Similarly, the single cell genomics field has used simulations also for dynamics for some time as evaluation at least implicitely.
-Please discuss advances and differences or similarities to those, eg.
-to simulations from (Ocone, Bioinf 2015) or (Pratapa, Nat Meth 2020).
-
-We expanded upon paragraph §4 and made a more direct comparison to the methods listed in Supplemental Table 1. Paragraph §4:
-
-> To this end, single-cell omics simulators emulate the technical procedures from single-cell omics protocols.
-Simulators such as Splatter [@zappia_splattersimulationsinglecell_2017], powsimR [@vieth_powsimrpoweranalysis_2017], PROSSTT [@papadopoulos_prossttprobabilisticsimulation_2019] and SymSim [@zhang_simulatingmultiplefaceted_2019]) have already been widely used to compare single-cell methods [@street_slingshotcelllineage_2018; @parra_reconstructingcomplexlineage_2019; @lummertzdarocha_reconstructioncomplexsinglecell_2018; @lin_scclassifysamplesize_2020] and perform independent benchmarks [@duo_systematicperformanceevaluation_2018; @saelens_comparisonsinglecelltrajectory_2019; @soneson_biasrobustnessscalability_2018].
-However, by focusing more on simulating the single-cell omics protocol (e.g.
-RNA capture, amplification, sequencing) and less on the underlying biology (e.g.
-transcription, splicing, translation), their applicability and reusability is limited towards the specific application for which they were designed (e.g.
-benchmarking clustering or differential expression methods), and extending these tools to include additional modalities or experimental conditions is challenging.
-
-We did not include the reference to Ocone Bioinf 2015 as the advances proposed by this research were mainly related to performing Network Inference.
-In terms of generating synthetic data, the simulator used by Ocone et al.
-offered little advantage over aforementioned bulk simulators.
-We also did not include a reference to Pratapa et al., as their methodology is borrowed heavily from dyngen itself and offers little advantage over it.
-
-## Comment 3
-> One of the main aims of any simulation framework should be to capture features of real biological data.
-I’d be great if the authors could demonstrate the fidelity of the generated data by comparing it to some real datasets, e.g.
-distribution of mean gene expression and number of zero counts per gene.
-How does it perform compared to other simulation frameworks?
-This would help me as well as community to evaluate how realistic the induced gene correlations/dependencies are, and what may be missing.
-
-A vignette has been added to the package and supplementary material which uses countsimQC [@soneson_unifiedqualityverification_2018] to explore the characteristics of an dyngen simulation and the reference dataset used by it ([dyngen.dynverse.org/articles/advanced_topics/comparison_reference.html](https://dyngen.dynverse.org/articles/advanced_topics/comparison_reference.html)).
-The distribution plots in this vignette, while not perfect, nicely demonstrates that the output generated by dyngen mimics the given reference dataset for the evaluated metrics.
-
-## Comment 4
-> Simulated data for the described key applications of dyngen (network inference, trajectory alignment, and RNA velocity) has already been generated within the respective works of lioness (Fig.
-S1+S4), cellAlign (Fig.
-S6), velocyto (Fig.
-S5), scvelo (Fig.
-S3), dynamo (Fig.
-1), many of which also use Gillespie.
-What does dyngen add conceptually?
-
-We modified paragraphs §5 and §6 to better highlight the scientific contributions dyngen makes:
-
-> We introduce dyngen, a method for simulating cellular dynamics at a single-cell, single-transcript resolution (Figure\ \ref{fig:overview}).
-This problem is tackled in three fully-configurable main steps.
-First, biological processes are mimicked by translating a gene regulatory network in a set of reactions (regulation, transcription, splicing, translation).
-Second, individual cells are simulated using Gillespie's stochastic simulation algorithm [@gillespie_exactstochasticsimulation_1977], which is designed to work well in low-molecule simulations.
-Finally, real reference datasets are used to emulate single-cell omics profiling protocols.
-> 
-> Throughout a simulation, dyngen tracks many layers of information, including the abundance of any molecule in the cell, the progression of the cell along a dynamic process, and the activation strength of individual regulatory interactions.
-In addition, dyngen can simulate a large variety of dynamic processes (e.g.
-cyclic, branching, disconnected) as well as a broad range of experimental conditions (e.g.
-batch effects and time-series, perturbation and single-cell knockdown experiments).
-For these reasons, dyngen can cater to a wide range of benchmarking applications, including trajectory inference, trajectory alignment, and trajectory differential expression (Table\ \ref{tab:comparison}).
-
-To the best of our knowledge, the works listed above use very different approaches to benchmarking their tool:
-
-* LIONESS, network inference method: Uses randomly generated networks & boolean networks to generate synthetic data.
-The networks used to generate datasets are very synthetic and completely devoid of any biological relevance.
-By using Gillespie's SSA algorithm instead of boolean networks, we can use dyngen to track many layers of information (e.g.
-the number of times a biological reaction has been triggered during a given time interval).
-This allows us to benchmark methods which require this information (e.g.
-RNA velocity).
-* cellAlign, trajectory alignment: Perturbs real datasets to assess the robustness of cellAlign.
-When benchmarking a tool, assessing its robustness is almost as important as assessing its accuracy.
-Robustness analysis can be performed without knowing the gold standard, whereas the gold standard is needed for assessing the accuracy of a tool.
-* velocyto, RNA velocity: Uses ODEs to simulate splicing mechanics.
-The code for reproducing these datasets couldn't be found.
-Based on the description, these simulations were limited to simulating only the splicing mechanics under a particular set of conditions, thereby limiting its applicability outside of this particular context.
-* scvelo, RNA velocity: Same as velocyto.
-* dynamo, RNA velocity: Uses Gillespie's SSA to simulate splicing reactions.
-However, this system is very limited, including only two genes.
-This application of Gillespie's SSA seems very tailored towards this particular use-case and would require major redesigns to be made applicable elsewhere.
-
-It should be noted that from the list above, only LIONESS really compares its accuracy performance with other methods using a quantitative metric.
-We would thus argue that these cases further highlight the need for broadly applicable synthetic data generators such as dyngen.
-
-## Comment 5
-> The authors evaluate several computational tools, including the afore-mentioned.
-Even if the authors mention not to aim for a large-scale benchmarking, they still should provide some insights into why some tools perform better than others.
-Having an interpretable simulation engine in hand, it is unfortunate that it yet lacks any explanation and the reader is left alone with performance statements that are not well-founded.
-Some intuition would be very desirable.
-Why does SCENIC outperform the other methods?
-Why does smoothing improve trajectory alignment?
-And what causes differences in velocity estimation?
-Could you, for instance, generate an example where the differences in velocity estimates become more obvious, where one estimation procedure fails to fully capture the known direction?
-
-Performing in-depth benchmarks on each of these use-cases can be considered separate studies on their own and is beyond the scope of this work.
-However, we did reanalyse the experiments and added some interpretation to the respective sections.
-
- * Trajectory alignment: We removed DTW+smoothing (as it was a custom method developed for the sake of this analysis) and added cellAlign to the comparison, as cellAlign also uses smoothing and DTW to get to the same objective, and has been developed specifically with single cell data in mind.
- * RNA velocity: We removed some parameter settings which are designed to be very similar and thus have no added benefit in the visualisation.
-We removed the 'difficulty' settings for the different datasets, as the 'easy' and 'medium' datasets were unrealistic; only the 'hard' settings were retained.
-While there are no large difference in performance between the different methods, we added interpretation on why methods obtain low scores on one metric and high scores on the other.
- * Cell-specific network inference: We added information on why certain methods significantly outperform others and provide possible solutions to problems, where possible.
-
-## Minor comment 1
-> Could the authors please discuss runtime and scalability, which obviously can easily become a weakness for such a complex model.
-
-A vignette has been added to the package and supplementary material explores the execution time of dyngen for an increasing number of cells and genes ([dyngen.dynverse.org/articles/advanced_topics/scalability_and_runtime.html](https://dyngen.dynverse.org/articles/advanced_topics/scalability_and_runtime.html)).
-These figures show that the execution time of dyngen scales linearly w.r.t.
-the number of cells and also linearly w.r.t.
-the number of genes.
-
-## Minor comment 2
-> Please state how many genes have been used in the examples.
-It seems like the number of genes used in the examples is rather small (< 100).
-Wouldn’t it be better to use a more realistic number of features (~10,000)?
-
-Unfortunately, the examples shown in the package documentation need to be very small to abide by CRAN package submission policies, as the complete testing of the software package (including but not limited to running all examples) can strictly take 10 minutes on CRAN's server using only one thread.
-
-However, the scalability vignette mentioned above also shows the execution time of each of the steps taken by dyngen in simulating a 10'000 by 10'000 dataset, which takes about 20 minutes to complete.
-
-## Minor comment 3
-> Congratulations for making the clearly documented package available.
-Installation and going through the vignette works fine.
-At several points during the simulation files are downloaded without asking permission.
-It may be better practice to ask the user for permission, or download into a temporary directory, or read the files directly without downloading, or include in the package etc.
-
-By default, required files are downloaded to a temporary directory.
-Unfortunately, these files cannot be included inside the package as they are reference datasets which are much larger than what is allowed by CRAN policies.
-
-Based on the reviewers suggestion, we modified the installation instructions such that it better reflects that files will be downloaded when using dyngen, and provides instructions on how to configure the system such that it will cache these files in a user-specified directory ([dyngen.dynverse.org/articles/installation.html](https://dyngen.dynverse.org/articles/installation.html)).
-
-## Minor comment 4
-> Further, it might be very useful to output the simulated data to a standard object, such as Seurat or SingleCellExperiment, or to provide instructions on how to convert it.
-
-Thank you for this suggestion.
-We agree that conversion functions from the typical dyngen output to the typical object types familiar to single-cell omics programmers would be very useful.
-Therefore, we included such functions in the latest version of dyngen to convert its output to the following formats: anndata, dyno, SingleCellExperiment, Seurat.
-
-[a]containing what?
-[b]todo
-
 # Supplementary Files
@@
 ``` r
-
 # the simulation is being sped up because rendering all vignettes with one core
-
 # for pkgdown can otherwise take a very long time

### 2021-03-29 manuscript/render_manuscript.R

@@ -20,4 +20,5 @@ render("manuscript/render.Rmd")
 system("pdftk manuscript/render.pdf cat 1-16 output manuscript/manuscript.pdf")
 system("pdftk manuscript/render.pdf cat 17-45 output manuscript/supplementary_files.pdf")
 
+# system("latexdiff manuscript/v3/render.tex manuscript/render.tex > manuscript/track_changes.tex")
 # file.remove("manuscript/render.pdf")

### 2021-05-12 analysis/summary/source_data.R
write statistics to file

@@ -8,12 +8,15 @@ out <- list()
 # Trajectory alignment
 exp_a <- start_analysis("usecase_trajectory_alignment")
 scores_a <- read_rds(exp_a$result("results.rds"))
+stats_a <- read_rds(exp_a$result("statistics.rds"))
 
 out$trajectory_alignment_summary <- scores_a %>% select(-mean_distance) %>% rename(dataset_id = id, method_id = method)
+out$trajectory_alignment_statistics <- stats_a %>% select(-label)
 
 # RNA velocity
 exp_b <- start_analysis("usecase_rna_velocity")
 scores_b <- read_rds(exp_b$result("scores.rds"))
+stats_b <- read_rds(exp_b$result("statistics.rds"))
 
 design_names <-
   tribble(
@@ -23,16 +26,18 @@ design_names <-
     "scvelo", "dynamical", "scvelo dynamical"
   )
 
-out$RNA_velocity_percell <- scores_b$per_cell %>% inner_join(design_names, by = c("method_id", "params_id")) %>% select(-params_id, -method_id) %>% rename(method_id = method_label) %>% select(method_id, everything())
+out$RNA_velocity_perfeature <- scores_b$per_cell %>% inner_join(design_names, by = c("method_id", "params_id")) %>% select(-params_id, -method_id) %>% rename(method_id = method_label) %>% select(method_id, everything())
 out$RNA_velocity_summary <- scores_b$summ %>% inner_join(design_names, by = c("method_id", "params_id")) %>% select(-params_id, -method_id, -mean_corr) %>% rename(method_id = method_label) %>% select(method_id, everything())
+out$RNA_velocity_statistics <- stats_b %>% select(-label)
 
 # Network inference
 exp_c <- start_analysis("usecase_network_inference")
-
 scores_c <- read_rds(exp_c$result("scores.rds"))
+stats_c <- read_rds(exp_c$result("statistics.rds"))
 
 out$CSNI_percell <- scores_c$aucs %>% filter(method == "casewise_casewise") %>% select(-method, -method_label, cni_method_id) %>% rename(method_id = cni_method_name)
 out$CSNI_summary <- scores_c$summ %>% filter(method == "casewise_casewise") %>% select(-method, -method_label, cni_method_id) %>% rename(method_id = cni_method_name)
+out$CSNI_statistics <- stats_c %>% select(-label)
 
 # write output
 openxlsx::write.xlsx(out, exp$result("Source Data.xlsx"))

### 2021-05-12 analysis/usecase_network_inference/4_supp_fig.R

@@ -49,6 +49,8 @@ pairwise <-
   ungroup() %>%
   mutate(label = gsub("\\[Holm-corrected\\]", "", label))
 
+write_rds(pairwise, exp$result("statistics.rds"), compress = "gz")
+
 # create metric plots
 g_metrics <- map2(c("auroc", "aupr"), c("mean AUROC", "mean AUPR"), function(metric, metric_name) {
   data <- results %>% filter(metric == !!metric)

### 2021-05-12 analysis/usecase_rna_velocity/4_plot_usecases.R

@@ -187,6 +187,8 @@ pairwise <-
   ungroup() %>%
   mutate(label = gsub("\\[Holm-corrected\\]", "", label))
 
+write_rds(pairwise, exp$result("statistics.rds"), compress = "gz")
+
 # create metric plots
 g_metrics <- map2(names(metric_labels), metric_labels, function(metric, metric_name) {
   data <- results %>% filter(metric == !!metric)

### 2021-05-12 analysis/usecase_trajectory_alignment/3_plots.R

@@ -257,6 +257,8 @@ pairwise <-
   ungroup() %>%
   mutate(label = gsub("\\[Holm-corrected\\]", "", label))
 
+write_rds(pairwise, exp$result("statistics.rds"), compress = "gz")
+
 metric_labels <- c(abwap = "ABWAP score")
 
 # create metric plots

### 2021-05-12 manuscript/manuscript.tex
rework manuscript

% Options for packages loaded elsewhere
\PassOptionsToPackage{unicode}{hyperref}
\PassOptionsToPackage{hyphens}{url}
%
\documentclass[
  table,
  10pt,
  a4paper]{article}
\usepackage{amsmath,amssymb}
\usepackage{lmodern}
\usepackage{ifxetex,ifluatex}
\ifnum 0\ifxetex 1\fi\ifluatex 1\fi=0 % if pdftex
  \usepackage[T1]{fontenc}
  \usepackage[utf8]{inputenc}
  \usepackage{textcomp} % provide euro and other symbols
\else % if luatex or xetex
  \usepackage{unicode-math}
  \defaultfontfeatures{Scale=MatchLowercase}
  \defaultfontfeatures[\rmfamily]{Ligatures=TeX,Scale=1}
\fi
% Use upquote if available, for straight quotes in verbatim environments
\IfFileExists{upquote.sty}{\usepackage{upquote}}{}
\IfFileExists{microtype.sty}{% use microtype if available
  \usepackage[]{microtype}
  \UseMicrotypeSet[protrusion]{basicmath} % disable protrusion for tt fonts
}{}
\makeatletter
\@ifundefined{KOMAClassName}{% if non-KOMA class
  \IfFileExists{parskip.sty}{%
    \usepackage{parskip}
  }{% else
    \setlength{\parindent}{0pt}
    \setlength{\parskip}{6pt plus 2pt minus 1pt}}
}{% if KOMA class
  \KOMAoptions{parskip=half}}
\makeatother
\usepackage{xcolor}
\IfFileExists{xurl.sty}{\usepackage{xurl}}{} % add URL line breaks if available
\IfFileExists{bookmark.sty}{\usepackage{bookmark}}{\usepackage{hyperref}}
\hypersetup{
  pdftitle={dyngen: a multi-modal simulator for spearheading new single-cell omics analyses},
  pdfauthor={Robrecht Cannoodt,1,2,3; Wouter Saelens,1,2; Louise Deconinck1,2; Yvan Saeys1,2,},
  hidelinks,
  pdfcreator={LaTeX via pandoc}}
\urlstyle{same} % disable monospaced font for URLs
\usepackage[margin=1in]{geometry}
\usepackage{graphicx}
\makeatletter
\def\maxwidth{\ifdim\Gin@nat@width>\linewidth\linewidth\else\Gin@nat@width\fi}
\def\maxheight{\ifdim\Gin@nat@height>\textheight\textheight\else\Gin@nat@height\fi}
\makeatother
% Scale images if necessary, so that they will not overflow the page
% margins by default, and it is still possible to overwrite the defaults
% using explicit options in \includegraphics[width, height, ...]{}
\setkeys{Gin}{width=\maxwidth,height=\maxheight,keepaspectratio}
% Set default figure placement to htbp
\makeatletter
\def\fps@figure{htbp}
\makeatother
\setlength{\emergencystretch}{3em} % prevent overfull lines
\providecommand{\tightlist}{%
  \setlength{\itemsep}{0pt}\setlength{\parskip}{0pt}}
\setcounter{secnumdepth}{-\maxdimen} % remove section numbering
\usepackage{tcolorbox}
\usepackage{colortbl}
\usepackage{booktabs}
\usepackage{tabularx}
\usepackage{fontspec}
\usepackage{pifont}
\usepackage{float}
\usepackage{caption}
\captionsetup[table]{position=bottom}
\setmainfont [Path = fonts/,
  UprightFont = *-300,
  ItalicFont = *-300-Italic,
  BoldFont = *-700,
  BoldItalicFont = *-700-Italic
]{MuseoSans}

\ifluatex
  \usepackage{selnolig}  % disable illegal ligatures
\fi
\usepackage[sorting=none,url=false,backend=bibtex]{biblatex}

\usepackage{filecontents}

\bibliography{\jobname}

\title{dyngen: a multi-modal simulator for spearheading new single-cell
omics analyses}
\author{Robrecht Cannoodt\textsuperscript{$\dagger{}$,1,2,3} \and Wouter
Saelens\textsuperscript{$\dagger{}$,1,2} \and Louise
Deconinck\textsuperscript{1,2} \and Yvan Saeys\textsuperscript{1,2,*}}
\date{12 May 2021}

\begin{document}
\maketitle

\textsuperscript{$\dagger{}$}
These authors contributed equally to this work.

\textsuperscript{1} Data Mining and Modelling for Biomedicine group, VIB
Center for Inflammation Research, Ghent, Belgium\\
\textsuperscript{2} Department of Applied Mathematics, Computer Science,
and Statistics, Ghent University, Ghent, Belgium\\
\textsuperscript{3} Data Intuitive, Lebbeke, Belgium

\textsuperscript{*} Correspondence:
\href{mailto:yvan.saeys@ugent.be}{Yvan Saeys
\textless{}\href{mailto:yvan.saeys@ugent.be}{\nolinkurl{yvan.saeys@ugent.be}}\textgreater{}}

\hypertarget{abstract}{%
\section{Abstract}\label{abstract}}

We present dyngen, a novel, multi-modal simulation engine for studying
dynamic cellular processes at single-cell resolution. dyngen is more
flexible than current single-cell simulation engines, and allows better
method development and benchmarking, thereby stimulating development and
testing of novel computational methods. We demonstrate its potential for
spearheading novel computational methods on three novel applications:
aligning cell developmental trajectories, cell-specific regulatory
network inference and estimation of RNA velocity.

\hypertarget{introduction}{%
\section{Introduction}\label{introduction}}

Single-cell simulation engines are becoming increasingly important for
testing and benchmarking computational methods, a pressing need in the
widely expanding field of single-cell biology. Complementary to real
biological data, synthetic data provides a valuable alternative where
the actual ground truth is completely known and thus can be compared to,
in order to make quantitative evaluations of computational methods that
aim to reconstruct this ground truth
\autocite{zappia_splattersimulationsinglecell_2017}. In addition,
simulation engines are more flexible when it comes to stress-testing
computational methods, for example by varying the parameters of the
simulation, such as the amount of noise, samples, and cells measured,
allowing benchmarking of methods over a wide range of possible
scenarios. In this way, they can even guide the design of real
biological experiments, finding out the best conditions to be used as
input for subsequent computational pipelines.

Another, more experimental use of simulation engines is their important
role in spearheading the development of novel computational methods,
possibly even before real data is available. In this way, simulation
engines can be used to assess the value of novel experimental protocols
or treatments. Simulation engines are also increasingly important when
it comes to finding alternatives to animal models, for example for drug
testing and precision medicine. In such scenarios, cellular simulations
can act as digital twins, offering unlimited experimentation \emph{in
silico} \autocite{bjornsson_digitaltwinspersonalize_2019}.

Simulating realistic data requires that the underlying biology is
recapitulated as best as possible, and in the case of transcriptomics
data this typically involves modelling the underlying gene regulatory
networks. Simulators of ``bulk'' microarray or RNA-sequencing profiles
simulate biological processes (e.g.~transcription, translation) by
translating a database of known regulatory interactions into a set of
ordinary differential equations (ODE)
\autocite{roy_systemgeneratingtranscription_2008,hache_gengesystematicgeneration_2009,schaffter_genenetweaversilicobenchmark_2011,vandenbulcke_syntrengeneratorsynthetic_2006}.
These methods have been instrumental in performing benchmarking studies
\autocite{prill_rigorousassessmentsystems_2010,marbach_revealingstrengthsweaknesses_2010,marbach_wisdomcrowdsrobust_2012}.
However, the advent of single-cell omics introduced several new types of
analyses (e.g.~trajectory inference, RNA velocity, cell-specific network
inference) which exploit the higher resolution of single-cell versus
bulk omics \autocite{luecken_currentbestpractices_2019}. In addition,
the data characteristics of single-cell omics are vastly different from
bulk omics, typically having much lower library sizes and a higher
dropout rate, but also a high number of profiles
\autocite{vallejos_normalizingsinglecellrna_2017}. The low library
sizes, in particular, are problematic as ODEs are ill-suited for
performing low-molecule simulations
\autocite{gillespie_exactstochasticsimulation_1977}. This necessitates
the development of new single-cell simulators.

To this end, single-cell omics simulators emulate the technical
procedures from single-cell omics protocols. Simulators such as Splatter
\autocite{zappia_splattersimulationsinglecell_2017}, powsimR
\autocite{vieth_powsimrpoweranalysis_2017}, PROSSTT
\autocite{papadopoulos_prossttprobabilisticsimulation_2019} and SymSim
\autocite{zhang_simulatingmultiplefaceted_2019}) have already been
widely used to compare single-cell methods
\autocite{street_slingshotcelllineage_2018,parra_reconstructingcomplexlineage_2019,lummertzdarocha_reconstructioncomplexsinglecell_2018,lin_scclassifysamplesize_2020}
and perform independent benchmarks
\autocite{duo_systematicperformanceevaluation_2018,saelens_comparisonsinglecelltrajectory_2019,soneson_biasrobustnessscalability_2018}.
However, by focusing more on simulating the single-cell omics protocol
(e.g.~RNA capture, amplification, sequencing) and less on the underlying
biology (e.g.~transcription, splicing, translation), their applicability
and reusability is limited towards the specific application for which
they were designed (e.g.~benchmarking clustering or differential
expression methods), and extending these tools to include additional
modalities or experimental conditions is challenging.

We introduce dyngen, a method for simulating cellular dynamics at a
single-cell, single-transcript resolution (Figure~\ref{fig:overview}).
This problem is tackled in three fully-configurable main steps. First,
biological processes are mimicked by translating a gene regulatory
network into a set of reactions (regulation, transcription, splicing,
translation). Second, individual cells are simulated using Gillespie's
stochastic simulation algorithm
\autocite{gillespie_exactstochasticsimulation_1977}, which is designed
to work well in low-molecule simulations. Finally, real reference
datasets are used to emulate single-cell omics profiling protocols.

Throughout a simulation, dyngen tracks many layers of information,
including the abundance of any molecule in the cell, the progression of
the cell along a dynamic process, and the activation strength of
individual regulatory interactions. In addition, dyngen can simulate a
large variety of dynamic processes (e.g.~cyclic, branching,
disconnected) as well as a broad range of experimental conditions
(e.g.~batch effects and time-series, perturbation and single-cell
knockdown experiments). For these reasons, dyngen can cater to a wide
range of benchmarking applications, including trajectory inference,
trajectory alignment, and trajectory differential expression
(Supplementary Table~1).

\begin{figure}[t!]
    \centering
    \includegraphics[width=\linewidth]{result_files/summary/figure_1_edited.pdf}
    \caption{
      \textbf{Showcase of dyngen functionality.}
      \textbf{A:} Changes in abundance levels are driven strictly by gene regulatory reactions.
      \textbf{B:} The input Gene Regulatory Network (GRN) is defined such that it models a dynamic process of interest.
      \textbf{C:} The reactions define how abundance levels of molecules change at any particular time point.
      \textbf{D:} Firing many reactions can significantly alter the cellular state over time.
      \textbf{E:} dyngen keeps track of the likelihood of a reaction firing during small intervals of time, called the propensity, as well as the actual number of firings.
      \textbf{F:} Similarly, dyngen can also keep track of the regulatory activity of every interaction.
      \textbf{G:} A benchmark of trajectory inference methods has already been performed using the cell state ground-truth \cite{saelens_comparisonsinglecelltrajectory_2019}.
      \textbf{H:} The cell state ground-truth enables evaluating trajectory alignment methods.
      \textbf{I:} The reaction propensity ground-truth enables evaluating RNA velocity methods.
      \textbf{J:} The cellwise regulatory network ground-truth enables evaluating cell-specific gene regulatory network inference methods.
    }
    \label{fig:overview}
\end{figure}

\hypertarget{results}{%
\section{Results}\label{results}}

We demonstrate dyngen's broad applicability by evaluating three novel
types of computational approaches for which no simulation engines exist
yet: cell-specific network inference, trajectory alignment and RNA
velocity (Figure~\ref{fig:applications}). We emphasize that our main aim
here is to illustrate the potential of dyngen for these evaluations,
rather than performing large-scale benchmarking, which would require
assessing many more quantitative and qualitative aspects of each method
\autocite{weber_essentialguidelinescomputational_2019}.

\begin{figure}[t!]
    \centering
    \includegraphics[width=\linewidth]{result_files/summary/figure_2.pdf}
    \caption{
      \textbf{dyngen provides ground-truth data for a variety of applications (left), which can be used to quantitatively evaluate methods (right).} Box plots denote the $Q_0$ to $Q_4$ quartile values.
      \textbf{A:} Trajectory alignment aligns two trajectories between samples. We evaluate dynamic time warping (DTW) and cellAlign when aligning two linear trajectories with different kinetic parameters based on the area differences between the worst possible alignment and the predicted alignment (Area Between Worst And Prediction, or ABWAP). 
      \textbf{B:} RNA velocity calculates for each cell the direction in which the expression of each gene is moving. We evaluated scVelo and velocyto by comparing these vectors with the known velocity vector (velocity correlation) and with the known direction of the cellular trajectory in a dimensionality reduction (velocity arrow cosine).
      \textbf{C:} Cell-specific network inference (CSNI) predicts the regulatory network of every individual cell. We evaluate each cell-specific regulatory network with typical metrics for network inference: the Area Under the Receiver Operating Characteristics-curve (AUROC) and Area Under the Precision Recall-curve (AUPR). We evaluate three CSNI methods by computing the mean AUROC and AUPR across all cells.
    }
    \label{fig:applications}
\end{figure}

Use-case ``trajectory alignment''. Trajectory alignment methods align
trajectories from different samples and allow studying the differences
between the different trajectories. For example, by comparing the
transcriptomic profiles of cells from a diseased patient to a healthy
control, it might be possible to detect transcriptomics differences
(differential expression) of particular cells along a developmental
process, or to detect an early stop of the trajectory of the diseased
patient. Currently, trajectory alignment is limited to aligning linear
trajectories, though other topologies of a trajectory could be aligned
as well. Dynamic Time Warping (DTW)
\autocite{giorgino_computingvisualizingdynamic_2009} is a method
designed for aligning temporal sequences for speech recognition but has
since been used to compare gene expression kinetics from many different
biological processes
\autocite{cacchiarelli_aligningsinglecelldevelopmental_2018,kanton_organoidsinglecellgenomic_2019,mcfaline-figueroa_pooledsinglecellgenetic_2019,alpert_alignmentsinglecelltrajectories_2018}.
cellAlign \autocite{alpert_alignmentsinglecelltrajectories_2018} uses
DTW to perform trajectory alignment, but also includes interpolation and
scaling of the single cell data as a preprocessing step. We evaluate the
performance of DTW and cellAlign by simulating 40 datasets, each
containing two linear trajectories generated with the same gene
regulatory network but with slightly different simulation kinetics. We
assess the accuracy of the obtained alignments by comparing the
generated alignment path with the worst possible alignment that could be
performed (Supplementary Fig. ~1D), named the Area Between Worst And
Prediction (ABWAP) score. Overall, cellAlign performs significantly
better than DTW (Supplementary Fig.~1), which is likely due to the
interpolation and scaling steps provided by cellAlign, reducing noise in
the data and improving the comparability of the trajectories. Note that,
in this comparison, only linear trajectory alignment is performed. While
dyngen can generate non-linear trajectories (e.g.~cyclic or branching),
both aligning non-linear trajectories and constructing a quantitative
accuracy metric for non-linear trajectory alignment is not trivial and
an avenue for future work.

Use-case ``RNA velocity''. RNA velocity methods use the relative ratio
between pre-mRNA and mature mRNA reads to predict the rate of
increase/decrease of RNA molecule abundance, as this can be used to
predict the directionality of single cell differentiation in
trajectories
\autocite{zeisel_coupledpremrnamrna_2011,lamanno_rnavelocitysingle_2018}.
Already two algorithms are currently available for estimating the RNA
velocity vector from spliced and unspliced counts: velocyto
\autocite{lamanno_rnavelocitysingle_2018} and scvelo
\autocite{bergen_generalizingrnavelocity_2020}. Yet, to date, no
quantitative assessment of their accuracy has been performed, mainly due
to the difficulty in obtaining real ground-truth data to do so. In
contrast, the ground-truth RNA velocity can be easily extracted from a
dyngen simulation, as it is possible to store the rate at which mRNA
molecules are being transcribed and degraded at any particular point in
time. We executed velocyto and scvelo (with 2 different parameter
settings, stochastic and dynamical) on 42 datasets with a variety of
backbones (including linear, bifurcating, cyclic, disconnected). We
evaluated the predictions using two metrics (Supplementary Fig.~2), one
which directly compares the predicted RNA velocity of each gene with the
ground-truth RNA velocity (called the ``velocity correlation''), and one
which compares the direction of the ground-truth trajectory embedded in
a dimensionality reduction with the average RNA velocity of cells in
that neighbourhood (called the ``velocity arrow cosine''). While both
velocyto and scvelo obtained high scores for the velocity arrow cosine
metric (overall 25th percentile = 0.606), the velocity correlation is
rather low (overall 75th percentile = 0.156). This means that predicting
the RNA velocity (i.e.~transcription rate minus the decay rate) for
particular individual genes can be challenging, but the combined
information is very informative in determining the directionality of
cell progression in the trajectory. In terms of velocity correlation, no
method performed significantly better than the other, whereas ``scvelo
stochastic'' performed slightly worse than ``scvelo dynamical'' and
velocyto in terms of velocity arrow cosine score. Note that, given that
some genes are more informative in determining the overall
directionality of cell progression, performing a feature selection
before computing the embedded dimensionality reduction might result in
significantly improved velocity arrow cosine scores.

Use-case ``Cell-specific network inference'' (CSNI). CSNI methods
predict not only which transcription factors regulate which target
genes, but also aim to identify how active each interaction is in each
of the cells, since interactions can be turned off and on depending on
the cellular state. While a few pioneering CSNI approaches have already
been developed
\autocite{aibar_scenicsinglecellregulatory_2017,kuijjer_estimatingsamplespecificregulatory_2019,liu_personalizedcharacterizationdiseases_2016},
a quantitative assessment of their performance is until now lacking.
This is not surprising, as neither real nor in silico datasets of
cell-specific or even cell-type-specific interactions exist that are
large enough so that it can be used as a ground-truth for evaluating
CSNI methods. Extracting the ground-truth dynamic network in dyngen is
straightforward though, given that we can calculate how target gene
expression would change without the regulator being present. We used
this ground-truth to compare the performance of three CSNI methods
(Supplementary Fig.~3): LIONESS
\autocite{kuijjer_estimatingsamplespecificregulatory_2019}, SSN
\autocite{liu_personalizedcharacterizationdiseases_2016} and SCENIC
\autocite{aibar_scenicsinglecellregulatory_2017}. For each dataset, we
computed the mean AUROC and AUPR scores of the individual cells.
Comparing the mean AUROC and AUPR showed that pySCENIC significantly
outperforms both LIONESS and SSN, and in turn that LIONESS significantly
outperforms SSN. The poor performance of SSN is expected, as its
methodology for predicting a cell-specific is simply computing the
difference in Pearson correlation values applied to the whole dataset
and the whole dataset minus one sample. This strategy performs poorly in
large datasets where cell correlations are high, as the removal of one
cell will not yield large differences in correlation values and will
result in mostly noise. Overall, pySCENIC almost always performs better
than LIONESS, except for a few datasets where LIONESS does manage to
obtain a higher AUROC score. However, by using a different internal
network inference (e.g.~GENIE3
\autocite{huynh-thu_inferringregulatorynetworks_2010} or pySCENIC's
GRNBoost2 \autocite{moerman_grnboost2arboretoefficient_2019}) could
significantly increase the performance obtained by LIONESS.

\hypertarget{discussion}{%
\section{Discussion}\label{discussion}}

dyngen's single-cell simulations can be used to evaluate common
single-cell omics computational methods such as clustering, batch
correction, trajectory inference, and network inference. However, the
framework is flexible enough to be adaptable to a broad range of
applications, including methods that integrate clustering, network
inference, and trajectory inference. In this respect, dyngen may promote
the development of new tools in the single-cell field similarly as other
simulators have done in the past
\autocite{schaffter_genenetweaversilicobenchmark_2011,ewing_combiningtumorgenome_2015}.
Additionally, one could anticipate technological developments in
single-cell multi-omics. In this way, dyngen allows designing and
evaluating the performance and robustness of new types of computational
analyses before experimental data becomes available, comparing which
experimental protocol is the most cost-effective in producing
qualitative and robust results in downstream analysis. One major
assumption of dyngen is that cells are assumed to be well-mixed and
independent from each other. Subdividing a cell into multiple 2D or 3D
subvolumes or allowing cells to exchange molecules, respectively, could
pave the way to better study key cellular processes such as cell
division, intercellular communication, and migration
\autocite{smith_spatialstochasticintracellular_2019}.

\newpage

\hypertarget{sec:dyngen-methods}{%
\section{Methods}\label{sec:dyngen-methods}}

The workflow to generate \emph{in silico} single-cell data consists of
six main steps (Supplementary Fig~4).

\hypertarget{sec:dyngen-modules}{%
\subsection{Defining the module network}\label{sec:dyngen-modules}}

One of the main processes involved in cellular dynamic processes is gene
regulation, where regulatory cascades and feedback loops lead to
progressive changes in expression and decision making. The exact way a
cell chooses a certain path during its differentiation is still an
active research field, although certain models have already emerged and
been tested \emph{in vivo}. One driver of bifurcation is mutual
antagonism, where two genes strongly repress each other
\autocite{rekhtman_directinteractionhematopoietic_1999,xu_regulationbifurcatingcell_2015},
forcing one of the two to become inactive
\autocite{graf_forcingcellschange_2009}. Such mutual antagonism can be
modelled and simulated
\autocite{wang_quantifyingwaddingtonlandscape_2011,ferrell_bistabilitybifurcationswaddington_2012}.
Although the two-gene model is simple and elegant, the reality is
frequently more complex, with multiple genes (grouped into modules)
repressing each other \autocite{yosef_dynamicregulatorynetwork_2013}.

To start a dyngen simulation, the user needs to define a module network.
The module network describes how sets of genes regulate each other and
is what mainly determines which dynamic processes occur within the
simulated cells.

A module network consists of modules connected together by regulatory
interactions, which can be either up- or down-regulating. A module may
have basal expression, which means genes in this module will be
transcribed without the presence of transcription factor molecules. A
module marked as ``active during the burn phase'' means that this module
will be allowed to generate expression of its genes during an initial
warm-up phase (See section \ref{sec:dyngen-simcell}). At the end of the
dyngen process, cells will not be sampled from the burn phase
simulations. Interactions between modules have a strength (which is a
positive integer) and an effect (+1 for upregulating, -1 for
downregulating).

Several examples of module networks are given in Supplementary Fig.~5. A
simple chain of modules (where one module upregulates the next) results
in a \emph{linear} process. By having the last module repress the first
module, the process becomes \emph{cyclic}. Two modules repressing each
other is the basis of a \emph{bifurcating} process, though several
chains of modules have to be attached in order to achieve progression
before and after the bifurcation process. Finally, a \emph{converging}
process has a bifurcation occurring during the burn phase, after which
any differences in module regulation is removed.

Note that these examples represent the bare minimum in terms of the
number of modules used. Using longer chains of modules is typically
desired. In addition, the fate decisions made in this example of a
bifurcation is reversible, meaning cells can be reprogrammed to go down
a different differentiation path. If this effect is undesirable, more
safeguards need to be put in place to prevent reprogramming from
occurring.

\hypertarget{sec:dyngen-grn}{%
\subsection{Generating the gene regulatory
network}\label{sec:dyngen-grn}}

The GRN is generated based on the given module network in four main
steps (Supplementary Fig.~6).

Step 1, sampling the transcription factors (TF). The TFs are the main
drivers of the molecular changes in the simulation. The user provides a
backbone and the number of TFs to generate. Each TF is assigned to a
module such that each module has at least \(x\) parameters (default
\(x=1\)). A TF inherits the `burn' and `basal expression' from the
module it belongs to.

Step 2, generating the TF interactions. Let each TF be regulated
according to the interactions in the backbone. These interactions
inherit the effect, strength, and independence parameters from the
interactions in the backbone. A TF can only be regulated by other TFs or
itself.

Step 3, sampling the target subnetwork. A user-defined number of target
genes are added to the GRN. Target genes are regulated by a TF or
another target gene, but are always downstream of at least one TF. To
sample the interactions between target genes, one of the many FANTOM5
\autocite{lizio_gatewaysfantom5promoter_2015} GRNs is sampled. The currently
existing TFs are mapped to regulators in the FANTOM5 GRN. The targets
are drawn from the FANTOM5 GRN weighted by their page rank value, to
create an induced GRN. For each target, at most \(x\) regulators are
sampled from the induced FANTOM5 GRN (default \(x=5\)). The interactions
connecting a target gene and its regulators are added to the GRN.

Step 4, sampling the housekeeping subnetwork. Housekeeping genes are
completely separate from any TFs or target genes. A user-defined set of
housekeeping genes is also sampled from the FANTOM5 GRN. The
interactions of the FANTOM5 GRN are first subsampled such that the
maximum in-degree of each gene is \(x\) (default \(x=5\)). A random gene
is sampled and a breadth-first-search is performed to sample the desired
number of housekeeping genes.

\hypertarget{sec:dyngen-reactions}{%
\subsection{Convert gene regulatory network to a set of
reactions}\label{sec:dyngen-reactions}}

\newcommand{\x}[1]{\text{x}_{#1}}
\newcommand{\y}[1]{\text{y}_{#1}}
\newcommand{\z}[1]{\text{z}_{#1}}

\newcommand{\rs}[1]{\text{R}_{#1}}
\newcommand{\rp}[1]{\text{R}^+_{#1}}
\newcommand{\rn}[1]{\text{R}^-_{#1}}

\newcommand{\xpr}[1]{\text{xpr}_{#1}}
\newcommand{\xhl}[1]{\text{xhl}_{#1}}
\newcommand{\ysr}[1]{\text{ysr}_{#1}}
\newcommand{\yhl}[1]{\text{yhl}_{#1}}
\newcommand{\ydr}[1]{\text{ydr}_{#1}}
\newcommand{\zpr}[1]{\text{zpr}_{#1}}
\newcommand{\zhl}[1]{\text{zhl}_{#1}}
\newcommand{\zdr}[1]{\text{zdr}_{#1}}

\newcommand{\str}[1]{\text{str}_{#1}}
\newcommand{\hill}[1]{\text{hill}_{#1}}
\newcommand{\ind}[1]{\text{ind}_{#1}}
\newcommand{\dis}[1]{\text{dis}_{#1}}
\newcommand{\buf}[1]{\text{bind}_{#1}}
\newcommand{\ba}[1]{\text{bas}_{#1}}

Simulating a cell's GRN makes use of a stochastic framework which tracks
the abundance levels of molecules over time in a discrete quantity. For
every gene \(G\), the abundance levels of three molecules are tracked,
namely of corresponding pre-mRNAs, mature mRNAs and proteins, which are
represented by the terms \(\text{x}_{G}\), \(\text{y}_{G}\) and
\(\text{z}_{G}\) respectively. The GRN defines how a reaction affects
the abundance levels of molecules and how likely it will occur. Gibson
and Bruck \autocite{gibson_probabilisticmodelprokaryotic_2000} provide a
good introduction to modelling gene regulation with stochastic
frameworks, on which many of the concepts below are based.

For every gene in the GRN a set of reactions are defined, namely
transcription, splicing, translation, and degradation. Each reaction
consists of a propensity function -- a formula \(f(.)\) to calculate the
probability \(f(.) \times \text{d}t\) of it occurring during a time
interval \(\text{d}t\) -- and the effect -- how it will affect the
current state if triggered.

The effects of each reaction mimic the respective biological processes
(Supplementary Table~2, middle). Transcription of gene \(G\) results in
the creation of a single pre-mRNA molecule \(\text{x}_{G}\). Splicing
turns one pre-mRNA \(\text{x}_{G}\) into a mature mRNA \(\text{x}_{G}\).
Translation uses a mature mRNA \(\text{y}_{G}\) to produce a protein
\(\text{z}_{G}\). Pre-mRNA, mRNA and protein degradation results in the
removal of a \(\text{x}_{G}\), \(\text{y}_{G}\), and \(\text{z}_{G}\)
molecule, respectively.

The propensity of all reactions except transcription are all linear
functions (Supplementary Table~2, right) of the abundance level of some
molecule multiplied by a per-gene constant (Supplementary Table~3). The
propensity of transcription of a gene \(G\) depends on the abundance
levels of its TFs. The per-gene and per-interaction constants are based
on the median reported production-rates and half-lives of molecules
measured of 5000 mammalian genes
\autocite{schwanhausser_globalquantificationmammalian_2011}, except that
the transcription rate has been amplified by a factor of 10.

\newcommand{\proptran}{f}
\newcommand{\ai}[2]{$S_{#1} = S_{#2b}$}
\newcommand{\zk}[1]{\frac{y_#1}{k_#1}^{c_#1}}
\newcommand{\wi}[1]{\nu_#1}

The propensity of the transcription of a gene \(G\) is inspired by
thermodynamic models of gene regulation
\autocite{schilstra_biologicgeneexpression_2008}, in which the promoter
of \(G\) can be bound or unbound by a set of \(N\) transcription factors
\(H_i\). Let \(f(\text{z}_{1}, \text{z}_{2}, \ldots, \text{z}_{N})\)
denote the propensity function of \(G\), in function of the abundance
levels of the transcription factors. The following subsections explain
and define the propensity function when \(N=1\), \(N=2\), and finally
for an arbitrary \(N\).

\hypertarget{propensity-of-transcription-when-n1}{%
\subsubsection{\texorpdfstring{Propensity of transcription when
\(N=1\)}{Propensity of transcription when N=1}}\label{propensity-of-transcription-when-n1}}

In the simplest case when \(N=1\), the promoter can be in one of two
states. In state \(S_0\), the promoter is not bound by any transcription
factors, and in state \(S_1\) the promoter is bound by \(H_1\). Each
state \(S_j\) is linked with a relative activation \(\alpha_j\), a
number between 0 and 1 representing the activity of the promoter at this
particular state. The propensity function is thus equal to the expected
value of the activity of the promoter multiplied by the pre-mRNA
production rate of \(G\).

\begin{align}
  f(y_1, y_2, \ldots, y_N) & = \text{xpr} \cdot \sum_{j = 0}^{2^N - 1} \alpha_j \cdot P(S_j) \label{eqn:activ0} 
\end{align}

For \(N=1\), \(P(S_1)\) is equal to the Hill equation, where \(k_i\)
represents the concentration of \(H_i\) at half-occupation and \(n_i\)
represents the Hill coefficient. Typically, \(n_i\) is between
{[}1,10{]}

\begin{align}
  P(S_1) & = \frac{y_1^{n_1}}{k_1^{n_1} + y_1^{n_1}} \\
             & = \frac{(y_1/k_1)^{n_1}}{1 + (y_1/k_1)^{n_1}}
\end{align}

The Hill equation can be simplified by letting
\(\nu_i = \left(\frac{y_i}{k_i}\right)^{n_i}\).

\begin{align}
P(S_1) & = \frac{\nu_1}{1 + \nu_1} \label{eqn:hillsimp}
\end{align}

Since \(P(S_0) = 1 - P(S_1)\), the activation function is formulated and
simplified as follows.

\begin{align}
f(y_1) & = \text{xpr} \cdot \left(\alpha_0 \cdot P(S_0) + \alpha_1 \cdot P(S_1)\right) \\
           & = \text{xpr} \cdot \left(\alpha_0 \cdot \frac{1}{1 + \nu_1} + \alpha_1 \cdot \frac{\nu_1}{1 + \nu_1}\right) \\
           & = \text{xpr} \cdot \frac{\alpha_0 + \alpha_1 \cdot \nu_1}{1 + \nu_1} \\
\end{align}

\hypertarget{propensity-of-transcription-when-n2}{%
\subsubsection{\texorpdfstring{Propensity of transcription when
\(N=2\)}{Propensity of transcription when N=2}}\label{propensity-of-transcription-when-n2}}

When \(N=2\), there are four states \(S_j\). The relative activations
\(\alpha_j\) can be defined such that \(H_1\) and \(H_2\) are
independent (additive) or synergistic (multiplicative). In order to
define the propensity of transcription \(f(.)\), the Hill equation
\(P(S_j)\) is extended for two transcription factors.

Let \(w_j\) be the numerator of \(P(S_j)\), defined as the product of
all transcription factors bound in that state:

\begin{align}
w_0 & = 1 \\
w_1 & = \nu_1 \\
w_2 & = \nu_2 \\
w_3 & = \nu_1 \cdot \nu_2
\end{align}

The denominator of \(P(S_j)\) is then equal to the sum of all \(w_j\).
The probability of state \(S_j\) is thus defined as:

\begin{align}
    P(S_j) & = \frac{w_j}{\sum_{j=0}^{j < 2^N} w_j} \\
               & = \frac{w_j}{1 + \nu_1 + \nu_2 + \nu_1 \cdot \nu_2} \\
               & = \frac{w_j}{\prod_{i=1}^{i \leq N} (\nu_i + 1)}
\end{align}

Substituting \(P(S_j)\) and \(w_j\) into \(f(.)\) results in the
following equation:

\begin{align}
f(y_1, y_2) & = \text{xpr} \cdot \sum_{j = 0}^{2^N - 1} \alpha_j \cdot P(S_j) \\
 & = \text{xpr} \cdot \frac{\sum_{j = 0}^{2^N - 1} \alpha_j \cdot w_j}{\prod_{i=1}^{i \leq N} (\nu_i + 1)} \\
 & = \text{xpr} \cdot \frac{\alpha_0 + \alpha_1 \cdot \nu_1 + \alpha_2 \cdot \nu_2 + \alpha_3 \cdot \nu_1 \cdot \nu_2}{(\nu_1 + 1) \cdot (\nu_2 + 1)} \\
\end{align}

\hypertarget{propensity-of-transcription-for-an-arbitrary-n}{%
\subsubsection{\texorpdfstring{Propensity of transcription for an
arbitrary
\(N\)}{Propensity of transcription for an arbitrary N}}\label{propensity-of-transcription-for-an-arbitrary-n}}

For an arbitrary \(N\), there are \(2^N\) states \(S_j\). The relative
activations \(\alpha_j\) can be defined such that \(H_1\) and \(H_2\)
are independent (additive) or synergistic (multiplicative). In order to
define the propensity of transcription \(f(.)\), the Hill equation
\(P(S_j)\) is extended for \(N\) transcription factors.

Let \(w_j\) be the numerator of \(P(S_j)\), defined as the product of
all transcription factors bound in that state:

\begin{align}
  w_j & = \prod_{i=1}^{i \leq N} (j \text{ mod } i) = 1 \text{ ? } \nu_i \text{ : } 1
\end{align}

The denominator of \(P(S_j)\) is then equal to the sum of all \(w_j\).
The probability of state \(S_j\) is thus defined as:

\begin{align}
P(S_j) & = \frac{w_j}{\sum_{j=0}^{j < 2^N} w_j} \\
& = \frac{w_j}{\prod_{i=1}^{i \leq N} (\nu_i + 1)}
\end{align}

Substituting \(P(S_j)\) into \(f(.)\) yields:

\begin{align}
f(y_1, y_2, \ldots, y_N) & = \text{xpr} \cdot \sum_{j = 0}^{2^N - 1} \alpha_j \cdot P(S_j) \\
& = \text{xpr} \cdot \frac{\sum_{j = 0}^{2^N - 1} \alpha_j \cdot w_j}{\prod_{i=1}^{i \leq N} (\nu_i + 1)} \label{eqn:prop2n}
\end{align}

\hypertarget{propensity-of-transcription-for-a-large-n}{%
\subsubsection{\texorpdfstring{Propensity of transcription for a large
\(N\)}{Propensity of transcription for a large N}}\label{propensity-of-transcription-for-a-large-n}}

For large values of \(N\), computing \(f(.)\) is practically infeasible
as it requires performing \(2^N\) summations. In order to greatly
simplify \(f(.)\), \(\alpha_j\) could be defined as 0 when one of the
regulators inhibits transcription and 1 otherwise.

\begin{equation}
\alpha_j = \begin{cases}
 0 & \text{ if } \exists i : j \text{ mod } i = 1 \text{ and } H_i \text{ represses } G \\
 1 & \text{otherwise}
\end{cases} \label{eqn:assalpha}
\end{equation}

Substituting equation \ref{eqn:assalpha} into equation \ref{eqn:prop2n}
and defining \(R = \{1, 2, \ldots, N\}\) and
\(R^+ = \{i | H_i \text{ activates } G\}\) yields the simplified
propensity function:

\begin{align}
f(y_1, y_2, \ldots, y_N) & = \text{xpr} \cdot \frac{\prod_{i \in R^+} (\nu_i + 1)}{\prod_{i \in R} (\nu_i + 1)}
\end{align}

\hypertarget{independence-synergism-and-basal-expression}{%
\subsubsection{Independence, synergism and basal
expression}\label{independence-synergism-and-basal-expression}}

The definition of \(\alpha_j\) as in equation \ref{eqn:assalpha}
presents two main limitations. Firstly, since \(\alpha_0 = 1\), it is
impossible to tweak the propensity of transcription when no
transcription factors are bound. Secondly, it is not possible to tweak
the independence and synergism of multiple regulators.

Let \(\text{ba} \in [0,1]\) denote the basal expression strength \(G\)
(i.e.~how much will \(G\) be expressed when no transcription factors are
bound), and \(\text{sy} \in [0,1]\) denote the synergism of regulators
\(H_i\) of \(G\), the transcription propensity becomes:

\begin{align}
f(y_1, y_2, \ldots, y_N) & = \text{xpr} \cdot \frac{\text{ba} - \text{sy}^{|R^+|} + \prod_{i \in R^+} (\nu_i + \text{sy})}{\prod_{i \in R} (\nu_i + 1)}
\end{align}

\hypertarget{sec:dyngen-simcell}{%
\subsection{Simulate single cells}\label{sec:dyngen-simcell}}

dyngen uses Gillespie's stochastic simulation algorithm (SSA)
\autocite{gillespie_exactstochasticsimulation_1977} to simulate dynamic
processes. An SSA simulation is an iterative process where at each
iteration one reaction is triggered.

Each reaction consists of its propensity -- a formula to calculate the
probability of the reaction occurring during an infinitesimal time
interval -- and the effect -- how it will affect the current state if
triggered. Each time a reaction is triggered, the simulation time is
incremented by
\(\tau = \frac{1}{\sum_j \textrm{prop}_j} \ln\left(\frac{1}{r}\right)\),
with \(r \in U(0, 1)\) and \(\textrm{prop}_j\) the propensity value of
the \(j\)th reaction for the current state of the simulation.

GillespieSSA2 is an optimised library for performing SSA simulations.
The propensity functions are compiled to C++ and SSA approximations can
be used which allow triggering many reactions simultaneously at each
iteration. The framework also allows storing the abundance levels of
molecules only after a specific interval has passed since the previous
census. By setting the census interval to 0, the whole simulation's
trajectory is retained but many of these time points will contain very
similar information. In addition to the abundance levels, also the
propensity values and the number of firings of each of the reactions at
each of the time steps can be retained.

\hypertarget{sec:dyngen-experiment}{%
\subsection{Simulate experiment}\label{sec:dyngen-experiment}}

From the SSA simulation we obtain the abundance levels of all the
molecules at every state. We need to replicate technical effects
introduced by experimental protocols in order to obtain data that is
similar to real data. For this, the cells are sampled from the
simulations and molecules are sampled for each of the cells. Gene
capture rates and library sizes are empirically derived from real
datasets as to match real technical variation.

\hypertarget{sample-cells}{%
\subsubsection{Sample cells}\label{sample-cells}}

In this step, \(N\) cells are sampled from the simulations. Two
approaches are implemented: sampling from an unsynchronised population
of single cells (snapshot) or sampling at multiple time points in a
synchronised population (time series).

\textbf{Snapshot} The backbone consists of several states linked
together by transition edges with length \(L_i\), to which the different
states in the different simulations have been mapped (Supplementary
Fig.~7A). From each transition, \(N_i = N / \frac{L_i}{\sum L_i}\) cells
are sampled uniformly, rounded such that \(\sum N_i = N\).

\textbf{Time series} Assuming that the final time of the simulation is
\(T\), the interval \([0, T]\) is divided into \(k\) equal intervals of
width \(w\) separated by \(k-1\) gaps of width \(g\). \(N_i = N / k\)
cells are sampled uniformly from each interval (Supplementary Fig.~7B),
rounded such that \(\sum N_i = N\). By default, \(k = 8\) and
\(g = 0.75\). For usual dyngen simulations, \(10 \leq T \leq 20\). For
larger values of \(T\), \(k\) and \(g\) should be increased accordingly.

\hypertarget{sample-molecules}{%
\subsubsection{Sample molecules}\label{sample-molecules}}

Molecules are sampled from the simulation to replicate how molecules are
experimentally sampled. A real dataset is downloaded from a repository
of single-cell RNA-seq datasets
\autocite{cannoodt_singlecellomicsdatasets_2018}. For each \emph{in
silico} cell \(i\), draw its library size \(ls_i\) from the distribution
of transcript counts per cell in the real dataset. The capture rate
\(cr_j\) of each \emph{in silico} molecule type \(j\) is drawn from
\(N(1, 0.05)\). Finally, for each cell \(i\), draw \(ls_i\) molecules
from the multinomial distribution with probabilities
\(cr_j \times ab_{i,j}\) with \(ab_{i,j}\) the molecule abundance level
of molecule \(j\) in cell \(i\).

\hypertarget{sec:dyngen-batcheffect}{%
\subsection{Simulating batch effects}\label{sec:dyngen-batcheffect}}

Simulating batch effects can be performed in multiple ways. One such way
is to perform the first two steps of the creation of a dyngen model
(defining the module network and generating the GRN). For each desired
batch, create a separate model for which random kinetics are generated
and perform all subsequent dyngen steps (convert to reactions, simulate
gold standard, simulate single cells, simulate experiment). Since each
separate model has different underlying kinetics, the combined output
will resemble having batch effects.

\hypertarget{sec:dyngen-groundtruth}{%
\subsection{Determining the ground-truth
trajectory}\label{sec:dyngen-groundtruth}}

To construct the ground-truth trajectory, the user needs to provide the
ground-truth state network alongside the initial module network
(Supplementary Fig.~8). Each edge in the state network specifies which
modules are allowed to change in expression in transitioning from one
state to another. For each edge, a simulation is run using the end state
of an upstream branch as the initial expression vector, and only
allowing the modules as predefined by the attribute to change.

As an example, consider the cyclic trajectory shown in Supplementary
Fig.~8. State S0 begins with an expression vector of all zero values. To
simulate the transition from S0 to S1, regulation of the genes in
modules A, B and C are turned on. After a predefined period of time, the
end state of this transition is considered the expression vector of
state S1. To simulate the transition from S1 to S2, regulation of the
genes in modules D and E are turned on, while the regulation of genes in
module C is turned off. During this simulation, the expression of genes
in modules A, B, D, and E is thus allowed to change. The end state of
the simulation is considered the expression vector of state S2.

For each of the branches in the state network, an expression matrix and
the corresponding progression time along that branch are retained. To
map a simulated cell to the ground-truth, the correlation between its
expression values and the expression matrix of the ground-truth
trajectory is calculated, and the cell is mapped to the position in the
ground-truth trajectory that has the highest correlation.

\hypertarget{sec:dyngen-extractgrn}{%
\subsection{Determining the cell-specific ground-truth regulatory
network}\label{sec:dyngen-extractgrn}}

Calculating the regulatory effect of a regulator \(R\) on a target \(T\)
(Supplementary Fig.~4F) requires determining the contribution of \(R\)
in the propensity function of the transcription of \(T\)
(section~\ref{sec:dyngen-reactions}) with respect to other regulators.
This information is useful, amongst others, for benchmarking
cell-specific network inference methods.

The regulatory effect of \(R\) on \(T\) at a particular state \(S\) is
defined as the change in the propensity of transcription when \(R\) is
set to zero, scaled by the inverse of the pre-mRNA production rate of
\(T\). More formally:

\begin{eqnarray*}
  \text{regeffect}_G & = \frac{\text{proptrans}_G(S) - \text{proptrans}_G(S[\text{z}_{T} \leftarrow 0])}{\text{xpr}_{G}}
\end{eqnarray*}

Determining the regulatory effect for all interactions and cells in the
dataset yields the complete cell-specific ground-truth GRN. The
regulatory effect lies between \([-1, 1]\), where -1 represents complete
inhibition of \(T\) by \(R\), 1 represents maximal activation of \(T\)
by \(R\), and 0 represents inactivity of the regulatory interaction
between \(R\) and \(T\).

\hypertarget{sec:dyngen-nicompare}{%
\subsection{Comparison of cell-specific network inference
methods}\label{sec:dyngen-nicompare}}

42 datasets were generated using the 14 different predefined backbones
and three different seeds. For every cell in the dataset, the
transcriptomics profile and the corresponding cell-specific ground-truth
regulatory network was determined (Section~\ref{sec:dyngen-extractgrn}).

We selected three cell-specific NI methods: SCENIC
\autocite{aibar_scenicsinglecellregulatory_2017}, LIONESS
\autocite{kuijjer_estimatingsamplespecificregulatory_2015,kuijjer_estimatingsamplespecificregulatory_2019},
and SSN \autocite{liu_personalizedcharacterizationdiseases_2016}.

LIONESS \autocite{kuijjer_estimatingsamplespecificregulatory_2019} runs
a NI method multiple times to construct cell-specific GRNs. LIONESS
first infers a GRN with all of the samples. A second GRN is inferred
with all samples except one particular profile. The cell-specific GRN
for that particular profile is defined as the difference between the two
GRN matrices. This process is repeated for all profiles, resulting in a
cell-specific GRN. By default, LIONESS uses PANDA
\autocite{glass_passingmessagesbiological_2013} to infer GRNs, but since
dyngen does not produce motif data and motif data is required by PANDA,
PANDA is inapplicable in this context. Instead, we used the lionessR
\autocite{kuijjer_lionessrsinglesample_2019} implementation of LIONESS,
which uses by default the Pearson correlation as a NI method. We marked
results from this implementation as ``LIONESS + Pearson''.

SSN \autocite{liu_personalizedcharacterizationdiseases_2016} follows, in
essence, the exact same methodology as LIONESS except that it
specifically only uses the Pearson correlation. It is worth noting that
the LIONESS preprint was released before the publication of SSN. Since
no implementation was provided by the authors, we implemented SSN in R
using basic R and tidyverse functions
\autocite{wickham_welcometidyverse_2019} and marked results from this
implementation as "SSN*".

SCENIC \autocite{aibar_scenicsinglecellregulatory_2017} consists of four
main steps. First, classical network inference is performed with
stochastic gradient boosting machines using \texttt{arboreto}
\autocite{moerman_grnboost2arboretoefficient_2019}. Second, the top 10
regulators of every target gene are selected. Interactions are grouped
together in `modules'; each module contains one regulator and all of its
targets. Next, the modules are filtered using motif analysis. Finally,
for each module and each cell, an activity score is calculated using
AUCell. As a post-processing of this output, all modules and the
corresponding activity scores are combined back into a cell-specific GRN
consisting of (cell, regulator, target, score) pairs. For this analysis,
the Python implementation of SCENIC was used, namely pySCENIC
\autocite{vandesande_scalablescenicworkflow_2020}. Since dyngen does not
generate motif data, step 3 in this analysis is skipped.

The Area Under the Receiver Operating Characteristic-curve (AUROC) and
Area Under the Precision-Recall curve (AUPR) metrics are common metrics
for evaluating a predicted GRN with a ground-truth GRN
\autocite{marbach_generatingrealisticsilico_2009}. To compare a
predicted cell-specific GRN with the ground-truth cell-specific GRN, the
top 10'000 interactions per cell is retained, and the mean AUROC and
AUPR scores are calculated across all cells.

We compared the mean AUROC and AUPR scores obtained by the three CSNI
methods across all datasets by performing pairwise non-parametric paired
two-sided Durbin-Conover tests
\autocite{conover_multiplecomparisonsprocedures_1979} using
\texttt{pairwiseComparisons}
\autocite{patil_pairwisecomparisonsmultiplepairwise_2019}. Test
statistics and p-values for the all pairwise combinations are reported
in the Source Data file. Reported p-values are adjusted for multiple
testing using Holm correction
\autocite{holm_simplesequentiallyrejective_1979}.

\hypertarget{sec:dyngen-velcompare}{%
\subsection{Comparison of RNA velocity
methods}\label{sec:dyngen-velcompare}}

Three datasets were generated for each of the 14 different predefined
backbones, resulting in a collection of 42 datasets. Throughout each of
the simulation, the propensity of the transcription and mRNA decay is
collected, as the RNA velocity of a gene at any point in the simulation
is the difference between the transcription propensity and the mRNA
decay propensity.

We applied two RNA velocity methods: velocyto
\autocite{lamanno_rnavelocitysingle_2018}, as implemented in the
\texttt{velocyto.py} package, and scvelo method
\autocite{bergen_generalizingrnavelocity_2020}, as implemented in the
\texttt{scvelo} package. For scvelo, we chose two parameter settings for
``mode'', namely ``stochastic'' and ``dynamical''. For both methods, we
used the same normalized data as provided by dyngen, with no extra cell
or feature filtering, but otherwise matched the parameters to their
respective tutorial vignettes as well as possible.

We compared each RNA velocity prediction to the ground-truth using two
metrics: the velocity correlation and the velocity arrow cosine. For the
velocity correlation, we extracted a ground truth RNA velocity by
subtracting for each mRNA molecule the propensity of its production by
the propensity of its degradation. If the expression of an mRNA will
increase in the future, this value is positive, while it is negative if
it is going to decrease. For each gene, we determined its velocity
correlation by calculating the Spearman rank correlation between the
ground truth velocity with the observed velocity. For the velocity arrow
cosine, we determined a set of 100 trajectory waypoints uniformly spread
on the trajectory. For each waypoint, we weighted each cell based on a
Gaussian kernel on its geodesic distance from the waypoint. These
weights were used to calculate a weighted average velocity vector of
each waypoint. We then calculated for each waypoint the cosine
similarity between this velocity vector and the known direction of the
trajectory.

We compared the velocity correlation and velocity arrow cosine scores
obtained by velocyto and scvelo across all datasets by performing
pairwise non-parametric paired two-sided Durbin-Conover tests
\autocite{conover_multiplecomparisonsprocedures_1979} using
\texttt{pairwiseComparisons}
\autocite{patil_pairwisecomparisonsmultiplepairwise_2019}. Test
statistics and p-values for the all pairwise combinations are reported
in the Source Data file. Reported p-values are adjusted for multiple
testing using Holm correction
\autocite{holm_simplesequentiallyrejective_1979}.

\hypertarget{sec:dyngen-tacompare}{%
\subsection{Comparison of trajectory
alignment}\label{sec:dyngen-tacompare}}

Four custom linear backbones of varying sizes were constructed. For each
of these backbones, 10 datasets were generated with 10 different seeds,
resulting in a total of 40 datasets. Every dataset is generated in three
main steps. First, the GRN is generated based on the given backbone.
Next, generating the kinetics, gold standard, and cells is performed
twice, resulting in two sub-datasets. Finally, the two sub-datasets are
combined and cells are sampled from the combined dataset. Since the two
sub-datasets were simulated with different kinetic parameters, the
combined dataset will contain two trajectories.

On each combined dataset we applied two trajectory alignment methods,
Dynamic Time Warping (DTW)
\autocite{giorgino_computingvisualizingdynamic_2009} and cellAlign
\autocite{alpert_alignmentsinglecelltrajectories_2018}. DTW is designed
to align temporal sequences by dilating or contracting the sequences to
best match each other. cellAlign uses DTW to perform this alignment, but
first interpolates and rescales the input data in order to better cope
with single-cell omics data.

To evaluate a trajectory alignment method on a combined dataset we
computed the geodesic distances of each cell from the start of the
trajectory, also called the \emph{pseudotime}. For each dataset, the
pseudotime values are rescaled between 0 and 1 to allow for easier
comparison. A trajectory alignment produces a sequence of index pairs
\([(i_0, j_0), (i_1, j_1), \ldots, (i_N, j_N)]\), where \(i_0\) and
\(j_0\) are equal to 0 (the first position in both pseudotime series),
\(i_N\) and \(j_N\) are equal to the respective last positions in the
pair of pseudotime series, and \([i_0, i_1, \ldots, i_N]\) and
\([j_0, j_1, \ldots, j_N]\) are in ascending order and can contain
duplicates values. The ABWAP metric is defined as follows, where
\(pt_1\) and \(pt_2\) are the unit pseudotime vectors. See Supplementary
Fig.~1D for a visual interpretation of this metric.

\begin{equation}
  \textrm{ABWAP} = 1 - \textrm{area\_under\_curve}(pt_1[i_0 .. i_N] + pt_2[j_0 .. j_N], abs(pt_1[i_0 .. i_N] - pt_2[j_0 .. j_N]))
\end{equation}

We compared the ABWAP scores obtained by DTW and cellAlign across all
datasets by performing pairwise non-parametric paired two-sided
Durbin-Conover tests
\autocite{conover_multiplecomparisonsprocedures_1979} using
\texttt{pairwiseComparisons}
\autocite{patil_pairwisecomparisonsmultiplepairwise_2019}. Test
statistics and p-values for the all pairwise combinations are reported
in the Source Data file. Reported p-values are adjusted for multiple
testing using Holm correction
\autocite{holm_simplesequentiallyrejective_1979}.

\hypertarget{data-availability}{%
\section{Data Availability}\label{data-availability}}

Source data for box plots in Figure \ref{fig:applications} and
Supplementary Figures~1, 2~and~3 are provided with this paper. All code
and data required to reproduce the analysis are available on GitHub at
\href{https://github.com/dynverse/dyngen_manuscript}{github.com/dynverse/dyngen\_manuscript}.
The datasets generated for the different use cases are available on
Zenodo with record number 4637926 (doi:
\href{https://doi.org/10.5281/zenodo.4637926}{10.5281/zenodo.4637926}).

\hypertarget{code-availability}{%
\section{Code Availability}\label{code-availability}}

Results in this manuscript were generated with R 4.0.3 and dyngen 1.0.0.
dyngen is available as an open-source software package at
\href{https://cran.r-project.org/package=dyngen}{cran.r-project.org/package=dyngen}
and also on Zenodo with record number 4751443 (doi:
\href{https://doi.org/10.5281/zenodo.4751443}{10.5281/zenodo.4751443}).
The analyses performed in this manuscript are available on GitHub at
\href{https://github.com/dynverse/dyngen_manuscript}{github.com/dynverse/dyngen\_manuscript}.

The version numbers of downstream dependencies of dyngen and
dyngen\_manuscript used in this study are: anndata 0.7.5.1, assertthat
0.2.1, babelwhale 1.0.1, bit 4.0.4, bit64 4.0.5, carrier 0.1.0,
cellAlign 0.1.0, codetools 0.2-18, colorspace 2.0-0, compiler 4.0.4,
crayon 1.4.1.9000, data.table 1.13.4, DBI 1.1.1, debugme 1.1.0, desc
1.2.0, digest 0.6.27, dplyr 1.0.5, dtw 1.22-3, dynparam 1.0.1, dynutils
1.0.6, dynwrap 1.2.2, ellipsis 0.3.1, fansi 0.4.2, farver 2.1.0, future
1.20.1, future.apply 1.7.0, generics 0.1.0, ggforce 0.3.2, ggplot2
3.3.3, ggraph 2.0.4, ggrepel 0.9.0, GillespieSSA2 0.2.7, globals 0.14.0,
glue 1.4.2, graphlayouts 0.7.1, grid 4.0.4, gridExtra 2.3, gtable 0.3.0,
gtools 3.8.2, hdf5r 1.3.3, hms 1.0.0, igraph 1.2.6, irlba 2.3.3,
jsonlite 1.7.2, lattice 0.20-41, lifecycle 1.0.0, lisi 1.0, listenv
0.8.0, lmds 0.1.0, magrittr 2.0.1, MASS 7.3-53, Matrix 1.3-2,
matrixStats 0.57.0, munsell 0.5.0, parallel 4.0.4, parallelly 1.21.0,
patchwork 1.1.1, pbapply 1.4-3, pheatmap 1.0.12, pillar 1.5.1, pkgconfig
2.0.3, plyr 1.8.6, polyclip 1.10-0, pracma 2.2.9, processx 3.4.5, proxy
0.4-24, proxyC 0.1.5, ps 1.6.0, purrr 0.3.4, R6 2.5.0, RANN 2.6.1,
rappdirs 0.3.3, RColorBrewer 1.1-2, Rcpp 1.0.6, RcppParallel 5.0.2,
RcppXPtrUtils 0.1.1, readr 1.4.0, remotes 2.2.0, reshape2 1.4.4,
reticulate 1.18-9007, rlang 0.4.10, rprojroot 2.0.2, scales 1.1.1,
sctransform 0.3.2, scvelo 0.1.0.9000, stringi 1.5.3, stringr 1.4.0,
tibble 3.0.5, tidygraph 1.2.0, tidyr 1.1.2, tidyselect 1.1.0, tools
4.0.4, tweenr 1.0.1, utf8 1.1.4, vctrs 0.3.6, viridis 0.5.1, viridisLite
0.3.0, yaml 2.2.1.

\hypertarget{author-contributions}{%
\section{Author contributions}\label{author-contributions}}

\begin{itemize}
\tightlist
\item
  W.S. and R.C. designed the study.
\item
  R.C., W.S., and L.D. performed the experiments and analysed the data.
\item
  R.C. and W.S. implemented the dyngen software package.
\item
  R.C., W.S., L.D., and Y.S. wrote the manuscript.
\item
  Y.S. supervised the project.
\end{itemize}

\printbibliography

\begin{filecontents}{\jobname.bib}
@Article{aibar_scenicsinglecellregulatory_2017,
	title = {{{SCENIC}}: Single-Cell Regulatory Network Inference and Clustering},
	author = {Sara Aibar and Carmen Bravo Gonz{\a'a}lez-Blas and Thomas Moerman and V{\^a}n Anh Huynh-Thu and Hana Imrichova and Gert Hulselmans and Florian Rambow and Jean-Christophe Marine and Pierre Geurts and Jan Aerts and Joost {van den Oord} and Zeynep Kalender Atak and Jasper Wouters and Stein Aerts},
	year = {2017},
	month = {oct},
	issn = {1548-7091},
	doi = {10.1038/nmeth.4463},
	journal = {Nature Methods},
}

@Article{alpert_alignmentsinglecelltrajectories_2018,
	title = {Alignment of Single-Cell Trajectories to Compare Cellular Expression Dynamics},
	author = {Ayelet Alpert and Lindsay S. Moore and Tania Dubovik and Shai S. Shen-Orr},
	year = {2018},
	month = {apr},
	volume = {15},
	pages = {267--270},
	issn = {1548-7105},
	doi = {10.1038/nmeth.4628},
	abstract = {cellAlign enables quantitative comparisons of expression dynamics within and between single-cell trajectories based on single-cell RNA-seq or mass cytometry data.},
	copyright = {2018 Nature Publishing Group, a division of Macmillan Publishers Limited. All Rights Reserved.},
	file = {/home/rcannood/Zotero/storage/QB7FH2T9/nmeth.html},
	journal = {Nature Methods},
	language = {en},
	number = {4},
}

@Article{bergen_generalizingrnavelocity_2020,
	title = {Generalizing {{RNA}} Velocity to Transient Cell States through Dynamical Modeling},
	author = {Volker Bergen and Marius Lange and Stefan Peidli and F. Alexander Wolf and Fabian J. Theis},
	year = {2020},
	month = {dec},
	volume = {38},
	pages = {1408--1414},
	issn = {1546-1696},
	doi = {10.1038/s41587-020-0591-3},
	abstract = {RNA velocity has opened up new ways of studying cellular differentiation in single-cell RNA-sequencing data. It describes the rate of gene expression change for an individual gene at a given time point based on the ratio of its spliced and unspliced messenger RNA (mRNA). However, errors in velocity estimates arise if the central assumptions of a common splicing rate and the observation of the full splicing dynamics with steady-state mRNA levels are violated. Here we present scVelo, a method that overcomes these limitations by solving the full transcriptional dynamics of splicing kinetics using a likelihood-based dynamical model. This generalizes RNA velocity to systems with transient cell states, which are common in development and in response to perturbations. We apply scVelo to disentangling subpopulation kinetics in neurogenesis and pancreatic endocrinogenesis. We infer gene-specific rates of transcription, splicing and degradation, recover each cell's position in the underlying differentiation processes and detect putative driver genes. scVelo will facilitate the study of lineage decisions and gene regulation.},
	journal = {Nature Biotechnology},
	number = {12},
}

@Article{bjornsson_digitaltwinspersonalize_2019,
	title = {Digital Twins to Personalize Medicine},
	author = {Bergthor Bj{\"o}rnsson and Carl Borrebaeck and Nils Elander and Thomas Gasslander and Danuta R. Gawel and Mika Gustafsson and Rebecka J{\"o}rnsten and Eun Jung Lee and Xinxiu Li and Sandra Lilja and David Mart\'inez-Enguita and Andreas Matussek and Per Sandstr{\"o}m and Samuel Sch{\"a}fer and Margaretha Stenmarker and X. F. Sun and Oleg Sysoev and Huan Zhang and Mikael Benson and {on behalf of the Swedish Digital Twin Consortium}},
	year = {2019},
	month = {dec},
	volume = {12},
	pages = {4},
	issn = {1756-994X},
	doi = {10.1186/s13073-019-0701-3},
	abstract = {Personalized medicine requires the integration and processing of vast amounts of data. Here, we propose a solution to this challenge that is based on constructing Digital Twins. These are high-resolution models of individual patients that are computationally treated with thousands of drugs to find the drug that is optimal for the patient.},
	journal = {Genome Medicine},
	number = {1},
}

@Article{cacchiarelli_aligningsinglecelldevelopmental_2018,
	title = {Aligning {{Single}}-{{Cell Developmental}} and {{Reprogramming Trajectories Identifies Molecular Determinants}} of {{Myogenic Reprogramming Outcome}}},
	author = {Davide Cacchiarelli and Xiaojie Qiu and Sanjay Srivatsan and Anna Manfredi and Michael Ziller and Eliah Overbey and Antonio Grimaldi and Jonna Grimsby and Prapti Pokharel and Kenneth J. Livak and Shuqiang Li and Alexander Meissner and Tarjei S. Mikkelsen and John L. Rinn and Cole Trapnell},
	year = {2018},
	month = {sep},
	volume = {7},
	pages = {258-268.e3},
	issn = {2405-4712},
	doi = {10.1016/j.cels.2018.07.006},
	abstract = {{$<$}h2{$>$}Summary{$<$}/h2{$><$}p{$>$}Cellular reprogramming through manipulation of defined factors holds great promise for large-scale production of cell types needed for use in therapy and for revealing principles of gene regulation. However, most reprogramming systems are inefficient, converting only a fraction of cells to the desired state. Here, we analyze MYOD-mediated reprogramming of human fibroblasts to myotubes, a well-characterized model system for direct conversion by defined factors, at pseudotemporal resolution using single-cell RNA-seq. To expose barriers to efficient conversion, we introduce a novel analytic technique, trajectory alignment, which enables quantitative comparison of gene expression kinetics across two biological processes. Reprogrammed cells navigate a trajectory with branch points that correspond to two alternative decision points, with cells that select incorrect branches terminating at aberrant or incomplete reprogramming outcomes. Analysis of these branch points revealed insulin and BMP signaling as crucial molecular determinants of reprogramming. Single-cell trajectory alignment enables rigorous quantitative comparisons between biological trajectories found in diverse processes in development, reprogramming, and other contexts.{$<$}/p{$>$}},
	file = {/home/rcannood/Zotero/storage/YQH8GP2I/Cacchiarelli et al. - 2018 - Aligning Single-Cell Developmental and Reprogrammi.pdf;/home/rcannood/Zotero/storage/Q5MUK95E/S2405-4712(18)30314-4.html},
	journal = {Cell Systems},
	language = {English},
	number = {3},
	pmid = {30195438},
}

@Article{cannoodt_singlecellomicsdatasets_2018,
	title = {Single-Cell -Omics Datasets Containing a Trajectory},
	author = {Robrecht Cannoodt and Wouter Saelens and Helena Todorov and Yvan Saeys},
	year = {2018},
	month = {oct},
	doi = {10.5281/zenodo.1211532},
	journal = {Zenodo},
}

@TechReport{conover_multiplecomparisonsprocedures_1979,
	title = {On Multiple-Comparisons Procedures},
	author = {William Jay Conover and Ronald L Iman},
	year = {1979},
	institution = {{Technical report, Los Alamos Scientific Laboratory}},
}

@Article{duo_systematicperformanceevaluation_2018,
	title = {A Systematic Performance Evaluation of Clustering Methods for Single-Cell {{RNA}}-Seq Data},
	author = {Angelo Du{\a`o} and Mark D. Robinson and Charlotte Soneson},
	year = {2018},
	volume = {7},
	pages = {1141},
	issn = {2046-1402},
	doi = {10.12688/f1000research.15666.2},
	abstract = {Subpopulation identification, usually via some form of unsupervised clustering, is a fundamental step in the analysis of many single-cell RNA-seq data sets. This has motivated the development and application of a broad range of clustering methods, based on various underlying algorithms. Here, we provide a systematic and extensible performance evaluation of 14 clustering algorithms implemented in R, including both methods developed explicitly for scRNA-seq data and more general-purpose methods. The methods were evaluated using nine publicly available scRNA-seq data sets as well as three simulations with varying degree of cluster separability. The same feature selection approaches were used for all methods, allowing us to focus on the investigation of the performance of the clustering algorithms themselves. We evaluated the ability of recovering known subpopulations, the stability and the run time and scalability of the methods. Additionally, we investigated whether the performance could be improved by generating consensus partitions from multiple individual clustering methods. We found substantial differences in the performance, run time and stability between the methods, with SC3 and Seurat showing the most favorable results. Additionally, we found that consensus clustering typically did not improve the performance compared to the best of the combined methods, but that several of the top-performing methods already perform some type of consensus clustering. All the code used for the evaluation is available on GitHub ( https://github.com/markrobinsonuzh/scRNAseq\_clustering\_comparison). In addition, an R package providing access to data and clustering results, thereby facilitating inclusion of new methods and data sets, is available from Bioconductor ( https://bioconductor.org/packages/DuoClustering2018).},
	file = {/home/rcannood/Zotero/storage/ASWVFH87/Duò et al. - 2018 - A systematic performance evaluation of clustering .pdf},
	journal = {F1000Research},
	keywords = {Algorithms,Benchmarking,Cluster Analysis,Clustering,Clustering methods,Gene Expression Profiling,Humans,RNA-seq,RNA; Small Cytoplasmic,Sequence Analysis; RNA,Single-Cell Analysis,Single-cell RNA-seq,Software},
	language = {eng},
	pmcid = {PMC6134335},
	pmid = {30271584},
}

@Article{ewing_combiningtumorgenome_2015,
	title = {Combining Tumor Genome Simulation with Crowdsourcing to Benchmark Somatic Single-Nucleotide-Variant Detection},
	author = {Adam D. Ewing and Kathleen E. Houlahan and Yin Hu and Kyle Ellrott and Cristian Caloian and Takafumi N. Yamaguchi and J. Christopher Bare and Christine P'ng and Daryl Waggott and Veronica Y. Sabelnykova and {ICGC-TCGA DREAM Somatic Mutation Calling Challenge Participants} and Liu Xi and Ninad Dewal and Yu Fan and Wenyi Wang and David Wheeler and Andreas Wilm and Grace Hui Ting and Chenhao Li and Denis Bertrand and Niranjan Nagarajan and Qing-Rong Chen and Chih-Hao Hsu and Ying Hu and Chunhua Yan and Warren Kibbe and Daoud Meerzaman and Kristian Cibulskis and Mara Rosenberg and Louis Bergelson and Adam Kiezun and Amie Radenbaugh and Anne-Sophie Sertier and Anthony Ferrari and Laurie Tonton and Kunal Bhutani and Nancy F. Hansen and Difei Wang and Lei Song and Zhongwu Lai and Yang Liao and Wei Shi and Jos{\a'e} Carbonell-Caballero and Joaqu\'in Dopazo and Cheryl C. K. Lau and Justin Guinney and Michael R. Kellen and Thea C. Norman and David Haussler and Stephen H. Friend and Gustavo Stolovitzky and Adam A. Margolin and Joshua M. Stuart and Paul C. Boutros},
	year = {2015},
	month = {jul},
	volume = {12},
	pages = {623--630},
	issn = {1548-7105},
	doi = {10.1038/nmeth.3407},
	abstract = {The detection of somatic mutations from cancer genome sequences is key to understanding the genetic basis of disease progression, patient survival and response to therapy. Benchmarking is needed for tool assessment and improvement but is complicated by a lack of gold standards, by extensive resource requirements and by difficulties in sharing personal genomic information. To resolve these issues, we launched the ICGC-TCGA DREAM Somatic Mutation Calling Challenge, a crowdsourced benchmark of somatic mutation detection algorithms. Here we report the BAMSurgeon tool for simulating cancer genomes and the results of 248 analyses of three in silico tumors created with it. Different algorithms exhibit characteristic error profiles, and, intriguingly, false positives show a trinucleotide profile very similar to one found in human tumors. Although the three simulated tumors differ in sequence contamination (deviation from normal cell sequence) and in subclonality, an ensemble of pipelines outperforms the best individual pipeline in all cases. BAMSurgeon is available at https://github.com/adamewing/bamsurgeon/.},
	copyright = {2015 Nature Publishing Group},
	file = {/home/rcannood/Zotero/storage/NBPKWHK3/Ewing et al. - 2015 - Combining tumor genome simulation with crowdsourci.pdf;/home/rcannood/Zotero/storage/TDZU2TL9/nmeth.html},
	journal = {Nature Methods},
	language = {en},
	number = {7},
}

@Article{ferrell_bistabilitybifurcationswaddington_2012,
	title = {Bistability, {{Bifurcations}}, and {{Waddington}}'s {{Epigenetic Landscape}}},
	author = {James E Ferrell},
	year = {2012},
	month = {jun},
	volume = {22},
	pages = {R458--R466},
	issn = {0960-9822},
	doi = {10.1016/j.cub.2012.03.045},
	abstract = {Waddington's epigenetic landscape is probably the most famous and most powerful metaphor in developmental biology. Cells, represented by balls, roll downhill through a landscape of bifurcating valleys. Each new valley represents a possible cell fate and the ridges between the valleys maintain the cell fate once it has been chosen. Here I examine models of two important developmental processes \$\textbackslash backslash\$textemdash\{\} cell-fate induction and lateral inhibition \$\textbackslash backslash\$textemdash\{\} and ask whether the landscapes for these models at least qualitatively resemble Waddington's picture. For cell-fate induction, the answer is no. The commitment of a cell to a new fate corresponds to the disappearance of a valley from the landscape, not the splitting of one valley into two, and it occurs through a type of bifurcation \$\textbackslash backslash\$textemdash\{\} a saddle-node bifurcation \$\textbackslash backslash\$textemdash\{\} that possesses an intrinsic irreversibility that is missing from Waddington's picture. Lateral inhibition, a symmetrical cell\$\textbackslash backslash\$textendash\{\}cell competition process, corresponds better to Waddington's picture, with one valley reversibly splitting into two through a pitchfork bifurcation. I propose an alternative epigenetic landscape that has numerous valleys and ridges right from the start, with the process of cell-fate commitment corresponding to the irreversible disappearance of some of these valleys and ridges, via cell-fate induction, complemented by the creation of new valleys and ridges through processes like cell\$\textbackslash backslash\$textendash\{\}cell competition.},
	file = {/home/rcannood/Zotero/storage/GK4TJ5BV/Ferrell - 2012 - Bistability, Bifurcations, and Waddington's Epigen.pdf;/home/rcannood/Zotero/storage/HUSMJA22/Ferrell - 2012 - Bistability, Bifurcations, and Waddington's Epigen.pdf;/home/rcannood/Zotero/storage/8VLPBTG2/S0960-9822(12)00332-6.html;/home/rcannood/Zotero/storage/XBTP42Y6/S0960982212003326.html},
	journal = {Current Biology},
	number = {11},
}

@Article{gibson_probabilisticmodelprokaryotic_2000,
	title = {A Probabilistic Model of a Prokaryotic Gene and Its Regulation},
	author = {Michael A. Gibson and Jehoshua Bruck},
	year = {2000},
	journal = {Computational Methods in Molecular Biology: From Genotype to Phenotype, MIT press, Cambridge},
}

@Article{gillespie_exactstochasticsimulation_1977,
	title = {Exact Stochastic Simulation of Coupled Chemical Reactions},
	author = {Daniel T. Gillespie},
	year = {1977},
	month = {dec},
	volume = {81},
	pages = {2340--2361},
	issn = {0022-3654},
	doi = {10.1021/j100540a008},
	file = {/home/rcannood/Zotero/storage/ELV9AQRF/j100540a008.html},
	journal = {The Journal of Physical Chemistry},
	number = {25},
}

@Article{giorgino_computingvisualizingdynamic_2009,
	title = {Computing and {{Visualizing Dynamic Time Warping Alignments}} in {{R}}: {{The}} Dtw {{Package}}},
	author = {Toni Giorgino},
	year = {2009},
	month = {sep},
	doi = {10.18637/jss.v031.i07},
	abstract = {Dynamic time warping is a popular technique for comparing time series, providing both a distance measure that is insensitive to local compression and stretches and the warping which optimally deforms one of the two input series onto the other. A variety of algorithms and constraints have been discussed in the literature. The dtw package provides an unification of them; it allows R users to compute time series alignments mixing freely a variety of continuity constraints, restriction windows, endpoints, local distance definitions, and so on. The package also provides functions for visualizing alignments and constraints using several classic diagram types.},
	journal = {Journal of Statistical Software},
	number = {7},
}

@Article{glass_passingmessagesbiological_2013,
	title = {Passing {{Messages}} between {{Biological Networks}} to {{Refine Predicted Interactions}}},
	author = {Kimberly Glass and Curtis Huttenhower and John Quackenbush and Guo-Cheng Yuan},
	year = {2013},
	month = {may},
	volume = {8},
	pages = {e64832},
	doi = {10.1371/journal.pone.0064832},
	abstract = {Regulatory network reconstruction is a fundamental problem in computational biology. There are significant limitations to such reconstruction using individual datasets, and increasingly people attempt to construct networks using multiple, independent datasets obtained from complementary sources, but methods for this integration are lacking. We developed PANDA (Passing Attributes between Networks for Data Assimilation), a message-passing model using multiple sources of information to predict regulatory relationships, and used it to integrate protein-protein interaction, gene expression, and sequence motif data to reconstruct genome-wide, condition-specific regulatory networks in yeast as a model. The resulting networks were not only more accurate than those produced using individual data sets and other existing methods, but they also captured information regarding specific biological mechanisms and pathways that were missed using other methodologies. PANDA is scalable to higher eukaryotes, applicable to specific tissue or cell type data and conceptually generalizable to include a variety of regulatory, interaction, expression, and other genome-scale data. An implementation of the PANDA algorithm is available at www.sourceforge.net/projects/panda-net.},
	journal = {PLOS ONE},
	number = {5},
}

@Article{graf_forcingcellschange_2009,
	title = {Forcing {{Cells}} to {{Change Lineages}}},
	author = {Thomas Graf and Tariq Enver},
	year = {2009},
	month = {dec},
	volume = {462},
	pages = {587},
	issn = {1476-4687},
	doi = {10.1038/nature08533},
	abstract = {\$\textbackslash textless\$p\$\textbackslash textgreater\$Though it is important for the success of multicellular organisms that their many different cell types remain stable once specified, advances in the production of induced pluripotent stem (iPS) cells from somatic cells by exposure to a transcription factor cocktail emphasize the potential of almost any cell to be reprogrammed. This raises the question of whether interconversion between cell types occurs physiologically or in the context of disease. Thomas Graf and Tariq Enver use examples from the blood cell system to illustrate the principles of transcription factor transdifferentiation, which they conclude is a process fundamentally similar to differentiation and different from iPS reprogramming. They postulate that conversions of one cell type to another can occur as a normal developmental process, and speculate on the prospects of forced lineage reprogramming in regenerative medicine.\$\textbackslash textless\$/p\$\textbackslash textgreater\$},
	journal = {Nature},
	number = {7273},
}

@Article{hache_gengesystematicgeneration_2009,
	title = {{{GeNGe}}: Systematic Generation of Gene Regulatory Networks},
	shorttitle = {{{GeNGe}}},
	author = {Hendrik Hache and Christoph Wierling and Hans Lehrach and Ralf Herwig},
	year = {2009},
	month = {may},
	volume = {25},
	pages = {1205--1207},
	issn = {1367-4803},
	doi = {10.1093/bioinformatics/btp115},
	abstract = {Summary: The analysis of gene regulatory networks (GRNs) is a central goal of bioinformatics highly accelerated by the advent of new experimental techniques, such as RNA interference. A battery of reverse engineering methods has been developed in recent years to reconstruct the underlying GRNs from these and other experimental data. However, the performance of the individual methods is poorly understood and validation of algorithmic performances is still missing to a large extent. To enable such systematic validation, we have developed the web application GeNGe (GEne Network GEnerator), a controlled framework for the automatic generation of GRNs. The theoretical model for a GRN is a non-linear differential equation system. Networks can be user-defined or constructed in a modular way with the option to introduce global and local network perturbations. Resulting data can be used, e.g. as benchmark data for evaluating GRN reconstruction methods or for predicting effects of perturbations as theoretical counterparts of biological experiments.Availability: Available online at http://genge.molgen.mpg.deContact:hache@molgen.mpg.deSupplementary information:Supplementary data are available at Bioinformatics online.},
	journal = {Bioinformatics},
	number = {9},
}

@Article{holm_simplesequentiallyrejective_1979,
	title = {A Simple Sequentially Rejective Multiple Test Procedure},
	author = {Sture Holm},
	year = {1979},
	pages = {65--70},
	journal = {Scandinavian journal of statistics},
}

@Article{huynh-thu_inferringregulatorynetworks_2010,
	title = {Inferring {{Regulatory Networks}} from {{Expression Data Using Tree}}-{{Based Methods}}},
	author = {V{\^a}n Anh Huynh-Thu and Irrthum Alexandre and Wehenkel Louis and Geurts Pierre and Alexandre Irrthum and Louis Wehenkel and Pierre Geurts},
	year = {2010},
	month = {jan},
	volume = {5},
	pages = {e12776},
	issn = {1932-6203},
	doi = {10.1371/journal.pone.0012776},
	abstract = {One of the pressing open problems of computational systems biology is the elucidation of the topology of genetic regulatory networks (GRNs) using high throughput genomic data, in particular microarray gene expression data. The Dialogue for Reverse Engineering Assessments and Methods (DREAM) challenge aims to evaluate the success of GRN inference algorithms on benchmarks of simulated data. In this article, we present GENIE3, a new algorithm for the inference of GRNs that was best performer in the DREAM4 In Silico Multifactorial challenge. GENIE3 decomposes the prediction of a regulatory network between p genes into p different regression problems. In each of the regression problems, the expression pattern of one of the genes (target gene) is predicted from the expression patterns of all the other genes (input genes), using tree-based ensemble methods Random Forests or Extra-Trees. The importance of an input gene in the prediction of the target gene expression pattern is taken as an indication of a putative regulatory link. Putative regulatory links are then aggregated over all genes to provide a ranking of interactions from which the whole network is reconstructed. In addition to performing well on the DREAM4 In Silico Multifactorial challenge simulated data, we show that GENIE3 compares favorably with existing algorithms to decipher the genetic regulatory network of Escherichia coli. It doesn't make any assumption about the nature of gene regulation, can deal with combinatorial and non-linear interactions, produces directed GRNs, and is fast and scalable. In conclusion, we propose a new algorithm for GRN inference that performs well on both synthetic and real gene expression data. The algorithm, based on feature selection with tree-based ensemble methods, is simple and generic, making it adaptable to other types of genomic data and interactions.},
	file = {/home/rcannood/Zotero/storage/3I6PLCQL/Huynh-Thu et al. - 2010 - Inferring Regulatory Networks from Expression Data Using Tree-Based Methods.pdf},
	journal = {PLoS ONE},
	keywords = {Algorithms,Computational Biology,Computational Biology: methods,Escherichia coli,Escherichia coli: genetics,Gene Expression Regulation,Gene Regulatory Networks,Oligonucleotide Array Sequence Analysis},
	number = {9},
	pmid = {20927193},
}

@Article{kanton_organoidsinglecellgenomic_2019,
	title = {Organoid Single-Cell Genomic Atlas Uncovers Human-Specific Features of Brain Development},
	author = {Sabina Kanton and Michael James Boyle and Zhisong He and Malgorzata Santel and Anne Weigert and F{\a'a}tima Sanch\'is-Calleja and Patricia Guijarro and Leila Sidow and Jonas Simon Fleck and Dingding Han and Zhengzong Qian and Michael Heide and Wieland B. Huttner and Philipp Khaitovich and Svante P{\"a}{{\"a}}bo and Barbara Treutlein and J. Gray Camp},
	year = {2019},
	month = {oct},
	volume = {574},
	pages = {418--422},
	issn = {1476-4687},
	doi = {10.1038/s41586-019-1654-9},
	abstract = {Species comparisons using single-cell transcriptomics and accessible chromatin profiling in stem cell-derived cerebral organoids are used to map dynamic gene-regulatory changes that are unique to humans.},
	copyright = {2019 The Author(s), under exclusive licence to Springer Nature Limited},
	file = {/home/rcannood/Zotero/storage/AVPYT7HY/s41586-019-1654-9.html},
	journal = {Nature},
	language = {en},
	number = {7778},
}

@Article{kuijjer_estimatingsamplespecificregulatory_2015,
	title = {Estimating Sample-Specific Regulatory Networks},
	author = {Marieke Lydia Kuijjer and Matthew Tung and GuoCheng Yuan and John Quackenbush and Kimberly Glass},
	year = {2015},
	pages = {1--19},
	abstract = {Biological systems are driven by intricate interactions among the complex array of molecules that comprise the cell. Many methods have been developed to reconstruct network models that attempt to capture those interactions. These methods often draw on large numbers of measured expression samples to tease out subtle signals and infer connections between genes (or gene products). The result is an aggregate network model representing a single estimate for edge likelihoods. While informative, aggregate models fail to capture the heterogeneity that is often represented in a population. Here we propose a method to reverse engineer sample-specific networks from aggregate network models. We demonstrate the accuracy and applicability of our approach in several datasets, including simulated data, microarray expression data from synchronized yeast cells, and RNA-seq data collected from human subjects. We show that these sample-specific networks can be used to study the evolution of network topology across time and to characterize shifts in gene regulation that may not be apparent in the expression data. We believe the ability to generate sample-specific networks will revolutionize the field of network biology and has the potential to usher in an era of precision network medicine.},
	file = {/home/rcannood/Zotero/storage/YECTK66M/Kuijjer et al. - 2015 - Estimating sample-specific regulatory networks.pdf},
}

@Article{kuijjer_estimatingsamplespecificregulatory_2019,
	title = {Estimating {{Sample}}-{{Specific Regulatory Networks}}},
	author = {Marieke Lydia Kuijjer and Matthew George Tung and GuoCheng Yuan and John Quackenbush and Kimberly Glass},
	year = {2019},
	month = {mar},
	volume = {14},
	pages = {226--240},
	issn = {2589-0042},
	doi = {10.1016/j.isci.2019.03.021},
	abstract = {Biological systems are driven by intricate interactions among molecules. Many methods have been developed that draw on large numbers of expression samples to infer connections between genes (or their products). The result is an aggregate network representing a single estimate for the likelihood of each interaction, or ``edge,'' in the network. Although informative, aggregate models fail to capture population heterogeneity. Here we propose a method to reverse engineer sample-specific networks from aggregate networks. We demonstrate our approach in several contexts, including simulated, yeast microarray, and human lymphoblastoid cell line RNA sequencing data. We use these sample-specific networks to study changes in network topology across time and to characterize shifts in gene regulation that were not apparent in the expression data. We believe that generating sample-specific networks will greatly facilitate the application of network methods to large, complex, and heterogeneous multi-omic datasets, supporting the emerging field of precision network medicine.,                                        \textbullet{}               We developed LIONESS to extract single-sample networks from aggregate models                                         \textbullet{}               We tested LIONESS using in silico, yeast, and human expression data                                         \textbullet{}               LIONESS-estimated networks are reproducible, accurate, and biologically meaningful                                         \textbullet{}               Single-sample network analysis highlights important biological processes                                 , Biological Sciences; Bioinformatics; Complex Systems},
	file = {/home/rcannood/Zotero/storage/ITHUQ7NH/Kuijjer et al. - 2019 - Estimating Sample-Specific Regulatory Networks.pdf},
	journal = {iScience},
	pmcid = {PMC6463816},
	pmid = {30981959},
}

@Article{kuijjer_lionessrsinglesample_2019,
	title = {{{lionessR}}: Single Sample Network Inference in {{R}}},
	shorttitle = {{{lionessR}}},
	author = {Marieke L. Kuijjer and Ping-Han Hsieh and John Quackenbush and Kimberly Glass},
	year = {2019},
	month = {oct},
	volume = {19},
	pages = {1003},
	issn = {1471-2407},
	doi = {10.1186/s12885-019-6235-7},
	abstract = {In biomedical research, network inference algorithms are typically used to infer complex association patterns between biological entities, such as between genes or proteins, using data from a population. This resulting aggregate network, in essence, averages over the networks of those individuals in the population. LIONESS (Linear Interpolation to Obtain Network Estimates for Single Samples) is a method that can be used together with a network inference algorithm to extract networks for individual samples in a population. The method's key characteristic is that, by modeling networks for individual samples in a data set, it can capture network heterogeneity in a population. LIONESS was originally made available as a function within the PANDA (Passing Attributes between Networks for Data Assimilation) regulatory network reconstruction framework. However, the LIONESS algorithm is generalizable and can be used to model single sample networks based on a wide range of network inference algorithms.},
	file = {/home/rcannood/Zotero/storage/RS448RZF/Kuijjer et al. - 2019 - lionessR single sample network inference in R.pdf;/home/rcannood/Zotero/storage/9STH6ZYN/s12885-019-6235-7.html},
	journal = {BMC Cancer},
	number = {1},
}

@Article{lamanno_rnavelocitysingle_2018,
	title = {{{RNA Velocity}} of {{Single Cells}}},
	author = {Gioele {La Manno} and Ruslan Soldatov and Amit Zeisel and Emelie Braun and Hannah Hochgerner and Viktor Petukhov and Katja Lidschreiber and Maria E Kastriti and Peter L{\"o}nnerberg and Alessandro Furlan and Jean Fan and Lars E Borm and Zehua Liu and David {van Bruggen} and Jimin Guo and Xiaoling He and Roger Barker and Erik Sundstr{\"o}m and Gon{\c c}alo Castelo-Branco and Patrick Cramer and Igor Adameyko and Sten Linnarsson and Peter V Kharchenko},
	year = {2018},
	month = {aug},
	volume = {560},
	pages = {494--498},
	issn = {1476-4687},
	doi = {10.1038/s41586-018-0414-6},
	abstract = {RNA velocity, estimated in single cells by comparison of spliced and unspliced mRNA, is a good indicator of transcriptome dynamics and will provide a useful tool for analysis of developmental lineage.},
	journal = {Nature},
	number = {7719},
}

@Article{lin_scclassifysamplesize_2020,
	title = {{{scClassify}}: Sample Size Estimation and Multiscale Classification of Cells Using Single and Multiple Reference},
	author = {Yingxin Lin and Yue Cao and Hani Jieun Kim and Agus Salim and Terence P Speed and David M Lin and Pengyi Yang and Jean Yee Hwa Yang},
	year = {2020},
	month = {jun},
	volume = {16},
	pages = {e9389},
	publisher = {{John Wiley \& Sons, Ltd}},
	issn = {1744-4292},
	doi = {10.15252/msb.20199389},
	abstract = {Abstract Automated cell type identification is a key computational challenge in single-cell RNA-sequencing (scRNA-seq) data. To capitalise on the large collection of well-annotated scRNA-seq datasets, we developed scClassify, a multiscale classification framework based on ensemble learning and cell type hierarchies constructed from single or multiple annotated datasets as references. scClassify enables the estimation of sample size required for accurate classification of cell types in a cell type hierarchy and allows joint classification of cells when multiple references are available. We show that scClassify consistently performs better than other supervised cell type classification methods across 114 pairs of reference and testing data, representing a diverse combination of sizes, technologies and levels of complexity, and further demonstrate the unique components of scClassify through simulations and compendia of experimental datasets. Finally, we demonstrate the scalability of scClassify on large single-cell atlases and highlight a novel application of identifying subpopulations of cells from the Tabula Muris data that were unidentified in the original publication. Together, scClassify represents state-of-the-art methodology in automated cell type identification from scRNA-seq data.},
	journal = {Molecular Systems Biology},
	keywords = {cell type hierarchy,cell type identification,multiscale classification,sample size estimation,single-cell},
	number = {6},
}

@Article{liu_personalizedcharacterizationdiseases_2016,
	title = {Personalized Characterization of Diseases Using Sample-Specific Networks},
	author = {Xiaoping Liu and Yuetong Wang and Hongbin Ji and Kazuyuki Aihara and Luonan Chen},
	year = {2016},
	volume = {44},
	pages = {e164--e164},
	issn = {0305-1048},
	doi = {10.1093/nar/gkw772},
	abstract = {A complex disease generally results not from malfunction of individual molecules but from dysfunction of the relevant system or network, which dynamically changes with time and conditions. Thus, estimating a condition-specific network from a single sample is crucial to elucidating the molecular mechanisms of complex diseases at the system level. However, there is currently no effective way to construct such an individual-specific network by expression profiling of a single sample because of the requirement of multiple samples for computing correlations. We developed here with a statistical method, i.e. a sample-specific network (SSN) method, which allows us to construct individual-specific networks based on molecular expressions of a single sample. Using this method, we can characterize various human diseases at a network level. In particular, such SSNs can lead to the identification of individual-specific disease modules as well as driver genes, even without gene sequencing information. Extensive analysis by using the Cancer Genome Atlas data not only demonstrated the effectiveness of the method, but also found new individual-specific driver genes and network patterns for various types of cancer. Biological experiments on drug resistance further validated one important advantage of our method over the traditional methods, i.e. we can even identify such drug resistance genes that actually have no clear differential expression between samples with and without the resistance, due to the additional network information.},
	file = {/home/rcannood/Zotero/storage/6MSYIL7R/Liu et al. - 2016 - Personalized characterization of diseases using sample-specific networks.pdf},
	journal = {Nucleic Acids Research},
	keywords = {cancer,drug resistance,genes,steroid-sensitive nephrotic syndrome},
	number = {22},
	pmid = {27596597},
}

@Article{luecken_currentbestpractices_2019,
	title = {Current Best Practices in Single-Cell {{RNA}}-Seq Analysis: A Tutorial},
	shorttitle = {Current Best Practices in Single-Cell {{RNA}}-Seq Analysis},
	author = {Malte D Luecken and Fabian J Theis},
	year = {2019},
	month = {jun},
	volume = {15},
	pages = {e8746},
	issn = {1744-4292},
	doi = {10.15252/msb.20188746},
	abstract = {Abstract Single-cell RNA-seq has enabled gene expression to be studied at an unprecedented resolution. The promise of this technology is attracting a growing user base for single-cell analysis methods. As more analysis tools are becoming available, it is becoming increasingly difficult to navigate this landscape and produce an up-to-date workflow to analyse one's data. Here, we detail the steps of a typical single-cell RNA-seq analysis, including pre-processing (quality control, normalization, data correction, feature selection, and dimensionality reduction) and cell- and gene-level downstream analysis. We formulate current best-practice recommendations for these steps based on independent comparison studies. We have integrated these best-practice recommendations into a workflow, which we apply to a public dataset to further illustrate how these steps work in practice. Our documented case study can be found at https://www.github.com/theislab/single-cell-tutorial. This review will serve as a workflow tutorial for new entrants into the field, and help established users update their analysis pipelines.},
	file = {/home/rcannood/Zotero/storage/MY4TZUJT/Luecken and Theis - 2019 - Current best practices in single-cell RNA-seq anal.pdf;/home/rcannood/Zotero/storage/7PFR3DKU/msb.html},
	journal = {Molecular Systems Biology},
	keywords = {analysis pipeline development,computational biology,data analysis tutorial,single-cell RNA-seq},
	number = {6},
}

@Article{lummertzdarocha_reconstructioncomplexsinglecell_2018,
	title = {Reconstruction of Complex Single-Cell Trajectories Using {{CellRouter}}},
	author = {Edroaldo {Lummertz da Rocha} and R. Grant Rowe and Vanessa Lundin and Mohan Malleshaiah and Deepak Kumar Jha and Carlos R. Rambo and Hu Li and Trista E. North and James J. Collins and George Q. Daley},
	year = {2018},
	month = {mar},
	volume = {9},
	pages = {892},
	issn = {2041-1723},
	doi = {10.1038/s41467-018-03214-y},
	abstract = {A better understanding of the cell-fate transitions that occur in complex cellular ecosystems in normal development and disease could inform cell engineering efforts and lead to improved therapies. However, a major challenge is to simultaneously identify new cell states, and their transitions, to elucidate the gene expression dynamics governing cell-type diversification. Here, we present CellRouter, a multifaceted single-cell analysis platform that identifies complex cell-state transition trajectories by using flow networks to explore the subpopulation structure of multi-dimensional, single-cell omics data. We demonstrate its versatility by applying CellRouter to single-cell RNA sequencing data sets to reconstruct cell-state transition trajectories during hematopoietic stem and progenitor cell (HSPC) differentiation to the erythroid, myeloid and lymphoid lineages, as well as during re-specification of cell identity by cellular reprogramming of monocytes and B-cells to HSPCs. CellRouter opens previously undescribed paths for in-depth characterization of complex cellular ecosystems and establishment of enhanced cell engineering approaches.},
	journal = {Nature Communications},
	number = {1},
}

@Article{marbach_generatingrealisticsilico_2009,
	title = {Generating {{Realistic In Silico Gene Networks}} for {{Performance Assessment}} of {{Reverse Engineering Methods}}},
	author = {Daniel Marbach and Thomas Schaffter and Claudio Mattiussi and Dario Floreano},
	year = {2009},
	month = {feb},
	volume = {16},
	pages = {229--239},
	issn = {1557-8666},
	doi = {10.1089/cmb.2008.09TT},
	abstract = {Reverse engineering methods are typically first tested on simulated data from in silico networks, for systematic and efficient performance assessment, before an application to real biological networks. In this paper, we present a method for generating biologically plausible in silico networks, which allow realistic performance assessment of network inference algorithms. Instead of using random graph models, which are known to only partly capture the structural properties of biological networks, we generate network structures by extracting modules from known biological interaction networks. Using the yeast transcriptional regulatory network as a test case, we show that extracted modules have a biologically plausible connectivity because they preserve functional and structural properties of the original network. Our method was selected to generate the {"}gold standard{"} networks for the gene network reverse engineering challenge of the third DREAM conference (Dialogue on Reverse Engineering Assessment and Methods 2008, Cambridge, MA).},
	file = {/home/rcannood/Zotero/storage/NXLVA76Q/Marbach et al. - 2009 - Generating Realistic In Silico Gene Networks for Performance Assessment of Reverse Engineering Methods.pdf},
	journal = {Journal of Computational Biology},
	keywords = {Algorithms,Computational Biology,Computational Biology: methods,Computer Simulation,DRE,Gene Expression Profiling,Gene Expression Profiling: methods,gene regulatory networks,Gene Regulatory Networks,Genetic,Models,reverse engineering,Saccharomyces cerevisiae,Saccharomyces cerevisiae: genetics,Saccharomyces cerevisiae: metabolism,Software},
	number = {2},
	pmid = {19183003},
}

@Article{marbach_revealingstrengthsweaknesses_2010,
	title = {Revealing Strengths and Weaknesses of Methods for Gene Network Inference},
	author = {Daniel Marbach and Robert J Prill and Thomas Schaffter and Claudio Mattiussi and Dario Floreano and Gustavo Stolovitzky},
	year = {2010},
	month = {apr},
	volume = {107},
	pages = {6286--6291},
	issn = {1091-6490},
	doi = {10.1073/pnas.0913357107},
	abstract = {Numerous methods have been developed for inferring gene regulatory networks from expression data, however, both their absolute and comparative performance remain poorly understood. In this paper, we introduce a framework for critical performance assessment of methods for gene network inference. We present an in silico benchmark suite that we provided as a blinded, community-wide challenge within the context of the DREAM (Dialogue on Reverse Engineering Assessment and Methods) project. We assess the performance of 29 gene-network-inference methods, which have been applied independently by participating teams. Performance profiling reveals that current inference methods are affected, to various degrees, by different types of systematic prediction errors. In particular, all but the best-performing method failed to accurately infer multiple regulatory inputs (combinatorial regulation) of genes. The results of this community-wide experiment show that reliable network inference from gene expression data remains an unsolved problem, and they indicate potential ways of network reconstruction improvements.},
	file = {/home/rcannood/Zotero/storage/784SY83Y/Marbach et al. - 2010 - Revealing strengths and weaknesses of methods for gene network inference.pdf},
	journal = {Proceedings of the National Academy of Sciences},
	keywords = {Biometry,community experiment,Computational Biology,Computational Biology: methods,DREAM challenge,Gene Expression Profiling,Gene Regulatory Networks,performance assessment,reverse engineering,transcriptional regulatory networks},
	number = {14},
	pmid = {20308593},
}

@Article{marbach_wisdomcrowdsrobust_2012,
	title = {Wisdom of {{Crowds}} for {{Robust Gene Network Inference}}},
	author = {Daniel Marbach and James C Costello and Robert K{\"u}ffner and Nicole M Nicci Vega and Robert J Prill and Diogo M Camacho and Kyle R Allison and Manolis Kellis and James J Collins and Gustavo Stolovitzky and {DREAM5 Consortium} and Manolis Kellis and James J Collins and Gustavo Stolovitzky},
	year = {2012},
	month = {jul},
	volume = {9},
	pages = {796--804},
	issn = {1548-7091},
	doi = {10.1038/nmeth.2016},
	abstract = {Reconstructing gene regulatory networks from high-throughput data is a long-standing problem. Through the DREAM project (Dialogue on Reverse Engineering Assessment and Methods), we performed a comprehensive blind assessment of over thirty network inference methods on Escherichia coli, Staphylococcus aureus, Saccharomyces cerevisiae, and in silico microarray data. We characterize performance, data requirements, and inherent biases of different inference approaches offering guidelines for both algorithm application and development. We observe that no single inference method performs optimally across all datasets. In contrast, integration of predictions from multiple inference methods shows robust and high performance across diverse datasets. Thereby, we construct high-confidence networks for E. coli and S. aureus, each comprising \$\textbackslash backslash\$textasciitilde\{\}1700 transcriptional interactions at an estimated precision of 50\%. We experimentally test 53 novel interactions in E. coli, of which 23 were supported (43\%). Our results establish community-based methods as a powerful and robust tool for the inference of transcriptional gene regulatory networks.},
	file = {/home/rcannood/Zotero/storage/AQRNS57R/Marbach et al. - 2012 - Wisdom of crowds for robust gene network inference.pdf;/home/rcannood/Zotero/storage/KGJVQ9WL/Marbach et al. - 2012 - Wisdom of crowds for robust gene network inference.pdf},
	journal = {Nature methods},
	keywords = {Algorithms,Bacterial,Bacterial: genetics,Computational Biology,Escherichia coli,Escherichia coli: genetics,Gene Expression Regulation,Gene Regulatory Networks,Genetic,Genetic: genetics,Oligonucleotide Array Sequence Analysis,Saccharomyces cerevisiae,Saccharomyces cerevisiae: genetics,Software,Staphylococcus aureus,Staphylococcus aureus: genetics,Transcription},
	number = {8},
	pmid = {22796662},
}

@Article{mcfaline-figueroa_pooledsinglecellgenetic_2019,
	title = {A Pooled Single-Cell Genetic Screen Identifies Regulatory Checkpoints in the Continuum of the Epithelial-to-Mesenchymal Transition},
	author = {Jos{\a'e} L. McFaline-Figueroa and Andrew J. Hill and Xiaojie Qiu and Dana Jackson and Jay Shendure and Cole Trapnell},
	year = {2019},
	month = {sep},
	volume = {51},
	pages = {1389--1398},
	issn = {1546-1718},
	doi = {10.1038/s41588-019-0489-5},
	abstract = {Transcriptional profiling of 61,052 cells undergoing epithelial-to-mesenchymal transition (EMT) shows continuous waves of gene regulation, not discrete stages. A CRISPR screen orders EMT regulators along EMT progression.},
	copyright = {2019 The Author(s), under exclusive licence to Springer Nature America, Inc.},
	file = {/home/rcannood/Zotero/storage/RAQRP6UC/s41588-019-0489-5.html},
	journal = {Nature Genetics},
	language = {en},
	number = {9},
}

@Article{moerman_grnboost2arboretoefficient_2019,
	title = {{{GRNBoost2}} and {{Arboreto}}: Efficient and Scalable Inference of Gene Regulatory Networks},
	author = {Thomas Moerman and Sara {Aibar Santos} and Carmen {Bravo Gonz{\a'a}lez-Blas} and Jaak Simm and Yves Moreau and Jan Aerts and Stein Aerts},
	year = {2019},
	month = {jun},
	volume = {35},
	pages = {2159--2161},
	issn = {1367-4803},
	doi = {10.1093/bioinformatics/bty916},
	abstract = {Inferring a Gene Regulatory Network (GRN) from gene expression data is a computationally expensive task, exacerbated by increasing data sizes due to advances in high-throughput gene profiling technology, such as single-cell RNA-seq. To equip researchers with a toolset to infer GRNs from large expression datasets, we propose GRNBoost2 and the Arboreto framework. GRNBoost2 is an efficient algorithm for regulatory network inference using gradient boosting, based on the GENIE3 architecture. Arboreto is a computational framework that scales up GRN inference algorithms complying with this architecture. Arboreto includes both GRNBoost2 and an improved implementation of GENIE3, as a user-friendly open source Python package.Arboreto is available under the 3-Clause BSD license at http://arboreto.readthedocs.io.Supplementary data are available at Bioinformatics online.},
	journal = {Bioinformatics},
	number = {12},
}

@Article{papadopoulos_prossttprobabilisticsimulation_2019,
	title = {{{PROSSTT}}: Probabilistic Simulation of Single-Cell {{RNA}}-Seq Data for Complex Differentiation Processes},
	shorttitle = {{{PROSSTT}}},
	author = {Nikolaos Papadopoulos and Parra R. Gonzalo and Johannes S{\"o}ding},
	year = {2019},
	month = {sep},
	volume = {35},
	pages = {3517--3519},
	issn = {1367-4811},
	doi = {10.1093/bioinformatics/btz078},
	abstract = {SUMMARY: Cellular lineage trees can be derived from single-cell RNA sequencing snapshots of differentiating cells. Currently, only datasets with simple topologies are available. To test and further develop tools for lineage tree reconstruction, we need test datasets with known complex topologies. PROSSTT can simulate scRNA-seq datasets for differentiation processes with lineage trees of any desired complexity, noise level, noise model and size. PROSSTT also provides scripts to quantify the quality of predicted lineage trees. AVAILABILITY AND IMPLEMENTATION: https://github.com/soedinglab/prosstt. SUPPLEMENTARY INFORMATION: Supplementary data are available at Bioinformatics online.},
	file = {/home/rcannood/Zotero/storage/UERERPJX/Papadopoulos et al. - 2019 - PROSSTT probabilistic simulation of single-cell R.pdf},
	journal = {Bioinformatics (Oxford, England)},
	keywords = {Cell Differentiation,Gene Expression Profiling,RNA-Seq,Single-Cell Analysis,Software},
	language = {eng},
	number = {18},
	pmcid = {PMC6748774},
	pmid = {30715210},
}

@Article{parra_reconstructingcomplexlineage_2019,
	title = {Reconstructing Complex Lineage Trees from {{scRNA}}-Seq Data Using {{MERLoT}}},
	author = {R. Gonzalo Parra and Nikolaos Papadopoulos and Laura Ahumada-Arranz and Jakob El Kholtei and Noah Mottelson and Yehor Horokhovsky and Barbara Treutlein and Johannes Soeding},
	year = {2019},
	month = {sep},
	volume = {47},
	pages = {8961--8974},
	issn = {1362-4962},
	doi = {10.1093/nar/gkz706},
	abstract = {Advances in single-cell transcriptomics techniques are revolutionizing studies of cellular differentiation and heterogeneity. It has become possible to track the trajectory of thousands of genes across the cellular lineage trees that represent the temporal emergence of cell types during dynamic processes. However, reconstruction of cellular lineage trees with more than a few cell fates has proved challenging. We present MERLoT (https://github.com/soedinglab/merlot), a flexible and user-friendly tool to reconstruct complex lineage trees from single-cell transcriptomics data. It can impute temporal gene expression profiles along the reconstructed tree. We show MERLoT's capabilities on various real cases and hundreds of simulated datasets.},
	file = {/home/rcannood/Zotero/storage/86VN2MRY/Parra et al. - 2019 - Reconstructing complex lineage trees from scRNA-se.pdf},
	journal = {Nucleic Acids Research},
	keywords = {Algorithms,Animals,Cell Differentiation,Cell Lineage,Cluster Analysis,Gene Expression Profiling,Hematopoietic Stem Cells,Humans,Mice,Mouse Embryonic Stem Cells,Sequence Analysis; RNA,Single-Cell Analysis,Software,Time Factors,Transcriptome},
	language = {eng},
	number = {17},
	pmcid = {PMC6755101},
	pmid = {31428793},
}

@Manual{patil_pairwisecomparisonsmultiplepairwise_2019,
	title = {{{pairwiseComparisons}}: {{Multiple}} Pairwise Comparison Tests},
	author = {Indrajeet Patil},
	year = {2019},
	type = {Manual},
}

@Article{prill_rigorousassessmentsystems_2010,
	title = {Towards a {{Rigorous Assessment}} of {{Systems Biology Models}}: {{The DREAM3 Challenges}}},
	author = {Robert J Prill and Marbach Daniel and Saez-Rodriguez Julio and Sorger Peter K. and Alexopoulos Leonidas G. and Xue Xiaowei and Clarke Neil D. and Altan-Bonnet Gregoire and Stolovitzky Gustavo and Daniel Marbach and Julio Saez-Rodriguez and Peter K Sorger and Leonidas G Alexopoulos and Xiaowei Xue and Neil D Clarke and Gregoire Altan-Bonnet and Gustavo Stolovitzky and Marbach Daniel and Saez-Rodriguez Julio and Sorger Peter K. and Alexopoulos Leonidas G. and Xue Xiaowei and Clarke Neil D. and Altan-Bonnet Gregoire and Stolovitzky Gustavo},
	year = {2010},
	month = {jan},
	volume = {5},
	pages = {e9202},
	issn = {1932-6203},
	doi = {10.1371/journal.pone.0009202},
	abstract = {BACKGROUND: Systems biology has embraced computational modeling in response to the quantitative nature and increasing scale of contemporary data sets. The onslaught of data is accelerating as molecular profiling technology evolves. The Dialogue for Reverse Engineering Assessments and Methods (DREAM) is a community effort to catalyze discussion about the design, application, and assessment of systems biology models through annual reverse-engineering challenges. METHODOLOGY AND PRINCIPAL FINDINGS: We describe our assessments of the four challenges associated with the third DREAM conference which came to be known as the DREAM3 challenges: signaling cascade identification, signaling response prediction, gene expression prediction, and the DREAM3 in silico network challenge. The challenges, based on anonymized data sets, tested participants in network inference and prediction of measurements. Forty teams submitted 413 predicted networks and measurement test sets. Overall, a handful of best-performer teams were identified, while a majority of teams made predictions that were equivalent to random. Counterintuitively, combining the predictions of multiple teams (including the weaker teams) can in some cases improve predictive power beyond that of any single method. CONCLUSIONS: DREAM provides valuable feedback to practitioners of systems biology modeling. Lessons learned from the predictions of the community provide much-needed context for interpreting claims of efficacy of algorithms described in the scientific literature.},
	file = {/home/rcannood/Zotero/storage/Q5E22H2F/Prill et al. - 2010 - Towards a Rigorous Assessment of Systems Biology Models The DREAM3 Challenges.pdf},
	journal = {PLoS ONE},
	keywords = {Algorithms,Animals,Biological,Cluster Analysis,Computational Biology,Computational Biology: methods,Gene Expression Profiling,Gene Expression Profiling: methods,Gene Regulatory Networks,Humans,Models,Protein Interaction Mapping,Protein Interaction Mapping: methods,Reproducibility of Results,Signal Transduction,Software,Systems Biology,Systems Biology: methods},
	number = {2},
	pmid = {20186320},
}

@Article{rekhtman_directinteractionhematopoietic_1999,
	title = {Direct Interaction of Hematopoietic Transcription Factors {{PU}}.1 and {{GATA}}-1: Functional Antagonism in Erythroid Cells},
	shorttitle = {Direct Interaction of Hematopoietic Transcription Factors {{PU}}.1 and {{GATA}}-1},
	author = {N. Rekhtman and F. Radparvar and T. Evans and A. I. Skoultchi},
	year = {1999},
	month = {jun},
	volume = {13},
	pages = {1398--1411},
	issn = {0890-9369},
	doi = {10.1101/gad.13.11.1398},
	abstract = {Malignant transformation usually inhibits terminal cell differentiation but the precise mechanisms involved are not understood. PU.1 is a hematopoietic-specific Ets family transcription factor that is required for development of some lymphoid and myeloid lineages. PU.1 can also act as an oncoprotein as activation of its expression in erythroid precursors by proviral insertion or transgenesis causes erythroleukemias in mice. Restoration of terminal differentiation in the mouse erythroleukemia (MEL) cells requires a decline in the level of PU.1, indicating that PU.1 can block erythroid differentiation. Here we investigate the mechanism by which PU.1 interferes with erythroid differentiation. We find that PU.1 interacts directly with GATA-1, a zinc finger transcription factor required for erythroid differentiation. Interaction between PU.1 and GATA-1 requires intact DNA-binding domains in both proteins. PU.1 represses GATA-1-mediated transcriptional activation. Both the DNA binding and transactivation domains of PU.1 are required for repression and both domains are also needed to block terminal differentiation in MEL cells. We also show that ectopic expression of PU.1 in Xenopus embryos is sufficient to block erythropoiesis during normal development. Furthermore, introduction of exogenous GATA-1 in both MEL cells and Xenopus embryos and explants relieves the block to erythroid differentiation imposed by PU.1. Our results indicate that the stoichiometry of directly interacting but opposing transcription factors may be a crucial determinant governing processes of normal differentiation and malignant transformation.},
	file = {/home/rcannood/Zotero/storage/77DJ34BY/Rekhtman et al. - 1999 - Direct interaction of hematopoietic transcription .pdf},
	journal = {Genes \& Development},
	keywords = {3T3 Cells,Animals,Binding Sites,Cell Differentiation,DNA-Binding Proteins,Erythroid Precursor Cells,Erythroid-Specific DNA-Binding Factors,GATA1 Transcription Factor,HeLa Cells,Hematopoiesis,Humans,Mice,Proto-Oncogene Proteins,Recombinant Fusion Proteins,Trans-Activators,Transcription Factors,Transcription; Genetic,Transcriptional Activation,Xenopus laevis,Zinc Fingers},
	language = {eng},
	number = {11},
	pmcid = {PMC316770},
	pmid = {10364157},
}

@Article{roy_systemgeneratingtranscription_2008,
	title = {A System for Generating Transcription Regulatory Networks with Combinatorial Control of Transcription},
	author = {Sushmita Roy and Margaret Werner-Washburne and Terran Lane},
	year = {2008},
	month = {may},
	volume = {24},
	pages = {1318--1320},
	issn = {1367-4803},
	doi = {10.1093/bioinformatics/btn126},
	abstract = {Summary: We have developed a new software system, REgulatory Network generator with COmbinatorial control (RENCO), for automatic generation of differential equations describing pre-transcriptional combinatorics in artificial regulatory networks. RENCO has the following benefits: (a) it explicitly models protein\textendash protein interactions among transcription factors, (b) it captures combinatorial control of transcription factors on target genes and (c) it produces output in Systems Biology Markup Language (SBML) format, which allows these equations to be directly imported into existing simulators. Explicit modeling of the protein interactions allows RENCO to incorporate greater mechanistic detail of the transcription machinery compared to existing models and can provide a better assessment of algorithms for regulatory network inference.Availability: RENCO is a C++ command line program, available at http://sourceforge.net/projects/renco/Contact:terran@cs.unm.eduSupplementary information: Supplementary data are available at Bioinformatics online.},
	journal = {Bioinformatics},
	number = {10},
}

@Article{saelens_comparisonsinglecelltrajectory_2019,
	title = {A Comparison of Single-Cell Trajectory Inference Methods},
	author = {Wouter Saelens and Robrecht Cannoodt and Helena Todorov and Yvan Saeys},
	year = {2019},
	volume = {37},
	issn = {15461696},
	doi = {10.1038/s41587-019-0071-9},
	abstract = {Trajectory inference approaches analyze genome-wide omics data from thousands of single cells and computationally infer the order of these cells along developmental trajectories. Although more than 70 trajectory inference tools have already been developed, it is challenging to compare their performance because the input they require and output models they produce vary substantially. Here, we benchmark 45 of these methods on 110 real and 229 synthetic datasets for cellular ordering, topology, scalability and usability. Our results highlight the complementarity of existing tools, and that the choice of method should depend mostly on the dataset dimensions and trajectory topology. Based on these results, we develop a set of guidelines to help users select the best method for their dataset. Our freely available data and evaluation pipeline ( https://benchmark.dynverse.org ) will aid in the development of improved tools designed to analyze increasingly large and complex single-cell datasets.},
	file = {/home/rcannood/Zotero/storage/JKZCFAVA/Saelens et al. - 2019 - A comparison of single-cell trajectory inference methods.pdf;/home/rcannood/Zotero/storage/T9HMNIDB/Saelens2019.pdf},
	journal = {Nature Biotechnology},
	number = {May},
}

@Article{schaffter_genenetweaversilicobenchmark_2011,
	title = {{{GeneNetWeaver}}: In Silico Benchmark Generation and Performance Profiling of Network Inference Methods.},
	shorttitle = {\{\vphantom\}{{GeneNetWeaver}}\vphantom\{\}},
	author = {Thomas Schaffter and Daniel Marbach and Dario Floreano},
	year = {2011},
	month = {aug},
	volume = {27},
	pages = {2263--2270},
	issn = {1367-4811},
	doi = {10.1093/bioinformatics/btr373},
	abstract = {MOTIVATION: Over the last decade, numerous methods have been developed for inference of regulatory networks from gene expression data. However, accurate and systematic evaluation of these methods is hampered by the difficulty of constructing adequate benchmarks and the lack of tools for a differentiated analysis of network predictions on such benchmarks. RESULTS: Here, we describe a novel and comprehensive method for in silico benchmark generation and performance profiling of network inference methods available to the community as an open-source software called GeneNetWeaver (GNW). In addition to the generation of detailed dynamical models of gene regulatory networks to be used as benchmarks, GNW provides a network motif analysis that reveals systematic prediction errors, thereby indicating potential ways of improving inference methods. The accuracy of network inference methods is evaluated using standard metrics such as precision-recall and receiver operating characteristic curves. We show how GNW can be used to assess the performance and identify the strengths and weaknesses of six inference methods. Furthermore, we used GNW to provide the international Dialogue for Reverse Engineering Assessments and Methods (DREAM) competition with three network inference challenges (DREAM3, DREAM4 and DREAM5). AVAILABILITY: GNW is available at http://gnw.sourceforge.net along with its Java source code, user manual and supporting data. SUPPLEMENTARY INFORMATION: Supplementary data are available at Bioinformatics online. CONTACT: dario.floreano@epfl.ch.},
	file = {/home/rcannood/Zotero/storage/P5JMJUBN/Schaffter, Marbach, Floreano - 2011 - GeneNetWeaver in silico benchmark generation and performance profiling of network inference method.pdf},
	journal = {Bioinformatics},
	keywords = {Benchmarking,Escherichia coli,Escherichia coli: genetics,Gene Expression,Gene Regulatory Networks,Genetic,Models,Software},
	number = {16},
	pmid = {21697125},
}

@Article{schilstra_biologicgeneexpression_2008,
	title = {Bio-{{Logic}}: {{Gene Expression}} and the {{Laws}} of {{Combinatorial Logic}}},
	shorttitle = {Bio-{{Logic}}},
	author = {Maria J. Schilstra and Chrystopher L. Nehaniv},
	year = {2008},
	month = {jan},
	volume = {14},
	pages = {121--133},
	issn = {1064-5462},
	doi = {10.1162/artl.2008.14.1.121},
	abstract = {Abstract At the heart of the development of fertilized eggs into fully formed organisms and the adaptation of cells to changed conditions are genetic regulatory networks (GRNs). In higher multicellular organisms, signal selection and multiplexing are performed at the cis-regulatory domains of genes, where combinations of transcription factors (TFs) regulate the rates at which the genes are transcribed into mRNA. To be able to act as activators or repressors of gene transcription, TFs must first bind to target sequences on the regulatory domains. Two TFs that act in concert may bind entirely independently of each other, but more often binding of the first one will alter the affinity of the other for its binding site. This article presents a systematic investigation into the effect of TF binding dependences on the predicted regulatory function of this bio-logic. Four extreme scenarios, commonly used to classify enzyme activation and inhibition patterns, for the binding of two TFs were explored: independent (the TFs bind without affecting each other's affinities), competitive (the TFs compete for the same binding site), ordered (the TFs bind in a compulsory order), and joint binding (the TFs either bind as a preformed complex, or binding of one is virtually impossible in the absence of the other). The conclusions are: (1) the laws of combinatorial logic hold only for systems with independently binding TFs; (2) systems formed according to the other scenarios can mimic the functions of their Boolean logical counterparts, but cannot be combined or decomposed in the same way; and (3) the continuously scaled output of systems consisting of competitively binding activators and repressors can be controlled more robustly than that of single TF or (quasi-)logical multi-TF systems.},
	file = {/home/rcannood/Zotero/storage/C7UWQGI3/Schilstra and Nehaniv - 2008 - Bio-Logic Gene Expression and the Laws of Combina.pdf;/home/rcannood/Zotero/storage/KXDLUMAG/artl.2008.14.1.html},
	journal = {Artificial Life},
	number = {1},
}

@Article{schwanhausser_globalquantificationmammalian_2011,
	title = {Global Quantification of Mammalian Gene Expression Control},
	author = {Bj{\"o}rn Schwanh{\"a}usser and Dorothea Busse and Na Li and Gunnar Dittmar and Johannes Schuchhardt and Jana Wolf and Wei Chen and Matthias Selbach},
	year = {2011},
	month = {may},
	volume = {473},
	pages = {337--342},
	issn = {1476-4687},
	doi = {10.1038/nature10098},
	abstract = {Gene expression is a multistep process that involves the transcription, translation and turnover of messenger RNAs and proteins. Although it is one of the most fundamental processes of life, the entire cascade has never been quantified on a genome-wide scale. Here we simultaneously measured absolute mRNA and protein abundance and turnover by parallel metabolic pulse labelling for more than 5,000 genes in mammalian cells. Whereas mRNA and protein levels correlated better than previously thought, corresponding half-lives showed no correlation. Using a quantitative model we have obtained the first genome-scale prediction of synthesis rates of mRNAs and proteins. We find that the cellular abundance of proteins is predominantly controlled at the level of translation. Genes with similar combinations of mRNA and protein stability shared functional properties, indicating that half-lives evolved under energetic and dynamic constraints. Quantitative information about all stages of gene expression provides a rich resource and helps to provide a greater understanding of the underlying design principles.},
	journal = {Nature},
	number = {7347},
}

@Article{smith_spatialstochasticintracellular_2019,
	title = {Spatial {{Stochastic Intracellular Kinetics}}: {{A Review}} of {{Modelling Approaches}}},
	author = {Stephen Smith and Ramon Grima},
	year = {2019},
	month = {aug},
	volume = {81},
	pages = {2960--3009},
	issn = {1522-9602},
	doi = {10.1007/s11538-018-0443-1},
	abstract = {Models of chemical kinetics that incorporate both stochasticity and diffusion are an increasingly common tool for studying biology. The variety of competing models is vast, but two stand out by virtue of their popularity: the reaction\textendash diffusion master equation and Brownian dynamics. In this review, we critically address a number of open questions surrounding these models: How can they be justified physically? How do they relate to each other? How do they fit into the wider landscape of chemical models, ranging from the rate equations to molecular dynamics? This review assumes no prior knowledge of modelling chemical kinetics and should be accessible to a wide range of readers.},
	journal = {Bulletin of Mathematical Biology},
	number = {8},
}

@Article{soneson_biasrobustnessscalability_2018,
	title = {Bias, Robustness and Scalability in Single-Cell Differential Expression Analysis},
	author = {Charlotte Soneson and Mark D. Robinson},
	year = {2018},
	month = {apr},
	volume = {15},
	pages = {255--261},
	issn = {1548-7105},
	doi = {10.1038/nmeth.4612},
	abstract = {Many methods have been used to determine differential gene expression from single-cell RNA (scRNA)-seq data. We evaluated 36 approaches using experimental and synthetic data and found considerable differences in the number and characteristics of the genes that are called differentially expressed. Prefiltering of lowly expressed genes has important effects, particularly for some of the methods developed for bulk RNA-seq data analysis. However, we found that bulk RNA-seq analysis methods do not generally perform worse than those developed specifically for scRNA-seq. We also present conquer, a repository of consistently processed, analysis-ready public scRNA-seq data sets that is aimed at simplifying method evaluation and reanalysis of published results. Each data set provides abundance estimates for both genes and transcripts, as well as quality control and exploratory analysis reports.},
	journal = {Nature Methods},
	keywords = {Computational Biology,Gene Expression Profiling,Gene Expression Regulation,High-Throughput Nucleotide Sequencing,RNA,Sequence Analysis; RNA,Single-Cell Analysis,Software},
	language = {eng},
	number = {4},
	pmid = {29481549},
}

@Article{street_slingshotcelllineage_2018,
	title = {Slingshot: {{Cell Lineage}} and {{Pseudotime Inference}} for {{Single}}-{{Cell Transcriptomics}}},
	shorttitle = {Slingshot},
	author = {Kelly Street and Davide Risso and Russell B Fletcher and Diya Das and John Ngai and Nir Yosef and Elizabeth Purdom and Sandrine Dudoit},
	year = {2018},
	month = {jun},
	volume = {19},
	pages = {477},
	issn = {1471-2164},
	doi = {10.1186/s12864-018-4772-0},
	abstract = {Single-cell transcriptomics allows researchers to investigate complex communities of heterogeneous cells. It can be applied to stem cells and their descendants in order to chart the progression from multipotent progenitors to fully differentiated cells. While a variety of statistical and computational methods have been proposed for inferring cell lineages, the problem of accurately characterizing multiple branching lineages remains difficult to solve.},
	journal = {BMC Genomics},
	number = {1},
}

@Article{vallejos_normalizingsinglecellrna_2017,
	title = {Normalizing Single-Cell {{RNA}} Sequencing Data: Challenges and Opportunities},
	shorttitle = {Normalizing Single-Cell {{RNA}} Sequencing Data},
	author = {Catalina A. Vallejos and Davide Risso and Antonio Scialdone and Sandrine Dudoit and John C. Marioni},
	year = {2017},
	month = {jun},
	volume = {14},
	pages = {565--571},
	publisher = {{Nature Publishing Group}},
	issn = {1548-7105},
	doi = {10.1038/nmeth.4292},
	abstract = {This Perspective examines single-cell RNA-seq data challenges and the need for normalization methods designed specifically for single-cell data in order to remove technical biases.},
	copyright = {2017 Nature Publishing Group, a division of Macmillan Publishers Limited. All Rights Reserved.},
	journal = {Nature Methods},
	language = {en},
	number = {6},
}

@Article{vandenbulcke_syntrengeneratorsynthetic_2006,
	title = {{{SynTReN}}: A Generator of Synthetic Gene Expression Data for Design and Analysis of Structure Learning Algorithms},
	shorttitle = {{{SynTReN}}},
	author = {Tim {Van den Bulcke} and Koenraad {Van Leemput} and Bart Naudts and Piet {van Remortel} and Hongwu Ma and Alain Verschoren and Bart {De Moor} and Kathleen Marchal},
	year = {2006},
	month = {jan},
	volume = {7},
	pages = {43},
	issn = {1471-2105},
	doi = {10.1186/1471-2105-7-43},
	abstract = {The development of algorithms to infer the structure of gene regulatory networks based on expression data is an important subject in bioinformatics research. Validation of these algorithms requires benchmark data sets for which the underlying network is known. Since experimental data sets of the appropriate size and design are usually not available, there is a clear need to generate well-characterized synthetic data sets that allow thorough testing of learning algorithms in a fast and reproducible manner.},
	journal = {BMC Bioinformatics},
	language = {en},
	number = {1},
}

@Article{vandesande_scalablescenicworkflow_2020,
	title = {A Scalable {{SCENIC}} Workflow for Single-Cell Gene Regulatory Network Analysis},
	author = {Bram {Van de Sande} and Christopher Flerin and Kristofer Davie and Maxime {De Waegeneer} and Gert Hulselmans and Sara Aibar and Ruth Seurinck and Wouter Saelens and Robrecht Cannoodt and Quentin Rouchon and Toni Verbeiren and Dries {De Maeyer} and Joke Reumers and Yvan Saeys and Stein Aerts},
	year = {2020},
	month = {jul},
	volume = {15},
	pages = {2247--2276},
	issn = {1750-2799},
	doi = {10.1038/s41596-020-0336-2},
	abstract = {This protocol explains how to perform a fast SCENIC analysis alongside standard best practices steps on single-cell RNA-sequencing data using software containers and Nextflow pipelines. SCENIC reconstructs regulons (i.e., transcription factors and their target genes) assesses the activity of these discovered regulons in individual cells and uses these cellular activity patterns to find meaningful clusters of cells. Here we present an improved version of SCENIC with several advances. SCENIC has been refactored and reimplemented in Python (pySCENIC), resulting in a tenfold increase in speed, and has been packaged into containers for ease of use. It is now also possible to use epigenomic track databases, as well as motifs, to refine regulons. In this protocol, we explain the different steps of SCENIC: the workflow starts from the count matrix depicting the gene abundances for all cells and consists of three stages. First, coexpression modules are inferred using a regression per-target approach (GRNBoost2). Next, the indirect targets are pruned from these modules using cis-regulatory motif discovery (cisTarget). Lastly, the activity of these regulons is quantified via an enrichment score for the regulon's target genes (AUCell). Nonlinear projection methods can be used to display visual groupings of cells based on the cellular activity patterns of these regulons. The results can be exported as a loom file and visualized in the SCope web application. This protocol is illustrated on two use cases: a peripheral blood mononuclear cell data set and a panel of single-cell RNA-sequencing cancer experiments. For a data set of 10,000 genes and 50,000 cells, the pipeline runs in {$<$}2 h.},
	journal = {Nature Protocols},
	number = {7},
}

@Article{vieth_powsimrpoweranalysis_2017,
	title = {{{powsimR}}: Power Analysis for Bulk and Single Cell {{RNA}}-Seq Experiments},
	shorttitle = {{{powsimR}}},
	author = {Beate Vieth and Christoph Ziegenhain and Swati Parekh and Wolfgang Enard and Ines Hellmann},
	year = {2017},
	month = {nov},
	volume = {33},
	pages = {3486--3488},
	issn = {1367-4803},
	doi = {10.1093/bioinformatics/btx435},
	abstract = {AbstractSummary.  Power analysis is essential to optimize the design of RNA-seq experiments and to assess and compare the power to detect differentially express},
	file = {/home/rcannood/Zotero/storage/MFRF86MZ/Vieth et al. - 2017 - powsimR power analysis for bulk and single cell R.pdf;/home/rcannood/Zotero/storage/HSCUJRI8/3952669.html},
	journal = {Bioinformatics},
	language = {en},
	number = {21},
}

@Article{wang_quantifyingwaddingtonlandscape_2011,
	title = {Quantifying the {{Waddington Landscape}} and {{Biological Paths}} for {{Development}} and {{Differentiation}}},
	author = {Jin Wang and Kun Zhang and Li Xu and Erkang Wang},
	year = {2011},
	month = {may},
	volume = {108},
	pages = {8257--8262},
	issn = {0027-8424, 1091-6490},
	doi = {10.1073/pnas.1017017108},
	abstract = {We developed a theoretical framework to prove the existence and quantify the Waddington landscape as well as chreode-biological paths for development and differentiation. The cells can have states with the higher probability ones giving the different cell types. Different cell types correspond to different basins of attractions of the probability landscape. We study how the cells develop from undifferentiated cells to differentiated cells from landscape perspectives. We quantified the Waddington landscape through construction of underlying probability landscape for cell development. We show the developmental process proceeds as moving from undifferentiated to the differentiated basins of attractions. The barrier height of the basins of attractions correlates with the escape time that determines the stability of cell types. We show that the developmental process can be quantitatively described and uncovered by the biological paths on the quantified Waddington landscape from undifferentiated to the differentiated cells. We found the dynamics of the developmental process is controlled by a combination of the gradient and curl force on the landscape. The biological paths often do not follow the steepest descent path on the landscape. The landscape framework also quantifies the possibility of reverse differentiation process such as cell reprogramming from differentiated cells back to the original stem cell. We show that the biological path of reverse differentiation is irreversible and different from the one for differentiation process. We found that the developmental process described by the underlying landscape and the associated biological paths is relatively stable and robust against the influences of environmental perturbations.},
	journal = {Proceedings of the National Academy of Sciences},
	number = {20},
	pmid = {21536909},
}

@Article{weber_essentialguidelinescomputational_2019,
	title = {Essential Guidelines for Computational Method Benchmarking},
	author = {Lukas M. Weber and Wouter Saelens and Robrecht Cannoodt and Charlotte Soneson and Alexander Hapfelmeier and Paul P. Gardner and Anne-Laure Boulesteix and Yvan Saeys and Mark D. Robinson},
	year = {2019},
	month = {jun},
	volume = {20},
	pages = {125},
	issn = {1474-760X},
	doi = {10.1186/s13059-019-1738-8},
	abstract = {In computational biology and other sciences, researchers are frequently faced with a choice between several computational methods for performing data analyses. Benchmarking studies aim to rigorously compare the performance of different methods using well-characterized benchmark datasets, to determine the strengths of each method or to provide recommendations regarding suitable choices of methods for an analysis. However, benchmarking studies must be carefully designed and implemented to provide accurate, unbiased, and informative results. Here, we summarize key practical guidelines and recommendations for performing high-quality benchmarking analyses, based on our experiences in computational biology.},
	file = {/home/rcannood/Zotero/storage/83LMG2SG/Weber et al. - 2019 - Essential guidelines for computational method benc.pdf;/home/rcannood/Zotero/storage/GL7G9MIR/s13059-019-1738-8.html},
	journal = {Genome Biology},
	number = {1},
}

@Article{wickham_welcometidyverse_2019,
	title = {Welcome to the {{Tidyverse}}},
	author = {Hadley Wickham and Mara Averick and Jennifer Bryan and Winston Chang and Lucy McGowan and Romain Fran{\c c}ois and Garrett Grolemund and Alex Hayes and Lionel Henry and Jim Hester and Max Kuhn and Thomas Pedersen and Evan Miller and Stephan Bache and Kirill M{\"u}ller and Jeroen Ooms and David Robinson and Dana Seidel and Vitalie Spinu and Kohske Takahashi and Davis Vaughan and Claus Wilke and Kara Woo and Hiroaki Yutani},
	year = {2019},
	month = {nov},
	doi = {10.21105/joss.01686},
	abstract = {Wickham et al., (2019). Welcome to the Tidyverse. Journal of Open Source Software, 4(43), 1686, https://doi.org/10.21105/joss.01686},
	file = {/home/rcannood/Zotero/storage/TVYSNC4J/Wickham et al. - 2019 - Welcome to the Tidyverse.pdf;/home/rcannood/Zotero/storage/EAQ2VYF8/joss.html},
	language = {en},
}

@Article{xu_regulationbifurcatingcell_2015,
	title = {Regulation of Bifurcating {{B}} Cell Trajectories by Mutual Antagonism between Transcription Factors {{IRF4}} and {{IRF8}}},
	author = {Heping Xu and Virendra K Chaudhri and Zhiguo Wu and Konstantinos Biliouris and Krista Dienger-Stambaugh and Yrina Rochman and Harinder Singh},
	year = {2015},
	month = {dec},
	volume = {16},
	pages = {1274--1281},
	abstract = {Upon recognition of antigen, B cells undertake a bifurcated response in which some cells rapidly differentiate into plasmablasts while others undergo affinity maturation in germinal centers (GCs). Here we identified a double-negative feedback loop between the transcription factors IRF4 and IRF8 that regulated the initial developmental bifurcation of activated B cells as well as the GC response. IRF8 dampened signaling via the B cell antigen receptor (BCR), facilitated antigen-specific interaction with helper T cells, and promoted antibody affinity maturation while antagonizing IRF4-driven differentiation of plasmablasts. Genomic analysis revealed concentration-dependent actions of IRF4 and IRF8 in regulating distinct gene-expression programs. Stochastic modeling suggested that the double-negative feedback was sufficient to initiate bifurcation of the B cell developmental trajectories.},
	journal = {Nat. Immunol.},
	number = {12},
}

@Article{yosef_dynamicregulatorynetwork_2013,
	title = {Dynamic Regulatory Network Controlling \{\vphantom\}{{TH17}}\vphantom\{\} Cell Differentiation},
	author = {Nir Yosef and Alex K Shalek and Jellert T Gaublomme and Hulin Jin and Youjin Lee and Amit Awasthi and Chuan Wu and Katarzyna Karwacz and Sheng Xiao and Marsela Jorgolli and David Gennert and Rahul Satija and Arvind Shakya and Diana Y Lu and John J Trombetta and Meenu R Pillai and Peter J Ratcliffe and Mathew L Coleman and Mark Bix and Dean Tantin and Hongkun Park and Vijay K Kuchroo and Aviv Regev},
	year = {2013},
	volume = {496},
	pages = {461--468},
	abstract = {Despite their importance, the molecular circuits that control the differentiation of naive T cells remain largely unknown. Recent studies that reconstructed regulatory networks in mammalian cells have focused on short-term responses and relied on perturbation-based approaches that cannot be readily applied to primary T cells. Here we combine transcriptional profiling at high temporal resolution, novel computational algorithms, and innovative nanowire-based perturbation tools to systematically derive and experimentally validate a model of the dynamic regulatory network that controls the differentiation of mouse TH17 cells, a proinflammatory T-cell subset that has been implicated in the pathogenesis of multiple autoimmune diseases. The TH17 transcriptional network consists of two self-reinforcing, but mutually antagonistic, modules, with 12 novel regulators, the coupled action of which may be essential for maintaining the balance between TH17 and other CD4(+) T-cell subsets. Our study identifies and validates 39 regulatory factors, embeds them within a comprehensive temporal network and reveals its organizational principles; it also highlights novel drug targets for controlling TH17 cell differentiation.},
	journal = {Nature},
	number = {7446},
}

@Article{zappia_splattersimulationsinglecell_2017,
	title = {Splatter: {{Simulation}} of {{Single}}-{{Cell RNA Sequencing Data}}},
	shorttitle = {Splatter},
	author = {Luke Zappia and Belinda Phipson and Alicia Oshlack},
	year = {2017},
	month = {sep},
	volume = {18},
	pages = {174},
	issn = {1474-760X},
	doi = {10.1186/s13059-017-1305-0},
	abstract = {As single-cell RNA sequencing (scRNA-seq) technologies have rapidly developed, so have analysis methods. Many methods have been tested, developed, and validated using simulated datasets. Unfortunately, current simulations are often poorly documented, their similarity to real data is not demonstrated, or reproducible code is not available. Here, we present the Splatter Bioconductor package for simple, reproducible, and well-documented simulation of scRNA-seq data. Splatter provides an interface to multiple simulation methods including Splat, our own simulation, based on a gamma-Poisson distribution. Splat can simulate single populations of cells, populations with multiple cell types, or differentiation paths.},
	journal = {Genome Biology},
	keywords = {RNA-seq,Simulation,Single-cell,Software},
}

@Article{zeisel_coupledpremrnamrna_2011,
	title = {Coupled Pre-{{mRNA}} and {{mRNA}} Dynamics Unveil Operational Strategies Underlying Transcriptional Responses to Stimuli},
	author = {Amit Zeisel and Wolfgang J K{\"o}stler and Natali Molotski and Jonathan M Tsai and Rita Krauthgamer and Jasmine Jacob-Hirsch and Gideon Rechavi and Yoav Soen and Steffen Jung and Yosef Yarden and Eytan Domany},
	year = {2011},
	month = {jan},
	volume = {7},
	pages = {529},
	issn = {1744-4292},
	doi = {10.1038/msb.2011.62},
	abstract = {Transcriptional responses to extracellular stimuli involve tuning the rates of transcript production and degradation. Here, we show that the time-dependent profiles of these rates can be inferred from simultaneous measurements of precursor mRNA (pre-mRNA) and mature mRNA profiles. Transcriptome-wide measurements demonstrate that genes with similar mRNA profiles often exhibit marked differences in the amplitude and onset of their production rate. The latter is characterized by a large dynamic range, with a group of genes exhibiting an unexpectedly strong transient production overshoot, thereby accelerating their induction and, when combined with time-dependent degradation, shaping transient responses with precise timing and amplitude.},
	file = {/home/rcannood/Zotero/storage/QRQW86Y9/Zeisel et al. - 2011 - Coupled pre-mRNA and mRNA dynamics unveil operatio.pdf;/home/rcannood/Zotero/storage/R8ZG26IL/msb.2011.html},
	journal = {Molecular Systems Biology},
	keywords = {half-life,operational strategy,pre-mRNA,transcriptional response},
	number = {1},
}

@Article{zhang_simulatingmultiplefaceted_2019,
	title = {Simulating Multiple Faceted Variability in Single Cell {{RNA}} Sequencing},
	author = {Xiuwei Zhang and Chenling Xu and Nir Yosef},
	year = {2019},
	month = {jun},
	volume = {10},
	pages = {1--16},
	issn = {2041-1723},
	doi = {10.1038/s41467-019-10500-w},
	abstract = {Simulated single cell RNA sequencing data is useful for method development and comparison. Here, the authors developed SymSim, a simulator that explicitly models the main factors of variation in single cell data.},
	copyright = {2019 The Author(s)},
	file = {/home/rcannood/Zotero/storage/YLQYCW6E/Zhang et al. - 2019 - Simulating multiple faceted variability in single .pdf;/home/rcannood/Zotero/storage/EXJ96Y59/s41467-019-10500-w.html},
	journal = {Nature Communications},
	language = {en},
	number = {1},
}
\end{filecontents}

\end{document}

### 2021-05-12 manuscript/supplementary_files.Rmd
rework manuscript

--- 2021-03-25 manuscript/render.Rmd
@@
 ---
-title: "Response to reviewers' comments"
 output:
   pdf_document:
-    keep_tex: yes
+    keep_tex: no
     latex_engine: xelatex
@@
 ---
-<!-- Notes (remove this before submitting)
-dyngen manuscript v4  
-github.com/dynverse/dyngen  
-github.com/dynverse/dyngen_manuscript
-dyngen.dynverse.org
 
-Antwoord naar editor: wordt gebruikt in dynverse, tradeseq, cellrank, BEELINE
--->
+```{r setup, include=FALSE}
+ # for lua files regarding author information, see https://stackoverflow.com/a/52919256
+knitr::opts_chunk$set(collapse=TRUE, results="hold")
+library(tidyverse)
+```
 
-# dyngen rebuttal
+\newcommand{\x}[1]{\text{x}_{#1}}
+\newcommand{\y}[1]{\text{y}_{#1}}
+\newcommand{\z}[1]{\text{z}_{#1}}
 
-We would like to thank the reviewer for the constructive comments, which has motivated us to greatly improve the experiments performed in this work and the documentation provided for the tool.
-Notable changes are:
+\newcommand{\rs}[1]{\text{R}_{#1}}
+\newcommand{\rp}[1]{\text{R}^+_{#1}}
+\newcommand{\rn}[1]{\text{R}^-_{#1}}
 
- * We created a documentation website available at [dyngen.dynverse.org](https://dyngen.dynverse.org). [a]Several insights 
- * We simplified and reran all use-case experiments, which allowed us to add some interpretation of the results for each of the use cases.
- * We performed new experiments assessing the scalability of dyngen and the similarity in output in comparison to a reference (real) dataset. The results are included in the supplementary material.
- As we think that several of the remarks raised by the reviewer are also interesting for the target audience of the paper, we not only include our responses to these remarks here, but in addition have created several tutorials that can be used to perform these experiments with dyngen.
-These tutorials now also allow users to reproduce some of the additional analyses requested by the reviewer, and they have been added to a new website that we created for our method, with a lot more documentation website ([dyngen.dynverse.org](https://dyngen.dynverse.org)).
-* … todo, add more ...[b]
+\newcommand{\xpr}[1]{\text{xpr}_{#1}}
+\newcommand{\xhl}[1]{\text{xhl}_{#1}}
+\newcommand{\ysr}[1]{\text{ysr}_{#1}}
+\newcommand{\yhl}[1]{\text{yhl}_{#1}}
+\newcommand{\ydr}[1]{\text{ydr}_{#1}}
+\newcommand{\zpr}[1]{\text{zpr}_{#1}}
+\newcommand{\zhl}[1]{\text{zhl}_{#1}}
+\newcommand{\zdr}[1]{\text{zdr}_{#1}}
 
-## Comment 1
-> Bioinformatics has been looking into GRNs, their simulation and estimation for more than a decade (in bulk at the time), and I miss some important references and insights from then.
-Such as alternate definitions of what a GRN is - eg does logic play a role, should you use boolean networks (see Albert&Othma, JTB 2003) -, do old-school bulk simulators such as GeneNetWeaver (Bioinformatics 2011) be used, can one use transformation methods (eg Wittmann et al, BMC Sysbio 2009) and how is this different from what has been introduced here (what is novelty here).
+\newcommand{\str}[1]{\text{str}_{#1}}
+\newcommand{\hill}[1]{\text{hill}_{#1}}
+\newcommand{\ind}[1]{\text{ind}_{#1}}
+\newcommand{\dis}[1]{\text{dis}_{#1}}
+\newcommand{\buf}[1]{\text{bind}_{#1}}
+\newcommand{\ba}[1]{\text{bas}_{#1}}
 
-Certainly simulating expression datasets using GRNs has been around for a long time and dyngen is inspired heavily by such software (GeneNetWeaver in particular).
-Nevertheless, there are great differences between bulk and single-cell simulators:  the data characteristics of single-cell omics are vastly different from bulk omics, typically having much lower library sizes and a higher dropout rate, but also a high number of profiles.
-The low library sizes, in particular, are problematic as ODEs are ill-suited for performing low-molecule simulations.
-This necessitates the development of new single-cell simulators.
+\newcommand{\proptran}{f}
+\newcommand{\ai}[2]{$S_{#1} = S_{#2b}$}
+\newcommand{\zk}[1]{\frac{y_#1}{k_#1}^{c_#1}}
+\newcommand{\wi}[1]{\nu_#1}
 
-We have added paragraph §3 to the introduction to discuss the relevance of bulk simulators and why they are not suitable for simulating single-cell expression data:
+# Supplementary Figures
+\renewcommand{\figurename}{Supplemental Figure}
+\renewcommand{\tablename}{Supplemental Table}
 
-> Simulating realistic data requires that the underlying biology is recapitulated as best as possible, and in the case of transcriptomics data this typically involves modelling the underlying gene regulatory networks.
-Simulators of "bulk" microarray or RNA-sequencing profiles simulate biological processes (e.g.
-transcription, translation) by translating a database of known regulatory interactions into a set of ordinary differential equations (ODE) [@roy_systemgeneratingtranscription_2008; @hache_gengesystematicgeneration_2009; @schaffter_genenetweaversilicobenchmark_2011; @vandenbulcke_syntrengeneratorsynthetic_2006].
-These methods have been instrumental in performing benchmarking studies [@prill_rigorousassessmentsystems_2010; @marbach_revealingstrengthsweaknesses_2010; @marbach_wisdomcrowdsrobust_2012].
-However, the advent of single-cell omics introduced several new types of analyses (e.g.
-trajectory inference, RNA velocity, cell-specific network inference) which exploit the higher resolution of single-cell versus bulk omics [@luecken_currentbestpractices_2019].
-In addition, the data characteristics of single-cell omics are vastly different from bulk omics, typically having much lower library sizes and a higher dropout rate, but also a high number of profiles [@vallejos_normalizingsinglecellrna_2017].
-The low library sizes, in particular, are problematic as ODEs are ill-suited for performing low-molecule simulations [@gillespie_exactstochasticsimulation_1977].
-This necessitates the development of new single-cell simulators.
+\newcommand{\yes}{\ding{51}}
+\newcommand{\no}{}
+\definecolor{light-gray}{rgb}{.85,.85,.85}
+\newcommand{\grayline}{\arrayrulecolor{light-gray}\cline{3-7}\arrayrulecolor{black}}
+\newcommand{\blackline}{\arrayrulecolor{black}\cline{3-7}}
+\newcommand{\method}[1]{#1}
 
-## Comment 2
-> Similarly, the single cell genomics field has used simulations also for dynamics for some time as evaluation at least implicitely.
-Please discuss advances and differences or similarities to those, eg.
-to simulations from (Ocone, Bioinf 2015) or (Pratapa, Nat Meth 2020).
+\begin{table}[H]
+    \caption{
+      \textbf{Feature comparison of single-cell synthetic data generators and simulators.} $ ^1$ As showcased by vignette.
+$^2$ As showcased by Van den Berge et al.
+\cite{vandenberge_trajectorybaseddifferentialexpression_2020}.
+    } \label{tab:comparison}
+    \small
+    \begin{tabular}{p{.25cm}l|*{5}{>{\centering\arraybackslash}p{1.25cm}|}}
+        \blackline
+        & \method{} & \method{splatter} & \method{powsimR} & \method{PROSSTT} & \method{SymSim} & \method{dyngen} \\
+        \blackline
+        \multicolumn{7}{l}{\textbf{Available modality outputs}} \\
+        \blackline
+        - & mRNA expression & \yes & \yes & \yes & \yes & \yes \\ \grayline
+        - & Pre-mRNA expression & \no & \no & \no & \no & \yes \\ \grayline
+        - & Protein expression & \no & \no & \no & \no & \yes \\ \grayline
+        - & Promotor activity & \no & \no & \no & \yes & \yes \\ \grayline
+        - & Reaction activity & \no & \no & \no & \no & \yes \\
+        \blackline
+        \multicolumn{7}{l}{\textbf{Available ground-truth outputs}} \\
+        \blackline
+        - & True counts & \yes & \no & \no & \yes & \yes \\ \grayline
+        - & Cluster labels & \yes & \yes & \no & \yes & \yes \\ \grayline
+        - & Trajectory & \yes & \no & \yes $ ^1$ & \yes & \yes \\ \grayline
+        - & Batch labels & \yes & \no & \no & \no & \yes $ ^1$\\ \grayline
+        - & Differential expression & \no & \no & \no & \yes & \yes $ ^2$ \\ \grayline
+        - & Knocked down regulators & \no & \no & \no & \no & \yes \\ \grayline
+        - & Regulatory network & \no & \no & \no & \yes & \yes \\ \grayline
+        - & Cell-specific regulatory network & \no & \no & \no & \no & \yes \\
+        \blackline
+        \multicolumn{7}{l}{\textbf{Emulate experimental effects}} \\
+        \blackline
+        - & Single-cell RNA sequencing & \yes & \yes & \yes & \yes & \yes \\ \grayline
+        - & Batch effects & \yes & \no & \no & \no & \yes $ ^1$ \\ \grayline
+        - & Knockdown experiment & \no & \no & \no & \no & \yes \\ \grayline
+        - & Time-series & \no & \no & \no & \no & \yes \\ \grayline
+        - & Snapshot & \no & \no & \no & \no & \yes \\
+        \blackline
+        \multicolumn{7}{l}{\textbf{Evaluation applications}} \\
+        \blackline
+        - & Clustering & \yes & \yes & \no & \yes & \no \\ \grayline
+        - & Trajectory inference & \yes & \no & \yes & \yes & \yes \\ \grayline
+        - & Network inference & \no & \no & \no & \yes & \yes \\ \grayline
+        - & Cell-specific network inference & \no & \no & \no & \no & \yes \\ \grayline
+        - & Differential expression & \yes & \no & \no & \no & \no \\ \grayline
+        - & Trajectory differential expression & \no & \no & \no & \no & \yes $ ^2$ \\ \grayline
+        - & Batch effect correction & \yes & \no & \no & \no & \yes \\ \grayline
+        - & RNA Velocity & \no & \no & \no & \no & \yes \\ \grayline
+        - & Trajectory alignment & \no & \no & \no & \no & \yes \\
+        \blackline
+        %\multicolumn{7}{l}{\textbf{Showcased dataset}} \\
+        %\blackline
+        %- & Number of cells &  &  &  &  & \\ \grayline
+        %- & Number of genes &  &  &  &  & \\ \grayline
+        %- & Execution time & & & & & \\
+        \blackline
+\end{tabular}
+\end{table}
 
-We expanded upon paragraph §4 and made a more direct comparison to the methods listed in Supplemental Table 1. Paragraph §4:
+\begin{figure}[H]
+    \centering
+    \includegraphics[width=\linewidth]{result_files/usecase_trajectory_alignment/supp_fig.pdf}
+    \caption{
+        \textbf{dyngen allows benchmarking of trajectory alignment methods.} 
+ \textbf{A}: An example linear dataset in need of trajectory alignment. Dashed lines represent the gold-standard alignment between the two trajectories according to the respective pseudotimes.
+  \textbf{B}: Result of the DTW alignment on the two trajectories. 
+  \textbf{C}: DTW calculates an accumulated distance matrix.
+In this matrix, a warping path (shown in black), following a valley in the matrix from the bottom left to the top right corner is found.
+This shows how the trajectories best match each other.
+  \textbf{D}: Illustration of the Area Between Worst and Prediction (ABWAP) metric.
+The warping path from subfigure C is mapped to the respective pseudotimes from both trajectories.
+The ABWAP score is equal to the area between the prediction and the worst possible prediction.
+  \textbf{E:} An evaluation of DTW versus cellAlign on 40 different linear trajectories, in which cellAlign significantly outperforms DTW.
+Box plot denotes the $Q_0$ to $Q_4$ quartile values.
+Significance of differences between methods was tested with a non-parametric paired two-sided Durbin-Conover test.
+Test statistics and p-values for all combinations are reported in the Source Data file.
+    }
+    \label{fig:traj_align}
+\end{figure}
 
-> To this end, single-cell omics simulators emulate the technical procedures from single-cell omics protocols.
-Simulators such as Splatter [@zappia_splattersimulationsinglecell_2017], powsimR [@vieth_powsimrpoweranalysis_2017], PROSSTT [@papadopoulos_prossttprobabilisticsimulation_2019] and SymSim [@zhang_simulatingmultiplefaceted_2019]) have already been widely used to compare single-cell methods [@street_slingshotcelllineage_2018; @parra_reconstructingcomplexlineage_2019; @lummertzdarocha_reconstructioncomplexsinglecell_2018; @lin_scclassifysamplesize_2020] and perform independent benchmarks [@duo_systematicperformanceevaluation_2018; @saelens_comparisonsinglecelltrajectory_2019; @soneson_biasrobustnessscalability_2018].
-However, by focusing more on simulating the single-cell omics protocol (e.g.
-RNA capture, amplification, sequencing) and less on the underlying biology (e.g.
-transcription, splicing, translation), their applicability and reusability is limited towards the specific application for which they were designed (e.g.
-benchmarking clustering or differential expression methods), and extending these tools to include additional modalities or experimental conditions is challenging.
+\begin{figure}[H]
+    \centering
+    \includegraphics[width=\linewidth]{result_files/usecase_rna_velocity/supp_fig.pdf}
+    \caption{
+        \textbf{dyngen allows benchmarking of RNA velocity methods.} 
+        \textbf{A:} The ground-truth information of a bifurcating dataset: ground-truth trajectory (left), gene expression of a gene B5\_TF1 (middle), and the RNA velocity of B5\_TF1 (right).
+\textbf{B:} The RNA velocity estimates of gene B5\_TF1 by the different methods.
+\textbf{C:} The velocity stream plots produced from the predictions of each method, as generated by scvelo.
+\textbf{D:} The predictions scored by two different metrics, the velocity correlation and the velocity arrow cosine.
+The velocity correlation is the correlation between the ground-truth velocity (A, right) and the predicted velocity (B).
+The velocity arrow cosine is the cosine similarity between the direction of segments of the ground-truth trajectory (A, left) and the RNA velocity values calculated at those points (C).
+Box plots denote the $Q_0$ to $Q_4$ quartile values.
+Significance of differences between methods was tested with pairwise non-parametric paired two-sided Durbin-Conover tests, Holm corrected for multiple testing.
+Test statistics and p-values for all combinations are reported in the Source Data file.
+    }
+    \label{fig:velocity}
+\end{figure}
 
-We did not include the reference to Ocone Bioinf 2015 as the advances proposed by this research were mainly related to performing Network Inference.
-In terms of generating synthetic data, the simulator used by Ocone et al.
-offered little advantage over aforementioned bulk simulators.
-We also did not include a reference to Pratapa et al., as their methodology is borrowed heavily from dyngen itself and offers little advantage over it.
+\begin{figure}[H]
+    \centering
+    \includegraphics[width=\linewidth]{result_files/usecase_network_inference/supp_fig.pdf}
+    \caption{
+        \textbf{dyngen allows benchmarking Cell-specific Network Inference (CSNI) methods.} 
+        \textbf{A:} A cell is simulated using the global gene regulatory network (GRN, top left).
+However, at any particular state in the simulation, only a fraction of the gene regulatory interactions are active.
+        \textbf{B:} CSNI methods were executed to predict the regulatory interactions that are active in each cell specifically.
+Using the ground-truth cell-specific GRN, the performance of each method was quantified on 42 dyngen datasets.
+Box plots denote the $Q_0$ to $Q_4$ quartile values.
+Significance of differences between methods was tested with pairwise non-parametric paired two-sided Durbin-Conover tests, Holm corrected for multiple testing.
+Test statistics and p-values for all combinations are reported in the Source Data file.
+    }
+    \label{fig:scgrn}
+\end{figure}
 
-## Comment 3
-> One of the main aims of any simulation framework should be to capture features of real biological data.
-I’d be great if the authors could demonstrate the fidelity of the generated data by comparing it to some real datasets, e.g.
-distribution of mean gene expression and number of zero counts per gene.
-How does it perform compared to other simulation frameworks?
-This would help me as well as community to evaluate how realistic the induced gene correlations/dependencies are, and what may be missing.
+\begin{figure}[H]
+    \centering
+    \includegraphics[width=\textwidth]{result_files/explain_methods}
+    \caption{\textbf{The workflow of dyngen consists of six main steps.} 
+    \textbf{A:} The user needs to specify the desired module network or use a predefined module network. The module network is what determines the dynamic behaviour of simulated cells.
+    \textbf{B:} The number of desired transcription factors (which drive the desired dynamic process) are amongst the given modules and adds regulatory interactions according to the module network.
+Additional target genes (which do not influence the dynamic process) are added by sampling interactions from GRN interaction databases.
+    \textbf{C:} Each gene regulatory interaction in the GRN is converted to a set of biochemical reactions. 
+    \textbf{D:} Along with the module network, the user also needs to specify the backbone structure of expected cell states.
+The average expression of each edge in the backbone is simulated by activating a restricted set of genes for each edge.
+    \textbf{E:} Multiple Gillespie SSA simulations are run using the reactions defined in step C.
+The counts of each of the molecules at each time step are extracted.
+Each time step is mapped to a point in the backbone.
+    \textbf{F:} The molecule levels of multiple simulations are shown over time (left).
+From each simulation, multiple cells are sampled (from left to middle).
+Technical noise from profiling is simulated by sampling molecules from the set of molecules inside each cell (from middle to right).
+    }
+    \label{fig:explain_methods}
+\end{figure}
 
-A vignette has been added to the package and supplementary material which uses countsimQC [@soneson_unifiedqualityverification_2018] to explore the characteristics of an dyngen simulation and the reference dataset used by it ([dyngen.dynverse.org/articles/advanced_topics/comparison_reference.html](https://dyngen.dynverse.org/articles/advanced_topics/comparison_reference.html)).
-The distribution plots in this vignette, while not perfect, nicely demonstrates that the output generated by dyngen mimics the given reference dataset for the evaluated metrics.
+\begin{figure}[H]
+    \centering
+    \includegraphics[width=0.8\textwidth]{result_files/example_backbones_onlymodules}
+    \caption{
+      \textbf{The module network determines the type of dynamic process which simulated cells will undergo.} A module network describes the regulatory interactions between sets of transcription factors which drive the desired dynamic process.
+    }
+    \label{fig:example_backbones_onlymodules}
+\end{figure}
 
-## Comment 4
-> Simulated data for the described key applications of dyngen (network inference, trajectory alignment, and RNA velocity) has already been generated within the respective works of lioness (Fig.
-S1+S4), cellAlign (Fig.
-S6), velocyto (Fig.
-S5), scvelo (Fig.
-S3), dynamo (Fig.
-1), many of which also use Gillespie.
-What does dyngen add conceptually?
+\begin{figure}[H]
+    \centering
+    \includegraphics[width=\linewidth]{result_files/gen_feature_network}
+    \caption{
+        \textbf{Generating the feature network from a backbone consists of four main steps.}
+    }
+    \label{fig:gen_feature_network}
+\end{figure}
 
-We modified paragraphs §5 and §6 to better highlight the scientific contributions dyngen makes:
+\begin{table}[H]
+    \caption{
+      \textbf{Reactions affecting the abundance levels of pre-mRNA $\x G$, mature mRNA $\y G$ and proteins $\z G$ of gene $G$.} Define the set of regulators of $G$ as $\rs{G}$, the set of upregulating regulators of $G$ as $\rp G$, and the set of down-regulating regulators of $G$ as $\rn G$.
+Parameters used in the propensity formulae are defined in Supplementary Table \ref{tab:reaction_params}.
+    } \label{tab:reaction_def}
+    \centering
+    \begin{tabular}{|lcc|}
+        \hline
+        Reaction & Effect & Propensity \\ \hline \hline
+        Transcription & $\emptyset \rightarrow \x G$ & $\xpr G \times \frac{\ba G - \ind{G}^{|\rp{G}|} + \prod\limits_{H \in \rp{G}}(\ind G + \buf{G,H})}{\prod\limits_{H \in \rs{G}}(1 + \buf{G,H})}$ \\
+        Splicing & $\x G \rightarrow \y G$ & $\ysr G \times \x G$ \\
+        Translation & $\y G \rightarrow \y G + \z G$ & $\zpr G \times \y G$ \\ \hline\hline
+        Pre-mRNA degradation & $\x G \rightarrow \emptyset$ & $\ydr G \times \x G$ \\
+        Mature mRNA degradation & $\y G \rightarrow \emptyset$ & $\ydr G \times \y G$ \\
+        Protein degradation & $\z G \rightarrow \emptyset$ & $\zdr G \times \z G$ \\ \hline
+    \end{tabular}
+\end{table}
 
-> We introduce dyngen, a method for simulating cellular dynamics at a single-cell, single-transcript resolution (Figure\ \ref{fig:overview}).
-This problem is tackled in three fully-configurable main steps.
-First, biological processes are mimicked by translating a gene regulatory network in a set of reactions (regulation, transcription, splicing, translation).
-Second, individual cells are simulated using Gillespie's stochastic simulation algorithm [@gillespie_exactstochasticsimulation_1977], which is designed to work well in low-molecule simulations.
-Finally, real reference datasets are used to emulate single-cell omics profiling protocols.
-> 
-> Throughout a simulation, dyngen tracks many layers of information, including the abundance of any molecule in the cell, the progression of the cell along a dynamic process, and the activation strength of individual regulatory interactions.
-In addition, dyngen can simulate a large variety of dynamic processes (e.g.
-cyclic, branching, disconnected) as well as a broad range of experimental conditions (e.g.
-batch effects and time-series, perturbation and single-cell knockdown experiments).
-For these reasons, dyngen can cater to a wide range of benchmarking applications, including trajectory inference, trajectory alignment, and trajectory differential expression (Table\ \ref{tab:comparison}).
+\begin{table}[H]
+    \caption{
+      \textbf{Default parameters defined for the calculation of reaction propensity functions.}
+    } \label{tab:reaction_params}
+    \centering
+    \begin{tabular}{|lrl|}
+        \hline
+        Parameter & Symbol & Definition \\ \hline \hline
+        Transcription rate & $\xpr{G}$ & $\in U(10, 20)$ \\
+        Splicing rate & $\ysr G$ & $= \ln(2)\ /\ 2$ \\
+        Translation rate & $\zpr{G}$ & $\in U(100, 150)$ \\
+        (Pre-)mRNA half-life & $\yhl{G}$ & $\in U(2.5, 5)$ \\
+        Protein half-life & $\zhl G$ & $\in U(5, 10)$ \\
+        Interaction strength & $\str{G,H}$ & $\in 10^{U(0, 2)}$ * \\
+        Hill coefficient & $\hill{G,H}$ & $\in U(0.5, 2)$ * \\
+        Independence factor & $\ind G$ & $\in U(0, 1)$ * \\ \hline\hline
+        (Pre-)mRNA degradation rate & $\ydr G$ & $= \ln(2)\ /\ \yhl G$ \\
+        Protein degradation rate & $\zdr G$ & $= \ln(2)\ /\ \zhl G$ \\
+        Dissociation constant & $\dis H$ & $= 0.5 \times \frac{\xpr H \times \ysr H \times \zpr H}{(\ydr H + \ysr H) \times \ydr H \times \zdr H}$ \\
+        Binding strength & $\buf{G,H}$ & $= \str{G,H} \times \left(\z H\ /\ \dis H\right) ^ {\hill{G,H}}$ \\
+        Basal expression & $\ba G$ & $= \begin{cases} 1 & \mbox{if } \rp{G} = \emptyset \\ 0.0001 & \mbox{if } \rn{G} = \emptyset \mbox{ and } \rp{G} \neq \emptyset \\ 0.5 & \mbox{otherwise} \end{cases}$ * \\ \hline
+        \multicolumn{3}{l}{*: unless $G$ is a TF, then the value is determined by the backbone.}
+    \end{tabular}
+\end{table}
 
-To the best of our knowledge, the works listed above use very different approaches to benchmarking their tool:
+\begin{figure}[H]
+    \centering
+    \includegraphics[width=.6\linewidth]{result_files/sample_cells.pdf}
+    \caption{
+        \textbf{Two approaches can be used to sample cells from simulations: snapshot and time-series.}
+    }
+    \label{fig:sample_cells}
+\end{figure}
 
-* LIONESS, network inference method: Uses randomly generated networks & boolean networks to generate synthetic data.
-The networks used to generate datasets are very synthetic and completely devoid of any biological relevance.
-By using Gillespie's SSA algorithm instead of boolean networks, we can use dyngen to track many layers of information (e.g.
-the number of times a biological reaction has been triggered during a given time interval).
-This allows us to benchmark methods which require this information (e.g.
-RNA velocity).
-* cellAlign, trajectory alignment: Perturbs real datasets to assess the robustness of cellAlign.
-When benchmarking a tool, assessing its robustness is almost as important as assessing its accuracy.
-Robustness analysis can be performed without knowing the gold standard, whereas the gold standard is needed for assessing the accuracy of a tool.
-* velocyto, RNA velocity: Uses ODEs to simulate splicing mechanics.
-The code for reproducing these datasets couldn't be found.
-Based on the description, these simulations were limited to simulating only the splicing mechanics under a particular set of conditions, thereby limiting its applicability outside of this particular context.
-* scvelo, RNA velocity: Same as velocyto.
-* dynamo, RNA velocity: Uses Gillespie's SSA to simulate splicing reactions.
-However, this system is very limited, including only two genes.
-This application of Gillespie's SSA seems very tailored towards this particular use-case and would require major redesigns to be made applicable elsewhere.
-
-It should be noted that from the list above, only LIONESS really compares its accuracy performance with other methods using a quantitative metric.
-We would thus argue that these cases further highlight the need for broadly applicable synthetic data generators such as dyngen.
-
-## Comment 5
-> The authors evaluate several computational tools, including the afore-mentioned.
-Even if the authors mention not to aim for a large-scale benchmarking, they still should provide some insights into why some tools perform better than others.
-Having an interpretable simulation engine in hand, it is unfortunate that it yet lacks any explanation and the reader is left alone with performance statements that are not well-founded.
-Some intuition would be very desirable.
-Why does SCENIC outperform the other methods?
-Why does smoothing improve trajectory alignment?
-And what causes differences in velocity estimation?
-Could you, for instance, generate an example where the differences in velocity estimates become more obvious, where one estimation procedure fails to fully capture the known direction?
-
-Performing in-depth benchmarks on each of these use-cases can be considered separate studies on their own and is beyond the scope of this work.
-However, we did reanalyse the experiments and added some interpretation to the respective sections.
-
- * Trajectory alignment: We removed DTW+smoothing (as it was a custom method developed for the sake of this analysis) and added cellAlign to the comparison, as cellAlign also uses smoothing and DTW to get to the same objective, and has been developed specifically with single cell data in mind.
- * RNA velocity: We removed some parameter settings which are designed to be very similar and thus have no added benefit in the visualisation.
-We removed the 'difficulty' settings for the different datasets, as the 'easy' and 'medium' datasets were unrealistic; only the 'hard' settings were retained.
-While there are no large difference in performance between the different methods, we added interpretation on why methods obtain low scores on one metric and high scores on the other.
- * Cell-specific network inference: We added information on why certain methods significantly outperform others and provide possible solutions to problems, where possible.
-
-## Minor comment 1
-> Could the authors please discuss runtime and scalability, which obviously can easily become a weakness for such a complex model.
-
-A vignette has been added to the package and supplementary material explores the execution time of dyngen for an increasing number of cells and genes ([dyngen.dynverse.org/articles/advanced_topics/scalability_and_runtime.html](https://dyngen.dynverse.org/articles/advanced_topics/scalability_and_runtime.html)).
-These figures show that the execution time of dyngen scales linearly w.r.t.
-the number of cells and also linearly w.r.t.
-the number of genes.
-
-## Minor comment 2
-> Please state how many genes have been used in the examples.
-It seems like the number of genes used in the examples is rather small (< 100).
-Wouldn’t it be better to use a more realistic number of features (~10,000)?
-
-Unfortunately, the examples shown in the package documentation need to be very small to abide by CRAN package submission policies, as the complete testing of the software package (including but not limited to running all examples) can strictly take 10 minutes on CRAN's server using only one thread.
-
-However, the scalability vignette mentioned above also shows the execution time of each of the steps taken by dyngen in simulating a 10'000 by 10'000 dataset, which takes about 20 minutes to complete.
-
-## Minor comment 3
-> Congratulations for making the clearly documented package available.
-Installation and going through the vignette works fine.
-At several points during the simulation files are downloaded without asking permission.
-It may be better practice to ask the user for permission, or download into a temporary directory, or read the files directly without downloading, or include in the package etc.
-
-By default, required files are downloaded to a temporary directory.
-Unfortunately, these files cannot be included inside the package as they are reference datasets which are much larger than what is allowed by CRAN policies.
-
-Based on the reviewers suggestion, we modified the installation instructions such that it better reflects that files will be downloaded when using dyngen, and provides instructions on how to configure the system such that it will cache these files in a user-specified directory ([dyngen.dynverse.org/articles/installation.html](https://dyngen.dynverse.org/articles/installation.html)).
-
-## Minor comment 4
-> Further, it might be very useful to output the simulated data to a standard object, such as Seurat or SingleCellExperiment, or to provide instructions on how to convert it.
-
-Thank you for this suggestion.
-We agree that conversion functions from the typical dyngen output to the typical object types familiar to single-cell omics programmers would be very useful.
-Therefore, we included such functions in the latest version of dyngen to convert its output to the following formats: anndata, dyno, SingleCellExperiment, Seurat.
-
-[a]containing what?
-[b]todo
+\begin{figure}[H]
+    \centering
+    \includegraphics[width=0.8\textwidth]{result_files/example_backbones}
+    \caption{\textbf{Examples of the ground-truth state networks which need to be provided alongside the module network.}}
+    \label{fig:example_backbones}
+\end{figure}
 

### 2021-05-12 manuscript/render_manuscript.R

@@ -5,20 +5,38 @@ library(textreadr)
 
 if (Sys.info()[["user"]] == "rcannood") {
   # httr::set_config(httr::config(http_version = 0)) # avoid http2 framing layer bug
-  drive <- drive_download(as_id("1gU6--vq988dQS3qYQRlIqC1QO96A4XM81s54K6YJKIk"), type = "text", overwrite = TRUE, path = tempfile())
+  drive <- drive_download(as_id("1aJTm23fUmUH4ad7Hl3eHoNFfK4Ltqj4LwLfP6pwmfi4"), type = "text", overwrite = TRUE, path = tempfile())
 
   # read docx
-  c(readr::read_lines(drive$local_path), readr::read_lines("manuscript/vignettes.Rmd")) %>%
+  readr::read_lines(drive$local_path) %>%
     str_replace_all("([^ ])(\\[@[^\\]]*\\])", "\\1 \\2") %>% # add spaces before citations
     str_replace_all("^#", "\n#") %>%
-    write_lines("manuscript/render.Rmd")
+    write_lines("manuscript/manuscript.Rmd")
 }
 
-# render the manuscript
-render("manuscript/render.Rmd")
+# take the relevant citatiosn and write it to a second file
+bib_keys <-
+  readr::read_lines(drive$local_path) %>%
+  paste(collapse = "\n") %>%
+  str_extract_all("\\[@[^\\]]*\\]") %>%
+  first() %>%
+  gsub("^\\[@", "", .) %>%
+  gsub("\\]$", "", .) %>%
+  str_split("; *") %>%
+  unlist() %>%
+  gsub("@", "", .) %>%
+  unique
 
-system("pdftk manuscript/render.pdf cat 1-16 output manuscript/manuscript.pdf")
-system("pdftk manuscript/render.pdf cat 17-45 output manuscript/supplementary_files.pdf")
+knitcitations::cite_options(check.entries = FALSE)
+bib <- knitcitations::read.bibtex("manuscript/library.bib")
+knitcitations::write.bibtex(bib[bib_keys], file = "manuscript/knitcitations.bib")
+
+# render the manuscript
+render("manuscript/manuscript.Rmd")
+
+
+
+render("manuscript/supplementary_files.Rmd")
 
 # system("latexdiff manuscript/v3/render.tex manuscript/render.tex > manuscript/track_changes.tex")
 # file.remove("manuscript/render.pdf")

### 2021-05-12 manuscript/render_manuscript.R
fix path

@@ -16,7 +16,7 @@ if (Sys.info()[["user"]] == "rcannood") {
 
 # take the relevant citatiosn and write it to a second file
 bib_keys <-
-  readr::read_lines(drive$local_path) %>%
+  readr::read_lines("manuscript/manuscript.Rmd") %>%
   paste(collapse = "\n") %>%
   str_extract_all("\\[@[^\\]]*\\]") %>%
   first() %>%

### 2021-05-12 manuscript/render_manuscript.R
fix typo

@@ -14,7 +14,7 @@ if (Sys.info()[["user"]] == "rcannood") {
     write_lines("manuscript/manuscript.Rmd")
 }
 
-# take the relevant citatiosn and write it to a second file
+# take the relevant citations and write it to a second file
 bib_keys <-
   readr::read_lines("manuscript/manuscript.Rmd") %>%
   paste(collapse = "\n") %>%

### 2021-05-12 manuscript/manuscript.Rmd
update

--- 2021-03-25 manuscript/render.Rmd
@@
 author:
-- Robrecht Cannoodt*
-- Wouter Saelens*
-- Louise Deconinck
-- Yvan Saeys
+- name: Robrecht Cannoodt
+  institute: [dambi, twist, di]
+  equal_contributor: true
+- name: Wouter Saelens
+  institute: [dambi, twist, epfl]
+  equal_contributor: true
+- name: Louise Deconinck
+  institute: [dambi, twist]
+- name: Yvan Saeys
+  institute: [dambi, twist]
+  correspondence: true
+  email: yvan.saeys@ugent.be
+institute: 
+  - dambi: Data Mining and Modelling for Biomedicine group, VIB Center for Inflammation Research, Ghent, Belgium
+  - twist: Department of Applied Mathematics, Computer Science, and Statistics, Ghent University, Ghent, Belgium
+  - di: Data Intuitive, Lebbeke, Belgium
+  - epfl: Institute of Bioengineering, School of Life Sciences, École Polytechnique Fédérale de Lausanne (EPFL), Lausanne, Switzerland
 date: "`r format(Sys.time(), '%d %B %Y')`"
@@
     citation_package: biblatex
-biblatexoptions: sorting=none,url=false
+    pandoc_args:
+      - '--lua-filter=lua/scholarly-metadata.lua'
+      - '--lua-filter=lua/author-info-blocks.lua'
+biblatexoptions: sorting=none,url=false,backend=bibtex
 classoption:
@@
 ```{r setup, include=FALSE}
+ # for lua files regarding author information, see https://stackoverflow.com/a/52919256
 knitr::opts_chunk$set(collapse=TRUE, results="hold")
@@
 
-# Main text
+# Introduction
 <!-- format: 1500 Words, no headings, max. 2 figures, max. 20 references-->
@@
 batch effects and time-series, perturbation and single-cell knockdown experiments).
-For these reasons, dyngen can cater to a wide range of benchmarking applications, including trajectory inference, trajectory alignment, and trajectory differential expression (Table\ \ref{tab:comparison}).
+For these reasons, dyngen can cater to a wide range of benchmarking applications, including trajectory inference, trajectory alignment, and trajectory differential expression (Supplementary Table\ 1).
 
@@
 
-<!-- RESULTS -->
+# Results
 We demonstrate dyngen's broad applicability by evaluating three novel types of computational approaches for which no simulation engines exist yet: cell-specific network inference, trajectory alignment and RNA velocity (Figure\ \ref{fig:applications}).
@@
     \caption{
-      \textbf{dyngen provides ground-truth data for a variety of applications (left), which can be used to quantitatively evaluate methods (right).}
+      \textbf{dyngen provides ground-truth data for a variety of applications (left), which can be used to quantitatively evaluate methods (right).} Box plots denote the $Q_0$ to $Q_4$ quartile values.
       \textbf{A:} Trajectory alignment aligns two trajectories between samples.
-We use dyngen to evaluate dynamic time warping (DTW) and cellAlign when aligning two linear trajectories with different kinetic parameters.
+We evaluate dynamic time warping (DTW) and cellAlign when aligning two linear trajectories with different kinetic parameters based on the area differences between the worst possible alignment and the predicted alignment (Area Between Worst And Prediction, or ABWAP).
       \textbf{B:} RNA velocity calculates for each cell the direction in which the expression of each gene is moving.
@@
 
-**Trajectory alignment** methods align trajectories from different samples and allow studying the differences between the different trajectories.
+Use-case "trajectory alignment".
+Trajectory alignment methods align trajectories from different samples and allow studying the differences between the different trajectories.
 For example, by comparing the transcriptomic profiles of cells from a diseased patient to a healthy control, it might be possible to detect transcriptomics differences (differential expression) of particular cells along a developmental process, or to detect an early stop of the trajectory of the diseased patient.
-Currently, trajectory alignment is limited to aligning linear trajectories, though other topologies of trajectory could be aligned as well.
+Currently, trajectory alignment is limited to aligning linear trajectories, though other topologies of a trajectory could be aligned as well.
 Dynamic Time Warping (DTW) [@giorgino_computingvisualizingdynamic_2009] is a method designed for aligning temporal sequences for speech recognition but has since been used to compare gene expression kinetics from many different biological processes [@cacchiarelli_aligningsinglecelldevelopmental_2018; @kanton_organoidsinglecellgenomic_2019; @mcfaline-figueroa_pooledsinglecellgenetic_2019; @alpert_alignmentsinglecelltrajectories_2018].
-We evaluate the performance of DTW by simulating two linear trajectories with slightly different simulation kinetics and assessing the accuracy of the alignment by DTW.
-We compare the performance of DTW against a version of DTW where gene expression is first smoothed, for varying degrees of noise added to the gene expression (Figure\ \ref{fig:traj_align}).
-We observe that DTW+smoother performs significantly better than the unsmoothed version across all levels of noise.
+cellAlign [@alpert_alignmentsinglecelltrajectories_2018] uses DTW to perform trajectory alignment, but also includes interpolation and scaling of the single cell data as a preprocessing step.
+We evaluate the performance of DTW and cellAlign by simulating 40 datasets, each containing two linear trajectories generated with the same gene regulatory network but with slightly different simulation kinetics.
+We assess the accuracy of the obtained alignments by comparing the generated alignment path with the worst possible alignment that could be performed (Supplementary Fig.
+\ 1D), named the Area Between Worst And Prediction (ABWAP) score.
+Overall, cellAlign performs significantly better than DTW (Supplementary Fig.\ 1), which is likely due to the interpolation and scaling steps provided by cellAlign, reducing noise in the data and improving the comparability of the trajectories.
 Note that, in this comparison, only linear trajectory alignment is performed.
 While dyngen can generate non-linear trajectories (e.g.
-cyclic or branching), constructing a quantitative accuracy metric for non-linear trajectory alignment is not trivial and an avenue for future work.[a]
+cyclic or branching), both aligning non-linear trajectories and constructing a quantitative accuracy metric for non-linear trajectory alignment is not trivial and an avenue for future work.
 
-**RNA velocity** methods use the relative ratio between pre-mRNA and mature mRNA reads to predict the rate of increase/decrease of RNA molecule abundance, as this can be used to predict the directionality of single cell differentiation in trajectories [@zeisel_coupledpremrnamrna_2011;@lamanno_rnavelocitysingle_2018].
+Use-case "RNA velocity".
+RNA velocity methods use the relative ratio between pre-mRNA and mature mRNA reads to predict the rate of increase/decrease of RNA molecule abundance, as this can be used to predict the directionality of single cell differentiation in trajectories [@zeisel_coupledpremrnamrna_2011;@lamanno_rnavelocitysingle_2018].
 Already two algorithms are currently available for estimating the RNA velocity vector from spliced and unspliced counts: velocyto [@lamanno_rnavelocitysingle_2018] and scvelo [@bergen_generalizingrnavelocity_2020].
@@
 We executed velocyto and scvelo (with 2 different parameter settings, stochastic and dynamical) on 42 datasets with a variety of backbones (including linear, bifurcating, cyclic, disconnected).
-We evaluated the predictions using two metrics (Figure\ \ref{fig:velocity}), one which directly compares the predicted RNA velocity of each gene with the ground-truth RNA velocity (called the "velocity correlation"), and one which compares the direction of the ground-truth trajectory embedded in a dimensionality reduction with the average RNA velocity of cells in that neighbourhood (called the "velocity arrow cosine").
+We evaluated the predictions using two metrics (Supplementary Fig.\ 2), one which directly compares the predicted RNA velocity of each gene with the ground-truth RNA velocity (called the "velocity correlation"), and one which compares the direction of the ground-truth trajectory embedded in a dimensionality reduction with the average RNA velocity of cells in that neighbourhood (called the "velocity arrow cosine").
 While both velocyto and scvelo obtained high scores for the velocity arrow cosine metric (overall 25th percentile = 0.606), the velocity correlation is rather low (overall 75th percentile = 0.156).
 This means that predicting the RNA velocity (i.e.
-transcription rate minus the decay rate) for any particular gene is challenging, but the combined information is very informative in determining the directionality of cell progression in the trajectory.
-In terms of velocity correlation, no method performed significantly better than the other, whereas scvelo stochastic performed slightly worse than scvelo dynamical and velocyto in terms of velocity arrow cosine score.
+transcription rate minus the decay rate) for particular individual genes can be challenging, but the combined information is very informative in determining the directionality of cell progression in the trajectory.
+In terms of velocity correlation, no method performed significantly better than the other, whereas "scvelo stochastic" performed slightly worse than "scvelo dynamical" and velocyto in terms of velocity arrow cosine score.
+Note that, given that some genes are more informative in determining the overall directionality of cell progression, performing a feature selection before computing the embedded dimensionality reduction might result in significantly improved velocity arrow cosine scores.
 
-**Cell-specific network inference** (CSNI) methods predict not only which transcription factors regulate which target genes, but also aim to identify how active each interaction is in each of the cells, since interactions can be turned off and on depending on the cellular state.
+Use-case "Cell-specific network inference" (CSNI).
+CSNI methods predict not only which transcription factors regulate which target genes, but also aim to identify how active each interaction is in each of the cells, since interactions can be turned off and on depending on the cellular state.
 While a few pioneering CSNI approaches have already been developed [@aibar_scenicsinglecellregulatory_2017; @kuijjer_estimatingsamplespecificregulatory_2019; @liu_personalizedcharacterizationdiseases_2016], a quantitative assessment of their performance is until now lacking.
@@
 Extracting the ground-truth dynamic network in dyngen is straightforward though, given that we can calculate how target gene expression would change without the regulator being present.
-We used this ground-truth to compare the performance of three CSNI methods (Figure\ \ref{fig:scgrn}): LIONESS [@kuijjer_estimatingsamplespecificregulatory_2019], SSN [@liu_personalizedcharacterizationdiseases_2016] and SCENIC [@aibar_scenicsinglecellregulatory_2017].
+We used this ground-truth to compare the performance of three CSNI methods (Supplementary Fig.\ 3): LIONESS [@kuijjer_estimatingsamplespecificregulatory_2019], SSN [@liu_personalizedcharacterizationdiseases_2016] and SCENIC [@aibar_scenicsinglecellregulatory_2017].
 For each dataset, we computed the mean AUROC and AUPR scores of the individual cells.
@@
 
-<!-- DISCUSSION -->
-In summary, dyngen's single-cell simulations can be used to evaluate common single-cell omics computational methods such as clustering, batch correction, trajectory inference, and network inference.
+# Discussion
+dyngen's single-cell simulations can be used to evaluate common single-cell omics computational methods such as clustering, batch correction, trajectory inference, and network inference.
 However, the framework is flexible enough to be adaptable to a broad range of applications, including methods that integrate clustering, network inference, and trajectory inference.
@@
 
-# Data Availability[b]
-
-# Code Availability
-dyngen is available as an open-source R package on CRAN at [cran.r-project.org/package=dyngen](https://cran.r-project.org/package=dyngen).
-The analyses performed in this manuscript are available on GitHub at [github.com/dynverse/dyngen_manuscript](https://github.com/dynverse/dyngen_manuscript).
-
-# Author contributions
-* W.S. and R.C. designed the study.
-* R.C., W.S., and L.D. performed the experiments and analysed the data.
-* R.C. and W.S. implemented the dyngen software package.
-* R.C., W.S., L.D., and Y.S. wrote the manuscript.
-* Y.S. supervised the project.
-
 \newpage
@@
 # Methods {#sec:dyngen-methods}
-The workflow to generate _in silico_ single-cell data consists of six main steps (Figure\ \ref{fig:explain_methods}).
+The workflow to generate _in silico_ single-cell data consists of six main steps (Supplementary Fig\ 4).
 
@@
 
-Several examples of module networks are given in Figure\ \ref{fig:example_backbones_onlymodules}.
+Several examples of module networks are given in Supplementary Fig.\ 5.
 A simple chain of modules (where one module upregulates the next) results in a _linear_ process.
@@
 ## Generating the gene regulatory network {#sec:dyngen-grn}
-The GRN is generated based on the given module network in four main steps (Figure\ \ref{fig:gen_feature_network}).
+The GRN is generated based on the given module network in four main steps (Supplementary Fig.\ 6).
 
-**Step 1, sampling the transcription factors (TF).** The TFs are the main drivers of the molecular changes in the simulation.
+Step 1, sampling the transcription factors (TF).
+The TFs are the main drivers of the molecular changes in the simulation.
 The user provides a backbone and the number of TFs to generate.
@@
 
-**Step 2, generating the TF interactions.** Let each TF be regulated according to the interactions in the backbone.
+Step 2, generating the TF interactions.
+Let each TF be regulated according to the interactions in the backbone.
 These interactions inherit the effect, strength, and independence parameters from the interactions in the backbone.
@@
 
-**Step 3, sampling the target subnetwork.**
+Step 3, sampling the target subnetwork.
 A user-defined number of target genes are added to the GRN.
 Target genes are regulated by a TF or another target gene, but are always downstream of at least one TF.
-To sample the interactions between target genes, one of the many FANTOM5 \cite{lizio_gatewaysfantom5promoter_2015} GRNs is sampled.
+To sample the interactions between target genes, one of the many FANTOM5 [@lizio_gatewaysfantom5promoter_2015] GRNs is sampled.
 The currently existing TFs are mapped to regulators in the FANTOM5 GRN.
@@
 
-**Step 4, sampling the housekeeping subnetwork.**
+Step 4, sampling the housekeeping subnetwork.
 Housekeeping genes are completely separate from any TFs or target genes.
@@
 
-The effects of each reaction mimic the respective biological processes (Table\ \ref{tab:reaction_def}, middle).
+The effects of each reaction mimic the respective biological processes (Supplementary Table\ 2, middle).
 Transcription of gene $G$ results in the creation of a single pre-mRNA molecule $\x G$.
@@
 
-The propensity of all reactions except transcription are all linear functions (Table\ \ref{tab:reaction_def}, right) of the abundance level of some molecule multiplied by a per-gene constant (Table\ \ref{tab:reaction_params}).
+The propensity of all reactions except transcription are all linear functions (Supplementary Table\ 2, right) of the abundance level of some molecule multiplied by a per-gene constant (Supplementary Table\ 3).
 The propensity of transcription of a gene $G$ depends on the abundance levels of its TFs.
@@
 \begin{align}
-  f(y_1, y_2, \ldots, y_N) & = \text{xpr} \cdot \sum_{j = 0}^{2^N - 1} \alpha_j \cdot P(S_j) \label{eqn:activ0} \\
+  f(y_1, y_2, \ldots, y_N) & = \text{xpr} \cdot \sum_{j = 0}^{2^N - 1} \alpha_j \cdot P(S_j) \label{eqn:activ0} 
 \end{align}
@@
 Each reaction consists of its propensity -- a formula to calculate the probability of the reaction occurring during an infinitesimal time interval -- and the effect -- how it will affect the current state if triggered.
-Each time a reaction is triggered, the simulation time is incremented by $\tau = \frac{1}{\sum_j prop_j} \ln\left(\frac{1}{r}\right)$, with $r \in U(0, 1)$ and $prop_j$ the propensity value of the $j$th reaction for the current state of the simulation.
+Each time a reaction is triggered, the simulation time is incremented by $\tau = \frac{1}{\sum_j \textrm{prop}_j} \ln\left(\frac{1}{r}\right)$, with $r \in U(0, 1)$ and $\textrm{prop}_j$ the propensity value of the $j$th reaction for the current state of the simulation.
 
@@
 By setting the census interval to 0, the whole simulation's trajectory is retained but many of these time points will contain very similar information.
-In addition to the abundance levels, also the propensity values and the number of firings of each of the reactions at each of the time steps can be retained, as well as specific sub-calculations of the propensity values, such as the regulator activity level $reg_{G,H}$.
+In addition to the abundance levels, also the propensity values and the number of firings of each of the reactions at each of the time steps can be retained.
 
@@
 
-**Snapshot** The backbone consists of several states linked together by transition edges with length $L_i$, to which the different states in the different simulations have been mapped (Figure\ \ref{fig:sample_cells}A).
+**Snapshot** The backbone consists of several states linked together by transition edges with length $L_i$, to which the different states in the different simulations have been mapped (Supplementary Fig.\ 7A).
 From each transition, $N_i = N / \frac{L_i}{\sum L_i}$ cells are sampled uniformly, rounded such that $\sum N_i = N$.
@@
 **Time series** Assuming that the final time of the simulation is $T$, the interval $[0, T]$ is divided into $k$ equal intervals of width $w$ separated by $k-1$ gaps of width $g$.
-$N_i = N / k$ cells are sampled uniformly from each interval (Figure\ \ref{fig:sample_cells}B), rounded such that $\sum N_i = N$.
+$N_i = N / k$ cells are sampled uniformly from each interval (Supplementary Fig.\ 7B), rounded such that $\sum N_i = N$.
 By default, $k = 8$ and $g = 0.75$.
@@
 ## Determining the ground-truth trajectory {#sec:dyngen-groundtruth}
-To construct the ground-truth trajectory, the user needs to provide the ground-truth state network alongside the initial module network (Figure\ \ref{fig:example_backbones}).
+To construct the ground-truth trajectory, the user needs to provide the ground-truth state network alongside the initial module network (Supplementary Fig.\ 8).
 Each edge in the state network specifies which modules are allowed to change in expression in transitioning from one state to another.
@@
 
-As an example, consider the cyclic trajectory shown in Figure\ \ref{fig:example_backbones}.
+As an example, consider the cyclic trajectory shown in Supplementary Fig.\ 8.
 State S0 begins with an expression vector of all zero values.
@@
 ## Determining the cell-specific ground-truth regulatory network {#sec:dyngen-extractgrn}
-Calculating the regulatory effect of a regulator $R$ on a target $T$ (Figure\ \ref{fig:explain_methods}F) requires determining the contribution of $R$ in the propensity function of the transcription of $T$ (section\ \ref{sec:dyngen-reactions}) with respect to other regulators.
+Calculating the regulatory effect of a regulator $R$ on a target $T$ (Supplementary Fig.\ 4F) requires determining the contribution of $R$ in the propensity function of the transcription of $T$ (section\ \ref{sec:dyngen-reactions}) with respect to other regulators.
 This information is useful, amongst others, for benchmarking cell-specific network inference methods.
@@
 
-We compared the mean AUROC and AUPR scores obtained by the three CSNI methods across all datasets by performing pairwise non-parametric paired Durbin-Conover tests [@conover_multiplecomparisonsprocedures_1979] using `pairwiseComparisons` [@patil_pairwisecomparisonsmultiplepairwise_2019].
+We compared the mean AUROC and AUPR scores obtained by the three CSNI methods across all datasets by performing pairwise non-parametric paired two-sided Durbin-Conover tests [@conover_multiplecomparisonsprocedures_1979] using `pairwiseComparisons` [@patil_pairwisecomparisonsmultiplepairwise_2019].
+Test statistics and p-values for the all pairwise combinations are reported in the Source Data file.
 Reported p-values are adjusted for multiple testing using Holm correction [@holm_simplesequentiallyrejective_1979].
@@
 
-We compared the velocity correlation and velocity arrow cosine scores obtained by velocyto and scvelo across all datasets by performing pairwise non-parametric paired Durbin-Conover tests [@conover_multiplecomparisonsprocedures_1979] using `pairwiseComparisons` [@patil_pairwisecomparisonsmultiplepairwise_2019].
+We compared the velocity correlation and velocity arrow cosine scores obtained by velocyto and scvelo across all datasets by performing pairwise non-parametric paired two-sided Durbin-Conover tests [@conover_multiplecomparisonsprocedures_1979] using `pairwiseComparisons` [@patil_pairwisecomparisonsmultiplepairwise_2019].
+Test statistics and p-values for the all pairwise combinations are reported in the Source Data file.
 Reported p-values are adjusted for multiple testing using Holm correction [@holm_simplesequentiallyrejective_1979].
 
-## Comparison of trajectory alignment {#sec:dyngen-tacompare}[c]
+## Comparison of trajectory alignment {#sec:dyngen-tacompare}
 
@@
 DTW is designed to align temporal sequences by dilating or contracting the sequences to best match each other.
-cellAlign …[d] .
+cellAlign uses DTW to perform this alignment, but first interpolates and rescales the input data in order to better cope with single-cell omics data.
 
-To evaluate a trajectory alignment method on a paired dataset we computed the geodesic distances of each cell from the start of the trajectory, also called the _pseudotime_.
+To evaluate a trajectory alignment method on a combined dataset we computed the geodesic distances of each cell from the start of the trajectory, also called the _pseudotime_.
 For each dataset, the pseudotime values are rescaled between 0 and 1 to allow for easier comparison.
 A trajectory alignment produces a sequence of index pairs $[(i_0, j_0), (i_1, j_1), \ldots, (i_N, j_N)]$, where $i_0$ and $j_0$ are equal to 0 (the first position in both pseudotime series), $i_N$ and $j_N$ are equal to the respective last positions in the pair of pseudotime series, and $[i_0, i_1, \ldots, i_N]$ and $[j_0, j_1, \ldots, j_N]$ are in ascending order and can contain duplicates values.
-The 'score'[e] metric is defined as follows, where a score of 0 represents the worst possible ordering and 1 represents the best possible ordering.
+The ABWAP metric is defined as follows, where $pt_1$ and $pt_2$ are the unit pseudotime vectors.
+See Supplementary Fig.\ 1D for a visual interpretation of this metric.
 
 \begin{equation}
-  \textrm{score} = 1 - \textrm{area\_under\_curve}(pt_1[i_0 .. i_N] + pt_2[j_0 .. j_N], abs(pt_1[i_0 .. i_N] - pt_2[j_0 .. j_N]))
+  \textrm{ABWAP} = 1 - \textrm{area\_under\_curve}(pt_1[i_0 .. i_N] + pt_2[j_0 .. j_N], abs(pt_1[i_0 .. i_N] - pt_2[j_0 .. j_N]))
 \end{equation}
 
-\printbibliography
-\newpage
+We compared the ABWAP scores obtained by DTW and cellAlign across all datasets by performing pairwise non-parametric paired two-sided Durbin-Conover tests [@conover_multiplecomparisonsprocedures_1979] using `pairwiseComparisons` [@patil_pairwisecomparisonsmultiplepairwise_2019].
+Test statistics and p-values for the all pairwise combinations are reported in the Source Data file.
+Reported p-values are adjusted for multiple testing using Holm correction [@holm_simplesequentiallyrejective_1979].
 
-# Supplementary Figures
-\setcounter{page}{1}
-\newrefsection
-\setcounter{table}{0}
-\renewcommand{\thetable}{S\arabic{table}}
-\setcounter{figure}{0}
-\renewcommand{\thefigure}{S\arabic{figure}}
+# Data Availability
+Source data for box plots in Figure \ref{fig:applications} and Supplementary Figures\ 1, 2\ and\ 3 are provided with this paper.
+All code and data required to reproduce the analysis are available on GitHub at [github.com/dynverse/dyngen_manuscript](https://github.com/dynverse/dyngen_manuscript).
+The datasets generated for the different use cases are available on Zenodo with record number 4637926 (doi: [10.5281/zenodo.4637926](https://doi.org/10.5281/zenodo.4637926)).
 
-\newcommand{\yes}{\ding{51}}
-\newcommand{\no}{}
-\definecolor{light-gray}{rgb}{.85,.85,.85}
-\newcommand{\grayline}{\arrayrulecolor{light-gray}\cline{3-7}\arrayrulecolor{black}}
-\newcommand{\blackline}{\arrayrulecolor{black}\cline{3-7}}
-\newcommand{\method}[1]{#1}
+# Code Availability
+Results in this manuscript were generated with R 4.0.3 and dyngen 1.0.0.
+dyngen is available as an open-source software package at [cran.r-project.org/package=dyngen](https://cran.r-project.org/package=dyngen) and also on Zenodo with record number 4751443 (doi: [10.5281/zenodo.4751443](https://doi.org/10.5281/zenodo.4751443)).
+The analyses performed in this manuscript are available on GitHub at [github.com/dynverse/dyngen_manuscript](https://github.com/dynverse/dyngen_manuscript).
 
-\begin{table}[H]
-    \caption{
-      \textbf{Feature comparison of single-cell synthetic data generators and simulators.} $ ^1$ As showcased by vignette.
-$^2$ As showcased by Van den Berge et al.
-\cite{vandenberge_trajectorybaseddifferentialexpression_2020}.
-    } \label{tab:comparison}
-    \small
-    \begin{tabular}{p{.25cm}l|*{5}{>{\centering\arraybackslash}p{1.25cm}|}}
-            \blackline
-            & \method{} & \method{splatter} & \method{powsimR} & \method{PROSSTT} & \method{SymSim} & \method{dyngen} \\
-            \blackline
-            \multicolumn{7}{l}{\textbf{Available modality outputs}} \\
-            \blackline
-            - & mRNA expression & \yes & \yes & \yes & \yes & \yes \\ \grayline
-            - & Pre-mRNA expression & \no & \no & \no & \no & \yes \\ \grayline
-            - & Protein expression & \no & \no & \no & \no & \yes \\ \grayline
-            - & Promotor activity & \no & \no & \no & \yes & \yes \\ \grayline
-            - & Reaction activity & \no & \no & \no & \no & \yes \\
-            \blackline
-            \multicolumn{7}{l}{\textbf{Available ground-truth outputs}} \\
-            \blackline
-            - & True counts & \yes & \no & \no & \yes & \yes \\ \grayline
-            - & Cluster labels & \yes & \yes & \no & \yes & \yes \\ \grayline
-            - & Trajectory & \yes & \no & \yes $ ^1$ & \yes & \yes \\ \grayline
-            - & Batch labels & \yes & \no & \no & \no & \yes $ ^1$\\ \grayline
-            - & Differential expression & \no & \no & \no & \yes & \yes $ ^2$ \\ \grayline
-            - & Knocked down regulators & \no & \no & \no & \no & \yes \\ \grayline
-            - & Regulatory network & \no & \no & \no & \yes & \yes \\ \grayline
-            - & Cell-specific regulatory network & \no & \no & \no & \no & \yes \\
-            \blackline
-            \multicolumn{7}{l}{\textbf{Emulate experimental effects}} \\
-            \blackline
-            - & Single-cell RNA sequencing & \yes & \yes & \yes & \yes & \yes \\ \grayline
-            - & Batch effects & \yes & \no & \no & \no & \yes $ ^1$ \\ \grayline
-            - & Knockdown experiment & \no & \no & \no & \no & \yes \\ \grayline
-            - & Time-series & \no & \no & \no & \no & \yes \\ \grayline
-            - & Snapshot & \no & \no & \no & \no & \yes \\
-            \blackline
-            \multicolumn{7}{l}{\textbf{Evaluation applications}} \\
-            \blackline
-            - & Clustering & \yes & \yes & \no & \yes & \no \\ \grayline
-            - & Trajectory inference & \yes & \no & \yes & \yes & \yes \\ \grayline
-            - & Network inference & \no & \no & \no & \yes & \yes \\ \grayline
-            - & Cell-specific network inference & \no & \no & \no & \no & \yes \\ \grayline
-            - & Differential expression & \yes & \no & \no & \no & \no \\ \grayline
-            - & Trajectory differential expression & \no & \no & \no & \no & \yes $ ^2$ \\ \grayline
-            - & Batch effect correction & \yes & \no & \no & \no & \yes \\ \grayline
-            - & RNA Velocity & \no & \no & \no & \no & \yes \\ \grayline
-            - & Trajectory alignment & \no & \no & \no & \no & \yes \\
-            \blackline
-            %\multicolumn{7}{l}{\textbf{Showcased dataset}} \\
-            %\blackline
-            %- & Number of cells &  &  &  &  & \\ \grayline
-            %- & Number of genes &  &  &  &  & \\ \grayline
-            %- & Execution time & & & & & \\
-            \blackline
-\end{tabular}
-\end{table}
+The version numbers of downstream dependencies of dyngen and dyngen_manuscript used in this study are: anndata 0.7.5.1, assertthat 0.2.1, babelwhale 1.0.1, bit 4.0.4, bit64 4.0.5, carrier 0.1.0, cellAlign 0.1.0, codetools 0.2-18, colorspace 2.0-0, compiler 4.0.4, crayon 1.4.1.9000, data.table 1.13.4, DBI 1.1.1, debugme 1.1.0, desc 1.2.0, digest 0.6.27, dplyr 1.0.5, dtw 1.22-3, dynparam 1.0.1, dynutils 1.0.6, dynwrap 1.2.2, ellipsis 0.3.1, fansi 0.4.2, farver 2.1.0, future 1.20.1, future.apply 1.7.0, generics 0.1.0, ggforce 0.3.2, ggplot2 3.3.3, ggraph 2.0.4, ggrepel 0.9.0, GillespieSSA2 0.2.7, globals 0.14.0, glue 1.4.2, graphlayouts 0.7.1, grid 4.0.4, gridExtra 2.3, gtable 0.3.0, gtools 3.8.2, hdf5r 1.3.3, hms 1.0.0, igraph 1.2.6, irlba 2.3.3, jsonlite 1.7.2, lattice 0.20-41, lifecycle 1.0.0, lisi 1.0, listenv 0.8.0, lmds 0.1.0, magrittr 2.0.1, MASS 7.3-53, Matrix 1.3-2, matrixStats 0.57.0, munsell 0.5.0, parallel 4.0.4, parallelly 1.21.0, patchwork 1.1.1, pbapply 1.4-3, pheatmap 1.0.12, pillar 1.5.1, pkgconfig 2.0.3, plyr 1.8.6, polyclip 1.10-0, pracma 2.2.9, processx 3.4.5, proxy 0.4-24, proxyC 0.1.5, ps 1.6.0, purrr 0.3.4, R6 2.5.0, RANN 2.6.1, rappdirs 0.3.3, RColorBrewer 1.1-2, Rcpp 1.0.6, RcppParallel 5.0.2, RcppXPtrUtils 0.1.1, readr 1.4.0, remotes 2.2.0, reshape2 1.4.4, reticulate 1.18-9007, rlang 0.4.10, rprojroot 2.0.2, scales 1.1.1, sctransform 0.3.2, scvelo 0.1.0.9000, stringi 1.5.3, stringr 1.4.0, tibble 3.0.5, tidygraph 1.2.0, tidyr 1.1.2, tidyselect 1.1.0, tools 4.0.4, tweenr 1.0.1, utf8 1.1.4, vctrs 0.3.6, viridis 0.5.1, viridisLite 0.3.0, yaml 2.2.1.
 
-\begin{figure}[H]
-    \centering
-    \includegraphics[width=\linewidth]{result_files/usecase_trajectory_alignment/supp_fig.pdf}
-    \caption{
-            \textbf{dyngen allows benchmarking of trajectory alignment methods.} 
- \textbf{A, B}: Two samples containing a linear trajectory, generated by dyngen.
-  \textbf{C, D}: Result of the DTW alignment on the samples in \textbf{A} and \textbf{B}.
-In \textbf{C}, the individual mappings of the alignment between cells are shown.
-In \textbf{D}, the accumulated distance matrix between the two trajectories is shown, including the black warping path, corresponding to these cell to cell alignments in \textbf{C}.
-  \textbf{E, F}: Shows the accumulated distance matrices obtained after using DTW on two trajectories where noise (noise level of 0.4) was added to the count matrix.
-In E the complete count matrices were used to perform the alignment.
-In F, smoothed pseudocells were used.
-  \textbf{G:} Shows the influence of added noise to the different processing methods. We can see that DTW + smoothing performs best in noisy circumstances.
-    }[f]
-    \label{fig:traj_align}
-\end{figure}
-
-\begin{figure}[H]
-    \centering
-    \includegraphics[width=\linewidth]{result_files/usecase_rna_velocity/supp_fig.pdf}
-    \caption{
-            \textbf{dyngen allows benchmarking of RNA velocity methods.} 
-            \textbf{A:} The ground-truth information of a bifurcating dataset: ground-truth trajectory (left), gene expression of a gene B5\_TF1 (middle), and the RNA velocity of B5\_TF1 (right).
-\textbf{B:} The RNA velocity estimates of gene B5\_TF1 by the different methods.
-\textbf{C:} The velocity stream plots produced from the predictions of each method, as generated by scvelo.
-\textbf{D:} The predictions scored by two different metrics, the velocity correlation and the velocity arrow cosine.
-The velocity correlation is the correlation between the ground-truth velocity (A, right) and the predicted velocity (B).
-The velocity arrow cosine is the cosine similarity between the direction of segments of the ground-truth trajectory (A, left) and the RNA velocity values calculated at those points (C).
-    }
-    \label{fig:velocity}
-\end{figure}
-
-\begin{figure}[H]
-    \centering
-    \includegraphics[width=\linewidth]{result_files/usecase_network_inference/supp_fig.pdf}
-    \caption{
-            \textbf{dyngen allows benchmarking Cell-specific Network Inference (CSNI) methods.} 
-            \textbf{A:} A cell is simulated using the global gene regulatory network (GRN, top left).
-However, at any particular state in the simulation, only a fraction of the gene regulatory interactions are active.
-            \textbf{B:} CSNI methods were executed to predict the regulatory interactions that are active in each cell specifically.
-Using the ground-truth cell-specific GRN, the performance of each method was quantified on 42 dyngen datasets.
-    }
-    \label{fig:scgrn}
-\end{figure}
-
-\begin{figure}[H]
-    \centering
-    \includegraphics[width=\textwidth]{result_files/explain_methods}
-    \caption{\textbf{The workflow of dyngen consists of six main steps.} 
-    \textbf{A:} The user needs to specify the desired module network or use a predefined module network. The module network is what determines the dynamic behaviour of simulated cells.
-    \textbf{B:} The number of desired transcription factors (which drive the desired dynamic process) are amongst the given modules and adds regulatory interactions according to the module network.
-Additional target genes (which do not influence the dynamic process) are added by sampling interactions from GRN interaction databases.
-    \textbf{C:} Each gene regulatory interaction in the GRN is converted to a set of biochemical reactions. 
-    \textbf{D:} Along with the module network, the user also needs to specify the backbone structure of expected cell states.
-The average expression of each edge in the backbone is simulated by activating a restricted set of genes for each edge.
-    \textbf{E:} Multiple Gillespie SSA simulations are run using the reactions defined in step C.
-The counts of each of the molecules at each time step are extracted.
-Each time step is mapped to a point in the backbone.
-    \textbf{F:} The molecule levels of multiple simulations are shown over time (left).
-From each simulation, multiple cells are sampled (from left to middle).
-Technical noise from profiling is simulated by sampling molecules from the set of molecules inside each cell (from middle to right).
-    }
-    \label{fig:explain_methods}
-\end{figure}
-
-\begin{figure}[H]
-    \centering
-    \includegraphics[width=0.8\textwidth]{result_files/example_backbones_onlymodules}
-    \caption{
-      \textbf{The module network determines the type of dynamic process which simulated cells will undergo.} A module network describes the regulatory interactions between sets of transcription factors which drive the desired dynamic process.
-    }
-    \label{fig:example_backbones_onlymodules}
-\end{figure}
-
-\begin{figure}[H]
-    \centering
-    \includegraphics[width=\linewidth]{result_files/gen_feature_network}
-    \caption{
-        \textbf{Generating the feature network from a backbone consists of four main steps.}
-    }
-    \label{fig:gen_feature_network}
-\end{figure}
-
-\begin{table}[H]
-    \caption{
-      \textbf{Reactions affecting the abundance levels of pre-mRNA $\x G$, mature mRNA $\y G$ and proteins $\z G$ of gene $G$.} Define the set of regulators of $G$ as $\rs{G}$, the set of upregulating regulators of $G$ as $\rp G$, and the set of down-regulating regulators of $G$ as $\rn G$.
-Parameters used in the propensity formulae are defined in Table \ref{tab:reaction_params}.
-    } \label{tab:reaction_def}
-    \centering
-    \begin{tabular}{|lcc|}
-        \hline
-        Reaction & Effect & Propensity \\ \hline \hline
-        Transcription & $\emptyset \rightarrow \x G$ & $\xpr G \times \frac{\ba G - \ind{G}^{|\rp{G}|} + \prod\limits_{H \in \rp{G}}(\ind G + \buf{G,H})}{\prod\limits_{H \in \rs{G}}(1 + \buf{G,H})}$ \\
-        Splicing & $\x G \rightarrow \y G$ & $\ysr G \times \x G$ \\
-        Translation & $\y G \rightarrow \y G + \z G$ & $\zpr G \times \y G$ \\ \hline\hline
-        Pre-mRNA degradation & $\x G \rightarrow \emptyset$ & $\ydr G \times \x G$ \\
-        Mature mRNA degradation & $\y G \rightarrow \emptyset$ & $\ydr G \times \y G$ \\
-        Protein degradation & $\z G \rightarrow \emptyset$ & $\zdr G \times \z G$ \\ \hline
-    \end{tabular}
-\end{table}
-
-\begin{table}[H]
-    \caption{
-      \textbf{Default parameters defined for the calculation of reaction propensity functions.}
-    } \label{tab:reaction_params}
-    \centering
-    \begin{tabular}{|lrl|}
-        \hline
-        Parameter & Symbol & Definition \\ \hline \hline
-        Transcription rate & $\xpr{G}$ & $\in U(10, 20)$ \\
-        Splicing rate & $\ysr G$ & $= \ln(2)\ /\ 2$ \\
-        Translation rate & $\zpr{G}$ & $\in U(100, 150)$ \\
-        (Pre-)mRNA half-life & $\yhl{G}$ & $\in U(2.5, 5)$ \\
-        Protein half-life & $\zhl G$ & $\in U(5, 10)$ \\
-        Interaction strength & $\str{G,H}$ & $\in 10^{U(0, 2)}$ * \\
-        Hill coefficient & $\hill{G,H}$ & $\in U(0.5, 2)$ * \\
-        Independence factor & $\ind G$ & $\in U(0, 1)$ * \\ \hline\hline
-        (Pre-)mRNA degradation rate & $\ydr G$ & $= \ln(2)\ /\ \yhl G$ \\
-        Protein degradation rate & $\zdr G$ & $= \ln(2)\ /\ \zhl G$ \\
-        Dissociation constant & $\dis H$ & $= 0.5 \times \frac{\xpr H \times \ysr H \times \zpr H}{(\ydr H + \ysr H) \times \ydr H \times \zdr H}$ \\
-        Binding strength & $\buf{G,H}$ & $= \str{G,H} \times \left(\z H\ /\ \dis H\right) ^ {\hill{G,H}}$ \\
-        Basal expression & $\ba G$ & $= \begin{cases} 1 & \mbox{if } \rp{G} = \emptyset \\ 0.0001 & \mbox{if } \rn{G} = \emptyset \mbox{ and } \rp{G} \neq \emptyset \\ 0.5 & \mbox{otherwise} \end{cases}$ * \\ \hline
-        \multicolumn{3}{l}{*: unless $G$ is a TF, then the value is determined by the backbone.}
-    \end{tabular}
-\end{table}
-
-\begin{figure}[H]
-    \centering
-    \includegraphics[width=.6\linewidth]{result_files/sample_cells.pdf}
-    \caption{
-        \textbf{Two approaches can be used to sample cells from simulations: snapshot and time-series.}
-    }
-    \label{fig:sample_cells}
-\end{figure}
-
-\begin{figure}[H]
-    \centering
-    \includegraphics[width=0.8\textwidth]{result_files/example_backbones}
-    \caption{\textbf{Examples of the ground-truth state networks which need to be provided alongside the module network.}}
-    \label{fig:example_backbones}
-\end{figure}
-
-[a]todo: update
-[b]// todo: upload temporary data to zenodo?
-[c]todo: update this section
-[d]@louise.deconinck@gmail.com could you add a description for cellAlign?
-_Assigned to Louise_
-[e]TODO: rename
-[f]TODO: update text
+# Author contributions
+* W.S. and R.C. designed the study.
+* R.C., W.S., and L.D. performed the experiments and analysed the data.
+* R.C. and W.S. implemented the dyngen software package.
+* R.C., W.S., L.D., and Y.S. wrote the manuscript.
+* Y.S. supervised the project.

### 2021-05-12 manuscript/manuscript.tex
update

@@
   pdftitle={dyngen: a multi-modal simulator for spearheading new single-cell omics analyses},
-  pdfauthor={Robrecht Cannoodt,1,2,3; Wouter Saelens,1,2; Louise Deconinck1,2; Yvan Saeys1,2,},
+  pdfauthor={Robrecht Cannoodt,1,2,3; Wouter Saelens,1,2,4; Louise Deconinck1,2; Yvan Saeys1,2,},
   hidelinks,
@@
 \usepackage[sorting=none,url=false,backend=bibtex]{biblatex}
-
-\usepackage{filecontents}
-
-\bibliography{\jobname}
+\addbibresource{\jobname.bib}
+\begin{filecontents}{\jobname.bib}
+@Article{aibar_scenicsinglecellregulatory_2017,
+  title = {{{SCENIC}}: Single-Cell Regulatory Network Inference and Clustering},
+  author = {Sara Aibar and Carmen Bravo Gonz{\a'a}lez-Blas and Thomas Moerman and V{\^a}n Anh Huynh-Thu and Hana Imrichova and Gert Hulselmans and Florian Rambow and Jean-Christophe Marine and Pierre Geurts and Jan Aerts and Joost {van den Oord} and Zeynep Kalender Atak and Jasper Wouters and Stein Aerts},
+  year = {2017},
+  month = {oct},
+  issn = {1548-7091},
+  doi = {10.1038/nmeth.4463},
+  journal = {Nature Methods},
+}
+
+@Article{alpert_alignmentsinglecelltrajectories_2018,
+  title = {Alignment of Single-Cell Trajectories to Compare Cellular Expression Dynamics},
+  author = {Ayelet Alpert and Lindsay S. Moore and Tania Dubovik and Shai S. Shen-Orr},
+  year = {2018},
+  month = {apr},
+  volume = {15},
+  pages = {267--270},
+  issn = {1548-7105},
+  doi = {10.1038/nmeth.4628},
+  abstract = {cellAlign enables quantitative comparisons of expression dynamics within and between single-cell trajectories based on single-cell RNA-seq or mass cytometry data.},
+  copyright = {2018 Nature Publishing Group, a division of Macmillan Publishers Limited. All Rights Reserved.},
+  file = {/home/rcannood/Zotero/storage/QB7FH2T9/nmeth.html},
+  journal = {Nature Methods},
+  language = {en},
+  number = {4},
+}
+
+@Article{bergen_generalizingrnavelocity_2020,
+  title = {Generalizing {{RNA}} Velocity to Transient Cell States through Dynamical Modeling},
+  author = {Volker Bergen and Marius Lange and Stefan Peidli and F. Alexander Wolf and Fabian J. Theis},
+  year = {2020},
+  month = {dec},
+  volume = {38},
+  pages = {1408--1414},
+  issn = {1546-1696},
+  doi = {10.1038/s41587-020-0591-3},
+  abstract = {RNA velocity has opened up new ways of studying cellular differentiation in single-cell RNA-sequencing data.
+It describes the rate of gene expression change for an individual gene at a given time point based on the ratio of its spliced and unspliced messenger RNA (mRNA).
+However, errors in velocity estimates arise if the central assumptions of a common splicing rate and the observation of the full splicing dynamics with steady-state mRNA levels are violated.
+Here we present scVelo, a method that overcomes these limitations by solving the full transcriptional dynamics of splicing kinetics using a likelihood-based dynamical model.
+This generalizes RNA velocity to systems with transient cell states, which are common in development and in response to perturbations.
+We apply scVelo to disentangling subpopulation kinetics in neurogenesis and pancreatic endocrinogenesis.
+We infer gene-specific rates of transcription, splicing and degradation, recover each cell's position in the underlying differentiation processes and detect putative driver genes.
+scVelo will facilitate the study of lineage decisions and gene regulation.},
+  journal = {Nature Biotechnology},
+  number = {12},
+}
+
+@Article{bjornsson_digitaltwinspersonalize_2019,
+  title = {Digital Twins to Personalize Medicine},
+  author = {Bergthor Bj{\"o}rnsson and Carl Borrebaeck and Nils Elander and Thomas Gasslander and Danuta R.
+Gawel and Mika Gustafsson and Rebecka J{\"o}rnsten and Eun Jung Lee and Xinxiu Li and Sandra Lilja and David Mart\'inez-Enguita and Andreas Matussek and Per Sandstr{\"o}m and Samuel Sch{\"a}fer and Margaretha Stenmarker and X.
+F.
+Sun and Oleg Sysoev and Huan Zhang and Mikael Benson and {on behalf of the Swedish Digital Twin Consortium}},
+  year = {2019},
+  month = {dec},
+  volume = {12},
+  pages = {4},
+  issn = {1756-994X},
+  doi = {10.1186/s13073-019-0701-3},
+  abstract = {Personalized medicine requires the integration and processing of vast amounts of data.
+Here, we propose a solution to this challenge that is based on constructing Digital Twins.
+These are high-resolution models of individual patients that are computationally treated with thousands of drugs to find the drug that is optimal for the patient.},
+  journal = {Genome Medicine},
+  number = {1},
+}
+
+@Article{cacchiarelli_aligningsinglecelldevelopmental_2018,
+  title = {Aligning {{Single}}-{{Cell Developmental}} and {{Reprogramming Trajectories Identifies Molecular Determinants}} of {{Myogenic Reprogramming Outcome}}},
+  author = {Davide Cacchiarelli and Xiaojie Qiu and Sanjay Srivatsan and Anna Manfredi and Michael Ziller and Eliah Overbey and Antonio Grimaldi and Jonna Grimsby and Prapti Pokharel and Kenneth J.
+Livak and Shuqiang Li and Alexander Meissner and Tarjei S.
+Mikkelsen and John L.
+Rinn and Cole Trapnell},
+  year = {2018},
+  month = {sep},
+  volume = {7},
+  pages = {258-268.e3},
+  issn = {2405-4712},
+  doi = {10.1016/j.cels.2018.07.006},
+  abstract = {{$<$}h2{$>$}Summary{$<$}/h2{$><$}p{$>$}Cellular reprogramming through manipulation of defined factors holds great promise for large-scale production of cell types needed for use in therapy and for revealing principles of gene regulation.
+However, most reprogramming systems are inefficient, converting only a fraction of cells to the desired state.
+Here, we analyze MYOD-mediated reprogramming of human fibroblasts to myotubes, a well-characterized model system for direct conversion by defined factors, at pseudotemporal resolution using single-cell RNA-seq.
+To expose barriers to efficient conversion, we introduce a novel analytic technique, trajectory alignment, which enables quantitative comparison of gene expression kinetics across two biological processes.
+Reprogrammed cells navigate a trajectory with branch points that correspond to two alternative decision points, with cells that select incorrect branches terminating at aberrant or incomplete reprogramming outcomes.
+Analysis of these branch points revealed insulin and BMP signaling as crucial molecular determinants of reprogramming.
+Single-cell trajectory alignment enables rigorous quantitative comparisons between biological trajectories found in diverse processes in development, reprogramming, and other contexts.{$<$}/p{$>$}},
+  file = {/home/rcannood/Zotero/storage/YQH8GP2I/Cacchiarelli et al. - 2018 - Aligning Single-Cell Developmental and Reprogrammi.pdf;/home/rcannood/Zotero/storage/Q5MUK95E/S2405-4712(18)30314-4.html},
+  journal = {Cell Systems},
+  language = {English},
+  number = {3},
+  pmid = {30195438},
+}
+
+@Article{cannoodt_singlecellomicsdatasets_2018,
+  title = {Single-Cell -Omics Datasets Containing a Trajectory},
+  author = {Robrecht Cannoodt and Wouter Saelens and Helena Todorov and Yvan Saeys},
+  year = {2018},
+  month = {oct},
+  doi = {10.5281/zenodo.1211532},
+  journal = {Zenodo},
+}
+
+@TechReport{conover_multiplecomparisonsprocedures_1979,
+  title = {On Multiple-Comparisons Procedures},
+  author = {William Jay Conover and Ronald L Iman},
+  year = {1979},
+  institution = {{Technical report, Los Alamos Scientific Laboratory}},
+}
+
+@Article{duo_systematicperformanceevaluation_2018,
+  title = {A Systematic Performance Evaluation of Clustering Methods for Single-Cell {{RNA}}-Seq Data},
+  author = {Angelo Du{\a`o} and Mark D. Robinson and Charlotte Soneson},
+  year = {2018},
+  volume = {7},
+  pages = {1141},
+  issn = {2046-1402},
+  doi = {10.12688/f1000research.15666.2},
+  abstract = {Subpopulation identification, usually via some form of unsupervised clustering, is a fundamental step in the analysis of many single-cell RNA-seq data sets.
+This has motivated the development and application of a broad range of clustering methods, based on various underlying algorithms.
+Here, we provide a systematic and extensible performance evaluation of 14 clustering algorithms implemented in R, including both methods developed explicitly for scRNA-seq data and more general-purpose methods.
+The methods were evaluated using nine publicly available scRNA-seq data sets as well as three simulations with varying degree of cluster separability.
+The same feature selection approaches were used for all methods, allowing us to focus on the investigation of the performance of the clustering algorithms themselves.
+We evaluated the ability of recovering known subpopulations, the stability and the run time and scalability of the methods.
+Additionally, we investigated whether the performance could be improved by generating consensus partitions from multiple individual clustering methods.
+We found substantial differences in the performance, run time and stability between the methods, with SC3 and Seurat showing the most favorable results.
+Additionally, we found that consensus clustering typically did not improve the performance compared to the best of the combined methods, but that several of the top-performing methods already perform some type of consensus clustering.
+All the code used for the evaluation is available on GitHub ( https://github.com/markrobinsonuzh/scRNAseq\_clustering\_comparison).
+In addition, an R package providing access to data and clustering results, thereby facilitating inclusion of new methods and data sets, is available from Bioconductor ( https://bioconductor.org/packages/DuoClustering2018).},
+  file = {/home/rcannood/Zotero/storage/ASWVFH87/Duò et al. - 2018 - A systematic performance evaluation of clustering .pdf},
+  journal = {F1000Research},
+  keywords = {Algorithms,Benchmarking,Cluster Analysis,Clustering,Clustering methods,Gene Expression Profiling,Humans,RNA-seq,RNA; Small Cytoplasmic,Sequence Analysis; RNA,Single-Cell Analysis,Single-cell RNA-seq,Software},
+  language = {eng},
+  pmcid = {PMC6134335},
+  pmid = {30271584},
+}
+
+@Article{ewing_combiningtumorgenome_2015,
+  title = {Combining Tumor Genome Simulation with Crowdsourcing to Benchmark Somatic Single-Nucleotide-Variant Detection},
+  author = {Adam D.
+Ewing and Kathleen E.
+Houlahan and Yin Hu and Kyle Ellrott and Cristian Caloian and Takafumi N.
+Yamaguchi and J.
+Christopher Bare and Christine P'ng and Daryl Waggott and Veronica Y.
+Sabelnykova and {ICGC-TCGA DREAM Somatic Mutation Calling Challenge Participants} and Liu Xi and Ninad Dewal and Yu Fan and Wenyi Wang and David Wheeler and Andreas Wilm and Grace Hui Ting and Chenhao Li and Denis Bertrand and Niranjan Nagarajan and Qing-Rong Chen and Chih-Hao Hsu and Ying Hu and Chunhua Yan and Warren Kibbe and Daoud Meerzaman and Kristian Cibulskis and Mara Rosenberg and Louis Bergelson and Adam Kiezun and Amie Radenbaugh and Anne-Sophie Sertier and Anthony Ferrari and Laurie Tonton and Kunal Bhutani and Nancy F.
+Hansen and Difei Wang and Lei Song and Zhongwu Lai and Yang Liao and Wei Shi and Jos{\a'e} Carbonell-Caballero and Joaqu\'in Dopazo and Cheryl C.
+K.
+Lau and Justin Guinney and Michael R.
+Kellen and Thea C.
+Norman and David Haussler and Stephen H.
+Friend and Gustavo Stolovitzky and Adam A.
+Margolin and Joshua M.
+Stuart and Paul C.
+Boutros},
+  year = {2015},
+  month = {jul},
+  volume = {12},
+  pages = {623--630},
+  issn = {1548-7105},
+  doi = {10.1038/nmeth.3407},
+  abstract = {The detection of somatic mutations from cancer genome sequences is key to understanding the genetic basis of disease progression, patient survival and response to therapy.
+Benchmarking is needed for tool assessment and improvement but is complicated by a lack of gold standards, by extensive resource requirements and by difficulties in sharing personal genomic information.
+To resolve these issues, we launched the ICGC-TCGA DREAM Somatic Mutation Calling Challenge, a crowdsourced benchmark of somatic mutation detection algorithms.
+Here we report the BAMSurgeon tool for simulating cancer genomes and the results of 248 analyses of three in silico tumors created with it.
+Different algorithms exhibit characteristic error profiles, and, intriguingly, false positives show a trinucleotide profile very similar to one found in human tumors.
+Although the three simulated tumors differ in sequence contamination (deviation from normal cell sequence) and in subclonality, an ensemble of pipelines outperforms the best individual pipeline in all cases.
+BAMSurgeon is available at https://github.com/adamewing/bamsurgeon/.},
+  copyright = {2015 Nature Publishing Group},
+  file = {/home/rcannood/Zotero/storage/NBPKWHK3/Ewing et al. - 2015 - Combining tumor genome simulation with crowdsourci.pdf;/home/rcannood/Zotero/storage/TDZU2TL9/nmeth.html},
+  journal = {Nature Methods},
+  language = {en},
+  number = {7},
+}
+
+@Article{ferrell_bistabilitybifurcationswaddington_2012,
+  title = {Bistability, {{Bifurcations}}, and {{Waddington}}'s {{Epigenetic Landscape}}},
+  author = {James E Ferrell},
+  year = {2012},
+  month = {jun},
+  volume = {22},
+  pages = {R458--R466},
+  issn = {0960-9822},
+  doi = {10.1016/j.cub.2012.03.045},
+  abstract = {Waddington's epigenetic landscape is probably the most famous and most powerful metaphor in developmental biology.
+Cells, represented by balls, roll downhill through a landscape of bifurcating valleys.
+Each new valley represents a possible cell fate and the ridges between the valleys maintain the cell fate once it has been chosen.
+Here I examine models of two important developmental processes \$\textbackslash backslash\$textemdash\{\} cell-fate induction and lateral inhibition \$\textbackslash backslash\$textemdash\{\} and ask whether the landscapes for these models at least qualitatively resemble Waddington's picture.
+For cell-fate induction, the answer is no.
+The commitment of a cell to a new fate corresponds to the disappearance of a valley from the landscape, not the splitting of one valley into two, and it occurs through a type of bifurcation \$\textbackslash backslash\$textemdash\{\} a saddle-node bifurcation \$\textbackslash backslash\$textemdash\{\} that possesses an intrinsic irreversibility that is missing from Waddington's picture.
+Lateral inhibition, a symmetrical cell\$\textbackslash backslash\$textendash\{\}cell competition process, corresponds better to Waddington's picture, with one valley reversibly splitting into two through a pitchfork bifurcation.
+I propose an alternative epigenetic landscape that has numerous valleys and ridges right from the start, with the process of cell-fate commitment corresponding to the irreversible disappearance of some of these valleys and ridges, via cell-fate induction, complemented by the creation of new valleys and ridges through processes like cell\$\textbackslash backslash\$textendash\{\}cell competition.},
+  file = {/home/rcannood/Zotero/storage/GK4TJ5BV/Ferrell - 2012 - Bistability, Bifurcations, and Waddington's Epigen.pdf;/home/rcannood/Zotero/storage/HUSMJA22/Ferrell - 2012 - Bistability, Bifurcations, and Waddington's Epigen.pdf;/home/rcannood/Zotero/storage/8VLPBTG2/S0960-9822(12)00332-6.html;/home/rcannood/Zotero/storage/XBTP42Y6/S0960982212003326.html},
+  journal = {Current Biology},
+  number = {11},
+}
+
+@Article{gibson_probabilisticmodelprokaryotic_2000,
+  title = {A Probabilistic Model of a Prokaryotic Gene and Its Regulation},
+  author = {Michael A. Gibson and Jehoshua Bruck},
+  year = {2000},
+  journal = {Computational Methods in Molecular Biology: From Genotype to Phenotype, MIT press, Cambridge},
+}
+
+@Article{gillespie_exactstochasticsimulation_1977,
+  title = {Exact Stochastic Simulation of Coupled Chemical Reactions},
+  author = {Daniel T. Gillespie},
+  year = {1977},
+  month = {dec},
+  volume = {81},
+  pages = {2340--2361},
+  issn = {0022-3654},
+  doi = {10.1021/j100540a008},
+  file = {/home/rcannood/Zotero/storage/ELV9AQRF/j100540a008.html},
+  journal = {The Journal of Physical Chemistry},
+  number = {25},
+}
+
+@Article{giorgino_computingvisualizingdynamic_2009,
+  title = {Computing and {{Visualizing Dynamic Time Warping Alignments}} in {{R}}: {{The}} Dtw {{Package}}},
+  author = {Toni Giorgino},
+  year = {2009},
+  month = {sep},
+  doi = {10.18637/jss.v031.i07},
+  abstract = {Dynamic time warping is a popular technique for comparing time series, providing both a distance measure that is insensitive to local compression and stretches and the warping which optimally deforms one of the two input series onto the other.
+A variety of algorithms and constraints have been discussed in the literature.
+The dtw package provides an unification of them; it allows R users to compute time series alignments mixing freely a variety of continuity constraints, restriction windows, endpoints, local distance definitions, and so on.
+The package also provides functions for visualizing alignments and constraints using several classic diagram types.},
+  journal = {Journal of Statistical Software},
+  number = {7},
+}
+
+@Article{glass_passingmessagesbiological_2013,
+  title = {Passing {{Messages}} between {{Biological Networks}} to {{Refine Predicted Interactions}}},
+  author = {Kimberly Glass and Curtis Huttenhower and John Quackenbush and Guo-Cheng Yuan},
+  year = {2013},
+  month = {may},
+  volume = {8},
+  pages = {e64832},
+  doi = {10.1371/journal.pone.0064832},
+  abstract = {Regulatory network reconstruction is a fundamental problem in computational biology.
+There are significant limitations to such reconstruction using individual datasets, and increasingly people attempt to construct networks using multiple, independent datasets obtained from complementary sources, but methods for this integration are lacking.
+We developed PANDA (Passing Attributes between Networks for Data Assimilation), a message-passing model using multiple sources of information to predict regulatory relationships, and used it to integrate protein-protein interaction, gene expression, and sequence motif data to reconstruct genome-wide, condition-specific regulatory networks in yeast as a model.
+The resulting networks were not only more accurate than those produced using individual data sets and other existing methods, but they also captured information regarding specific biological mechanisms and pathways that were missed using other methodologies.
+PANDA is scalable to higher eukaryotes, applicable to specific tissue or cell type data and conceptually generalizable to include a variety of regulatory, interaction, expression, and other genome-scale data.
+An implementation of the PANDA algorithm is available at www.sourceforge.net/projects/panda-net.},
+  journal = {PLOS ONE},
+  number = {5},
+}
+
+@Article{graf_forcingcellschange_2009,
+  title = {Forcing {{Cells}} to {{Change Lineages}}},
+  author = {Thomas Graf and Tariq Enver},
+  year = {2009},
+  month = {dec},
+  volume = {462},
+  pages = {587},
+  issn = {1476-4687},
+  doi = {10.1038/nature08533},
+  abstract = {\$\textbackslash textless\$p\$\textbackslash textgreater\$Though it is important for the success of multicellular organisms that their many different cell types remain stable once specified, advances in the production of induced pluripotent stem (iPS) cells from somatic cells by exposure to a transcription factor cocktail emphasize the potential of almost any cell to be reprogrammed.
+This raises the question of whether interconversion between cell types occurs physiologically or in the context of disease.
+Thomas Graf and Tariq Enver use examples from the blood cell system to illustrate the principles of transcription factor transdifferentiation, which they conclude is a process fundamentally similar to differentiation and different from iPS reprogramming.
+They postulate that conversions of one cell type to another can occur as a normal developmental process, and speculate on the prospects of forced lineage reprogramming in regenerative medicine.\$\textbackslash textless\$/p\$\textbackslash textgreater\$},
+  journal = {Nature},
+  number = {7273},
+}
+
+@Article{hache_gengesystematicgeneration_2009,
+  title = {{{GeNGe}}: Systematic Generation of Gene Regulatory Networks},
+  shorttitle = {{{GeNGe}}},
+  author = {Hendrik Hache and Christoph Wierling and Hans Lehrach and Ralf Herwig},
+  year = {2009},
+  month = {may},
+  volume = {25},
+  pages = {1205--1207},
+  issn = {1367-4803},
+  doi = {10.1093/bioinformatics/btp115},
+  abstract = {Summary: The analysis of gene regulatory networks (GRNs) is a central goal of bioinformatics highly accelerated by the advent of new experimental techniques, such as RNA interference.
+A battery of reverse engineering methods has been developed in recent years to reconstruct the underlying GRNs from these and other experimental data.
+However, the performance of the individual methods is poorly understood and validation of algorithmic performances is still missing to a large extent.
+To enable such systematic validation, we have developed the web application GeNGe (GEne Network GEnerator), a controlled framework for the automatic generation of GRNs.
+The theoretical model for a GRN is a non-linear differential equation system.
+Networks can be user-defined or constructed in a modular way with the option to introduce global and local network perturbations.
+Resulting data can be used, e.g.
+as benchmark data for evaluating GRN reconstruction methods or for predicting effects of perturbations as theoretical counterparts of biological experiments.Availability: Available online at http://genge.molgen.mpg.deContact:hache@molgen.mpg.deSupplementary information:Supplementary data are available at Bioinformatics online.},
+  journal = {Bioinformatics},
+  number = {9},
+}
+
+@Article{holm_simplesequentiallyrejective_1979,
+  title = {A Simple Sequentially Rejective Multiple Test Procedure},
+  author = {Sture Holm},
+  year = {1979},
+  pages = {65--70},
+  journal = {Scandinavian journal of statistics},
+}
+
+@Article{huynh-thu_inferringregulatorynetworks_2010,
+  title = {Inferring {{Regulatory Networks}} from {{Expression Data Using Tree}}-{{Based Methods}}},
+  author = {V{\^a}n Anh Huynh-Thu and Irrthum Alexandre and Wehenkel Louis and Geurts Pierre and Alexandre Irrthum and Louis Wehenkel and Pierre Geurts},
+  year = {2010},
+  month = {jan},
+  volume = {5},
+  pages = {e12776},
+  issn = {1932-6203},
+  doi = {10.1371/journal.pone.0012776},
+  abstract = {One of the pressing open problems of computational systems biology is the elucidation of the topology of genetic regulatory networks (GRNs) using high throughput genomic data, in particular microarray gene expression data.
+The Dialogue for Reverse Engineering Assessments and Methods (DREAM) challenge aims to evaluate the success of GRN inference algorithms on benchmarks of simulated data.
+In this article, we present GENIE3, a new algorithm for the inference of GRNs that was best performer in the DREAM4 In Silico Multifactorial challenge.
+GENIE3 decomposes the prediction of a regulatory network between p genes into p different regression problems.
+In each of the regression problems, the expression pattern of one of the genes (target gene) is predicted from the expression patterns of all the other genes (input genes), using tree-based ensemble methods Random Forests or Extra-Trees.
+The importance of an input gene in the prediction of the target gene expression pattern is taken as an indication of a putative regulatory link.
+Putative regulatory links are then aggregated over all genes to provide a ranking of interactions from which the whole network is reconstructed.
+In addition to performing well on the DREAM4 In Silico Multifactorial challenge simulated data, we show that GENIE3 compares favorably with existing algorithms to decipher the genetic regulatory network of Escherichia coli.
+It doesn't make any assumption about the nature of gene regulation, can deal with combinatorial and non-linear interactions, produces directed GRNs, and is fast and scalable.
+In conclusion, we propose a new algorithm for GRN inference that performs well on both synthetic and real gene expression data.
+The algorithm, based on feature selection with tree-based ensemble methods, is simple and generic, making it adaptable to other types of genomic data and interactions.},
+  file = {/home/rcannood/Zotero/storage/3I6PLCQL/Huynh-Thu et al. - 2010 - Inferring Regulatory Networks from Expression Data Using Tree-Based Methods.pdf},
+  journal = {PLoS ONE},
+  keywords = {Algorithms,Computational Biology,Computational Biology: methods,Escherichia coli,Escherichia coli: genetics,Gene Expression Regulation,Gene Regulatory Networks,Oligonucleotide Array Sequence Analysis},
+  number = {9},
+  pmid = {20927193},
+}
+
+@Article{kanton_organoidsinglecellgenomic_2019,
+  title = {Organoid Single-Cell Genomic Atlas Uncovers Human-Specific Features of Brain Development},
+  author = {Sabina Kanton and Michael James Boyle and Zhisong He and Malgorzata Santel and Anne Weigert and F{\a'a}tima Sanch\'is-Calleja and Patricia Guijarro and Leila Sidow and Jonas Simon Fleck and Dingding Han and Zhengzong Qian and Michael Heide and Wieland B.
+Huttner and Philipp Khaitovich and Svante P{\"a}{{\"a}}bo and Barbara Treutlein and J.
+Gray Camp},
+  year = {2019},
+  month = {oct},
+  volume = {574},
+  pages = {418--422},
+  issn = {1476-4687},
+  doi = {10.1038/s41586-019-1654-9},
+  abstract = {Species comparisons using single-cell transcriptomics and accessible chromatin profiling in stem cell-derived cerebral organoids are used to map dynamic gene-regulatory changes that are unique to humans.},
+  copyright = {2019 The Author(s), under exclusive licence to Springer Nature Limited},
+  file = {/home/rcannood/Zotero/storage/AVPYT7HY/s41586-019-1654-9.html},
+  journal = {Nature},
+  language = {en},
+  number = {7778},
+}
+
+@Article{kuijjer_estimatingsamplespecificregulatory_2015,
+  title = {Estimating Sample-Specific Regulatory Networks},
+  author = {Marieke Lydia Kuijjer and Matthew Tung and GuoCheng Yuan and John Quackenbush and Kimberly Glass},
+  year = {2015},
+  pages = {1--19},
+  abstract = {Biological systems are driven by intricate interactions among the complex array of molecules that comprise the cell.
+Many methods have been developed to reconstruct network models that attempt to capture those interactions.
+These methods often draw on large numbers of measured expression samples to tease out subtle signals and infer connections between genes (or gene products).
+The result is an aggregate network model representing a single estimate for edge likelihoods.
+While informative, aggregate models fail to capture the heterogeneity that is often represented in a population.
+Here we propose a method to reverse engineer sample-specific networks from aggregate network models.
+We demonstrate the accuracy and applicability of our approach in several datasets, including simulated data, microarray expression data from synchronized yeast cells, and RNA-seq data collected from human subjects.
+We show that these sample-specific networks can be used to study the evolution of network topology across time and to characterize shifts in gene regulation that may not be apparent in the expression data.
+We believe the ability to generate sample-specific networks will revolutionize the field of network biology and has the potential to usher in an era of precision network medicine.},
+  file = {/home/rcannood/Zotero/storage/YECTK66M/Kuijjer et al. - 2015 - Estimating sample-specific regulatory networks.pdf},
+}
+
+@Article{kuijjer_estimatingsamplespecificregulatory_2019,
+  title = {Estimating {{Sample}}-{{Specific Regulatory Networks}}},
+  author = {Marieke Lydia Kuijjer and Matthew George Tung and GuoCheng Yuan and John Quackenbush and Kimberly Glass},
+  year = {2019},
+  month = {mar},
+  volume = {14},
+  pages = {226--240},
+  issn = {2589-0042},
+  doi = {10.1016/j.isci.2019.03.021},
+  abstract = {Biological systems are driven by intricate interactions among molecules.
+Many methods have been developed that draw on large numbers of expression samples to infer connections between genes (or their products).
+The result is an aggregate network representing a single estimate for the likelihood of each interaction, or ``edge,'' in the network.
+Although informative, aggregate models fail to capture population heterogeneity.
+Here we propose a method to reverse engineer sample-specific networks from aggregate networks.
+We demonstrate our approach in several contexts, including simulated, yeast microarray, and human lymphoblastoid cell line RNA sequencing data.
+We use these sample-specific networks to study changes in network topology across time and to characterize shifts in gene regulation that were not apparent in the expression data.
+We believe that generating sample-specific networks will greatly facilitate the application of network methods to large, complex, and heterogeneous multi-omic datasets, supporting the emerging field of precision network medicine.,                                        \textbullet{}               We developed LIONESS to extract single-sample networks from aggregate models                                         \textbullet{}               We tested LIONESS using in silico, yeast, and human expression data                                         \textbullet{}               LIONESS-estimated networks are reproducible, accurate, and biologically meaningful                                         \textbullet{}               Single-sample network analysis highlights important biological processes                                 , Biological Sciences; Bioinformatics; Complex Systems},
+  file = {/home/rcannood/Zotero/storage/ITHUQ7NH/Kuijjer et al. - 2019 - Estimating Sample-Specific Regulatory Networks.pdf},
+  journal = {iScience},
+  pmcid = {PMC6463816},
+  pmid = {30981959},
+}
+
+@Article{kuijjer_lionessrsinglesample_2019,
+  title = {{{lionessR}}: Single Sample Network Inference in {{R}}},
+  shorttitle = {{{lionessR}}},
+  author = {Marieke L. Kuijjer and Ping-Han Hsieh and John Quackenbush and Kimberly Glass},
+  year = {2019},
+  month = {oct},
+  volume = {19},
+  pages = {1003},
+  issn = {1471-2407},
+  doi = {10.1186/s12885-019-6235-7},
+  abstract = {In biomedical research, network inference algorithms are typically used to infer complex association patterns between biological entities, such as between genes or proteins, using data from a population.
+This resulting aggregate network, in essence, averages over the networks of those individuals in the population.
+LIONESS (Linear Interpolation to Obtain Network Estimates for Single Samples) is a method that can be used together with a network inference algorithm to extract networks for individual samples in a population.
+The method's key characteristic is that, by modeling networks for individual samples in a data set, it can capture network heterogeneity in a population.
+LIONESS was originally made available as a function within the PANDA (Passing Attributes between Networks for Data Assimilation) regulatory network reconstruction framework.
+However, the LIONESS algorithm is generalizable and can be used to model single sample networks based on a wide range of network inference algorithms.},
+  file = {/home/rcannood/Zotero/storage/RS448RZF/Kuijjer et al. - 2019 - lionessR single sample network inference in R.pdf;/home/rcannood/Zotero/storage/9STH6ZYN/s12885-019-6235-7.html},
+  journal = {BMC Cancer},
+  number = {1},
+}
+
+@Article{lamanno_rnavelocitysingle_2018,
+  title = {{{RNA Velocity}} of {{Single Cells}}},
+  author = {Gioele {La Manno} and Ruslan Soldatov and Amit Zeisel and Emelie Braun and Hannah Hochgerner and Viktor Petukhov and Katja Lidschreiber and Maria E Kastriti and Peter L{\"o}nnerberg and Alessandro Furlan and Jean Fan and Lars E Borm and Zehua Liu and David {van Bruggen} and Jimin Guo and Xiaoling He and Roger Barker and Erik Sundstr{\"o}m and Gon{\c c}alo Castelo-Branco and Patrick Cramer and Igor Adameyko and Sten Linnarsson and Peter V Kharchenko},
+  year = {2018},
+  month = {aug},
+  volume = {560},
+  pages = {494--498},
+  issn = {1476-4687},
+  doi = {10.1038/s41586-018-0414-6},
+  abstract = {RNA velocity, estimated in single cells by comparison of spliced and unspliced mRNA, is a good indicator of transcriptome dynamics and will provide a useful tool for analysis of developmental lineage.},
+  journal = {Nature},
+  number = {7719},
+}
+
+@Article{lin_scclassifysamplesize_2020,
+  title = {{{scClassify}}: Sample Size Estimation and Multiscale Classification of Cells Using Single and Multiple Reference},
+  author = {Yingxin Lin and Yue Cao and Hani Jieun Kim and Agus Salim and Terence P Speed and David M Lin and Pengyi Yang and Jean Yee Hwa Yang},
+  year = {2020},
+  month = {jun},
+  volume = {16},
+  pages = {e9389},
+  publisher = {{John Wiley \& Sons, Ltd}},
+  issn = {1744-4292},
+  doi = {10.15252/msb.20199389},
+  abstract = {Abstract Automated cell type identification is a key computational challenge in single-cell RNA-sequencing (scRNA-seq) data.
+To capitalise on the large collection of well-annotated scRNA-seq datasets, we developed scClassify, a multiscale classification framework based on ensemble learning and cell type hierarchies constructed from single or multiple annotated datasets as references.
+scClassify enables the estimation of sample size required for accurate classification of cell types in a cell type hierarchy and allows joint classification of cells when multiple references are available.
+We show that scClassify consistently performs better than other supervised cell type classification methods across 114 pairs of reference and testing data, representing a diverse combination of sizes, technologies and levels of complexity, and further demonstrate the unique components of scClassify through simulations and compendia of experimental datasets.
+Finally, we demonstrate the scalability of scClassify on large single-cell atlases and highlight a novel application of identifying subpopulations of cells from the Tabula Muris data that were unidentified in the original publication.
+Together, scClassify represents state-of-the-art methodology in automated cell type identification from scRNA-seq data.},
+  journal = {Molecular Systems Biology},
+  keywords = {cell type hierarchy,cell type identification,multiscale classification,sample size estimation,single-cell},
+  number = {6},
+}
+
+@Article{liu_personalizedcharacterizationdiseases_2016,
+  title = {Personalized Characterization of Diseases Using Sample-Specific Networks},
+  author = {Xiaoping Liu and Yuetong Wang and Hongbin Ji and Kazuyuki Aihara and Luonan Chen},
+  year = {2016},
+  volume = {44},
+  pages = {e164--e164},
+  issn = {0305-1048},
+  doi = {10.1093/nar/gkw772},
+  abstract = {A complex disease generally results not from malfunction of individual molecules but from dysfunction of the relevant system or network, which dynamically changes with time and conditions.
+Thus, estimating a condition-specific network from a single sample is crucial to elucidating the molecular mechanisms of complex diseases at the system level.
+However, there is currently no effective way to construct such an individual-specific network by expression profiling of a single sample because of the requirement of multiple samples for computing correlations.
+We developed here with a statistical method, i.e.
+a sample-specific network (SSN) method, which allows us to construct individual-specific networks based on molecular expressions of a single sample.
+Using this method, we can characterize various human diseases at a network level.
+In particular, such SSNs can lead to the identification of individual-specific disease modules as well as driver genes, even without gene sequencing information.
+Extensive analysis by using the Cancer Genome Atlas data not only demonstrated the effectiveness of the method, but also found new individual-specific driver genes and network patterns for various types of cancer.
+Biological experiments on drug resistance further validated one important advantage of our method over the traditional methods, i.e.
+we can even identify such drug resistance genes that actually have no clear differential expression between samples with and without the resistance, due to the additional network information.},
+  file = {/home/rcannood/Zotero/storage/6MSYIL7R/Liu et al. - 2016 - Personalized characterization of diseases using sample-specific networks.pdf},
+  journal = {Nucleic Acids Research},
+  keywords = {cancer,drug resistance,genes,steroid-sensitive nephrotic syndrome},
+  number = {22},
+  pmid = {27596597},
+}
+
+@Article{lizio_gatewaysfantom5promoter_2015,
+  title = {Gateways to the {{FANTOM5}} Promoter Level Mammalian Expression Atlas},
+  author = {Marina Lizio and Jayson Harshbarger and Hisashi Shimoji and Jessica Severin and Takeya Kasukawa and Serkan Sahin and Imad Abugessaisa and Shiro Fukuda and Fumi Hori and Sachi Ishikawa-Kato and Christopher J.
+Mungall and Erik Arner and J.
+Kenneth Baillie and Nicolas Bertin and Hidemasa Bono and Michiel {de Hoon} and Alexander D.
+Diehl and Emmanuel Dimont and Tom C.
+Freeman and Kaori Fujieda and Winston Hide and Rajaram Kaliyaperumal and Toshiaki Katayama and Timo Lassmann and Terrence F.
+Meehan and Koro Nishikata and Hiromasa Ono and Michael Rehli and Albin Sandelin and Erik A.
+Schultes and Peter AC {`t Hoen} and Zuotian Tatum and Mark Thompson and Tetsuro Toyoda and Derek W.
+Wright and Carsten O.
+Daub and Masayoshi Itoh and Piero Carninci and Yoshihide Hayashizaki and Alistair RR Forrest and Hideya Kawaji and {the FANTOM consortium}},
+  year = {2015},
+  month = {jan},
+  volume = {16},
+  pages = {22},
+  issn = {1465-6906},
+  doi = {10.1186/s13059-014-0560-6},
+  abstract = {The FANTOM5 project investigates transcription initiation activities in more than 1,000 human and mouse primary cells, cell lines and tissues using CAGE.
+Based on manual curation of sample information and development of an ontology for sample classification, we assemble the resulting data into a centralized data resource (http://fantom.gsc.riken.jp/5/).
+This resource contains web-based tools and data-access points for the research community to search and extract data related to samples, genes, promoter activities, transcription factors and enhancers across the FANTOM5 atlas.},
+  journal = {Genome Biology},
+  number = {1},
+}
+
+@Article{luecken_currentbestpractices_2019,
+  title = {Current Best Practices in Single-Cell {{RNA}}-Seq Analysis: A Tutorial},
+  shorttitle = {Current Best Practices in Single-Cell {{RNA}}-Seq Analysis},
+  author = {Malte D Luecken and Fabian J Theis},
+  year = {2019},
+  month = {jun},
+  volume = {15},
+  pages = {e8746},
+  issn = {1744-4292},
+  doi = {10.15252/msb.20188746},
+  abstract = {Abstract Single-cell RNA-seq has enabled gene expression to be studied at an unprecedented resolution.
+The promise of this technology is attracting a growing user base for single-cell analysis methods.
+As more analysis tools are becoming available, it is becoming increasingly difficult to navigate this landscape and produce an up-to-date workflow to analyse one's data.
+Here, we detail the steps of a typical single-cell RNA-seq analysis, including pre-processing (quality control, normalization, data correction, feature selection, and dimensionality reduction) and cell- and gene-level downstream analysis.
+We formulate current best-practice recommendations for these steps based on independent comparison studies.
+We have integrated these best-practice recommendations into a workflow, which we apply to a public dataset to further illustrate how these steps work in practice.
+Our documented case study can be found at https://www.github.com/theislab/single-cell-tutorial.
+This review will serve as a workflow tutorial for new entrants into the field, and help established users update their analysis pipelines.},
+  file = {/home/rcannood/Zotero/storage/MY4TZUJT/Luecken and Theis - 2019 - Current best practices in single-cell RNA-seq anal.pdf;/home/rcannood/Zotero/storage/7PFR3DKU/msb.html},
+  journal = {Molecular Systems Biology},
+  keywords = {analysis pipeline development,computational biology,data analysis tutorial,single-cell RNA-seq},
+  number = {6},
+}
+
+@Article{lummertzdarocha_reconstructioncomplexsinglecell_2018,
+  title = {Reconstruction of Complex Single-Cell Trajectories Using {{CellRouter}}},
+  author = {Edroaldo {Lummertz da Rocha} and R.
+Grant Rowe and Vanessa Lundin and Mohan Malleshaiah and Deepak Kumar Jha and Carlos R.
+Rambo and Hu Li and Trista E.
+North and James J.
+Collins and George Q.
+Daley},
+  year = {2018},
+  month = {mar},
+  volume = {9},
+  pages = {892},
+  issn = {2041-1723},
+  doi = {10.1038/s41467-018-03214-y},
+  abstract = {A better understanding of the cell-fate transitions that occur in complex cellular ecosystems in normal development and disease could inform cell engineering efforts and lead to improved therapies.
+However, a major challenge is to simultaneously identify new cell states, and their transitions, to elucidate the gene expression dynamics governing cell-type diversification.
+Here, we present CellRouter, a multifaceted single-cell analysis platform that identifies complex cell-state transition trajectories by using flow networks to explore the subpopulation structure of multi-dimensional, single-cell omics data.
+We demonstrate its versatility by applying CellRouter to single-cell RNA sequencing data sets to reconstruct cell-state transition trajectories during hematopoietic stem and progenitor cell (HSPC) differentiation to the erythroid, myeloid and lymphoid lineages, as well as during re-specification of cell identity by cellular reprogramming of monocytes and B-cells to HSPCs.
+CellRouter opens previously undescribed paths for in-depth characterization of complex cellular ecosystems and establishment of enhanced cell engineering approaches.},
+  journal = {Nature Communications},
+  number = {1},
+}
+
+@Article{marbach_generatingrealisticsilico_2009,
+  title = {Generating {{Realistic In Silico Gene Networks}} for {{Performance Assessment}} of {{Reverse Engineering Methods}}},
+  author = {Daniel Marbach and Thomas Schaffter and Claudio Mattiussi and Dario Floreano},
+  year = {2009},
+  month = {feb},
+  volume = {16},
+  pages = {229--239},
+  issn = {1557-8666},
+  doi = {10.1089/cmb.2008.09TT},
+  abstract = {Reverse engineering methods are typically first tested on simulated data from in silico networks, for systematic and efficient performance assessment, before an application to real biological networks.
+In this paper, we present a method for generating biologically plausible in silico networks, which allow realistic performance assessment of network inference algorithms.
+Instead of using random graph models, which are known to only partly capture the structural properties of biological networks, we generate network structures by extracting modules from known biological interaction networks.
+Using the yeast transcriptional regulatory network as a test case, we show that extracted modules have a biologically plausible connectivity because they preserve functional and structural properties of the original network.
+Our method was selected to generate the {"}gold standard{"} networks for the gene network reverse engineering challenge of the third DREAM conference (Dialogue on Reverse Engineering Assessment and Methods 2008, Cambridge, MA).},
+  file = {/home/rcannood/Zotero/storage/NXLVA76Q/Marbach et al. - 2009 - Generating Realistic In Silico Gene Networks for Performance Assessment of Reverse Engineering Methods.pdf},
+  journal = {Journal of Computational Biology},
+  keywords = {Algorithms,Computational Biology,Computational Biology: methods,Computer Simulation,DRE,Gene Expression Profiling,Gene Expression Profiling: methods,gene regulatory networks,Gene Regulatory Networks,Genetic,Models,reverse engineering,Saccharomyces cerevisiae,Saccharomyces cerevisiae: genetics,Saccharomyces cerevisiae: metabolism,Software},
+  number = {2},
+  pmid = {19183003},
+}
+
+@Article{marbach_revealingstrengthsweaknesses_2010,
+  title = {Revealing Strengths and Weaknesses of Methods for Gene Network Inference},
+  author = {Daniel Marbach and Robert J Prill and Thomas Schaffter and Claudio Mattiussi and Dario Floreano and Gustavo Stolovitzky},
+  year = {2010},
+  month = {apr},
+  volume = {107},
+  pages = {6286--6291},
+  issn = {1091-6490},
+  doi = {10.1073/pnas.0913357107},
+  abstract = {Numerous methods have been developed for inferring gene regulatory networks from expression data, however, both their absolute and comparative performance remain poorly understood.
+In this paper, we introduce a framework for critical performance assessment of methods for gene network inference.
+We present an in silico benchmark suite that we provided as a blinded, community-wide challenge within the context of the DREAM (Dialogue on Reverse Engineering Assessment and Methods) project.
+We assess the performance of 29 gene-network-inference methods, which have been applied independently by participating teams.
+Performance profiling reveals that current inference methods are affected, to various degrees, by different types of systematic prediction errors.
+In particular, all but the best-performing method failed to accurately infer multiple regulatory inputs (combinatorial regulation) of genes.
+The results of this community-wide experiment show that reliable network inference from gene expression data remains an unsolved problem, and they indicate potential ways of network reconstruction improvements.},
+  file = {/home/rcannood/Zotero/storage/784SY83Y/Marbach et al. - 2010 - Revealing strengths and weaknesses of methods for gene network inference.pdf},
+  journal = {Proceedings of the National Academy of Sciences},
+  keywords = {Biometry,community experiment,Computational Biology,Computational Biology: methods,DREAM challenge,Gene Expression Profiling,Gene Regulatory Networks,performance assessment,reverse engineering,transcriptional regulatory networks},
+  number = {14},
+  pmid = {20308593},
+}
+
+@Article{marbach_wisdomcrowdsrobust_2012,
+  title = {Wisdom of {{Crowds}} for {{Robust Gene Network Inference}}},
+  author = {Daniel Marbach and James C Costello and Robert K{\"u}ffner and Nicole M Nicci Vega and Robert J Prill and Diogo M Camacho and Kyle R Allison and Manolis Kellis and James J Collins and Gustavo Stolovitzky and {DREAM5 Consortium} and Manolis Kellis and James J Collins and Gustavo Stolovitzky},
+  year = {2012},
+  month = {jul},
+  volume = {9},
+  pages = {796--804},
+  issn = {1548-7091},
+  doi = {10.1038/nmeth.2016},
+  abstract = {Reconstructing gene regulatory networks from high-throughput data is a long-standing problem.
+Through the DREAM project (Dialogue on Reverse Engineering Assessment and Methods), we performed a comprehensive blind assessment of over thirty network inference methods on Escherichia coli, Staphylococcus aureus, Saccharomyces cerevisiae, and in silico microarray data.
+We characterize performance, data requirements, and inherent biases of different inference approaches offering guidelines for both algorithm application and development.
+We observe that no single inference method performs optimally across all datasets.
+In contrast, integration of predictions from multiple inference methods shows robust and high performance across diverse datasets.
+Thereby, we construct high-confidence networks for E.
+coli and S.
+aureus, each comprising \$\textbackslash backslash\$textasciitilde\{\}1700 transcriptional interactions at an estimated precision of 50\%.
+We experimentally test 53 novel interactions in E.
+coli, of which 23 were supported (43\%).
+Our results establish community-based methods as a powerful and robust tool for the inference of transcriptional gene regulatory networks.},
+  file = {/home/rcannood/Zotero/storage/AQRNS57R/Marbach et al.
+- 2012 - Wisdom of crowds for robust gene network inference.pdf;/home/rcannood/Zotero/storage/KGJVQ9WL/Marbach et al.
+- 2012 - Wisdom of crowds for robust gene network inference.pdf},
+  journal = {Nature methods},
+  keywords = {Algorithms,Bacterial,Bacterial: genetics,Computational Biology,Escherichia coli,Escherichia coli: genetics,Gene Expression Regulation,Gene Regulatory Networks,Genetic,Genetic: genetics,Oligonucleotide Array Sequence Analysis,Saccharomyces cerevisiae,Saccharomyces cerevisiae: genetics,Software,Staphylococcus aureus,Staphylococcus aureus: genetics,Transcription},
+  number = {8},
+  pmid = {22796662},
+}
+
+@Article{mcfaline-figueroa_pooledsinglecellgenetic_2019,
+  title = {A Pooled Single-Cell Genetic Screen Identifies Regulatory Checkpoints in the Continuum of the Epithelial-to-Mesenchymal Transition},
+  author = {Jos{\a'e} L. McFaline-Figueroa and Andrew J. Hill and Xiaojie Qiu and Dana Jackson and Jay Shendure and Cole Trapnell},
+  year = {2019},
+  month = {sep},
+  volume = {51},
+  pages = {1389--1398},
+  issn = {1546-1718},
+  doi = {10.1038/s41588-019-0489-5},
+  abstract = {Transcriptional profiling of 61,052 cells undergoing epithelial-to-mesenchymal transition (EMT) shows continuous waves of gene regulation, not discrete stages.
+A CRISPR screen orders EMT regulators along EMT progression.},
+  copyright = {2019 The Author(s), under exclusive licence to Springer Nature America, Inc.},
+  file = {/home/rcannood/Zotero/storage/RAQRP6UC/s41588-019-0489-5.html},
+  journal = {Nature Genetics},
+  language = {en},
+  number = {9},
+}
+
+@Article{moerman_grnboost2arboretoefficient_2019,
+  title = {{{GRNBoost2}} and {{Arboreto}}: Efficient and Scalable Inference of Gene Regulatory Networks},
+  author = {Thomas Moerman and Sara {Aibar Santos} and Carmen {Bravo Gonz{\a'a}lez-Blas} and Jaak Simm and Yves Moreau and Jan Aerts and Stein Aerts},
+  year = {2019},
+  month = {jun},
+  volume = {35},
+  pages = {2159--2161},
+  issn = {1367-4803},
+  doi = {10.1093/bioinformatics/bty916},
+  abstract = {Inferring a Gene Regulatory Network (GRN) from gene expression data is a computationally expensive task, exacerbated by increasing data sizes due to advances in high-throughput gene profiling technology, such as single-cell RNA-seq.
+To equip researchers with a toolset to infer GRNs from large expression datasets, we propose GRNBoost2 and the Arboreto framework.
+GRNBoost2 is an efficient algorithm for regulatory network inference using gradient boosting, based on the GENIE3 architecture.
+Arboreto is a computational framework that scales up GRN inference algorithms complying with this architecture.
+Arboreto includes both GRNBoost2 and an improved implementation of GENIE3, as a user-friendly open source Python package.Arboreto is available under the 3-Clause BSD license at http://arboreto.readthedocs.io.Supplementary data are available at Bioinformatics online.},
+  journal = {Bioinformatics},
+  number = {12},
+}
+
+@Article{papadopoulos_prossttprobabilisticsimulation_2019,
+  title = {{{PROSSTT}}: Probabilistic Simulation of Single-Cell {{RNA}}-Seq Data for Complex Differentiation Processes},
+  shorttitle = {{{PROSSTT}}},
+  author = {Nikolaos Papadopoulos and Parra R. Gonzalo and Johannes S{\"o}ding},
+  year = {2019},
+  month = {sep},
+  volume = {35},
+  pages = {3517--3519},
+  issn = {1367-4811},
+  doi = {10.1093/bioinformatics/btz078},
+  abstract = {SUMMARY: Cellular lineage trees can be derived from single-cell RNA sequencing snapshots of differentiating cells.
+Currently, only datasets with simple topologies are available.
+To test and further develop tools for lineage tree reconstruction, we need test datasets with known complex topologies.
+PROSSTT can simulate scRNA-seq datasets for differentiation processes with lineage trees of any desired complexity, noise level, noise model and size.
+PROSSTT also provides scripts to quantify the quality of predicted lineage trees.
+AVAILABILITY AND IMPLEMENTATION: https://github.com/soedinglab/prosstt.
+SUPPLEMENTARY INFORMATION: Supplementary data are available at Bioinformatics online.},
+  file = {/home/rcannood/Zotero/storage/UERERPJX/Papadopoulos et al. - 2019 - PROSSTT probabilistic simulation of single-cell R.pdf},
+  journal = {Bioinformatics (Oxford, England)},
+  keywords = {Cell Differentiation,Gene Expression Profiling,RNA-Seq,Single-Cell Analysis,Software},
+  language = {eng},
+  number = {18},
+  pmcid = {PMC6748774},
+  pmid = {30715210},
+}
+
+@Article{parra_reconstructingcomplexlineage_2019,
+  title = {Reconstructing Complex Lineage Trees from {{scRNA}}-Seq Data Using {{MERLoT}}},
+  author = {R. Gonzalo Parra and Nikolaos Papadopoulos and Laura Ahumada-Arranz and Jakob El Kholtei and Noah Mottelson and Yehor Horokhovsky and Barbara Treutlein and Johannes Soeding},
+  year = {2019},
+  month = {sep},
+  volume = {47},
+  pages = {8961--8974},
+  issn = {1362-4962},
+  doi = {10.1093/nar/gkz706},
+  abstract = {Advances in single-cell transcriptomics techniques are revolutionizing studies of cellular differentiation and heterogeneity.
+It has become possible to track the trajectory of thousands of genes across the cellular lineage trees that represent the temporal emergence of cell types during dynamic processes.
+However, reconstruction of cellular lineage trees with more than a few cell fates has proved challenging.
+We present MERLoT (https://github.com/soedinglab/merlot), a flexible and user-friendly tool to reconstruct complex lineage trees from single-cell transcriptomics data.
+It can impute temporal gene expression profiles along the reconstructed tree.
+We show MERLoT's capabilities on various real cases and hundreds of simulated datasets.},
+  file = {/home/rcannood/Zotero/storage/86VN2MRY/Parra et al. - 2019 - Reconstructing complex lineage trees from scRNA-se.pdf},
+  journal = {Nucleic Acids Research},
+  keywords = {Algorithms,Animals,Cell Differentiation,Cell Lineage,Cluster Analysis,Gene Expression Profiling,Hematopoietic Stem Cells,Humans,Mice,Mouse Embryonic Stem Cells,Sequence Analysis; RNA,Single-Cell Analysis,Software,Time Factors,Transcriptome},
+  language = {eng},
+  number = {17},
+  pmcid = {PMC6755101},
+  pmid = {31428793},
+}
+
+@Manual{patil_pairwisecomparisonsmultiplepairwise_2019,
+  title = {{{pairwiseComparisons}}: {{Multiple}} Pairwise Comparison Tests},
+  author = {Indrajeet Patil},
+  year = {2019},
+  type = {Manual},
+}
+
+@Article{prill_rigorousassessmentsystems_2010,
+  title = {Towards a {{Rigorous Assessment}} of {{Systems Biology Models}}: {{The DREAM3 Challenges}}},
+  author = {Robert J Prill and Marbach Daniel and Saez-Rodriguez Julio and Sorger Peter K.
+and Alexopoulos Leonidas G.
+and Xue Xiaowei and Clarke Neil D.
+and Altan-Bonnet Gregoire and Stolovitzky Gustavo and Daniel Marbach and Julio Saez-Rodriguez and Peter K Sorger and Leonidas G Alexopoulos and Xiaowei Xue and Neil D Clarke and Gregoire Altan-Bonnet and Gustavo Stolovitzky and Marbach Daniel and Saez-Rodriguez Julio and Sorger Peter K.
+and Alexopoulos Leonidas G.
+and Xue Xiaowei and Clarke Neil D.
+and Altan-Bonnet Gregoire and Stolovitzky Gustavo},
+  year = {2010},
+  month = {jan},
+  volume = {5},
+  pages = {e9202},
+  issn = {1932-6203},
+  doi = {10.1371/journal.pone.0009202},
+  abstract = {BACKGROUND: Systems biology has embraced computational modeling in response to the quantitative nature and increasing scale of contemporary data sets.
+The onslaught of data is accelerating as molecular profiling technology evolves.
+The Dialogue for Reverse Engineering Assessments and Methods (DREAM) is a community effort to catalyze discussion about the design, application, and assessment of systems biology models through annual reverse-engineering challenges.
+METHODOLOGY AND PRINCIPAL FINDINGS: We describe our assessments of the four challenges associated with the third DREAM conference which came to be known as the DREAM3 challenges: signaling cascade identification, signaling response prediction, gene expression prediction, and the DREAM3 in silico network challenge.
+The challenges, based on anonymized data sets, tested participants in network inference and prediction of measurements.
+Forty teams submitted 413 predicted networks and measurement test sets.
+Overall, a handful of best-performer teams were identified, while a majority of teams made predictions that were equivalent to random.
+Counterintuitively, combining the predictions of multiple teams (including the weaker teams) can in some cases improve predictive power beyond that of any single method.
+CONCLUSIONS: DREAM provides valuable feedback to practitioners of systems biology modeling.
+Lessons learned from the predictions of the community provide much-needed context for interpreting claims of efficacy of algorithms described in the scientific literature.},
+  file = {/home/rcannood/Zotero/storage/Q5E22H2F/Prill et al. - 2010 - Towards a Rigorous Assessment of Systems Biology Models The DREAM3 Challenges.pdf},
+  journal = {PLoS ONE},
+  keywords = {Algorithms,Animals,Biological,Cluster Analysis,Computational Biology,Computational Biology: methods,Gene Expression Profiling,Gene Expression Profiling: methods,Gene Regulatory Networks,Humans,Models,Protein Interaction Mapping,Protein Interaction Mapping: methods,Reproducibility of Results,Signal Transduction,Software,Systems Biology,Systems Biology: methods},
+  number = {2},
+  pmid = {20186320},
+}
+
+@Article{rekhtman_directinteractionhematopoietic_1999,
+  title = {Direct Interaction of Hematopoietic Transcription Factors {{PU}}.1 and {{GATA}}-1: Functional Antagonism in Erythroid Cells},
+  shorttitle = {Direct Interaction of Hematopoietic Transcription Factors {{PU}}.1 and {{GATA}}-1},
+  author = {N. Rekhtman and F. Radparvar and T. Evans and A. I. Skoultchi},
+  year = {1999},
+  month = {jun},
+  volume = {13},
+  pages = {1398--1411},
+  issn = {0890-9369},
+  doi = {10.1101/gad.13.11.1398},
+  abstract = {Malignant transformation usually inhibits terminal cell differentiation but the precise mechanisms involved are not understood.
+PU.1 is a hematopoietic-specific Ets family transcription factor that is required for development of some lymphoid and myeloid lineages.
+PU.1 can also act as an oncoprotein as activation of its expression in erythroid precursors by proviral insertion or transgenesis causes erythroleukemias in mice.
+Restoration of terminal differentiation in the mouse erythroleukemia (MEL) cells requires a decline in the level of PU.1, indicating that PU.1 can block erythroid differentiation.
+Here we investigate the mechanism by which PU.1 interferes with erythroid differentiation.
+We find that PU.1 interacts directly with GATA-1, a zinc finger transcription factor required for erythroid differentiation.
+Interaction between PU.1 and GATA-1 requires intact DNA-binding domains in both proteins.
+PU.1 represses GATA-1-mediated transcriptional activation.
+Both the DNA binding and transactivation domains of PU.1 are required for repression and both domains are also needed to block terminal differentiation in MEL cells.
+We also show that ectopic expression of PU.1 in Xenopus embryos is sufficient to block erythropoiesis during normal development.
+Furthermore, introduction of exogenous GATA-1 in both MEL cells and Xenopus embryos and explants relieves the block to erythroid differentiation imposed by PU.1.
+Our results indicate that the stoichiometry of directly interacting but opposing transcription factors may be a crucial determinant governing processes of normal differentiation and malignant transformation.},
+  file = {/home/rcannood/Zotero/storage/77DJ34BY/Rekhtman et al. - 1999 - Direct interaction of hematopoietic transcription .pdf},
+  journal = {Genes \& Development},
+  keywords = {3T3 Cells,Animals,Binding Sites,Cell Differentiation,DNA-Binding Proteins,Erythroid Precursor Cells,Erythroid-Specific DNA-Binding Factors,GATA1 Transcription Factor,HeLa Cells,Hematopoiesis,Humans,Mice,Proto-Oncogene Proteins,Recombinant Fusion Proteins,Trans-Activators,Transcription Factors,Transcription; Genetic,Transcriptional Activation,Xenopus laevis,Zinc Fingers},
+  language = {eng},
+  number = {11},
+  pmcid = {PMC316770},
+  pmid = {10364157},
+}
+
+@Article{roy_systemgeneratingtranscription_2008,
+  title = {A System for Generating Transcription Regulatory Networks with Combinatorial Control of Transcription},
+  author = {Sushmita Roy and Margaret Werner-Washburne and Terran Lane},
+  year = {2008},
+  month = {may},
+  volume = {24},
+  pages = {1318--1320},
+  issn = {1367-4803},
+  doi = {10.1093/bioinformatics/btn126},
+  abstract = {Summary: We have developed a new software system, REgulatory Network generator with COmbinatorial control (RENCO), for automatic generation of differential equations describing pre-transcriptional combinatorics in artificial regulatory networks.
+RENCO has the following benefits: (a) it explicitly models protein\textendash protein interactions among transcription factors, (b) it captures combinatorial control of transcription factors on target genes and (c) it produces output in Systems Biology Markup Language (SBML) format, which allows these equations to be directly imported into existing simulators.
+Explicit modeling of the protein interactions allows RENCO to incorporate greater mechanistic detail of the transcription machinery compared to existing models and can provide a better assessment of algorithms for regulatory network inference.Availability: RENCO is a C++ command line program, available at http://sourceforge.net/projects/renco/Contact:terran@cs.unm.eduSupplementary information: Supplementary data are available at Bioinformatics online.},
+  journal = {Bioinformatics},
+  number = {10},
+}
+
+@Article{saelens_comparisonsinglecelltrajectory_2019,
+  title = {A Comparison of Single-Cell Trajectory Inference Methods},
+  author = {Wouter Saelens and Robrecht Cannoodt and Helena Todorov and Yvan Saeys},
+  year = {2019},
+  volume = {37},
+  issn = {15461696},
+  doi = {10.1038/s41587-019-0071-9},
+  abstract = {Trajectory inference approaches analyze genome-wide omics data from thousands of single cells and computationally infer the order of these cells along developmental trajectories.
+Although more than 70 trajectory inference tools have already been developed, it is challenging to compare their performance because the input they require and output models they produce vary substantially.
+Here, we benchmark 45 of these methods on 110 real and 229 synthetic datasets for cellular ordering, topology, scalability and usability.
+Our results highlight the complementarity of existing tools, and that the choice of method should depend mostly on the dataset dimensions and trajectory topology.
+Based on these results, we develop a set of guidelines to help users select the best method for their dataset.
+Our freely available data and evaluation pipeline ( https://benchmark.dynverse.org ) will aid in the development of improved tools designed to analyze increasingly large and complex single-cell datasets.},
+  file = {/home/rcannood/Zotero/storage/JKZCFAVA/Saelens et al. - 2019 - A comparison of single-cell trajectory inference methods.pdf;/home/rcannood/Zotero/storage/T9HMNIDB/Saelens2019.pdf},
+  journal = {Nature Biotechnology},
+  number = {May},
+}
+
+@Article{schaffter_genenetweaversilicobenchmark_2011,
+  title = {{{GeneNetWeaver}}: In Silico Benchmark Generation and Performance Profiling of Network Inference Methods.},
+  shorttitle = {\{\vphantom\}{{GeneNetWeaver}}\vphantom\{\}},
+  author = {Thomas Schaffter and Daniel Marbach and Dario Floreano},
+  year = {2011},
+  month = {aug},
+  volume = {27},
+  pages = {2263--2270},
+  issn = {1367-4811},
+  doi = {10.1093/bioinformatics/btr373},
+  abstract = {MOTIVATION: Over the last decade, numerous methods have been developed for inference of regulatory networks from gene expression data.
+However, accurate and systematic evaluation of these methods is hampered by the difficulty of constructing adequate benchmarks and the lack of tools for a differentiated analysis of network predictions on such benchmarks.
+RESULTS: Here, we describe a novel and comprehensive method for in silico benchmark generation and performance profiling of network inference methods available to the community as an open-source software called GeneNetWeaver (GNW).
+In addition to the generation of detailed dynamical models of gene regulatory networks to be used as benchmarks, GNW provides a network motif analysis that reveals systematic prediction errors, thereby indicating potential ways of improving inference methods.
+The accuracy of network inference methods is evaluated using standard metrics such as precision-recall and receiver operating characteristic curves.
+We show how GNW can be used to assess the performance and identify the strengths and weaknesses of six inference methods.
+Furthermore, we used GNW to provide the international Dialogue for Reverse Engineering Assessments and Methods (DREAM) competition with three network inference challenges (DREAM3, DREAM4 and DREAM5).
+AVAILABILITY: GNW is available at http://gnw.sourceforge.net along with its Java source code, user manual and supporting data.
+SUPPLEMENTARY INFORMATION: Supplementary data are available at Bioinformatics online.
+CONTACT: dario.floreano@epfl.ch.},
+  file = {/home/rcannood/Zotero/storage/P5JMJUBN/Schaffter, Marbach, Floreano - 2011 - GeneNetWeaver in silico benchmark generation and performance profiling of network inference method.pdf},
+  journal = {Bioinformatics},
+  keywords = {Benchmarking,Escherichia coli,Escherichia coli: genetics,Gene Expression,Gene Regulatory Networks,Genetic,Models,Software},
+  number = {16},
+  pmid = {21697125},
+}
+
+@Article{schilstra_biologicgeneexpression_2008,
+  title = {Bio-{{Logic}}: {{Gene Expression}} and the {{Laws}} of {{Combinatorial Logic}}},
+  shorttitle = {Bio-{{Logic}}},
+  author = {Maria J. Schilstra and Chrystopher L. Nehaniv},
+  year = {2008},
+  month = {jan},
+  volume = {14},
+  pages = {121--133},
+  issn = {1064-5462},
+  doi = {10.1162/artl.2008.14.1.121},
+  abstract = {Abstract At the heart of the development of fertilized eggs into fully formed organisms and the adaptation of cells to changed conditions are genetic regulatory networks (GRNs).
+In higher multicellular organisms, signal selection and multiplexing are performed at the cis-regulatory domains of genes, where combinations of transcription factors (TFs) regulate the rates at which the genes are transcribed into mRNA.
+To be able to act as activators or repressors of gene transcription, TFs must first bind to target sequences on the regulatory domains.
+Two TFs that act in concert may bind entirely independently of each other, but more often binding of the first one will alter the affinity of the other for its binding site.
+This article presents a systematic investigation into the effect of TF binding dependences on the predicted regulatory function of this bio-logic.
+Four extreme scenarios, commonly used to classify enzyme activation and inhibition patterns, for the binding of two TFs were explored: independent (the TFs bind without affecting each other's affinities), competitive (the TFs compete for the same binding site), ordered (the TFs bind in a compulsory order), and joint binding (the TFs either bind as a preformed complex, or binding of one is virtually impossible in the absence of the other).
+The conclusions are: (1) the laws of combinatorial logic hold only for systems with independently binding TFs; (2) systems formed according to the other scenarios can mimic the functions of their Boolean logical counterparts, but cannot be combined or decomposed in the same way; and (3) the continuously scaled output of systems consisting of competitively binding activators and repressors can be controlled more robustly than that of single TF or (quasi-)logical multi-TF systems.},
+  file = {/home/rcannood/Zotero/storage/C7UWQGI3/Schilstra and Nehaniv - 2008 - Bio-Logic Gene Expression and the Laws of Combina.pdf;/home/rcannood/Zotero/storage/KXDLUMAG/artl.2008.14.1.html},
+  journal = {Artificial Life},
+  number = {1},
+}
+
+@Article{schwanhausser_globalquantificationmammalian_2011,
+  title = {Global Quantification of Mammalian Gene Expression Control},
+  author = {Bj{\"o}rn Schwanh{\"a}usser and Dorothea Busse and Na Li and Gunnar Dittmar and Johannes Schuchhardt and Jana Wolf and Wei Chen and Matthias Selbach},
+  year = {2011},
+  month = {may},
+  volume = {473},
+  pages = {337--342},
+  issn = {1476-4687},
+  doi = {10.1038/nature10098},
+  abstract = {Gene expression is a multistep process that involves the transcription, translation and turnover of messenger RNAs and proteins.
+Although it is one of the most fundamental processes of life, the entire cascade has never been quantified on a genome-wide scale.
+Here we simultaneously measured absolute mRNA and protein abundance and turnover by parallel metabolic pulse labelling for more than 5,000 genes in mammalian cells.
+Whereas mRNA and protein levels correlated better than previously thought, corresponding half-lives showed no correlation.
+Using a quantitative model we have obtained the first genome-scale prediction of synthesis rates of mRNAs and proteins.
+We find that the cellular abundance of proteins is predominantly controlled at the level of translation.
+Genes with similar combinations of mRNA and protein stability shared functional properties, indicating that half-lives evolved under energetic and dynamic constraints.
+Quantitative information about all stages of gene expression provides a rich resource and helps to provide a greater understanding of the underlying design principles.},
+  journal = {Nature},
+  number = {7347},
+}
+
+@Article{smith_spatialstochasticintracellular_2019,
+  title = {Spatial {{Stochastic Intracellular Kinetics}}: {{A Review}} of {{Modelling Approaches}}},
+  author = {Stephen Smith and Ramon Grima},
+  year = {2019},
+  month = {aug},
+  volume = {81},
+  pages = {2960--3009},
+  issn = {1522-9602},
+  doi = {10.1007/s11538-018-0443-1},
+  abstract = {Models of chemical kinetics that incorporate both stochasticity and diffusion are an increasingly common tool for studying biology.
+The variety of competing models is vast, but two stand out by virtue of their popularity: the reaction\textendash diffusion master equation and Brownian dynamics.
+In this review, we critically address a number of open questions surrounding these models: How can they be justified physically?
+How do they relate to each other?
+How do they fit into the wider landscape of chemical models, ranging from the rate equations to molecular dynamics?
+This review assumes no prior knowledge of modelling chemical kinetics and should be accessible to a wide range of readers.},
+  journal = {Bulletin of Mathematical Biology},
+  number = {8},
+}
+
+@Article{soneson_biasrobustnessscalability_2018,
+  title = {Bias, Robustness and Scalability in Single-Cell Differential Expression Analysis},
+  author = {Charlotte Soneson and Mark D. Robinson},
+  year = {2018},
+  month = {apr},
+  volume = {15},
+  pages = {255--261},
+  issn = {1548-7105},
+  doi = {10.1038/nmeth.4612},
+  abstract = {Many methods have been used to determine differential gene expression from single-cell RNA (scRNA)-seq data.
+We evaluated 36 approaches using experimental and synthetic data and found considerable differences in the number and characteristics of the genes that are called differentially expressed.
+Prefiltering of lowly expressed genes has important effects, particularly for some of the methods developed for bulk RNA-seq data analysis.
+However, we found that bulk RNA-seq analysis methods do not generally perform worse than those developed specifically for scRNA-seq.
+We also present conquer, a repository of consistently processed, analysis-ready public scRNA-seq data sets that is aimed at simplifying method evaluation and reanalysis of published results.
+Each data set provides abundance estimates for both genes and transcripts, as well as quality control and exploratory analysis reports.},
+  journal = {Nature Methods},
+  keywords = {Computational Biology,Gene Expression Profiling,Gene Expression Regulation,High-Throughput Nucleotide Sequencing,RNA,Sequence Analysis; RNA,Single-Cell Analysis,Software},
+  language = {eng},
+  number = {4},
+  pmid = {29481549},
+}
+
+@Article{street_slingshotcelllineage_2018,
+  title = {Slingshot: {{Cell Lineage}} and {{Pseudotime Inference}} for {{Single}}-{{Cell Transcriptomics}}},
+  shorttitle = {Slingshot},
+  author = {Kelly Street and Davide Risso and Russell B Fletcher and Diya Das and John Ngai and Nir Yosef and Elizabeth Purdom and Sandrine Dudoit},
+  year = {2018},
+  month = {jun},
+  volume = {19},
+  pages = {477},
+  issn = {1471-2164},
+  doi = {10.1186/s12864-018-4772-0},
+  abstract = {Single-cell transcriptomics allows researchers to investigate complex communities of heterogeneous cells.
+It can be applied to stem cells and their descendants in order to chart the progression from multipotent progenitors to fully differentiated cells.
+While a variety of statistical and computational methods have been proposed for inferring cell lineages, the problem of accurately characterizing multiple branching lineages remains difficult to solve.},
+  journal = {BMC Genomics},
+  number = {1},
+}
+
+@Article{vallejos_normalizingsinglecellrna_2017,
+  title = {Normalizing Single-Cell {{RNA}} Sequencing Data: Challenges and Opportunities},
+  shorttitle = {Normalizing Single-Cell {{RNA}} Sequencing Data},
+  author = {Catalina A. Vallejos and Davide Risso and Antonio Scialdone and Sandrine Dudoit and John C. Marioni},
+  year = {2017},
+  month = {jun},
+  volume = {14},
+  pages = {565--571},
+  publisher = {{Nature Publishing Group}},
+  issn = {1548-7105},
+  doi = {10.1038/nmeth.4292},
+  abstract = {This Perspective examines single-cell RNA-seq data challenges and the need for normalization methods designed specifically for single-cell data in order to remove technical biases.},
+  copyright = {2017 Nature Publishing Group, a division of Macmillan Publishers Limited. All Rights Reserved.},
+  journal = {Nature Methods},
+  language = {en},
+  number = {6},
+}
+
+@Article{vandenbulcke_syntrengeneratorsynthetic_2006,
+  title = {{{SynTReN}}: A Generator of Synthetic Gene Expression Data for Design and Analysis of Structure Learning Algorithms},
+  shorttitle = {{{SynTReN}}},
+  author = {Tim {Van den Bulcke} and Koenraad {Van Leemput} and Bart Naudts and Piet {van Remortel} and Hongwu Ma and Alain Verschoren and Bart {De Moor} and Kathleen Marchal},
+  year = {2006},
+  month = {jan},
+  volume = {7},
+  pages = {43},
+  issn = {1471-2105},
+  doi = {10.1186/1471-2105-7-43},
+  abstract = {The development of algorithms to infer the structure of gene regulatory networks based on expression data is an important subject in bioinformatics research.
+Validation of these algorithms requires benchmark data sets for which the underlying network is known.
+Since experimental data sets of the appropriate size and design are usually not available, there is a clear need to generate well-characterized synthetic data sets that allow thorough testing of learning algorithms in a fast and reproducible manner.},
+  journal = {BMC Bioinformatics},
+  language = {en},
+  number = {1},
+}
+
+@Article{vandesande_scalablescenicworkflow_2020,
+  title = {A Scalable {{SCENIC}} Workflow for Single-Cell Gene Regulatory Network Analysis},
+  author = {Bram {Van de Sande} and Christopher Flerin and Kristofer Davie and Maxime {De Waegeneer} and Gert Hulselmans and Sara Aibar and Ruth Seurinck and Wouter Saelens and Robrecht Cannoodt and Quentin Rouchon and Toni Verbeiren and Dries {De Maeyer} and Joke Reumers and Yvan Saeys and Stein Aerts},
+  year = {2020},
+  month = {jul},
+  volume = {15},
+  pages = {2247--2276},
+  issn = {1750-2799},
+  doi = {10.1038/s41596-020-0336-2},
+  abstract = {This protocol explains how to perform a fast SCENIC analysis alongside standard best practices steps on single-cell RNA-sequencing data using software containers and Nextflow pipelines.
+SCENIC reconstructs regulons (i.e., transcription factors and their target genes) assesses the activity of these discovered regulons in individual cells and uses these cellular activity patterns to find meaningful clusters of cells.
+Here we present an improved version of SCENIC with several advances.
+SCENIC has been refactored and reimplemented in Python (pySCENIC), resulting in a tenfold increase in speed, and has been packaged into containers for ease of use.
+It is now also possible to use epigenomic track databases, as well as motifs, to refine regulons.
+In this protocol, we explain the different steps of SCENIC: the workflow starts from the count matrix depicting the gene abundances for all cells and consists of three stages.
+First, coexpression modules are inferred using a regression per-target approach (GRNBoost2).
+Next, the indirect targets are pruned from these modules using cis-regulatory motif discovery (cisTarget).
+Lastly, the activity of these regulons is quantified via an enrichment score for the regulon's target genes (AUCell).
+Nonlinear projection methods can be used to display visual groupings of cells based on the cellular activity patterns of these regulons.
+The results can be exported as a loom file and visualized in the SCope web application.
+This protocol is illustrated on two use cases: a peripheral blood mononuclear cell data set and a panel of single-cell RNA-sequencing cancer experiments.
+For a data set of 10,000 genes and 50,000 cells, the pipeline runs in {$<$}2 h.},
+  journal = {Nature Protocols},
+  number = {7},
+}
+
+@Article{vieth_powsimrpoweranalysis_2017,
+  title = {{{powsimR}}: Power Analysis for Bulk and Single Cell {{RNA}}-Seq Experiments},
+  shorttitle = {{{powsimR}}},
+  author = {Beate Vieth and Christoph Ziegenhain and Swati Parekh and Wolfgang Enard and Ines Hellmann},
+  year = {2017},
+  month = {nov},
+  volume = {33},
+  pages = {3486--3488},
+  issn = {1367-4803},
+  doi = {10.1093/bioinformatics/btx435},
+  abstract = {AbstractSummary.  Power analysis is essential to optimize the design of RNA-seq experiments and to assess and compare the power to detect differentially express},
+  file = {/home/rcannood/Zotero/storage/MFRF86MZ/Vieth et al. - 2017 - powsimR power analysis for bulk and single cell R.pdf;/home/rcannood/Zotero/storage/HSCUJRI8/3952669.html},
+  journal = {Bioinformatics},
+  language = {en},
+  number = {21},
+}
+
+@Article{wang_quantifyingwaddingtonlandscape_2011,
+  title = {Quantifying the {{Waddington Landscape}} and {{Biological Paths}} for {{Development}} and {{Differentiation}}},
+  author = {Jin Wang and Kun Zhang and Li Xu and Erkang Wang},
+  year = {2011},
+  month = {may},
+  volume = {108},
+  pages = {8257--8262},
+  issn = {0027-8424, 1091-6490},
+  doi = {10.1073/pnas.1017017108},
+  abstract = {We developed a theoretical framework to prove the existence and quantify the Waddington landscape as well as chreode-biological paths for development and differentiation.
+The cells can have states with the higher probability ones giving the different cell types.
+Different cell types correspond to different basins of attractions of the probability landscape.
+We study how the cells develop from undifferentiated cells to differentiated cells from landscape perspectives.
+We quantified the Waddington landscape through construction of underlying probability landscape for cell development.
+We show the developmental process proceeds as moving from undifferentiated to the differentiated basins of attractions.
+The barrier height of the basins of attractions correlates with the escape time that determines the stability of cell types.
+We show that the developmental process can be quantitatively described and uncovered by the biological paths on the quantified Waddington landscape from undifferentiated to the differentiated cells.
+We found the dynamics of the developmental process is controlled by a combination of the gradient and curl force on the landscape.
+The biological paths often do not follow the steepest descent path on the landscape.
+The landscape framework also quantifies the possibility of reverse differentiation process such as cell reprogramming from differentiated cells back to the original stem cell.
+We show that the biological path of reverse differentiation is irreversible and different from the one for differentiation process.
+We found that the developmental process described by the underlying landscape and the associated biological paths is relatively stable and robust against the influences of environmental perturbations.},
+  journal = {Proceedings of the National Academy of Sciences},
+  number = {20},
+  pmid = {21536909},
+}
+
+@Article{weber_essentialguidelinescomputational_2019,
+  title = {Essential Guidelines for Computational Method Benchmarking},
+  author = {Lukas M. Weber and Wouter Saelens and Robrecht Cannoodt and Charlotte Soneson and Alexander Hapfelmeier and Paul P. Gardner and Anne-Laure Boulesteix and Yvan Saeys and Mark D. Robinson},
+  year = {2019},
+  month = {jun},
+  volume = {20},
+  pages = {125},
+  issn = {1474-760X},
+  doi = {10.1186/s13059-019-1738-8},
+  abstract = {In computational biology and other sciences, researchers are frequently faced with a choice between several computational methods for performing data analyses.
+Benchmarking studies aim to rigorously compare the performance of different methods using well-characterized benchmark datasets, to determine the strengths of each method or to provide recommendations regarding suitable choices of methods for an analysis.
+However, benchmarking studies must be carefully designed and implemented to provide accurate, unbiased, and informative results.
+Here, we summarize key practical guidelines and recommendations for performing high-quality benchmarking analyses, based on our experiences in computational biology.},
+  file = {/home/rcannood/Zotero/storage/83LMG2SG/Weber et al. - 2019 - Essential guidelines for computational method benc.pdf;/home/rcannood/Zotero/storage/GL7G9MIR/s13059-019-1738-8.html},
+  journal = {Genome Biology},
+  number = {1},
+}
+
+@Article{wickham_welcometidyverse_2019,
+  title = {Welcome to the {{Tidyverse}}},
+  author = {Hadley Wickham and Mara Averick and Jennifer Bryan and Winston Chang and Lucy McGowan and Romain Fran{\c c}ois and Garrett Grolemund and Alex Hayes and Lionel Henry and Jim Hester and Max Kuhn and Thomas Pedersen and Evan Miller and Stephan Bache and Kirill M{\"u}ller and Jeroen Ooms and David Robinson and Dana Seidel and Vitalie Spinu and Kohske Takahashi and Davis Vaughan and Claus Wilke and Kara Woo and Hiroaki Yutani},
+  year = {2019},
+  month = {nov},
+  doi = {10.21105/joss.01686},
+  abstract = {Wickham et al., (2019). Welcome to the Tidyverse. Journal of Open Source Software, 4(43), 1686, https://doi.org/10.21105/joss.01686},
+  file = {/home/rcannood/Zotero/storage/TVYSNC4J/Wickham et al. - 2019 - Welcome to the Tidyverse.pdf;/home/rcannood/Zotero/storage/EAQ2VYF8/joss.html},
+  language = {en},
+}
+
+@Article{xu_regulationbifurcatingcell_2015,
+  title = {Regulation of Bifurcating {{B}} Cell Trajectories by Mutual Antagonism between Transcription Factors {{IRF4}} and {{IRF8}}},
+  author = {Heping Xu and Virendra K Chaudhri and Zhiguo Wu and Konstantinos Biliouris and Krista Dienger-Stambaugh and Yrina Rochman and Harinder Singh},
+  year = {2015},
+  month = {dec},
+  volume = {16},
+  pages = {1274--1281},
+  abstract = {Upon recognition of antigen, B cells undertake a bifurcated response in which some cells rapidly differentiate into plasmablasts while others undergo affinity maturation in germinal centers (GCs).
+Here we identified a double-negative feedback loop between the transcription factors IRF4 and IRF8 that regulated the initial developmental bifurcation of activated B cells as well as the GC response.
+IRF8 dampened signaling via the B cell antigen receptor (BCR), facilitated antigen-specific interaction with helper T cells, and promoted antibody affinity maturation while antagonizing IRF4-driven differentiation of plasmablasts.
+Genomic analysis revealed concentration-dependent actions of IRF4 and IRF8 in regulating distinct gene-expression programs.
+Stochastic modeling suggested that the double-negative feedback was sufficient to initiate bifurcation of the B cell developmental trajectories.},
+  journal = {Nat. Immunol.},
+  number = {12},
+}
+
+@Article{yosef_dynamicregulatorynetwork_2013,
+  title = {Dynamic Regulatory Network Controlling \{\vphantom\}{{TH17}}\vphantom\{\} Cell Differentiation},
+  author = {Nir Yosef and Alex K Shalek and Jellert T Gaublomme and Hulin Jin and Youjin Lee and Amit Awasthi and Chuan Wu and Katarzyna Karwacz and Sheng Xiao and Marsela Jorgolli and David Gennert and Rahul Satija and Arvind Shakya and Diana Y Lu and John J Trombetta and Meenu R Pillai and Peter J Ratcliffe and Mathew L Coleman and Mark Bix and Dean Tantin and Hongkun Park and Vijay K Kuchroo and Aviv Regev},
+  year = {2013},
+  volume = {496},
+  pages = {461--468},
+  abstract = {Despite their importance, the molecular circuits that control the differentiation of naive T cells remain largely unknown.
+Recent studies that reconstructed regulatory networks in mammalian cells have focused on short-term responses and relied on perturbation-based approaches that cannot be readily applied to primary T cells.
+Here we combine transcriptional profiling at high temporal resolution, novel computational algorithms, and innovative nanowire-based perturbation tools to systematically derive and experimentally validate a model of the dynamic regulatory network that controls the differentiation of mouse TH17 cells, a proinflammatory T-cell subset that has been implicated in the pathogenesis of multiple autoimmune diseases.
+The TH17 transcriptional network consists of two self-reinforcing, but mutually antagonistic, modules, with 12 novel regulators, the coupled action of which may be essential for maintaining the balance between TH17 and other CD4(+) T-cell subsets.
+Our study identifies and validates 39 regulatory factors, embeds them within a comprehensive temporal network and reveals its organizational principles; it also highlights novel drug targets for controlling TH17 cell differentiation.},
+  journal = {Nature},
+  number = {7446},
+}
+
+@Article{zappia_splattersimulationsinglecell_2017,
+  title = {Splatter: {{Simulation}} of {{Single}}-{{Cell RNA Sequencing Data}}},
+  shorttitle = {Splatter},
+  author = {Luke Zappia and Belinda Phipson and Alicia Oshlack},
+  year = {2017},
+  month = {sep},
+  volume = {18},
+  pages = {174},
+  issn = {1474-760X},
+  doi = {10.1186/s13059-017-1305-0},
+  abstract = {As single-cell RNA sequencing (scRNA-seq) technologies have rapidly developed, so have analysis methods.
+Many methods have been tested, developed, and validated using simulated datasets.
+Unfortunately, current simulations are often poorly documented, their similarity to real data is not demonstrated, or reproducible code is not available.
+Here, we present the Splatter Bioconductor package for simple, reproducible, and well-documented simulation of scRNA-seq data.
+Splatter provides an interface to multiple simulation methods including Splat, our own simulation, based on a gamma-Poisson distribution.
+Splat can simulate single populations of cells, populations with multiple cell types, or differentiation paths.},
+  journal = {Genome Biology},
+  keywords = {RNA-seq,Simulation,Single-cell,Software},
+}
+
+@Article{zeisel_coupledpremrnamrna_2011,
+  title = {Coupled Pre-{{mRNA}} and {{mRNA}} Dynamics Unveil Operational Strategies Underlying Transcriptional Responses to Stimuli},
+  author = {Amit Zeisel and Wolfgang J K{\"o}stler and Natali Molotski and Jonathan M Tsai and Rita Krauthgamer and Jasmine Jacob-Hirsch and Gideon Rechavi and Yoav Soen and Steffen Jung and Yosef Yarden and Eytan Domany},
+  year = {2011},
+  month = {jan},
+  volume = {7},
+  pages = {529},
+  issn = {1744-4292},
+  doi = {10.1038/msb.2011.62},
+  abstract = {Transcriptional responses to extracellular stimuli involve tuning the rates of transcript production and degradation.
+Here, we show that the time-dependent profiles of these rates can be inferred from simultaneous measurements of precursor mRNA (pre-mRNA) and mature mRNA profiles.
+Transcriptome-wide measurements demonstrate that genes with similar mRNA profiles often exhibit marked differences in the amplitude and onset of their production rate.
+The latter is characterized by a large dynamic range, with a group of genes exhibiting an unexpectedly strong transient production overshoot, thereby accelerating their induction and, when combined with time-dependent degradation, shaping transient responses with precise timing and amplitude.},
+  file = {/home/rcannood/Zotero/storage/QRQW86Y9/Zeisel et al. - 2011 - Coupled pre-mRNA and mRNA dynamics unveil operatio.pdf;/home/rcannood/Zotero/storage/R8ZG26IL/msb.2011.html},
+  journal = {Molecular Systems Biology},
+  keywords = {half-life,operational strategy,pre-mRNA,transcriptional response},
+  number = {1},
+}
+
+@Article{zhang_simulatingmultiplefaceted_2019,
+  title = {Simulating Multiple Faceted Variability in Single Cell {{RNA}} Sequencing},
+  author = {Xiuwei Zhang and Chenling Xu and Nir Yosef},
+  year = {2019},
+  month = {jun},
+  volume = {10},
+  pages = {1--16},
+  issn = {2041-1723},
+  doi = {10.1038/s41467-019-10500-w},
+  abstract = {Simulated single cell RNA sequencing data is useful for method development and comparison.
+Here, the authors developed SymSim, a simulator that explicitly models the main factors of variation in single cell data.},
+  copyright = {2019 The Author(s)},
+  file = {/home/rcannood/Zotero/storage/YLQYCW6E/Zhang et al. - 2019 - Simulating multiple faceted variability in single .pdf;/home/rcannood/Zotero/storage/EXJ96Y59/s41467-019-10500-w.html},
+  journal = {Nature Communications},
+  language = {en},
+  number = {1},
+}
+\end{filecontents}
 
@@
 \author{Robrecht Cannoodt\textsuperscript{$\dagger{}$,1,2,3} \and Wouter
-Saelens\textsuperscript{$\dagger{}$,1,2} \and Louise
+Saelens\textsuperscript{$\dagger{}$,1,2,4} \and Louise
 Deconinck\textsuperscript{1,2} \and Yvan Saeys\textsuperscript{1,2,*}}
@@
 and Statistics, Ghent University, Ghent, Belgium\\
-\textsuperscript{3} Data Intuitive, Lebbeke, Belgium
+\textsuperscript{3} Data Intuitive, Lebbeke, Belgium\\
+\textsuperscript{4} Institute of Bioengineering, School of Life
+Sciences, École Polytechnique Fédérale de Lausanne (EPFL), Lausanne,
+Switzerland
 
@@
 sample the interactions between target genes, one of the many FANTOM5
-\autocite{lizio_gatewaysfantom5promoter_2015} GRNs is sampled. The currently
-existing TFs are mapped to regulators in the FANTOM5 GRN. The targets
-are drawn from the FANTOM5 GRN weighted by their page rank value, to
-create an induced GRN. For each target, at most \(x\) regulators are
-sampled from the induced FANTOM5 GRN (default \(x=5\)). The interactions
-connecting a target gene and its regulators are added to the GRN.
+\autocite{lizio_gatewaysfantom5promoter_2015} GRNs is sampled. The
+currently existing TFs are mapped to regulators in the FANTOM5 GRN. The
+targets are drawn from the FANTOM5 GRN weighted by their page rank
+value, to create an induced GRN. For each target, at most \(x\)
+regulators are sampled from the induced FANTOM5 GRN (default \(x=5\)).
+The interactions connecting a target gene and its regulators are added
+to the GRN.
 
@@
 
-\begin{filecontents}{\jobname.bib}
-@Article{aibar_scenicsinglecellregulatory_2017,
-	title = {{{SCENIC}}: Single-Cell Regulatory Network Inference and Clustering},
-	author = {Sara Aibar and Carmen Bravo Gonz{\a'a}lez-Blas and Thomas Moerman and V{\^a}n Anh Huynh-Thu and Hana Imrichova and Gert Hulselmans and Florian Rambow and Jean-Christophe Marine and Pierre Geurts and Jan Aerts and Joost {van den Oord} and Zeynep Kalender Atak and Jasper Wouters and Stein Aerts},
-	year = {2017},
-	month = {oct},
-	issn = {1548-7091},
-	doi = {10.1038/nmeth.4463},
-	journal = {Nature Methods},
-}
-
-@Article{alpert_alignmentsinglecelltrajectories_2018,
-	title = {Alignment of Single-Cell Trajectories to Compare Cellular Expression Dynamics},
-	author = {Ayelet Alpert and Lindsay S. Moore and Tania Dubovik and Shai S. Shen-Orr},
-	year = {2018},
-	month = {apr},
-	volume = {15},
-	pages = {267--270},
-	issn = {1548-7105},
-	doi = {10.1038/nmeth.4628},
-	abstract = {cellAlign enables quantitative comparisons of expression dynamics within and between single-cell trajectories based on single-cell RNA-seq or mass cytometry data.},
-	copyright = {2018 Nature Publishing Group, a division of Macmillan Publishers Limited. All Rights Reserved.},
-	file = {/home/rcannood/Zotero/storage/QB7FH2T9/nmeth.html},
-	journal = {Nature Methods},
-	language = {en},
-	number = {4},
-}
-
-@Article{bergen_generalizingrnavelocity_2020,
-	title = {Generalizing {{RNA}} Velocity to Transient Cell States through Dynamical Modeling},
-	author = {Volker Bergen and Marius Lange and Stefan Peidli and F. Alexander Wolf and Fabian J. Theis},
-	year = {2020},
-	month = {dec},
-	volume = {38},
-	pages = {1408--1414},
-	issn = {1546-1696},
-	doi = {10.1038/s41587-020-0591-3},
-	abstract = {RNA velocity has opened up new ways of studying cellular differentiation in single-cell RNA-sequencing data.
-It describes the rate of gene expression change for an individual gene at a given time point based on the ratio of its spliced and unspliced messenger RNA (mRNA).
-However, errors in velocity estimates arise if the central assumptions of a common splicing rate and the observation of the full splicing dynamics with steady-state mRNA levels are violated.
-Here we present scVelo, a method that overcomes these limitations by solving the full transcriptional dynamics of splicing kinetics using a likelihood-based dynamical model.
-This generalizes RNA velocity to systems with transient cell states, which are common in development and in response to perturbations.
-We apply scVelo to disentangling subpopulation kinetics in neurogenesis and pancreatic endocrinogenesis.
-We infer gene-specific rates of transcription, splicing and degradation, recover each cell's position in the underlying differentiation processes and detect putative driver genes.
-scVelo will facilitate the study of lineage decisions and gene regulation.},
-	journal = {Nature Biotechnology},
-	number = {12},
-}
-
-@Article{bjornsson_digitaltwinspersonalize_2019,
-	title = {Digital Twins to Personalize Medicine},
-	author = {Bergthor Bj{\"o}rnsson and Carl Borrebaeck and Nils Elander and Thomas Gasslander and Danuta R.
-Gawel and Mika Gustafsson and Rebecka J{\"o}rnsten and Eun Jung Lee and Xinxiu Li and Sandra Lilja and David Mart\'inez-Enguita and Andreas Matussek and Per Sandstr{\"o}m and Samuel Sch{\"a}fer and Margaretha Stenmarker and X.
-F.
-Sun and Oleg Sysoev and Huan Zhang and Mikael Benson and {on behalf of the Swedish Digital Twin Consortium}},
-	year = {2019},
-	month = {dec},
-	volume = {12},
-	pages = {4},
-	issn = {1756-994X},
-	doi = {10.1186/s13073-019-0701-3},
-	abstract = {Personalized medicine requires the integration and processing of vast amounts of data.
-Here, we propose a solution to this challenge that is based on constructing Digital Twins.
-These are high-resolution models of individual patients that are computationally treated with thousands of drugs to find the drug that is optimal for the patient.},
-	journal = {Genome Medicine},
-	number = {1},
-}
-
-@Article{cacchiarelli_aligningsinglecelldevelopmental_2018,
-	title = {Aligning {{Single}}-{{Cell Developmental}} and {{Reprogramming Trajectories Identifies Molecular Determinants}} of {{Myogenic Reprogramming Outcome}}},
-	author = {Davide Cacchiarelli and Xiaojie Qiu and Sanjay Srivatsan and Anna Manfredi and Michael Ziller and Eliah Overbey and Antonio Grimaldi and Jonna Grimsby and Prapti Pokharel and Kenneth J.
-Livak and Shuqiang Li and Alexander Meissner and Tarjei S.
-Mikkelsen and John L.
-Rinn and Cole Trapnell},
-	year = {2018},
-	month = {sep},
-	volume = {7},
-	pages = {258-268.e3},
-	issn = {2405-4712},
-	doi = {10.1016/j.cels.2018.07.006},
-	abstract = {{$<$}h2{$>$}Summary{$<$}/h2{$><$}p{$>$}Cellular reprogramming through manipulation of defined factors holds great promise for large-scale production of cell types needed for use in therapy and for revealing principles of gene regulation.
-However, most reprogramming systems are inefficient, converting only a fraction of cells to the desired state.
-Here, we analyze MYOD-mediated reprogramming of human fibroblasts to myotubes, a well-characterized model system for direct conversion by defined factors, at pseudotemporal resolution using single-cell RNA-seq.
-To expose barriers to efficient conversion, we introduce a novel analytic technique, trajectory alignment, which enables quantitative comparison of gene expression kinetics across two biological processes.
-Reprogrammed cells navigate a trajectory with branch points that correspond to two alternative decision points, with cells that select incorrect branches terminating at aberrant or incomplete reprogramming outcomes.
-Analysis of these branch points revealed insulin and BMP signaling as crucial molecular determinants of reprogramming.
-Single-cell trajectory alignment enables rigorous quantitative comparisons between biological trajectories found in diverse processes in development, reprogramming, and other contexts.{$<$}/p{$>$}},
-	file = {/home/rcannood/Zotero/storage/YQH8GP2I/Cacchiarelli et al. - 2018 - Aligning Single-Cell Developmental and Reprogrammi.pdf;/home/rcannood/Zotero/storage/Q5MUK95E/S2405-4712(18)30314-4.html},
-	journal = {Cell Systems},
-	language = {English},
-	number = {3},
-	pmid = {30195438},
-}
-
-@Article{cannoodt_singlecellomicsdatasets_2018,
-	title = {Single-Cell -Omics Datasets Containing a Trajectory},
-	author = {Robrecht Cannoodt and Wouter Saelens and Helena Todorov and Yvan Saeys},
-	year = {2018},
-	month = {oct},
-	doi = {10.5281/zenodo.1211532},
-	journal = {Zenodo},
-}
-
-@TechReport{conover_multiplecomparisonsprocedures_1979,
-	title = {On Multiple-Comparisons Procedures},
-	author = {William Jay Conover and Ronald L Iman},
-	year = {1979},
-	institution = {{Technical report, Los Alamos Scientific Laboratory}},
-}
-
-@Article{duo_systematicperformanceevaluation_2018,
-	title = {A Systematic Performance Evaluation of Clustering Methods for Single-Cell {{RNA}}-Seq Data},
-	author = {Angelo Du{\a`o} and Mark D. Robinson and Charlotte Soneson},
-	year = {2018},
-	volume = {7},
-	pages = {1141},
-	issn = {2046-1402},
-	doi = {10.12688/f1000research.15666.2},
-	abstract = {Subpopulation identification, usually via some form of unsupervised clustering, is a fundamental step in the analysis of many single-cell RNA-seq data sets.
-This has motivated the development and application of a broad range of clustering methods, based on various underlying algorithms.
-Here, we provide a systematic and extensible performance evaluation of 14 clustering algorithms implemented in R, including both methods developed explicitly for scRNA-seq data and more general-purpose methods.
-The methods were evaluated using nine publicly available scRNA-seq data sets as well as three simulations with varying degree of cluster separability.
-The same feature selection approaches were used for all methods, allowing us to focus on the investigation of the performance of the clustering algorithms themselves.
-We evaluated the ability of recovering known subpopulations, the stability and the run time and scalability of the methods.
-Additionally, we investigated whether the performance could be improved by generating consensus partitions from multiple individual clustering methods.
-We found substantial differences in the performance, run time and stability between the methods, with SC3 and Seurat showing the most favorable results.
-Additionally, we found that consensus clustering typically did not improve the performance compared to the best of the combined methods, but that several of the top-performing methods already perform some type of consensus clustering.
-All the code used for the evaluation is available on GitHub ( https://github.com/markrobinsonuzh/scRNAseq\_clustering\_comparison).
-In addition, an R package providing access to data and clustering results, thereby facilitating inclusion of new methods and data sets, is available from Bioconductor ( https://bioconductor.org/packages/DuoClustering2018).},
-	file = {/home/rcannood/Zotero/storage/ASWVFH87/Duò et al. - 2018 - A systematic performance evaluation of clustering .pdf},
-	journal = {F1000Research},
-	keywords = {Algorithms,Benchmarking,Cluster Analysis,Clustering,Clustering methods,Gene Expression Profiling,Humans,RNA-seq,RNA; Small Cytoplasmic,Sequence Analysis; RNA,Single-Cell Analysis,Single-cell RNA-seq,Software},
-	language = {eng},
-	pmcid = {PMC6134335},
-	pmid = {30271584},
-}
-
-@Article{ewing_combiningtumorgenome_2015,
-	title = {Combining Tumor Genome Simulation with Crowdsourcing to Benchmark Somatic Single-Nucleotide-Variant Detection},
-	author = {Adam D.
-Ewing and Kathleen E.
-Houlahan and Yin Hu and Kyle Ellrott and Cristian Caloian and Takafumi N.
-Yamaguchi and J.
-Christopher Bare and Christine P'ng and Daryl Waggott and Veronica Y.
-Sabelnykova and {ICGC-TCGA DREAM Somatic Mutation Calling Challenge Participants} and Liu Xi and Ninad Dewal and Yu Fan and Wenyi Wang and David Wheeler and Andreas Wilm and Grace Hui Ting and Chenhao Li and Denis Bertrand and Niranjan Nagarajan and Qing-Rong Chen and Chih-Hao Hsu and Ying Hu and Chunhua Yan and Warren Kibbe and Daoud Meerzaman and Kristian Cibulskis and Mara Rosenberg and Louis Bergelson and Adam Kiezun and Amie Radenbaugh and Anne-Sophie Sertier and Anthony Ferrari and Laurie Tonton and Kunal Bhutani and Nancy F.
-Hansen and Difei Wang and Lei Song and Zhongwu Lai and Yang Liao and Wei Shi and Jos{\a'e} Carbonell-Caballero and Joaqu\'in Dopazo and Cheryl C.
-K.
-Lau and Justin Guinney and Michael R.
-Kellen and Thea C.
-Norman and David Haussler and Stephen H.
-Friend and Gustavo Stolovitzky and Adam A.
-Margolin and Joshua M.
-Stuart and Paul C.
-Boutros},
-	year = {2015},
-	month = {jul},
-	volume = {12},
-	pages = {623--630},
-	issn = {1548-7105},
-	doi = {10.1038/nmeth.3407},
-	abstract = {The detection of somatic mutations from cancer genome sequences is key to understanding the genetic basis of disease progression, patient survival and response to therapy.
-Benchmarking is needed for tool assessment and improvement but is complicated by a lack of gold standards, by extensive resource requirements and by difficulties in sharing personal genomic information.
-To resolve these issues, we launched the ICGC-TCGA DREAM Somatic Mutation Calling Challenge, a crowdsourced benchmark of somatic mutation detection algorithms.
-Here we report the BAMSurgeon tool for simulating cancer genomes and the results of 248 analyses of three in silico tumors created with it.
-Different algorithms exhibit characteristic error profiles, and, intriguingly, false positives show a trinucleotide profile very similar to one found in human tumors.
-Although the three simulated tumors differ in sequence contamination (deviation from normal cell sequence) and in subclonality, an ensemble of pipelines outperforms the best individual pipeline in all cases.
-BAMSurgeon is available at https://github.com/adamewing/bamsurgeon/.},
-	copyright = {2015 Nature Publishing Group},
-	file = {/home/rcannood/Zotero/storage/NBPKWHK3/Ewing et al. - 2015 - Combining tumor genome simulation with crowdsourci.pdf;/home/rcannood/Zotero/storage/TDZU2TL9/nmeth.html},
-	journal = {Nature Methods},
-	language = {en},
-	number = {7},
-}
-
-@Article{ferrell_bistabilitybifurcationswaddington_2012,
-	title = {Bistability, {{Bifurcations}}, and {{Waddington}}'s {{Epigenetic Landscape}}},
-	author = {James E Ferrell},
-	year = {2012},
-	month = {jun},
-	volume = {22},
-	pages = {R458--R466},
-	issn = {0960-9822},
-	doi = {10.1016/j.cub.2012.03.045},
-	abstract = {Waddington's epigenetic landscape is probably the most famous and most powerful metaphor in developmental biology.
-Cells, represented by balls, roll downhill through a landscape of bifurcating valleys.
-Each new valley represents a possible cell fate and the ridges between the valleys maintain the cell fate once it has been chosen.
-Here I examine models of two important developmental processes \$\textbackslash backslash\$textemdash\{\} cell-fate induction and lateral inhibition \$\textbackslash backslash\$textemdash\{\} and ask whether the landscapes for these models at least qualitatively resemble Waddington's picture.
-For cell-fate induction, the answer is no.
-The commitment of a cell to a new fate corresponds to the disappearance of a valley from the landscape, not the splitting of one valley into two, and it occurs through a type of bifurcation \$\textbackslash backslash\$textemdash\{\} a saddle-node bifurcation \$\textbackslash backslash\$textemdash\{\} that possesses an intrinsic irreversibility that is missing from Waddington's picture.
-Lateral inhibition, a symmetrical cell\$\textbackslash backslash\$textendash\{\}cell competition process, corresponds better to Waddington's picture, with one valley reversibly splitting into two through a pitchfork bifurcation.
-I propose an alternative epigenetic landscape that has numerous valleys and ridges right from the start, with the process of cell-fate commitment corresponding to the irreversible disappearance of some of these valleys and ridges, via cell-fate induction, complemented by the creation of new valleys and ridges through processes like cell\$\textbackslash backslash\$textendash\{\}cell competition.},
-	file = {/home/rcannood/Zotero/storage/GK4TJ5BV/Ferrell - 2012 - Bistability, Bifurcations, and Waddington's Epigen.pdf;/home/rcannood/Zotero/storage/HUSMJA22/Ferrell - 2012 - Bistability, Bifurcations, and Waddington's Epigen.pdf;/home/rcannood/Zotero/storage/8VLPBTG2/S0960-9822(12)00332-6.html;/home/rcannood/Zotero/storage/XBTP42Y6/S0960982212003326.html},
-	journal = {Current Biology},
-	number = {11},
-}
-
-@Article{gibson_probabilisticmodelprokaryotic_2000,
-	title = {A Probabilistic Model of a Prokaryotic Gene and Its Regulation},
-	author = {Michael A. Gibson and Jehoshua Bruck},
-	year = {2000},
-	journal = {Computational Methods in Molecular Biology: From Genotype to Phenotype, MIT press, Cambridge},
-}
-
-@Article{gillespie_exactstochasticsimulation_1977,
-	title = {Exact Stochastic Simulation of Coupled Chemical Reactions},
-	author = {Daniel T. Gillespie},
-	year = {1977},
-	month = {dec},
-	volume = {81},
-	pages = {2340--2361},
-	issn = {0022-3654},
-	doi = {10.1021/j100540a008},
-	file = {/home/rcannood/Zotero/storage/ELV9AQRF/j100540a008.html},
-	journal = {The Journal of Physical Chemistry},
-	number = {25},
-}
-
-@Article{giorgino_computingvisualizingdynamic_2009,
-	title = {Computing and {{Visualizing Dynamic Time Warping Alignments}} in {{R}}: {{The}} Dtw {{Package}}},
-	author = {Toni Giorgino},
-	year = {2009},
-	month = {sep},
-	doi = {10.18637/jss.v031.i07},
-	abstract = {Dynamic time warping is a popular technique for comparing time series, providing both a distance measure that is insensitive to local compression and stretches and the warping which optimally deforms one of the two input series onto the other.
-A variety of algorithms and constraints have been discussed in the literature.
-The dtw package provides an unification of them; it allows R users to compute time series alignments mixing freely a variety of continuity constraints, restriction windows, endpoints, local distance definitions, and so on.
-The package also provides functions for visualizing alignments and constraints using several classic diagram types.},
-	journal = {Journal of Statistical Software},
-	number = {7},
-}
-
-@Article{glass_passingmessagesbiological_2013,
-	title = {Passing {{Messages}} between {{Biological Networks}} to {{Refine Predicted Interactions}}},
-	author = {Kimberly Glass and Curtis Huttenhower and John Quackenbush and Guo-Cheng Yuan},
-	year = {2013},
-	month = {may},
-	volume = {8},
-	pages = {e64832},
-	doi = {10.1371/journal.pone.0064832},
-	abstract = {Regulatory network reconstruction is a fundamental problem in computational biology.
-There are significant limitations to such reconstruction using individual datasets, and increasingly people attempt to construct networks using multiple, independent datasets obtained from complementary sources, but methods for this integration are lacking.
-We developed PANDA (Passing Attributes between Networks for Data Assimilation), a message-passing model using multiple sources of information to predict regulatory relationships, and used it to integrate protein-protein interaction, gene expression, and sequence motif data to reconstruct genome-wide, condition-specific regulatory networks in yeast as a model.
-The resulting networks were not only more accurate than those produced using individual data sets and other existing methods, but they also captured information regarding specific biological mechanisms and pathways that were missed using other methodologies.
-PANDA is scalable to higher eukaryotes, applicable to specific tissue or cell type data and conceptually generalizable to include a variety of regulatory, interaction, expression, and other genome-scale data.
-An implementation of the PANDA algorithm is available at www.sourceforge.net/projects/panda-net.},
-	journal = {PLOS ONE},
-	number = {5},
-}
-
-@Article{graf_forcingcellschange_2009,
-	title = {Forcing {{Cells}} to {{Change Lineages}}},
-	author = {Thomas Graf and Tariq Enver},
-	year = {2009},
-	month = {dec},
-	volume = {462},
-	pages = {587},
-	issn = {1476-4687},
-	doi = {10.1038/nature08533},
-	abstract = {\$\textbackslash textless\$p\$\textbackslash textgreater\$Though it is important for the success of multicellular organisms that their many different cell types remain stable once specified, advances in the production of induced pluripotent stem (iPS) cells from somatic cells by exposure to a transcription factor cocktail emphasize the potential of almost any cell to be reprogrammed.
-This raises the question of whether interconversion between cell types occurs physiologically or in the context of disease.
-Thomas Graf and Tariq Enver use examples from the blood cell system to illustrate the principles of transcription factor transdifferentiation, which they conclude is a process fundamentally similar to differentiation and different from iPS reprogramming.
-They postulate that conversions of one cell type to another can occur as a normal developmental process, and speculate on the prospects of forced lineage reprogramming in regenerative medicine.\$\textbackslash textless\$/p\$\textbackslash textgreater\$},
-	journal = {Nature},
-	number = {7273},
-}
-
-@Article{hache_gengesystematicgeneration_2009,
-	title = {{{GeNGe}}: Systematic Generation of Gene Regulatory Networks},
-	shorttitle = {{{GeNGe}}},
-	author = {Hendrik Hache and Christoph Wierling and Hans Lehrach and Ralf Herwig},
-	year = {2009},
-	month = {may},
-	volume = {25},
-	pages = {1205--1207},
-	issn = {1367-4803},
-	doi = {10.1093/bioinformatics/btp115},
-	abstract = {Summary: The analysis of gene regulatory networks (GRNs) is a central goal of bioinformatics highly accelerated by the advent of new experimental techniques, such as RNA interference.
-A battery of reverse engineering methods has been developed in recent years to reconstruct the underlying GRNs from these and other experimental data.
-However, the performance of the individual methods is poorly understood and validation of algorithmic performances is still missing to a large extent.
-To enable such systematic validation, we have developed the web application GeNGe (GEne Network GEnerator), a controlled framework for the automatic generation of GRNs.
-The theoretical model for a GRN is a non-linear differential equation system.
-Networks can be user-defined or constructed in a modular way with the option to introduce global and local network perturbations.
-Resulting data can be used, e.g.
-as benchmark data for evaluating GRN reconstruction methods or for predicting effects of perturbations as theoretical counterparts of biological experiments.Availability: Available online at http://genge.molgen.mpg.deContact:hache@molgen.mpg.deSupplementary information:Supplementary data are available at Bioinformatics online.},
-	journal = {Bioinformatics},
-	number = {9},
-}
-
-@Article{holm_simplesequentiallyrejective_1979,
-	title = {A Simple Sequentially Rejective Multiple Test Procedure},
-	author = {Sture Holm},
-	year = {1979},
-	pages = {65--70},
-	journal = {Scandinavian journal of statistics},
-}
-
-@Article{huynh-thu_inferringregulatorynetworks_2010,
-	title = {Inferring {{Regulatory Networks}} from {{Expression Data Using Tree}}-{{Based Methods}}},
-	author = {V{\^a}n Anh Huynh-Thu and Irrthum Alexandre and Wehenkel Louis and Geurts Pierre and Alexandre Irrthum and Louis Wehenkel and Pierre Geurts},
-	year = {2010},
-	month = {jan},
-	volume = {5},
-	pages = {e12776},
-	issn = {1932-6203},
-	doi = {10.1371/journal.pone.0012776},
-	abstract = {One of the pressing open problems of computational systems biology is the elucidation of the topology of genetic regulatory networks (GRNs) using high throughput genomic data, in particular microarray gene expression data.
-The Dialogue for Reverse Engineering Assessments and Methods (DREAM) challenge aims to evaluate the success of GRN inference algorithms on benchmarks of simulated data.
-In this article, we present GENIE3, a new algorithm for the inference of GRNs that was best performer in the DREAM4 In Silico Multifactorial challenge.
-GENIE3 decomposes the prediction of a regulatory network between p genes into p different regression problems.
-In each of the regression problems, the expression pattern of one of the genes (target gene) is predicted from the expression patterns of all the other genes (input genes), using tree-based ensemble methods Random Forests or Extra-Trees.
-The importance of an input gene in the prediction of the target gene expression pattern is taken as an indication of a putative regulatory link.
-Putative regulatory links are then aggregated over all genes to provide a ranking of interactions from which the whole network is reconstructed.
-In addition to performing well on the DREAM4 In Silico Multifactorial challenge simulated data, we show that GENIE3 compares favorably with existing algorithms to decipher the genetic regulatory network of Escherichia coli.
-It doesn't make any assumption about the nature of gene regulation, can deal with combinatorial and non-linear interactions, produces directed GRNs, and is fast and scalable.
-In conclusion, we propose a new algorithm for GRN inference that performs well on both synthetic and real gene expression data.
-The algorithm, based on feature selection with tree-based ensemble methods, is simple and generic, making it adaptable to other types of genomic data and interactions.},
-	file = {/home/rcannood/Zotero/storage/3I6PLCQL/Huynh-Thu et al. - 2010 - Inferring Regulatory Networks from Expression Data Using Tree-Based Methods.pdf},
-	journal = {PLoS ONE},
-	keywords = {Algorithms,Computational Biology,Computational Biology: methods,Escherichia coli,Escherichia coli: genetics,Gene Expression Regulation,Gene Regulatory Networks,Oligonucleotide Array Sequence Analysis},
-	number = {9},
-	pmid = {20927193},
-}
-
-@Article{kanton_organoidsinglecellgenomic_2019,
-	title = {Organoid Single-Cell Genomic Atlas Uncovers Human-Specific Features of Brain Development},
-	author = {Sabina Kanton and Michael James Boyle and Zhisong He and Malgorzata Santel and Anne Weigert and F{\a'a}tima Sanch\'is-Calleja and Patricia Guijarro and Leila Sidow and Jonas Simon Fleck and Dingding Han and Zhengzong Qian and Michael Heide and Wieland B.
-Huttner and Philipp Khaitovich and Svante P{\"a}{{\"a}}bo and Barbara Treutlein and J.
-Gray Camp},
-	year = {2019},
-	month = {oct},
-	volume = {574},
-	pages = {418--422},
-	issn = {1476-4687},
-	doi = {10.1038/s41586-019-1654-9},
-	abstract = {Species comparisons using single-cell transcriptomics and accessible chromatin profiling in stem cell-derived cerebral organoids are used to map dynamic gene-regulatory changes that are unique to humans.},
-	copyright = {2019 The Author(s), under exclusive licence to Springer Nature Limited},
-	file = {/home/rcannood/Zotero/storage/AVPYT7HY/s41586-019-1654-9.html},
-	journal = {Nature},
-	language = {en},
-	number = {7778},
-}
-
-@Article{kuijjer_estimatingsamplespecificregulatory_2015,
-	title = {Estimating Sample-Specific Regulatory Networks},
-	author = {Marieke Lydia Kuijjer and Matthew Tung and GuoCheng Yuan and John Quackenbush and Kimberly Glass},
-	year = {2015},
-	pages = {1--19},
-	abstract = {Biological systems are driven by intricate interactions among the complex array of molecules that comprise the cell.
-Many methods have been developed to reconstruct network models that attempt to capture those interactions.
-These methods often draw on large numbers of measured expression samples to tease out subtle signals and infer connections between genes (or gene products).
-The result is an aggregate network model representing a single estimate for edge likelihoods.
-While informative, aggregate models fail to capture the heterogeneity that is often represented in a population.
-Here we propose a method to reverse engineer sample-specific networks from aggregate network models.
-We demonstrate the accuracy and applicability of our approach in several datasets, including simulated data, microarray expression data from synchronized yeast cells, and RNA-seq data collected from human subjects.
-We show that these sample-specific networks can be used to study the evolution of network topology across time and to characterize shifts in gene regulation that may not be apparent in the expression data.
-We believe the ability to generate sample-specific networks will revolutionize the field of network biology and has the potential to usher in an era of precision network medicine.},
-	file = {/home/rcannood/Zotero/storage/YECTK66M/Kuijjer et al. - 2015 - Estimating sample-specific regulatory networks.pdf},
-}
-
-@Article{kuijjer_estimatingsamplespecificregulatory_2019,
-	title = {Estimating {{Sample}}-{{Specific Regulatory Networks}}},
-	author = {Marieke Lydia Kuijjer and Matthew George Tung and GuoCheng Yuan and John Quackenbush and Kimberly Glass},
-	year = {2019},
-	month = {mar},
-	volume = {14},
-	pages = {226--240},
-	issn = {2589-0042},
-	doi = {10.1016/j.isci.2019.03.021},
-	abstract = {Biological systems are driven by intricate interactions among molecules.
-Many methods have been developed that draw on large numbers of expression samples to infer connections between genes (or their products).
-The result is an aggregate network representing a single estimate for the likelihood of each interaction, or ``edge,'' in the network.
-Although informative, aggregate models fail to capture population heterogeneity.
-Here we propose a method to reverse engineer sample-specific networks from aggregate networks.
-We demonstrate our approach in several contexts, including simulated, yeast microarray, and human lymphoblastoid cell line RNA sequencing data.
-We use these sample-specific networks to study changes in network topology across time and to characterize shifts in gene regulation that were not apparent in the expression data.
-We believe that generating sample-specific networks will greatly facilitate the application of network methods to large, complex, and heterogeneous multi-omic datasets, supporting the emerging field of precision network medicine.,                                        \textbullet{}               We developed LIONESS to extract single-sample networks from aggregate models                                         \textbullet{}               We tested LIONESS using in silico, yeast, and human expression data                                         \textbullet{}               LIONESS-estimated networks are reproducible, accurate, and biologically meaningful                                         \textbullet{}               Single-sample network analysis highlights important biological processes                                 , Biological Sciences; Bioinformatics; Complex Systems},
-	file = {/home/rcannood/Zotero/storage/ITHUQ7NH/Kuijjer et al. - 2019 - Estimating Sample-Specific Regulatory Networks.pdf},
-	journal = {iScience},
-	pmcid = {PMC6463816},
-	pmid = {30981959},
-}
-
-@Article{kuijjer_lionessrsinglesample_2019,
-	title = {{{lionessR}}: Single Sample Network Inference in {{R}}},
-	shorttitle = {{{lionessR}}},
-	author = {Marieke L. Kuijjer and Ping-Han Hsieh and John Quackenbush and Kimberly Glass},
-	year = {2019},
-	month = {oct},
-	volume = {19},
-	pages = {1003},
-	issn = {1471-2407},
-	doi = {10.1186/s12885-019-6235-7},
-	abstract = {In biomedical research, network inference algorithms are typically used to infer complex association patterns between biological entities, such as between genes or proteins, using data from a population.
-This resulting aggregate network, in essence, averages over the networks of those individuals in the population.
-LIONESS (Linear Interpolation to Obtain Network Estimates for Single Samples) is a method that can be used together with a network inference algorithm to extract networks for individual samples in a population.
-The method's key characteristic is that, by modeling networks for individual samples in a data set, it can capture network heterogeneity in a population.
-LIONESS was originally made available as a function within the PANDA (Passing Attributes between Networks for Data Assimilation) regulatory network reconstruction framework.
-However, the LIONESS algorithm is generalizable and can be used to model single sample networks based on a wide range of network inference algorithms.},
-	file = {/home/rcannood/Zotero/storage/RS448RZF/Kuijjer et al. - 2019 - lionessR single sample network inference in R.pdf;/home/rcannood/Zotero/storage/9STH6ZYN/s12885-019-6235-7.html},
-	journal = {BMC Cancer},
-	number = {1},
-}
-
-@Article{lamanno_rnavelocitysingle_2018,
-	title = {{{RNA Velocity}} of {{Single Cells}}},
-	author = {Gioele {La Manno} and Ruslan Soldatov and Amit Zeisel and Emelie Braun and Hannah Hochgerner and Viktor Petukhov and Katja Lidschreiber and Maria E Kastriti and Peter L{\"o}nnerberg and Alessandro Furlan and Jean Fan and Lars E Borm and Zehua Liu and David {van Bruggen} and Jimin Guo and Xiaoling He and Roger Barker and Erik Sundstr{\"o}m and Gon{\c c}alo Castelo-Branco and Patrick Cramer and Igor Adameyko and Sten Linnarsson and Peter V Kharchenko},
-	year = {2018},
-	month = {aug},
-	volume = {560},
-	pages = {494--498},
-	issn = {1476-4687},
-	doi = {10.1038/s41586-018-0414-6},
-	abstract = {RNA velocity, estimated in single cells by comparison of spliced and unspliced mRNA, is a good indicator of transcriptome dynamics and will provide a useful tool for analysis of developmental lineage.},
-	journal = {Nature},
-	number = {7719},
-}
-
-@Article{lin_scclassifysamplesize_2020,
-	title = {{{scClassify}}: Sample Size Estimation and Multiscale Classification of Cells Using Single and Multiple Reference},
-	author = {Yingxin Lin and Yue Cao and Hani Jieun Kim and Agus Salim and Terence P Speed and David M Lin and Pengyi Yang and Jean Yee Hwa Yang},
-	year = {2020},
-	month = {jun},
-	volume = {16},
-	pages = {e9389},
-	publisher = {{John Wiley \& Sons, Ltd}},
-	issn = {1744-4292},
-	doi = {10.15252/msb.20199389},
-	abstract = {Abstract Automated cell type identification is a key computational challenge in single-cell RNA-sequencing (scRNA-seq) data.
-To capitalise on the large collection of well-annotated scRNA-seq datasets, we developed scClassify, a multiscale classification framework based on ensemble learning and cell type hierarchies constructed from single or multiple annotated datasets as references.
-scClassify enables the estimation of sample size required for accurate classification of cell types in a cell type hierarchy and allows joint classification of cells when multiple references are available.
-We show that scClassify consistently performs better than other supervised cell type classification methods across 114 pairs of reference and testing data, representing a diverse combination of sizes, technologies and levels of complexity, and further demonstrate the unique components of scClassify through simulations and compendia of experimental datasets.
-Finally, we demonstrate the scalability of scClassify on large single-cell atlases and highlight a novel application of identifying subpopulations of cells from the Tabula Muris data that were unidentified in the original publication.
-Together, scClassify represents state-of-the-art methodology in automated cell type identification from scRNA-seq data.},
-	journal = {Molecular Systems Biology},
-	keywords = {cell type hierarchy,cell type identification,multiscale classification,sample size estimation,single-cell},
-	number = {6},
-}
-
-@Article{liu_personalizedcharacterizationdiseases_2016,
-	title = {Personalized Characterization of Diseases Using Sample-Specific Networks},
-	author = {Xiaoping Liu and Yuetong Wang and Hongbin Ji and Kazuyuki Aihara and Luonan Chen},
-	year = {2016},
-	volume = {44},
-	pages = {e164--e164},
-	issn = {0305-1048},
-	doi = {10.1093/nar/gkw772},
-	abstract = {A complex disease generally results not from malfunction of individual molecules but from dysfunction of the relevant system or network, which dynamically changes with time and conditions.
-Thus, estimating a condition-specific network from a single sample is crucial to elucidating the molecular mechanisms of complex diseases at the system level.
-However, there is currently no effective way to construct such an individual-specific network by expression profiling of a single sample because of the requirement of multiple samples for computing correlations.
-We developed here with a statistical method, i.e.
-a sample-specific network (SSN) method, which allows us to construct individual-specific networks based on molecular expressions of a single sample.
-Using this method, we can characterize various human diseases at a network level.
-In particular, such SSNs can lead to the identification of individual-specific disease modules as well as driver genes, even without gene sequencing information.
-Extensive analysis by using the Cancer Genome Atlas data not only demonstrated the effectiveness of the method, but also found new individual-specific driver genes and network patterns for various types of cancer.
-Biological experiments on drug resistance further validated one important advantage of our method over the traditional methods, i.e.
-we can even identify such drug resistance genes that actually have no clear differential expression between samples with and without the resistance, due to the additional network information.},
-	file = {/home/rcannood/Zotero/storage/6MSYIL7R/Liu et al. - 2016 - Personalized characterization of diseases using sample-specific networks.pdf},
-	journal = {Nucleic Acids Research},
-	keywords = {cancer,drug resistance,genes,steroid-sensitive nephrotic syndrome},
-	number = {22},
-	pmid = {27596597},
-}
-
-@Article{luecken_currentbestpractices_2019,
-	title = {Current Best Practices in Single-Cell {{RNA}}-Seq Analysis: A Tutorial},
-	shorttitle = {Current Best Practices in Single-Cell {{RNA}}-Seq Analysis},
-	author = {Malte D Luecken and Fabian J Theis},
-	year = {2019},
-	month = {jun},
-	volume = {15},
-	pages = {e8746},
-	issn = {1744-4292},
-	doi = {10.15252/msb.20188746},
-	abstract = {Abstract Single-cell RNA-seq has enabled gene expression to be studied at an unprecedented resolution.
-The promise of this technology is attracting a growing user base for single-cell analysis methods.
-As more analysis tools are becoming available, it is becoming increasingly difficult to navigate this landscape and produce an up-to-date workflow to analyse one's data.
-Here, we detail the steps of a typical single-cell RNA-seq analysis, including pre-processing (quality control, normalization, data correction, feature selection, and dimensionality reduction) and cell- and gene-level downstream analysis.
-We formulate current best-practice recommendations for these steps based on independent comparison studies.
-We have integrated these best-practice recommendations into a workflow, which we apply to a public dataset to further illustrate how these steps work in practice.
-Our documented case study can be found at https://www.github.com/theislab/single-cell-tutorial.
-This review will serve as a workflow tutorial for new entrants into the field, and help established users update their analysis pipelines.},
-	file = {/home/rcannood/Zotero/storage/MY4TZUJT/Luecken and Theis - 2019 - Current best practices in single-cell RNA-seq anal.pdf;/home/rcannood/Zotero/storage/7PFR3DKU/msb.html},
-	journal = {Molecular Systems Biology},
-	keywords = {analysis pipeline development,computational biology,data analysis tutorial,single-cell RNA-seq},
-	number = {6},
-}
-
-@Article{lummertzdarocha_reconstructioncomplexsinglecell_2018,
-	title = {Reconstruction of Complex Single-Cell Trajectories Using {{CellRouter}}},
-	author = {Edroaldo {Lummertz da Rocha} and R.
-Grant Rowe and Vanessa Lundin and Mohan Malleshaiah and Deepak Kumar Jha and Carlos R.
-Rambo and Hu Li and Trista E.
-North and James J.
-Collins and George Q.
-Daley},
-	year = {2018},
-	month = {mar},
-	volume = {9},
-	pages = {892},
-	issn = {2041-1723},
-	doi = {10.1038/s41467-018-03214-y},
-	abstract = {A better understanding of the cell-fate transitions that occur in complex cellular ecosystems in normal development and disease could inform cell engineering efforts and lead to improved therapies.
-However, a major challenge is to simultaneously identify new cell states, and their transitions, to elucidate the gene expression dynamics governing cell-type diversification.
-Here, we present CellRouter, a multifaceted single-cell analysis platform that identifies complex cell-state transition trajectories by using flow networks to explore the subpopulation structure of multi-dimensional, single-cell omics data.
-We demonstrate its versatility by applying CellRouter to single-cell RNA sequencing data sets to reconstruct cell-state transition trajectories during hematopoietic stem and progenitor cell (HSPC) differentiation to the erythroid, myeloid and lymphoid lineages, as well as during re-specification of cell identity by cellular reprogramming of monocytes and B-cells to HSPCs.
-CellRouter opens previously undescribed paths for in-depth characterization of complex cellular ecosystems and establishment of enhanced cell engineering approaches.},
-	journal = {Nature Communications},
-	number = {1},
-}
-
-@Article{marbach_generatingrealisticsilico_2009,
-	title = {Generating {{Realistic In Silico Gene Networks}} for {{Performance Assessment}} of {{Reverse Engineering Methods}}},
-	author = {Daniel Marbach and Thomas Schaffter and Claudio Mattiussi and Dario Floreano},
-	year = {2009},
-	month = {feb},
-	volume = {16},
-	pages = {229--239},
-	issn = {1557-8666},
-	doi = {10.1089/cmb.2008.09TT},
-	abstract = {Reverse engineering methods are typically first tested on simulated data from in silico networks, for systematic and efficient performance assessment, before an application to real biological networks.
-In this paper, we present a method for generating biologically plausible in silico networks, which allow realistic performance assessment of network inference algorithms.
-Instead of using random graph models, which are known to only partly capture the structural properties of biological networks, we generate network structures by extracting modules from known biological interaction networks.
-Using the yeast transcriptional regulatory network as a test case, we show that extracted modules have a biologically plausible connectivity because they preserve functional and structural properties of the original network.
-Our method was selected to generate the {"}gold standard{"} networks for the gene network reverse engineering challenge of the third DREAM conference (Dialogue on Reverse Engineering Assessment and Methods 2008, Cambridge, MA).},
-	file = {/home/rcannood/Zotero/storage/NXLVA76Q/Marbach et al. - 2009 - Generating Realistic In Silico Gene Networks for Performance Assessment of Reverse Engineering Methods.pdf},
-	journal = {Journal of Computational Biology},
-	keywords = {Algorithms,Computational Biology,Computational Biology: methods,Computer Simulation,DRE,Gene Expression Profiling,Gene Expression Profiling: methods,gene regulatory networks,Gene Regulatory Networks,Genetic,Models,reverse engineering,Saccharomyces cerevisiae,Saccharomyces cerevisiae: genetics,Saccharomyces cerevisiae: metabolism,Software},
-	number = {2},
-	pmid = {19183003},
-}
-
-@Article{marbach_revealingstrengthsweaknesses_2010,
-	title = {Revealing Strengths and Weaknesses of Methods for Gene Network Inference},
-	author = {Daniel Marbach and Robert J Prill and Thomas Schaffter and Claudio Mattiussi and Dario Floreano and Gustavo Stolovitzky},
-	year = {2010},
-	month = {apr},
-	volume = {107},
-	pages = {6286--6291},
-	issn = {1091-6490},
-	doi = {10.1073/pnas.0913357107},
-	abstract = {Numerous methods have been developed for inferring gene regulatory networks from expression data, however, both their absolute and comparative performance remain poorly understood.
-In this paper, we introduce a framework for critical performance assessment of methods for gene network inference.
-We present an in silico benchmark suite that we provided as a blinded, community-wide challenge within the context of the DREAM (Dialogue on Reverse Engineering Assessment and Methods) project.
-We assess the performance of 29 gene-network-inference methods, which have been applied independently by participating teams.
-Performance profiling reveals that current inference methods are affected, to various degrees, by different types of systematic prediction errors.
-In particular, all but the best-performing method failed to accurately infer multiple regulatory inputs (combinatorial regulation) of genes.
-The results of this community-wide experiment show that reliable network inference from gene expression data remains an unsolved problem, and they indicate potential ways of network reconstruction improvements.},
-	file = {/home/rcannood/Zotero/storage/784SY83Y/Marbach et al. - 2010 - Revealing strengths and weaknesses of methods for gene network inference.pdf},
-	journal = {Proceedings of the National Academy of Sciences},
-	keywords = {Biometry,community experiment,Computational Biology,Computational Biology: methods,DREAM challenge,Gene Expression Profiling,Gene Regulatory Networks,performance assessment,reverse engineering,transcriptional regulatory networks},
-	number = {14},
-	pmid = {20308593},
-}
-
-@Article{marbach_wisdomcrowdsrobust_2012,
-	title = {Wisdom of {{Crowds}} for {{Robust Gene Network Inference}}},
-	author = {Daniel Marbach and James C Costello and Robert K{\"u}ffner and Nicole M Nicci Vega and Robert J Prill and Diogo M Camacho and Kyle R Allison and Manolis Kellis and James J Collins and Gustavo Stolovitzky and {DREAM5 Consortium} and Manolis Kellis and James J Collins and Gustavo Stolovitzky},
-	year = {2012},
-	month = {jul},
-	volume = {9},
-	pages = {796--804},
-	issn = {1548-7091},
-	doi = {10.1038/nmeth.2016},
-	abstract = {Reconstructing gene regulatory networks from high-throughput data is a long-standing problem.
-Through the DREAM project (Dialogue on Reverse Engineering Assessment and Methods), we performed a comprehensive blind assessment of over thirty network inference methods on Escherichia coli, Staphylococcus aureus, Saccharomyces cerevisiae, and in silico microarray data.
-We characterize performance, data requirements, and inherent biases of different inference approaches offering guidelines for both algorithm application and development.
-We observe that no single inference method performs optimally across all datasets.
-In contrast, integration of predictions from multiple inference methods shows robust and high performance across diverse datasets.
-Thereby, we construct high-confidence networks for E.
-coli and S.
-aureus, each comprising \$\textbackslash backslash\$textasciitilde\{\}1700 transcriptional interactions at an estimated precision of 50\%.
-We experimentally test 53 novel interactions in E.
-coli, of which 23 were supported (43\%).
-Our results establish community-based methods as a powerful and robust tool for the inference of transcriptional gene regulatory networks.},
-	file = {/home/rcannood/Zotero/storage/AQRNS57R/Marbach et al.
-- 2012 - Wisdom of crowds for robust gene network inference.pdf;/home/rcannood/Zotero/storage/KGJVQ9WL/Marbach et al.
-- 2012 - Wisdom of crowds for robust gene network inference.pdf},
-	journal = {Nature methods},
-	keywords = {Algorithms,Bacterial,Bacterial: genetics,Computational Biology,Escherichia coli,Escherichia coli: genetics,Gene Expression Regulation,Gene Regulatory Networks,Genetic,Genetic: genetics,Oligonucleotide Array Sequence Analysis,Saccharomyces cerevisiae,Saccharomyces cerevisiae: genetics,Software,Staphylococcus aureus,Staphylococcus aureus: genetics,Transcription},
-	number = {8},
-	pmid = {22796662},
-}
-
-@Article{mcfaline-figueroa_pooledsinglecellgenetic_2019,
-	title = {A Pooled Single-Cell Genetic Screen Identifies Regulatory Checkpoints in the Continuum of the Epithelial-to-Mesenchymal Transition},
-	author = {Jos{\a'e} L. McFaline-Figueroa and Andrew J. Hill and Xiaojie Qiu and Dana Jackson and Jay Shendure and Cole Trapnell},
-	year = {2019},
-	month = {sep},
-	volume = {51},
-	pages = {1389--1398},
-	issn = {1546-1718},
-	doi = {10.1038/s41588-019-0489-5},
-	abstract = {Transcriptional profiling of 61,052 cells undergoing epithelial-to-mesenchymal transition (EMT) shows continuous waves of gene regulation, not discrete stages.
-A CRISPR screen orders EMT regulators along EMT progression.},
-	copyright = {2019 The Author(s), under exclusive licence to Springer Nature America, Inc.},
-	file = {/home/rcannood/Zotero/storage/RAQRP6UC/s41588-019-0489-5.html},
-	journal = {Nature Genetics},
-	language = {en},
-	number = {9},
-}
-
-@Article{moerman_grnboost2arboretoefficient_2019,
-	title = {{{GRNBoost2}} and {{Arboreto}}: Efficient and Scalable Inference of Gene Regulatory Networks},
-	author = {Thomas Moerman and Sara {Aibar Santos} and Carmen {Bravo Gonz{\a'a}lez-Blas} and Jaak Simm and Yves Moreau and Jan Aerts and Stein Aerts},
-	year = {2019},
-	month = {jun},
-	volume = {35},
-	pages = {2159--2161},
-	issn = {1367-4803},
-	doi = {10.1093/bioinformatics/bty916},
-	abstract = {Inferring a Gene Regulatory Network (GRN) from gene expression data is a computationally expensive task, exacerbated by increasing data sizes due to advances in high-throughput gene profiling technology, such as single-cell RNA-seq.
-To equip researchers with a toolset to infer GRNs from large expression datasets, we propose GRNBoost2 and the Arboreto framework.
-GRNBoost2 is an efficient algorithm for regulatory network inference using gradient boosting, based on the GENIE3 architecture.
-Arboreto is a computational framework that scales up GRN inference algorithms complying with this architecture.
-Arboreto includes both GRNBoost2 and an improved implementation of GENIE3, as a user-friendly open source Python package.Arboreto is available under the 3-Clause BSD license at http://arboreto.readthedocs.io.Supplementary data are available at Bioinformatics online.},
-	journal = {Bioinformatics},
-	number = {12},
-}
-
-@Article{papadopoulos_prossttprobabilisticsimulation_2019,
-	title = {{{PROSSTT}}: Probabilistic Simulation of Single-Cell {{RNA}}-Seq Data for Complex Differentiation Processes},
-	shorttitle = {{{PROSSTT}}},
-	author = {Nikolaos Papadopoulos and Parra R. Gonzalo and Johannes S{\"o}ding},
-	year = {2019},
-	month = {sep},
-	volume = {35},
-	pages = {3517--3519},
-	issn = {1367-4811},
-	doi = {10.1093/bioinformatics/btz078},
-	abstract = {SUMMARY: Cellular lineage trees can be derived from single-cell RNA sequencing snapshots of differentiating cells.
-Currently, only datasets with simple topologies are available.
-To test and further develop tools for lineage tree reconstruction, we need test datasets with known complex topologies.
-PROSSTT can simulate scRNA-seq datasets for differentiation processes with lineage trees of any desired complexity, noise level, noise model and size.
-PROSSTT also provides scripts to quantify the quality of predicted lineage trees.
-AVAILABILITY AND IMPLEMENTATION: https://github.com/soedinglab/prosstt.
-SUPPLEMENTARY INFORMATION: Supplementary data are available at Bioinformatics online.},
-	file = {/home/rcannood/Zotero/storage/UERERPJX/Papadopoulos et al. - 2019 - PROSSTT probabilistic simulation of single-cell R.pdf},
-	journal = {Bioinformatics (Oxford, England)},
-	keywords = {Cell Differentiation,Gene Expression Profiling,RNA-Seq,Single-Cell Analysis,Software},
-	language = {eng},
-	number = {18},
-	pmcid = {PMC6748774},
-	pmid = {30715210},
-}
-
-@Article{parra_reconstructingcomplexlineage_2019,
-	title = {Reconstructing Complex Lineage Trees from {{scRNA}}-Seq Data Using {{MERLoT}}},
-	author = {R. Gonzalo Parra and Nikolaos Papadopoulos and Laura Ahumada-Arranz and Jakob El Kholtei and Noah Mottelson and Yehor Horokhovsky and Barbara Treutlein and Johannes Soeding},
-	year = {2019},
-	month = {sep},
-	volume = {47},
-	pages = {8961--8974},
-	issn = {1362-4962},
-	doi = {10.1093/nar/gkz706},
-	abstract = {Advances in single-cell transcriptomics techniques are revolutionizing studies of cellular differentiation and heterogeneity.
-It has become possible to track the trajectory of thousands of genes across the cellular lineage trees that represent the temporal emergence of cell types during dynamic processes.
-However, reconstruction of cellular lineage trees with more than a few cell fates has proved challenging.
-We present MERLoT (https://github.com/soedinglab/merlot), a flexible and user-friendly tool to reconstruct complex lineage trees from single-cell transcriptomics data.
-It can impute temporal gene expression profiles along the reconstructed tree.
-We show MERLoT's capabilities on various real cases and hundreds of simulated datasets.},
-	file = {/home/rcannood/Zotero/storage/86VN2MRY/Parra et al. - 2019 - Reconstructing complex lineage trees from scRNA-se.pdf},
-	journal = {Nucleic Acids Research},
-	keywords = {Algorithms,Animals,Cell Differentiation,Cell Lineage,Cluster Analysis,Gene Expression Profiling,Hematopoietic Stem Cells,Humans,Mice,Mouse Embryonic Stem Cells,Sequence Analysis; RNA,Single-Cell Analysis,Software,Time Factors,Transcriptome},
-	language = {eng},
-	number = {17},
-	pmcid = {PMC6755101},
-	pmid = {31428793},
-}
-
-@Manual{patil_pairwisecomparisonsmultiplepairwise_2019,
-	title = {{{pairwiseComparisons}}: {{Multiple}} Pairwise Comparison Tests},
-	author = {Indrajeet Patil},
-	year = {2019},
-	type = {Manual},
-}
-
-@Article{prill_rigorousassessmentsystems_2010,
-	title = {Towards a {{Rigorous Assessment}} of {{Systems Biology Models}}: {{The DREAM3 Challenges}}},
-	author = {Robert J Prill and Marbach Daniel and Saez-Rodriguez Julio and Sorger Peter K.
-and Alexopoulos Leonidas G.
-and Xue Xiaowei and Clarke Neil D.
-and Altan-Bonnet Gregoire and Stolovitzky Gustavo and Daniel Marbach and Julio Saez-Rodriguez and Peter K Sorger and Leonidas G Alexopoulos and Xiaowei Xue and Neil D Clarke and Gregoire Altan-Bonnet and Gustavo Stolovitzky and Marbach Daniel and Saez-Rodriguez Julio and Sorger Peter K.
-and Alexopoulos Leonidas G.
-and Xue Xiaowei and Clarke Neil D.
-and Altan-Bonnet Gregoire and Stolovitzky Gustavo},
-	year = {2010},
-	month = {jan},
-	volume = {5},
-	pages = {e9202},
-	issn = {1932-6203},
-	doi = {10.1371/journal.pone.0009202},
-	abstract = {BACKGROUND: Systems biology has embraced computational modeling in response to the quantitative nature and increasing scale of contemporary data sets.
-The onslaught of data is accelerating as molecular profiling technology evolves.
-The Dialogue for Reverse Engineering Assessments and Methods (DREAM) is a community effort to catalyze discussion about the design, application, and assessment of systems biology models through annual reverse-engineering challenges.
-METHODOLOGY AND PRINCIPAL FINDINGS: We describe our assessments of the four challenges associated with the third DREAM conference which came to be known as the DREAM3 challenges: signaling cascade identification, signaling response prediction, gene expression prediction, and the DREAM3 in silico network challenge.
-The challenges, based on anonymized data sets, tested participants in network inference and prediction of measurements.
-Forty teams submitted 413 predicted networks and measurement test sets.
-Overall, a handful of best-performer teams were identified, while a majority of teams made predictions that were equivalent to random.
-Counterintuitively, combining the predictions of multiple teams (including the weaker teams) can in some cases improve predictive power beyond that of any single method.
-CONCLUSIONS: DREAM provides valuable feedback to practitioners of systems biology modeling.
-Lessons learned from the predictions of the community provide much-needed context for interpreting claims of efficacy of algorithms described in the scientific literature.},
-	file = {/home/rcannood/Zotero/storage/Q5E22H2F/Prill et al. - 2010 - Towards a Rigorous Assessment of Systems Biology Models The DREAM3 Challenges.pdf},
-	journal = {PLoS ONE},
-	keywords = {Algorithms,Animals,Biological,Cluster Analysis,Computational Biology,Computational Biology: methods,Gene Expression Profiling,Gene Expression Profiling: methods,Gene Regulatory Networks,Humans,Models,Protein Interaction Mapping,Protein Interaction Mapping: methods,Reproducibility of Results,Signal Transduction,Software,Systems Biology,Systems Biology: methods},
-	number = {2},
-	pmid = {20186320},
-}
-
-@Article{rekhtman_directinteractionhematopoietic_1999,
-	title = {Direct Interaction of Hematopoietic Transcription Factors {{PU}}.1 and {{GATA}}-1: Functional Antagonism in Erythroid Cells},
-	shorttitle = {Direct Interaction of Hematopoietic Transcription Factors {{PU}}.1 and {{GATA}}-1},
-	author = {N. Rekhtman and F. Radparvar and T. Evans and A. I. Skoultchi},
-	year = {1999},
-	month = {jun},
-	volume = {13},
-	pages = {1398--1411},
-	issn = {0890-9369},
-	doi = {10.1101/gad.13.11.1398},
-	abstract = {Malignant transformation usually inhibits terminal cell differentiation but the precise mechanisms involved are not understood.
-PU.1 is a hematopoietic-specific Ets family transcription factor that is required for development of some lymphoid and myeloid lineages.
-PU.1 can also act as an oncoprotein as activation of its expression in erythroid precursors by proviral insertion or transgenesis causes erythroleukemias in mice.
-Restoration of terminal differentiation in the mouse erythroleukemia (MEL) cells requires a decline in the level of PU.1, indicating that PU.1 can block erythroid differentiation.
-Here we investigate the mechanism by which PU.1 interferes with erythroid differentiation.
-We find that PU.1 interacts directly with GATA-1, a zinc finger transcription factor required for erythroid differentiation.
-Interaction between PU.1 and GATA-1 requires intact DNA-binding domains in both proteins.
-PU.1 represses GATA-1-mediated transcriptional activation.
-Both the DNA binding and transactivation domains of PU.1 are required for repression and both domains are also needed to block terminal differentiation in MEL cells.
-We also show that ectopic expression of PU.1 in Xenopus embryos is sufficient to block erythropoiesis during normal development.
-Furthermore, introduction of exogenous GATA-1 in both MEL cells and Xenopus embryos and explants relieves the block to erythroid differentiation imposed by PU.1.
-Our results indicate that the stoichiometry of directly interacting but opposing transcription factors may be a crucial determinant governing processes of normal differentiation and malignant transformation.},
-	file = {/home/rcannood/Zotero/storage/77DJ34BY/Rekhtman et al. - 1999 - Direct interaction of hematopoietic transcription .pdf},
-	journal = {Genes \& Development},
-	keywords = {3T3 Cells,Animals,Binding Sites,Cell Differentiation,DNA-Binding Proteins,Erythroid Precursor Cells,Erythroid-Specific DNA-Binding Factors,GATA1 Transcription Factor,HeLa Cells,Hematopoiesis,Humans,Mice,Proto-Oncogene Proteins,Recombinant Fusion Proteins,Trans-Activators,Transcription Factors,Transcription; Genetic,Transcriptional Activation,Xenopus laevis,Zinc Fingers},
-	language = {eng},
-	number = {11},
-	pmcid = {PMC316770},
-	pmid = {10364157},
-}
-
-@Article{roy_systemgeneratingtranscription_2008,
-	title = {A System for Generating Transcription Regulatory Networks with Combinatorial Control of Transcription},
-	author = {Sushmita Roy and Margaret Werner-Washburne and Terran Lane},
-	year = {2008},
-	month = {may},
-	volume = {24},
-	pages = {1318--1320},
-	issn = {1367-4803},
-	doi = {10.1093/bioinformatics/btn126},
-	abstract = {Summary: We have developed a new software system, REgulatory Network generator with COmbinatorial control (RENCO), for automatic generation of differential equations describing pre-transcriptional combinatorics in artificial regulatory networks.
-RENCO has the following benefits: (a) it explicitly models protein\textendash protein interactions among transcription factors, (b) it captures combinatorial control of transcription factors on target genes and (c) it produces output in Systems Biology Markup Language (SBML) format, which allows these equations to be directly imported into existing simulators.
-Explicit modeling of the protein interactions allows RENCO to incorporate greater mechanistic detail of the transcription machinery compared to existing models and can provide a better assessment of algorithms for regulatory network inference.Availability: RENCO is a C++ command line program, available at http://sourceforge.net/projects/renco/Contact:terran@cs.unm.eduSupplementary information: Supplementary data are available at Bioinformatics online.},
-	journal = {Bioinformatics},
-	number = {10},
-}
-
-@Article{saelens_comparisonsinglecelltrajectory_2019,
-	title = {A Comparison of Single-Cell Trajectory Inference Methods},
-	author = {Wouter Saelens and Robrecht Cannoodt and Helena Todorov and Yvan Saeys},
-	year = {2019},
-	volume = {37},
-	issn = {15461696},
-	doi = {10.1038/s41587-019-0071-9},
-	abstract = {Trajectory inference approaches analyze genome-wide omics data from thousands of single cells and computationally infer the order of these cells along developmental trajectories.
-Although more than 70 trajectory inference tools have already been developed, it is challenging to compare their performance because the input they require and output models they produce vary substantially.
-Here, we benchmark 45 of these methods on 110 real and 229 synthetic datasets for cellular ordering, topology, scalability and usability.
-Our results highlight the complementarity of existing tools, and that the choice of method should depend mostly on the dataset dimensions and trajectory topology.
-Based on these results, we develop a set of guidelines to help users select the best method for their dataset.
-Our freely available data and evaluation pipeline ( https://benchmark.dynverse.org ) will aid in the development of improved tools designed to analyze increasingly large and complex single-cell datasets.},
-	file = {/home/rcannood/Zotero/storage/JKZCFAVA/Saelens et al. - 2019 - A comparison of single-cell trajectory inference methods.pdf;/home/rcannood/Zotero/storage/T9HMNIDB/Saelens2019.pdf},
-	journal = {Nature Biotechnology},
-	number = {May},
-}
-
-@Article{schaffter_genenetweaversilicobenchmark_2011,
-	title = {{{GeneNetWeaver}}: In Silico Benchmark Generation and Performance Profiling of Network Inference Methods.},
-	shorttitle = {\{\vphantom\}{{GeneNetWeaver}}\vphantom\{\}},
-	author = {Thomas Schaffter and Daniel Marbach and Dario Floreano},
-	year = {2011},
-	month = {aug},
-	volume = {27},
-	pages = {2263--2270},
-	issn = {1367-4811},
-	doi = {10.1093/bioinformatics/btr373},
-	abstract = {MOTIVATION: Over the last decade, numerous methods have been developed for inference of regulatory networks from gene expression data.
-However, accurate and systematic evaluation of these methods is hampered by the difficulty of constructing adequate benchmarks and the lack of tools for a differentiated analysis of network predictions on such benchmarks.
-RESULTS: Here, we describe a novel and comprehensive method for in silico benchmark generation and performance profiling of network inference methods available to the community as an open-source software called GeneNetWeaver (GNW).
-In addition to the generation of detailed dynamical models of gene regulatory networks to be used as benchmarks, GNW provides a network motif analysis that reveals systematic prediction errors, thereby indicating potential ways of improving inference methods.
-The accuracy of network inference methods is evaluated using standard metrics such as precision-recall and receiver operating characteristic curves.
-We show how GNW can be used to assess the performance and identify the strengths and weaknesses of six inference methods.
-Furthermore, we used GNW to provide the international Dialogue for Reverse Engineering Assessments and Methods (DREAM) competition with three network inference challenges (DREAM3, DREAM4 and DREAM5).
-AVAILABILITY: GNW is available at http://gnw.sourceforge.net along with its Java source code, user manual and supporting data.
-SUPPLEMENTARY INFORMATION: Supplementary data are available at Bioinformatics online.
-CONTACT: dario.floreano@epfl.ch.},
-	file = {/home/rcannood/Zotero/storage/P5JMJUBN/Schaffter, Marbach, Floreano - 2011 - GeneNetWeaver in silico benchmark generation and performance profiling of network inference method.pdf},
-	journal = {Bioinformatics},
-	keywords = {Benchmarking,Escherichia coli,Escherichia coli: genetics,Gene Expression,Gene Regulatory Networks,Genetic,Models,Software},
-	number = {16},
-	pmid = {21697125},
-}
-
-@Article{schilstra_biologicgeneexpression_2008,
-	title = {Bio-{{Logic}}: {{Gene Expression}} and the {{Laws}} of {{Combinatorial Logic}}},
-	shorttitle = {Bio-{{Logic}}},
-	author = {Maria J. Schilstra and Chrystopher L. Nehaniv},
-	year = {2008},
-	month = {jan},
-	volume = {14},
-	pages = {121--133},
-	issn = {1064-5462},
-	doi = {10.1162/artl.2008.14.1.121},
-	abstract = {Abstract At the heart of the development of fertilized eggs into fully formed organisms and the adaptation of cells to changed conditions are genetic regulatory networks (GRNs).
-In higher multicellular organisms, signal selection and multiplexing are performed at the cis-regulatory domains of genes, where combinations of transcription factors (TFs) regulate the rates at which the genes are transcribed into mRNA.
-To be able to act as activators or repressors of gene transcription, TFs must first bind to target sequences on the regulatory domains.
-Two TFs that act in concert may bind entirely independently of each other, but more often binding of the first one will alter the affinity of the other for its binding site.
-This article presents a systematic investigation into the effect of TF binding dependences on the predicted regulatory function of this bio-logic.
-Four extreme scenarios, commonly used to classify enzyme activation and inhibition patterns, for the binding of two TFs were explored: independent (the TFs bind without affecting each other's affinities), competitive (the TFs compete for the same binding site), ordered (the TFs bind in a compulsory order), and joint binding (the TFs either bind as a preformed complex, or binding of one is virtually impossible in the absence of the other).
-The conclusions are: (1) the laws of combinatorial logic hold only for systems with independently binding TFs; (2) systems formed according to the other scenarios can mimic the functions of their Boolean logical counterparts, but cannot be combined or decomposed in the same way; and (3) the continuously scaled output of systems consisting of competitively binding activators and repressors can be controlled more robustly than that of single TF or (quasi-)logical multi-TF systems.},
-	file = {/home/rcannood/Zotero/storage/C7UWQGI3/Schilstra and Nehaniv - 2008 - Bio-Logic Gene Expression and the Laws of Combina.pdf;/home/rcannood/Zotero/storage/KXDLUMAG/artl.2008.14.1.html},
-	journal = {Artificial Life},
-	number = {1},
-}
-
-@Article{schwanhausser_globalquantificationmammalian_2011,
-	title = {Global Quantification of Mammalian Gene Expression Control},
-	author = {Bj{\"o}rn Schwanh{\"a}usser and Dorothea Busse and Na Li and Gunnar Dittmar and Johannes Schuchhardt and Jana Wolf and Wei Chen and Matthias Selbach},
-	year = {2011},
-	month = {may},
-	volume = {473},
-	pages = {337--342},
-	issn = {1476-4687},
-	doi = {10.1038/nature10098},
-	abstract = {Gene expression is a multistep process that involves the transcription, translation and turnover of messenger RNAs and proteins.
-Although it is one of the most fundamental processes of life, the entire cascade has never been quantified on a genome-wide scale.
-Here we simultaneously measured absolute mRNA and protein abundance and turnover by parallel metabolic pulse labelling for more than 5,000 genes in mammalian cells.
-Whereas mRNA and protein levels correlated better than previously thought, corresponding half-lives showed no correlation.
-Using a quantitative model we have obtained the first genome-scale prediction of synthesis rates of mRNAs and proteins.
-We find that the cellular abundance of proteins is predominantly controlled at the level of translation.
-Genes with similar combinations of mRNA and protein stability shared functional properties, indicating that half-lives evolved under energetic and dynamic constraints.
-Quantitative information about all stages of gene expression provides a rich resource and helps to provide a greater understanding of the underlying design principles.},
-	journal = {Nature},
-	number = {7347},
-}
-
-@Article{smith_spatialstochasticintracellular_2019,
-	title = {Spatial {{Stochastic Intracellular Kinetics}}: {{A Review}} of {{Modelling Approaches}}},
-	author = {Stephen Smith and Ramon Grima},
-	year = {2019},
-	month = {aug},
-	volume = {81},
-	pages = {2960--3009},
-	issn = {1522-9602},
-	doi = {10.1007/s11538-018-0443-1},
-	abstract = {Models of chemical kinetics that incorporate both stochasticity and diffusion are an increasingly common tool for studying biology.
-The variety of competing models is vast, but two stand out by virtue of their popularity: the reaction\textendash diffusion master equation and Brownian dynamics.
-In this review, we critically address a number of open questions surrounding these models: How can they be justified physically?
-How do they relate to each other?
-How do they fit into the wider landscape of chemical models, ranging from the rate equations to molecular dynamics?
-This review assumes no prior knowledge of modelling chemical kinetics and should be accessible to a wide range of readers.},
-	journal = {Bulletin of Mathematical Biology},
-	number = {8},
-}
-
-@Article{soneson_biasrobustnessscalability_2018,
-	title = {Bias, Robustness and Scalability in Single-Cell Differential Expression Analysis},
-	author = {Charlotte Soneson and Mark D. Robinson},
-	year = {2018},
-	month = {apr},
-	volume = {15},
-	pages = {255--261},
-	issn = {1548-7105},
-	doi = {10.1038/nmeth.4612},
-	abstract = {Many methods have been used to determine differential gene expression from single-cell RNA (scRNA)-seq data.
-We evaluated 36 approaches using experimental and synthetic data and found considerable differences in the number and characteristics of the genes that are called differentially expressed.
-Prefiltering of lowly expressed genes has important effects, particularly for some of the methods developed for bulk RNA-seq data analysis.
-However, we found that bulk RNA-seq analysis methods do not generally perform worse than those developed specifically for scRNA-seq.
-We also present conquer, a repository of consistently processed, analysis-ready public scRNA-seq data sets that is aimed at simplifying method evaluation and reanalysis of published results.
-Each data set provides abundance estimates for both genes and transcripts, as well as quality control and exploratory analysis reports.},
-	journal = {Nature Methods},
-	keywords = {Computational Biology,Gene Expression Profiling,Gene Expression Regulation,High-Throughput Nucleotide Sequencing,RNA,Sequence Analysis; RNA,Single-Cell Analysis,Software},
-	language = {eng},
-	number = {4},
-	pmid = {29481549},
-}
-
-@Article{street_slingshotcelllineage_2018,
-	title = {Slingshot: {{Cell Lineage}} and {{Pseudotime Inference}} for {{Single}}-{{Cell Transcriptomics}}},
-	shorttitle = {Slingshot},
-	author = {Kelly Street and Davide Risso and Russell B Fletcher and Diya Das and John Ngai and Nir Yosef and Elizabeth Purdom and Sandrine Dudoit},
-	year = {2018},
-	month = {jun},
-	volume = {19},
-	pages = {477},
-	issn = {1471-2164},
-	doi = {10.1186/s12864-018-4772-0},
-	abstract = {Single-cell transcriptomics allows researchers to investigate complex communities of heterogeneous cells.
-It can be applied to stem cells and their descendants in order to chart the progression from multipotent progenitors to fully differentiated cells.
-While a variety of statistical and computational methods have been proposed for inferring cell lineages, the problem of accurately characterizing multiple branching lineages remains difficult to solve.},
-	journal = {BMC Genomics},
-	number = {1},
-}
-
-@Article{vallejos_normalizingsinglecellrna_2017,
-	title = {Normalizing Single-Cell {{RNA}} Sequencing Data: Challenges and Opportunities},
-	shorttitle = {Normalizing Single-Cell {{RNA}} Sequencing Data},
-	author = {Catalina A. Vallejos and Davide Risso and Antonio Scialdone and Sandrine Dudoit and John C. Marioni},
-	year = {2017},
-	month = {jun},
-	volume = {14},
-	pages = {565--571},
-	publisher = {{Nature Publishing Group}},
-	issn = {1548-7105},
-	doi = {10.1038/nmeth.4292},
-	abstract = {This Perspective examines single-cell RNA-seq data challenges and the need for normalization methods designed specifically for single-cell data in order to remove technical biases.},
-	copyright = {2017 Nature Publishing Group, a division of Macmillan Publishers Limited. All Rights Reserved.},
-	journal = {Nature Methods},
-	language = {en},
-	number = {6},
-}
-
-@Article{vandenbulcke_syntrengeneratorsynthetic_2006,
-	title = {{{SynTReN}}: A Generator of Synthetic Gene Expression Data for Design and Analysis of Structure Learning Algorithms},
-	shorttitle = {{{SynTReN}}},
-	author = {Tim {Van den Bulcke} and Koenraad {Van Leemput} and Bart Naudts and Piet {van Remortel} and Hongwu Ma and Alain Verschoren and Bart {De Moor} and Kathleen Marchal},
-	year = {2006},
-	month = {jan},
-	volume = {7},
-	pages = {43},
-	issn = {1471-2105},
-	doi = {10.1186/1471-2105-7-43},
-	abstract = {The development of algorithms to infer the structure of gene regulatory networks based on expression data is an important subject in bioinformatics research.
-Validation of these algorithms requires benchmark data sets for which the underlying network is known.
-Since experimental data sets of the appropriate size and design are usually not available, there is a clear need to generate well-characterized synthetic data sets that allow thorough testing of learning algorithms in a fast and reproducible manner.},
-	journal = {BMC Bioinformatics},
-	language = {en},
-	number = {1},
-}
-
-@Article{vandesande_scalablescenicworkflow_2020,
-	title = {A Scalable {{SCENIC}} Workflow for Single-Cell Gene Regulatory Network Analysis},
-	author = {Bram {Van de Sande} and Christopher Flerin and Kristofer Davie and Maxime {De Waegeneer} and Gert Hulselmans and Sara Aibar and Ruth Seurinck and Wouter Saelens and Robrecht Cannoodt and Quentin Rouchon and Toni Verbeiren and Dries {De Maeyer} and Joke Reumers and Yvan Saeys and Stein Aerts},
-	year = {2020},
-	month = {jul},
-	volume = {15},
-	pages = {2247--2276},
-	issn = {1750-2799},
-	doi = {10.1038/s41596-020-0336-2},
-	abstract = {This protocol explains how to perform a fast SCENIC analysis alongside standard best practices steps on single-cell RNA-sequencing data using software containers and Nextflow pipelines.
-SCENIC reconstructs regulons (i.e., transcription factors and their target genes) assesses the activity of these discovered regulons in individual cells and uses these cellular activity patterns to find meaningful clusters of cells.
-Here we present an improved version of SCENIC with several advances.
-SCENIC has been refactored and reimplemented in Python (pySCENIC), resulting in a tenfold increase in speed, and has been packaged into containers for ease of use.
-It is now also possible to use epigenomic track databases, as well as motifs, to refine regulons.
-In this protocol, we explain the different steps of SCENIC: the workflow starts from the count matrix depicting the gene abundances for all cells and consists of three stages.
-First, coexpression modules are inferred using a regression per-target approach (GRNBoost2).
-Next, the indirect targets are pruned from these modules using cis-regulatory motif discovery (cisTarget).
-Lastly, the activity of these regulons is quantified via an enrichment score for the regulon's target genes (AUCell).
-Nonlinear projection methods can be used to display visual groupings of cells based on the cellular activity patterns of these regulons.
-The results can be exported as a loom file and visualized in the SCope web application.
-This protocol is illustrated on two use cases: a peripheral blood mononuclear cell data set and a panel of single-cell RNA-sequencing cancer experiments.
-For a data set of 10,000 genes and 50,000 cells, the pipeline runs in {$<$}2 h.},
-	journal = {Nature Protocols},
-	number = {7},
-}
-
-@Article{vieth_powsimrpoweranalysis_2017,
-	title = {{{powsimR}}: Power Analysis for Bulk and Single Cell {{RNA}}-Seq Experiments},
-	shorttitle = {{{powsimR}}},
-	author = {Beate Vieth and Christoph Ziegenhain and Swati Parekh and Wolfgang Enard and Ines Hellmann},
-	year = {2017},
-	month = {nov},
-	volume = {33},
-	pages = {3486--3488},
-	issn = {1367-4803},
-	doi = {10.1093/bioinformatics/btx435},
-	abstract = {AbstractSummary.  Power analysis is essential to optimize the design of RNA-seq experiments and to assess and compare the power to detect differentially express},
-	file = {/home/rcannood/Zotero/storage/MFRF86MZ/Vieth et al. - 2017 - powsimR power analysis for bulk and single cell R.pdf;/home/rcannood/Zotero/storage/HSCUJRI8/3952669.html},
-	journal = {Bioinformatics},
-	language = {en},
-	number = {21},
-}
-
-@Article{wang_quantifyingwaddingtonlandscape_2011,
-	title = {Quantifying the {{Waddington Landscape}} and {{Biological Paths}} for {{Development}} and {{Differentiation}}},
-	author = {Jin Wang and Kun Zhang and Li Xu and Erkang Wang},
-	year = {2011},
-	month = {may},
-	volume = {108},
-	pages = {8257--8262},
-	issn = {0027-8424, 1091-6490},
-	doi = {10.1073/pnas.1017017108},
-	abstract = {We developed a theoretical framework to prove the existence and quantify the Waddington landscape as well as chreode-biological paths for development and differentiation.
-The cells can have states with the higher probability ones giving the different cell types.
-Different cell types correspond to different basins of attractions of the probability landscape.
-We study how the cells develop from undifferentiated cells to differentiated cells from landscape perspectives.
-We quantified the Waddington landscape through construction of underlying probability landscape for cell development.
-We show the developmental process proceeds as moving from undifferentiated to the differentiated basins of attractions.
-The barrier height of the basins of attractions correlates with the escape time that determines the stability of cell types.
-We show that the developmental process can be quantitatively described and uncovered by the biological paths on the quantified Waddington landscape from undifferentiated to the differentiated cells.
-We found the dynamics of the developmental process is controlled by a combination of the gradient and curl force on the landscape.
-The biological paths often do not follow the steepest descent path on the landscape.
-The landscape framework also quantifies the possibility of reverse differentiation process such as cell reprogramming from differentiated cells back to the original stem cell.
-We show that the biological path of reverse differentiation is irreversible and different from the one for differentiation process.
-We found that the developmental process described by the underlying landscape and the associated biological paths is relatively stable and robust against the influences of environmental perturbations.},
-	journal = {Proceedings of the National Academy of Sciences},
-	number = {20},
-	pmid = {21536909},
-}
-
-@Article{weber_essentialguidelinescomputational_2019,
-	title = {Essential Guidelines for Computational Method Benchmarking},
-	author = {Lukas M. Weber and Wouter Saelens and Robrecht Cannoodt and Charlotte Soneson and Alexander Hapfelmeier and Paul P. Gardner and Anne-Laure Boulesteix and Yvan Saeys and Mark D. Robinson},
-	year = {2019},
-	month = {jun},
-	volume = {20},
-	pages = {125},
-	issn = {1474-760X},
-	doi = {10.1186/s13059-019-1738-8},
-	abstract = {In computational biology and other sciences, researchers are frequently faced with a choice between several computational methods for performing data analyses.
-Benchmarking studies aim to rigorously compare the performance of different methods using well-characterized benchmark datasets, to determine the strengths of each method or to provide recommendations regarding suitable choices of methods for an analysis.
-However, benchmarking studies must be carefully designed and implemented to provide accurate, unbiased, and informative results.
-Here, we summarize key practical guidelines and recommendations for performing high-quality benchmarking analyses, based on our experiences in computational biology.},
-	file = {/home/rcannood/Zotero/storage/83LMG2SG/Weber et al. - 2019 - Essential guidelines for computational method benc.pdf;/home/rcannood/Zotero/storage/GL7G9MIR/s13059-019-1738-8.html},
-	journal = {Genome Biology},
-	number = {1},
-}
-
-@Article{wickham_welcometidyverse_2019,
-	title = {Welcome to the {{Tidyverse}}},
-	author = {Hadley Wickham and Mara Averick and Jennifer Bryan and Winston Chang and Lucy McGowan and Romain Fran{\c c}ois and Garrett Grolemund and Alex Hayes and Lionel Henry and Jim Hester and Max Kuhn and Thomas Pedersen and Evan Miller and Stephan Bache and Kirill M{\"u}ller and Jeroen Ooms and David Robinson and Dana Seidel and Vitalie Spinu and Kohske Takahashi and Davis Vaughan and Claus Wilke and Kara Woo and Hiroaki Yutani},
-	year = {2019},
-	month = {nov},
-	doi = {10.21105/joss.01686},
-	abstract = {Wickham et al., (2019). Welcome to the Tidyverse. Journal of Open Source Software, 4(43), 1686, https://doi.org/10.21105/joss.01686},
-	file = {/home/rcannood/Zotero/storage/TVYSNC4J/Wickham et al. - 2019 - Welcome to the Tidyverse.pdf;/home/rcannood/Zotero/storage/EAQ2VYF8/joss.html},
-	language = {en},
-}
-
-@Article{xu_regulationbifurcatingcell_2015,
-	title = {Regulation of Bifurcating {{B}} Cell Trajectories by Mutual Antagonism between Transcription Factors {{IRF4}} and {{IRF8}}},
-	author = {Heping Xu and Virendra K Chaudhri and Zhiguo Wu and Konstantinos Biliouris and Krista Dienger-Stambaugh and Yrina Rochman and Harinder Singh},
-	year = {2015},
-	month = {dec},
-	volume = {16},
-	pages = {1274--1281},
-	abstract = {Upon recognition of antigen, B cells undertake a bifurcated response in which some cells rapidly differentiate into plasmablasts while others undergo affinity maturation in germinal centers (GCs).
-Here we identified a double-negative feedback loop between the transcription factors IRF4 and IRF8 that regulated the initial developmental bifurcation of activated B cells as well as the GC response.
-IRF8 dampened signaling via the B cell antigen receptor (BCR), facilitated antigen-specific interaction with helper T cells, and promoted antibody affinity maturation while antagonizing IRF4-driven differentiation of plasmablasts.
-Genomic analysis revealed concentration-dependent actions of IRF4 and IRF8 in regulating distinct gene-expression programs.
-Stochastic modeling suggested that the double-negative feedback was sufficient to initiate bifurcation of the B cell developmental trajectories.},
-	journal = {Nat. Immunol.},
-	number = {12},
-}
-
-@Article{yosef_dynamicregulatorynetwork_2013,
-	title = {Dynamic Regulatory Network Controlling \{\vphantom\}{{TH17}}\vphantom\{\} Cell Differentiation},
-	author = {Nir Yosef and Alex K Shalek and Jellert T Gaublomme and Hulin Jin and Youjin Lee and Amit Awasthi and Chuan Wu and Katarzyna Karwacz and Sheng Xiao and Marsela Jorgolli and David Gennert and Rahul Satija and Arvind Shakya and Diana Y Lu and John J Trombetta and Meenu R Pillai and Peter J Ratcliffe and Mathew L Coleman and Mark Bix and Dean Tantin and Hongkun Park and Vijay K Kuchroo and Aviv Regev},
-	year = {2013},
-	volume = {496},
-	pages = {461--468},
-	abstract = {Despite their importance, the molecular circuits that control the differentiation of naive T cells remain largely unknown.
-Recent studies that reconstructed regulatory networks in mammalian cells have focused on short-term responses and relied on perturbation-based approaches that cannot be readily applied to primary T cells.
-Here we combine transcriptional profiling at high temporal resolution, novel computational algorithms, and innovative nanowire-based perturbation tools to systematically derive and experimentally validate a model of the dynamic regulatory network that controls the differentiation of mouse TH17 cells, a proinflammatory T-cell subset that has been implicated in the pathogenesis of multiple autoimmune diseases.
-The TH17 transcriptional network consists of two self-reinforcing, but mutually antagonistic, modules, with 12 novel regulators, the coupled action of which may be essential for maintaining the balance between TH17 and other CD4(+) T-cell subsets.
-Our study identifies and validates 39 regulatory factors, embeds them within a comprehensive temporal network and reveals its organizational principles; it also highlights novel drug targets for controlling TH17 cell differentiation.},
-	journal = {Nature},
-	number = {7446},
-}
-
-@Article{zappia_splattersimulationsinglecell_2017,
-	title = {Splatter: {{Simulation}} of {{Single}}-{{Cell RNA Sequencing Data}}},
-	shorttitle = {Splatter},
-	author = {Luke Zappia and Belinda Phipson and Alicia Oshlack},
-	year = {2017},
-	month = {sep},
-	volume = {18},
-	pages = {174},
-	issn = {1474-760X},
-	doi = {10.1186/s13059-017-1305-0},
-	abstract = {As single-cell RNA sequencing (scRNA-seq) technologies have rapidly developed, so have analysis methods.
-Many methods have been tested, developed, and validated using simulated datasets.
-Unfortunately, current simulations are often poorly documented, their similarity to real data is not demonstrated, or reproducible code is not available.
-Here, we present the Splatter Bioconductor package for simple, reproducible, and well-documented simulation of scRNA-seq data.
-Splatter provides an interface to multiple simulation methods including Splat, our own simulation, based on a gamma-Poisson distribution.
-Splat can simulate single populations of cells, populations with multiple cell types, or differentiation paths.},
-	journal = {Genome Biology},
-	keywords = {RNA-seq,Simulation,Single-cell,Software},
-}
-
-@Article{zeisel_coupledpremrnamrna_2011,
-	title = {Coupled Pre-{{mRNA}} and {{mRNA}} Dynamics Unveil Operational Strategies Underlying Transcriptional Responses to Stimuli},
-	author = {Amit Zeisel and Wolfgang J K{\"o}stler and Natali Molotski and Jonathan M Tsai and Rita Krauthgamer and Jasmine Jacob-Hirsch and Gideon Rechavi and Yoav Soen and Steffen Jung and Yosef Yarden and Eytan Domany},
-	year = {2011},
-	month = {jan},
-	volume = {7},
-	pages = {529},
-	issn = {1744-4292},
-	doi = {10.1038/msb.2011.62},
-	abstract = {Transcriptional responses to extracellular stimuli involve tuning the rates of transcript production and degradation.
-Here, we show that the time-dependent profiles of these rates can be inferred from simultaneous measurements of precursor mRNA (pre-mRNA) and mature mRNA profiles.
-Transcriptome-wide measurements demonstrate that genes with similar mRNA profiles often exhibit marked differences in the amplitude and onset of their production rate.
-The latter is characterized by a large dynamic range, with a group of genes exhibiting an unexpectedly strong transient production overshoot, thereby accelerating their induction and, when combined with time-dependent degradation, shaping transient responses with precise timing and amplitude.},
-	file = {/home/rcannood/Zotero/storage/QRQW86Y9/Zeisel et al. - 2011 - Coupled pre-mRNA and mRNA dynamics unveil operatio.pdf;/home/rcannood/Zotero/storage/R8ZG26IL/msb.2011.html},
-	journal = {Molecular Systems Biology},
-	keywords = {half-life,operational strategy,pre-mRNA,transcriptional response},
-	number = {1},
-}
-
-@Article{zhang_simulatingmultiplefaceted_2019,
-	title = {Simulating Multiple Faceted Variability in Single Cell {{RNA}} Sequencing},
-	author = {Xiuwei Zhang and Chenling Xu and Nir Yosef},
-	year = {2019},
-	month = {jun},
-	volume = {10},
-	pages = {1--16},
-	issn = {2041-1723},
-	doi = {10.1038/s41467-019-10500-w},
-	abstract = {Simulated single cell RNA sequencing data is useful for method development and comparison.
-Here, the authors developed SymSim, a simulator that explicitly models the main factors of variation in single cell data.},
-	copyright = {2019 The Author(s)},
-	file = {/home/rcannood/Zotero/storage/YLQYCW6E/Zhang et al. - 2019 - Simulating multiple faceted variability in single .pdf;/home/rcannood/Zotero/storage/EXJ96Y59/s41467-019-10500-w.html},
-	journal = {Nature Communications},
-	language = {en},
-	number = {1},
-}
-\end{filecontents}
-
 \end{document}

### 2021-05-12 manuscript/render_manuscript.R

@@ -31,10 +31,15 @@ knitcitations::cite_options(check.entries = FALSE)
 bib <- knitcitations::read.bibtex("manuscript/library.bib")
 knitcitations::write.bibtex(bib[bib_keys], file = "manuscript/knitcitations.bib")
 
+
 # render the manuscript
 render("manuscript/manuscript.Rmd")
 
+bib_replace <- paste(c("\\addbibresource{\\jobname.bib}", "\\begin{filecontents}{\\jobname.bib}", readr::read_lines("manuscript/knitcitations.bib"), "\\end{filecontents}"), collapse = "\n")
 
+read_lines("manuscript/manuscript.tex") %>%
+  gsub("\\addbibresource{library.bib}", bib_replace, ., fixed = TRUE) %>%
+  write_lines("manuscript/manuscript.tex")
 
 render("manuscript/supplementary_files.Rmd")
 

### 2021-05-12 manuscript/v5/manuscript.tex
update for v5

--- 2021-05-12 manuscript/manuscript.tex
@@
-% Options for packages loaded elsewhere
-\PassOptionsToPackage{unicode}{hyperref}
-\PassOptionsToPackage{hyphens}{url}
-%
-\documentclass[
-  table,
-  10pt,
-  a4paper]{article}
+\documentclass[10pt, a4paper]{article}
 \usepackage{amsmath,amssymb}
 \usepackage{lmodern}
-\usepackage{ifxetex,ifluatex}
-\ifnum 0\ifxetex 1\fi\ifluatex 1\fi=0 % if pdftex
 \usepackage[T1]{fontenc}
 \usepackage[utf8]{inputenc}
-  \usepackage{textcomp} % provide euro and other symbols
-\else % if luatex or xetex
-  \usepackage{unicode-math}
-  \defaultfontfeatures{Scale=MatchLowercase}
-  \defaultfontfeatures[\rmfamily]{Ligatures=TeX,Scale=1}
-\fi
-% Use upquote if available, for straight quotes in verbatim environments
-\IfFileExists{upquote.sty}{\usepackage{upquote}}{}
-\IfFileExists{microtype.sty}{% use microtype if available
-  \usepackage[]{microtype}
-  \UseMicrotypeSet[protrusion]{basicmath} % disable protrusion for tt fonts
-}{}
-\makeatletter
-\@ifundefined{KOMAClassName}{% if non-KOMA class
-  \IfFileExists{parskip.sty}{%
-    \usepackage{parskip}
-  }{% else
-    \setlength{\parindent}{0pt}
-    \setlength{\parskip}{6pt plus 2pt minus 1pt}}
-}{% if KOMA class
-  \KOMAoptions{parskip=half}}
-\makeatother
-\usepackage{xcolor}
-\IfFileExists{xurl.sty}{\usepackage{xurl}}{} % add URL line breaks if available
-\IfFileExists{bookmark.sty}{\usepackage{bookmark}}{\usepackage{hyperref}}
+\usepackage{textcomp}
+
+\usepackage{hyperref}
 \hypersetup{
   pdftitle={dyngen: a multi-modal simulator for spearheading new single-cell omics analyses},
-  pdfauthor={Robrecht Cannoodt,1,2,3; Wouter Saelens,1,2; Louise Deconinck1,2; Yvan Saeys1,2,},
-  hidelinks,
-  pdfcreator={LaTeX via pandoc}}
-\urlstyle{same} % disable monospaced font for URLs
+  pdfauthor={Robrecht Cannoodt; Wouter Saelens; Louise Deconinck; Yvan Saeys},
+  hidelinks}
 \usepackage[margin=1in]{geometry}
 \usepackage{graphicx}
-\makeatletter
-\def\maxwidth{\ifdim\Gin@nat@width>\linewidth\linewidth\else\Gin@nat@width\fi}
-\def\maxheight{\ifdim\Gin@nat@height>\textheight\textheight\else\Gin@nat@height\fi}
-\makeatother
-% Scale images if necessary, so that they will not overflow the page
-% margins by default, and it is still possible to overwrite the defaults
-% using explicit options in \includegraphics[width, height, ...]{}
-\setkeys{Gin}{width=\maxwidth,height=\maxheight,keepaspectratio}
-% Set default figure placement to htbp
-\makeatletter
-\def\fps@figure{htbp}
-\makeatother
-\setlength{\emergencystretch}{3em} % prevent overfull lines
-\providecommand{\tightlist}{%
-  \setlength{\itemsep}{0pt}\setlength{\parskip}{0pt}}
-\setcounter{secnumdepth}{-\maxdimen} % remove section numbering
-\usepackage{tcolorbox}
-\usepackage{colortbl}
-\usepackage{booktabs}
-\usepackage{tabularx}
-\usepackage{fontspec}
-\usepackage{pifont}
-\usepackage{float}
-\usepackage{caption}
-\captionsetup[table]{position=bottom}
-\setmainfont [Path = fonts/,
-  UprightFont = *-300,
-  ItalicFont = *-300-Italic,
-  BoldFont = *-700,
-  BoldItalicFont = *-700-Italic
-]{MuseoSans}
-
-\ifluatex
-  \usepackage{selnolig}  % disable illegal ligatures
-\fi
-\usepackage[sorting=none,url=false,backend=bibtex]{biblatex}
-
-\usepackage{filecontents}
-
-\bibliography{\jobname}
-
-\title{dyngen: a multi-modal simulator for spearheading new single-cell
-omics analyses}
-\author{Robrecht Cannoodt\textsuperscript{$\dagger{}$,1,2,3} \and Wouter
-Saelens\textsuperscript{$\dagger{}$,1,2} \and Louise
-Deconinck\textsuperscript{1,2} \and Yvan Saeys\textsuperscript{1,2,*}}
+
+\title{dyngen: a multi-modal simulator for spearheading new single-cell omics analyses}
+\author{Robrecht Cannoodt\textsuperscript{$\dagger{}$,1,2,3} \and Wouter Saelens\textsuperscript{$\dagger{}$,1,2,4} \and Louise Deconinck\textsuperscript{1,2} \and Yvan Saeys\textsuperscript{1,2,*}}
 \date{12 May 2021}
@@
 
-\textsuperscript{$\dagger{}$}
-These authors contributed equally to this work.
-
 \textsuperscript{1} Data Mining and Modelling for Biomedicine group, VIB
-Center for Inflammation Research, Ghent, Belgium\\
+Center for Inflammation Research, Ghent, Belgium
+
 \textsuperscript{2} Department of Applied Mathematics, Computer Science,
-and Statistics, Ghent University, Ghent, Belgium\\
+and Statistics, Ghent University, Ghent, Belgium
+
 \textsuperscript{3} Data Intuitive, Lebbeke, Belgium
 
-\textsuperscript{*} Correspondence:
-\href{mailto:yvan.saeys@ugent.be}{Yvan Saeys
-\textless{}\href{mailto:yvan.saeys@ugent.be}{\nolinkurl{yvan.saeys@ugent.be}}\textgreater{}}
-
-\hypertarget{abstract}{%
-\section{Abstract}\label{abstract}}
+\textsuperscript{4} Institute of Bioengineering, School of Life
+Sciences, École Polytechnique Fédérale de Lausanne (EPFL), Lausanne,
+Switzerland
+
+\textsuperscript{$\dagger{}$} These authors contributed equally to this work.
+
+\textsuperscript{*} Correspondence: \texttt{yvan.saeys@ugent.be}
+
+\section{Abstract}\label{abstract}
 
@@
 
-\hypertarget{introduction}{%
-\section{Introduction}\label{introduction}}
+\section{Introduction}\label{introduction}
 
@@
 aim to reconstruct this ground truth
-\autocite{zappia_splattersimulationsinglecell_2017}. In addition,
+\cite{zappia_splattersimulationsinglecell_2017}. In addition,
 simulation engines are more flexible when it comes to stress-testing
@@
 can act as digital twins, offering unlimited experimentation \emph{in
-silico} \autocite{bjornsson_digitaltwinspersonalize_2019}.
+silico} \cite{bjornsson_digitaltwinspersonalize_2019}.
 
@@
 ordinary differential equations (ODE)
-\autocite{roy_systemgeneratingtranscription_2008,hache_gengesystematicgeneration_2009,schaffter_genenetweaversilicobenchmark_2011,vandenbulcke_syntrengeneratorsynthetic_2006}.
+\cite{roy_systemgeneratingtranscription_2008,hache_gengesystematicgeneration_2009,schaffter_genenetweaversilicobenchmark_2011,vandenbulcke_syntrengeneratorsynthetic_2006}.
 These methods have been instrumental in performing benchmarking studies
-\autocite{prill_rigorousassessmentsystems_2010,marbach_revealingstrengthsweaknesses_2010,marbach_wisdomcrowdsrobust_2012}.
+\cite{prill_rigorousassessmentsystems_2010,marbach_revealingstrengthsweaknesses_2010,marbach_wisdomcrowdsrobust_2012}.
 However, the advent of single-cell omics introduced several new types of
@@
 inference) which exploit the higher resolution of single-cell versus
-bulk omics \autocite{luecken_currentbestpractices_2019}. In addition,
+bulk omics \cite{luecken_currentbestpractices_2019}. In addition,
 the data characteristics of single-cell omics are vastly different from
@@
 dropout rate, but also a high number of profiles
-\autocite{vallejos_normalizingsinglecellrna_2017}. The low library
+\cite{vallejos_normalizingsinglecellrna_2017}. The low library
 sizes, in particular, are problematic as ODEs are ill-suited for
 performing low-molecule simulations
-\autocite{gillespie_exactstochasticsimulation_1977}. This necessitates
+\cite{gillespie_exactstochasticsimulation_1977}. This necessitates
 the development of new single-cell simulators.
@@
 procedures from single-cell omics protocols. Simulators such as Splatter
-\autocite{zappia_splattersimulationsinglecell_2017}, powsimR
-\autocite{vieth_powsimrpoweranalysis_2017}, PROSSTT
-\autocite{papadopoulos_prossttprobabilisticsimulation_2019} and SymSim
-\autocite{zhang_simulatingmultiplefaceted_2019}) have already been
+\cite{zappia_splattersimulationsinglecell_2017}, powsimR
+\cite{vieth_powsimrpoweranalysis_2017}, PROSSTT
+\cite{papadopoulos_prossttprobabilisticsimulation_2019} and SymSim
+\cite{zhang_simulatingmultiplefaceted_2019}) have already been
 widely used to compare single-cell methods
-\autocite{street_slingshotcelllineage_2018,parra_reconstructingcomplexlineage_2019,lummertzdarocha_reconstructioncomplexsinglecell_2018,lin_scclassifysamplesize_2020}
+\cite{street_slingshotcelllineage_2018,parra_reconstructingcomplexlineage_2019,lummertzdarocha_reconstructioncomplexsinglecell_2018,lin_scclassifysamplesize_2020}
 and perform independent benchmarks
-\autocite{duo_systematicperformanceevaluation_2018,saelens_comparisonsinglecelltrajectory_2019,soneson_biasrobustnessscalability_2018}.
+\cite{duo_systematicperformanceevaluation_2018,saelens_comparisonsinglecelltrajectory_2019,soneson_biasrobustnessscalability_2018}.
 However, by focusing more on simulating the single-cell omics protocol
@@
 stochastic simulation algorithm
-\autocite{gillespie_exactstochasticsimulation_1977}, which is designed
+\cite{gillespie_exactstochasticsimulation_1977}, which is designed
 to work well in low-molecule simulations. Finally, real reference
@@
 
-\begin{figure}[t!]
+\begin{figure}[htbp]
     \centering
-    \includegraphics[width=\linewidth]{result_files/summary/figure_1_edited.pdf}
+    \includegraphics[width=\linewidth]{figure_1.pdf}
     \caption{
@@
 
-\hypertarget{results}{%
-\section{Results}\label{results}}
+\section{Results}\label{results}
 
@@
 assessing many more quantitative and qualitative aspects of each method
-\autocite{weber_essentialguidelinescomputational_2019}.
-
-\begin{figure}[t!]
+\cite{weber_essentialguidelinescomputational_2019}.
+
+\begin{figure}[htbp]
     \centering
-    \includegraphics[width=\linewidth]{result_files/summary/figure_2.pdf}
+    \includegraphics[width=\linewidth]{figure_2.pdf}
     \caption{
@@
 as well. Dynamic Time Warping (DTW)
-\autocite{giorgino_computingvisualizingdynamic_2009} is a method
+\cite{giorgino_computingvisualizingdynamic_2009} is a method
 designed for aligning temporal sequences for speech recognition but has
@@
 biological processes
-\autocite{cacchiarelli_aligningsinglecelldevelopmental_2018,kanton_organoidsinglecellgenomic_2019,mcfaline-figueroa_pooledsinglecellgenetic_2019,alpert_alignmentsinglecelltrajectories_2018}.
-cellAlign \autocite{alpert_alignmentsinglecelltrajectories_2018} uses
+\cite{cacchiarelli_aligningsinglecelldevelopmental_2018,kanton_organoidsinglecellgenomic_2019,mcfaline-figueroa_pooledsinglecellgenetic_2019,alpert_alignmentsinglecelltrajectories_2018}.
+cellAlign \cite{alpert_alignmentsinglecelltrajectories_2018} uses
 DTW to perform trajectory alignment, but also includes interpolation and
@@
 trajectories
-\autocite{zeisel_coupledpremrnamrna_2011,lamanno_rnavelocitysingle_2018}.
+\cite{zeisel_coupledpremrnamrna_2011,lamanno_rnavelocitysingle_2018}.
 Already two algorithms are currently available for estimating the RNA
 velocity vector from spliced and unspliced counts: velocyto
-\autocite{lamanno_rnavelocitysingle_2018} and scvelo
-\autocite{bergen_generalizingrnavelocity_2020}. Yet, to date, no
+\cite{lamanno_rnavelocitysingle_2018} and scvelo
+\cite{bergen_generalizingrnavelocity_2020}. Yet, to date, no
 quantitative assessment of their accuracy has been performed, mainly due
@@
 been developed
-\autocite{aibar_scenicsinglecellregulatory_2017,kuijjer_estimatingsamplespecificregulatory_2019,liu_personalizedcharacterizationdiseases_2016},
+\cite{aibar_scenicsinglecellregulatory_2017,kuijjer_estimatingsamplespecificregulatory_2019,liu_personalizedcharacterizationdiseases_2016},
 a quantitative assessment of their performance is until now lacking.
@@
 (Supplementary Fig.~3): LIONESS
-\autocite{kuijjer_estimatingsamplespecificregulatory_2019}, SSN
-\autocite{liu_personalizedcharacterizationdiseases_2016} and SCENIC
-\autocite{aibar_scenicsinglecellregulatory_2017}. For each dataset, we
+\cite{kuijjer_estimatingsamplespecificregulatory_2019}, SSN
+\cite{liu_personalizedcharacterizationdiseases_2016} and SCENIC
+\cite{aibar_scenicsinglecellregulatory_2017}. For each dataset, we
 computed the mean AUROC and AUPR scores of the individual cells.
@@
 network inference (e.g.~GENIE3
-\autocite{huynh-thu_inferringregulatorynetworks_2010} or pySCENIC's
-GRNBoost2 \autocite{moerman_grnboost2arboretoefficient_2019}) could
+\cite{huynh-thu_inferringregulatorynetworks_2010} or pySCENIC's
+GRNBoost2 \cite{moerman_grnboost2arboretoefficient_2019}) could
 significantly increase the performance obtained by LIONESS.
 
-\hypertarget{discussion}{%
-\section{Discussion}\label{discussion}}
+\section{Discussion}\label{discussion}
 
@@
 simulators have done in the past
-\autocite{schaffter_genenetweaversilicobenchmark_2011,ewing_combiningtumorgenome_2015}.
+\cite{schaffter_genenetweaversilicobenchmark_2011,ewing_combiningtumorgenome_2015}.
 Additionally, one could anticipate technological developments in
@@
 division, intercellular communication, and migration
-\autocite{smith_spatialstochasticintracellular_2019}.
+\cite{smith_spatialstochasticintracellular_2019}.
 
@@
 
-\hypertarget{sec:dyngen-methods}{%
-\section{Methods}\label{sec:dyngen-methods}}
+\section{Methods}\label{sec:dyngen-methods}
 
@@
 
-\hypertarget{sec:dyngen-modules}{%
-\subsection{Defining the module network}\label{sec:dyngen-modules}}
+\subsection{Defining the module network}\label{sec:dyngen-modules}
 
@@
 antagonism, where two genes strongly repress each other
-\autocite{rekhtman_directinteractionhematopoietic_1999,xu_regulationbifurcatingcell_2015},
+\cite{rekhtman_directinteractionhematopoietic_1999,xu_regulationbifurcatingcell_2015},
 forcing one of the two to become inactive
-\autocite{graf_forcingcellschange_2009}. Such mutual antagonism can be
+\cite{graf_forcingcellschange_2009}. Such mutual antagonism can be
 modelled and simulated
-\autocite{wang_quantifyingwaddingtonlandscape_2011,ferrell_bistabilitybifurcationswaddington_2012}.
+\cite{wang_quantifyingwaddingtonlandscape_2011,ferrell_bistabilitybifurcationswaddington_2012}.
 Although the two-gene model is simple and elegant, the reality is
 frequently more complex, with multiple genes (grouped into modules)
-repressing each other \autocite{yosef_dynamicregulatorynetwork_2013}.
+repressing each other \cite{yosef_dynamicregulatorynetwork_2013}.
 
@@
 will be allowed to generate expression of its genes during an initial
-warm-up phase (See section \ref{sec:dyngen-simcell}). At the end of the
+warm-up phase. At the end of the
 dyngen process, cells will not be sampled from the burn phase
@@
 
-\hypertarget{sec:dyngen-grn}{%
-\subsection{Generating the gene regulatory
-network}\label{sec:dyngen-grn}}
+\subsection{Generating the gene regulatory network}\label{sec:dyngen-grn}
 
@@
 sample the interactions between target genes, one of the many FANTOM5
-\autocite{lizio_gatewaysfantom5promoter_2015} GRNs is sampled. The currently
-existing TFs are mapped to regulators in the FANTOM5 GRN. The targets
-are drawn from the FANTOM5 GRN weighted by their page rank value, to
-create an induced GRN. For each target, at most \(x\) regulators are
-sampled from the induced FANTOM5 GRN (default \(x=5\)). The interactions
-connecting a target gene and its regulators are added to the GRN.
+\cite{lizio_gatewaysfantom5promoter_2015} GRNs is sampled. The
+currently existing TFs are mapped to regulators in the FANTOM5 GRN. The
+targets are drawn from the FANTOM5 GRN weighted by their page rank
+value, to create an induced GRN. For each target, at most \(x\)
+regulators are sampled from the induced FANTOM5 GRN (default \(x=5\)).
+The interactions connecting a target gene and its regulators are added
+to the GRN.
 
@@
 
-\hypertarget{sec:dyngen-reactions}{%
-\subsection{Convert gene regulatory network to a set of
-reactions}\label{sec:dyngen-reactions}}
+\subsection{Convert gene regulatory network to a set of reactions}\label{sec:dyngen-reactions}
 
@@
 the abundance levels of molecules and how likely it will occur. Gibson
-and Bruck \autocite{gibson_probabilisticmodelprokaryotic_2000} provide a
+and Bruck \cite{gibson_probabilisticmodelprokaryotic_2000} provide a
 good introduction to modelling gene regulation with stochastic
@@
 measured of 5000 mammalian genes
-\autocite{schwanhausser_globalquantificationmammalian_2011}, except that
+\cite{schwanhausser_globalquantificationmammalian_2011}, except that
 the transcription rate has been amplified by a factor of 10.
@@
 thermodynamic models of gene regulation
-\autocite{schilstra_biologicgeneexpression_2008}, in which the promoter
+\cite{schilstra_biologicgeneexpression_2008}, in which the promoter
 of \(G\) can be bound or unbound by a set of \(N\) transcription factors
@@
 
-\hypertarget{propensity-of-transcription-when-n1}{%
-\subsubsection{\texorpdfstring{Propensity of transcription when
-\(N=1\)}{Propensity of transcription when N=1}}\label{propensity-of-transcription-when-n1}}
+\subsubsection{\texorpdfstring{Propensity of transcription when \(N=1\)}{Propensity of transcription when N=1}}\label{propensity-of-transcription-when-n1}
 
@@
 
-\hypertarget{propensity-of-transcription-when-n2}{%
-\subsubsection{\texorpdfstring{Propensity of transcription when
-\(N=2\)}{Propensity of transcription when N=2}}\label{propensity-of-transcription-when-n2}}
+\subsubsection{\texorpdfstring{Propensity of transcription when \(N=2\)}{Propensity of transcription when N=2}}\label{propensity-of-transcription-when-n2}
 
@@
 
-\hypertarget{propensity-of-transcription-for-an-arbitrary-n}{%
-\subsubsection{\texorpdfstring{Propensity of transcription for an
-arbitrary
-\(N\)}{Propensity of transcription for an arbitrary N}}\label{propensity-of-transcription-for-an-arbitrary-n}}
+\subsubsection{\texorpdfstring{Propensity of transcription for an arbitrary \(N\)}{Propensity of transcription for an arbitrary N}}\label{propensity-of-transcription-for-an-arbitrary-n}
 
@@
 
-\hypertarget{propensity-of-transcription-for-a-large-n}{%
-\subsubsection{\texorpdfstring{Propensity of transcription for a large
-\(N\)}{Propensity of transcription for a large N}}\label{propensity-of-transcription-for-a-large-n}}
+\subsubsection{\texorpdfstring{Propensity of transcription for a large \(N\)}{Propensity of transcription for a large N}}\label{propensity-of-transcription-for-a-large-n}
 
@@
 
-\hypertarget{independence-synergism-and-basal-expression}{%
-\subsubsection{Independence, synergism and basal
-expression}\label{independence-synergism-and-basal-expression}}
+\subsubsection{Independence, synergism and basal expression}\label{independence-synergism-and-basal-expression}
 
@@
 
-\hypertarget{sec:dyngen-simcell}{%
-\subsection{Simulate single cells}\label{sec:dyngen-simcell}}
+\subsection{Simulate single cells}\label{dyngen-simcell}
 
 dyngen uses Gillespie's stochastic simulation algorithm (SSA)
-\autocite{gillespie_exactstochasticsimulation_1977} to simulate dynamic
+\cite{gillespie_exactstochasticsimulation_1977} to simulate dynamic
 processes. An SSA simulation is an iterative process where at each
@@
 
-\hypertarget{sec:dyngen-experiment}{%
-\subsection{Simulate experiment}\label{sec:dyngen-experiment}}
+\subsection{Simulate experiment}\label{sec:dyngen-experiment}
 
@@
 
-\hypertarget{sample-cells}{%
-\subsubsection{Sample cells}\label{sample-cells}}
+\subsubsection{Sample cells}\label{sample-cells}
 
@@
 
-\hypertarget{sample-molecules}{%
-\subsubsection{Sample molecules}\label{sample-molecules}}
+\subsubsection{Sample molecules}\label{sample-molecules}
 
@@
 of single-cell RNA-seq datasets
-\autocite{cannoodt_singlecellomicsdatasets_2018}. For each \emph{in
+\cite{cannoodt_singlecellomicsdatasets_2018}. For each \emph{in
 silico} cell \(i\), draw its library size \(ls_i\) from the distribution
@@
 
-\hypertarget{sec:dyngen-batcheffect}{%
-\subsection{Simulating batch effects}\label{sec:dyngen-batcheffect}}
+\subsection{Simulating batch effects}\label{sec:dyngen-batcheffect}
 
@@
 
-\hypertarget{sec:dyngen-groundtruth}{%
-\subsection{Determining the ground-truth
-trajectory}\label{sec:dyngen-groundtruth}}
+\subsection{Determining the ground-truth trajectory}\label{sec:dyngen-groundtruth}
 
@@
 
-\hypertarget{sec:dyngen-extractgrn}{%
-\subsection{Determining the cell-specific ground-truth regulatory
-network}\label{sec:dyngen-extractgrn}}
+\subsection{Determining the cell-specific ground-truth regulatory network}\label{sec:dyngen-extractgrn}
 
@@
 
-\begin{eqnarray*}
+\begin{eqnarray}
   \text{regeffect}_G & = \frac{\text{proptrans}_G(S) - \text{proptrans}_G(S[\text{z}_{T} \leftarrow 0])}{\text{xpr}_{G}}
-\end{eqnarray*}
+\end{eqnarray}
 
@@
 
-\hypertarget{sec:dyngen-nicompare}{%
-\subsection{Comparison of cell-specific network inference
-methods}\label{sec:dyngen-nicompare}}
+\subsection{Comparison of cell-specific network inference methods}\label{sec:dyngen-nicompare}
 
@@
 We selected three cell-specific NI methods: SCENIC
-\autocite{aibar_scenicsinglecellregulatory_2017}, LIONESS
-\autocite{kuijjer_estimatingsamplespecificregulatory_2015,kuijjer_estimatingsamplespecificregulatory_2019},
-and SSN \autocite{liu_personalizedcharacterizationdiseases_2016}.
-
-LIONESS \autocite{kuijjer_estimatingsamplespecificregulatory_2019} runs
+\cite{aibar_scenicsinglecellregulatory_2017}, LIONESS
+\cite{kuijjer_estimatingsamplespecificregulatory_2015,kuijjer_estimatingsamplespecificregulatory_2019},
+and SSN \cite{liu_personalizedcharacterizationdiseases_2016}.
+
+LIONESS \cite{kuijjer_estimatingsamplespecificregulatory_2019} runs
 a NI method multiple times to construct cell-specific GRNs. LIONESS
@@
 cell-specific GRN. By default, LIONESS uses PANDA
-\autocite{glass_passingmessagesbiological_2013} to infer GRNs, but since
+\cite{glass_passingmessagesbiological_2013} to infer GRNs, but since
 dyngen does not produce motif data and motif data is required by PANDA,
 PANDA is inapplicable in this context. Instead, we used the lionessR
-\autocite{kuijjer_lionessrsinglesample_2019} implementation of LIONESS,
+\cite{kuijjer_lionessrsinglesample_2019} implementation of LIONESS,
 which uses by default the Pearson correlation as a NI method. We marked
@@
 
-SSN \autocite{liu_personalizedcharacterizationdiseases_2016} follows, in
+SSN \cite{liu_personalizedcharacterizationdiseases_2016} follows, in
 essence, the exact same methodology as LIONESS except that it
@@
 using basic R and tidyverse functions
-\autocite{wickham_welcometidyverse_2019} and marked results from this
+\cite{wickham_welcometidyverse_2019} and marked results from this
 implementation as "SSN*".
 
-SCENIC \autocite{aibar_scenicsinglecellregulatory_2017} consists of four
+SCENIC \cite{aibar_scenicsinglecellregulatory_2017} consists of four
 main steps. First, classical network inference is performed with
 stochastic gradient boosting machines using \texttt{arboreto}
-\autocite{moerman_grnboost2arboretoefficient_2019}. Second, the top 10
+\cite{moerman_grnboost2arboretoefficient_2019}. Second, the top 10
 regulators of every target gene are selected. Interactions are grouped
@@
 the Python implementation of SCENIC was used, namely pySCENIC
-\autocite{vandesande_scalablescenicworkflow_2020}. Since dyngen does not
+\cite{vandesande_scalablescenicworkflow_2020}. Since dyngen does not
 generate motif data, step 3 in this analysis is skipped.
@@
 for evaluating a predicted GRN with a ground-truth GRN
-\autocite{marbach_generatingrealisticsilico_2009}. To compare a
+\cite{marbach_generatingrealisticsilico_2009}. To compare a
 predicted cell-specific GRN with the ground-truth cell-specific GRN, the
@@
 two-sided Durbin-Conover tests
-\autocite{conover_multiplecomparisonsprocedures_1979} using
+\cite{conover_multiplecomparisonsprocedures_1979} using
 \texttt{pairwiseComparisons}
-\autocite{patil_pairwisecomparisonsmultiplepairwise_2019}. Test
+\cite{patil_pairwisecomparisonsmultiplepairwise_2019}. Test
 statistics and p-values for the all pairwise combinations are reported
@@
 testing using Holm correction
-\autocite{holm_simplesequentiallyrejective_1979}.
-
-\hypertarget{sec:dyngen-velcompare}{%
-\subsection{Comparison of RNA velocity
-methods}\label{sec:dyngen-velcompare}}
+\cite{holm_simplesequentiallyrejective_1979}.
+
+\subsection{Comparison of RNA velocity methods}\label{sec:dyngen-velcompare}
 
@@
 We applied two RNA velocity methods: velocyto
-\autocite{lamanno_rnavelocitysingle_2018}, as implemented in the
+\cite{lamanno_rnavelocitysingle_2018}, as implemented in the
 \texttt{velocyto.py} package, and scvelo method
-\autocite{bergen_generalizingrnavelocity_2020}, as implemented in the
+\cite{bergen_generalizingrnavelocity_2020}, as implemented in the
 \texttt{scvelo} package. For scvelo, we chose two parameter settings for
@@
 pairwise non-parametric paired two-sided Durbin-Conover tests
-\autocite{conover_multiplecomparisonsprocedures_1979} using
+\cite{conover_multiplecomparisonsprocedures_1979} using
 \texttt{pairwiseComparisons}
-\autocite{patil_pairwisecomparisonsmultiplepairwise_2019}. Test
+\cite{patil_pairwisecomparisonsmultiplepairwise_2019}. Test
 statistics and p-values for the all pairwise combinations are reported
@@
 testing using Holm correction
-\autocite{holm_simplesequentiallyrejective_1979}.
-
-\hypertarget{sec:dyngen-tacompare}{%
-\subsection{Comparison of trajectory
-alignment}\label{sec:dyngen-tacompare}}
+\cite{holm_simplesequentiallyrejective_1979}.
+
+ \subsection{Comparison of trajectory alignment}\label{sec:dyngen-tacompare}
 
@@
 Dynamic Time Warping (DTW)
-\autocite{giorgino_computingvisualizingdynamic_2009} and cellAlign
-\autocite{alpert_alignmentsinglecelltrajectories_2018}. DTW is designed
+\cite{giorgino_computingvisualizingdynamic_2009} and cellAlign
+\cite{alpert_alignmentsinglecelltrajectories_2018}. DTW is designed
 to align temporal sequences by dilating or contracting the sequences to
@@
 Durbin-Conover tests
-\autocite{conover_multiplecomparisonsprocedures_1979} using
+\cite{conover_multiplecomparisonsprocedures_1979} using
 \texttt{pairwiseComparisons}
-\autocite{patil_pairwisecomparisonsmultiplepairwise_2019}. Test
+\cite{patil_pairwisecomparisonsmultiplepairwise_2019}. Test
 statistics and p-values for the all pairwise combinations are reported
@@
 testing using Holm correction
-\autocite{holm_simplesequentiallyrejective_1979}.
-
-\hypertarget{data-availability}{%
-\section{Data Availability}\label{data-availability}}
+\cite{holm_simplesequentiallyrejective_1979}.
+
+\section{Data Availability}\label{data-availability}
 
@@
 
-\hypertarget{code-availability}{%
-\section{Code Availability}\label{code-availability}}
+\section{Code Availability}\label{code-availability}
 
@@
 
-\hypertarget{author-contributions}{%
-\section{Author contributions}\label{author-contributions}}
+\section{Author contributions}\label{author-contributions}
 
 \begin{itemize}
-\tightlist
-\item
-  W.S. and R.C. designed the study.
-\item
-  R.C., W.S., and L.D. performed the experiments and analysed the data.
-\item
-  R.C. and W.S. implemented the dyngen software package.
-\item
-  R.C., W.S., L.D., and Y.S. wrote the manuscript.
-\item
-  Y.S. supervised the project.
+\item W.S. and R.C. designed the study.
+\item R.C., W.S., and L.D. performed the experiments and analysed the data.
+\item R.C. and W.S. implemented the dyngen software package.
+\item R.C., W.S., L.D., and Y.S. wrote the manuscript.
+\item Y.S. supervised the project.
 \end{itemize}
 
-\printbibliography
-
-\begin{filecontents}{\jobname.bib}
-@Article{aibar_scenicsinglecellregulatory_2017,
-	title = {{{SCENIC}}: Single-Cell Regulatory Network Inference and Clustering},
-	author = {Sara Aibar and Carmen Bravo Gonz{\a'a}lez-Blas and Thomas Moerman and V{\^a}n Anh Huynh-Thu and Hana Imrichova and Gert Hulselmans and Florian Rambow and Jean-Christophe Marine and Pierre Geurts and Jan Aerts and Joost {van den Oord} and Zeynep Kalender Atak and Jasper Wouters and Stein Aerts},
-	year = {2017},
-	month = {oct},
-	issn = {1548-7091},
-	doi = {10.1038/nmeth.4463},
-	journal = {Nature Methods},
-}
-
-@Article{alpert_alignmentsinglecelltrajectories_2018,
-	title = {Alignment of Single-Cell Trajectories to Compare Cellular Expression Dynamics},
-	author = {Ayelet Alpert and Lindsay S. Moore and Tania Dubovik and Shai S. Shen-Orr},
-	year = {2018},
-	month = {apr},
-	volume = {15},
-	pages = {267--270},
-	issn = {1548-7105},
-	doi = {10.1038/nmeth.4628},
-	abstract = {cellAlign enables quantitative comparisons of expression dynamics within and between single-cell trajectories based on single-cell RNA-seq or mass cytometry data.},
-	copyright = {2018 Nature Publishing Group, a division of Macmillan Publishers Limited. All Rights Reserved.},
-	file = {/home/rcannood/Zotero/storage/QB7FH2T9/nmeth.html},
-	journal = {Nature Methods},
-	language = {en},
-	number = {4},
-}
-
-@Article{bergen_generalizingrnavelocity_2020,
-	title = {Generalizing {{RNA}} Velocity to Transient Cell States through Dynamical Modeling},
-	author = {Volker Bergen and Marius Lange and Stefan Peidli and F. Alexander Wolf and Fabian J. Theis},
-	year = {2020},
-	month = {dec},
-	volume = {38},
-	pages = {1408--1414},
-	issn = {1546-1696},
-	doi = {10.1038/s41587-020-0591-3},
-	abstract = {RNA velocity has opened up new ways of studying cellular differentiation in single-cell RNA-sequencing data.
-It describes the rate of gene expression change for an individual gene at a given time point based on the ratio of its spliced and unspliced messenger RNA (mRNA).
-However, errors in velocity estimates arise if the central assumptions of a common splicing rate and the observation of the full splicing dynamics with steady-state mRNA levels are violated.
-Here we present scVelo, a method that overcomes these limitations by solving the full transcriptional dynamics of splicing kinetics using a likelihood-based dynamical model.
-This generalizes RNA velocity to systems with transient cell states, which are common in development and in response to perturbations.
-We apply scVelo to disentangling subpopulation kinetics in neurogenesis and pancreatic endocrinogenesis.
-We infer gene-specific rates of transcription, splicing and degradation, recover each cell's position in the underlying differentiation processes and detect putative driver genes.
-scVelo will facilitate the study of lineage decisions and gene regulation.},
-	journal = {Nature Biotechnology},
-	number = {12},
-}
-
-@Article{bjornsson_digitaltwinspersonalize_2019,
-	title = {Digital Twins to Personalize Medicine},
-	author = {Bergthor Bj{\"o}rnsson and Carl Borrebaeck and Nils Elander and Thomas Gasslander and Danuta R.
-Gawel and Mika Gustafsson and Rebecka J{\"o}rnsten and Eun Jung Lee and Xinxiu Li and Sandra Lilja and David Mart\'inez-Enguita and Andreas Matussek and Per Sandstr{\"o}m and Samuel Sch{\"a}fer and Margaretha Stenmarker and X.
-F.
-Sun and Oleg Sysoev and Huan Zhang and Mikael Benson and {on behalf of the Swedish Digital Twin Consortium}},
-	year = {2019},
-	month = {dec},
-	volume = {12},
-	pages = {4},
-	issn = {1756-994X},
-	doi = {10.1186/s13073-019-0701-3},
-	abstract = {Personalized medicine requires the integration and processing of vast amounts of data.
-Here, we propose a solution to this challenge that is based on constructing Digital Twins.
-These are high-resolution models of individual patients that are computationally treated with thousands of drugs to find the drug that is optimal for the patient.},
-	journal = {Genome Medicine},
-	number = {1},
-}
-
-@Article{cacchiarelli_aligningsinglecelldevelopmental_2018,
-	title = {Aligning {{Single}}-{{Cell Developmental}} and {{Reprogramming Trajectories Identifies Molecular Determinants}} of {{Myogenic Reprogramming Outcome}}},
-	author = {Davide Cacchiarelli and Xiaojie Qiu and Sanjay Srivatsan and Anna Manfredi and Michael Ziller and Eliah Overbey and Antonio Grimaldi and Jonna Grimsby and Prapti Pokharel and Kenneth J.
-Livak and Shuqiang Li and Alexander Meissner and Tarjei S.
-Mikkelsen and John L.
-Rinn and Cole Trapnell},
-	year = {2018},
-	month = {sep},
-	volume = {7},
-	pages = {258-268.e3},
-	issn = {2405-4712},
-	doi = {10.1016/j.cels.2018.07.006},
-	abstract = {{$<$}h2{$>$}Summary{$<$}/h2{$><$}p{$>$}Cellular reprogramming through manipulation of defined factors holds great promise for large-scale production of cell types needed for use in therapy and for revealing principles of gene regulation.
-However, most reprogramming systems are inefficient, converting only a fraction of cells to the desired state.
-Here, we analyze MYOD-mediated reprogramming of human fibroblasts to myotubes, a well-characterized model system for direct conversion by defined factors, at pseudotemporal resolution using single-cell RNA-seq.
-To expose barriers to efficient conversion, we introduce a novel analytic technique, trajectory alignment, which enables quantitative comparison of gene expression kinetics across two biological processes.
-Reprogrammed cells navigate a trajectory with branch points that correspond to two alternative decision points, with cells that select incorrect branches terminating at aberrant or incomplete reprogramming outcomes.
-Analysis of these branch points revealed insulin and BMP signaling as crucial molecular determinants of reprogramming.
-Single-cell trajectory alignment enables rigorous quantitative comparisons between biological trajectories found in diverse processes in development, reprogramming, and other contexts.{$<$}/p{$>$}},
-	file = {/home/rcannood/Zotero/storage/YQH8GP2I/Cacchiarelli et al. - 2018 - Aligning Single-Cell Developmental and Reprogrammi.pdf;/home/rcannood/Zotero/storage/Q5MUK95E/S2405-4712(18)30314-4.html},
-	journal = {Cell Systems},
-	language = {English},
-	number = {3},
-	pmid = {30195438},
-}
-
-@Article{cannoodt_singlecellomicsdatasets_2018,
-	title = {Single-Cell -Omics Datasets Containing a Trajectory},
-	author = {Robrecht Cannoodt and Wouter Saelens and Helena Todorov and Yvan Saeys},
-	year = {2018},
-	month = {oct},
-	doi = {10.5281/zenodo.1211532},
-	journal = {Zenodo},
-}
-
-@TechReport{conover_multiplecomparisonsprocedures_1979,
-	title = {On Multiple-Comparisons Procedures},
-	author = {William Jay Conover and Ronald L Iman},
-	year = {1979},
-	institution = {{Technical report, Los Alamos Scientific Laboratory}},
-}
-
-@Article{duo_systematicperformanceevaluation_2018,
-	title = {A Systematic Performance Evaluation of Clustering Methods for Single-Cell {{RNA}}-Seq Data},
-	author = {Angelo Du{\a`o} and Mark D. Robinson and Charlotte Soneson},
-	year = {2018},
-	volume = {7},
-	pages = {1141},
-	issn = {2046-1402},
-	doi = {10.12688/f1000research.15666.2},
-	abstract = {Subpopulation identification, usually via some form of unsupervised clustering, is a fundamental step in the analysis of many single-cell RNA-seq data sets.
-This has motivated the development and application of a broad range of clustering methods, based on various underlying algorithms.
-Here, we provide a systematic and extensible performance evaluation of 14 clustering algorithms implemented in R, including both methods developed explicitly for scRNA-seq data and more general-purpose methods.
-The methods were evaluated using nine publicly available scRNA-seq data sets as well as three simulations with varying degree of cluster separability.
-The same feature selection approaches were used for all methods, allowing us to focus on the investigation of the performance of the clustering algorithms themselves.
-We evaluated the ability of recovering known subpopulations, the stability and the run time and scalability of the methods.
-Additionally, we investigated whether the performance could be improved by generating consensus partitions from multiple individual clustering methods.
-We found substantial differences in the performance, run time and stability between the methods, with SC3 and Seurat showing the most favorable results.
-Additionally, we found that consensus clustering typically did not improve the performance compared to the best of the combined methods, but that several of the top-performing methods already perform some type of consensus clustering.
-All the code used for the evaluation is available on GitHub ( https://github.com/markrobinsonuzh/scRNAseq\_clustering\_comparison).
-In addition, an R package providing access to data and clustering results, thereby facilitating inclusion of new methods and data sets, is available from Bioconductor ( https://bioconductor.org/packages/DuoClustering2018).},
-	file = {/home/rcannood/Zotero/storage/ASWVFH87/Duò et al. - 2018 - A systematic performance evaluation of clustering .pdf},
-	journal = {F1000Research},
-	keywords = {Algorithms,Benchmarking,Cluster Analysis,Clustering,Clustering methods,Gene Expression Profiling,Humans,RNA-seq,RNA; Small Cytoplasmic,Sequence Analysis; RNA,Single-Cell Analysis,Single-cell RNA-seq,Software},
-	language = {eng},
-	pmcid = {PMC6134335},
-	pmid = {30271584},
-}
-
-@Article{ewing_combiningtumorgenome_2015,
-	title = {Combining Tumor Genome Simulation with Crowdsourcing to Benchmark Somatic Single-Nucleotide-Variant Detection},
-	author = {Adam D.
-Ewing and Kathleen E.
-Houlahan and Yin Hu and Kyle Ellrott and Cristian Caloian and Takafumi N.
-Yamaguchi and J.
-Christopher Bare and Christine P'ng and Daryl Waggott and Veronica Y.
-Sabelnykova and {ICGC-TCGA DREAM Somatic Mutation Calling Challenge Participants} and Liu Xi and Ninad Dewal and Yu Fan and Wenyi Wang and David Wheeler and Andreas Wilm and Grace Hui Ting and Chenhao Li and Denis Bertrand and Niranjan Nagarajan and Qing-Rong Chen and Chih-Hao Hsu and Ying Hu and Chunhua Yan and Warren Kibbe and Daoud Meerzaman and Kristian Cibulskis and Mara Rosenberg and Louis Bergelson and Adam Kiezun and Amie Radenbaugh and Anne-Sophie Sertier and Anthony Ferrari and Laurie Tonton and Kunal Bhutani and Nancy F.
-Hansen and Difei Wang and Lei Song and Zhongwu Lai and Yang Liao and Wei Shi and Jos{\a'e} Carbonell-Caballero and Joaqu\'in Dopazo and Cheryl C.
-K.
-Lau and Justin Guinney and Michael R.
-Kellen and Thea C.
-Norman and David Haussler and Stephen H.
-Friend and Gustavo Stolovitzky and Adam A.
-Margolin and Joshua M.
-Stuart and Paul C.
-Boutros},
-	year = {2015},
-	month = {jul},
-	volume = {12},
-	pages = {623--630},
-	issn = {1548-7105},
-	doi = {10.1038/nmeth.3407},
-	abstract = {The detection of somatic mutations from cancer genome sequences is key to understanding the genetic basis of disease progression, patient survival and response to therapy.
-Benchmarking is needed for tool assessment and improvement but is complicated by a lack of gold standards, by extensive resource requirements and by difficulties in sharing personal genomic information.
-To resolve these issues, we launched the ICGC-TCGA DREAM Somatic Mutation Calling Challenge, a crowdsourced benchmark of somatic mutation detection algorithms.
-Here we report the BAMSurgeon tool for simulating cancer genomes and the results of 248 analyses of three in silico tumors created with it.
-Different algorithms exhibit characteristic error profiles, and, intriguingly, false positives show a trinucleotide profile very similar to one found in human tumors.
-Although the three simulated tumors differ in sequence contamination (deviation from normal cell sequence) and in subclonality, an ensemble of pipelines outperforms the best individual pipeline in all cases.
-BAMSurgeon is available at https://github.com/adamewing/bamsurgeon/.},
-	copyright = {2015 Nature Publishing Group},
-	file = {/home/rcannood/Zotero/storage/NBPKWHK3/Ewing et al. - 2015 - Combining tumor genome simulation with crowdsourci.pdf;/home/rcannood/Zotero/storage/TDZU2TL9/nmeth.html},
-	journal = {Nature Methods},
-	language = {en},
-	number = {7},
-}
-
-@Article{ferrell_bistabilitybifurcationswaddington_2012,
-	title = {Bistability, {{Bifurcations}}, and {{Waddington}}'s {{Epigenetic Landscape}}},
-	author = {James E Ferrell},
-	year = {2012},
-	month = {jun},
-	volume = {22},
-	pages = {R458--R466},
-	issn = {0960-9822},
-	doi = {10.1016/j.cub.2012.03.045},
-	abstract = {Waddington's epigenetic landscape is probably the most famous and most powerful metaphor in developmental biology.
-Cells, represented by balls, roll downhill through a landscape of bifurcating valleys.
-Each new valley represents a possible cell fate and the ridges between the valleys maintain the cell fate once it has been chosen.
-Here I examine models of two important developmental processes \$\textbackslash backslash\$textemdash\{\} cell-fate induction and lateral inhibition \$\textbackslash backslash\$textemdash\{\} and ask whether the landscapes for these models at least qualitatively resemble Waddington's picture.
-For cell-fate induction, the answer is no.
-The commitment of a cell to a new fate corresponds to the disappearance of a valley from the landscape, not the splitting of one valley into two, and it occurs through a type of bifurcation \$\textbackslash backslash\$textemdash\{\} a saddle-node bifurcation \$\textbackslash backslash\$textemdash\{\} that possesses an intrinsic irreversibility that is missing from Waddington's picture.
-Lateral inhibition, a symmetrical cell\$\textbackslash backslash\$textendash\{\}cell competition process, corresponds better to Waddington's picture, with one valley reversibly splitting into two through a pitchfork bifurcation.
-I propose an alternative epigenetic landscape that has numerous valleys and ridges right from the start, with the process of cell-fate commitment corresponding to the irreversible disappearance of some of these valleys and ridges, via cell-fate induction, complemented by the creation of new valleys and ridges through processes like cell\$\textbackslash backslash\$textendash\{\}cell competition.},
-	file = {/home/rcannood/Zotero/storage/GK4TJ5BV/Ferrell - 2012 - Bistability, Bifurcations, and Waddington's Epigen.pdf;/home/rcannood/Zotero/storage/HUSMJA22/Ferrell - 2012 - Bistability, Bifurcations, and Waddington's Epigen.pdf;/home/rcannood/Zotero/storage/8VLPBTG2/S0960-9822(12)00332-6.html;/home/rcannood/Zotero/storage/XBTP42Y6/S0960982212003326.html},
-	journal = {Current Biology},
-	number = {11},
-}
-
-@Article{gibson_probabilisticmodelprokaryotic_2000,
-	title = {A Probabilistic Model of a Prokaryotic Gene and Its Regulation},
-	author = {Michael A. Gibson and Jehoshua Bruck},
-	year = {2000},
-	journal = {Computational Methods in Molecular Biology: From Genotype to Phenotype, MIT press, Cambridge},
-}
-
-@Article{gillespie_exactstochasticsimulation_1977,
-	title = {Exact Stochastic Simulation of Coupled Chemical Reactions},
-	author = {Daniel T. Gillespie},
-	year = {1977},
-	month = {dec},
-	volume = {81},
-	pages = {2340--2361},
-	issn = {0022-3654},
-	doi = {10.1021/j100540a008},
-	file = {/home/rcannood/Zotero/storage/ELV9AQRF/j100540a008.html},
-	journal = {The Journal of Physical Chemistry},
-	number = {25},
-}
-
-@Article{giorgino_computingvisualizingdynamic_2009,
-	title = {Computing and {{Visualizing Dynamic Time Warping Alignments}} in {{R}}: {{The}} Dtw {{Package}}},
-	author = {Toni Giorgino},
-	year = {2009},
-	month = {sep},
-	doi = {10.18637/jss.v031.i07},
-	abstract = {Dynamic time warping is a popular technique for comparing time series, providing both a distance measure that is insensitive to local compression and stretches and the warping which optimally deforms one of the two input series onto the other.
-A variety of algorithms and constraints have been discussed in the literature.
-The dtw package provides an unification of them; it allows R users to compute time series alignments mixing freely a variety of continuity constraints, restriction windows, endpoints, local distance definitions, and so on.
-The package also provides functions for visualizing alignments and constraints using several classic diagram types.},
-	journal = {Journal of Statistical Software},
-	number = {7},
-}
-
-@Article{glass_passingmessagesbiological_2013,
-	title = {Passing {{Messages}} between {{Biological Networks}} to {{Refine Predicted Interactions}}},
-	author = {Kimberly Glass and Curtis Huttenhower and John Quackenbush and Guo-Cheng Yuan},
-	year = {2013},
-	month = {may},
-	volume = {8},
-	pages = {e64832},
-	doi = {10.1371/journal.pone.0064832},
-	abstract = {Regulatory network reconstruction is a fundamental problem in computational biology.
-There are significant limitations to such reconstruction using individual datasets, and increasingly people attempt to construct networks using multiple, independent datasets obtained from complementary sources, but methods for this integration are lacking.
-We developed PANDA (Passing Attributes between Networks for Data Assimilation), a message-passing model using multiple sources of information to predict regulatory relationships, and used it to integrate protein-protein interaction, gene expression, and sequence motif data to reconstruct genome-wide, condition-specific regulatory networks in yeast as a model.
-The resulting networks were not only more accurate than those produced using individual data sets and other existing methods, but they also captured information regarding specific biological mechanisms and pathways that were missed using other methodologies.
-PANDA is scalable to higher eukaryotes, applicable to specific tissue or cell type data and conceptually generalizable to include a variety of regulatory, interaction, expression, and other genome-scale data.
-An implementation of the PANDA algorithm is available at www.sourceforge.net/projects/panda-net.},
-	journal = {PLOS ONE},
-	number = {5},
-}
-
-@Article{graf_forcingcellschange_2009,
-	title = {Forcing {{Cells}} to {{Change Lineages}}},
-	author = {Thomas Graf and Tariq Enver},
-	year = {2009},
-	month = {dec},
-	volume = {462},
-	pages = {587},
-	issn = {1476-4687},
-	doi = {10.1038/nature08533},
-	abstract = {\$\textbackslash textless\$p\$\textbackslash textgreater\$Though it is important for the success of multicellular organisms that their many different cell types remain stable once specified, advances in the production of induced pluripotent stem (iPS) cells from somatic cells by exposure to a transcription factor cocktail emphasize the potential of almost any cell to be reprogrammed.
-This raises the question of whether interconversion between cell types occurs physiologically or in the context of disease.
-Thomas Graf and Tariq Enver use examples from the blood cell system to illustrate the principles of transcription factor transdifferentiation, which they conclude is a process fundamentally similar to differentiation and different from iPS reprogramming.
-They postulate that conversions of one cell type to another can occur as a normal developmental process, and speculate on the prospects of forced lineage reprogramming in regenerative medicine.\$\textbackslash textless\$/p\$\textbackslash textgreater\$},
-	journal = {Nature},
-	number = {7273},
-}
-
-@Article{hache_gengesystematicgeneration_2009,
-	title = {{{GeNGe}}: Systematic Generation of Gene Regulatory Networks},
-	shorttitle = {{{GeNGe}}},
-	author = {Hendrik Hache and Christoph Wierling and Hans Lehrach and Ralf Herwig},
-	year = {2009},
-	month = {may},
-	volume = {25},
-	pages = {1205--1207},
-	issn = {1367-4803},
-	doi = {10.1093/bioinformatics/btp115},
-	abstract = {Summary: The analysis of gene regulatory networks (GRNs) is a central goal of bioinformatics highly accelerated by the advent of new experimental techniques, such as RNA interference.
-A battery of reverse engineering methods has been developed in recent years to reconstruct the underlying GRNs from these and other experimental data.
-However, the performance of the individual methods is poorly understood and validation of algorithmic performances is still missing to a large extent.
-To enable such systematic validation, we have developed the web application GeNGe (GEne Network GEnerator), a controlled framework for the automatic generation of GRNs.
-The theoretical model for a GRN is a non-linear differential equation system.
-Networks can be user-defined or constructed in a modular way with the option to introduce global and local network perturbations.
-Resulting data can be used, e.g.
-as benchmark data for evaluating GRN reconstruction methods or for predicting effects of perturbations as theoretical counterparts of biological experiments.Availability: Available online at http://genge.molgen.mpg.deContact:hache@molgen.mpg.deSupplementary information:Supplementary data are available at Bioinformatics online.},
-	journal = {Bioinformatics},
-	number = {9},
-}
-
-@Article{holm_simplesequentiallyrejective_1979,
-	title = {A Simple Sequentially Rejective Multiple Test Procedure},
-	author = {Sture Holm},
-	year = {1979},
-	pages = {65--70},
-	journal = {Scandinavian journal of statistics},
-}
-
-@Article{huynh-thu_inferringregulatorynetworks_2010,
-	title = {Inferring {{Regulatory Networks}} from {{Expression Data Using Tree}}-{{Based Methods}}},
-	author = {V{\^a}n Anh Huynh-Thu and Irrthum Alexandre and Wehenkel Louis and Geurts Pierre and Alexandre Irrthum and Louis Wehenkel and Pierre Geurts},
-	year = {2010},
-	month = {jan},
-	volume = {5},
-	pages = {e12776},
-	issn = {1932-6203},
-	doi = {10.1371/journal.pone.0012776},
-	abstract = {One of the pressing open problems of computational systems biology is the elucidation of the topology of genetic regulatory networks (GRNs) using high throughput genomic data, in particular microarray gene expression data.
-The Dialogue for Reverse Engineering Assessments and Methods (DREAM) challenge aims to evaluate the success of GRN inference algorithms on benchmarks of simulated data.
-In this article, we present GENIE3, a new algorithm for the inference of GRNs that was best performer in the DREAM4 In Silico Multifactorial challenge.
-GENIE3 decomposes the prediction of a regulatory network between p genes into p different regression problems.
-In each of the regression problems, the expression pattern of one of the genes (target gene) is predicted from the expression patterns of all the other genes (input genes), using tree-based ensemble methods Random Forests or Extra-Trees.
-The importance of an input gene in the prediction of the target gene expression pattern is taken as an indication of a putative regulatory link.
-Putative regulatory links are then aggregated over all genes to provide a ranking of interactions from which the whole network is reconstructed.
-In addition to performing well on the DREAM4 In Silico Multifactorial challenge simulated data, we show that GENIE3 compares favorably with existing algorithms to decipher the genetic regulatory network of Escherichia coli.
-It doesn't make any assumption about the nature of gene regulation, can deal with combinatorial and non-linear interactions, produces directed GRNs, and is fast and scalable.
-In conclusion, we propose a new algorithm for GRN inference that performs well on both synthetic and real gene expression data.
-The algorithm, based on feature selection with tree-based ensemble methods, is simple and generic, making it adaptable to other types of genomic data and interactions.},
-	file = {/home/rcannood/Zotero/storage/3I6PLCQL/Huynh-Thu et al. - 2010 - Inferring Regulatory Networks from Expression Data Using Tree-Based Methods.pdf},
-	journal = {PLoS ONE},
-	keywords = {Algorithms,Computational Biology,Computational Biology: methods,Escherichia coli,Escherichia coli: genetics,Gene Expression Regulation,Gene Regulatory Networks,Oligonucleotide Array Sequence Analysis},
-	number = {9},
-	pmid = {20927193},
-}
-
-@Article{kanton_organoidsinglecellgenomic_2019,
-	title = {Organoid Single-Cell Genomic Atlas Uncovers Human-Specific Features of Brain Development},
-	author = {Sabina Kanton and Michael James Boyle and Zhisong He and Malgorzata Santel and Anne Weigert and F{\a'a}tima Sanch\'is-Calleja and Patricia Guijarro and Leila Sidow and Jonas Simon Fleck and Dingding Han and Zhengzong Qian and Michael Heide and Wieland B.
-Huttner and Philipp Khaitovich and Svante P{\"a}{{\"a}}bo and Barbara Treutlein and J.
-Gray Camp},
-	year = {2019},
-	month = {oct},
-	volume = {574},
-	pages = {418--422},
-	issn = {1476-4687},
-	doi = {10.1038/s41586-019-1654-9},
-	abstract = {Species comparisons using single-cell transcriptomics and accessible chromatin profiling in stem cell-derived cerebral organoids are used to map dynamic gene-regulatory changes that are unique to humans.},
-	copyright = {2019 The Author(s), under exclusive licence to Springer Nature Limited},
-	file = {/home/rcannood/Zotero/storage/AVPYT7HY/s41586-019-1654-9.html},
-	journal = {Nature},
-	language = {en},
-	number = {7778},
-}
-
-@Article{kuijjer_estimatingsamplespecificregulatory_2015,
-	title = {Estimating Sample-Specific Regulatory Networks},
-	author = {Marieke Lydia Kuijjer and Matthew Tung and GuoCheng Yuan and John Quackenbush and Kimberly Glass},
-	year = {2015},
-	pages = {1--19},
-	abstract = {Biological systems are driven by intricate interactions among the complex array of molecules that comprise the cell.
-Many methods have been developed to reconstruct network models that attempt to capture those interactions.
-These methods often draw on large numbers of measured expression samples to tease out subtle signals and infer connections between genes (or gene products).
-The result is an aggregate network model representing a single estimate for edge likelihoods.
-While informative, aggregate models fail to capture the heterogeneity that is often represented in a population.
-Here we propose a method to reverse engineer sample-specific networks from aggregate network models.
-We demonstrate the accuracy and applicability of our approach in several datasets, including simulated data, microarray expression data from synchronized yeast cells, and RNA-seq data collected from human subjects.
-We show that these sample-specific networks can be used to study the evolution of network topology across time and to characterize shifts in gene regulation that may not be apparent in the expression data.
-We believe the ability to generate sample-specific networks will revolutionize the field of network biology and has the potential to usher in an era of precision network medicine.},
-	file = {/home/rcannood/Zotero/storage/YECTK66M/Kuijjer et al. - 2015 - Estimating sample-specific regulatory networks.pdf},
-}
-
-@Article{kuijjer_estimatingsamplespecificregulatory_2019,
-	title = {Estimating {{Sample}}-{{Specific Regulatory Networks}}},
-	author = {Marieke Lydia Kuijjer and Matthew George Tung and GuoCheng Yuan and John Quackenbush and Kimberly Glass},
-	year = {2019},
-	month = {mar},
-	volume = {14},
-	pages = {226--240},
-	issn = {2589-0042},
-	doi = {10.1016/j.isci.2019.03.021},
-	abstract = {Biological systems are driven by intricate interactions among molecules.
-Many methods have been developed that draw on large numbers of expression samples to infer connections between genes (or their products).
-The result is an aggregate network representing a single estimate for the likelihood of each interaction, or ``edge,'' in the network.
-Although informative, aggregate models fail to capture population heterogeneity.
-Here we propose a method to reverse engineer sample-specific networks from aggregate networks.
-We demonstrate our approach in several contexts, including simulated, yeast microarray, and human lymphoblastoid cell line RNA sequencing data.
-We use these sample-specific networks to study changes in network topology across time and to characterize shifts in gene regulation that were not apparent in the expression data.
-We believe that generating sample-specific networks will greatly facilitate the application of network methods to large, complex, and heterogeneous multi-omic datasets, supporting the emerging field of precision network medicine.,                                        \textbullet{}               We developed LIONESS to extract single-sample networks from aggregate models                                         \textbullet{}               We tested LIONESS using in silico, yeast, and human expression data                                         \textbullet{}               LIONESS-estimated networks are reproducible, accurate, and biologically meaningful                                         \textbullet{}               Single-sample network analysis highlights important biological processes                                 , Biological Sciences; Bioinformatics; Complex Systems},
-	file = {/home/rcannood/Zotero/storage/ITHUQ7NH/Kuijjer et al. - 2019 - Estimating Sample-Specific Regulatory Networks.pdf},
-	journal = {iScience},
-	pmcid = {PMC6463816},
-	pmid = {30981959},
-}
-
-@Article{kuijjer_lionessrsinglesample_2019,
-	title = {{{lionessR}}: Single Sample Network Inference in {{R}}},
-	shorttitle = {{{lionessR}}},
-	author = {Marieke L. Kuijjer and Ping-Han Hsieh and John Quackenbush and Kimberly Glass},
-	year = {2019},
-	month = {oct},
-	volume = {19},
-	pages = {1003},
-	issn = {1471-2407},
-	doi = {10.1186/s12885-019-6235-7},
-	abstract = {In biomedical research, network inference algorithms are typically used to infer complex association patterns between biological entities, such as between genes or proteins, using data from a population.
-This resulting aggregate network, in essence, averages over the networks of those individuals in the population.
-LIONESS (Linear Interpolation to Obtain Network Estimates for Single Samples) is a method that can be used together with a network inference algorithm to extract networks for individual samples in a population.
-The method's key characteristic is that, by modeling networks for individual samples in a data set, it can capture network heterogeneity in a population.
-LIONESS was originally made available as a function within the PANDA (Passing Attributes between Networks for Data Assimilation) regulatory network reconstruction framework.
-However, the LIONESS algorithm is generalizable and can be used to model single sample networks based on a wide range of network inference algorithms.},
-	file = {/home/rcannood/Zotero/storage/RS448RZF/Kuijjer et al. - 2019 - lionessR single sample network inference in R.pdf;/home/rcannood/Zotero/storage/9STH6ZYN/s12885-019-6235-7.html},
-	journal = {BMC Cancer},
-	number = {1},
-}
-
-@Article{lamanno_rnavelocitysingle_2018,
-	title = {{{RNA Velocity}} of {{Single Cells}}},
-	author = {Gioele {La Manno} and Ruslan Soldatov and Amit Zeisel and Emelie Braun and Hannah Hochgerner and Viktor Petukhov and Katja Lidschreiber and Maria E Kastriti and Peter L{\"o}nnerberg and Alessandro Furlan and Jean Fan and Lars E Borm and Zehua Liu and David {van Bruggen} and Jimin Guo and Xiaoling He and Roger Barker and Erik Sundstr{\"o}m and Gon{\c c}alo Castelo-Branco and Patrick Cramer and Igor Adameyko and Sten Linnarsson and Peter V Kharchenko},
-	year = {2018},
-	month = {aug},
-	volume = {560},
-	pages = {494--498},
-	issn = {1476-4687},
-	doi = {10.1038/s41586-018-0414-6},
-	abstract = {RNA velocity, estimated in single cells by comparison of spliced and unspliced mRNA, is a good indicator of transcriptome dynamics and will provide a useful tool for analysis of developmental lineage.},
-	journal = {Nature},
-	number = {7719},
-}
-
-@Article{lin_scclassifysamplesize_2020,
-	title = {{{scClassify}}: Sample Size Estimation and Multiscale Classification of Cells Using Single and Multiple Reference},
-	author = {Yingxin Lin and Yue Cao and Hani Jieun Kim and Agus Salim and Terence P Speed and David M Lin and Pengyi Yang and Jean Yee Hwa Yang},
-	year = {2020},
-	month = {jun},
-	volume = {16},
-	pages = {e9389},
-	publisher = {{John Wiley \& Sons, Ltd}},
-	issn = {1744-4292},
-	doi = {10.15252/msb.20199389},
-	abstract = {Abstract Automated cell type identification is a key computational challenge in single-cell RNA-sequencing (scRNA-seq) data.
-To capitalise on the large collection of well-annotated scRNA-seq datasets, we developed scClassify, a multiscale classification framework based on ensemble learning and cell type hierarchies constructed from single or multiple annotated datasets as references.
-scClassify enables the estimation of sample size required for accurate classification of cell types in a cell type hierarchy and allows joint classification of cells when multiple references are available.
-We show that scClassify consistently performs better than other supervised cell type classification methods across 114 pairs of reference and testing data, representing a diverse combination of sizes, technologies and levels of complexity, and further demonstrate the unique components of scClassify through simulations and compendia of experimental datasets.
-Finally, we demonstrate the scalability of scClassify on large single-cell atlases and highlight a novel application of identifying subpopulations of cells from the Tabula Muris data that were unidentified in the original publication.
-Together, scClassify represents state-of-the-art methodology in automated cell type identification from scRNA-seq data.},
-	journal = {Molecular Systems Biology},
-	keywords = {cell type hierarchy,cell type identification,multiscale classification,sample size estimation,single-cell},
-	number = {6},
-}
-
-@Article{liu_personalizedcharacterizationdiseases_2016,
-	title = {Personalized Characterization of Diseases Using Sample-Specific Networks},
-	author = {Xiaoping Liu and Yuetong Wang and Hongbin Ji and Kazuyuki Aihara and Luonan Chen},
-	year = {2016},
-	volume = {44},
-	pages = {e164--e164},
-	issn = {0305-1048},
-	doi = {10.1093/nar/gkw772},
-	abstract = {A complex disease generally results not from malfunction of individual molecules but from dysfunction of the relevant system or network, which dynamically changes with time and conditions.
-Thus, estimating a condition-specific network from a single sample is crucial to elucidating the molecular mechanisms of complex diseases at the system level.
-However, there is currently no effective way to construct such an individual-specific network by expression profiling of a single sample because of the requirement of multiple samples for computing correlations.
-We developed here with a statistical method, i.e.
-a sample-specific network (SSN) method, which allows us to construct individual-specific networks based on molecular expressions of a single sample.
-Using this method, we can characterize various human diseases at a network level.
-In particular, such SSNs can lead to the identification of individual-specific disease modules as well as driver genes, even without gene sequencing information.
-Extensive analysis by using the Cancer Genome Atlas data not only demonstrated the effectiveness of the method, but also found new individual-specific driver genes and network patterns for various types of cancer.
-Biological experiments on drug resistance further validated one important advantage of our method over the traditional methods, i.e.
-we can even identify such drug resistance genes that actually have no clear differential expression between samples with and without the resistance, due to the additional network information.},
-	file = {/home/rcannood/Zotero/storage/6MSYIL7R/Liu et al. - 2016 - Personalized characterization of diseases using sample-specific networks.pdf},
-	journal = {Nucleic Acids Research},
-	keywords = {cancer,drug resistance,genes,steroid-sensitive nephrotic syndrome},
-	number = {22},
-	pmid = {27596597},
-}
-
-@Article{luecken_currentbestpractices_2019,
-	title = {Current Best Practices in Single-Cell {{RNA}}-Seq Analysis: A Tutorial},
-	shorttitle = {Current Best Practices in Single-Cell {{RNA}}-Seq Analysis},
-	author = {Malte D Luecken and Fabian J Theis},
-	year = {2019},
-	month = {jun},
-	volume = {15},
-	pages = {e8746},
-	issn = {1744-4292},
-	doi = {10.15252/msb.20188746},
-	abstract = {Abstract Single-cell RNA-seq has enabled gene expression to be studied at an unprecedented resolution.
-The promise of this technology is attracting a growing user base for single-cell analysis methods.
-As more analysis tools are becoming available, it is becoming increasingly difficult to navigate this landscape and produce an up-to-date workflow to analyse one's data.
-Here, we detail the steps of a typical single-cell RNA-seq analysis, including pre-processing (quality control, normalization, data correction, feature selection, and dimensionality reduction) and cell- and gene-level downstream analysis.
-We formulate current best-practice recommendations for these steps based on independent comparison studies.
-We have integrated these best-practice recommendations into a workflow, which we apply to a public dataset to further illustrate how these steps work in practice.
-Our documented case study can be found at https://www.github.com/theislab/single-cell-tutorial.
-This review will serve as a workflow tutorial for new entrants into the field, and help established users update their analysis pipelines.},
-	file = {/home/rcannood/Zotero/storage/MY4TZUJT/Luecken and Theis - 2019 - Current best practices in single-cell RNA-seq anal.pdf;/home/rcannood/Zotero/storage/7PFR3DKU/msb.html},
-	journal = {Molecular Systems Biology},
-	keywords = {analysis pipeline development,computational biology,data analysis tutorial,single-cell RNA-seq},
-	number = {6},
-}
-
-@Article{lummertzdarocha_reconstructioncomplexsinglecell_2018,
-	title = {Reconstruction of Complex Single-Cell Trajectories Using {{CellRouter}}},
-	author = {Edroaldo {Lummertz da Rocha} and R.
-Grant Rowe and Vanessa Lundin and Mohan Malleshaiah and Deepak Kumar Jha and Carlos R.
-Rambo and Hu Li and Trista E.
-North and James J.
-Collins and George Q.
-Daley},
-	year = {2018},
-	month = {mar},
-	volume = {9},
-	pages = {892},
-	issn = {2041-1723},
-	doi = {10.1038/s41467-018-03214-y},
-	abstract = {A better understanding of the cell-fate transitions that occur in complex cellular ecosystems in normal development and disease could inform cell engineering efforts and lead to improved therapies.
-However, a major challenge is to simultaneously identify new cell states, and their transitions, to elucidate the gene expression dynamics governing cell-type diversification.
-Here, we present CellRouter, a multifaceted single-cell analysis platform that identifies complex cell-state transition trajectories by using flow networks to explore the subpopulation structure of multi-dimensional, single-cell omics data.
-We demonstrate its versatility by applying CellRouter to single-cell RNA sequencing data sets to reconstruct cell-state transition trajectories during hematopoietic stem and progenitor cell (HSPC) differentiation to the erythroid, myeloid and lymphoid lineages, as well as during re-specification of cell identity by cellular reprogramming of monocytes and B-cells to HSPCs.
-CellRouter opens previously undescribed paths for in-depth characterization of complex cellular ecosystems and establishment of enhanced cell engineering approaches.},
-	journal = {Nature Communications},
-	number = {1},
-}
-
-@Article{marbach_generatingrealisticsilico_2009,
-	title = {Generating {{Realistic In Silico Gene Networks}} for {{Performance Assessment}} of {{Reverse Engineering Methods}}},
-	author = {Daniel Marbach and Thomas Schaffter and Claudio Mattiussi and Dario Floreano},
-	year = {2009},
-	month = {feb},
-	volume = {16},
-	pages = {229--239},
-	issn = {1557-8666},
-	doi = {10.1089/cmb.2008.09TT},
-	abstract = {Reverse engineering methods are typically first tested on simulated data from in silico networks, for systematic and efficient performance assessment, before an application to real biological networks.
-In this paper, we present a method for generating biologically plausible in silico networks, which allow realistic performance assessment of network inference algorithms.
-Instead of using random graph models, which are known to only partly capture the structural properties of biological networks, we generate network structures by extracting modules from known biological interaction networks.
-Using the yeast transcriptional regulatory network as a test case, we show that extracted modules have a biologically plausible connectivity because they preserve functional and structural properties of the original network.
-Our method was selected to generate the {"}gold standard{"} networks for the gene network reverse engineering challenge of the third DREAM conference (Dialogue on Reverse Engineering Assessment and Methods 2008, Cambridge, MA).},
-	file = {/home/rcannood/Zotero/storage/NXLVA76Q/Marbach et al. - 2009 - Generating Realistic In Silico Gene Networks for Performance Assessment of Reverse Engineering Methods.pdf},
-	journal = {Journal of Computational Biology},
-	keywords = {Algorithms,Computational Biology,Computational Biology: methods,Computer Simulation,DRE,Gene Expression Profiling,Gene Expression Profiling: methods,gene regulatory networks,Gene Regulatory Networks,Genetic,Models,reverse engineering,Saccharomyces cerevisiae,Saccharomyces cerevisiae: genetics,Saccharomyces cerevisiae: metabolism,Software},
-	number = {2},
-	pmid = {19183003},
-}
-
-@Article{marbach_revealingstrengthsweaknesses_2010,
-	title = {Revealing Strengths and Weaknesses of Methods for Gene Network Inference},
-	author = {Daniel Marbach and Robert J Prill and Thomas Schaffter and Claudio Mattiussi and Dario Floreano and Gustavo Stolovitzky},
-	year = {2010},
-	month = {apr},
-	volume = {107},
-	pages = {6286--6291},
-	issn = {1091-6490},
-	doi = {10.1073/pnas.0913357107},
-	abstract = {Numerous methods have been developed for inferring gene regulatory networks from expression data, however, both their absolute and comparative performance remain poorly understood.
-In this paper, we introduce a framework for critical performance assessment of methods for gene network inference.
-We present an in silico benchmark suite that we provided as a blinded, community-wide challenge within the context of the DREAM (Dialogue on Reverse Engineering Assessment and Methods) project.
-We assess the performance of 29 gene-network-inference methods, which have been applied independently by participating teams.
-Performance profiling reveals that current inference methods are affected, to various degrees, by different types of systematic prediction errors.
-In particular, all but the best-performing method failed to accurately infer multiple regulatory inputs (combinatorial regulation) of genes.
-The results of this community-wide experiment show that reliable network inference from gene expression data remains an unsolved problem, and they indicate potential ways of network reconstruction improvements.},
-	file = {/home/rcannood/Zotero/storage/784SY83Y/Marbach et al. - 2010 - Revealing strengths and weaknesses of methods for gene network inference.pdf},
-	journal = {Proceedings of the National Academy of Sciences},
-	keywords = {Biometry,community experiment,Computational Biology,Computational Biology: methods,DREAM challenge,Gene Expression Profiling,Gene Regulatory Networks,performance assessment,reverse engineering,transcriptional regulatory networks},
-	number = {14},
-	pmid = {20308593},
-}
-
-@Article{marbach_wisdomcrowdsrobust_2012,
-	title = {Wisdom of {{Crowds}} for {{Robust Gene Network Inference}}},
-	author = {Daniel Marbach and James C Costello and Robert K{\"u}ffner and Nicole M Nicci Vega and Robert J Prill and Diogo M Camacho and Kyle R Allison and Manolis Kellis and James J Collins and Gustavo Stolovitzky and {DREAM5 Consortium} and Manolis Kellis and James J Collins and Gustavo Stolovitzky},
-	year = {2012},
-	month = {jul},
-	volume = {9},
-	pages = {796--804},
-	issn = {1548-7091},
-	doi = {10.1038/nmeth.2016},
-	abstract = {Reconstructing gene regulatory networks from high-throughput data is a long-standing problem.
-Through the DREAM project (Dialogue on Reverse Engineering Assessment and Methods), we performed a comprehensive blind assessment of over thirty network inference methods on Escherichia coli, Staphylococcus aureus, Saccharomyces cerevisiae, and in silico microarray data.
-We characterize performance, data requirements, and inherent biases of different inference approaches offering guidelines for both algorithm application and development.
-We observe that no single inference method performs optimally across all datasets.
-In contrast, integration of predictions from multiple inference methods shows robust and high performance across diverse datasets.
-Thereby, we construct high-confidence networks for E.
-coli and S.
-aureus, each comprising \$\textbackslash backslash\$textasciitilde\{\}1700 transcriptional interactions at an estimated precision of 50\%.
-We experimentally test 53 novel interactions in E.
-coli, of which 23 were supported (43\%).
-Our results establish community-based methods as a powerful and robust tool for the inference of transcriptional gene regulatory networks.},
-	file = {/home/rcannood/Zotero/storage/AQRNS57R/Marbach et al.
-- 2012 - Wisdom of crowds for robust gene network inference.pdf;/home/rcannood/Zotero/storage/KGJVQ9WL/Marbach et al.
-- 2012 - Wisdom of crowds for robust gene network inference.pdf},
-	journal = {Nature methods},
-	keywords = {Algorithms,Bacterial,Bacterial: genetics,Computational Biology,Escherichia coli,Escherichia coli: genetics,Gene Expression Regulation,Gene Regulatory Networks,Genetic,Genetic: genetics,Oligonucleotide Array Sequence Analysis,Saccharomyces cerevisiae,Saccharomyces cerevisiae: genetics,Software,Staphylococcus aureus,Staphylococcus aureus: genetics,Transcription},
-	number = {8},
-	pmid = {22796662},
-}
-
-@Article{mcfaline-figueroa_pooledsinglecellgenetic_2019,
-	title = {A Pooled Single-Cell Genetic Screen Identifies Regulatory Checkpoints in the Continuum of the Epithelial-to-Mesenchymal Transition},
-	author = {Jos{\a'e} L. McFaline-Figueroa and Andrew J. Hill and Xiaojie Qiu and Dana Jackson and Jay Shendure and Cole Trapnell},
-	year = {2019},
-	month = {sep},
-	volume = {51},
-	pages = {1389--1398},
-	issn = {1546-1718},
-	doi = {10.1038/s41588-019-0489-5},
-	abstract = {Transcriptional profiling of 61,052 cells undergoing epithelial-to-mesenchymal transition (EMT) shows continuous waves of gene regulation, not discrete stages.
-A CRISPR screen orders EMT regulators along EMT progression.},
-	copyright = {2019 The Author(s), under exclusive licence to Springer Nature America, Inc.},
-	file = {/home/rcannood/Zotero/storage/RAQRP6UC/s41588-019-0489-5.html},
-	journal = {Nature Genetics},
-	language = {en},
-	number = {9},
-}
-
-@Article{moerman_grnboost2arboretoefficient_2019,
-	title = {{{GRNBoost2}} and {{Arboreto}}: Efficient and Scalable Inference of Gene Regulatory Networks},
-	author = {Thomas Moerman and Sara {Aibar Santos} and Carmen {Bravo Gonz{\a'a}lez-Blas} and Jaak Simm and Yves Moreau and Jan Aerts and Stein Aerts},
-	year = {2019},
-	month = {jun},
-	volume = {35},
-	pages = {2159--2161},
-	issn = {1367-4803},
-	doi = {10.1093/bioinformatics/bty916},
-	abstract = {Inferring a Gene Regulatory Network (GRN) from gene expression data is a computationally expensive task, exacerbated by increasing data sizes due to advances in high-throughput gene profiling technology, such as single-cell RNA-seq.
-To equip researchers with a toolset to infer GRNs from large expression datasets, we propose GRNBoost2 and the Arboreto framework.
-GRNBoost2 is an efficient algorithm for regulatory network inference using gradient boosting, based on the GENIE3 architecture.
-Arboreto is a computational framework that scales up GRN inference algorithms complying with this architecture.
-Arboreto includes both GRNBoost2 and an improved implementation of GENIE3, as a user-friendly open source Python package.Arboreto is available under the 3-Clause BSD license at http://arboreto.readthedocs.io.Supplementary data are available at Bioinformatics online.},
-	journal = {Bioinformatics},
-	number = {12},
-}
-
-@Article{papadopoulos_prossttprobabilisticsimulation_2019,
-	title = {{{PROSSTT}}: Probabilistic Simulation of Single-Cell {{RNA}}-Seq Data for Complex Differentiation Processes},
-	shorttitle = {{{PROSSTT}}},
-	author = {Nikolaos Papadopoulos and Parra R. Gonzalo and Johannes S{\"o}ding},
-	year = {2019},
-	month = {sep},
-	volume = {35},
-	pages = {3517--3519},
-	issn = {1367-4811},
-	doi = {10.1093/bioinformatics/btz078},
-	abstract = {SUMMARY: Cellular lineage trees can be derived from single-cell RNA sequencing snapshots of differentiating cells.
-Currently, only datasets with simple topologies are available.
-To test and further develop tools for lineage tree reconstruction, we need test datasets with known complex topologies.
-PROSSTT can simulate scRNA-seq datasets for differentiation processes with lineage trees of any desired complexity, noise level, noise model and size.
-PROSSTT also provides scripts to quantify the quality of predicted lineage trees.
-AVAILABILITY AND IMPLEMENTATION: https://github.com/soedinglab/prosstt.
-SUPPLEMENTARY INFORMATION: Supplementary data are available at Bioinformatics online.},
-	file = {/home/rcannood/Zotero/storage/UERERPJX/Papadopoulos et al. - 2019 - PROSSTT probabilistic simulation of single-cell R.pdf},
-	journal = {Bioinformatics (Oxford, England)},
-	keywords = {Cell Differentiation,Gene Expression Profiling,RNA-Seq,Single-Cell Analysis,Software},
-	language = {eng},
-	number = {18},
-	pmcid = {PMC6748774},
-	pmid = {30715210},
-}
-
-@Article{parra_reconstructingcomplexlineage_2019,
-	title = {Reconstructing Complex Lineage Trees from {{scRNA}}-Seq Data Using {{MERLoT}}},
-	author = {R. Gonzalo Parra and Nikolaos Papadopoulos and Laura Ahumada-Arranz and Jakob El Kholtei and Noah Mottelson and Yehor Horokhovsky and Barbara Treutlein and Johannes Soeding},
-	year = {2019},
-	month = {sep},
-	volume = {47},
-	pages = {8961--8974},
-	issn = {1362-4962},
-	doi = {10.1093/nar/gkz706},
-	abstract = {Advances in single-cell transcriptomics techniques are revolutionizing studies of cellular differentiation and heterogeneity.
-It has become possible to track the trajectory of thousands of genes across the cellular lineage trees that represent the temporal emergence of cell types during dynamic processes.
-However, reconstruction of cellular lineage trees with more than a few cell fates has proved challenging.
-We present MERLoT (https://github.com/soedinglab/merlot), a flexible and user-friendly tool to reconstruct complex lineage trees from single-cell transcriptomics data.
-It can impute temporal gene expression profiles along the reconstructed tree.
-We show MERLoT's capabilities on various real cases and hundreds of simulated datasets.},
-	file = {/home/rcannood/Zotero/storage/86VN2MRY/Parra et al. - 2019 - Reconstructing complex lineage trees from scRNA-se.pdf},
-	journal = {Nucleic Acids Research},
-	keywords = {Algorithms,Animals,Cell Differentiation,Cell Lineage,Cluster Analysis,Gene Expression Profiling,Hematopoietic Stem Cells,Humans,Mice,Mouse Embryonic Stem Cells,Sequence Analysis; RNA,Single-Cell Analysis,Software,Time Factors,Transcriptome},
-	language = {eng},
-	number = {17},
-	pmcid = {PMC6755101},
-	pmid = {31428793},
-}
-
-@Manual{patil_pairwisecomparisonsmultiplepairwise_2019,
-	title = {{{pairwiseComparisons}}: {{Multiple}} Pairwise Comparison Tests},
-	author = {Indrajeet Patil},
-	year = {2019},
-	type = {Manual},
-}
-
-@Article{prill_rigorousassessmentsystems_2010,
-	title = {Towards a {{Rigorous Assessment}} of {{Systems Biology Models}}: {{The DREAM3 Challenges}}},
-	author = {Robert J Prill and Marbach Daniel and Saez-Rodriguez Julio and Sorger Peter K.
-and Alexopoulos Leonidas G.
-and Xue Xiaowei and Clarke Neil D.
-and Altan-Bonnet Gregoire and Stolovitzky Gustavo and Daniel Marbach and Julio Saez-Rodriguez and Peter K Sorger and Leonidas G Alexopoulos and Xiaowei Xue and Neil D Clarke and Gregoire Altan-Bonnet and Gustavo Stolovitzky and Marbach Daniel and Saez-Rodriguez Julio and Sorger Peter K.
-and Alexopoulos Leonidas G.
-and Xue Xiaowei and Clarke Neil D.
-and Altan-Bonnet Gregoire and Stolovitzky Gustavo},
-	year = {2010},
-	month = {jan},
-	volume = {5},
-	pages = {e9202},
-	issn = {1932-6203},
-	doi = {10.1371/journal.pone.0009202},
-	abstract = {BACKGROUND: Systems biology has embraced computational modeling in response to the quantitative nature and increasing scale of contemporary data sets.
-The onslaught of data is accelerating as molecular profiling technology evolves.
-The Dialogue for Reverse Engineering Assessments and Methods (DREAM) is a community effort to catalyze discussion about the design, application, and assessment of systems biology models through annual reverse-engineering challenges.
-METHODOLOGY AND PRINCIPAL FINDINGS: We describe our assessments of the four challenges associated with the third DREAM conference which came to be known as the DREAM3 challenges: signaling cascade identification, signaling response prediction, gene expression prediction, and the DREAM3 in silico network challenge.
-The challenges, based on anonymized data sets, tested participants in network inference and prediction of measurements.
-Forty teams submitted 413 predicted networks and measurement test sets.
-Overall, a handful of best-performer teams were identified, while a majority of teams made predictions that were equivalent to random.
-Counterintuitively, combining the predictions of multiple teams (including the weaker teams) can in some cases improve predictive power beyond that of any single method.
-CONCLUSIONS: DREAM provides valuable feedback to practitioners of systems biology modeling.
-Lessons learned from the predictions of the community provide much-needed context for interpreting claims of efficacy of algorithms described in the scientific literature.},
-	file = {/home/rcannood/Zotero/storage/Q5E22H2F/Prill et al. - 2010 - Towards a Rigorous Assessment of Systems Biology Models The DREAM3 Challenges.pdf},
-	journal = {PLoS ONE},
-	keywords = {Algorithms,Animals,Biological,Cluster Analysis,Computational Biology,Computational Biology: methods,Gene Expression Profiling,Gene Expression Profiling: methods,Gene Regulatory Networks,Humans,Models,Protein Interaction Mapping,Protein Interaction Mapping: methods,Reproducibility of Results,Signal Transduction,Software,Systems Biology,Systems Biology: methods},
-	number = {2},
-	pmid = {20186320},
-}
-
-@Article{rekhtman_directinteractionhematopoietic_1999,
-	title = {Direct Interaction of Hematopoietic Transcription Factors {{PU}}.1 and {{GATA}}-1: Functional Antagonism in Erythroid Cells},
-	shorttitle = {Direct Interaction of Hematopoietic Transcription Factors {{PU}}.1 and {{GATA}}-1},
-	author = {N. Rekhtman and F. Radparvar and T. Evans and A. I. Skoultchi},
-	year = {1999},
-	month = {jun},
-	volume = {13},
-	pages = {1398--1411},
-	issn = {0890-9369},
-	doi = {10.1101/gad.13.11.1398},
-	abstract = {Malignant transformation usually inhibits terminal cell differentiation but the precise mechanisms involved are not understood.
-PU.1 is a hematopoietic-specific Ets family transcription factor that is required for development of some lymphoid and myeloid lineages.
-PU.1 can also act as an oncoprotein as activation of its expression in erythroid precursors by proviral insertion or transgenesis causes erythroleukemias in mice.
-Restoration of terminal differentiation in the mouse erythroleukemia (MEL) cells requires a decline in the level of PU.1, indicating that PU.1 can block erythroid differentiation.
-Here we investigate the mechanism by which PU.1 interferes with erythroid differentiation.
-We find that PU.1 interacts directly with GATA-1, a zinc finger transcription factor required for erythroid differentiation.
-Interaction between PU.1 and GATA-1 requires intact DNA-binding domains in both proteins.
-PU.1 represses GATA-1-mediated transcriptional activation.
-Both the DNA binding and transactivation domains of PU.1 are required for repression and both domains are also needed to block terminal differentiation in MEL cells.
-We also show that ectopic expression of PU.1 in Xenopus embryos is sufficient to block erythropoiesis during normal development.
-Furthermore, introduction of exogenous GATA-1 in both MEL cells and Xenopus embryos and explants relieves the block to erythroid differentiation imposed by PU.1.
-Our results indicate that the stoichiometry of directly interacting but opposing transcription factors may be a crucial determinant governing processes of normal differentiation and malignant transformation.},
-	file = {/home/rcannood/Zotero/storage/77DJ34BY/Rekhtman et al. - 1999 - Direct interaction of hematopoietic transcription .pdf},
-	journal = {Genes \& Development},
-	keywords = {3T3 Cells,Animals,Binding Sites,Cell Differentiation,DNA-Binding Proteins,Erythroid Precursor Cells,Erythroid-Specific DNA-Binding Factors,GATA1 Transcription Factor,HeLa Cells,Hematopoiesis,Humans,Mice,Proto-Oncogene Proteins,Recombinant Fusion Proteins,Trans-Activators,Transcription Factors,Transcription; Genetic,Transcriptional Activation,Xenopus laevis,Zinc Fingers},
-	language = {eng},
-	number = {11},
-	pmcid = {PMC316770},
-	pmid = {10364157},
-}
-
-@Article{roy_systemgeneratingtranscription_2008,
-	title = {A System for Generating Transcription Regulatory Networks with Combinatorial Control of Transcription},
-	author = {Sushmita Roy and Margaret Werner-Washburne and Terran Lane},
-	year = {2008},
-	month = {may},
-	volume = {24},
-	pages = {1318--1320},
-	issn = {1367-4803},
-	doi = {10.1093/bioinformatics/btn126},
-	abstract = {Summary: We have developed a new software system, REgulatory Network generator with COmbinatorial control (RENCO), for automatic generation of differential equations describing pre-transcriptional combinatorics in artificial regulatory networks.
-RENCO has the following benefits: (a) it explicitly models protein\textendash protein interactions among transcription factors, (b) it captures combinatorial control of transcription factors on target genes and (c) it produces output in Systems Biology Markup Language (SBML) format, which allows these equations to be directly imported into existing simulators.
-Explicit modeling of the protein interactions allows RENCO to incorporate greater mechanistic detail of the transcription machinery compared to existing models and can provide a better assessment of algorithms for regulatory network inference.Availability: RENCO is a C++ command line program, available at http://sourceforge.net/projects/renco/Contact:terran@cs.unm.eduSupplementary information: Supplementary data are available at Bioinformatics online.},
-	journal = {Bioinformatics},
-	number = {10},
-}
-
-@Article{saelens_comparisonsinglecelltrajectory_2019,
-	title = {A Comparison of Single-Cell Trajectory Inference Methods},
-	author = {Wouter Saelens and Robrecht Cannoodt and Helena Todorov and Yvan Saeys},
-	year = {2019},
-	volume = {37},
-	issn = {15461696},
-	doi = {10.1038/s41587-019-0071-9},
-	abstract = {Trajectory inference approaches analyze genome-wide omics data from thousands of single cells and computationally infer the order of these cells along developmental trajectories.
-Although more than 70 trajectory inference tools have already been developed, it is challenging to compare their performance because the input they require and output models they produce vary substantially.
-Here, we benchmark 45 of these methods on 110 real and 229 synthetic datasets for cellular ordering, topology, scalability and usability.
-Our results highlight the complementarity of existing tools, and that the choice of method should depend mostly on the dataset dimensions and trajectory topology.
-Based on these results, we develop a set of guidelines to help users select the best method for their dataset.
-Our freely available data and evaluation pipeline ( https://benchmark.dynverse.org ) will aid in the development of improved tools designed to analyze increasingly large and complex single-cell datasets.},
-	file = {/home/rcannood/Zotero/storage/JKZCFAVA/Saelens et al. - 2019 - A comparison of single-cell trajectory inference methods.pdf;/home/rcannood/Zotero/storage/T9HMNIDB/Saelens2019.pdf},
-	journal = {Nature Biotechnology},
-	number = {May},
-}
-
-@Article{schaffter_genenetweaversilicobenchmark_2011,
-	title = {{{GeneNetWeaver}}: In Silico Benchmark Generation and Performance Profiling of Network Inference Methods.},
-	shorttitle = {\{\vphantom\}{{GeneNetWeaver}}\vphantom\{\}},
-	author = {Thomas Schaffter and Daniel Marbach and Dario Floreano},
-	year = {2011},
-	month = {aug},
-	volume = {27},
-	pages = {2263--2270},
-	issn = {1367-4811},
-	doi = {10.1093/bioinformatics/btr373},
-	abstract = {MOTIVATION: Over the last decade, numerous methods have been developed for inference of regulatory networks from gene expression data.
-However, accurate and systematic evaluation of these methods is hampered by the difficulty of constructing adequate benchmarks and the lack of tools for a differentiated analysis of network predictions on such benchmarks.
-RESULTS: Here, we describe a novel and comprehensive method for in silico benchmark generation and performance profiling of network inference methods available to the community as an open-source software called GeneNetWeaver (GNW).
-In addition to the generation of detailed dynamical models of gene regulatory networks to be used as benchmarks, GNW provides a network motif analysis that reveals systematic prediction errors, thereby indicating potential ways of improving inference methods.
-The accuracy of network inference methods is evaluated using standard metrics such as precision-recall and receiver operating characteristic curves.
-We show how GNW can be used to assess the performance and identify the strengths and weaknesses of six inference methods.
-Furthermore, we used GNW to provide the international Dialogue for Reverse Engineering Assessments and Methods (DREAM) competition with three network inference challenges (DREAM3, DREAM4 and DREAM5).
-AVAILABILITY: GNW is available at http://gnw.sourceforge.net along with its Java source code, user manual and supporting data.
-SUPPLEMENTARY INFORMATION: Supplementary data are available at Bioinformatics online.
-CONTACT: dario.floreano@epfl.ch.},
-	file = {/home/rcannood/Zotero/storage/P5JMJUBN/Schaffter, Marbach, Floreano - 2011 - GeneNetWeaver in silico benchmark generation and performance profiling of network inference method.pdf},
-	journal = {Bioinformatics},
-	keywords = {Benchmarking,Escherichia coli,Escherichia coli: genetics,Gene Expression,Gene Regulatory Networks,Genetic,Models,Software},
-	number = {16},
-	pmid = {21697125},
-}
-
-@Article{schilstra_biologicgeneexpression_2008,
-	title = {Bio-{{Logic}}: {{Gene Expression}} and the {{Laws}} of {{Combinatorial Logic}}},
-	shorttitle = {Bio-{{Logic}}},
-	author = {Maria J. Schilstra and Chrystopher L. Nehaniv},
-	year = {2008},
-	month = {jan},
-	volume = {14},
-	pages = {121--133},
-	issn = {1064-5462},
-	doi = {10.1162/artl.2008.14.1.121},
-	abstract = {Abstract At the heart of the development of fertilized eggs into fully formed organisms and the adaptation of cells to changed conditions are genetic regulatory networks (GRNs).
-In higher multicellular organisms, signal selection and multiplexing are performed at the cis-regulatory domains of genes, where combinations of transcription factors (TFs) regulate the rates at which the genes are transcribed into mRNA.
-To be able to act as activators or repressors of gene transcription, TFs must first bind to target sequences on the regulatory domains.
-Two TFs that act in concert may bind entirely independently of each other, but more often binding of the first one will alter the affinity of the other for its binding site.
-This article presents a systematic investigation into the effect of TF binding dependences on the predicted regulatory function of this bio-logic.
-Four extreme scenarios, commonly used to classify enzyme activation and inhibition patterns, for the binding of two TFs were explored: independent (the TFs bind without affecting each other's affinities), competitive (the TFs compete for the same binding site), ordered (the TFs bind in a compulsory order), and joint binding (the TFs either bind as a preformed complex, or binding of one is virtually impossible in the absence of the other).
-The conclusions are: (1) the laws of combinatorial logic hold only for systems with independently binding TFs; (2) systems formed according to the other scenarios can mimic the functions of their Boolean logical counterparts, but cannot be combined or decomposed in the same way; and (3) the continuously scaled output of systems consisting of competitively binding activators and repressors can be controlled more robustly than that of single TF or (quasi-)logical multi-TF systems.},
-	file = {/home/rcannood/Zotero/storage/C7UWQGI3/Schilstra and Nehaniv - 2008 - Bio-Logic Gene Expression and the Laws of Combina.pdf;/home/rcannood/Zotero/storage/KXDLUMAG/artl.2008.14.1.html},
-	journal = {Artificial Life},
-	number = {1},
-}
-
-@Article{schwanhausser_globalquantificationmammalian_2011,
-	title = {Global Quantification of Mammalian Gene Expression Control},
-	author = {Bj{\"o}rn Schwanh{\"a}usser and Dorothea Busse and Na Li and Gunnar Dittmar and Johannes Schuchhardt and Jana Wolf and Wei Chen and Matthias Selbach},
-	year = {2011},
-	month = {may},
-	volume = {473},
-	pages = {337--342},
-	issn = {1476-4687},
-	doi = {10.1038/nature10098},
-	abstract = {Gene expression is a multistep process that involves the transcription, translation and turnover of messenger RNAs and proteins.
-Although it is one of the most fundamental processes of life, the entire cascade has never been quantified on a genome-wide scale.
-Here we simultaneously measured absolute mRNA and protein abundance and turnover by parallel metabolic pulse labelling for more than 5,000 genes in mammalian cells.
-Whereas mRNA and protein levels correlated better than previously thought, corresponding half-lives showed no correlation.
-Using a quantitative model we have obtained the first genome-scale prediction of synthesis rates of mRNAs and proteins.
-We find that the cellular abundance of proteins is predominantly controlled at the level of translation.
-Genes with similar combinations of mRNA and protein stability shared functional properties, indicating that half-lives evolved under energetic and dynamic constraints.
-Quantitative information about all stages of gene expression provides a rich resource and helps to provide a greater understanding of the underlying design principles.},
-	journal = {Nature},
-	number = {7347},
-}
-
-@Article{smith_spatialstochasticintracellular_2019,
-	title = {Spatial {{Stochastic Intracellular Kinetics}}: {{A Review}} of {{Modelling Approaches}}},
-	author = {Stephen Smith and Ramon Grima},
-	year = {2019},
-	month = {aug},
-	volume = {81},
-	pages = {2960--3009},
-	issn = {1522-9602},
-	doi = {10.1007/s11538-018-0443-1},
-	abstract = {Models of chemical kinetics that incorporate both stochasticity and diffusion are an increasingly common tool for studying biology.
-The variety of competing models is vast, but two stand out by virtue of their popularity: the reaction\textendash diffusion master equation and Brownian dynamics.
-In this review, we critically address a number of open questions surrounding these models: How can they be justified physically?
-How do they relate to each other?
-How do they fit into the wider landscape of chemical models, ranging from the rate equations to molecular dynamics?
-This review assumes no prior knowledge of modelling chemical kinetics and should be accessible to a wide range of readers.},
-	journal = {Bulletin of Mathematical Biology},
-	number = {8},
-}
-
-@Article{soneson_biasrobustnessscalability_2018,
-	title = {Bias, Robustness and Scalability in Single-Cell Differential Expression Analysis},
-	author = {Charlotte Soneson and Mark D. Robinson},
-	year = {2018},
-	month = {apr},
-	volume = {15},
-	pages = {255--261},
-	issn = {1548-7105},
-	doi = {10.1038/nmeth.4612},
-	abstract = {Many methods have been used to determine differential gene expression from single-cell RNA (scRNA)-seq data.
-We evaluated 36 approaches using experimental and synthetic data and found considerable differences in the number and characteristics of the genes that are called differentially expressed.
-Prefiltering of lowly expressed genes has important effects, particularly for some of the methods developed for bulk RNA-seq data analysis.
-However, we found that bulk RNA-seq analysis methods do not generally perform worse than those developed specifically for scRNA-seq.
-We also present conquer, a repository of consistently processed, analysis-ready public scRNA-seq data sets that is aimed at simplifying method evaluation and reanalysis of published results.
-Each data set provides abundance estimates for both genes and transcripts, as well as quality control and exploratory analysis reports.},
-	journal = {Nature Methods},
-	keywords = {Computational Biology,Gene Expression Profiling,Gene Expression Regulation,High-Throughput Nucleotide Sequencing,RNA,Sequence Analysis; RNA,Single-Cell Analysis,Software},
-	language = {eng},
-	number = {4},
-	pmid = {29481549},
-}
-
-@Article{street_slingshotcelllineage_2018,
-	title = {Slingshot: {{Cell Lineage}} and {{Pseudotime Inference}} for {{Single}}-{{Cell Transcriptomics}}},
-	shorttitle = {Slingshot},
-	author = {Kelly Street and Davide Risso and Russell B Fletcher and Diya Das and John Ngai and Nir Yosef and Elizabeth Purdom and Sandrine Dudoit},
-	year = {2018},
-	month = {jun},
-	volume = {19},
-	pages = {477},
-	issn = {1471-2164},
-	doi = {10.1186/s12864-018-4772-0},
-	abstract = {Single-cell transcriptomics allows researchers to investigate complex communities of heterogeneous cells.
-It can be applied to stem cells and their descendants in order to chart the progression from multipotent progenitors to fully differentiated cells.
-While a variety of statistical and computational methods have been proposed for inferring cell lineages, the problem of accurately characterizing multiple branching lineages remains difficult to solve.},
-	journal = {BMC Genomics},
-	number = {1},
-}
-
-@Article{vallejos_normalizingsinglecellrna_2017,
-	title = {Normalizing Single-Cell {{RNA}} Sequencing Data: Challenges and Opportunities},
-	shorttitle = {Normalizing Single-Cell {{RNA}} Sequencing Data},
-	author = {Catalina A. Vallejos and Davide Risso and Antonio Scialdone and Sandrine Dudoit and John C. Marioni},
-	year = {2017},
-	month = {jun},
-	volume = {14},
-	pages = {565--571},
-	publisher = {{Nature Publishing Group}},
-	issn = {1548-7105},
-	doi = {10.1038/nmeth.4292},
-	abstract = {This Perspective examines single-cell RNA-seq data challenges and the need for normalization methods designed specifically for single-cell data in order to remove technical biases.},
-	copyright = {2017 Nature Publishing Group, a division of Macmillan Publishers Limited. All Rights Reserved.},
-	journal = {Nature Methods},
-	language = {en},
-	number = {6},
-}
-
-@Article{vandenbulcke_syntrengeneratorsynthetic_2006,
-	title = {{{SynTReN}}: A Generator of Synthetic Gene Expression Data for Design and Analysis of Structure Learning Algorithms},
-	shorttitle = {{{SynTReN}}},
-	author = {Tim {Van den Bulcke} and Koenraad {Van Leemput} and Bart Naudts and Piet {van Remortel} and Hongwu Ma and Alain Verschoren and Bart {De Moor} and Kathleen Marchal},
-	year = {2006},
-	month = {jan},
-	volume = {7},
-	pages = {43},
-	issn = {1471-2105},
-	doi = {10.1186/1471-2105-7-43},
-	abstract = {The development of algorithms to infer the structure of gene regulatory networks based on expression data is an important subject in bioinformatics research.
-Validation of these algorithms requires benchmark data sets for which the underlying network is known.
-Since experimental data sets of the appropriate size and design are usually not available, there is a clear need to generate well-characterized synthetic data sets that allow thorough testing of learning algorithms in a fast and reproducible manner.},
-	journal = {BMC Bioinformatics},
-	language = {en},
-	number = {1},
-}
-
-@Article{vandesande_scalablescenicworkflow_2020,
-	title = {A Scalable {{SCENIC}} Workflow for Single-Cell Gene Regulatory Network Analysis},
-	author = {Bram {Van de Sande} and Christopher Flerin and Kristofer Davie and Maxime {De Waegeneer} and Gert Hulselmans and Sara Aibar and Ruth Seurinck and Wouter Saelens and Robrecht Cannoodt and Quentin Rouchon and Toni Verbeiren and Dries {De Maeyer} and Joke Reumers and Yvan Saeys and Stein Aerts},
-	year = {2020},
-	month = {jul},
-	volume = {15},
-	pages = {2247--2276},
-	issn = {1750-2799},
-	doi = {10.1038/s41596-020-0336-2},
-	abstract = {This protocol explains how to perform a fast SCENIC analysis alongside standard best practices steps on single-cell RNA-sequencing data using software containers and Nextflow pipelines.
-SCENIC reconstructs regulons (i.e., transcription factors and their target genes) assesses the activity of these discovered regulons in individual cells and uses these cellular activity patterns to find meaningful clusters of cells.
-Here we present an improved version of SCENIC with several advances.
-SCENIC has been refactored and reimplemented in Python (pySCENIC), resulting in a tenfold increase in speed, and has been packaged into containers for ease of use.
-It is now also possible to use epigenomic track databases, as well as motifs, to refine regulons.
-In this protocol, we explain the different steps of SCENIC: the workflow starts from the count matrix depicting the gene abundances for all cells and consists of three stages.
-First, coexpression modules are inferred using a regression per-target approach (GRNBoost2).
-Next, the indirect targets are pruned from these modules using cis-regulatory motif discovery (cisTarget).
-Lastly, the activity of these regulons is quantified via an enrichment score for the regulon's target genes (AUCell).
-Nonlinear projection methods can be used to display visual groupings of cells based on the cellular activity patterns of these regulons.
-The results can be exported as a loom file and visualized in the SCope web application.
-This protocol is illustrated on two use cases: a peripheral blood mononuclear cell data set and a panel of single-cell RNA-sequencing cancer experiments.
-For a data set of 10,000 genes and 50,000 cells, the pipeline runs in {$<$}2 h.},
-	journal = {Nature Protocols},
-	number = {7},
-}
-
-@Article{vieth_powsimrpoweranalysis_2017,
-	title = {{{powsimR}}: Power Analysis for Bulk and Single Cell {{RNA}}-Seq Experiments},
-	shorttitle = {{{powsimR}}},
-	author = {Beate Vieth and Christoph Ziegenhain and Swati Parekh and Wolfgang Enard and Ines Hellmann},
-	year = {2017},
-	month = {nov},
-	volume = {33},
-	pages = {3486--3488},
-	issn = {1367-4803},
-	doi = {10.1093/bioinformatics/btx435},
-	abstract = {AbstractSummary.  Power analysis is essential to optimize the design of RNA-seq experiments and to assess and compare the power to detect differentially express},
-	file = {/home/rcannood/Zotero/storage/MFRF86MZ/Vieth et al. - 2017 - powsimR power analysis for bulk and single cell R.pdf;/home/rcannood/Zotero/storage/HSCUJRI8/3952669.html},
-	journal = {Bioinformatics},
-	language = {en},
-	number = {21},
-}
-
-@Article{wang_quantifyingwaddingtonlandscape_2011,
-	title = {Quantifying the {{Waddington Landscape}} and {{Biological Paths}} for {{Development}} and {{Differentiation}}},
-	author = {Jin Wang and Kun Zhang and Li Xu and Erkang Wang},
-	year = {2011},
-	month = {may},
-	volume = {108},
-	pages = {8257--8262},
-	issn = {0027-8424, 1091-6490},
-	doi = {10.1073/pnas.1017017108},
-	abstract = {We developed a theoretical framework to prove the existence and quantify the Waddington landscape as well as chreode-biological paths for development and differentiation.
-The cells can have states with the higher probability ones giving the different cell types.
-Different cell types correspond to different basins of attractions of the probability landscape.
-We study how the cells develop from undifferentiated cells to differentiated cells from landscape perspectives.
-We quantified the Waddington landscape through construction of underlying probability landscape for cell development.
-We show the developmental process proceeds as moving from undifferentiated to the differentiated basins of attractions.
-The barrier height of the basins of attractions correlates with the escape time that determines the stability of cell types.
-We show that the developmental process can be quantitatively described and uncovered by the biological paths on the quantified Waddington landscape from undifferentiated to the differentiated cells.
-We found the dynamics of the developmental process is controlled by a combination of the gradient and curl force on the landscape.
-The biological paths often do not follow the steepest descent path on the landscape.
-The landscape framework also quantifies the possibility of reverse differentiation process such as cell reprogramming from differentiated cells back to the original stem cell.
-We show that the biological path of reverse differentiation is irreversible and different from the one for differentiation process.
-We found that the developmental process described by the underlying landscape and the associated biological paths is relatively stable and robust against the influences of environmental perturbations.},
-	journal = {Proceedings of the National Academy of Sciences},
-	number = {20},
-	pmid = {21536909},
-}
-
-@Article{weber_essentialguidelinescomputational_2019,
-	title = {Essential Guidelines for Computational Method Benchmarking},
-	author = {Lukas M. Weber and Wouter Saelens and Robrecht Cannoodt and Charlotte Soneson and Alexander Hapfelmeier and Paul P. Gardner and Anne-Laure Boulesteix and Yvan Saeys and Mark D. Robinson},
-	year = {2019},
-	month = {jun},
-	volume = {20},
-	pages = {125},
-	issn = {1474-760X},
-	doi = {10.1186/s13059-019-1738-8},
-	abstract = {In computational biology and other sciences, researchers are frequently faced with a choice between several computational methods for performing data analyses.
-Benchmarking studies aim to rigorously compare the performance of different methods using well-characterized benchmark datasets, to determine the strengths of each method or to provide recommendations regarding suitable choices of methods for an analysis.
-However, benchmarking studies must be carefully designed and implemented to provide accurate, unbiased, and informative results.
-Here, we summarize key practical guidelines and recommendations for performing high-quality benchmarking analyses, based on our experiences in computational biology.},
-	file = {/home/rcannood/Zotero/storage/83LMG2SG/Weber et al. - 2019 - Essential guidelines for computational method benc.pdf;/home/rcannood/Zotero/storage/GL7G9MIR/s13059-019-1738-8.html},
-	journal = {Genome Biology},
-	number = {1},
-}
-
-@Article{wickham_welcometidyverse_2019,
-	title = {Welcome to the {{Tidyverse}}},
-	author = {Hadley Wickham and Mara Averick and Jennifer Bryan and Winston Chang and Lucy McGowan and Romain Fran{\c c}ois and Garrett Grolemund and Alex Hayes and Lionel Henry and Jim Hester and Max Kuhn and Thomas Pedersen and Evan Miller and Stephan Bache and Kirill M{\"u}ller and Jeroen Ooms and David Robinson and Dana Seidel and Vitalie Spinu and Kohske Takahashi and Davis Vaughan and Claus Wilke and Kara Woo and Hiroaki Yutani},
-	year = {2019},
-	month = {nov},
-	doi = {10.21105/joss.01686},
-	abstract = {Wickham et al., (2019). Welcome to the Tidyverse. Journal of Open Source Software, 4(43), 1686, https://doi.org/10.21105/joss.01686},
-	file = {/home/rcannood/Zotero/storage/TVYSNC4J/Wickham et al. - 2019 - Welcome to the Tidyverse.pdf;/home/rcannood/Zotero/storage/EAQ2VYF8/joss.html},
-	language = {en},
-}
-
-@Article{xu_regulationbifurcatingcell_2015,
-	title = {Regulation of Bifurcating {{B}} Cell Trajectories by Mutual Antagonism between Transcription Factors {{IRF4}} and {{IRF8}}},
-	author = {Heping Xu and Virendra K Chaudhri and Zhiguo Wu and Konstantinos Biliouris and Krista Dienger-Stambaugh and Yrina Rochman and Harinder Singh},
-	year = {2015},
-	month = {dec},
-	volume = {16},
-	pages = {1274--1281},
-	abstract = {Upon recognition of antigen, B cells undertake a bifurcated response in which some cells rapidly differentiate into plasmablasts while others undergo affinity maturation in germinal centers (GCs).
-Here we identified a double-negative feedback loop between the transcription factors IRF4 and IRF8 that regulated the initial developmental bifurcation of activated B cells as well as the GC response.
-IRF8 dampened signaling via the B cell antigen receptor (BCR), facilitated antigen-specific interaction with helper T cells, and promoted antibody affinity maturation while antagonizing IRF4-driven differentiation of plasmablasts.
-Genomic analysis revealed concentration-dependent actions of IRF4 and IRF8 in regulating distinct gene-expression programs.
-Stochastic modeling suggested that the double-negative feedback was sufficient to initiate bifurcation of the B cell developmental trajectories.},
-	journal = {Nat. Immunol.},
-	number = {12},
-}
-
-@Article{yosef_dynamicregulatorynetwork_2013,
-	title = {Dynamic Regulatory Network Controlling \{\vphantom\}{{TH17}}\vphantom\{\} Cell Differentiation},
-	author = {Nir Yosef and Alex K Shalek and Jellert T Gaublomme and Hulin Jin and Youjin Lee and Amit Awasthi and Chuan Wu and Katarzyna Karwacz and Sheng Xiao and Marsela Jorgolli and David Gennert and Rahul Satija and Arvind Shakya and Diana Y Lu and John J Trombetta and Meenu R Pillai and Peter J Ratcliffe and Mathew L Coleman and Mark Bix and Dean Tantin and Hongkun Park and Vijay K Kuchroo and Aviv Regev},
-	year = {2013},
-	volume = {496},
-	pages = {461--468},
-	abstract = {Despite their importance, the molecular circuits that control the differentiation of naive T cells remain largely unknown.
-Recent studies that reconstructed regulatory networks in mammalian cells have focused on short-term responses and relied on perturbation-based approaches that cannot be readily applied to primary T cells.
-Here we combine transcriptional profiling at high temporal resolution, novel computational algorithms, and innovative nanowire-based perturbation tools to systematically derive and experimentally validate a model of the dynamic regulatory network that controls the differentiation of mouse TH17 cells, a proinflammatory T-cell subset that has been implicated in the pathogenesis of multiple autoimmune diseases.
-The TH17 transcriptional network consists of two self-reinforcing, but mutually antagonistic, modules, with 12 novel regulators, the coupled action of which may be essential for maintaining the balance between TH17 and other CD4(+) T-cell subsets.
-Our study identifies and validates 39 regulatory factors, embeds them within a comprehensive temporal network and reveals its organizational principles; it also highlights novel drug targets for controlling TH17 cell differentiation.},
-	journal = {Nature},
-	number = {7446},
-}
-
-@Article{zappia_splattersimulationsinglecell_2017,
-	title = {Splatter: {{Simulation}} of {{Single}}-{{Cell RNA Sequencing Data}}},
-	shorttitle = {Splatter},
-	author = {Luke Zappia and Belinda Phipson and Alicia Oshlack},
-	year = {2017},
-	month = {sep},
-	volume = {18},
-	pages = {174},
-	issn = {1474-760X},
-	doi = {10.1186/s13059-017-1305-0},
-	abstract = {As single-cell RNA sequencing (scRNA-seq) technologies have rapidly developed, so have analysis methods.
-Many methods have been tested, developed, and validated using simulated datasets.
-Unfortunately, current simulations are often poorly documented, their similarity to real data is not demonstrated, or reproducible code is not available.
-Here, we present the Splatter Bioconductor package for simple, reproducible, and well-documented simulation of scRNA-seq data.
-Splatter provides an interface to multiple simulation methods including Splat, our own simulation, based on a gamma-Poisson distribution.
-Splat can simulate single populations of cells, populations with multiple cell types, or differentiation paths.},
-	journal = {Genome Biology},
-	keywords = {RNA-seq,Simulation,Single-cell,Software},
-}
-
-@Article{zeisel_coupledpremrnamrna_2011,
-	title = {Coupled Pre-{{mRNA}} and {{mRNA}} Dynamics Unveil Operational Strategies Underlying Transcriptional Responses to Stimuli},
-	author = {Amit Zeisel and Wolfgang J K{\"o}stler and Natali Molotski and Jonathan M Tsai and Rita Krauthgamer and Jasmine Jacob-Hirsch and Gideon Rechavi and Yoav Soen and Steffen Jung and Yosef Yarden and Eytan Domany},
-	year = {2011},
-	month = {jan},
-	volume = {7},
-	pages = {529},
-	issn = {1744-4292},
-	doi = {10.1038/msb.2011.62},
-	abstract = {Transcriptional responses to extracellular stimuli involve tuning the rates of transcript production and degradation.
-Here, we show that the time-dependent profiles of these rates can be inferred from simultaneous measurements of precursor mRNA (pre-mRNA) and mature mRNA profiles.
-Transcriptome-wide measurements demonstrate that genes with similar mRNA profiles often exhibit marked differences in the amplitude and onset of their production rate.
-The latter is characterized by a large dynamic range, with a group of genes exhibiting an unexpectedly strong transient production overshoot, thereby accelerating their induction and, when combined with time-dependent degradation, shaping transient responses with precise timing and amplitude.},
-	file = {/home/rcannood/Zotero/storage/QRQW86Y9/Zeisel et al. - 2011 - Coupled pre-mRNA and mRNA dynamics unveil operatio.pdf;/home/rcannood/Zotero/storage/R8ZG26IL/msb.2011.html},
-	journal = {Molecular Systems Biology},
-	keywords = {half-life,operational strategy,pre-mRNA,transcriptional response},
-	number = {1},
-}
-
-@Article{zhang_simulatingmultiplefaceted_2019,
-	title = {Simulating Multiple Faceted Variability in Single Cell {{RNA}} Sequencing},
-	author = {Xiuwei Zhang and Chenling Xu and Nir Yosef},
-	year = {2019},
-	month = {jun},
-	volume = {10},
-	pages = {1--16},
-	issn = {2041-1723},
-	doi = {10.1038/s41467-019-10500-w},
-	abstract = {Simulated single cell RNA sequencing data is useful for method development and comparison.
-Here, the authors developed SymSim, a simulator that explicitly models the main factors of variation in single cell data.},
-	copyright = {2019 The Author(s)},
-	file = {/home/rcannood/Zotero/storage/YLQYCW6E/Zhang et al. - 2019 - Simulating multiple faceted variability in single .pdf;/home/rcannood/Zotero/storage/EXJ96Y59/s41467-019-10500-w.html},
-	journal = {Nature Communications},
-	language = {en},
-	number = {1},
-}
-\end{filecontents}
+\begin{thebibliography}{10}
+
+	\bibitem{aibar_scenicsinglecellregulatory_2017}
+	Sara Aibar, Carmen~Bravo Gonz{\a'a}lez-Blas, Thomas Moerman, V{\^a}n~Anh
+	Huynh-Thu, Hana Imrichova, Gert Hulselmans, Florian Rambow, Jean-Christophe
+	Marine, Pierre Geurts, Jan Aerts, Joost {van den Oord}, Zeynep~Kalender Atak,
+	Jasper Wouters, and Stein Aerts.
+	\newblock {{SCENIC}}: Single-cell regulatory network inference and clustering.
+	\newblock {\em Nature Methods}, 10 2017.
+
+	\bibitem{alpert_alignmentsinglecelltrajectories_2018}
+	Ayelet Alpert, Lindsay~S. Moore, Tania Dubovik, and Shai~S. Shen-Orr.
+	\newblock Alignment of single-cell trajectories to compare cellular expression
+	dynamics.
+	\newblock {\em Nature Methods}, 15(4):267--270, 4 2018.
+
+	\bibitem{bergen_generalizingrnavelocity_2020}
+	Volker Bergen, Marius Lange, Stefan Peidli, F.~Alexander Wolf, and Fabian~J.
+	Theis.
+	\newblock Generalizing {{RNA}} velocity to transient cell states through
+	dynamical modeling.
+	\newblock {\em Nature Biotechnology}, 38(12):1408--1414, 12 2020.
+
+	\bibitem{bjornsson_digitaltwinspersonalize_2019}
+	Bergthor Bj{\"o}rnsson, Carl Borrebaeck, Nils Elander, Thomas Gasslander,
+	Danuta~R. Gawel, Mika Gustafsson, Rebecka J{\"o}rnsten, Eun~Jung Lee, Xinxiu
+	Li, Sandra Lilja, David Mart\'inez-Enguita, Andreas Matussek, Per
+	Sandstr{\"o}m, Samuel Sch{\"a}fer, Margaretha Stenmarker, X.~F. Sun, Oleg
+	Sysoev, Huan Zhang, Mikael Benson, and {on behalf of the Swedish Digital Twin
+		Consortium}.
+	\newblock Digital twins to personalize medicine.
+	\newblock {\em Genome Medicine}, 12(1):4, 12 2019.
+
+	\bibitem{cacchiarelli_aligningsinglecelldevelopmental_2018}
+	Davide Cacchiarelli, Xiaojie Qiu, Sanjay Srivatsan, Anna Manfredi, Michael
+	Ziller, Eliah Overbey, Antonio Grimaldi, Jonna Grimsby, Prapti Pokharel,
+	Kenneth~J. Livak, Shuqiang Li, Alexander Meissner, Tarjei~S. Mikkelsen,
+	John~L. Rinn, and Cole Trapnell.
+	\newblock Aligning {{Single}}-{{Cell Developmental}} and {{Reprogramming
+			Trajectories Identifies Molecular Determinants}} of {{Myogenic Reprogramming
+			Outcome}}.
+	\newblock {\em Cell Systems}, 7(3):258--268.e3, 9 2018.
+
+	\bibitem{cannoodt_singlecellomicsdatasets_2018}
+	Robrecht Cannoodt, Wouter Saelens, Helena Todorov, and Yvan Saeys.
+	\newblock Single-cell -omics datasets containing a trajectory.
+	\newblock {\em Zenodo}, 10 2018.
+
+	\bibitem{conover_multiplecomparisonsprocedures_1979}
+	William~Jay Conover and Ronald~L Iman.
+	\newblock On multiple-comparisons procedures.
+	\newblock Technical report, {Technical report, Los Alamos Scientific
+		Laboratory}, 1979.
+
+	\bibitem{duo_systematicperformanceevaluation_2018}
+	Angelo Du{\a`o}, Mark~D. Robinson, and Charlotte Soneson.
+	\newblock A systematic performance evaluation of clustering methods for
+	single-cell {{RNA}}-seq data.
+	\newblock {\em F1000Research}, 7:1141, 2018.
+
+	\bibitem{ewing_combiningtumorgenome_2015}
+	Adam~D. Ewing, Kathleen~E. Houlahan, Yin Hu, Kyle Ellrott, Cristian Caloian,
+	Takafumi~N. Yamaguchi, J.~Christopher Bare, Christine P'ng, Daryl Waggott,
+	Veronica~Y. Sabelnykova, {ICGC-TCGA DREAM Somatic Mutation Calling Challenge
+		Participants}, Liu Xi, Ninad Dewal, Yu~Fan, Wenyi Wang, David Wheeler,
+	Andreas Wilm, Grace~Hui Ting, Chenhao Li, Denis Bertrand, Niranjan Nagarajan,
+	Qing-Rong Chen, Chih-Hao Hsu, Ying Hu, Chunhua Yan, Warren Kibbe, Daoud
+	Meerzaman, Kristian Cibulskis, Mara Rosenberg, Louis Bergelson, Adam Kiezun,
+	Amie Radenbaugh, Anne-Sophie Sertier, Anthony Ferrari, Laurie Tonton, Kunal
+	Bhutani, Nancy~F. Hansen, Difei Wang, Lei Song, Zhongwu Lai, Yang Liao, Wei
+	Shi, Jos{\a'e} Carbonell-Caballero, Joaqu\'in Dopazo, Cheryl C.~K. Lau,
+	Justin Guinney, Michael~R. Kellen, Thea~C. Norman, David Haussler, Stephen~H.
+	Friend, Gustavo Stolovitzky, Adam~A. Margolin, Joshua~M. Stuart, and Paul~C.
+	Boutros.
+	\newblock Combining tumor genome simulation with crowdsourcing to benchmark
+	somatic single-nucleotide-variant detection.
+	\newblock {\em Nature Methods}, 12(7):623--630, 7 2015.
+
+	\bibitem{ferrell_bistabilitybifurcationswaddington_2012}
+	James~E Ferrell.
+	\newblock Bistability, {{Bifurcations}}, and {{Waddington}}'s {{Epigenetic
+			Landscape}}.
+	\newblock {\em Current Biology}, 22(11):R458--R466, 6 2012.
+
+	\bibitem{gibson_probabilisticmodelprokaryotic_2000}
+	Michael~A. Gibson and Jehoshua Bruck.
+	\newblock A probabilistic model of a prokaryotic gene and its regulation.
+	\newblock {\em Computational Methods in Molecular Biology: From Genotype to
+		Phenotype, MIT press, Cambridge}, 2000.
+
+	\bibitem{gillespie_exactstochasticsimulation_1977}
+	Daniel~T. Gillespie.
+	\newblock Exact stochastic simulation of coupled chemical reactions.
+	\newblock {\em The Journal of Physical Chemistry}, 81(25):2340--2361, 12 1977.
+
+	\bibitem{giorgino_computingvisualizingdynamic_2009}
+	Toni Giorgino.
+	\newblock Computing and {{Visualizing Dynamic Time Warping Alignments}} in
+	{{R}}: {{The}} dtw {{Package}}.
+	\newblock {\em Journal of Statistical Software}, (7), 9 2009.
+
+	\bibitem{glass_passingmessagesbiological_2013}
+	Kimberly Glass, Curtis Huttenhower, John Quackenbush, and Guo-Cheng Yuan.
+	\newblock Passing {{Messages}} between {{Biological Networks}} to {{Refine
+			Predicted Interactions}}.
+	\newblock {\em PLOS ONE}, 8(5):e64832, 5 2013.
+
+	\bibitem{graf_forcingcellschange_2009}
+	Thomas Graf and Tariq Enver.
+	\newblock Forcing {{Cells}} to {{Change Lineages}}.
+	\newblock {\em Nature}, 462(7273):587, 12 2009.
+
+	\bibitem{hache_gengesystematicgeneration_2009}
+	Hendrik Hache, Christoph Wierling, Hans Lehrach, and Ralf Herwig.
+	\newblock {{GeNGe}}: Systematic generation of gene regulatory networks.
+	\newblock {\em Bioinformatics}, 25(9):1205--1207, 5 2009.
+
+	\bibitem{holm_simplesequentiallyrejective_1979}
+	Sture Holm.
+	\newblock A simple sequentially rejective multiple test procedure.
+	\newblock {\em Scandinavian journal of statistics}, pages 65--70, 1979.
+
+	\bibitem{huynh-thu_inferringregulatorynetworks_2010}
+	V{\^a}n~Anh Huynh-Thu, Irrthum Alexandre, Wehenkel Louis, Geurts Pierre,
+	Alexandre Irrthum, Louis Wehenkel, and Pierre Geurts.
+	\newblock Inferring {{Regulatory Networks}} from {{Expression Data Using
+			Tree}}-{{Based Methods}}.
+	\newblock {\em PLoS ONE}, 5(9):e12776, 1 2010.
+
+	\bibitem{kanton_organoidsinglecellgenomic_2019}
+	Sabina Kanton, Michael~James Boyle, Zhisong He, Malgorzata Santel, Anne
+	Weigert, F{\a'a}tima Sanch\'is-Calleja, Patricia Guijarro, Leila Sidow,
+	Jonas~Simon Fleck, Dingding Han, Zhengzong Qian, Michael Heide, Wieland~B.
+	Huttner, Philipp Khaitovich, Svante P{\"a}{{\"a}}bo, Barbara Treutlein, and
+	J.~Gray Camp.
+	\newblock Organoid single-cell genomic atlas uncovers human-specific features
+	of brain development.
+	\newblock {\em Nature}, 574(7778):418--422, 10 2019.
+
+	\bibitem{kuijjer_lionessrsinglesample_2019}
+	Marieke~L. Kuijjer, Ping-Han Hsieh, John Quackenbush, and Kimberly Glass.
+	\newblock {{lionessR}}: Single sample network inference in {{R}}.
+	\newblock {\em BMC Cancer}, 19(1):1003, 10 2019.
+
+	\bibitem{kuijjer_estimatingsamplespecificregulatory_2015}
+	Marieke~Lydia Kuijjer, Matthew Tung, GuoCheng Yuan, John Quackenbush, and
+	Kimberly Glass.
+	\newblock Estimating sample-specific regulatory networks.
+	\newblock pages 1--19, 2015.
+
+	\bibitem{kuijjer_estimatingsamplespecificregulatory_2019}
+	Marieke~Lydia Kuijjer, Matthew~George Tung, GuoCheng Yuan, John Quackenbush,
+	and Kimberly Glass.
+	\newblock Estimating {{Sample}}-{{Specific Regulatory Networks}}.
+	\newblock {\em iScience}, 14:226--240, 3 2019.
+
+	\bibitem{lamanno_rnavelocitysingle_2018}
+	Gioele {La Manno}, Ruslan Soldatov, Amit Zeisel, Emelie Braun, Hannah
+	Hochgerner, Viktor Petukhov, Katja Lidschreiber, Maria~E Kastriti, Peter
+	L{\"o}nnerberg, Alessandro Furlan, Jean Fan, Lars~E Borm, Zehua Liu, David
+	{van Bruggen}, Jimin Guo, Xiaoling He, Roger Barker, Erik Sundstr{\"o}m,
+	Gon{\c c}alo Castelo-Branco, Patrick Cramer, Igor Adameyko, Sten Linnarsson,
+	and Peter~V Kharchenko.
+	\newblock {{RNA Velocity}} of {{Single Cells}}.
+	\newblock {\em Nature}, 560(7719):494--498, 8 2018.
+
+	\bibitem{lin_scclassifysamplesize_2020}
+	Yingxin Lin, Yue Cao, Hani~Jieun Kim, Agus Salim, Terence~P Speed, David~M Lin,
+	Pengyi Yang, and Jean Yee~Hwa Yang.
+	\newblock {{scClassify}}: Sample size estimation and multiscale classification
+	of cells using single and multiple reference.
+	\newblock {\em Molecular Systems Biology}, 16(6):e9389, 6 2020.
+
+	\bibitem{liu_personalizedcharacterizationdiseases_2016}
+	Xiaoping Liu, Yuetong Wang, Hongbin Ji, Kazuyuki Aihara, and Luonan Chen.
+	\newblock Personalized characterization of diseases using sample-specific
+	networks.
+	\newblock {\em Nucleic Acids Research}, 44(22):e164--e164, 2016.
+
+	\bibitem{lizio_gatewaysfantom5promoter_2015}
+	Marina Lizio, Jayson Harshbarger, Hisashi Shimoji, Jessica Severin, Takeya
+	Kasukawa, Serkan Sahin, Imad Abugessaisa, Shiro Fukuda, Fumi Hori, Sachi
+	Ishikawa-Kato, Christopher~J. Mungall, Erik Arner, J.~Kenneth Baillie,
+	Nicolas Bertin, Hidemasa Bono, Michiel {de Hoon}, Alexander~D. Diehl,
+	Emmanuel Dimont, Tom~C. Freeman, Kaori Fujieda, Winston Hide, Rajaram
+	Kaliyaperumal, Toshiaki Katayama, Timo Lassmann, Terrence~F. Meehan, Koro
+	Nishikata, Hiromasa Ono, Michael Rehli, Albin Sandelin, Erik~A. Schultes,
+	Peter~AC {`t Hoen}, Zuotian Tatum, Mark Thompson, Tetsuro Toyoda, Derek~W.
+	Wright, Carsten~O. Daub, Masayoshi Itoh, Piero Carninci, Yoshihide
+	Hayashizaki, Alistair~RR Forrest, Hideya Kawaji, and {the FANTOM consortium}.
+	\newblock Gateways to the {{FANTOM5}} promoter level mammalian expression
+	atlas.
+	\newblock {\em Genome Biology}, 16(1):22, 1 2015.
+
+	\bibitem{luecken_currentbestpractices_2019}
+	Malte~D Luecken and Fabian~J Theis.
+	\newblock Current best practices in single-cell {{RNA}}-seq analysis: A
+	tutorial.
+	\newblock {\em Molecular Systems Biology}, 15(6):e8746, 6 2019.
+
+	\bibitem{lummertzdarocha_reconstructioncomplexsinglecell_2018}
+	Edroaldo {Lummertz da Rocha}, R.~Grant Rowe, Vanessa Lundin, Mohan Malleshaiah,
+	Deepak~Kumar Jha, Carlos~R. Rambo, Hu~Li, Trista~E. North, James~J. Collins,
+	and George~Q. Daley.
+	\newblock Reconstruction of complex single-cell trajectories using
+	{{CellRouter}}.
+	\newblock {\em Nature Communications}, 9(1):892, 3 2018.
+
+	\bibitem{marbach_wisdomcrowdsrobust_2012}
+	Daniel Marbach, James~C Costello, Robert K{\"u}ffner, Nicole M~Nicci Vega,
+	Robert~J Prill, Diogo~M Camacho, Kyle~R Allison, Manolis Kellis, James~J
+	Collins, Gustavo Stolovitzky, {DREAM5 Consortium}, Manolis Kellis, James~J
+	Collins, and Gustavo Stolovitzky.
+	\newblock Wisdom of {{Crowds}} for {{Robust Gene Network Inference}}.
+	\newblock {\em Nature methods}, 9(8):796--804, 7 2012.
+
+	\bibitem{marbach_revealingstrengthsweaknesses_2010}
+	Daniel Marbach, Robert~J Prill, Thomas Schaffter, Claudio Mattiussi, Dario
+	Floreano, and Gustavo Stolovitzky.
+	\newblock Revealing strengths and weaknesses of methods for gene network
+	inference.
+	\newblock {\em Proceedings of the National Academy of Sciences},
+	107(14):6286--6291, 4 2010.
+
+	\bibitem{marbach_generatingrealisticsilico_2009}
+	Daniel Marbach, Thomas Schaffter, Claudio Mattiussi, and Dario Floreano.
+	\newblock Generating {{Realistic In Silico Gene Networks}} for {{Performance
+			Assessment}} of {{Reverse Engineering Methods}}.
+	\newblock {\em Journal of Computational Biology}, 16(2):229--239, 2 2009.
+
+	\bibitem{mcfaline-figueroa_pooledsinglecellgenetic_2019}
+	Jos{\a'e}~L. McFaline-Figueroa, Andrew~J. Hill, Xiaojie Qiu, Dana Jackson, Jay
+	Shendure, and Cole Trapnell.
+	\newblock A pooled single-cell genetic screen identifies regulatory checkpoints
+	in the continuum of the epithelial-to-mesenchymal transition.
+	\newblock {\em Nature Genetics}, 51(9):1389--1398, 9 2019.
+
+	\bibitem{moerman_grnboost2arboretoefficient_2019}
+	Thomas Moerman, Sara {Aibar Santos}, Carmen {Bravo Gonz{\a'a}lez-Blas}, Jaak
+	Simm, Yves Moreau, Jan Aerts, and Stein Aerts.
+	\newblock {{GRNBoost2}} and {{Arboreto}}: Efficient and scalable inference of
+	gene regulatory networks.
+	\newblock {\em Bioinformatics}, 35(12):2159--2161, 6 2019.
+
+	\bibitem{papadopoulos_prossttprobabilisticsimulation_2019}
+	Nikolaos Papadopoulos, Parra~R. Gonzalo, and Johannes S{\"o}ding.
+	\newblock {{PROSSTT}}: Probabilistic simulation of single-cell {{RNA}}-seq data
+	for complex differentiation processes.
+	\newblock {\em Bioinformatics (Oxford, England)}, 35(18):3517--3519, 9 2019.
+
+	\bibitem{parra_reconstructingcomplexlineage_2019}
+	R.~Gonzalo Parra, Nikolaos Papadopoulos, Laura Ahumada-Arranz, Jakob~El
+	Kholtei, Noah Mottelson, Yehor Horokhovsky, Barbara Treutlein, and Johannes
+	Soeding.
+	\newblock Reconstructing complex lineage trees from {{scRNA}}-seq data using
+	{{MERLoT}}.
+	\newblock {\em Nucleic Acids Research}, 47(17):8961--8974, 9 2019.
+
+	\bibitem{patil_pairwisecomparisonsmultiplepairwise_2019}
+	Indrajeet Patil.
+	\newblock {\em {{pairwiseComparisons}}: {{Multiple}} Pairwise Comparison
+		Tests}, 2019.
+
+	\bibitem{prill_rigorousassessmentsystems_2010}
+	Robert~J Prill, Marbach Daniel, Saez-Rodriguez Julio, Sorger~Peter K.,
+	Alexopoulos~Leonidas G., Xue Xiaowei, Clarke~Neil D., Altan-Bonnet Gregoire,
+	Stolovitzky Gustavo, Daniel Marbach, Julio Saez-Rodriguez, Peter~K Sorger,
+	Leonidas~G Alexopoulos, Xiaowei Xue, Neil~D Clarke, Gregoire Altan-Bonnet,
+	Gustavo Stolovitzky, Marbach Daniel, Saez-Rodriguez Julio, Sorger~Peter K.,
+	Alexopoulos~Leonidas G., Xue Xiaowei, Clarke~Neil D., Altan-Bonnet Gregoire,
+	and Stolovitzky Gustavo.
+	\newblock Towards a {{Rigorous Assessment}} of {{Systems Biology Models}}:
+	{{The DREAM3 Challenges}}.
+	\newblock {\em PLoS ONE}, 5(2):e9202, 1 2010.
+
+	\bibitem{rekhtman_directinteractionhematopoietic_1999}
+	N.~Rekhtman, F.~Radparvar, T.~Evans, and A.~I. Skoultchi.
+	\newblock Direct interaction of hematopoietic transcription factors {{PU}}.1
+	and {{GATA}}-1: Functional antagonism in erythroid cells.
+	\newblock {\em Genes \& Development}, 13(11):1398--1411, 6 1999.
+
+	\bibitem{roy_systemgeneratingtranscription_2008}
+	Sushmita Roy, Margaret Werner-Washburne, and Terran Lane.
+	\newblock A system for generating transcription regulatory networks with
+	combinatorial control of transcription.
+	\newblock {\em Bioinformatics}, 24(10):1318--1320, 5 2008.
+
+	\bibitem{saelens_comparisonsinglecelltrajectory_2019}
+	Wouter Saelens, Robrecht Cannoodt, Helena Todorov, and Yvan Saeys.
+	\newblock A comparison of single-cell trajectory inference methods.
+	\newblock {\em Nature Biotechnology}, 37(5), 2019.
+
+	\bibitem{schaffter_genenetweaversilicobenchmark_2011}
+	Thomas Schaffter, Daniel Marbach, and Dario Floreano.
+	\newblock {{GeneNetWeaver}}: In silico benchmark generation and performance
+	profiling of network inference methods.
+	\newblock {\em Bioinformatics}, 27(16):2263--2270, 8 2011.
+
+	\bibitem{schilstra_biologicgeneexpression_2008}
+	Maria~J. Schilstra and Chrystopher~L. Nehaniv.
+	\newblock Bio-{{Logic}}: {{Gene Expression}} and the {{Laws}} of
+	{{Combinatorial Logic}}.
+	\newblock {\em Artificial Life}, 14(1):121--133, 1 2008.
+
+	\bibitem{schwanhausser_globalquantificationmammalian_2011}
+	Bj{\"o}rn Schwanh{\"a}usser, Dorothea Busse, Na~Li, Gunnar Dittmar, Johannes
+	Schuchhardt, Jana Wolf, Wei Chen, and Matthias Selbach.
+	\newblock Global quantification of mammalian gene expression control.
+	\newblock {\em Nature}, 473(7347):337--342, 5 2011.
+
+	\bibitem{smith_spatialstochasticintracellular_2019}
+	Stephen Smith and Ramon Grima.
+	\newblock Spatial {{Stochastic Intracellular Kinetics}}: {{A Review}} of
+	{{Modelling Approaches}}.
+	\newblock {\em Bulletin of Mathematical Biology}, 81(8):2960--3009, 8 2019.
+
+	\bibitem{soneson_biasrobustnessscalability_2018}
+	Charlotte Soneson and Mark~D. Robinson.
+	\newblock Bias, robustness and scalability in single-cell differential
+	expression analysis.
+	\newblock {\em Nature Methods}, 15(4):255--261, 4 2018.
+
+	\bibitem{street_slingshotcelllineage_2018}
+	Kelly Street, Davide Risso, Russell~B Fletcher, Diya Das, John Ngai, Nir Yosef,
+	Elizabeth Purdom, and Sandrine Dudoit.
+	\newblock Slingshot: {{Cell Lineage}} and {{Pseudotime Inference}} for
+	{{Single}}-{{Cell Transcriptomics}}.
+	\newblock {\em BMC Genomics}, 19(1):477, 6 2018.
+
+	\bibitem{vallejos_normalizingsinglecellrna_2017}
+	Catalina~A. Vallejos, Davide Risso, Antonio Scialdone, Sandrine Dudoit, and
+	John~C. Marioni.
+	\newblock Normalizing single-cell {{RNA}} sequencing data: Challenges and
+	opportunities.
+	\newblock {\em Nature Methods}, 14(6):565--571, 6 2017.
+
+	\bibitem{vandesande_scalablescenicworkflow_2020}
+	Bram {Van de Sande}, Christopher Flerin, Kristofer Davie, Maxime {De
+		Waegeneer}, Gert Hulselmans, Sara Aibar, Ruth Seurinck, Wouter Saelens,
+	Robrecht Cannoodt, Quentin Rouchon, Toni Verbeiren, Dries {De Maeyer}, Joke
+	Reumers, Yvan Saeys, and Stein Aerts.
+	\newblock A scalable {{SCENIC}} workflow for single-cell gene regulatory
+	network analysis.
+	\newblock {\em Nature Protocols}, 15(7):2247--2276, 7 2020.
+
+	\bibitem{vandenbulcke_syntrengeneratorsynthetic_2006}
+	Tim {Van den Bulcke}, Koenraad {Van Leemput}, Bart Naudts, Piet {van Remortel},
+	Hongwu Ma, Alain Verschoren, Bart {De Moor}, and Kathleen Marchal.
+	\newblock {{SynTReN}}: A generator of synthetic gene expression data for design
+	and analysis of structure learning algorithms.
+	\newblock {\em BMC Bioinformatics}, 7(1):43, 1 2006.
+
+	\bibitem{vieth_powsimrpoweranalysis_2017}
+	Beate Vieth, Christoph Ziegenhain, Swati Parekh, Wolfgang Enard, and Ines
+	Hellmann.
+	\newblock {{powsimR}}: Power analysis for bulk and single cell {{RNA}}-seq
+	experiments.
+	\newblock {\em Bioinformatics}, 33(21):3486--3488, 11 2017.
+
+	\bibitem{wang_quantifyingwaddingtonlandscape_2011}
+	Jin Wang, Kun Zhang, Li~Xu, and Erkang Wang.
+	\newblock Quantifying the {{Waddington Landscape}} and {{Biological Paths}} for
+	{{Development}} and {{Differentiation}}.
+	\newblock {\em Proceedings of the National Academy of Sciences},
+	108(20):8257--8262, 5 2011.
+
+	\bibitem{weber_essentialguidelinescomputational_2019}
+	Lukas~M. Weber, Wouter Saelens, Robrecht Cannoodt, Charlotte Soneson, Alexander
+	Hapfelmeier, Paul~P. Gardner, Anne-Laure Boulesteix, Yvan Saeys, and Mark~D.
+	Robinson.
+	\newblock Essential guidelines for computational method benchmarking.
+	\newblock {\em Genome Biology}, 20(1):125, 6 2019.
+
+	\bibitem{wickham_welcometidyverse_2019}
+	Hadley Wickham, Mara Averick, Jennifer Bryan, Winston Chang, Lucy McGowan,
+	Romain Fran{\c c}ois, Garrett Grolemund, Alex Hayes, Lionel Henry, Jim
+	Hester, Max Kuhn, Thomas Pedersen, Evan Miller, Stephan Bache, Kirill
+	M{\"u}ller, Jeroen Ooms, David Robinson, Dana Seidel, Vitalie Spinu, Kohske
+	Takahashi, Davis Vaughan, Claus Wilke, Kara Woo, and Hiroaki Yutani.
+	\newblock Welcome to the {{Tidyverse}}.
+	\newblock 11 2019.
+
+	\bibitem{xu_regulationbifurcatingcell_2015}
+	Heping Xu, Virendra~K Chaudhri, Zhiguo Wu, Konstantinos Biliouris, Krista
+	Dienger-Stambaugh, Yrina Rochman, and Harinder Singh.
+	\newblock Regulation of bifurcating {{B}} cell trajectories by mutual
+	antagonism between transcription factors {{IRF4}} and {{IRF8}}.
+	\newblock {\em Nat. Immunol.}, 16(12):1274--1281, 12 2015.
+
+	\bibitem{yosef_dynamicregulatorynetwork_2013}
+	Nir Yosef, Alex~K Shalek, Jellert~T Gaublomme, Hulin Jin, Youjin Lee, Amit
+	Awasthi, Chuan Wu, Katarzyna Karwacz, Sheng Xiao, Marsela Jorgolli, David
+	Gennert, Rahul Satija, Arvind Shakya, Diana~Y Lu, John~J Trombetta, Meenu~R
+	Pillai, Peter~J Ratcliffe, Mathew~L Coleman, Mark Bix, Dean Tantin, Hongkun
+	Park, Vijay~K Kuchroo, and Aviv Regev.
+	\newblock Dynamic regulatory network controlling
+	\{\vphantom\}{{TH17}}\vphantom\{\} cell differentiation.
+	\newblock {\em Nature}, 496(7446):461--468, 2013.
+
+	\bibitem{zappia_splattersimulationsinglecell_2017}
+	Luke Zappia, Belinda Phipson, and Alicia Oshlack.
+	\newblock Splatter: {{Simulation}} of {{Single}}-{{Cell RNA Sequencing Data}}.
+	\newblock {\em Genome Biology}, 18:174, 9 2017.
+
+	\bibitem{zeisel_coupledpremrnamrna_2011}
+	Amit Zeisel, Wolfgang~J K{\"o}stler, Natali Molotski, Jonathan~M Tsai, Rita
+	Krauthgamer, Jasmine Jacob-Hirsch, Gideon Rechavi, Yoav Soen, Steffen Jung,
+	Yosef Yarden, and Eytan Domany.
+	\newblock Coupled pre-{{mRNA}} and {{mRNA}} dynamics unveil operational
+	strategies underlying transcriptional responses to stimuli.
+	\newblock {\em Molecular Systems Biology}, 7(1):529, 1 2011.
+
+	\bibitem{zhang_simulatingmultiplefaceted_2019}
+	Xiuwei Zhang, Chenling Xu, and Nir Yosef.
+	\newblock Simulating multiple faceted variability in single cell {{RNA}}
+	sequencing.
+	\newblock {\em Nature Communications}, 10(1):1--16, 6 2019.
+
+\end{thebibliography}
 

### 2021-05-12 manuscript/render_manuscript.R

@@ -35,12 +35,6 @@ knitcitations::write.bibtex(bib[bib_keys], file = "manuscript/knitcitations.bib"
 # render the manuscript
 render("manuscript/manuscript.Rmd")
 
-bib_replace <- paste(c("\\addbibresource{\\jobname.bib}", "\\begin{filecontents}{\\jobname.bib}", readr::read_lines("manuscript/knitcitations.bib"), "\\end{filecontents}"), collapse = "\n")
-
-read_lines("manuscript/manuscript.tex") %>%
-  gsub("\\addbibresource{library.bib}", bib_replace, ., fixed = TRUE) %>%
-  write_lines("manuscript/manuscript.tex")
-
 render("manuscript/supplementary_files.Rmd")
 
 # system("latexdiff manuscript/v3/render.tex manuscript/render.tex > manuscript/track_changes.tex")

### 2021-05-17 manuscript/manuscript.Rmd
update for v5
update

@@
       - '--lua-filter=lua/author-info-blocks.lua'
-biblatexoptions: sorting=none,url=false,backend=bibtex
+biblatexoptions: sorting=none,url=false
 classoption:
@@
   \captionsetup[table]{position=bottom}
-  \setmainfont [Path = fonts/,
-    UprightFont = *-300,
-    ItalicFont = *-300-Italic,
-    BoldFont = *-700,
-    BoldItalicFont = *-700-Italic
-  ]{MuseoSans}
 
@@
 
-\begin{eqnarray*}
+\begin{eqnarray}
   \text{regeffect}_G & = \frac{\text{proptrans}_G(S) - \text{proptrans}_G(S[\z T \leftarrow 0])}{\xpr G}
-\end{eqnarray*}
+\end{eqnarray}
 
@@
 * Y.S. supervised the project.
+
+# Acknowledgements
+This project has been made possible in part by grant number 2020-218899 from the Chan Zuckerberg
+Initiative DAF, an advised fund of Silicon Valley Community Foundation.

### 2021-05-17 manuscript/manuscript.tex
update

--- 2021-05-12 manuscript/manuscript.tex
@@
 \captionsetup[table]{position=bottom}
-\setmainfont [Path = fonts/,
-  UprightFont = *-300,
-  ItalicFont = *-300-Italic,
-  BoldFont = *-700,
-  BoldItalicFont = *-700-Italic
-]{MuseoSans}
 
@@
 \fi
-\usepackage[sorting=none,url=false,backend=bibtex]{biblatex}
-
-\usepackage{filecontents}
-
-\bibliography{\jobname}
+\usepackage[sorting=none,url=false]{biblatex}
+\addbibresource{library.bib}
 
@@
 \author{Robrecht Cannoodt\textsuperscript{$\dagger{}$,1,2,3} \and Wouter
-Saelens\textsuperscript{$\dagger{}$,1,2} \and Louise
+Saelens\textsuperscript{$\dagger{}$,1,2,4} \and Louise
 Deconinck\textsuperscript{1,2} \and Yvan Saeys\textsuperscript{1,2,*}}
-\date{12 May 2021}
+\date{17 May 2021}
 
@@
 
+\hypertarget{acknowledgements}{%
+\section{Acknowledgements}\label{acknowledgements}}
+
+This project has been made possible in part by grant number 2020-218899
+from the Chan Zuckerberg Initiative DAF, an advised fund of Silicon
+Valley Community Foundation.
+
 \printbibliography
 
-\begin{filecontents}{\jobname.bib}
-@Article{aibar_scenicsinglecellregulatory_2017,
-	title = {{{SCENIC}}: Single-Cell Regulatory Network Inference and Clustering},
-	author = {Sara Aibar and Carmen Bravo Gonz{\a'a}lez-Blas and Thomas Moerman and V{\^a}n Anh Huynh-Thu and Hana Imrichova and Gert Hulselmans and Florian Rambow and Jean-Christophe Marine and Pierre Geurts and Jan Aerts and Joost {van den Oord} and Zeynep Kalender Atak and Jasper Wouters and Stein Aerts},
-	year = {2017},
-	month = {oct},
-	issn = {1548-7091},
-	doi = {10.1038/nmeth.4463},
-	journal = {Nature Methods},
-}
-
-@Article{alpert_alignmentsinglecelltrajectories_2018,
-	title = {Alignment of Single-Cell Trajectories to Compare Cellular Expression Dynamics},
-	author = {Ayelet Alpert and Lindsay S. Moore and Tania Dubovik and Shai S. Shen-Orr},
-	year = {2018},
-	month = {apr},
-	volume = {15},
-	pages = {267--270},
-	issn = {1548-7105},
-	doi = {10.1038/nmeth.4628},
-	abstract = {cellAlign enables quantitative comparisons of expression dynamics within and between single-cell trajectories based on single-cell RNA-seq or mass cytometry data.},
-	copyright = {2018 Nature Publishing Group, a division of Macmillan Publishers Limited. All Rights Reserved.},
-	file = {/home/rcannood/Zotero/storage/QB7FH2T9/nmeth.html},
-	journal = {Nature Methods},
-	language = {en},
-	number = {4},
-}
-
-@Article{bergen_generalizingrnavelocity_2020,
-	title = {Generalizing {{RNA}} Velocity to Transient Cell States through Dynamical Modeling},
-	author = {Volker Bergen and Marius Lange and Stefan Peidli and F. Alexander Wolf and Fabian J. Theis},
-	year = {2020},
-	month = {dec},
-	volume = {38},
-	pages = {1408--1414},
-	issn = {1546-1696},
-	doi = {10.1038/s41587-020-0591-3},
-	abstract = {RNA velocity has opened up new ways of studying cellular differentiation in single-cell RNA-sequencing data.
-It describes the rate of gene expression change for an individual gene at a given time point based on the ratio of its spliced and unspliced messenger RNA (mRNA).
-However, errors in velocity estimates arise if the central assumptions of a common splicing rate and the observation of the full splicing dynamics with steady-state mRNA levels are violated.
-Here we present scVelo, a method that overcomes these limitations by solving the full transcriptional dynamics of splicing kinetics using a likelihood-based dynamical model.
-This generalizes RNA velocity to systems with transient cell states, which are common in development and in response to perturbations.
-We apply scVelo to disentangling subpopulation kinetics in neurogenesis and pancreatic endocrinogenesis.
-We infer gene-specific rates of transcription, splicing and degradation, recover each cell's position in the underlying differentiation processes and detect putative driver genes.
-scVelo will facilitate the study of lineage decisions and gene regulation.},
-	journal = {Nature Biotechnology},
-	number = {12},
-}
-
-@Article{bjornsson_digitaltwinspersonalize_2019,
-	title = {Digital Twins to Personalize Medicine},
-	author = {Bergthor Bj{\"o}rnsson and Carl Borrebaeck and Nils Elander and Thomas Gasslander and Danuta R.
-Gawel and Mika Gustafsson and Rebecka J{\"o}rnsten and Eun Jung Lee and Xinxiu Li and Sandra Lilja and David Mart\'inez-Enguita and Andreas Matussek and Per Sandstr{\"o}m and Samuel Sch{\"a}fer and Margaretha Stenmarker and X.
-F.
-Sun and Oleg Sysoev and Huan Zhang and Mikael Benson and {on behalf of the Swedish Digital Twin Consortium}},
-	year = {2019},
-	month = {dec},
-	volume = {12},
-	pages = {4},
-	issn = {1756-994X},
-	doi = {10.1186/s13073-019-0701-3},
-	abstract = {Personalized medicine requires the integration and processing of vast amounts of data.
-Here, we propose a solution to this challenge that is based on constructing Digital Twins.
-These are high-resolution models of individual patients that are computationally treated with thousands of drugs to find the drug that is optimal for the patient.},
-	journal = {Genome Medicine},
-	number = {1},
-}
-
-@Article{cacchiarelli_aligningsinglecelldevelopmental_2018,
-	title = {Aligning {{Single}}-{{Cell Developmental}} and {{Reprogramming Trajectories Identifies Molecular Determinants}} of {{Myogenic Reprogramming Outcome}}},
-	author = {Davide Cacchiarelli and Xiaojie Qiu and Sanjay Srivatsan and Anna Manfredi and Michael Ziller and Eliah Overbey and Antonio Grimaldi and Jonna Grimsby and Prapti Pokharel and Kenneth J.
-Livak and Shuqiang Li and Alexander Meissner and Tarjei S.
-Mikkelsen and John L.
-Rinn and Cole Trapnell},
-	year = {2018},
-	month = {sep},
-	volume = {7},
-	pages = {258-268.e3},
-	issn = {2405-4712},
-	doi = {10.1016/j.cels.2018.07.006},
-	abstract = {{$<$}h2{$>$}Summary{$<$}/h2{$><$}p{$>$}Cellular reprogramming through manipulation of defined factors holds great promise for large-scale production of cell types needed for use in therapy and for revealing principles of gene regulation.
-However, most reprogramming systems are inefficient, converting only a fraction of cells to the desired state.
-Here, we analyze MYOD-mediated reprogramming of human fibroblasts to myotubes, a well-characterized model system for direct conversion by defined factors, at pseudotemporal resolution using single-cell RNA-seq.
-To expose barriers to efficient conversion, we introduce a novel analytic technique, trajectory alignment, which enables quantitative comparison of gene expression kinetics across two biological processes.
-Reprogrammed cells navigate a trajectory with branch points that correspond to two alternative decision points, with cells that select incorrect branches terminating at aberrant or incomplete reprogramming outcomes.
-Analysis of these branch points revealed insulin and BMP signaling as crucial molecular determinants of reprogramming.
-Single-cell trajectory alignment enables rigorous quantitative comparisons between biological trajectories found in diverse processes in development, reprogramming, and other contexts.{$<$}/p{$>$}},
-	file = {/home/rcannood/Zotero/storage/YQH8GP2I/Cacchiarelli et al. - 2018 - Aligning Single-Cell Developmental and Reprogrammi.pdf;/home/rcannood/Zotero/storage/Q5MUK95E/S2405-4712(18)30314-4.html},
-	journal = {Cell Systems},
-	language = {English},
-	number = {3},
-	pmid = {30195438},
-}
-
-@Article{cannoodt_singlecellomicsdatasets_2018,
-	title = {Single-Cell -Omics Datasets Containing a Trajectory},
-	author = {Robrecht Cannoodt and Wouter Saelens and Helena Todorov and Yvan Saeys},
-	year = {2018},
-	month = {oct},
-	doi = {10.5281/zenodo.1211532},
-	journal = {Zenodo},
-}
-
-@TechReport{conover_multiplecomparisonsprocedures_1979,
-	title = {On Multiple-Comparisons Procedures},
-	author = {William Jay Conover and Ronald L Iman},
-	year = {1979},
-	institution = {{Technical report, Los Alamos Scientific Laboratory}},
-}
-
-@Article{duo_systematicperformanceevaluation_2018,
-	title = {A Systematic Performance Evaluation of Clustering Methods for Single-Cell {{RNA}}-Seq Data},
-	author = {Angelo Du{\a`o} and Mark D. Robinson and Charlotte Soneson},
-	year = {2018},
-	volume = {7},
-	pages = {1141},
-	issn = {2046-1402},
-	doi = {10.12688/f1000research.15666.2},
-	abstract = {Subpopulation identification, usually via some form of unsupervised clustering, is a fundamental step in the analysis of many single-cell RNA-seq data sets.
-This has motivated the development and application of a broad range of clustering methods, based on various underlying algorithms.
-Here, we provide a systematic and extensible performance evaluation of 14 clustering algorithms implemented in R, including both methods developed explicitly for scRNA-seq data and more general-purpose methods.
-The methods were evaluated using nine publicly available scRNA-seq data sets as well as three simulations with varying degree of cluster separability.
-The same feature selection approaches were used for all methods, allowing us to focus on the investigation of the performance of the clustering algorithms themselves.
-We evaluated the ability of recovering known subpopulations, the stability and the run time and scalability of the methods.
-Additionally, we investigated whether the performance could be improved by generating consensus partitions from multiple individual clustering methods.
-We found substantial differences in the performance, run time and stability between the methods, with SC3 and Seurat showing the most favorable results.
-Additionally, we found that consensus clustering typically did not improve the performance compared to the best of the combined methods, but that several of the top-performing methods already perform some type of consensus clustering.
-All the code used for the evaluation is available on GitHub ( https://github.com/markrobinsonuzh/scRNAseq\_clustering\_comparison).
-In addition, an R package providing access to data and clustering results, thereby facilitating inclusion of new methods and data sets, is available from Bioconductor ( https://bioconductor.org/packages/DuoClustering2018).},
-	file = {/home/rcannood/Zotero/storage/ASWVFH87/Duò et al. - 2018 - A systematic performance evaluation of clustering .pdf},
-	journal = {F1000Research},
-	keywords = {Algorithms,Benchmarking,Cluster Analysis,Clustering,Clustering methods,Gene Expression Profiling,Humans,RNA-seq,RNA; Small Cytoplasmic,Sequence Analysis; RNA,Single-Cell Analysis,Single-cell RNA-seq,Software},
-	language = {eng},
-	pmcid = {PMC6134335},
-	pmid = {30271584},
-}
-
-@Article{ewing_combiningtumorgenome_2015,
-	title = {Combining Tumor Genome Simulation with Crowdsourcing to Benchmark Somatic Single-Nucleotide-Variant Detection},
-	author = {Adam D.
-Ewing and Kathleen E.
-Houlahan and Yin Hu and Kyle Ellrott and Cristian Caloian and Takafumi N.
-Yamaguchi and J.
-Christopher Bare and Christine P'ng and Daryl Waggott and Veronica Y.
-Sabelnykova and {ICGC-TCGA DREAM Somatic Mutation Calling Challenge Participants} and Liu Xi and Ninad Dewal and Yu Fan and Wenyi Wang and David Wheeler and Andreas Wilm and Grace Hui Ting and Chenhao Li and Denis Bertrand and Niranjan Nagarajan and Qing-Rong Chen and Chih-Hao Hsu and Ying Hu and Chunhua Yan and Warren Kibbe and Daoud Meerzaman and Kristian Cibulskis and Mara Rosenberg and Louis Bergelson and Adam Kiezun and Amie Radenbaugh and Anne-Sophie Sertier and Anthony Ferrari and Laurie Tonton and Kunal Bhutani and Nancy F.
-Hansen and Difei Wang and Lei Song and Zhongwu Lai and Yang Liao and Wei Shi and Jos{\a'e} Carbonell-Caballero and Joaqu\'in Dopazo and Cheryl C.
-K.
-Lau and Justin Guinney and Michael R.
-Kellen and Thea C.
-Norman and David Haussler and Stephen H.
-Friend and Gustavo Stolovitzky and Adam A.
-Margolin and Joshua M.
-Stuart and Paul C.
-Boutros},
-	year = {2015},
-	month = {jul},
-	volume = {12},
-	pages = {623--630},
-	issn = {1548-7105},
-	doi = {10.1038/nmeth.3407},
-	abstract = {The detection of somatic mutations from cancer genome sequences is key to understanding the genetic basis of disease progression, patient survival and response to therapy.
-Benchmarking is needed for tool assessment and improvement but is complicated by a lack of gold standards, by extensive resource requirements and by difficulties in sharing personal genomic information.
-To resolve these issues, we launched the ICGC-TCGA DREAM Somatic Mutation Calling Challenge, a crowdsourced benchmark of somatic mutation detection algorithms.
-Here we report the BAMSurgeon tool for simulating cancer genomes and the results of 248 analyses of three in silico tumors created with it.
-Different algorithms exhibit characteristic error profiles, and, intriguingly, false positives show a trinucleotide profile very similar to one found in human tumors.
-Although the three simulated tumors differ in sequence contamination (deviation from normal cell sequence) and in subclonality, an ensemble of pipelines outperforms the best individual pipeline in all cases.
-BAMSurgeon is available at https://github.com/adamewing/bamsurgeon/.},
-	copyright = {2015 Nature Publishing Group},
-	file = {/home/rcannood/Zotero/storage/NBPKWHK3/Ewing et al. - 2015 - Combining tumor genome simulation with crowdsourci.pdf;/home/rcannood/Zotero/storage/TDZU2TL9/nmeth.html},
-	journal = {Nature Methods},
-	language = {en},
-	number = {7},
-}
-
-@Article{ferrell_bistabilitybifurcationswaddington_2012,
-	title = {Bistability, {{Bifurcations}}, and {{Waddington}}'s {{Epigenetic Landscape}}},
-	author = {James E Ferrell},
-	year = {2012},
-	month = {jun},
-	volume = {22},
-	pages = {R458--R466},
-	issn = {0960-9822},
-	doi = {10.1016/j.cub.2012.03.045},
-	abstract = {Waddington's epigenetic landscape is probably the most famous and most powerful metaphor in developmental biology.
-Cells, represented by balls, roll downhill through a landscape of bifurcating valleys.
-Each new valley represents a possible cell fate and the ridges between the valleys maintain the cell fate once it has been chosen.
-Here I examine models of two important developmental processes \$\textbackslash backslash\$textemdash\{\} cell-fate induction and lateral inhibition \$\textbackslash backslash\$textemdash\{\} and ask whether the landscapes for these models at least qualitatively resemble Waddington's picture.
-For cell-fate induction, the answer is no.
-The commitment of a cell to a new fate corresponds to the disappearance of a valley from the landscape, not the splitting of one valley into two, and it occurs through a type of bifurcation \$\textbackslash backslash\$textemdash\{\} a saddle-node bifurcation \$\textbackslash backslash\$textemdash\{\} that possesses an intrinsic irreversibility that is missing from Waddington's picture.
-Lateral inhibition, a symmetrical cell\$\textbackslash backslash\$textendash\{\}cell competition process, corresponds better to Waddington's picture, with one valley reversibly splitting into two through a pitchfork bifurcation.
-I propose an alternative epigenetic landscape that has numerous valleys and ridges right from the start, with the process of cell-fate commitment corresponding to the irreversible disappearance of some of these valleys and ridges, via cell-fate induction, complemented by the creation of new valleys and ridges through processes like cell\$\textbackslash backslash\$textendash\{\}cell competition.},
-	file = {/home/rcannood/Zotero/storage/GK4TJ5BV/Ferrell - 2012 - Bistability, Bifurcations, and Waddington's Epigen.pdf;/home/rcannood/Zotero/storage/HUSMJA22/Ferrell - 2012 - Bistability, Bifurcations, and Waddington's Epigen.pdf;/home/rcannood/Zotero/storage/8VLPBTG2/S0960-9822(12)00332-6.html;/home/rcannood/Zotero/storage/XBTP42Y6/S0960982212003326.html},
-	journal = {Current Biology},
-	number = {11},
-}
-
-@Article{gibson_probabilisticmodelprokaryotic_2000,
-	title = {A Probabilistic Model of a Prokaryotic Gene and Its Regulation},
-	author = {Michael A. Gibson and Jehoshua Bruck},
-	year = {2000},
-	journal = {Computational Methods in Molecular Biology: From Genotype to Phenotype, MIT press, Cambridge},
-}
-
-@Article{gillespie_exactstochasticsimulation_1977,
-	title = {Exact Stochastic Simulation of Coupled Chemical Reactions},
-	author = {Daniel T. Gillespie},
-	year = {1977},
-	month = {dec},
-	volume = {81},
-	pages = {2340--2361},
-	issn = {0022-3654},
-	doi = {10.1021/j100540a008},
-	file = {/home/rcannood/Zotero/storage/ELV9AQRF/j100540a008.html},
-	journal = {The Journal of Physical Chemistry},
-	number = {25},
-}
-
-@Article{giorgino_computingvisualizingdynamic_2009,
-	title = {Computing and {{Visualizing Dynamic Time Warping Alignments}} in {{R}}: {{The}} Dtw {{Package}}},
-	author = {Toni Giorgino},
-	year = {2009},
-	month = {sep},
-	doi = {10.18637/jss.v031.i07},
-	abstract = {Dynamic time warping is a popular technique for comparing time series, providing both a distance measure that is insensitive to local compression and stretches and the warping which optimally deforms one of the two input series onto the other.
-A variety of algorithms and constraints have been discussed in the literature.
-The dtw package provides an unification of them; it allows R users to compute time series alignments mixing freely a variety of continuity constraints, restriction windows, endpoints, local distance definitions, and so on.
-The package also provides functions for visualizing alignments and constraints using several classic diagram types.},
-	journal = {Journal of Statistical Software},
-	number = {7},
-}
-
-@Article{glass_passingmessagesbiological_2013,
-	title = {Passing {{Messages}} between {{Biological Networks}} to {{Refine Predicted Interactions}}},
-	author = {Kimberly Glass and Curtis Huttenhower and John Quackenbush and Guo-Cheng Yuan},
-	year = {2013},
-	month = {may},
-	volume = {8},
-	pages = {e64832},
-	doi = {10.1371/journal.pone.0064832},
-	abstract = {Regulatory network reconstruction is a fundamental problem in computational biology.
-There are significant limitations to such reconstruction using individual datasets, and increasingly people attempt to construct networks using multiple, independent datasets obtained from complementary sources, but methods for this integration are lacking.
-We developed PANDA (Passing Attributes between Networks for Data Assimilation), a message-passing model using multiple sources of information to predict regulatory relationships, and used it to integrate protein-protein interaction, gene expression, and sequence motif data to reconstruct genome-wide, condition-specific regulatory networks in yeast as a model.
-The resulting networks were not only more accurate than those produced using individual data sets and other existing methods, but they also captured information regarding specific biological mechanisms and pathways that were missed using other methodologies.
-PANDA is scalable to higher eukaryotes, applicable to specific tissue or cell type data and conceptually generalizable to include a variety of regulatory, interaction, expression, and other genome-scale data.
-An implementation of the PANDA algorithm is available at www.sourceforge.net/projects/panda-net.},
-	journal = {PLOS ONE},
-	number = {5},
-}
-
-@Article{graf_forcingcellschange_2009,
-	title = {Forcing {{Cells}} to {{Change Lineages}}},
-	author = {Thomas Graf and Tariq Enver},
-	year = {2009},
-	month = {dec},
-	volume = {462},
-	pages = {587},
-	issn = {1476-4687},
-	doi = {10.1038/nature08533},
-	abstract = {\$\textbackslash textless\$p\$\textbackslash textgreater\$Though it is important for the success of multicellular organisms that their many different cell types remain stable once specified, advances in the production of induced pluripotent stem (iPS) cells from somatic cells by exposure to a transcription factor cocktail emphasize the potential of almost any cell to be reprogrammed.
-This raises the question of whether interconversion between cell types occurs physiologically or in the context of disease.
-Thomas Graf and Tariq Enver use examples from the blood cell system to illustrate the principles of transcription factor transdifferentiation, which they conclude is a process fundamentally similar to differentiation and different from iPS reprogramming.
-They postulate that conversions of one cell type to another can occur as a normal developmental process, and speculate on the prospects of forced lineage reprogramming in regenerative medicine.\$\textbackslash textless\$/p\$\textbackslash textgreater\$},
-	journal = {Nature},
-	number = {7273},
-}
-
-@Article{hache_gengesystematicgeneration_2009,
-	title = {{{GeNGe}}: Systematic Generation of Gene Regulatory Networks},
-	shorttitle = {{{GeNGe}}},
-	author = {Hendrik Hache and Christoph Wierling and Hans Lehrach and Ralf Herwig},
-	year = {2009},
-	month = {may},
-	volume = {25},
-	pages = {1205--1207},
-	issn = {1367-4803},
-	doi = {10.1093/bioinformatics/btp115},
-	abstract = {Summary: The analysis of gene regulatory networks (GRNs) is a central goal of bioinformatics highly accelerated by the advent of new experimental techniques, such as RNA interference.
-A battery of reverse engineering methods has been developed in recent years to reconstruct the underlying GRNs from these and other experimental data.
-However, the performance of the individual methods is poorly understood and validation of algorithmic performances is still missing to a large extent.
-To enable such systematic validation, we have developed the web application GeNGe (GEne Network GEnerator), a controlled framework for the automatic generation of GRNs.
-The theoretical model for a GRN is a non-linear differential equation system.
-Networks can be user-defined or constructed in a modular way with the option to introduce global and local network perturbations.
-Resulting data can be used, e.g.
-as benchmark data for evaluating GRN reconstruction methods or for predicting effects of perturbations as theoretical counterparts of biological experiments.Availability: Available online at http://genge.molgen.mpg.deContact:hache@molgen.mpg.deSupplementary information:Supplementary data are available at Bioinformatics online.},
-	journal = {Bioinformatics},
-	number = {9},
-}
-
-@Article{holm_simplesequentiallyrejective_1979,
-	title = {A Simple Sequentially Rejective Multiple Test Procedure},
-	author = {Sture Holm},
-	year = {1979},
-	pages = {65--70},
-	journal = {Scandinavian journal of statistics},
-}
-
-@Article{huynh-thu_inferringregulatorynetworks_2010,
-	title = {Inferring {{Regulatory Networks}} from {{Expression Data Using Tree}}-{{Based Methods}}},
-	author = {V{\^a}n Anh Huynh-Thu and Irrthum Alexandre and Wehenkel Louis and Geurts Pierre and Alexandre Irrthum and Louis Wehenkel and Pierre Geurts},
-	year = {2010},
-	month = {jan},
-	volume = {5},
-	pages = {e12776},
-	issn = {1932-6203},
-	doi = {10.1371/journal.pone.0012776},
-	abstract = {One of the pressing open problems of computational systems biology is the elucidation of the topology of genetic regulatory networks (GRNs) using high throughput genomic data, in particular microarray gene expression data.
-The Dialogue for Reverse Engineering Assessments and Methods (DREAM) challenge aims to evaluate the success of GRN inference algorithms on benchmarks of simulated data.
-In this article, we present GENIE3, a new algorithm for the inference of GRNs that was best performer in the DREAM4 In Silico Multifactorial challenge.
-GENIE3 decomposes the prediction of a regulatory network between p genes into p different regression problems.
-In each of the regression problems, the expression pattern of one of the genes (target gene) is predicted from the expression patterns of all the other genes (input genes), using tree-based ensemble methods Random Forests or Extra-Trees.
-The importance of an input gene in the prediction of the target gene expression pattern is taken as an indication of a putative regulatory link.
-Putative regulatory links are then aggregated over all genes to provide a ranking of interactions from which the whole network is reconstructed.
-In addition to performing well on the DREAM4 In Silico Multifactorial challenge simulated data, we show that GENIE3 compares favorably with existing algorithms to decipher the genetic regulatory network of Escherichia coli.
-It doesn't make any assumption about the nature of gene regulation, can deal with combinatorial and non-linear interactions, produces directed GRNs, and is fast and scalable.
-In conclusion, we propose a new algorithm for GRN inference that performs well on both synthetic and real gene expression data.
-The algorithm, based on feature selection with tree-based ensemble methods, is simple and generic, making it adaptable to other types of genomic data and interactions.},
-	file = {/home/rcannood/Zotero/storage/3I6PLCQL/Huynh-Thu et al. - 2010 - Inferring Regulatory Networks from Expression Data Using Tree-Based Methods.pdf},
-	journal = {PLoS ONE},
-	keywords = {Algorithms,Computational Biology,Computational Biology: methods,Escherichia coli,Escherichia coli: genetics,Gene Expression Regulation,Gene Regulatory Networks,Oligonucleotide Array Sequence Analysis},
-	number = {9},
-	pmid = {20927193},
-}
-
-@Article{kanton_organoidsinglecellgenomic_2019,
-	title = {Organoid Single-Cell Genomic Atlas Uncovers Human-Specific Features of Brain Development},
-	author = {Sabina Kanton and Michael James Boyle and Zhisong He and Malgorzata Santel and Anne Weigert and F{\a'a}tima Sanch\'is-Calleja and Patricia Guijarro and Leila Sidow and Jonas Simon Fleck and Dingding Han and Zhengzong Qian and Michael Heide and Wieland B.
-Huttner and Philipp Khaitovich and Svante P{\"a}{{\"a}}bo and Barbara Treutlein and J.
-Gray Camp},
-	year = {2019},
-	month = {oct},
-	volume = {574},
-	pages = {418--422},
-	issn = {1476-4687},
-	doi = {10.1038/s41586-019-1654-9},
-	abstract = {Species comparisons using single-cell transcriptomics and accessible chromatin profiling in stem cell-derived cerebral organoids are used to map dynamic gene-regulatory changes that are unique to humans.},
-	copyright = {2019 The Author(s), under exclusive licence to Springer Nature Limited},
-	file = {/home/rcannood/Zotero/storage/AVPYT7HY/s41586-019-1654-9.html},
-	journal = {Nature},
-	language = {en},
-	number = {7778},
-}
-
-@Article{kuijjer_estimatingsamplespecificregulatory_2015,
-	title = {Estimating Sample-Specific Regulatory Networks},
-	author = {Marieke Lydia Kuijjer and Matthew Tung and GuoCheng Yuan and John Quackenbush and Kimberly Glass},
-	year = {2015},
-	pages = {1--19},
-	abstract = {Biological systems are driven by intricate interactions among the complex array of molecules that comprise the cell.
-Many methods have been developed to reconstruct network models that attempt to capture those interactions.
-These methods often draw on large numbers of measured expression samples to tease out subtle signals and infer connections between genes (or gene products).
-The result is an aggregate network model representing a single estimate for edge likelihoods.
-While informative, aggregate models fail to capture the heterogeneity that is often represented in a population.
-Here we propose a method to reverse engineer sample-specific networks from aggregate network models.
-We demonstrate the accuracy and applicability of our approach in several datasets, including simulated data, microarray expression data from synchronized yeast cells, and RNA-seq data collected from human subjects.
-We show that these sample-specific networks can be used to study the evolution of network topology across time and to characterize shifts in gene regulation that may not be apparent in the expression data.
-We believe the ability to generate sample-specific networks will revolutionize the field of network biology and has the potential to usher in an era of precision network medicine.},
-	file = {/home/rcannood/Zotero/storage/YECTK66M/Kuijjer et al. - 2015 - Estimating sample-specific regulatory networks.pdf},
-}
-
-@Article{kuijjer_estimatingsamplespecificregulatory_2019,
-	title = {Estimating {{Sample}}-{{Specific Regulatory Networks}}},
-	author = {Marieke Lydia Kuijjer and Matthew George Tung and GuoCheng Yuan and John Quackenbush and Kimberly Glass},
-	year = {2019},
-	month = {mar},
-	volume = {14},
-	pages = {226--240},
-	issn = {2589-0042},
-	doi = {10.1016/j.isci.2019.03.021},
-	abstract = {Biological systems are driven by intricate interactions among molecules.
-Many methods have been developed that draw on large numbers of expression samples to infer connections between genes (or their products).
-The result is an aggregate network representing a single estimate for the likelihood of each interaction, or ``edge,'' in the network.
-Although informative, aggregate models fail to capture population heterogeneity.
-Here we propose a method to reverse engineer sample-specific networks from aggregate networks.
-We demonstrate our approach in several contexts, including simulated, yeast microarray, and human lymphoblastoid cell line RNA sequencing data.
-We use these sample-specific networks to study changes in network topology across time and to characterize shifts in gene regulation that were not apparent in the expression data.
-We believe that generating sample-specific networks will greatly facilitate the application of network methods to large, complex, and heterogeneous multi-omic datasets, supporting the emerging field of precision network medicine.,                                        \textbullet{}               We developed LIONESS to extract single-sample networks from aggregate models                                         \textbullet{}               We tested LIONESS using in silico, yeast, and human expression data                                         \textbullet{}               LIONESS-estimated networks are reproducible, accurate, and biologically meaningful                                         \textbullet{}               Single-sample network analysis highlights important biological processes                                 , Biological Sciences; Bioinformatics; Complex Systems},
-	file = {/home/rcannood/Zotero/storage/ITHUQ7NH/Kuijjer et al. - 2019 - Estimating Sample-Specific Regulatory Networks.pdf},
-	journal = {iScience},
-	pmcid = {PMC6463816},
-	pmid = {30981959},
-}
-
-@Article{kuijjer_lionessrsinglesample_2019,
-	title = {{{lionessR}}: Single Sample Network Inference in {{R}}},
-	shorttitle = {{{lionessR}}},
-	author = {Marieke L. Kuijjer and Ping-Han Hsieh and John Quackenbush and Kimberly Glass},
-	year = {2019},
-	month = {oct},
-	volume = {19},
-	pages = {1003},
-	issn = {1471-2407},
-	doi = {10.1186/s12885-019-6235-7},
-	abstract = {In biomedical research, network inference algorithms are typically used to infer complex association patterns between biological entities, such as between genes or proteins, using data from a population.
-This resulting aggregate network, in essence, averages over the networks of those individuals in the population.
-LIONESS (Linear Interpolation to Obtain Network Estimates for Single Samples) is a method that can be used together with a network inference algorithm to extract networks for individual samples in a population.
-The method's key characteristic is that, by modeling networks for individual samples in a data set, it can capture network heterogeneity in a population.
-LIONESS was originally made available as a function within the PANDA (Passing Attributes between Networks for Data Assimilation) regulatory network reconstruction framework.
-However, the LIONESS algorithm is generalizable and can be used to model single sample networks based on a wide range of network inference algorithms.},
-	file = {/home/rcannood/Zotero/storage/RS448RZF/Kuijjer et al. - 2019 - lionessR single sample network inference in R.pdf;/home/rcannood/Zotero/storage/9STH6ZYN/s12885-019-6235-7.html},
-	journal = {BMC Cancer},
-	number = {1},
-}
-
-@Article{lamanno_rnavelocitysingle_2018,
-	title = {{{RNA Velocity}} of {{Single Cells}}},
-	author = {Gioele {La Manno} and Ruslan Soldatov and Amit Zeisel and Emelie Braun and Hannah Hochgerner and Viktor Petukhov and Katja Lidschreiber and Maria E Kastriti and Peter L{\"o}nnerberg and Alessandro Furlan and Jean Fan and Lars E Borm and Zehua Liu and David {van Bruggen} and Jimin Guo and Xiaoling He and Roger Barker and Erik Sundstr{\"o}m and Gon{\c c}alo Castelo-Branco and Patrick Cramer and Igor Adameyko and Sten Linnarsson and Peter V Kharchenko},
-	year = {2018},
-	month = {aug},
-	volume = {560},
-	pages = {494--498},
-	issn = {1476-4687},
-	doi = {10.1038/s41586-018-0414-6},
-	abstract = {RNA velocity, estimated in single cells by comparison of spliced and unspliced mRNA, is a good indicator of transcriptome dynamics and will provide a useful tool for analysis of developmental lineage.},
-	journal = {Nature},
-	number = {7719},
-}
-
-@Article{lin_scclassifysamplesize_2020,
-	title = {{{scClassify}}: Sample Size Estimation and Multiscale Classification of Cells Using Single and Multiple Reference},
-	author = {Yingxin Lin and Yue Cao and Hani Jieun Kim and Agus Salim and Terence P Speed and David M Lin and Pengyi Yang and Jean Yee Hwa Yang},
-	year = {2020},
-	month = {jun},
-	volume = {16},
-	pages = {e9389},
-	publisher = {{John Wiley \& Sons, Ltd}},
-	issn = {1744-4292},
-	doi = {10.15252/msb.20199389},
-	abstract = {Abstract Automated cell type identification is a key computational challenge in single-cell RNA-sequencing (scRNA-seq) data.
-To capitalise on the large collection of well-annotated scRNA-seq datasets, we developed scClassify, a multiscale classification framework based on ensemble learning and cell type hierarchies constructed from single or multiple annotated datasets as references.
-scClassify enables the estimation of sample size required for accurate classification of cell types in a cell type hierarchy and allows joint classification of cells when multiple references are available.
-We show that scClassify consistently performs better than other supervised cell type classification methods across 114 pairs of reference and testing data, representing a diverse combination of sizes, technologies and levels of complexity, and further demonstrate the unique components of scClassify through simulations and compendia of experimental datasets.
-Finally, we demonstrate the scalability of scClassify on large single-cell atlases and highlight a novel application of identifying subpopulations of cells from the Tabula Muris data that were unidentified in the original publication.
-Together, scClassify represents state-of-the-art methodology in automated cell type identification from scRNA-seq data.},
-	journal = {Molecular Systems Biology},
-	keywords = {cell type hierarchy,cell type identification,multiscale classification,sample size estimation,single-cell},
-	number = {6},
-}
-
-@Article{liu_personalizedcharacterizationdiseases_2016,
-	title = {Personalized Characterization of Diseases Using Sample-Specific Networks},
-	author = {Xiaoping Liu and Yuetong Wang and Hongbin Ji and Kazuyuki Aihara and Luonan Chen},
-	year = {2016},
-	volume = {44},
-	pages = {e164--e164},
-	issn = {0305-1048},
-	doi = {10.1093/nar/gkw772},
-	abstract = {A complex disease generally results not from malfunction of individual molecules but from dysfunction of the relevant system or network, which dynamically changes with time and conditions.
-Thus, estimating a condition-specific network from a single sample is crucial to elucidating the molecular mechanisms of complex diseases at the system level.
-However, there is currently no effective way to construct such an individual-specific network by expression profiling of a single sample because of the requirement of multiple samples for computing correlations.
-We developed here with a statistical method, i.e.
-a sample-specific network (SSN) method, which allows us to construct individual-specific networks based on molecular expressions of a single sample.
-Using this method, we can characterize various human diseases at a network level.
-In particular, such SSNs can lead to the identification of individual-specific disease modules as well as driver genes, even without gene sequencing information.
-Extensive analysis by using the Cancer Genome Atlas data not only demonstrated the effectiveness of the method, but also found new individual-specific driver genes and network patterns for various types of cancer.
-Biological experiments on drug resistance further validated one important advantage of our method over the traditional methods, i.e.
-we can even identify such drug resistance genes that actually have no clear differential expression between samples with and without the resistance, due to the additional network information.},
-	file = {/home/rcannood/Zotero/storage/6MSYIL7R/Liu et al. - 2016 - Personalized characterization of diseases using sample-specific networks.pdf},
-	journal = {Nucleic Acids Research},
-	keywords = {cancer,drug resistance,genes,steroid-sensitive nephrotic syndrome},
-	number = {22},
-	pmid = {27596597},
-}
-
-@Article{luecken_currentbestpractices_2019,
-	title = {Current Best Practices in Single-Cell {{RNA}}-Seq Analysis: A Tutorial},
-	shorttitle = {Current Best Practices in Single-Cell {{RNA}}-Seq Analysis},
-	author = {Malte D Luecken and Fabian J Theis},
-	year = {2019},
-	month = {jun},
-	volume = {15},
-	pages = {e8746},
-	issn = {1744-4292},
-	doi = {10.15252/msb.20188746},
-	abstract = {Abstract Single-cell RNA-seq has enabled gene expression to be studied at an unprecedented resolution.
-The promise of this technology is attracting a growing user base for single-cell analysis methods.
-As more analysis tools are becoming available, it is becoming increasingly difficult to navigate this landscape and produce an up-to-date workflow to analyse one's data.
-Here, we detail the steps of a typical single-cell RNA-seq analysis, including pre-processing (quality control, normalization, data correction, feature selection, and dimensionality reduction) and cell- and gene-level downstream analysis.
-We formulate current best-practice recommendations for these steps based on independent comparison studies.
-We have integrated these best-practice recommendations into a workflow, which we apply to a public dataset to further illustrate how these steps work in practice.
-Our documented case study can be found at https://www.github.com/theislab/single-cell-tutorial.
-This review will serve as a workflow tutorial for new entrants into the field, and help established users update their analysis pipelines.},
-	file = {/home/rcannood/Zotero/storage/MY4TZUJT/Luecken and Theis - 2019 - Current best practices in single-cell RNA-seq anal.pdf;/home/rcannood/Zotero/storage/7PFR3DKU/msb.html},
-	journal = {Molecular Systems Biology},
-	keywords = {analysis pipeline development,computational biology,data analysis tutorial,single-cell RNA-seq},
-	number = {6},
-}
-
-@Article{lummertzdarocha_reconstructioncomplexsinglecell_2018,
-	title = {Reconstruction of Complex Single-Cell Trajectories Using {{CellRouter}}},
-	author = {Edroaldo {Lummertz da Rocha} and R.
-Grant Rowe and Vanessa Lundin and Mohan Malleshaiah and Deepak Kumar Jha and Carlos R.
-Rambo and Hu Li and Trista E.
-North and James J.
-Collins and George Q.
-Daley},
-	year = {2018},
-	month = {mar},
-	volume = {9},
-	pages = {892},
-	issn = {2041-1723},
-	doi = {10.1038/s41467-018-03214-y},
-	abstract = {A better understanding of the cell-fate transitions that occur in complex cellular ecosystems in normal development and disease could inform cell engineering efforts and lead to improved therapies.
-However, a major challenge is to simultaneously identify new cell states, and their transitions, to elucidate the gene expression dynamics governing cell-type diversification.
-Here, we present CellRouter, a multifaceted single-cell analysis platform that identifies complex cell-state transition trajectories by using flow networks to explore the subpopulation structure of multi-dimensional, single-cell omics data.
-We demonstrate its versatility by applying CellRouter to single-cell RNA sequencing data sets to reconstruct cell-state transition trajectories during hematopoietic stem and progenitor cell (HSPC) differentiation to the erythroid, myeloid and lymphoid lineages, as well as during re-specification of cell identity by cellular reprogramming of monocytes and B-cells to HSPCs.
-CellRouter opens previously undescribed paths for in-depth characterization of complex cellular ecosystems and establishment of enhanced cell engineering approaches.},
-	journal = {Nature Communications},
-	number = {1},
-}
-
-@Article{marbach_generatingrealisticsilico_2009,
-	title = {Generating {{Realistic In Silico Gene Networks}} for {{Performance Assessment}} of {{Reverse Engineering Methods}}},
-	author = {Daniel Marbach and Thomas Schaffter and Claudio Mattiussi and Dario Floreano},
-	year = {2009},
-	month = {feb},
-	volume = {16},
-	pages = {229--239},
-	issn = {1557-8666},
-	doi = {10.1089/cmb.2008.09TT},
-	abstract = {Reverse engineering methods are typically first tested on simulated data from in silico networks, for systematic and efficient performance assessment, before an application to real biological networks.
-In this paper, we present a method for generating biologically plausible in silico networks, which allow realistic performance assessment of network inference algorithms.
-Instead of using random graph models, which are known to only partly capture the structural properties of biological networks, we generate network structures by extracting modules from known biological interaction networks.
-Using the yeast transcriptional regulatory network as a test case, we show that extracted modules have a biologically plausible connectivity because they preserve functional and structural properties of the original network.
-Our method was selected to generate the {"}gold standard{"} networks for the gene network reverse engineering challenge of the third DREAM conference (Dialogue on Reverse Engineering Assessment and Methods 2008, Cambridge, MA).},
-	file = {/home/rcannood/Zotero/storage/NXLVA76Q/Marbach et al. - 2009 - Generating Realistic In Silico Gene Networks for Performance Assessment of Reverse Engineering Methods.pdf},
-	journal = {Journal of Computational Biology},
-	keywords = {Algorithms,Computational Biology,Computational Biology: methods,Computer Simulation,DRE,Gene Expression Profiling,Gene Expression Profiling: methods,gene regulatory networks,Gene Regulatory Networks,Genetic,Models,reverse engineering,Saccharomyces cerevisiae,Saccharomyces cerevisiae: genetics,Saccharomyces cerevisiae: metabolism,Software},
-	number = {2},
-	pmid = {19183003},
-}
-
-@Article{marbach_revealingstrengthsweaknesses_2010,
-	title = {Revealing Strengths and Weaknesses of Methods for Gene Network Inference},
-	author = {Daniel Marbach and Robert J Prill and Thomas Schaffter and Claudio Mattiussi and Dario Floreano and Gustavo Stolovitzky},
-	year = {2010},
-	month = {apr},
-	volume = {107},
-	pages = {6286--6291},
-	issn = {1091-6490},
-	doi = {10.1073/pnas.0913357107},
-	abstract = {Numerous methods have been developed for inferring gene regulatory networks from expression data, however, both their absolute and comparative performance remain poorly understood.
-In this paper, we introduce a framework for critical performance assessment of methods for gene network inference.
-We present an in silico benchmark suite that we provided as a blinded, community-wide challenge within the context of the DREAM (Dialogue on Reverse Engineering Assessment and Methods) project.
-We assess the performance of 29 gene-network-inference methods, which have been applied independently by participating teams.
-Performance profiling reveals that current inference methods are affected, to various degrees, by different types of systematic prediction errors.
-In particular, all but the best-performing method failed to accurately infer multiple regulatory inputs (combinatorial regulation) of genes.
-The results of this community-wide experiment show that reliable network inference from gene expression data remains an unsolved problem, and they indicate potential ways of network reconstruction improvements.},
-	file = {/home/rcannood/Zotero/storage/784SY83Y/Marbach et al. - 2010 - Revealing strengths and weaknesses of methods for gene network inference.pdf},
-	journal = {Proceedings of the National Academy of Sciences},
-	keywords = {Biometry,community experiment,Computational Biology,Computational Biology: methods,DREAM challenge,Gene Expression Profiling,Gene Regulatory Networks,performance assessment,reverse engineering,transcriptional regulatory networks},
-	number = {14},
-	pmid = {20308593},
-}
-
-@Article{marbach_wisdomcrowdsrobust_2012,
-	title = {Wisdom of {{Crowds}} for {{Robust Gene Network Inference}}},
-	author = {Daniel Marbach and James C Costello and Robert K{\"u}ffner and Nicole M Nicci Vega and Robert J Prill and Diogo M Camacho and Kyle R Allison and Manolis Kellis and James J Collins and Gustavo Stolovitzky and {DREAM5 Consortium} and Manolis Kellis and James J Collins and Gustavo Stolovitzky},
-	year = {2012},
-	month = {jul},
-	volume = {9},
-	pages = {796--804},
-	issn = {1548-7091},
-	doi = {10.1038/nmeth.2016},
-	abstract = {Reconstructing gene regulatory networks from high-throughput data is a long-standing problem.
-Through the DREAM project (Dialogue on Reverse Engineering Assessment and Methods), we performed a comprehensive blind assessment of over thirty network inference methods on Escherichia coli, Staphylococcus aureus, Saccharomyces cerevisiae, and in silico microarray data.
-We characterize performance, data requirements, and inherent biases of different inference approaches offering guidelines for both algorithm application and development.
-We observe that no single inference method performs optimally across all datasets.
-In contrast, integration of predictions from multiple inference methods shows robust and high performance across diverse datasets.
-Thereby, we construct high-confidence networks for E.
-coli and S.
-aureus, each comprising \$\textbackslash backslash\$textasciitilde\{\}1700 transcriptional interactions at an estimated precision of 50\%.
-We experimentally test 53 novel interactions in E.
-coli, of which 23 were supported (43\%).
-Our results establish community-based methods as a powerful and robust tool for the inference of transcriptional gene regulatory networks.},
-	file = {/home/rcannood/Zotero/storage/AQRNS57R/Marbach et al.
-- 2012 - Wisdom of crowds for robust gene network inference.pdf;/home/rcannood/Zotero/storage/KGJVQ9WL/Marbach et al.
-- 2012 - Wisdom of crowds for robust gene network inference.pdf},
-	journal = {Nature methods},
-	keywords = {Algorithms,Bacterial,Bacterial: genetics,Computational Biology,Escherichia coli,Escherichia coli: genetics,Gene Expression Regulation,Gene Regulatory Networks,Genetic,Genetic: genetics,Oligonucleotide Array Sequence Analysis,Saccharomyces cerevisiae,Saccharomyces cerevisiae: genetics,Software,Staphylococcus aureus,Staphylococcus aureus: genetics,Transcription},
-	number = {8},
-	pmid = {22796662},
-}
-
-@Article{mcfaline-figueroa_pooledsinglecellgenetic_2019,
-	title = {A Pooled Single-Cell Genetic Screen Identifies Regulatory Checkpoints in the Continuum of the Epithelial-to-Mesenchymal Transition},
-	author = {Jos{\a'e} L. McFaline-Figueroa and Andrew J. Hill and Xiaojie Qiu and Dana Jackson and Jay Shendure and Cole Trapnell},
-	year = {2019},
-	month = {sep},
-	volume = {51},
-	pages = {1389--1398},
-	issn = {1546-1718},
-	doi = {10.1038/s41588-019-0489-5},
-	abstract = {Transcriptional profiling of 61,052 cells undergoing epithelial-to-mesenchymal transition (EMT) shows continuous waves of gene regulation, not discrete stages.
-A CRISPR screen orders EMT regulators along EMT progression.},
-	copyright = {2019 The Author(s), under exclusive licence to Springer Nature America, Inc.},
-	file = {/home/rcannood/Zotero/storage/RAQRP6UC/s41588-019-0489-5.html},
-	journal = {Nature Genetics},
-	language = {en},
-	number = {9},
-}
-
-@Article{moerman_grnboost2arboretoefficient_2019,
-	title = {{{GRNBoost2}} and {{Arboreto}}: Efficient and Scalable Inference of Gene Regulatory Networks},
-	author = {Thomas Moerman and Sara {Aibar Santos} and Carmen {Bravo Gonz{\a'a}lez-Blas} and Jaak Simm and Yves Moreau and Jan Aerts and Stein Aerts},
-	year = {2019},
-	month = {jun},
-	volume = {35},
-	pages = {2159--2161},
-	issn = {1367-4803},
-	doi = {10.1093/bioinformatics/bty916},
-	abstract = {Inferring a Gene Regulatory Network (GRN) from gene expression data is a computationally expensive task, exacerbated by increasing data sizes due to advances in high-throughput gene profiling technology, such as single-cell RNA-seq.
-To equip researchers with a toolset to infer GRNs from large expression datasets, we propose GRNBoost2 and the Arboreto framework.
-GRNBoost2 is an efficient algorithm for regulatory network inference using gradient boosting, based on the GENIE3 architecture.
-Arboreto is a computational framework that scales up GRN inference algorithms complying with this architecture.
-Arboreto includes both GRNBoost2 and an improved implementation of GENIE3, as a user-friendly open source Python package.Arboreto is available under the 3-Clause BSD license at http://arboreto.readthedocs.io.Supplementary data are available at Bioinformatics online.},
-	journal = {Bioinformatics},
-	number = {12},
-}
-
-@Article{papadopoulos_prossttprobabilisticsimulation_2019,
-	title = {{{PROSSTT}}: Probabilistic Simulation of Single-Cell {{RNA}}-Seq Data for Complex Differentiation Processes},
-	shorttitle = {{{PROSSTT}}},
-	author = {Nikolaos Papadopoulos and Parra R. Gonzalo and Johannes S{\"o}ding},
-	year = {2019},
-	month = {sep},
-	volume = {35},
-	pages = {3517--3519},
-	issn = {1367-4811},
-	doi = {10.1093/bioinformatics/btz078},
-	abstract = {SUMMARY: Cellular lineage trees can be derived from single-cell RNA sequencing snapshots of differentiating cells.
-Currently, only datasets with simple topologies are available.
-To test and further develop tools for lineage tree reconstruction, we need test datasets with known complex topologies.
-PROSSTT can simulate scRNA-seq datasets for differentiation processes with lineage trees of any desired complexity, noise level, noise model and size.
-PROSSTT also provides scripts to quantify the quality of predicted lineage trees.
-AVAILABILITY AND IMPLEMENTATION: https://github.com/soedinglab/prosstt.
-SUPPLEMENTARY INFORMATION: Supplementary data are available at Bioinformatics online.},
-	file = {/home/rcannood/Zotero/storage/UERERPJX/Papadopoulos et al. - 2019 - PROSSTT probabilistic simulation of single-cell R.pdf},
-	journal = {Bioinformatics (Oxford, England)},
-	keywords = {Cell Differentiation,Gene Expression Profiling,RNA-Seq,Single-Cell Analysis,Software},
-	language = {eng},
-	number = {18},
-	pmcid = {PMC6748774},
-	pmid = {30715210},
-}
-
-@Article{parra_reconstructingcomplexlineage_2019,
-	title = {Reconstructing Complex Lineage Trees from {{scRNA}}-Seq Data Using {{MERLoT}}},
-	author = {R. Gonzalo Parra and Nikolaos Papadopoulos and Laura Ahumada-Arranz and Jakob El Kholtei and Noah Mottelson and Yehor Horokhovsky and Barbara Treutlein and Johannes Soeding},
-	year = {2019},
-	month = {sep},
-	volume = {47},
-	pages = {8961--8974},
-	issn = {1362-4962},
-	doi = {10.1093/nar/gkz706},
-	abstract = {Advances in single-cell transcriptomics techniques are revolutionizing studies of cellular differentiation and heterogeneity.
-It has become possible to track the trajectory of thousands of genes across the cellular lineage trees that represent the temporal emergence of cell types during dynamic processes.
-However, reconstruction of cellular lineage trees with more than a few cell fates has proved challenging.
-We present MERLoT (https://github.com/soedinglab/merlot), a flexible and user-friendly tool to reconstruct complex lineage trees from single-cell transcriptomics data.
-It can impute temporal gene expression profiles along the reconstructed tree.
-We show MERLoT's capabilities on various real cases and hundreds of simulated datasets.},
-	file = {/home/rcannood/Zotero/storage/86VN2MRY/Parra et al. - 2019 - Reconstructing complex lineage trees from scRNA-se.pdf},
-	journal = {Nucleic Acids Research},
-	keywords = {Algorithms,Animals,Cell Differentiation,Cell Lineage,Cluster Analysis,Gene Expression Profiling,Hematopoietic Stem Cells,Humans,Mice,Mouse Embryonic Stem Cells,Sequence Analysis; RNA,Single-Cell Analysis,Software,Time Factors,Transcriptome},
-	language = {eng},
-	number = {17},
-	pmcid = {PMC6755101},
-	pmid = {31428793},
-}
-
-@Manual{patil_pairwisecomparisonsmultiplepairwise_2019,
-	title = {{{pairwiseComparisons}}: {{Multiple}} Pairwise Comparison Tests},
-	author = {Indrajeet Patil},
-	year = {2019},
-	type = {Manual},
-}
-
-@Article{prill_rigorousassessmentsystems_2010,
-	title = {Towards a {{Rigorous Assessment}} of {{Systems Biology Models}}: {{The DREAM3 Challenges}}},
-	author = {Robert J Prill and Marbach Daniel and Saez-Rodriguez Julio and Sorger Peter K.
-and Alexopoulos Leonidas G.
-and Xue Xiaowei and Clarke Neil D.
-and Altan-Bonnet Gregoire and Stolovitzky Gustavo and Daniel Marbach and Julio Saez-Rodriguez and Peter K Sorger and Leonidas G Alexopoulos and Xiaowei Xue and Neil D Clarke and Gregoire Altan-Bonnet and Gustavo Stolovitzky and Marbach Daniel and Saez-Rodriguez Julio and Sorger Peter K.
-and Alexopoulos Leonidas G.
-and Xue Xiaowei and Clarke Neil D.
-and Altan-Bonnet Gregoire and Stolovitzky Gustavo},
-	year = {2010},
-	month = {jan},
-	volume = {5},
-	pages = {e9202},
-	issn = {1932-6203},
-	doi = {10.1371/journal.pone.0009202},
-	abstract = {BACKGROUND: Systems biology has embraced computational modeling in response to the quantitative nature and increasing scale of contemporary data sets.
-The onslaught of data is accelerating as molecular profiling technology evolves.
-The Dialogue for Reverse Engineering Assessments and Methods (DREAM) is a community effort to catalyze discussion about the design, application, and assessment of systems biology models through annual reverse-engineering challenges.
-METHODOLOGY AND PRINCIPAL FINDINGS: We describe our assessments of the four challenges associated with the third DREAM conference which came to be known as the DREAM3 challenges: signaling cascade identification, signaling response prediction, gene expression prediction, and the DREAM3 in silico network challenge.
-The challenges, based on anonymized data sets, tested participants in network inference and prediction of measurements.
-Forty teams submitted 413 predicted networks and measurement test sets.
-Overall, a handful of best-performer teams were identified, while a majority of teams made predictions that were equivalent to random.
-Counterintuitively, combining the predictions of multiple teams (including the weaker teams) can in some cases improve predictive power beyond that of any single method.
-CONCLUSIONS: DREAM provides valuable feedback to practitioners of systems biology modeling.
-Lessons learned from the predictions of the community provide much-needed context for interpreting claims of efficacy of algorithms described in the scientific literature.},
-	file = {/home/rcannood/Zotero/storage/Q5E22H2F/Prill et al. - 2010 - Towards a Rigorous Assessment of Systems Biology Models The DREAM3 Challenges.pdf},
-	journal = {PLoS ONE},
-	keywords = {Algorithms,Animals,Biological,Cluster Analysis,Computational Biology,Computational Biology: methods,Gene Expression Profiling,Gene Expression Profiling: methods,Gene Regulatory Networks,Humans,Models,Protein Interaction Mapping,Protein Interaction Mapping: methods,Reproducibility of Results,Signal Transduction,Software,Systems Biology,Systems Biology: methods},
-	number = {2},
-	pmid = {20186320},
-}
-
-@Article{rekhtman_directinteractionhematopoietic_1999,
-	title = {Direct Interaction of Hematopoietic Transcription Factors {{PU}}.1 and {{GATA}}-1: Functional Antagonism in Erythroid Cells},
-	shorttitle = {Direct Interaction of Hematopoietic Transcription Factors {{PU}}.1 and {{GATA}}-1},
-	author = {N. Rekhtman and F. Radparvar and T. Evans and A. I. Skoultchi},
-	year = {1999},
-	month = {jun},
-	volume = {13},
-	pages = {1398--1411},
-	issn = {0890-9369},
-	doi = {10.1101/gad.13.11.1398},
-	abstract = {Malignant transformation usually inhibits terminal cell differentiation but the precise mechanisms involved are not understood.
-PU.1 is a hematopoietic-specific Ets family transcription factor that is required for development of some lymphoid and myeloid lineages.
-PU.1 can also act as an oncoprotein as activation of its expression in erythroid precursors by proviral insertion or transgenesis causes erythroleukemias in mice.
-Restoration of terminal differentiation in the mouse erythroleukemia (MEL) cells requires a decline in the level of PU.1, indicating that PU.1 can block erythroid differentiation.
-Here we investigate the mechanism by which PU.1 interferes with erythroid differentiation.
-We find that PU.1 interacts directly with GATA-1, a zinc finger transcription factor required for erythroid differentiation.
-Interaction between PU.1 and GATA-1 requires intact DNA-binding domains in both proteins.
-PU.1 represses GATA-1-mediated transcriptional activation.
-Both the DNA binding and transactivation domains of PU.1 are required for repression and both domains are also needed to block terminal differentiation in MEL cells.
-We also show that ectopic expression of PU.1 in Xenopus embryos is sufficient to block erythropoiesis during normal development.
-Furthermore, introduction of exogenous GATA-1 in both MEL cells and Xenopus embryos and explants relieves the block to erythroid differentiation imposed by PU.1.
-Our results indicate that the stoichiometry of directly interacting but opposing transcription factors may be a crucial determinant governing processes of normal differentiation and malignant transformation.},
-	file = {/home/rcannood/Zotero/storage/77DJ34BY/Rekhtman et al. - 1999 - Direct interaction of hematopoietic transcription .pdf},
-	journal = {Genes \& Development},
-	keywords = {3T3 Cells,Animals,Binding Sites,Cell Differentiation,DNA-Binding Proteins,Erythroid Precursor Cells,Erythroid-Specific DNA-Binding Factors,GATA1 Transcription Factor,HeLa Cells,Hematopoiesis,Humans,Mice,Proto-Oncogene Proteins,Recombinant Fusion Proteins,Trans-Activators,Transcription Factors,Transcription; Genetic,Transcriptional Activation,Xenopus laevis,Zinc Fingers},
-	language = {eng},
-	number = {11},
-	pmcid = {PMC316770},
-	pmid = {10364157},
-}
-
-@Article{roy_systemgeneratingtranscription_2008,
-	title = {A System for Generating Transcription Regulatory Networks with Combinatorial Control of Transcription},
-	author = {Sushmita Roy and Margaret Werner-Washburne and Terran Lane},
-	year = {2008},
-	month = {may},
-	volume = {24},
-	pages = {1318--1320},
-	issn = {1367-4803},
-	doi = {10.1093/bioinformatics/btn126},
-	abstract = {Summary: We have developed a new software system, REgulatory Network generator with COmbinatorial control (RENCO), for automatic generation of differential equations describing pre-transcriptional combinatorics in artificial regulatory networks.
-RENCO has the following benefits: (a) it explicitly models protein\textendash protein interactions among transcription factors, (b) it captures combinatorial control of transcription factors on target genes and (c) it produces output in Systems Biology Markup Language (SBML) format, which allows these equations to be directly imported into existing simulators.
-Explicit modeling of the protein interactions allows RENCO to incorporate greater mechanistic detail of the transcription machinery compared to existing models and can provide a better assessment of algorithms for regulatory network inference.Availability: RENCO is a C++ command line program, available at http://sourceforge.net/projects/renco/Contact:terran@cs.unm.eduSupplementary information: Supplementary data are available at Bioinformatics online.},
-	journal = {Bioinformatics},
-	number = {10},
-}
-
-@Article{saelens_comparisonsinglecelltrajectory_2019,
-	title = {A Comparison of Single-Cell Trajectory Inference Methods},
-	author = {Wouter Saelens and Robrecht Cannoodt and Helena Todorov and Yvan Saeys},
-	year = {2019},
-	volume = {37},
-	issn = {15461696},
-	doi = {10.1038/s41587-019-0071-9},
-	abstract = {Trajectory inference approaches analyze genome-wide omics data from thousands of single cells and computationally infer the order of these cells along developmental trajectories.
-Although more than 70 trajectory inference tools have already been developed, it is challenging to compare their performance because the input they require and output models they produce vary substantially.
-Here, we benchmark 45 of these methods on 110 real and 229 synthetic datasets for cellular ordering, topology, scalability and usability.
-Our results highlight the complementarity of existing tools, and that the choice of method should depend mostly on the dataset dimensions and trajectory topology.
-Based on these results, we develop a set of guidelines to help users select the best method for their dataset.
-Our freely available data and evaluation pipeline ( https://benchmark.dynverse.org ) will aid in the development of improved tools designed to analyze increasingly large and complex single-cell datasets.},
-	file = {/home/rcannood/Zotero/storage/JKZCFAVA/Saelens et al. - 2019 - A comparison of single-cell trajectory inference methods.pdf;/home/rcannood/Zotero/storage/T9HMNIDB/Saelens2019.pdf},
-	journal = {Nature Biotechnology},
-	number = {May},
-}
-
-@Article{schaffter_genenetweaversilicobenchmark_2011,
-	title = {{{GeneNetWeaver}}: In Silico Benchmark Generation and Performance Profiling of Network Inference Methods.},
-	shorttitle = {\{\vphantom\}{{GeneNetWeaver}}\vphantom\{\}},
-	author = {Thomas Schaffter and Daniel Marbach and Dario Floreano},
-	year = {2011},
-	month = {aug},
-	volume = {27},
-	pages = {2263--2270},
-	issn = {1367-4811},
-	doi = {10.1093/bioinformatics/btr373},
-	abstract = {MOTIVATION: Over the last decade, numerous methods have been developed for inference of regulatory networks from gene expression data.
-However, accurate and systematic evaluation of these methods is hampered by the difficulty of constructing adequate benchmarks and the lack of tools for a differentiated analysis of network predictions on such benchmarks.
-RESULTS: Here, we describe a novel and comprehensive method for in silico benchmark generation and performance profiling of network inference methods available to the community as an open-source software called GeneNetWeaver (GNW).
-In addition to the generation of detailed dynamical models of gene regulatory networks to be used as benchmarks, GNW provides a network motif analysis that reveals systematic prediction errors, thereby indicating potential ways of improving inference methods.
-The accuracy of network inference methods is evaluated using standard metrics such as precision-recall and receiver operating characteristic curves.
-We show how GNW can be used to assess the performance and identify the strengths and weaknesses of six inference methods.
-Furthermore, we used GNW to provide the international Dialogue for Reverse Engineering Assessments and Methods (DREAM) competition with three network inference challenges (DREAM3, DREAM4 and DREAM5).
-AVAILABILITY: GNW is available at http://gnw.sourceforge.net along with its Java source code, user manual and supporting data.
-SUPPLEMENTARY INFORMATION: Supplementary data are available at Bioinformatics online.
-CONTACT: dario.floreano@epfl.ch.},
-	file = {/home/rcannood/Zotero/storage/P5JMJUBN/Schaffter, Marbach, Floreano - 2011 - GeneNetWeaver in silico benchmark generation and performance profiling of network inference method.pdf},
-	journal = {Bioinformatics},
-	keywords = {Benchmarking,Escherichia coli,Escherichia coli: genetics,Gene Expression,Gene Regulatory Networks,Genetic,Models,Software},
-	number = {16},
-	pmid = {21697125},
-}
-
-@Article{schilstra_biologicgeneexpression_2008,
-	title = {Bio-{{Logic}}: {{Gene Expression}} and the {{Laws}} of {{Combinatorial Logic}}},
-	shorttitle = {Bio-{{Logic}}},
-	author = {Maria J. Schilstra and Chrystopher L. Nehaniv},
-	year = {2008},
-	month = {jan},
-	volume = {14},
-	pages = {121--133},
-	issn = {1064-5462},
-	doi = {10.1162/artl.2008.14.1.121},
-	abstract = {Abstract At the heart of the development of fertilized eggs into fully formed organisms and the adaptation of cells to changed conditions are genetic regulatory networks (GRNs).
-In higher multicellular organisms, signal selection and multiplexing are performed at the cis-regulatory domains of genes, where combinations of transcription factors (TFs) regulate the rates at which the genes are transcribed into mRNA.
-To be able to act as activators or repressors of gene transcription, TFs must first bind to target sequences on the regulatory domains.
-Two TFs that act in concert may bind entirely independently of each other, but more often binding of the first one will alter the affinity of the other for its binding site.
-This article presents a systematic investigation into the effect of TF binding dependences on the predicted regulatory function of this bio-logic.
-Four extreme scenarios, commonly used to classify enzyme activation and inhibition patterns, for the binding of two TFs were explored: independent (the TFs bind without affecting each other's affinities), competitive (the TFs compete for the same binding site), ordered (the TFs bind in a compulsory order), and joint binding (the TFs either bind as a preformed complex, or binding of one is virtually impossible in the absence of the other).
-The conclusions are: (1) the laws of combinatorial logic hold only for systems with independently binding TFs; (2) systems formed according to the other scenarios can mimic the functions of their Boolean logical counterparts, but cannot be combined or decomposed in the same way; and (3) the continuously scaled output of systems consisting of competitively binding activators and repressors can be controlled more robustly than that of single TF or (quasi-)logical multi-TF systems.},
-	file = {/home/rcannood/Zotero/storage/C7UWQGI3/Schilstra and Nehaniv - 2008 - Bio-Logic Gene Expression and the Laws of Combina.pdf;/home/rcannood/Zotero/storage/KXDLUMAG/artl.2008.14.1.html},
-	journal = {Artificial Life},
-	number = {1},
-}
-
-@Article{schwanhausser_globalquantificationmammalian_2011,
-	title = {Global Quantification of Mammalian Gene Expression Control},
-	author = {Bj{\"o}rn Schwanh{\"a}usser and Dorothea Busse and Na Li and Gunnar Dittmar and Johannes Schuchhardt and Jana Wolf and Wei Chen and Matthias Selbach},
-	year = {2011},
-	month = {may},
-	volume = {473},
-	pages = {337--342},
-	issn = {1476-4687},
-	doi = {10.1038/nature10098},
-	abstract = {Gene expression is a multistep process that involves the transcription, translation and turnover of messenger RNAs and proteins.
-Although it is one of the most fundamental processes of life, the entire cascade has never been quantified on a genome-wide scale.
-Here we simultaneously measured absolute mRNA and protein abundance and turnover by parallel metabolic pulse labelling for more than 5,000 genes in mammalian cells.
-Whereas mRNA and protein levels correlated better than previously thought, corresponding half-lives showed no correlation.
-Using a quantitative model we have obtained the first genome-scale prediction of synthesis rates of mRNAs and proteins.
-We find that the cellular abundance of proteins is predominantly controlled at the level of translation.
-Genes with similar combinations of mRNA and protein stability shared functional properties, indicating that half-lives evolved under energetic and dynamic constraints.
-Quantitative information about all stages of gene expression provides a rich resource and helps to provide a greater understanding of the underlying design principles.},
-	journal = {Nature},
-	number = {7347},
-}
-
-@Article{smith_spatialstochasticintracellular_2019,
-	title = {Spatial {{Stochastic Intracellular Kinetics}}: {{A Review}} of {{Modelling Approaches}}},
-	author = {Stephen Smith and Ramon Grima},
-	year = {2019},
-	month = {aug},
-	volume = {81},
-	pages = {2960--3009},
-	issn = {1522-9602},
-	doi = {10.1007/s11538-018-0443-1},
-	abstract = {Models of chemical kinetics that incorporate both stochasticity and diffusion are an increasingly common tool for studying biology.
-The variety of competing models is vast, but two stand out by virtue of their popularity: the reaction\textendash diffusion master equation and Brownian dynamics.
-In this review, we critically address a number of open questions surrounding these models: How can they be justified physically?
-How do they relate to each other?
-How do they fit into the wider landscape of chemical models, ranging from the rate equations to molecular dynamics?
-This review assumes no prior knowledge of modelling chemical kinetics and should be accessible to a wide range of readers.},
-	journal = {Bulletin of Mathematical Biology},
-	number = {8},
-}
-
-@Article{soneson_biasrobustnessscalability_2018,
-	title = {Bias, Robustness and Scalability in Single-Cell Differential Expression Analysis},
-	author = {Charlotte Soneson and Mark D. Robinson},
-	year = {2018},
-	month = {apr},
-	volume = {15},
-	pages = {255--261},
-	issn = {1548-7105},
-	doi = {10.1038/nmeth.4612},
-	abstract = {Many methods have been used to determine differential gene expression from single-cell RNA (scRNA)-seq data.
-We evaluated 36 approaches using experimental and synthetic data and found considerable differences in the number and characteristics of the genes that are called differentially expressed.
-Prefiltering of lowly expressed genes has important effects, particularly for some of the methods developed for bulk RNA-seq data analysis.
-However, we found that bulk RNA-seq analysis methods do not generally perform worse than those developed specifically for scRNA-seq.
-We also present conquer, a repository of consistently processed, analysis-ready public scRNA-seq data sets that is aimed at simplifying method evaluation and reanalysis of published results.
-Each data set provides abundance estimates for both genes and transcripts, as well as quality control and exploratory analysis reports.},
-	journal = {Nature Methods},
-	keywords = {Computational Biology,Gene Expression Profiling,Gene Expression Regulation,High-Throughput Nucleotide Sequencing,RNA,Sequence Analysis; RNA,Single-Cell Analysis,Software},
-	language = {eng},
-	number = {4},
-	pmid = {29481549},
-}
-
-@Article{street_slingshotcelllineage_2018,
-	title = {Slingshot: {{Cell Lineage}} and {{Pseudotime Inference}} for {{Single}}-{{Cell Transcriptomics}}},
-	shorttitle = {Slingshot},
-	author = {Kelly Street and Davide Risso and Russell B Fletcher and Diya Das and John Ngai and Nir Yosef and Elizabeth Purdom and Sandrine Dudoit},
-	year = {2018},
-	month = {jun},
-	volume = {19},
-	pages = {477},
-	issn = {1471-2164},
-	doi = {10.1186/s12864-018-4772-0},
-	abstract = {Single-cell transcriptomics allows researchers to investigate complex communities of heterogeneous cells.
-It can be applied to stem cells and their descendants in order to chart the progression from multipotent progenitors to fully differentiated cells.
-While a variety of statistical and computational methods have been proposed for inferring cell lineages, the problem of accurately characterizing multiple branching lineages remains difficult to solve.},
-	journal = {BMC Genomics},
-	number = {1},
-}
-
-@Article{vallejos_normalizingsinglecellrna_2017,
-	title = {Normalizing Single-Cell {{RNA}} Sequencing Data: Challenges and Opportunities},
-	shorttitle = {Normalizing Single-Cell {{RNA}} Sequencing Data},
-	author = {Catalina A. Vallejos and Davide Risso and Antonio Scialdone and Sandrine Dudoit and John C. Marioni},
-	year = {2017},
-	month = {jun},
-	volume = {14},
-	pages = {565--571},
-	publisher = {{Nature Publishing Group}},
-	issn = {1548-7105},
-	doi = {10.1038/nmeth.4292},
-	abstract = {This Perspective examines single-cell RNA-seq data challenges and the need for normalization methods designed specifically for single-cell data in order to remove technical biases.},
-	copyright = {2017 Nature Publishing Group, a division of Macmillan Publishers Limited. All Rights Reserved.},
-	journal = {Nature Methods},
-	language = {en},
-	number = {6},
-}
-
-@Article{vandenbulcke_syntrengeneratorsynthetic_2006,
-	title = {{{SynTReN}}: A Generator of Synthetic Gene Expression Data for Design and Analysis of Structure Learning Algorithms},
-	shorttitle = {{{SynTReN}}},
-	author = {Tim {Van den Bulcke} and Koenraad {Van Leemput} and Bart Naudts and Piet {van Remortel} and Hongwu Ma and Alain Verschoren and Bart {De Moor} and Kathleen Marchal},
-	year = {2006},
-	month = {jan},
-	volume = {7},
-	pages = {43},
-	issn = {1471-2105},
-	doi = {10.1186/1471-2105-7-43},
-	abstract = {The development of algorithms to infer the structure of gene regulatory networks based on expression data is an important subject in bioinformatics research.
-Validation of these algorithms requires benchmark data sets for which the underlying network is known.
-Since experimental data sets of the appropriate size and design are usually not available, there is a clear need to generate well-characterized synthetic data sets that allow thorough testing of learning algorithms in a fast and reproducible manner.},
-	journal = {BMC Bioinformatics},
-	language = {en},
-	number = {1},
-}
-
-@Article{vandesande_scalablescenicworkflow_2020,
-	title = {A Scalable {{SCENIC}} Workflow for Single-Cell Gene Regulatory Network Analysis},
-	author = {Bram {Van de Sande} and Christopher Flerin and Kristofer Davie and Maxime {De Waegeneer} and Gert Hulselmans and Sara Aibar and Ruth Seurinck and Wouter Saelens and Robrecht Cannoodt and Quentin Rouchon and Toni Verbeiren and Dries {De Maeyer} and Joke Reumers and Yvan Saeys and Stein Aerts},
-	year = {2020},
-	month = {jul},
-	volume = {15},
-	pages = {2247--2276},
-	issn = {1750-2799},
-	doi = {10.1038/s41596-020-0336-2},
-	abstract = {This protocol explains how to perform a fast SCENIC analysis alongside standard best practices steps on single-cell RNA-sequencing data using software containers and Nextflow pipelines.
-SCENIC reconstructs regulons (i.e., transcription factors and their target genes) assesses the activity of these discovered regulons in individual cells and uses these cellular activity patterns to find meaningful clusters of cells.
-Here we present an improved version of SCENIC with several advances.
-SCENIC has been refactored and reimplemented in Python (pySCENIC), resulting in a tenfold increase in speed, and has been packaged into containers for ease of use.
-It is now also possible to use epigenomic track databases, as well as motifs, to refine regulons.
-In this protocol, we explain the different steps of SCENIC: the workflow starts from the count matrix depicting the gene abundances for all cells and consists of three stages.
-First, coexpression modules are inferred using a regression per-target approach (GRNBoost2).
-Next, the indirect targets are pruned from these modules using cis-regulatory motif discovery (cisTarget).
-Lastly, the activity of these regulons is quantified via an enrichment score for the regulon's target genes (AUCell).
-Nonlinear projection methods can be used to display visual groupings of cells based on the cellular activity patterns of these regulons.
-The results can be exported as a loom file and visualized in the SCope web application.
-This protocol is illustrated on two use cases: a peripheral blood mononuclear cell data set and a panel of single-cell RNA-sequencing cancer experiments.
-For a data set of 10,000 genes and 50,000 cells, the pipeline runs in {$<$}2 h.},
-	journal = {Nature Protocols},
-	number = {7},
-}
-
-@Article{vieth_powsimrpoweranalysis_2017,
-	title = {{{powsimR}}: Power Analysis for Bulk and Single Cell {{RNA}}-Seq Experiments},
-	shorttitle = {{{powsimR}}},
-	author = {Beate Vieth and Christoph Ziegenhain and Swati Parekh and Wolfgang Enard and Ines Hellmann},
-	year = {2017},
-	month = {nov},
-	volume = {33},
-	pages = {3486--3488},
-	issn = {1367-4803},
-	doi = {10.1093/bioinformatics/btx435},
-	abstract = {AbstractSummary.  Power analysis is essential to optimize the design of RNA-seq experiments and to assess and compare the power to detect differentially express},
-	file = {/home/rcannood/Zotero/storage/MFRF86MZ/Vieth et al. - 2017 - powsimR power analysis for bulk and single cell R.pdf;/home/rcannood/Zotero/storage/HSCUJRI8/3952669.html},
-	journal = {Bioinformatics},
-	language = {en},
-	number = {21},
-}
-
-@Article{wang_quantifyingwaddingtonlandscape_2011,
-	title = {Quantifying the {{Waddington Landscape}} and {{Biological Paths}} for {{Development}} and {{Differentiation}}},
-	author = {Jin Wang and Kun Zhang and Li Xu and Erkang Wang},
-	year = {2011},
-	month = {may},
-	volume = {108},
-	pages = {8257--8262},
-	issn = {0027-8424, 1091-6490},
-	doi = {10.1073/pnas.1017017108},
-	abstract = {We developed a theoretical framework to prove the existence and quantify the Waddington landscape as well as chreode-biological paths for development and differentiation.
-The cells can have states with the higher probability ones giving the different cell types.
-Different cell types correspond to different basins of attractions of the probability landscape.
-We study how the cells develop from undifferentiated cells to differentiated cells from landscape perspectives.
-We quantified the Waddington landscape through construction of underlying probability landscape for cell development.
-We show the developmental process proceeds as moving from undifferentiated to the differentiated basins of attractions.
-The barrier height of the basins of attractions correlates with the escape time that determines the stability of cell types.
-We show that the developmental process can be quantitatively described and uncovered by the biological paths on the quantified Waddington landscape from undifferentiated to the differentiated cells.
-We found the dynamics of the developmental process is controlled by a combination of the gradient and curl force on the landscape.
-The biological paths often do not follow the steepest descent path on the landscape.
-The landscape framework also quantifies the possibility of reverse differentiation process such as cell reprogramming from differentiated cells back to the original stem cell.
-We show that the biological path of reverse differentiation is irreversible and different from the one for differentiation process.
-We found that the developmental process described by the underlying landscape and the associated biological paths is relatively stable and robust against the influences of environmental perturbations.},
-	journal = {Proceedings of the National Academy of Sciences},
-	number = {20},
-	pmid = {21536909},
-}
-
-@Article{weber_essentialguidelinescomputational_2019,
-	title = {Essential Guidelines for Computational Method Benchmarking},
-	author = {Lukas M. Weber and Wouter Saelens and Robrecht Cannoodt and Charlotte Soneson and Alexander Hapfelmeier and Paul P. Gardner and Anne-Laure Boulesteix and Yvan Saeys and Mark D. Robinson},
-	year = {2019},
-	month = {jun},
-	volume = {20},
-	pages = {125},
-	issn = {1474-760X},
-	doi = {10.1186/s13059-019-1738-8},
-	abstract = {In computational biology and other sciences, researchers are frequently faced with a choice between several computational methods for performing data analyses.
-Benchmarking studies aim to rigorously compare the performance of different methods using well-characterized benchmark datasets, to determine the strengths of each method or to provide recommendations regarding suitable choices of methods for an analysis.
-However, benchmarking studies must be carefully designed and implemented to provide accurate, unbiased, and informative results.
-Here, we summarize key practical guidelines and recommendations for performing high-quality benchmarking analyses, based on our experiences in computational biology.},
-	file = {/home/rcannood/Zotero/storage/83LMG2SG/Weber et al. - 2019 - Essential guidelines for computational method benc.pdf;/home/rcannood/Zotero/storage/GL7G9MIR/s13059-019-1738-8.html},
-	journal = {Genome Biology},
-	number = {1},
-}
-
-@Article{wickham_welcometidyverse_2019,
-	title = {Welcome to the {{Tidyverse}}},
-	author = {Hadley Wickham and Mara Averick and Jennifer Bryan and Winston Chang and Lucy McGowan and Romain Fran{\c c}ois and Garrett Grolemund and Alex Hayes and Lionel Henry and Jim Hester and Max Kuhn and Thomas Pedersen and Evan Miller and Stephan Bache and Kirill M{\"u}ller and Jeroen Ooms and David Robinson and Dana Seidel and Vitalie Spinu and Kohske Takahashi and Davis Vaughan and Claus Wilke and Kara Woo and Hiroaki Yutani},
-	year = {2019},
-	month = {nov},
-	doi = {10.21105/joss.01686},
-	abstract = {Wickham et al., (2019). Welcome to the Tidyverse. Journal of Open Source Software, 4(43), 1686, https://doi.org/10.21105/joss.01686},
-	file = {/home/rcannood/Zotero/storage/TVYSNC4J/Wickham et al. - 2019 - Welcome to the Tidyverse.pdf;/home/rcannood/Zotero/storage/EAQ2VYF8/joss.html},
-	language = {en},
-}
-
-@Article{xu_regulationbifurcatingcell_2015,
-	title = {Regulation of Bifurcating {{B}} Cell Trajectories by Mutual Antagonism between Transcription Factors {{IRF4}} and {{IRF8}}},
-	author = {Heping Xu and Virendra K Chaudhri and Zhiguo Wu and Konstantinos Biliouris and Krista Dienger-Stambaugh and Yrina Rochman and Harinder Singh},
-	year = {2015},
-	month = {dec},
-	volume = {16},
-	pages = {1274--1281},
-	abstract = {Upon recognition of antigen, B cells undertake a bifurcated response in which some cells rapidly differentiate into plasmablasts while others undergo affinity maturation in germinal centers (GCs).
-Here we identified a double-negative feedback loop between the transcription factors IRF4 and IRF8 that regulated the initial developmental bifurcation of activated B cells as well as the GC response.
-IRF8 dampened signaling via the B cell antigen receptor (BCR), facilitated antigen-specific interaction with helper T cells, and promoted antibody affinity maturation while antagonizing IRF4-driven differentiation of plasmablasts.
-Genomic analysis revealed concentration-dependent actions of IRF4 and IRF8 in regulating distinct gene-expression programs.
-Stochastic modeling suggested that the double-negative feedback was sufficient to initiate bifurcation of the B cell developmental trajectories.},
-	journal = {Nat. Immunol.},
-	number = {12},
-}
-
-@Article{yosef_dynamicregulatorynetwork_2013,
-	title = {Dynamic Regulatory Network Controlling \{\vphantom\}{{TH17}}\vphantom\{\} Cell Differentiation},
-	author = {Nir Yosef and Alex K Shalek and Jellert T Gaublomme and Hulin Jin and Youjin Lee and Amit Awasthi and Chuan Wu and Katarzyna Karwacz and Sheng Xiao and Marsela Jorgolli and David Gennert and Rahul Satija and Arvind Shakya and Diana Y Lu and John J Trombetta and Meenu R Pillai and Peter J Ratcliffe and Mathew L Coleman and Mark Bix and Dean Tantin and Hongkun Park and Vijay K Kuchroo and Aviv Regev},
-	year = {2013},
-	volume = {496},
-	pages = {461--468},
-	abstract = {Despite their importance, the molecular circuits that control the differentiation of naive T cells remain largely unknown.
-Recent studies that reconstructed regulatory networks in mammalian cells have focused on short-term responses and relied on perturbation-based approaches that cannot be readily applied to primary T cells.
-Here we combine transcriptional profiling at high temporal resolution, novel computational algorithms, and innovative nanowire-based perturbation tools to systematically derive and experimentally validate a model of the dynamic regulatory network that controls the differentiation of mouse TH17 cells, a proinflammatory T-cell subset that has been implicated in the pathogenesis of multiple autoimmune diseases.
-The TH17 transcriptional network consists of two self-reinforcing, but mutually antagonistic, modules, with 12 novel regulators, the coupled action of which may be essential for maintaining the balance between TH17 and other CD4(+) T-cell subsets.
-Our study identifies and validates 39 regulatory factors, embeds them within a comprehensive temporal network and reveals its organizational principles; it also highlights novel drug targets for controlling TH17 cell differentiation.},
-	journal = {Nature},
-	number = {7446},
-}
-
-@Article{zappia_splattersimulationsinglecell_2017,
-	title = {Splatter: {{Simulation}} of {{Single}}-{{Cell RNA Sequencing Data}}},
-	shorttitle = {Splatter},
-	author = {Luke Zappia and Belinda Phipson and Alicia Oshlack},
-	year = {2017},
-	month = {sep},
-	volume = {18},
-	pages = {174},
-	issn = {1474-760X},
-	doi = {10.1186/s13059-017-1305-0},
-	abstract = {As single-cell RNA sequencing (scRNA-seq) technologies have rapidly developed, so have analysis methods.
-Many methods have been tested, developed, and validated using simulated datasets.
-Unfortunately, current simulations are often poorly documented, their similarity to real data is not demonstrated, or reproducible code is not available.
-Here, we present the Splatter Bioconductor package for simple, reproducible, and well-documented simulation of scRNA-seq data.
-Splatter provides an interface to multiple simulation methods including Splat, our own simulation, based on a gamma-Poisson distribution.
-Splat can simulate single populations of cells, populations with multiple cell types, or differentiation paths.},
-	journal = {Genome Biology},
-	keywords = {RNA-seq,Simulation,Single-cell,Software},
-}
-
-@Article{zeisel_coupledpremrnamrna_2011,
-	title = {Coupled Pre-{{mRNA}} and {{mRNA}} Dynamics Unveil Operational Strategies Underlying Transcriptional Responses to Stimuli},
-	author = {Amit Zeisel and Wolfgang J K{\"o}stler and Natali Molotski and Jonathan M Tsai and Rita Krauthgamer and Jasmine Jacob-Hirsch and Gideon Rechavi and Yoav Soen and Steffen Jung and Yosef Yarden and Eytan Domany},
-	year = {2011},
-	month = {jan},
-	volume = {7},
-	pages = {529},
-	issn = {1744-4292},
-	doi = {10.1038/msb.2011.62},
-	abstract = {Transcriptional responses to extracellular stimuli involve tuning the rates of transcript production and degradation.
-Here, we show that the time-dependent profiles of these rates can be inferred from simultaneous measurements of precursor mRNA (pre-mRNA) and mature mRNA profiles.
-Transcriptome-wide measurements demonstrate that genes with similar mRNA profiles often exhibit marked differences in the amplitude and onset of their production rate.
-The latter is characterized by a large dynamic range, with a group of genes exhibiting an unexpectedly strong transient production overshoot, thereby accelerating their induction and, when combined with time-dependent degradation, shaping transient responses with precise timing and amplitude.},
-	file = {/home/rcannood/Zotero/storage/QRQW86Y9/Zeisel et al. - 2011 - Coupled pre-mRNA and mRNA dynamics unveil operatio.pdf;/home/rcannood/Zotero/storage/R8ZG26IL/msb.2011.html},
-	journal = {Molecular Systems Biology},
-	keywords = {half-life,operational strategy,pre-mRNA,transcriptional response},
-	number = {1},
-}
-
-@Article{zhang_simulatingmultiplefaceted_2019,
-	title = {Simulating Multiple Faceted Variability in Single Cell {{RNA}} Sequencing},
-	author = {Xiuwei Zhang and Chenling Xu and Nir Yosef},
-	year = {2019},
-	month = {jun},
-	volume = {10},
-	pages = {1--16},
-	issn = {2041-1723},
-	doi = {10.1038/s41467-019-10500-w},
-	abstract = {Simulated single cell RNA sequencing data is useful for method development and comparison.
-Here, the authors developed SymSim, a simulator that explicitly models the main factors of variation in single cell data.},
-	copyright = {2019 The Author(s)},
-	file = {/home/rcannood/Zotero/storage/YLQYCW6E/Zhang et al. - 2019 - Simulating multiple faceted variability in single .pdf;/home/rcannood/Zotero/storage/EXJ96Y59/s41467-019-10500-w.html},
-	journal = {Nature Communications},
-	language = {en},
-	number = {1},
-}
-\end{filecontents}
-
 \end{document}

### 2021-05-17 manuscript/supplementary_files.Rmd
update

@@
 data("realcounts", package = "dyngen")
-name_realcounts <- "zenodo_1443566_real_silver_bone-marrow-mesenchyme-erythrocyte-differentiation_mca"
+name_realcounts <- paste0(
+  "zenodo_1443566_real_silver_", 
+  "bone-marrow-mesenchyme-erythrocyte-differentiation_mca"
+)
 url_realcounts <- realcounts %>% filter(name == name_realcounts) %>% pull(url)
-realcount <- dyngen:::.download_cacheable_file(url_realcounts, getOption("dyngen_download_cache_dir"), verbose = FALSE)
+cache_dir <- getOption("dyngen_download_cache_dir")
+realcount <- 
+  dyngen:::.download_cacheable_file(url_realcounts, cache_dir, verbose = FALSE)
 ```
@@
 ) %>% 
-  mutate(filename = paste0(save_dir, "/cells", num_cells, "_feats", num_features, "_rep", rep, ".rds"))
+  mutate(filename = paste0(save_dir, "/cells", num_cells, "_feats", num_features, 
+    "_rep", rep, ".rds"))
 
@@
 
-    cat("Running num_cells: ", num_cells, ", num_features: ", num_features, ", rep: ", rep, "\n", sep = "")
+    cat("Running num_cells: ", num_cells, ", num_features: ", num_features, 
+      ", rep: ", rep, "\n", sep = "")
     num_tfs <- nrow(backbone$module_info)
@@
 ggplot(timings1) + 
-  geom_bar(aes(x = forcats::fct_inorder(as.character(num_cells)), y = time_elapsed, fill = forcats::fct_inorder(group)), stat = "identity") +
+  geom_bar(aes(
+    x = forcats::fct_inorder(as.character(num_cells)), 
+    y = time_elapsed,
+    fill = forcats::fct_inorder(group)),
+    stat = "identity") +
   theme_classic() +
@@
 ggplot(timings2) + 
-  geom_bar(aes(x = forcats::fct_inorder(as.character(num_features)), y = time_elapsed, fill = forcats::fct_inorder(group)), stat = "identity") +
+  geom_bar(aes(
+    x = forcats::fct_inorder(as.character(num_features)), 
+    y = time_elapsed, 
+    fill = forcats::fct_inorder(group)),
+    stat = "identity") +
   theme_classic() +

### 2021-05-17 manuscript/render_manuscript.R

@@ -14,22 +14,22 @@ if (Sys.info()[["user"]] == "rcannood") {
     write_lines("manuscript/manuscript.Rmd")
 }
 
-# take the relevant citations and write it to a second file
-bib_keys <-
-  readr::read_lines("manuscript/manuscript.Rmd") %>%
-  paste(collapse = "\n") %>%
-  str_extract_all("\\[@[^\\]]*\\]") %>%
-  first() %>%
-  gsub("^\\[@", "", .) %>%
-  gsub("\\]$", "", .) %>%
-  str_split("; *") %>%
-  unlist() %>%
-  gsub("@", "", .) %>%
-  unique
-
-knitcitations::cite_options(check.entries = FALSE)
-bib <- knitcitations::read.bibtex("manuscript/library.bib")
-knitcitations::write.bibtex(bib[bib_keys], file = "manuscript/knitcitations.bib")
+# # take the relevant citations and write it to a second file
+# bib_keys <-
+#   readr::read_lines("manuscript/manuscript.Rmd") %>%
+#   paste(collapse = "\n") %>%
+#   str_extract_all("\\[@[^\\]]*\\]") %>%
+#   first() %>%
+#   gsub("^\\[@", "", .) %>%
+#   gsub("\\]$", "", .) %>%
+#   str_split("; *") %>%
+#   unlist() %>%
+#   gsub("@", "", .) %>%
+#   unique
+#
+# knitcitations::cite_options(check.entries = FALSE)
+# bib <- knitcitations::read.bibtex("manuscript/library.bib")
+# knitcitations::write.bibtex(bib[bib_keys], file = "manuscript/knitcitations.bib")
 
 
 # render the manuscript

### 2021-05-20 manuscript/manuscript.Rmd
update

@@
 ---
-title: "dyngen: a multi-modal simulator for spearheading new single-cell omics analyses"
+title: "dyngen is a multi-modal simulator for spearheading future single-cell omics analyses"
 author:
@@
   pdf_document:
+    number_sections: true
     keep_tex: yes
@@
 <!-- format: max. 70 words -->
-We present dyngen, a novel, multi-modal simulation engine for studying dynamic cellular processes at single-cell resolution.
-dyngen is more flexible than current single-cell simulation engines, and allows better method development and benchmarking, thereby stimulating development and testing of novel computational methods.
-We demonstrate its potential for spearheading novel computational methods on three novel applications: aligning cell developmental trajectories, cell-specific regulatory network inference and estimation of RNA velocity.
+We present dyngen, a multi-modal simulation engine for studying dynamic cellular processes at single-cell resolution.
+dyngen is more flexible than current single-cell simulation engines, and allows better method development and benchmarking, thereby stimulating development and testing of computational methods.
+We demonstrate its potential for spearheading computational methods on three applications: aligning cell developmental trajectories, cell-specific regulatory network inference and estimation of RNA velocity.
 
@@
 
-Another, more experimental use of simulation engines is their important role in spearheading the development of novel computational methods, possibly even before real data is available.
+Another, more experimental use of simulation engines is their important role in spearheading the development of computational methods, possibly even before real data is available.
 In this way, simulation engines can be used to assess the value of novel experimental protocols or treatments.
@@
 The low library sizes, in particular, are problematic as ODEs are ill-suited for performing low-molecule simulations [@gillespie_exactstochasticsimulation_1977].
-This necessitates the development of new single-cell simulators.
+This necessitates the development of single-cell simulators.
 
@@
 However, the framework is flexible enough to be adaptable to a broad range of applications, including methods that integrate clustering, network inference, and trajectory inference.
-In this respect, dyngen may promote the development of new tools in the single-cell field similarly as other simulators have done in the past [@schaffter_genenetweaversilicobenchmark_2011; @ewing_combiningtumorgenome_2015].
+In this respect, dyngen may promote the development of tools in the single-cell field similarly as other simulators have done in the past [@schaffter_genenetweaversilicobenchmark_2011; @ewing_combiningtumorgenome_2015].
 Additionally, one could anticipate technological developments in single-cell multi-omics.
@@
 A module may have basal expression, which means genes in this module will be transcribed without the presence of transcription factor molecules.
-A module marked as "active during the burn phase" means that this module will be allowed to generate expression of its genes during an initial warm-up phase (See section \ref{sec:dyngen-simcell}).
+A module marked as "active during the burn phase" means that this module will be allowed to generate expression of its genes during an initial warm-up phase.
 At the end of the dyngen process, cells will not be sampled from the burn phase simulations.
@@
 For this, the cells are sampled from the simulations and molecules are sampled for each of the cells.
-Gene capture rates and library sizes are empirically derived from real datasets as to match real technical variation.
+Gene capture rates and library sizes are empirically derived from real datasets to match real technical variation.
 
@@
 Finally, for each cell $i$, draw $ls_i$ molecules from the multinomial distribution with probabilities $cr_j \times ab_{i,j}$ with $ab_{i,j}$ the molecule abundance level of molecule $j$ in cell $i$.
+
+### Comparison between a dyngen and a reference dataset
+Comparison between a dyngen dataset and the reference dataset it used in terms of characteristic single-cell omics features showed that dyngen produces datasets with highly similar data characteristics (Supplementary Note 1).
+Supplementary Note 1 was generated using countsimQC [@soneson_unifiedqualityverification_2018].
+.
 
@@
 ## Determining the cell-specific ground-truth regulatory network {#sec:dyngen-extractgrn}
-Calculating the regulatory effect of a regulator $R$ on a target $T$ (Supplementary Fig.\ 4F) requires determining the contribution of $R$ in the propensity function of the transcription of $T$ (section\ \ref{sec:dyngen-reactions}) with respect to other regulators.
+Calculating the regulatory effect of a regulator $R$ on a target $T$ (Supplementary Fig.\ 4F) requires determining the contribution of $R$ in the propensity function of the transcription of $T$ with respect to other regulators.
 This information is useful, amongst others, for benchmarking cell-specific network inference methods.
@@
 42 datasets were generated using the 14 different predefined backbones and three different seeds.
-For every cell in the dataset, the transcriptomics profile and the corresponding cell-specific ground-truth regulatory network was determined (Section\ \ref{sec:dyngen-extractgrn}).
+For every cell in the dataset, the transcriptomics profile and the corresponding cell-specific ground-truth regulatory network was determined.
 
@@
 
+## Comparison of scalability and runtime
+Simulating a bifurcating cycle dataset with 10'000 genes and 10'000 required in total 1147 seconds (Supplementary Note 2).
+Fixing the number of genes and varying the number of cells showed that the execution time of dyngen scales linearly w.r.t.
+the number of cells (Supplementary Note 2).
+Fixing the number of cells and varying the number of genes also showed that the execution time of dyngen scales linearly w.r.t.
+the number of genes (Supplementary Note 2).
+These timings were measured using 30 (out of 32) threads using a AMD Ryzen 9 5950X clocked at 3.4GHz.
+
 # Data Availability
@@
 
+# Acknowledgements
+This project has been made possible in part by grant number 2020-218899 from the Chan Zuckerberg Initiative DAF, an advised fund of Silicon Valley Community Foundation.
+This research received funding from the Flemish Government under the “Onderzoeksprogramma Artificiële Intelligentie (AI) Vlaanderen” program.
+
 # Author contributions
@@
 
-# Acknowledgements
-This project has been made possible in part by grant number 2020-218899 from the Chan Zuckerberg
-Initiative DAF, an advised fund of Silicon Valley Community Foundation.
+# Competing interests
+The authors declare no competing interests.

### 2021-05-20 manuscript/manuscript.tex
update

@@
 \hypersetup{
-  pdftitle={dyngen: a multi-modal simulator for spearheading new single-cell omics analyses},
+  pdftitle={dyngen is a multi-modal simulator for spearheading future single-cell omics analyses},
   pdfauthor={Robrecht Cannoodt,1,2,3; Wouter Saelens,1,2,4; Louise Deconinck1,2; Yvan Saeys1,2,},
@@
   \setlength{\itemsep}{0pt}\setlength{\parskip}{0pt}}
-\setcounter{secnumdepth}{-\maxdimen} % remove section numbering
+\setcounter{secnumdepth}{5}
 \usepackage{tcolorbox}
@@
 
-\title{dyngen: a multi-modal simulator for spearheading new single-cell
-omics analyses}
+\title{dyngen is a multi-modal simulator for spearheading future
+single-cell omics analyses}
 \author{Robrecht Cannoodt\textsuperscript{$\dagger{}$,1,2,3} \and Wouter
@@
 Deconinck\textsuperscript{1,2} \and Yvan Saeys\textsuperscript{1,2,*}}
-\date{17 May 2021}
+\date{20 May 2021}
 
@@
 
-We present dyngen, a novel, multi-modal simulation engine for studying
-dynamic cellular processes at single-cell resolution. dyngen is more
-flexible than current single-cell simulation engines, and allows better
-method development and benchmarking, thereby stimulating development and
-testing of novel computational methods. We demonstrate its potential for
-spearheading novel computational methods on three novel applications:
-aligning cell developmental trajectories, cell-specific regulatory
-network inference and estimation of RNA velocity.
+We present dyngen, a multi-modal simulation engine for studying dynamic
+cellular processes at single-cell resolution. dyngen is more flexible
+than current single-cell simulation engines, and allows better method
+development and benchmarking, thereby stimulating development and
+testing of computational methods. We demonstrate its potential for
+spearheading computational methods on three applications: aligning cell
+developmental trajectories, cell-specific regulatory network inference
+and estimation of RNA velocity.
 
@@
 Another, more experimental use of simulation engines is their important
-role in spearheading the development of novel computational methods,
-possibly even before real data is available. In this way, simulation
-engines can be used to assess the value of novel experimental protocols
-or treatments. Simulation engines are also increasingly important when
-it comes to finding alternatives to animal models, for example for drug
+role in spearheading the development of computational methods, possibly
+even before real data is available. In this way, simulation engines can
+be used to assess the value of novel experimental protocols or
+treatments. Simulation engines are also increasingly important when it
+comes to finding alternatives to animal models, for example for drug
 testing and precision medicine. In such scenarios, cellular simulations
@@
 \autocite{gillespie_exactstochasticsimulation_1977}. This necessitates
-the development of new single-cell simulators.
+the development of single-cell simulators.
 
@@
 inference, and trajectory inference. In this respect, dyngen may promote
-the development of new tools in the single-cell field similarly as other
+the development of tools in the single-cell field similarly as other
 simulators have done in the past
@@
 will be allowed to generate expression of its genes during an initial
-warm-up phase (See section \ref{sec:dyngen-simcell}). At the end of the
-dyngen process, cells will not be sampled from the burn phase
-simulations. Interactions between modules have a strength (which is a
-positive integer) and an effect (+1 for upregulating, -1 for
-downregulating).
+warm-up phase. At the end of the dyngen process, cells will not be
+sampled from the burn phase simulations. Interactions between modules
+have a strength (which is a positive integer) and an effect (+1 for
+upregulating, -1 for downregulating).
 
@@
 capture rates and library sizes are empirically derived from real
-datasets as to match real technical variation.
+datasets to match real technical variation.
 
@@
 
+\hypertarget{comparison-between-a-dyngen-and-a-reference-dataset}{%
+\subsubsection{Comparison between a dyngen and a reference
+dataset}\label{comparison-between-a-dyngen-and-a-reference-dataset}}
+
+Comparison between a dyngen dataset and the reference dataset it used in
+terms of characteristic single-cell omics features showed that dyngen
+produces datasets with highly similar data characteristics
+(Supplementary Note 1). Supplementary Note 1 was generated using
+countsimQC \autocite{soneson_unifiedqualityverification_2018}. .
+
 \hypertarget{sec:dyngen-batcheffect}{%
@@
 (Supplementary Fig.~4F) requires determining the contribution of \(R\)
-in the propensity function of the transcription of \(T\)
-(section~\ref{sec:dyngen-reactions}) with respect to other regulators.
-This information is useful, amongst others, for benchmarking
-cell-specific network inference methods.
+in the propensity function of the transcription of \(T\) with respect to
+other regulators. This information is useful, amongst others, for
+benchmarking cell-specific network inference methods.
 
@@
 transcriptomics profile and the corresponding cell-specific ground-truth
-regulatory network was determined (Section~\ref{sec:dyngen-extractgrn}).
+regulatory network was determined.
 
@@
 \autocite{holm_simplesequentiallyrejective_1979}.
+
+\hypertarget{comparison-of-scalability-and-runtime}{%
+\subsection{Comparison of scalability and
+runtime}\label{comparison-of-scalability-and-runtime}}
+
+Simulating a bifurcating cycle dataset with 10'000 genes and 10'000
+required in total 1147 seconds (Supplementary Note 2). Fixing the number
+of genes and varying the number of cells showed that the execution time
+of dyngen scales linearly w.r.t. the number of cells (Supplementary Note
+2). Fixing the number of cells and varying the number of genes also
+showed that the execution time of dyngen scales linearly w.r.t. the
+number of genes (Supplementary Note 2). These timings were measured
+using 30 (out of 32) threads using a AMD Ryzen 9 5950X clocked at
+3.4GHz.
 
@@
 
+\hypertarget{acknowledgements}{%
+\section{Acknowledgements}\label{acknowledgements}}
+
+This project has been made possible in part by grant number 2020-218899
+from the Chan Zuckerberg Initiative DAF, an advised fund of Silicon
+Valley Community Foundation. This research received funding from the
+Flemish Government under the ``Onderzoeksprogramma Artificiële
+Intelligentie (AI) Vlaanderen'' program.
+
 \hypertarget{author-contributions}{%
@@
 
-\hypertarget{acknowledgements}{%
-\section{Acknowledgements}\label{acknowledgements}}
-
-This project has been made possible in part by grant number 2020-218899
-from the Chan Zuckerberg Initiative DAF, an advised fund of Silicon
-Valley Community Foundation.
+\hypertarget{competing-interests}{%
+\section{Competing interests}\label{competing-interests}}
+
+The authors declare no competing interests.
 

### 2021-05-20 manuscript/v5/manuscript.tex
update

@@
 \hypersetup{
-  pdftitle={dyngen: a multi-modal simulator for spearheading new single-cell omics analyses},
+  pdftitle={dyngen is a multi-modal simulator for spearheading future single-cell omics analyses},
   pdfauthor={Robrecht Cannoodt; Wouter Saelens; Louise Deconinck; Yvan Saeys},
@@
 \usepackage{graphicx}
-
-\title{dyngen: a multi-modal simulator for spearheading new single-cell omics analyses}
+\usepackage[numbers]{natbib}
+
+\title{dyngen is a multi-modal simulator for spearheading future single-cell omics analyses}
 \author{Robrecht Cannoodt\textsuperscript{$\dagger{}$,1,2,3} \and Wouter Saelens\textsuperscript{$\dagger{}$,1,2,4} \and Louise Deconinck\textsuperscript{1,2} \and Yvan Saeys\textsuperscript{1,2,*}}
-\date{12 May 2021}
+\date{20 May 2021}
 
@@
 
-\section{Abstract}\label{abstract}
-
-We present dyngen, a novel, multi-modal simulation engine for studying
-dynamic cellular processes at single-cell resolution. dyngen is more
-flexible than current single-cell simulation engines, and allows better
-method development and benchmarking, thereby stimulating development and
-testing of novel computational methods. We demonstrate its potential for
-spearheading novel computational methods on three novel applications:
-aligning cell developmental trajectories, cell-specific regulatory
-network inference and estimation of RNA velocity.
-
-\section{Introduction}\label{introduction}
+\hypertarget{abstract}{%
+	\section{Abstract}\label{abstract}}
+
+We present dyngen, a multi-modal simulation engine for studying dynamic
+cellular processes at single-cell resolution. dyngen is more flexible
+than current single-cell simulation engines, and allows better method
+development and benchmarking, thereby stimulating development and
+testing of computational methods. We demonstrate its potential for
+spearheading computational methods on three applications: aligning cell
+developmental trajectories, cell-specific regulatory network inference
+and estimation of RNA velocity.
+
+\hypertarget{introduction}{%
+	\section{Introduction}\label{introduction}}
 
@@
 \cite{gillespie_exactstochasticsimulation_1977}. This necessitates
-the development of new single-cell simulators.
+the development of single-cell simulators.
 
@@
 
-\begin{figure}[htbp]
+\begin{figure}[t!]
 	\centering
@@
 
-\section{Results}\label{results}
+\hypertarget{results}{%
+	\section{Results}\label{results}}
 
@@
 
-\section{Discussion}\label{discussion}
+\hypertarget{discussion}{%
+	\section{Discussion}\label{discussion}}
 
@@
 
-\section{Methods}\label{sec:dyngen-methods}
+\hypertarget{sec:dyngen-methods}{%
+	\section{Methods}\label{sec:dyngen-methods}}
 
@@
 
-\subsection{Defining the module network}\label{sec:dyngen-modules}
+\hypertarget{sec:dyngen-modules}{%
+	\subsection{Defining the module network}\label{sec:dyngen-modules}}
 
@@
 will be allowed to generate expression of its genes during an initial
-warm-up phase. At the end of the
-dyngen process, cells will not be sampled from the burn phase
-simulations. Interactions between modules have a strength (which is a
-positive integer) and an effect (+1 for upregulating, -1 for
-downregulating).
+warm-up phase. At the end of the dyngen process, cells will not be
+sampled from the burn phase simulations. Interactions between modules
+have a strength (which is a positive integer) and an effect (+1 for
+upregulating, -1 for downregulating).
 
@@
 
-\subsection{Generating the gene regulatory network}\label{sec:dyngen-grn}
+\hypertarget{sec:dyngen-grn}{%
+	\subsection{Generating the gene regulatory
+		network}\label{sec:dyngen-grn}}
 
@@
 
-\subsection{Convert gene regulatory network to a set of reactions}\label{sec:dyngen-reactions}
+\hypertarget{sec:dyngen-reactions}{%
+	\subsection{Convert gene regulatory network to a set of
+		reactions}\label{sec:dyngen-reactions}}
 
@@
 
-\subsubsection{\texorpdfstring{Propensity of transcription when \(N=1\)}{Propensity of transcription when N=1}}\label{propensity-of-transcription-when-n1}
+\hypertarget{propensity-of-transcription-when-n1}{%
+	\subsubsection{\texorpdfstring{Propensity of transcription when
+			\(N=1\)}{Propensity of transcription when N=1}}\label{propensity-of-transcription-when-n1}}
 
@@
 
-\subsubsection{\texorpdfstring{Propensity of transcription when \(N=2\)}{Propensity of transcription when N=2}}\label{propensity-of-transcription-when-n2}
+\hypertarget{propensity-of-transcription-when-n2}{%
+	\subsubsection{\texorpdfstring{Propensity of transcription when
+			\(N=2\)}{Propensity of transcription when N=2}}\label{propensity-of-transcription-when-n2}}
 
@@
 
-\subsubsection{\texorpdfstring{Propensity of transcription for an arbitrary \(N\)}{Propensity of transcription for an arbitrary N}}\label{propensity-of-transcription-for-an-arbitrary-n}
+\hypertarget{propensity-of-transcription-for-an-arbitrary-n}{%
+	\subsubsection{\texorpdfstring{Propensity of transcription for an
+			arbitrary
+			\(N\)}{Propensity of transcription for an arbitrary N}}\label{propensity-of-transcription-for-an-arbitrary-n}}
 
@@
 
-\subsubsection{\texorpdfstring{Propensity of transcription for a large \(N\)}{Propensity of transcription for a large N}}\label{propensity-of-transcription-for-a-large-n}
+\hypertarget{propensity-of-transcription-for-a-large-n}{%
+	\subsubsection{\texorpdfstring{Propensity of transcription for a large
+			\(N\)}{Propensity of transcription for a large N}}\label{propensity-of-transcription-for-a-large-n}}
 
@@
 
-\subsubsection{Independence, synergism and basal expression}\label{independence-synergism-and-basal-expression}
+\hypertarget{independence-synergism-and-basal-expression}{%
+	\subsubsection{Independence, synergism and basal
+		expression}\label{independence-synergism-and-basal-expression}}
 
@@
 
-\subsection{Simulate single cells}\label{dyngen-simcell}
+\hypertarget{sec:dyngen-simcell}{%
+	\subsection{Simulate single cells}\label{sec:dyngen-simcell}}
 
@@
 
-\subsection{Simulate experiment}\label{sec:dyngen-experiment}
+\hypertarget{sec:dyngen-experiment}{%
+	\subsection{Simulate experiment}\label{sec:dyngen-experiment}}
 
@@
 capture rates and library sizes are empirically derived from real
-datasets as to match real technical variation.
-
-\subsubsection{Sample cells}\label{sample-cells}
+datasets to match real technical variation.
+
+\hypertarget{sample-cells}{%
+	\subsubsection{Sample cells}\label{sample-cells}}
 
@@
 
-\subsubsection{Sample molecules}\label{sample-molecules}
+\hypertarget{sample-molecules}{%
+	\subsubsection{Sample molecules}\label{sample-molecules}}
 
@@
 
-\subsection{Simulating batch effects}\label{sec:dyngen-batcheffect}
+\hypertarget{comparison-between-a-dyngen-and-a-reference-dataset}{%
+	\subsubsection{Comparison between a dyngen and a reference
+		dataset}\label{comparison-between-a-dyngen-and-a-reference-dataset}}
+
+Comparison between a dyngen dataset and the reference dataset it used in
+terms of characteristic single-cell omics features showed that dyngen
+produces datasets with highly similar data characteristics
+(Supplementary Note 1). Supplementary Note 1 was generated using
+countsimQC \cite{soneson_unifiedqualityverification_2018}. .
+
+\hypertarget{sec:dyngen-batcheffect}{%
+	\subsection{Simulating batch effects}\label{sec:dyngen-batcheffect}}
 
@@
 
-\subsection{Determining the ground-truth trajectory}\label{sec:dyngen-groundtruth}
+\hypertarget{sec:dyngen-groundtruth}{%
+	\subsection{Determining the ground-truth
+		trajectory}\label{sec:dyngen-groundtruth}}
 
@@
 
-\subsection{Determining the cell-specific ground-truth regulatory network}\label{sec:dyngen-extractgrn}
+\hypertarget{sec:dyngen-extractgrn}{%
+	\subsection{Determining the cell-specific ground-truth regulatory
+		network}\label{sec:dyngen-extractgrn}}
 
@@
 
-\subsection{Comparison of cell-specific network inference methods}\label{sec:dyngen-nicompare}
+\hypertarget{sec:dyngen-nicompare}{%
+	\subsection{Comparison of cell-specific network inference
+		methods}\label{sec:dyngen-nicompare}}
 
@@
 
-\subsection{Comparison of RNA velocity methods}\label{sec:dyngen-velcompare}
+\hypertarget{sec:dyngen-velcompare}{%
+	\subsection{Comparison of RNA velocity
+		methods}\label{sec:dyngen-velcompare}}
 
@@
 
- \subsection{Comparison of trajectory alignment}\label{sec:dyngen-tacompare}
+\hypertarget{sec:dyngen-tacompare}{%
+	\subsection{Comparison of trajectory
+		alignment}\label{sec:dyngen-tacompare}}
 
@@
 
-\section{Data Availability}\label{data-availability}
+\hypertarget{comparison-of-scalability-and-runtime}{%
+	\subsection{Comparison of scalability and
+		runtime}\label{comparison-of-scalability-and-runtime}}
+
+Simulating a bifurcating cycle dataset with 10'000 genes and 10'000
+required in total 1147 seconds (Supplementary Note 2). Fixing the number
+of genes and varying the number of cells showed that the execution time
+of dyngen scales linearly w.r.t. the number of cells (Supplementary Note
+2). Fixing the number of cells and varying the number of genes also
+showed that the execution time of dyngen scales linearly w.r.t. the
+number of genes (Supplementary Note 2). These timings were measured
+using 30 (out of 32) threads using a AMD Ryzen 9 5950X clocked at
+3.4GHz.
+
+\hypertarget{data-availability}{%
+	\section{Data Availability}\label{data-availability}}
 
@@
 
-\section{Code Availability}\label{code-availability}
+\hypertarget{code-availability}{%
+	\section{Code Availability}\label{code-availability}}
 
@@
 
-\section{Author contributions}\label{author-contributions}
+\hypertarget{acknowledgements}{%
+	\section{Acknowledgements}\label{acknowledgements}}
+
+This project has been made possible in part by grant number 2020-218899
+from the Chan Zuckerberg Initiative DAF, an advised fund of Silicon
+Valley Community Foundation. This research received funding from the
+Flemish Government under the ``Onderzoeksprogramma Artificiële
+Intelligentie (AI) Vlaanderen'' program.
+
+\hypertarget{author-contributions}{%
+	\section{Author contributions}\label{author-contributions}}
 
 \begin{itemize}
-\item W.S. and R.C. designed the study.
-\item R.C., W.S., and L.D. performed the experiments and analysed the data.
-\item R.C. and W.S. implemented the dyngen software package.
-\item R.C., W.S., L.D., and Y.S. wrote the manuscript.
-\item Y.S. supervised the project.
+	\item
+	W.S. and R.C. designed the study.
+	\item
+	R.C., W.S., and L.D. performed the experiments and analysed the data.
+	\item
+	R.C. and W.S. implemented the dyngen software package.
+	\item
+	R.C., W.S., L.D., and Y.S. wrote the manuscript.
+	\item
+	Y.S. supervised the project.
 \end{itemize}
 
-\begin{thebibliography}{10}
-
-	\bibitem{aibar_scenicsinglecellregulatory_2017}
-	Sara Aibar, Carmen~Bravo Gonz{\a'a}lez-Blas, Thomas Moerman, V{\^a}n~Anh
-	Huynh-Thu, Hana Imrichova, Gert Hulselmans, Florian Rambow, Jean-Christophe
-	Marine, Pierre Geurts, Jan Aerts, Joost {van den Oord}, Zeynep~Kalender Atak,
-	Jasper Wouters, and Stein Aerts.
-	\newblock {{SCENIC}}: Single-cell regulatory network inference and clustering.
-	\newblock {\em Nature Methods}, 10 2017.
-
-	\bibitem{alpert_alignmentsinglecelltrajectories_2018}
-	Ayelet Alpert, Lindsay~S. Moore, Tania Dubovik, and Shai~S. Shen-Orr.
-	\newblock Alignment of single-cell trajectories to compare cellular expression
-	dynamics.
-	\newblock {\em Nature Methods}, 15(4):267--270, 4 2018.
-
-	\bibitem{bergen_generalizingrnavelocity_2020}
-	Volker Bergen, Marius Lange, Stefan Peidli, F.~Alexander Wolf, and Fabian~J.
-	Theis.
-	\newblock Generalizing {{RNA}} velocity to transient cell states through
-	dynamical modeling.
-	\newblock {\em Nature Biotechnology}, 38(12):1408--1414, 12 2020.
-
-	\bibitem{bjornsson_digitaltwinspersonalize_2019}
+\hypertarget{competing-interests}{%
+	\section{Competing interests}\label{competing-interests}}
+
+The authors declare no competing interests.
+
+%\bibliographystyle{unsrtnat}
+%\bibliography{../library}
+
+\begin{thebibliography}{59}
+	\providecommand{\natexlab}[1]{#1}
+	\providecommand{\url}[1]{\texttt{#1}}
+	\expandafter\ifx\csname urlstyle\endcsname\relax
+	\providecommand{\doi}[1]{doi: #1}\else
+	\providecommand{\doi}{doi: \begingroup \urlstyle{rm}\Url}\fi
+
+	\bibitem[Zappia et~al.(2017)Zappia, Phipson, and
+	Oshlack]{zappia_splattersimulationsinglecell_2017}
+	Luke Zappia, Belinda Phipson, and Alicia Oshlack.
+	\newblock Splatter: {{Simulation}} of {{Single}}-{{Cell RNA Sequencing Data}}.
+	\newblock \emph{Genome Biology}, 18:\penalty0 174, September 2017.
+	\newblock ISSN 1474-760X.
+	\newblock \doi{10.1186/s13059-017-1305-0}.
+
+	\bibitem[Bj{\"o}rnsson et~al.(2019)Bj{\"o}rnsson, Borrebaeck, Elander,
+	Gasslander, Gawel, Gustafsson, J{\"o}rnsten, Lee, Li, Lilja,
+	{Mart{\'i}nez-Enguita}, Matussek, Sandstr{\"o}m, Sch{\"a}fer, Stenmarker,
+	Sun, Sysoev, Zhang, Benson, and {on behalf of the Swedish Digital Twin
+		Consortium}]{bjornsson_digitaltwinspersonalize_2019}
 	Bergthor Bj{\"o}rnsson, Carl Borrebaeck, Nils Elander, Thomas Gasslander,
 	Danuta~R. Gawel, Mika Gustafsson, Rebecka J{\"o}rnsten, Eun~Jung Lee, Xinxiu
-	Li, Sandra Lilja, David Mart\'inez-Enguita, Andreas Matussek, Per
+	Li, Sandra Lilja, David {Mart{\'i}nez-Enguita}, Andreas Matussek, Per
 	Sandstr{\"o}m, Samuel Sch{\"a}fer, Margaretha Stenmarker, X.~F. Sun, Oleg
@@
 	\newblock Digital twins to personalize medicine.
-	\newblock {\em Genome Medicine}, 12(1):4, 12 2019.
-
-	\bibitem{cacchiarelli_aligningsinglecelldevelopmental_2018}
+	\newblock \emph{Genome Medicine}, 12\penalty0 (1):\penalty0 4, December 2019.
+	\newblock ISSN 1756-994X.
+	\newblock \doi{10.1186/s13073-019-0701-3}.
+
+	\bibitem[Roy et~al.(2008)Roy, {Werner-Washburne}, and
+	Lane]{roy_systemgeneratingtranscription_2008}
+	Sushmita Roy, Margaret {Werner-Washburne}, and Terran Lane.
+	\newblock A system for generating transcription regulatory networks with
+	combinatorial control of transcription.
+	\newblock \emph{Bioinformatics}, 24\penalty0 (10):\penalty0 1318--1320, May
+	2008.
+	\newblock ISSN 1367-4803.
+	\newblock \doi{10.1093/bioinformatics/btn126}.
+
+	\bibitem[Hache et~al.(2009)Hache, Wierling, Lehrach, and
+	Herwig]{hache_gengesystematicgeneration_2009}
+	Hendrik Hache, Christoph Wierling, Hans Lehrach, and Ralf Herwig.
+	\newblock {{GeNGe}}: Systematic generation of gene regulatory networks.
+	\newblock \emph{Bioinformatics}, 25\penalty0 (9):\penalty0 1205--1207, May
+	2009.
+	\newblock ISSN 1367-4803.
+	\newblock \doi{10.1093/bioinformatics/btp115}.
+
+	\bibitem[Schaffter et~al.(2011)Schaffter, Marbach, and
+	Floreano]{schaffter_genenetweaversilicobenchmark_2011}
+	Thomas Schaffter, Daniel Marbach, and Dario Floreano.
+	\newblock {{GeneNetWeaver}}: In silico benchmark generation and performance
+	profiling of network inference methods.
+	\newblock \emph{Bioinformatics}, 27\penalty0 (16):\penalty0 2263--2270, August
+	2011.
+	\newblock ISSN 1367-4811.
+	\newblock \doi{10.1093/bioinformatics/btr373}.
+
+	\bibitem[{Van den Bulcke} et~al.(2006){Van den Bulcke}, Van~Leemput, Naudts,
+	{van Remortel}, Ma, Verschoren, De~Moor, and
+	Marchal]{vandenbulcke_syntrengeneratorsynthetic_2006}
+	Tim {Van den Bulcke}, Koenraad Van~Leemput, Bart Naudts, Piet {van Remortel},
+	Hongwu Ma, Alain Verschoren, Bart De~Moor, and Kathleen Marchal.
+	\newblock {{SynTReN}}: A generator of synthetic gene expression data for design
+	and analysis of structure learning algorithms.
+	\newblock \emph{BMC Bioinformatics}, 7\penalty0 (1):\penalty0 43, January 2006.
+	\newblock ISSN 1471-2105.
+	\newblock \doi{10.1186/1471-2105-7-43}.
+
+	\bibitem[Prill et~al.(2010)Prill, Daniel, Julio, K., G., Xiaowei, D., Gregoire,
+	Gustavo, Marbach, {Saez-Rodriguez}, Sorger, Alexopoulos, Xue, Clarke,
+	{Altan-Bonnet}, Stolovitzky, Daniel, Julio, K., G., Xiaowei, D., Gregoire,
+	and Gustavo]{prill_rigorousassessmentsystems_2010}
+	Robert~J Prill, Marbach Daniel, Saez-Rodriguez Julio, Sorger~Peter K.,
+	Alexopoulos~Leonidas G., Xue Xiaowei, Clarke~Neil D., Altan-Bonnet Gregoire,
+	Stolovitzky Gustavo, Daniel Marbach, Julio {Saez-Rodriguez}, Peter~K Sorger,
+	Leonidas~G Alexopoulos, Xiaowei Xue, Neil~D Clarke, Gregoire {Altan-Bonnet},
+	Gustavo Stolovitzky, Marbach Daniel, Saez-Rodriguez Julio, Sorger~Peter K.,
+	Alexopoulos~Leonidas G., Xue Xiaowei, Clarke~Neil D., Altan-Bonnet Gregoire,
+	and Stolovitzky Gustavo.
+	\newblock Towards a {{Rigorous Assessment}} of {{Systems Biology Models}}:
+	{{The DREAM3 Challenges}}.
+	\newblock \emph{PLoS ONE}, 5\penalty0 (2):\penalty0 e9202, January 2010.
+	\newblock ISSN 1932-6203.
+	\newblock \doi{10.1371/journal.pone.0009202}.
+
+	\bibitem[Marbach et~al.(2010)Marbach, Prill, Schaffter, Mattiussi, Floreano,
+	and Stolovitzky]{marbach_revealingstrengthsweaknesses_2010}
+	Daniel Marbach, Robert~J Prill, Thomas Schaffter, Claudio Mattiussi, Dario
+	Floreano, and Gustavo Stolovitzky.
+	\newblock Revealing strengths and weaknesses of methods for gene network
+	inference.
+	\newblock \emph{Proceedings of the National Academy of Sciences}, 107\penalty0
+	(14):\penalty0 6286--6291, April 2010.
+	\newblock ISSN 1091-6490.
+	\newblock \doi{10.1073/pnas.0913357107}.
+
+	\bibitem[Marbach et~al.(2012)Marbach, Costello, K{\"u}ffner, Vega, Prill,
+	Camacho, Allison, Kellis, Collins, Stolovitzky, {DREAM5 Consortium}, Kellis,
+	Collins, and Stolovitzky]{marbach_wisdomcrowdsrobust_2012}
+	Daniel Marbach, James~C Costello, Robert K{\"u}ffner, Nicole M~Nicci Vega,
+	Robert~J Prill, Diogo~M Camacho, Kyle~R Allison, Manolis Kellis, James~J
+	Collins, Gustavo Stolovitzky, {DREAM5 Consortium}, Manolis Kellis, James~J
+	Collins, and Gustavo Stolovitzky.
+	\newblock Wisdom of {{Crowds}} for {{Robust Gene Network Inference}}.
+	\newblock \emph{Nature methods}, 9\penalty0 (8):\penalty0 796--804, July 2012.
+	\newblock ISSN 1548-7091.
+	\newblock \doi{10.1038/nmeth.2016}.
+
+	\bibitem[Luecken and Theis(2019)]{luecken_currentbestpractices_2019}
+	Malte~D Luecken and Fabian~J Theis.
+	\newblock Current best practices in single-cell {{RNA}}-seq analysis: A
+	tutorial.
+	\newblock \emph{Molecular Systems Biology}, 15\penalty0 (6):\penalty0 e8746,
+	June 2019.
+	\newblock ISSN 1744-4292.
+	\newblock \doi{10.15252/msb.20188746}.
+
+	\bibitem[Vallejos et~al.(2017)Vallejos, Risso, Scialdone, Dudoit, and
+	Marioni]{vallejos_normalizingsinglecellrna_2017}
+	Catalina~A. Vallejos, Davide Risso, Antonio Scialdone, Sandrine Dudoit, and
+	John~C. Marioni.
+	\newblock Normalizing single-cell {{RNA}} sequencing data: Challenges and
+	opportunities.
+	\newblock \emph{Nature Methods}, 14\penalty0 (6):\penalty0 565--571, June 2017.
+	\newblock ISSN 1548-7105.
+	\newblock \doi{10.1038/nmeth.4292}.
+
+	\bibitem[Gillespie(1977)]{gillespie_exactstochasticsimulation_1977}
+	Daniel~T. Gillespie.
+	\newblock Exact stochastic simulation of coupled chemical reactions.
+	\newblock \emph{The Journal of Physical Chemistry}, 81\penalty0 (25):\penalty0
+	2340--2361, December 1977.
+	\newblock ISSN 0022-3654.
+	\newblock \doi{10.1021/j100540a008}.
+
+	\bibitem[Vieth et~al.(2017)Vieth, Ziegenhain, Parekh, Enard, and
+	Hellmann]{vieth_powsimrpoweranalysis_2017}
+	Beate Vieth, Christoph Ziegenhain, Swati Parekh, Wolfgang Enard, and Ines
+	Hellmann.
+	\newblock {{powsimR}}: Power analysis for bulk and single cell {{RNA}}-seq
+	experiments.
+	\newblock \emph{Bioinformatics}, 33\penalty0 (21):\penalty0 3486--3488,
+	November 2017.
+	\newblock ISSN 1367-4803.
+	\newblock \doi{10.1093/bioinformatics/btx435}.
+
+	\bibitem[Papadopoulos et~al.(2019)Papadopoulos, Gonzalo, and
+	S{\"o}ding]{papadopoulos_prossttprobabilisticsimulation_2019}
+	Nikolaos Papadopoulos, Parra~R. Gonzalo, and Johannes S{\"o}ding.
+	\newblock {{PROSSTT}}: Probabilistic simulation of single-cell {{RNA}}-seq data
+	for complex differentiation processes.
+	\newblock \emph{Bioinformatics (Oxford, England)}, 35\penalty0 (18):\penalty0
+	3517--3519, September 2019.
+	\newblock ISSN 1367-4811.
+	\newblock \doi{10.1093/bioinformatics/btz078}.
+
+	\bibitem[Zhang et~al.(2019)Zhang, Xu, and
+	Yosef]{zhang_simulatingmultiplefaceted_2019}
+	Xiuwei Zhang, Chenling Xu, and Nir Yosef.
+	\newblock Simulating multiple faceted variability in single cell {{RNA}}
+	sequencing.
+	\newblock \emph{Nature Communications}, 10\penalty0 (1):\penalty0 1--16, June
+	2019.
+	\newblock ISSN 2041-1723.
+	\newblock \doi{10.1038/s41467-019-10500-w}.
+
+	\bibitem[Street et~al.(2018)Street, Risso, Fletcher, Das, Ngai, Yosef, Purdom,
+	and Dudoit]{street_slingshotcelllineage_2018}
+	Kelly Street, Davide Risso, Russell~B Fletcher, Diya Das, John Ngai, Nir Yosef,
+	Elizabeth Purdom, and Sandrine Dudoit.
+	\newblock Slingshot: {{Cell Lineage}} and {{Pseudotime Inference}} for
+	{{Single}}-{{Cell Transcriptomics}}.
+	\newblock \emph{BMC Genomics}, 19\penalty0 (1):\penalty0 477, June 2018.
+	\newblock ISSN 1471-2164.
+	\newblock \doi{10.1186/s12864-018-4772-0}.
+
+	\bibitem[Parra et~al.(2019)Parra, Papadopoulos, {Ahumada-Arranz}, Kholtei,
+	Mottelson, Horokhovsky, Treutlein, and
+	Soeding]{parra_reconstructingcomplexlineage_2019}
+	R.~Gonzalo Parra, Nikolaos Papadopoulos, Laura {Ahumada-Arranz}, Jakob~El
+	Kholtei, Noah Mottelson, Yehor Horokhovsky, Barbara Treutlein, and Johannes
+	Soeding.
+	\newblock Reconstructing complex lineage trees from {{scRNA}}-seq data using
+	{{MERLoT}}.
+	\newblock \emph{Nucleic Acids Research}, 47\penalty0 (17):\penalty0 8961--8974,
+	September 2019.
+	\newblock ISSN 1362-4962.
+	\newblock \doi{10.1093/nar/gkz706}.
+
+	\bibitem[{Lummertz da Rocha} et~al.(2018){Lummertz da Rocha}, Rowe, Lundin,
+	Malleshaiah, Jha, Rambo, Li, North, Collins, and
+	Daley]{lummertzdarocha_reconstructioncomplexsinglecell_2018}
+	Edroaldo {Lummertz da Rocha}, R.~Grant Rowe, Vanessa Lundin, Mohan Malleshaiah,
+	Deepak~Kumar Jha, Carlos~R. Rambo, Hu~Li, Trista~E. North, James~J. Collins,
+	and George~Q. Daley.
+	\newblock Reconstruction of complex single-cell trajectories using
+	{{CellRouter}}.
+	\newblock \emph{Nature Communications}, 9\penalty0 (1):\penalty0 892, March
+	2018.
+	\newblock ISSN 2041-1723.
+	\newblock \doi{10.1038/s41467-018-03214-y}.
+
+	\bibitem[Lin et~al.(2020)Lin, Cao, Kim, Salim, Speed, Lin, Yang, and
+	Yang]{lin_scclassifysamplesize_2020}
+	Yingxin Lin, Yue Cao, Hani~Jieun Kim, Agus Salim, Terence~P Speed, David~M Lin,
+	Pengyi Yang, and Jean Yee~Hwa Yang.
+	\newblock {{scClassify}}: Sample size estimation and multiscale classification
+	of cells using single and multiple reference.
+	\newblock \emph{Molecular Systems Biology}, 16\penalty0 (6):\penalty0 e9389,
+	June 2020.
+	\newblock ISSN 1744-4292.
+	\newblock \doi{10.15252/msb.20199389}.
+
+	\bibitem[Du{\`o} et~al.(2018)Du{\`o}, Robinson, and
+	Soneson]{duo_systematicperformanceevaluation_2018}
+	Angelo Du{\`o}, Mark~D. Robinson, and Charlotte Soneson.
+	\newblock A systematic performance evaluation of clustering methods for
+	single-cell {{RNA}}-seq data.
+	\newblock \emph{F1000Research}, 7:\penalty0 1141, 2018.
+	\newblock ISSN 2046-1402.
+	\newblock \doi{10.12688/f1000research.15666.2}.
+
+	\bibitem[Saelens et~al.(2019)Saelens, Cannoodt, Todorov, and
+	Saeys]{saelens_comparisonsinglecelltrajectory_2019}
+	Wouter Saelens, Robrecht Cannoodt, Helena Todorov, and Yvan Saeys.
+	\newblock A comparison of single-cell trajectory inference methods.
+	\newblock \emph{Nature Biotechnology}, 37\penalty0 (May), 2019.
+	\newblock ISSN 15461696.
+	\newblock \doi{10.1038/s41587-019-0071-9}.
+
+	\bibitem[Soneson and
+	Robinson(2018{\natexlab{a}})]{soneson_biasrobustnessscalability_2018}
+	Charlotte Soneson and Mark~D. Robinson.
+	\newblock Bias, robustness and scalability in single-cell differential
+	expression analysis.
+	\newblock \emph{Nature Methods}, 15\penalty0 (4):\penalty0 255--261, April
+	2018{\natexlab{a}}.
+	\newblock ISSN 1548-7105.
+	\newblock \doi{10.1038/nmeth.4612}.
+
+	\bibitem[Weber et~al.(2019)Weber, Saelens, Cannoodt, Soneson, Hapfelmeier,
+	Gardner, Boulesteix, Saeys, and
+	Robinson]{weber_essentialguidelinescomputational_2019}
+	Lukas~M. Weber, Wouter Saelens, Robrecht Cannoodt, Charlotte Soneson, Alexander
+	Hapfelmeier, Paul~P. Gardner, Anne-Laure Boulesteix, Yvan Saeys, and Mark~D.
+	Robinson.
+	\newblock Essential guidelines for computational method benchmarking.
+	\newblock \emph{Genome Biology}, 20\penalty0 (1):\penalty0 125, June 2019.
+	\newblock ISSN 1474-760X.
+	\newblock \doi{10.1186/s13059-019-1738-8}.
+
+	\bibitem[Giorgino(2009)]{giorgino_computingvisualizingdynamic_2009}
+	Toni Giorgino.
+	\newblock Computing and {{Visualizing Dynamic Time Warping Alignments}} in
+	{{R}}: {{The}} dtw {{Package}}.
+	\newblock \emph{Journal of Statistical Software}, \penalty0 (7), September
+	2009.
+	\newblock \doi{10.18637/jss.v031.i07}.
+
+	\bibitem[Cacchiarelli et~al.(2018)Cacchiarelli, Qiu, Srivatsan, Manfredi,
+	Ziller, Overbey, Grimaldi, Grimsby, Pokharel, Livak, Li, Meissner, Mikkelsen,
+	Rinn, and Trapnell]{cacchiarelli_aligningsinglecelldevelopmental_2018}
 	Davide Cacchiarelli, Xiaojie Qiu, Sanjay Srivatsan, Anna Manfredi, Michael
@@
 			Outcome}}.
-	\newblock {\em Cell Systems}, 7(3):258--268.e3, 9 2018.
-
-	\bibitem{cannoodt_singlecellomicsdatasets_2018}
-	Robrecht Cannoodt, Wouter Saelens, Helena Todorov, and Yvan Saeys.
-	\newblock Single-cell -omics datasets containing a trajectory.
-	\newblock {\em Zenodo}, 10 2018.
-
-	\bibitem{conover_multiplecomparisonsprocedures_1979}
-	William~Jay Conover and Ronald~L Iman.
-	\newblock On multiple-comparisons procedures.
-	\newblock Technical report, {Technical report, Los Alamos Scientific
-		Laboratory}, 1979.
-
-	\bibitem{duo_systematicperformanceevaluation_2018}
-	Angelo Du{\a`o}, Mark~D. Robinson, and Charlotte Soneson.
-	\newblock A systematic performance evaluation of clustering methods for
-	single-cell {{RNA}}-seq data.
-	\newblock {\em F1000Research}, 7:1141, 2018.
-
-	\bibitem{ewing_combiningtumorgenome_2015}
+	\newblock \emph{Cell Systems}, 7\penalty0 (3):\penalty0 258--268.e3, September
+	2018.
+	\newblock ISSN 2405-4712.
+	\newblock \doi{10.1016/j.cels.2018.07.006}.
+
+	\bibitem[Kanton et~al.(2019)Kanton, Boyle, He, Santel, Weigert,
+	{Sanch{\'i}s-Calleja}, Guijarro, Sidow, Fleck, Han, Qian, Heide, Huttner,
+	Khaitovich, P{\"a}{\"a}bo, Treutlein, and
+	Camp]{kanton_organoidsinglecellgenomic_2019}
+	Sabina Kanton, Michael~James Boyle, Zhisong He, Malgorzata Santel, Anne
+	Weigert, F{\'a}tima {Sanch{\'i}s-Calleja}, Patricia Guijarro, Leila Sidow,
+	Jonas~Simon Fleck, Dingding Han, Zhengzong Qian, Michael Heide, Wieland~B.
+	Huttner, Philipp Khaitovich, Svante P{\"a}{\"a}bo, Barbara Treutlein, and
+	J.~Gray Camp.
+	\newblock Organoid single-cell genomic atlas uncovers human-specific features
+	of brain development.
+	\newblock \emph{Nature}, 574\penalty0 (7778):\penalty0 418--422, October 2019.
+	\newblock ISSN 1476-4687.
+	\newblock \doi{10.1038/s41586-019-1654-9}.
+
+	\bibitem[{McFaline-Figueroa} et~al.(2019){McFaline-Figueroa}, Hill, Qiu,
+	Jackson, Shendure, and
+	Trapnell]{mcfaline-figueroa_pooledsinglecellgenetic_2019}
+	Jos{\'e}~L. {McFaline-Figueroa}, Andrew~J. Hill, Xiaojie Qiu, Dana Jackson, Jay
+	Shendure, and Cole Trapnell.
+	\newblock A pooled single-cell genetic screen identifies regulatory checkpoints
+	in the continuum of the epithelial-to-mesenchymal transition.
+	\newblock \emph{Nature Genetics}, 51\penalty0 (9):\penalty0 1389--1398,
+	September 2019.
+	\newblock ISSN 1546-1718.
+	\newblock \doi{10.1038/s41588-019-0489-5}.
+
+	\bibitem[Alpert et~al.(2018)Alpert, Moore, Dubovik, and
+	{Shen-Orr}]{alpert_alignmentsinglecelltrajectories_2018}
+	Ayelet Alpert, Lindsay~S. Moore, Tania Dubovik, and Shai~S. {Shen-Orr}.
+	\newblock Alignment of single-cell trajectories to compare cellular expression
+	dynamics.
+	\newblock \emph{Nature Methods}, 15\penalty0 (4):\penalty0 267--270, April
+	2018.
+	\newblock ISSN 1548-7105.
+	\newblock \doi{10.1038/nmeth.4628}.
+
+	\bibitem[Zeisel et~al.(2011)Zeisel, K{\"o}stler, Molotski, Tsai, Krauthgamer,
+	{Jacob-Hirsch}, Rechavi, Soen, Jung, Yarden, and
+	Domany]{zeisel_coupledpremrnamrna_2011}
+	Amit Zeisel, Wolfgang~J K{\"o}stler, Natali Molotski, Jonathan~M Tsai, Rita
+	Krauthgamer, Jasmine {Jacob-Hirsch}, Gideon Rechavi, Yoav Soen, Steffen Jung,
+	Yosef Yarden, and Eytan Domany.
+	\newblock Coupled pre-{{mRNA}} and {{mRNA}} dynamics unveil operational
+	strategies underlying transcriptional responses to stimuli.
+	\newblock \emph{Molecular Systems Biology}, 7\penalty0 (1):\penalty0 529,
+	January 2011.
+	\newblock ISSN 1744-4292.
+	\newblock \doi{10.1038/msb.2011.62}.
+
+	\bibitem[La~Manno et~al.(2018)La~Manno, Soldatov, Zeisel, Braun, Hochgerner,
+	Petukhov, Lidschreiber, Kastriti, L{\"o}nnerberg, Furlan, Fan, Borm, Liu,
+	{van Bruggen}, Guo, He, Barker, Sundstr{\"o}m, {Castelo-Branco}, Cramer,
+	Adameyko, Linnarsson, and Kharchenko]{lamanno_rnavelocitysingle_2018}
+	Gioele La~Manno, Ruslan Soldatov, Amit Zeisel, Emelie Braun, Hannah Hochgerner,
+	Viktor Petukhov, Katja Lidschreiber, Maria~E Kastriti, Peter L{\"o}nnerberg,
+	Alessandro Furlan, Jean Fan, Lars~E Borm, Zehua Liu, David {van Bruggen},
+	Jimin Guo, Xiaoling He, Roger Barker, Erik Sundstr{\"o}m, Gon{\c c}alo
+	{Castelo-Branco}, Patrick Cramer, Igor Adameyko, Sten Linnarsson, and Peter~V
+	Kharchenko.
+	\newblock {{RNA Velocity}} of {{Single Cells}}.
+	\newblock \emph{Nature}, 560\penalty0 (7719):\penalty0 494--498, August 2018.
+	\newblock ISSN 1476-4687.
+	\newblock \doi{10.1038/s41586-018-0414-6}.
+
+	\bibitem[Bergen et~al.(2020)Bergen, Lange, Peidli, Wolf, and
+	Theis]{bergen_generalizingrnavelocity_2020}
+	Volker Bergen, Marius Lange, Stefan Peidli, F.~Alexander Wolf, and Fabian~J.
+	Theis.
+	\newblock Generalizing {{RNA}} velocity to transient cell states through
+	dynamical modeling.
+	\newblock \emph{Nature Biotechnology}, 38\penalty0 (12):\penalty0 1408--1414,
+	December 2020.
+	\newblock ISSN 1546-1696.
+	\newblock \doi{10.1038/s41587-020-0591-3}.
+
+	\bibitem[Aibar et~al.(2017)Aibar, {Gonz{\'a}lez-Blas}, Moerman, {Huynh-Thu},
+	Imrichova, Hulselmans, Rambow, Marine, Geurts, Aerts, {van den Oord}, Atak,
+	Wouters, and Aerts]{aibar_scenicsinglecellregulatory_2017}
+	Sara Aibar, Carmen~Bravo {Gonz{\'a}lez-Blas}, Thomas Moerman, V{\^a}n~Anh
+	{Huynh-Thu}, Hana Imrichova, Gert Hulselmans, Florian Rambow, Jean-Christophe
+	Marine, Pierre Geurts, Jan Aerts, Joost {van den Oord}, Zeynep~Kalender Atak,
+	Jasper Wouters, and Stein Aerts.
+	\newblock {{SCENIC}}: Single-cell regulatory network inference and clustering.
+	\newblock \emph{Nature Methods}, October 2017.
+	\newblock ISSN 1548-7091.
+	\newblock \doi{10.1038/nmeth.4463}.
+
+	\bibitem[Kuijjer et~al.(2019{\natexlab{a}})Kuijjer, Tung, Yuan, Quackenbush,
+	and Glass]{kuijjer_estimatingsamplespecificregulatory_2019}
+	Marieke~Lydia Kuijjer, Matthew~George Tung, GuoCheng Yuan, John Quackenbush,
+	and Kimberly Glass.
+	\newblock Estimating {{Sample}}-{{Specific Regulatory Networks}}.
+	\newblock \emph{iScience}, 14:\penalty0 226--240, March 2019{\natexlab{a}}.
+	\newblock ISSN 2589-0042.
+	\newblock \doi{10.1016/j.isci.2019.03.021}.
+
+	\bibitem[Liu et~al.(2016)Liu, Wang, Ji, Aihara, and
+	Chen]{liu_personalizedcharacterizationdiseases_2016}
+	Xiaoping Liu, Yuetong Wang, Hongbin Ji, Kazuyuki Aihara, and Luonan Chen.
+	\newblock Personalized characterization of diseases using sample-specific
+	networks.
+	\newblock \emph{Nucleic Acids Research}, 44\penalty0 (22):\penalty0 e164--e164,
+	2016.
+	\newblock ISSN 0305-1048.
+	\newblock \doi{10.1093/nar/gkw772}.
+
+	\bibitem[{Huynh-Thu} et~al.(2010){Huynh-Thu}, Alexandre, Louis, Pierre,
+	Irrthum, Wehenkel, and Geurts]{huynh-thu_inferringregulatorynetworks_2010}
+	V{\^a}n~Anh {Huynh-Thu}, Irrthum Alexandre, Wehenkel Louis, Geurts Pierre,
+	Alexandre Irrthum, Louis Wehenkel, and Pierre Geurts.
+	\newblock Inferring {{Regulatory Networks}} from {{Expression Data Using
+			Tree}}-{{Based Methods}}.
+	\newblock \emph{PLoS ONE}, 5\penalty0 (9):\penalty0 e12776, January 2010.
+	\newblock ISSN 1932-6203.
+	\newblock \doi{10.1371/journal.pone.0012776}.
+
+	\bibitem[Moerman et~al.(2019)Moerman, Aibar~Santos, {Bravo Gonz{\'a}lez-Blas},
+	Simm, Moreau, Aerts, and Aerts]{moerman_grnboost2arboretoefficient_2019}
+	Thomas Moerman, Sara Aibar~Santos, Carmen {Bravo Gonz{\'a}lez-Blas}, Jaak Simm,
+	Yves Moreau, Jan Aerts, and Stein Aerts.
+	\newblock {{GRNBoost2}} and {{Arboreto}}: Efficient and scalable inference of
+	gene regulatory networks.
+	\newblock \emph{Bioinformatics}, 35\penalty0 (12):\penalty0 2159--2161, June
+	2019.
+	\newblock ISSN 1367-4803.
+	\newblock \doi{10.1093/bioinformatics/bty916}.
+
+	\bibitem[Ewing et~al.(2015)Ewing, Houlahan, Hu, Ellrott, Caloian, Yamaguchi,
+	Bare, P'ng, Waggott, Sabelnykova, {ICGC-TCGA DREAM Somatic Mutation Calling
+		Challenge Participants}, Xi, Dewal, Fan, Wang, Wheeler, Wilm, Ting, Li,
+	Bertrand, Nagarajan, Chen, Hsu, Hu, Yan, Kibbe, Meerzaman, Cibulskis,
+	Rosenberg, Bergelson, Kiezun, Radenbaugh, Sertier, Ferrari, Tonton, Bhutani,
+	Hansen, Wang, Song, Lai, Liao, Shi, {Carbonell-Caballero}, Dopazo, Lau,
+	Guinney, Kellen, Norman, Haussler, Friend, Stolovitzky, Margolin, Stuart, and
+	Boutros]{ewing_combiningtumorgenome_2015}
 	Adam~D. Ewing, Kathleen~E. Houlahan, Yin Hu, Kyle Ellrott, Cristian Caloian,
@@
 	Bhutani, Nancy~F. Hansen, Difei Wang, Lei Song, Zhongwu Lai, Yang Liao, Wei
-	Shi, Jos{\a'e} Carbonell-Caballero, Joaqu\'in Dopazo, Cheryl C.~K. Lau,
+	Shi, Jos{\'e} {Carbonell-Caballero}, Joaqu{\'i}n Dopazo, Cheryl C.~K. Lau,
 	Justin Guinney, Michael~R. Kellen, Thea~C. Norman, David Haussler, Stephen~H.
@@
 	somatic single-nucleotide-variant detection.
-	\newblock {\em Nature Methods}, 12(7):623--630, 7 2015.
-
-	\bibitem{ferrell_bistabilitybifurcationswaddington_2012}
+	\newblock \emph{Nature Methods}, 12\penalty0 (7):\penalty0 623--630, July 2015.
+	\newblock ISSN 1548-7105.
+	\newblock \doi{10.1038/nmeth.3407}.
+
+	\bibitem[Smith and Grima(2019)]{smith_spatialstochasticintracellular_2019}
+	Stephen Smith and Ramon Grima.
+	\newblock Spatial {{Stochastic Intracellular Kinetics}}: {{A Review}} of
+	{{Modelling Approaches}}.
+	\newblock \emph{Bulletin of Mathematical Biology}, 81\penalty0 (8):\penalty0
+	2960--3009, August 2019.
+	\newblock ISSN 1522-9602.
+	\newblock \doi{10.1007/s11538-018-0443-1}.
+
+	\bibitem[Rekhtman et~al.(1999)Rekhtman, Radparvar, Evans, and
+	Skoultchi]{rekhtman_directinteractionhematopoietic_1999}
+	N.~Rekhtman, F.~Radparvar, T.~Evans, and A.~I. Skoultchi.
+	\newblock Direct interaction of hematopoietic transcription factors {{PU}}.1
+	and {{GATA}}-1: Functional antagonism in erythroid cells.
+	\newblock \emph{Genes \& Development}, 13\penalty0 (11):\penalty0 1398--1411,
+	June 1999.
+	\newblock ISSN 0890-9369.
+	\newblock \doi{10.1101/gad.13.11.1398}.
+
+	\bibitem[Xu et~al.(2015)Xu, Chaudhri, Wu, Biliouris, {Dienger-Stambaugh},
+	Rochman, and Singh]{xu_regulationbifurcatingcell_2015}
+	Heping Xu, Virendra~K Chaudhri, Zhiguo Wu, Konstantinos Biliouris, Krista
+	{Dienger-Stambaugh}, Yrina Rochman, and Harinder Singh.
+	\newblock Regulation of bifurcating {{B}} cell trajectories by mutual
+	antagonism between transcription factors {{IRF4}} and {{IRF8}}.
+	\newblock \emph{Nat. Immunol.}, 16\penalty0 (12):\penalty0 1274--1281, December
+	2015.
+
+	\bibitem[Graf and Enver(2009)]{graf_forcingcellschange_2009}
+	Thomas Graf and Tariq Enver.
+	\newblock Forcing {{Cells}} to {{Change Lineages}}.
+	\newblock \emph{Nature}, 462\penalty0 (7273):\penalty0 587, December 2009.
+	\newblock ISSN 1476-4687.
+	\newblock \doi{10.1038/nature08533}.
+
+	\bibitem[Wang et~al.(2011)Wang, Zhang, Xu, and
+	Wang]{wang_quantifyingwaddingtonlandscape_2011}
+	Jin Wang, Kun Zhang, Li~Xu, and Erkang Wang.
+	\newblock Quantifying the {{Waddington Landscape}} and {{Biological Paths}} for
+	{{Development}} and {{Differentiation}}.
+	\newblock \emph{Proceedings of the National Academy of Sciences}, 108\penalty0
+	(20):\penalty0 8257--8262, May 2011.
+	\newblock ISSN 0027-8424, 1091-6490.
+	\newblock \doi{10.1073/pnas.1017017108}.
+
+	\bibitem[Ferrell(2012)]{ferrell_bistabilitybifurcationswaddington_2012}
 	James~E Ferrell.
@@
 			Landscape}}.
-	\newblock {\em Current Biology}, 22(11):R458--R466, 6 2012.
-
-	\bibitem{gibson_probabilisticmodelprokaryotic_2000}
-	Michael~A. Gibson and Jehoshua Bruck.
-	\newblock A probabilistic model of a prokaryotic gene and its regulation.
-	\newblock {\em Computational Methods in Molecular Biology: From Genotype to
-		Phenotype, MIT press, Cambridge}, 2000.
-
-	\bibitem{gillespie_exactstochasticsimulation_1977}
-	Daniel~T. Gillespie.
-	\newblock Exact stochastic simulation of coupled chemical reactions.
-	\newblock {\em The Journal of Physical Chemistry}, 81(25):2340--2361, 12 1977.
-
-	\bibitem{giorgino_computingvisualizingdynamic_2009}
-	Toni Giorgino.
-	\newblock Computing and {{Visualizing Dynamic Time Warping Alignments}} in
-	{{R}}: {{The}} dtw {{Package}}.
-	\newblock {\em Journal of Statistical Software}, (7), 9 2009.
-
-	\bibitem{glass_passingmessagesbiological_2013}
-	Kimberly Glass, Curtis Huttenhower, John Quackenbush, and Guo-Cheng Yuan.
-	\newblock Passing {{Messages}} between {{Biological Networks}} to {{Refine
-			Predicted Interactions}}.
-	\newblock {\em PLOS ONE}, 8(5):e64832, 5 2013.
-
-	\bibitem{graf_forcingcellschange_2009}
-	Thomas Graf and Tariq Enver.
-	\newblock Forcing {{Cells}} to {{Change Lineages}}.
-	\newblock {\em Nature}, 462(7273):587, 12 2009.
-
-	\bibitem{hache_gengesystematicgeneration_2009}
-	Hendrik Hache, Christoph Wierling, Hans Lehrach, and Ralf Herwig.
-	\newblock {{GeNGe}}: Systematic generation of gene regulatory networks.
-	\newblock {\em Bioinformatics}, 25(9):1205--1207, 5 2009.
-
-	\bibitem{holm_simplesequentiallyrejective_1979}
-	Sture Holm.
-	\newblock A simple sequentially rejective multiple test procedure.
-	\newblock {\em Scandinavian journal of statistics}, pages 65--70, 1979.
-
-	\bibitem{huynh-thu_inferringregulatorynetworks_2010}
-	V{\^a}n~Anh Huynh-Thu, Irrthum Alexandre, Wehenkel Louis, Geurts Pierre,
-	Alexandre Irrthum, Louis Wehenkel, and Pierre Geurts.
-	\newblock Inferring {{Regulatory Networks}} from {{Expression Data Using
-			Tree}}-{{Based Methods}}.
-	\newblock {\em PLoS ONE}, 5(9):e12776, 1 2010.
-
-	\bibitem{kanton_organoidsinglecellgenomic_2019}
-	Sabina Kanton, Michael~James Boyle, Zhisong He, Malgorzata Santel, Anne
-	Weigert, F{\a'a}tima Sanch\'is-Calleja, Patricia Guijarro, Leila Sidow,
-	Jonas~Simon Fleck, Dingding Han, Zhengzong Qian, Michael Heide, Wieland~B.
-	Huttner, Philipp Khaitovich, Svante P{\"a}{{\"a}}bo, Barbara Treutlein, and
-	J.~Gray Camp.
-	\newblock Organoid single-cell genomic atlas uncovers human-specific features
-	of brain development.
-	\newblock {\em Nature}, 574(7778):418--422, 10 2019.
-
-	\bibitem{kuijjer_lionessrsinglesample_2019}
-	Marieke~L. Kuijjer, Ping-Han Hsieh, John Quackenbush, and Kimberly Glass.
-	\newblock {{lionessR}}: Single sample network inference in {{R}}.
-	\newblock {\em BMC Cancer}, 19(1):1003, 10 2019.
-
-	\bibitem{kuijjer_estimatingsamplespecificregulatory_2015}
-	Marieke~Lydia Kuijjer, Matthew Tung, GuoCheng Yuan, John Quackenbush, and
-	Kimberly Glass.
-	\newblock Estimating sample-specific regulatory networks.
-	\newblock pages 1--19, 2015.
-
-	\bibitem{kuijjer_estimatingsamplespecificregulatory_2019}
-	Marieke~Lydia Kuijjer, Matthew~George Tung, GuoCheng Yuan, John Quackenbush,
-	and Kimberly Glass.
-	\newblock Estimating {{Sample}}-{{Specific Regulatory Networks}}.
-	\newblock {\em iScience}, 14:226--240, 3 2019.
-
-	\bibitem{lamanno_rnavelocitysingle_2018}
-	Gioele {La Manno}, Ruslan Soldatov, Amit Zeisel, Emelie Braun, Hannah
-	Hochgerner, Viktor Petukhov, Katja Lidschreiber, Maria~E Kastriti, Peter
-	L{\"o}nnerberg, Alessandro Furlan, Jean Fan, Lars~E Borm, Zehua Liu, David
-	{van Bruggen}, Jimin Guo, Xiaoling He, Roger Barker, Erik Sundstr{\"o}m,
-	Gon{\c c}alo Castelo-Branco, Patrick Cramer, Igor Adameyko, Sten Linnarsson,
-	and Peter~V Kharchenko.
-	\newblock {{RNA Velocity}} of {{Single Cells}}.
-	\newblock {\em Nature}, 560(7719):494--498, 8 2018.
-
-	\bibitem{lin_scclassifysamplesize_2020}
-	Yingxin Lin, Yue Cao, Hani~Jieun Kim, Agus Salim, Terence~P Speed, David~M Lin,
-	Pengyi Yang, and Jean Yee~Hwa Yang.
-	\newblock {{scClassify}}: Sample size estimation and multiscale classification
-	of cells using single and multiple reference.
-	\newblock {\em Molecular Systems Biology}, 16(6):e9389, 6 2020.
-
-	\bibitem{liu_personalizedcharacterizationdiseases_2016}
-	Xiaoping Liu, Yuetong Wang, Hongbin Ji, Kazuyuki Aihara, and Luonan Chen.
-	\newblock Personalized characterization of diseases using sample-specific
-	networks.
-	\newblock {\em Nucleic Acids Research}, 44(22):e164--e164, 2016.
-
-	\bibitem{lizio_gatewaysfantom5promoter_2015}
-	Marina Lizio, Jayson Harshbarger, Hisashi Shimoji, Jessica Severin, Takeya
-	Kasukawa, Serkan Sahin, Imad Abugessaisa, Shiro Fukuda, Fumi Hori, Sachi
-	Ishikawa-Kato, Christopher~J. Mungall, Erik Arner, J.~Kenneth Baillie,
-	Nicolas Bertin, Hidemasa Bono, Michiel {de Hoon}, Alexander~D. Diehl,
-	Emmanuel Dimont, Tom~C. Freeman, Kaori Fujieda, Winston Hide, Rajaram
-	Kaliyaperumal, Toshiaki Katayama, Timo Lassmann, Terrence~F. Meehan, Koro
-	Nishikata, Hiromasa Ono, Michael Rehli, Albin Sandelin, Erik~A. Schultes,
-	Peter~AC {`t Hoen}, Zuotian Tatum, Mark Thompson, Tetsuro Toyoda, Derek~W.
-	Wright, Carsten~O. Daub, Masayoshi Itoh, Piero Carninci, Yoshihide
-	Hayashizaki, Alistair~RR Forrest, Hideya Kawaji, and {the FANTOM consortium}.
-	\newblock Gateways to the {{FANTOM5}} promoter level mammalian expression
-	atlas.
-	\newblock {\em Genome Biology}, 16(1):22, 1 2015.
-
-	\bibitem{luecken_currentbestpractices_2019}
-	Malte~D Luecken and Fabian~J Theis.
-	\newblock Current best practices in single-cell {{RNA}}-seq analysis: A
-	tutorial.
-	\newblock {\em Molecular Systems Biology}, 15(6):e8746, 6 2019.
-
-	\bibitem{lummertzdarocha_reconstructioncomplexsinglecell_2018}
-	Edroaldo {Lummertz da Rocha}, R.~Grant Rowe, Vanessa Lundin, Mohan Malleshaiah,
-	Deepak~Kumar Jha, Carlos~R. Rambo, Hu~Li, Trista~E. North, James~J. Collins,
-	and George~Q. Daley.
-	\newblock Reconstruction of complex single-cell trajectories using
-	{{CellRouter}}.
-	\newblock {\em Nature Communications}, 9(1):892, 3 2018.
-
-	\bibitem{marbach_wisdomcrowdsrobust_2012}
-	Daniel Marbach, James~C Costello, Robert K{\"u}ffner, Nicole M~Nicci Vega,
-	Robert~J Prill, Diogo~M Camacho, Kyle~R Allison, Manolis Kellis, James~J
-	Collins, Gustavo Stolovitzky, {DREAM5 Consortium}, Manolis Kellis, James~J
-	Collins, and Gustavo Stolovitzky.
-	\newblock Wisdom of {{Crowds}} for {{Robust Gene Network Inference}}.
-	\newblock {\em Nature methods}, 9(8):796--804, 7 2012.
-
-	\bibitem{marbach_revealingstrengthsweaknesses_2010}
-	Daniel Marbach, Robert~J Prill, Thomas Schaffter, Claudio Mattiussi, Dario
-	Floreano, and Gustavo Stolovitzky.
-	\newblock Revealing strengths and weaknesses of methods for gene network
-	inference.
-	\newblock {\em Proceedings of the National Academy of Sciences},
-	107(14):6286--6291, 4 2010.
-
-	\bibitem{marbach_generatingrealisticsilico_2009}
-	Daniel Marbach, Thomas Schaffter, Claudio Mattiussi, and Dario Floreano.
-	\newblock Generating {{Realistic In Silico Gene Networks}} for {{Performance
-			Assessment}} of {{Reverse Engineering Methods}}.
-	\newblock {\em Journal of Computational Biology}, 16(2):229--239, 2 2009.
-
-	\bibitem{mcfaline-figueroa_pooledsinglecellgenetic_2019}
-	Jos{\a'e}~L. McFaline-Figueroa, Andrew~J. Hill, Xiaojie Qiu, Dana Jackson, Jay
-	Shendure, and Cole Trapnell.
-	\newblock A pooled single-cell genetic screen identifies regulatory checkpoints
-	in the continuum of the epithelial-to-mesenchymal transition.
-	\newblock {\em Nature Genetics}, 51(9):1389--1398, 9 2019.
-
-	\bibitem{moerman_grnboost2arboretoefficient_2019}
-	Thomas Moerman, Sara {Aibar Santos}, Carmen {Bravo Gonz{\a'a}lez-Blas}, Jaak
-	Simm, Yves Moreau, Jan Aerts, and Stein Aerts.
-	\newblock {{GRNBoost2}} and {{Arboreto}}: Efficient and scalable inference of
-	gene regulatory networks.
-	\newblock {\em Bioinformatics}, 35(12):2159--2161, 6 2019.
-
-	\bibitem{papadopoulos_prossttprobabilisticsimulation_2019}
-	Nikolaos Papadopoulos, Parra~R. Gonzalo, and Johannes S{\"o}ding.
-	\newblock {{PROSSTT}}: Probabilistic simulation of single-cell {{RNA}}-seq data
-	for complex differentiation processes.
-	\newblock {\em Bioinformatics (Oxford, England)}, 35(18):3517--3519, 9 2019.
-
-	\bibitem{parra_reconstructingcomplexlineage_2019}
-	R.~Gonzalo Parra, Nikolaos Papadopoulos, Laura Ahumada-Arranz, Jakob~El
-	Kholtei, Noah Mottelson, Yehor Horokhovsky, Barbara Treutlein, and Johannes
-	Soeding.
-	\newblock Reconstructing complex lineage trees from {{scRNA}}-seq data using
-	{{MERLoT}}.
-	\newblock {\em Nucleic Acids Research}, 47(17):8961--8974, 9 2019.
-
-	\bibitem{patil_pairwisecomparisonsmultiplepairwise_2019}
-	Indrajeet Patil.
-	\newblock {\em {{pairwiseComparisons}}: {{Multiple}} Pairwise Comparison
-		Tests}, 2019.
-
-	\bibitem{prill_rigorousassessmentsystems_2010}
-	Robert~J Prill, Marbach Daniel, Saez-Rodriguez Julio, Sorger~Peter K.,
-	Alexopoulos~Leonidas G., Xue Xiaowei, Clarke~Neil D., Altan-Bonnet Gregoire,
-	Stolovitzky Gustavo, Daniel Marbach, Julio Saez-Rodriguez, Peter~K Sorger,
-	Leonidas~G Alexopoulos, Xiaowei Xue, Neil~D Clarke, Gregoire Altan-Bonnet,
-	Gustavo Stolovitzky, Marbach Daniel, Saez-Rodriguez Julio, Sorger~Peter K.,
-	Alexopoulos~Leonidas G., Xue Xiaowei, Clarke~Neil D., Altan-Bonnet Gregoire,
-	and Stolovitzky Gustavo.
-	\newblock Towards a {{Rigorous Assessment}} of {{Systems Biology Models}}:
-	{{The DREAM3 Challenges}}.
-	\newblock {\em PLoS ONE}, 5(2):e9202, 1 2010.
-
-	\bibitem{rekhtman_directinteractionhematopoietic_1999}
-	N.~Rekhtman, F.~Radparvar, T.~Evans, and A.~I. Skoultchi.
-	\newblock Direct interaction of hematopoietic transcription factors {{PU}}.1
-	and {{GATA}}-1: Functional antagonism in erythroid cells.
-	\newblock {\em Genes \& Development}, 13(11):1398--1411, 6 1999.
-
-	\bibitem{roy_systemgeneratingtranscription_2008}
-	Sushmita Roy, Margaret Werner-Washburne, and Terran Lane.
-	\newblock A system for generating transcription regulatory networks with
-	combinatorial control of transcription.
-	\newblock {\em Bioinformatics}, 24(10):1318--1320, 5 2008.
-
-	\bibitem{saelens_comparisonsinglecelltrajectory_2019}
-	Wouter Saelens, Robrecht Cannoodt, Helena Todorov, and Yvan Saeys.
-	\newblock A comparison of single-cell trajectory inference methods.
-	\newblock {\em Nature Biotechnology}, 37(5), 2019.
-
-	\bibitem{schaffter_genenetweaversilicobenchmark_2011}
-	Thomas Schaffter, Daniel Marbach, and Dario Floreano.
-	\newblock {{GeneNetWeaver}}: In silico benchmark generation and performance
-	profiling of network inference methods.
-	\newblock {\em Bioinformatics}, 27(16):2263--2270, 8 2011.
-
-	\bibitem{schilstra_biologicgeneexpression_2008}
-	Maria~J. Schilstra and Chrystopher~L. Nehaniv.
-	\newblock Bio-{{Logic}}: {{Gene Expression}} and the {{Laws}} of
-	{{Combinatorial Logic}}.
-	\newblock {\em Artificial Life}, 14(1):121--133, 1 2008.
-
-	\bibitem{schwanhausser_globalquantificationmammalian_2011}
-	Bj{\"o}rn Schwanh{\"a}usser, Dorothea Busse, Na~Li, Gunnar Dittmar, Johannes
-	Schuchhardt, Jana Wolf, Wei Chen, and Matthias Selbach.
-	\newblock Global quantification of mammalian gene expression control.
-	\newblock {\em Nature}, 473(7347):337--342, 5 2011.
-
-	\bibitem{smith_spatialstochasticintracellular_2019}
-	Stephen Smith and Ramon Grima.
-	\newblock Spatial {{Stochastic Intracellular Kinetics}}: {{A Review}} of
-	{{Modelling Approaches}}.
-	\newblock {\em Bulletin of Mathematical Biology}, 81(8):2960--3009, 8 2019.
-
-	\bibitem{soneson_biasrobustnessscalability_2018}
-	Charlotte Soneson and Mark~D. Robinson.
-	\newblock Bias, robustness and scalability in single-cell differential
-	expression analysis.
-	\newblock {\em Nature Methods}, 15(4):255--261, 4 2018.
-
-	\bibitem{street_slingshotcelllineage_2018}
-	Kelly Street, Davide Risso, Russell~B Fletcher, Diya Das, John Ngai, Nir Yosef,
-	Elizabeth Purdom, and Sandrine Dudoit.
-	\newblock Slingshot: {{Cell Lineage}} and {{Pseudotime Inference}} for
-	{{Single}}-{{Cell Transcriptomics}}.
-	\newblock {\em BMC Genomics}, 19(1):477, 6 2018.
-
-	\bibitem{vallejos_normalizingsinglecellrna_2017}
-	Catalina~A. Vallejos, Davide Risso, Antonio Scialdone, Sandrine Dudoit, and
-	John~C. Marioni.
-	\newblock Normalizing single-cell {{RNA}} sequencing data: Challenges and
-	opportunities.
-	\newblock {\em Nature Methods}, 14(6):565--571, 6 2017.
-
-	\bibitem{vandesande_scalablescenicworkflow_2020}
-	Bram {Van de Sande}, Christopher Flerin, Kristofer Davie, Maxime {De
-		Waegeneer}, Gert Hulselmans, Sara Aibar, Ruth Seurinck, Wouter Saelens,
-	Robrecht Cannoodt, Quentin Rouchon, Toni Verbeiren, Dries {De Maeyer}, Joke
-	Reumers, Yvan Saeys, and Stein Aerts.
-	\newblock A scalable {{SCENIC}} workflow for single-cell gene regulatory
-	network analysis.
-	\newblock {\em Nature Protocols}, 15(7):2247--2276, 7 2020.
-
-	\bibitem{vandenbulcke_syntrengeneratorsynthetic_2006}
-	Tim {Van den Bulcke}, Koenraad {Van Leemput}, Bart Naudts, Piet {van Remortel},
-	Hongwu Ma, Alain Verschoren, Bart {De Moor}, and Kathleen Marchal.
-	\newblock {{SynTReN}}: A generator of synthetic gene expression data for design
-	and analysis of structure learning algorithms.
-	\newblock {\em BMC Bioinformatics}, 7(1):43, 1 2006.
-
-	\bibitem{vieth_powsimrpoweranalysis_2017}
-	Beate Vieth, Christoph Ziegenhain, Swati Parekh, Wolfgang Enard, and Ines
-	Hellmann.
-	\newblock {{powsimR}}: Power analysis for bulk and single cell {{RNA}}-seq
-	experiments.
-	\newblock {\em Bioinformatics}, 33(21):3486--3488, 11 2017.
-
-	\bibitem{wang_quantifyingwaddingtonlandscape_2011}
-	Jin Wang, Kun Zhang, Li~Xu, and Erkang Wang.
-	\newblock Quantifying the {{Waddington Landscape}} and {{Biological Paths}} for
-	{{Development}} and {{Differentiation}}.
-	\newblock {\em Proceedings of the National Academy of Sciences},
-	108(20):8257--8262, 5 2011.
-
-	\bibitem{weber_essentialguidelinescomputational_2019}
-	Lukas~M. Weber, Wouter Saelens, Robrecht Cannoodt, Charlotte Soneson, Alexander
-	Hapfelmeier, Paul~P. Gardner, Anne-Laure Boulesteix, Yvan Saeys, and Mark~D.
-	Robinson.
-	\newblock Essential guidelines for computational method benchmarking.
-	\newblock {\em Genome Biology}, 20(1):125, 6 2019.
-
-	\bibitem{wickham_welcometidyverse_2019}
-	Hadley Wickham, Mara Averick, Jennifer Bryan, Winston Chang, Lucy McGowan,
-	Romain Fran{\c c}ois, Garrett Grolemund, Alex Hayes, Lionel Henry, Jim
-	Hester, Max Kuhn, Thomas Pedersen, Evan Miller, Stephan Bache, Kirill
-	M{\"u}ller, Jeroen Ooms, David Robinson, Dana Seidel, Vitalie Spinu, Kohske
-	Takahashi, Davis Vaughan, Claus Wilke, Kara Woo, and Hiroaki Yutani.
-	\newblock Welcome to the {{Tidyverse}}.
-	\newblock 11 2019.
-
-	\bibitem{xu_regulationbifurcatingcell_2015}
-	Heping Xu, Virendra~K Chaudhri, Zhiguo Wu, Konstantinos Biliouris, Krista
-	Dienger-Stambaugh, Yrina Rochman, and Harinder Singh.
-	\newblock Regulation of bifurcating {{B}} cell trajectories by mutual
-	antagonism between transcription factors {{IRF4}} and {{IRF8}}.
-	\newblock {\em Nat. Immunol.}, 16(12):1274--1281, 12 2015.
-
-	\bibitem{yosef_dynamicregulatorynetwork_2013}
+	\newblock \emph{Current Biology}, 22\penalty0 (11):\penalty0 R458--R466, June
+	2012.
+	\newblock ISSN 0960-9822.
+	\newblock \doi{10.1016/j.cub.2012.03.045}.
+
+	\bibitem[Yosef et~al.(2013)Yosef, Shalek, Gaublomme, Jin, Lee, Awasthi, Wu,
+	Karwacz, Xiao, Jorgolli, Gennert, Satija, Shakya, Lu, Trombetta, Pillai,
+	Ratcliffe, Coleman, Bix, Tantin, Park, Kuchroo, and
+	Regev]{yosef_dynamicregulatorynetwork_2013}
 	Nir Yosef, Alex~K Shalek, Jellert~T Gaublomme, Hulin Jin, Youjin Lee, Amit
@@
 	\{\vphantom\}{{TH17}}\vphantom\{\} cell differentiation.
-	\newblock {\em Nature}, 496(7446):461--468, 2013.
-
-	\bibitem{zappia_splattersimulationsinglecell_2017}
-	Luke Zappia, Belinda Phipson, and Alicia Oshlack.
-	\newblock Splatter: {{Simulation}} of {{Single}}-{{Cell RNA Sequencing Data}}.
-	\newblock {\em Genome Biology}, 18:174, 9 2017.
-
-	\bibitem{zeisel_coupledpremrnamrna_2011}
-	Amit Zeisel, Wolfgang~J K{\"o}stler, Natali Molotski, Jonathan~M Tsai, Rita
-	Krauthgamer, Jasmine Jacob-Hirsch, Gideon Rechavi, Yoav Soen, Steffen Jung,
-	Yosef Yarden, and Eytan Domany.
-	\newblock Coupled pre-{{mRNA}} and {{mRNA}} dynamics unveil operational
-	strategies underlying transcriptional responses to stimuli.
-	\newblock {\em Molecular Systems Biology}, 7(1):529, 1 2011.
-
-	\bibitem{zhang_simulatingmultiplefaceted_2019}
-	Xiuwei Zhang, Chenling Xu, and Nir Yosef.
-	\newblock Simulating multiple faceted variability in single cell {{RNA}}
-	sequencing.
-	\newblock {\em Nature Communications}, 10(1):1--16, 6 2019.
+	\newblock \emph{Nature}, 496\penalty0 (7446):\penalty0 461--468, 2013.
+
+	\bibitem[Lizio et~al.(2015)Lizio, Harshbarger, Shimoji, Severin, Kasukawa,
+	Sahin, Abugessaisa, Fukuda, Hori, {Ishikawa-Kato}, Mungall, Arner, Baillie,
+	Bertin, Bono, {de Hoon}, Diehl, Dimont, Freeman, Fujieda, Hide,
+	Kaliyaperumal, Katayama, Lassmann, Meehan, Nishikata, Ono, Rehli, Sandelin,
+	Schultes, `t~Hoen, Tatum, Thompson, Toyoda, Wright, Daub, Itoh, Carninci,
+	Hayashizaki, Forrest, Kawaji, and {the FANTOM
+		consortium}]{lizio_gatewaysfantom5promoter_2015}
+	Marina Lizio, Jayson Harshbarger, Hisashi Shimoji, Jessica Severin, Takeya
+	Kasukawa, Serkan Sahin, Imad Abugessaisa, Shiro Fukuda, Fumi Hori, Sachi
+	{Ishikawa-Kato}, Christopher~J. Mungall, Erik Arner, J.~Kenneth Baillie,
+	Nicolas Bertin, Hidemasa Bono, Michiel {de Hoon}, Alexander~D. Diehl,
+	Emmanuel Dimont, Tom~C. Freeman, Kaori Fujieda, Winston Hide, Rajaram
+	Kaliyaperumal, Toshiaki Katayama, Timo Lassmann, Terrence~F. Meehan, Koro
+	Nishikata, Hiromasa Ono, Michael Rehli, Albin Sandelin, Erik~A. Schultes,
+	Peter~AC `t~Hoen, Zuotian Tatum, Mark Thompson, Tetsuro Toyoda, Derek~W.
+	Wright, Carsten~O. Daub, Masayoshi Itoh, Piero Carninci, Yoshihide
+	Hayashizaki, Alistair~RR Forrest, Hideya Kawaji, and {the FANTOM consortium}.
+	\newblock Gateways to the {{FANTOM5}} promoter level mammalian expression
+	atlas.
+	\newblock \emph{Genome Biology}, 16\penalty0 (1):\penalty0 22, January 2015.
+	\newblock ISSN 1465-6906.
+	\newblock \doi{10.1186/s13059-014-0560-6}.
+
+	\bibitem[Gibson and Bruck(2000)]{gibson_probabilisticmodelprokaryotic_2000}
+	Michael~A. Gibson and Jehoshua Bruck.
+	\newblock A probabilistic model of a prokaryotic gene and its regulation.
+	\newblock \emph{Computational Methods in Molecular Biology: From Genotype to
+		Phenotype, MIT press, Cambridge}, 2000.
+
+	\bibitem[Schwanh{\"a}usser et~al.(2011)Schwanh{\"a}usser, Busse, Li, Dittmar,
+	Schuchhardt, Wolf, Chen, and
+	Selbach]{schwanhausser_globalquantificationmammalian_2011}
+	Bj{\"o}rn Schwanh{\"a}usser, Dorothea Busse, Na~Li, Gunnar Dittmar, Johannes
+	Schuchhardt, Jana Wolf, Wei Chen, and Matthias Selbach.
+	\newblock Global quantification of mammalian gene expression control.
+	\newblock \emph{Nature}, 473\penalty0 (7347):\penalty0 337--342, May 2011.
+	\newblock ISSN 1476-4687.
+	\newblock \doi{10.1038/nature10098}.
+
+	\bibitem[Schilstra and Nehaniv(2008)]{schilstra_biologicgeneexpression_2008}
+	Maria~J. Schilstra and Chrystopher~L. Nehaniv.
+	\newblock Bio-{{Logic}}: {{Gene Expression}} and the {{Laws}} of
+	{{Combinatorial Logic}}.
+	\newblock \emph{Artificial Life}, 14\penalty0 (1):\penalty0 121--133, January
+	2008.
+	\newblock ISSN 1064-5462.
+	\newblock \doi{10.1162/artl.2008.14.1.121}.
+
+	\bibitem[Cannoodt et~al.(2018)Cannoodt, Saelens, Todorov, and
+	Saeys]{cannoodt_singlecellomicsdatasets_2018}
+	Robrecht Cannoodt, Wouter Saelens, Helena Todorov, and Yvan Saeys.
+	\newblock Single-cell -omics datasets containing a trajectory.
+	\newblock \emph{Zenodo}, October 2018.
+	\newblock \doi{10.5281/zenodo.1211532}.
+
+	\bibitem[Soneson and
+	Robinson(2018{\natexlab{b}})]{soneson_unifiedqualityverification_2018}
+	Charlotte Soneson and Mark~D. Robinson.
+	\newblock Towards unified quality verification of synthetic count data with
+	{{countsimQC}}.
+	\newblock \emph{Bioinformatics}, 34\penalty0 (4):\penalty0 691--692, February
+	2018{\natexlab{b}}.
+	\newblock ISSN 1367-4803.
+	\newblock \doi{10.1093/bioinformatics/btx631}.
+
+	\bibitem[Kuijjer et~al.(2015)Kuijjer, Tung, Yuan, Quackenbush, and
+	Glass]{kuijjer_estimatingsamplespecificregulatory_2015}
+	Marieke~Lydia Kuijjer, Matthew Tung, GuoCheng Yuan, John Quackenbush, and
+	Kimberly Glass.
+	\newblock Estimating sample-specific regulatory networks.
+	\newblock pages 1--19, 2015.
+
+	\bibitem[Glass et~al.(2013)Glass, Huttenhower, Quackenbush, and
+	Yuan]{glass_passingmessagesbiological_2013}
+	Kimberly Glass, Curtis Huttenhower, John Quackenbush, and Guo-Cheng Yuan.
+	\newblock Passing {{Messages}} between {{Biological Networks}} to {{Refine
+			Predicted Interactions}}.
+	\newblock \emph{PLOS ONE}, 8\penalty0 (5):\penalty0 e64832, May 2013.
+	\newblock \doi{10.1371/journal.pone.0064832}.
+
+	\bibitem[Kuijjer et~al.(2019{\natexlab{b}})Kuijjer, Hsieh, Quackenbush, and
+	Glass]{kuijjer_lionessrsinglesample_2019}
+	Marieke~L. Kuijjer, Ping-Han Hsieh, John Quackenbush, and Kimberly Glass.
+	\newblock {{lionessR}}: Single sample network inference in {{R}}.
+	\newblock \emph{BMC Cancer}, 19\penalty0 (1):\penalty0 1003, October
+	2019{\natexlab{b}}.
+	\newblock ISSN 1471-2407.
+	\newblock \doi{10.1186/s12885-019-6235-7}.
+
+	\bibitem[Wickham et~al.(2019)Wickham, Averick, Bryan, Chang, McGowan, Fran{\c
+		c}ois, Grolemund, Hayes, Henry, Hester, Kuhn, Pedersen, Miller, Bache,
+	M{\"u}ller, Ooms, Robinson, Seidel, Spinu, Takahashi, Vaughan, Wilke, Woo,
+	and Yutani]{wickham_welcometidyverse_2019}
+	Hadley Wickham, Mara Averick, Jennifer Bryan, Winston Chang, Lucy McGowan,
+	Romain Fran{\c c}ois, Garrett Grolemund, Alex Hayes, Lionel Henry, Jim
+	Hester, Max Kuhn, Thomas Pedersen, Evan Miller, Stephan Bache, Kirill
+	M{\"u}ller, Jeroen Ooms, David Robinson, Dana Seidel, Vitalie Spinu, Kohske
+	Takahashi, Davis Vaughan, Claus Wilke, Kara Woo, and Hiroaki Yutani.
+	\newblock Welcome to the {{Tidyverse}}.
+	\newblock November 2019.
+	\newblock \doi{10.21105/joss.01686}.
+
+	\bibitem[{Van de Sande} et~al.(2020){Van de Sande}, Flerin, Davie,
+	De~Waegeneer, Hulselmans, Aibar, Seurinck, Saelens, Cannoodt, Rouchon,
+	Verbeiren, De~Maeyer, Reumers, Saeys, and
+	Aerts]{vandesande_scalablescenicworkflow_2020}
+	Bram {Van de Sande}, Christopher Flerin, Kristofer Davie, Maxime De~Waegeneer,
+	Gert Hulselmans, Sara Aibar, Ruth Seurinck, Wouter Saelens, Robrecht
+	Cannoodt, Quentin Rouchon, Toni Verbeiren, Dries De~Maeyer, Joke Reumers,
+	Yvan Saeys, and Stein Aerts.
+	\newblock A scalable {{SCENIC}} workflow for single-cell gene regulatory
+	network analysis.
+	\newblock \emph{Nature Protocols}, 15\penalty0 (7):\penalty0 2247--2276, July
+	2020.
+	\newblock ISSN 1750-2799.
+	\newblock \doi{10.1038/s41596-020-0336-2}.
+
+	\bibitem[Marbach et~al.(2009)Marbach, Schaffter, Mattiussi, and
+	Floreano]{marbach_generatingrealisticsilico_2009}
+	Daniel Marbach, Thomas Schaffter, Claudio Mattiussi, and Dario Floreano.
+	\newblock Generating {{Realistic In Silico Gene Networks}} for {{Performance
+			Assessment}} of {{Reverse Engineering Methods}}.
+	\newblock \emph{Journal of Computational Biology}, 16\penalty0 (2):\penalty0
+	229--239, February 2009.
+	\newblock ISSN 1557-8666.
+	\newblock \doi{10.1089/cmb.2008.09TT}.
+
+	\bibitem[Conover and Iman(1979)]{conover_multiplecomparisonsprocedures_1979}
+	William~Jay Conover and Ronald~L Iman.
+	\newblock On multiple-comparisons procedures.
+	\newblock Technical report, {Technical report, Los Alamos Scientific
+		Laboratory}, 1979.
+
+	\bibitem[Patil(2019)]{patil_pairwisecomparisonsmultiplepairwise_2019}
+	Indrajeet Patil.
+	\newblock \emph{{{pairwiseComparisons}}: {{Multiple}} Pairwise Comparison
+		Tests}, 2019.
+
+	\bibitem[Holm(1979)]{holm_simplesequentiallyrejective_1979}
+	Sture Holm.
+	\newblock A simple sequentially rejective multiple test procedure.
+	\newblock \emph{Scandinavian journal of statistics}, pages 65--70, 1979.
 

### 2021-05-20 manuscript/render_manuscript.R

@@ -35,7 +35,7 @@ if (Sys.info()[["user"]] == "rcannood") {
 # render the manuscript
 render("manuscript/manuscript.Rmd")
 
-render("manuscript/supplementary_files.Rmd")
+# render("manuscript/supplementary_files.Rmd")
 
 # system("latexdiff manuscript/v3/render.tex manuscript/render.tex > manuscript/track_changes.tex")
 # file.remove("manuscript/render.pdf")

### 2021-05-21 manuscript/v5/manuscript.tex
update title

@@
 \hypersetup{
-  pdftitle={dyngen is a multi-modal simulator for spearheading future single-cell omics analyses},
+  pdftitle={Spearheading future omics analyses using dyngen, a multi-modal simulator of single cells},
   pdfauthor={Robrecht Cannoodt; Wouter Saelens; Louise Deconinck; Yvan Saeys},
@@
 
-\title{dyngen is a multi-modal simulator for spearheading future single-cell omics analyses}
+\title{Spearheading future omics analyses using dyngen, a multi-modal simulator of single cells}
 \author{Robrecht Cannoodt\textsuperscript{$\dagger{}$,1,2,3} \and Wouter Saelens\textsuperscript{$\dagger{}$,1,2,4} \and Louise Deconinck\textsuperscript{1,2} \and Yvan Saeys\textsuperscript{1,2,*}}
@@
 \hypertarget{abstract}{%
-	\section{Abstract}\label{abstract}}
+	\section*{Abstract}\label{abstract}}
 
@@
 \hypertarget{introduction}{%
-	\section{Introduction}\label{introduction}}
+	\section*{Introduction}\label{introduction}}
 
@@
 \hypertarget{results}{%
-	\section{Results}\label{results}}
+	\section*{Results}\label{results}}
 
@@
 \hypertarget{discussion}{%
-	\section{Discussion}\label{discussion}}
+	\section*{Discussion}\label{discussion}}
 
@@
 \hypertarget{sec:dyngen-methods}{%
-	\section{Methods}\label{sec:dyngen-methods}}
+	\section*{Methods}\label{sec:dyngen-methods}}
 
@@
 \hypertarget{sec:dyngen-modules}{%
-	\subsection{Defining the module network}\label{sec:dyngen-modules}}
+	\subsection*{Defining the module network}\label{sec:dyngen-modules}}
 
@@
 \hypertarget{sec:dyngen-grn}{%
-	\subsection{Generating the gene regulatory
+	\subsection*{Generating the gene regulatory
 		network}\label{sec:dyngen-grn}}
@@
 \hypertarget{sec:dyngen-reactions}{%
-	\subsection{Convert gene regulatory network to a set of
+	\subsection*{Convert gene regulatory network to a set of
 		reactions}\label{sec:dyngen-reactions}}
@@
 \hypertarget{propensity-of-transcription-when-n1}{%
-	\subsubsection{\texorpdfstring{Propensity of transcription when
+	\subsubsection*{\texorpdfstring{Propensity of transcription when
 			\(N=1\)}{Propensity of transcription when N=1}}\label{propensity-of-transcription-when-n1}}
@@
 \hypertarget{propensity-of-transcription-when-n2}{%
-	\subsubsection{\texorpdfstring{Propensity of transcription when
+	\subsubsection*{\texorpdfstring{Propensity of transcription when
 			\(N=2\)}{Propensity of transcription when N=2}}\label{propensity-of-transcription-when-n2}}
@@
 \hypertarget{propensity-of-transcription-for-an-arbitrary-n}{%
-	\subsubsection{\texorpdfstring{Propensity of transcription for an
+	\subsubsection*{\texorpdfstring{Propensity of transcription for an
 			arbitrary
@@
 \hypertarget{propensity-of-transcription-for-a-large-n}{%
-	\subsubsection{\texorpdfstring{Propensity of transcription for a large
+	\subsubsection*{\texorpdfstring{Propensity of transcription for a large
 			\(N\)}{Propensity of transcription for a large N}}\label{propensity-of-transcription-for-a-large-n}}
@@
 \hypertarget{independence-synergism-and-basal-expression}{%
-	\subsubsection{Independence, synergism and basal
+	\subsubsection*{Independence, synergism and basal
 		expression}\label{independence-synergism-and-basal-expression}}
@@
 \hypertarget{sec:dyngen-simcell}{%
-	\subsection{Simulate single cells}\label{sec:dyngen-simcell}}
+	\subsection*{Simulate single cells}\label{sec:dyngen-simcell}}
 
@@
 \hypertarget{sec:dyngen-experiment}{%
-	\subsection{Simulate experiment}\label{sec:dyngen-experiment}}
+	\subsection*{Simulate experiment}\label{sec:dyngen-experiment}}
 
@@
 \hypertarget{sample-cells}{%
-	\subsubsection{Sample cells}\label{sample-cells}}
+	\subsubsection*{Sample cells}\label{sample-cells}}
 
@@
 \hypertarget{sample-molecules}{%
-	\subsubsection{Sample molecules}\label{sample-molecules}}
+	\subsubsection*{Sample molecules}\label{sample-molecules}}
 
@@
 \hypertarget{comparison-between-a-dyngen-and-a-reference-dataset}{%
-	\subsubsection{Comparison between a dyngen and a reference
+	\subsubsection*{Comparison between a dyngen and a reference
 		dataset}\label{comparison-between-a-dyngen-and-a-reference-dataset}}
@@
 \hypertarget{sec:dyngen-batcheffect}{%
-	\subsection{Simulating batch effects}\label{sec:dyngen-batcheffect}}
+	\subsection*{Simulating batch effects}\label{sec:dyngen-batcheffect}}
 
@@
 \hypertarget{sec:dyngen-groundtruth}{%
-	\subsection{Determining the ground-truth
+	\subsection*{Determining the ground-truth
 		trajectory}\label{sec:dyngen-groundtruth}}
@@
 \hypertarget{sec:dyngen-extractgrn}{%
-	\subsection{Determining the cell-specific ground-truth regulatory
+	\subsection*{Determining the cell-specific ground-truth regulatory
 		network}\label{sec:dyngen-extractgrn}}
@@
 \hypertarget{sec:dyngen-nicompare}{%
-	\subsection{Comparison of cell-specific network inference
+	\subsection*{Comparison of cell-specific network inference
 		methods}\label{sec:dyngen-nicompare}}
@@
 \hypertarget{sec:dyngen-velcompare}{%
-	\subsection{Comparison of RNA velocity
+	\subsection*{Comparison of RNA velocity
 		methods}\label{sec:dyngen-velcompare}}
@@
 \hypertarget{sec:dyngen-tacompare}{%
-	\subsection{Comparison of trajectory
+	\subsection*{Comparison of trajectory
 		alignment}\label{sec:dyngen-tacompare}}
@@
 \hypertarget{comparison-of-scalability-and-runtime}{%
-	\subsection{Comparison of scalability and
+	\subsection*{Comparison of scalability and
 		runtime}\label{comparison-of-scalability-and-runtime}}
@@
 \hypertarget{data-availability}{%
-	\section{Data Availability}\label{data-availability}}
+	\section*{Data Availability}\label{data-availability}}
 
@@
 \hypertarget{code-availability}{%
-	\section{Code Availability}\label{code-availability}}
+	\section*{Code Availability}\label{code-availability}}
 
@@
 \hypertarget{acknowledgements}{%
-	\section{Acknowledgements}\label{acknowledgements}}
+	\section*{Acknowledgements}\label{acknowledgements}}
 
@@
 \hypertarget{author-contributions}{%
-	\section{Author contributions}\label{author-contributions}}
+	\section*{Author contributions}\label{author-contributions}}
 
@@
 \hypertarget{competing-interests}{%
-	\section{Competing interests}\label{competing-interests}}
+	\section*{Competing interests}\label{competing-interests}}
 
